### 2015-01-08 material/iti.tex
materials

%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[11pt,twocolumn,english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{hyperref}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{titlesec}
\usepackage{todonotes}
\usepackage{graphicx}

\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection.\ }{0em}{}

%\usepackage{todo}
\pagestyle{empty}

\makeatletter
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\setlength{\topmargin}{-35pt}
\setlength{\textheight}{690pt}
\setlength{\textwidth}{445pt}
\setlength{\columnsep}{18pt}
\setlength{\oddsidemargin}{3pt}
\setlength{\evensidemargin}{3pt}
\setlength{\parindent}{0.5cm}

%
%         New commands and environments
%

\newcommand{\bigast}{\mbox{\small$\ast$}}
\newcommand{\real}{\mbox{\bf R}}
\newcommand{\RR}{\mathbb{R}}

%
%         Data for Title
%
%A.I. Laboratory
%Jo¡ zef Stefan Institute
%Ljubljana, Slovenia
\title{\vspace*{-25pt}\Large\bf Cross-Lingual Document Similarity} %for Large, Multi-lingual Data \\ based on Wikipedia multilingual alignment}
\author{\vspace*{-21pt} \ \\ \large {Andrej Muhi\v c, Jan Rupnik, Primo\v z \v Skraba}  \\
         {\em A.I. Laboratory, Jo\v zef Stefan Institute} \\
         {\em Jamova 39, 10000 Ljubljana, Slovenia} \\
         {\em E-mail: andrej.muhic@ijs.si, jan.rupnik@ijs.si, primoz.skraba@ijs.si}}
\date{ }

\usepackage{babel}
\makeatother
\begin{document}
\thispagestyle{empty}
\maketitle

\noindent \textbf{\emph{\large Abstract.}}
\emph{In this paper we investigated how to compute similarities between documents written in different languages based
on a weekly aligned multi-lingual collection of documents. Computing the cross-lingual similarities is based on an aligned set of basis
vectors obtained by either latent semantic indexing or the $k$-means algorithm on an aligned multi-lingual corpus.
We evaluated the methods on two data sets: Wikipedia and European Parliament Proceedings Parallel Corpus.
%compare two methods for computing language independent representations of documents.
%Such representations allow us to compare documents written in different languages.
%low rank approximation methods for data
%obtained by Wikipedia multilingual alignemnt.
%%We describe the procedure used to obtain Wikipedia multilingual corpus in detail.
%The resulting corpus is only weakly aligned as correspondences are just links and not direct translations.
%Its special properties are discussed. %Troubles with parsing Wikipedia data dumps in special markup language are discussed. Rather than focusing only on bilingual retrieval, we
%We examine the structure in up to 10 chosen languages. The algorithms we
%choose to compare are $k$-means and cross-lingual latent semantic
%indexing(CL-LSI).%, and multi-view cannonical correlation analysis
%%(mCCA). We test these methods on the European Parliament
%%Proceedings Parallel Corpus.
\ }\\ \\
 \emph{} \textbf{\large Keywords. }cross-lingual, similarity, LSI, k-means, Wikipedia, information retrieval

 \section{Introduction}
%While the Internet
%originally had English content almost exclusively, now there is a
%diversity of different languages well beyond bi-lingual into
%truly multi-lingual territory. A notable example of phenomenon is the Wikipedia.
%When extracting topics from documents in multiple languages, we
%would like to find topics that are not only important within a
%language but have a corresponding and equivalent representation
%in the other languages. Machine translation tools are available for well
%established languages but almost no tools exist for some less represented ones.
The globalization process increased the availability of the multi-lingual information sources and the need for automatic cross-lingual processing tools.
The prime example of the process is the Wikipedia - while the majority of the pages in 2001 were written in English, the percentage of English articles dropped to 14\% by 2012.
Our goal is to find language independent representation of the documents without using machine translation tools which may not always be available.
Finding a representation which is valid over multiple languages, we can use well-established machine learning tools designed for monolingual text-mining tasks.
The representations can be interpreted as multi-lingual topics, which can be used as proxies to compute cross-lingual similarities between documents.
%Wikipedia collection is a huge %Document collections are a typically represented as high dimensional data set
%, and one of the prime candidates for lower dimensional representations.

Our work focuses on studying non-probabilistic approaches to analysis of
multi-lingual document collections. The methods we compare
are described in section~\ref{sec:alg}.
%Recently a probabilistic
%approach based on latent Dirichlet allocation was proposed
%\cite{xLDA}, however no implementations were readily available. We plan to implement it and include it in subsequent studies.

The paper is organized as follows: we first introduce the setting
of multilingual data, then describe the algorithms and
datasets. We conclude with the evaluation and discussion.
\begin{comment}
Concepts and... are important across countries. However, much of
what we do is still determined by language. While the Internet
originally had English content almost exclusively, now there is a
diversity of different languages well beyond bi-lingual into
truly multi-lingual territory.

The goal when doing dimensionality reduction on data is to
capture the ``important parts'' of an underlying dataset. These
will preserve the properties of the original high dimensional
data, but faster to deal with algorithmically and highlight the
components (or mixtures thereof) which best explain the data
making it useful for interpretation during analysis as well.

metric analysis

The most common representation of the

metric space....

multilingual is important....

Initially distance of kernel similarity matrices are not only
have an approximate low rank structure, but for this study we
look at matrix with additional structure.

We can view our

 Special structure of multi-lingual data
\end{comment}

\section{Multilingual Data}

Our data is a collection of documents in multiple languages along
with an alignment with correspondences across
languages. Individual documents are represented as vectors by using the standard vector space model where each term corresponds to a word or a phrase in a fixed vocabulary. An aligned corpus is represented by a set of matrices $D^J$ defined as $D^J := [d_1^J , \ldots , d_{m_j}^J ] \in \RR^{n_J \times m_J}$ where $m_J$ is the number of documents in language $J$ and $n_J$ is the dictionary size. Let $N = \sum_{J=1}^M n_J$, where $M$ is the number of languages. Each multi-lingual document of the aligned corpus is formed as a block vector
$d_k = [ d_k^1; \dots; d_k^M] \in \RR^{N}$, and documents are indexed so that $d_k^{J_1}$ and $d_k^{J_2}$ is an aligned document pair for all $k, J_1, J_2$.
%additionally identified by its corresponding language.
%Each language has also its own dictionary independent of other languages.
%The individual
%term-document matrices are grouped according to language giving $
%\{ D_1, D_2, \ldots, D_\ell\}$.

Our main interest are similarities between documents. We can measure the similarities between documents of the same language
 by using the cosine similarity. We note that given a
transformation function between dictionaries, we could compute
similarities between documents in different languages. However,
bilingual-dictionaries are not available for many language pairs in the Wikipedia corpus
and we will rely on document correspondences to compute document similarities across
languages in the present paper. Note that this implies a correspondence between
columns in each $D^K$ for all $K$. Translations between
dictionaries would give us row-based correspondences, but we will
investigate this approach in future work.

Terms in the dictionary are generally not equally important in
determining similarity between documents, so we must preprocess
the documents. We first use term frequency ($TF$), to prune away
infrequent terms. Rather than take a fixed number of top terms in
each document we use an adaptive measure.

Let $f(n)$ be a map
which returns numbers of terms appearing at least $n$ times.  For each language $k$
we find maximal $n$ such that
$\max(10^5, N_k)\leq f(n) \leq 6\cdot 10^5,$ where $N_k$ is number of documents in complete
dictionary of language $k.$ As we do not apply any stemming and lemmatization this ensures that we get well balanced corpus and retain at least of
$10^5$ words for languages with rather small Wikipedias but potentially rich vocabularies.
%find $n$ such that
%\begin{equation}
%\frac{|f(n+1) - f(n)|}{\mbox{original \# of terms}} < 0.001
%\end{equation}
%and used these as the terms for the document.
Once this pruning
step is complete, we further re-weight the remaining terms. A term
weight should correspond to the importance of the term for the
given corpus.  The common weighting scheme is called Term
Frequency Inverse Document Frequency (TFIDF) weighting. An
Inverse Document Frequency (IDF) weight for the dictionary term $j$
is defined as $w_j = \log( DF_j )$, where $DF_j$ is the number of
documents in the corpus which contain term $j$.  A document
TFIDF vector is its original vector multiplied element-wise by the
weights. The $j$-th element of a document vector is given by $
TF_j \log( DF_j )$. Finally, we re-normalize each vector to have
Euclidean norm equal to 1.

\section{Algorithms}
\label{sec:alg}
We compute the low rank approximation of the term-document
matrix using two algorithms: $k$-means\cite{h-ca-75} and
cross-lingual latent semantic indexing(CL-LSI)\cite{cl_lsi}.
%, and
%multi-view canonical correlation analysis (mCCA)\cite{Kettenring}.

The $k$-means algorithm is perhaps the most well-known and used clustering
algorithm. In order to apply the algorithm we first merge
all the term-document matrices into a single matrix by stacking
the individual term-document matrices and discarding nonaligned documents.
\begin{equation}
D_{\mbox{Total}} = \begin{bmatrix}D^T_1 ,D^T_2, \cdots,D^T_\ell\
\end{bmatrix}
^T
\end{equation}
such that the columns respect the alignment of the
documents. Therefore, each document  is represented by a long
vector indexed by the terms in all languages. These vectors
determine the similarity on which the $k$-means is computed.
%In the implementation we never form the matrix $D_{\mbox{Total}}$ explicitly but rather implement
%implicit multiplication as
%$D_{\mbox{Total}} x = \sum_{i=1}^\ell D_i x$ and multiplication
%with transpose as
%$D_{\mbox{Total}}^T y = \sum_{i=1}^\ell D_i^T y_i,$ where
%$y$ is partitioned accordingly to $D_{\mbox{Total}}.$
%That lowers memory requirement and compacts sparse indexing, leading to faster element access.
The algorithm outputs a set of centroids that are then separated to form an aligned basis.
As a final step the the projected documents are computed as the least squares solutions.

The next method is CL-LSI which is a variant of LSI~\cite{lsi}
for more than one language. The method is based on computing the singular value decomposition of $D_{Total}$. Since the matrix can be large we
can use an iterative method like the Lanczos~\cite{matrix_comp}
algorithm with reorthogonalization to find the left singular vectors corresponding to the
largest singular values. It turns out that Lanczos method converges slowly as the ratio of leading singular eigenvalues
is close to one. Moreover Lanczos method is hard to parallelize.
Instead we use randomized version of the singular value decomposition (SVD) described in [1] than can be viewed as block
Lanczos method.
That enables us to use parallelization and can speed up the computation considerably when multiprocessing is available.

At the end we obtain low rank approximation $D_{\mbox{Total}} \approx U_k \Sigma_k V_k^T,$
where $U_k$ are basis vectors of interest and $\Sigma_k$ is truncated diagonal matrix of singular eigenvalues.
Each column $u_i$ of $U_k$ consists of block vectors $u_i = \begin{bmatrix} u^{1^T}_i & \dots & u^{\ell^T}_i \end{bmatrix}^T.$
We do not normalize each block $j$  as this would destroy the low rank approximation.
%Each block $j$ is normalized $u^{j^1}_i = u^{j^1}_i/||u^{j^1}_i||$ to reduce the bias.
We obtain the
aligned reduced basis $U_{\textrm{aligned}} = \begin{bmatrix} {U^1}^T &\dots & {U^\ell}^T \end{bmatrix}^T,$
where $D_j =  U^j \Sigma_k V_k.$
The reduced language free representation for language $j$ and document $d$ is computed as the weighted least square solution $\Sigma_k^{-1} {U^j}^+ d,$
where $+$ denotes the pseudo inverse of a matrix that is is used because columns of $U^j$ do not form an orthogonal basis.
The numerical implementation of the least squares is done by QR algorithm.

%Finally, we test mCCA is specifically designed to consider data
%from multiple sources (in this case languages). Therefore, we do
%not merge the individual term-document matrices into one as in
%the other two methods. For each language (view), we estimate the
%pairwise correlation coefficients, then we try to find vectors
%which maximize the sum of all pairwise correlations over all
%languages. This can be written as the following optimization
%\begin{equation}
%\label{eq:opt1}
%\max_{w_1,\ldots, w_\ell} \sum_{i=1}^\ell\sum_{j=i+1}^\ell  \frac{w_i^T D_i D_j^T w_j}{\sqrt{ w_i^T D_i D_i^T w_i\vphantom{ D_j^T } }\sqrt{ w_j^T D_j D_j^T w_j}}
%\end{equation}
%However, this allows only for a 1-dimensional representation
%(i.e. one vector per language). Since we would like to allow for
%more vectors per language, we denote the double sum in
%Equation~\ref{eq:opt1} by $\mbox{mCCA}(w_1,\ldots,w_\ell)$. If
%$\mbox{corr}$ denotes the correlation, to find $M$ vectors, the
%optimization objective function  becomes
%\begin{equation}
%\max_{\substack{w_i^{(j)}\;;\;i=1,\ldots,\ell\\j=1,\ldots,M }}\sum_{s=1}^{\mbox{M}} \mbox{mCCA}\left(w_1^{(s)},\ldots,w_\ell^{(s)}\right),
%\qquad\mbox{s.t.} \qquad \mbox{corr} \left(w_i^{(s)},w_i^{(t)}\right) = 0
%\quad \forall s\neq t
%\end{equation}
%We require that the set of $M$ vectors we return for each
%language are uncorrelated (so that we do not get copies
%of a vector) which together maximize the pairwise correlation
%between languages.

\section{Data Set}

To investigate the empirical performance of the low rank
approximations we will test the algorithms on a large-scale,
real-world multilingual dataset that we extracted from Wikipedia by using inter-language links as an alignment. This  results in a large number of weakly comparable
documents in more than $200$ languages.
Wikipedia is a large source of multilingual data that is especially important for
 the languages for which no translation tools, multilingual dictionaries as Eurovoc, or strongly aligned multilingual
 corpora as Europarl are available. Documents in different languages are related with so called 'inter-language' links that can be found on the left of
 the Wikipedia page. The Wikipedia is constantly growing. There are currently four Wikipedias with more than $10^6$ articles, $40$ with more than $10^5$ articles, $100$ with more than $10^4$ articles, and $216$ with more than $1000$ articles.
 Wikipedia uses special user-friendly markup language that is very easy to write but very hard to parse. Simplicity of language can cause ambiguities and moreover it is
 constantly changing. For example, separator | can be used in different contexts.

 Wikipedia raw xml dumps of all currently $270$ active editions were downloaded from the Wikipedia dump page\href{http://dumps.wikimedia.org}.
 The xml files are too large to be parsed with DOM like parser that needs to store the whole xml tree in the memory, instead we implemented Sax like parser
 that tries to simulate behavior of Wikipedia official parser and is as simple, fast and error prone as possible.
 We parse all Wikipedia markup but do not extend the templates.
 Each Wikipedia page is embedded in the page tag. First we check if the title of the page consists of any special namespace and do not process such
 pages. Then we check if this is a redirection page and we store the redirect link as inter-language links can point to redirection link also. If nothing of the above applies we extract the text and parse the Wikipedia markup. Currently all the markup is removed.

 We get inter-language link matrix using previously stored redirection links and inter-language links.
 If inter-language link points to the redirection we replace it with the redirection target link. It turns out that we obtain the matrix $M$ that is not symmetric, consequently the underlying graph is not symmetric.
 That means that existence of the inter-language link in one way (i.e. English to German) does not guarantee that there is an inter-language link in the reverse direction (German to English).
 To correct this we transform this matrix to symmetric by computing $M+M^T$ and obtaining an undirected graph. In the rare case that we have multiple links pointing from the document, we pick the first one that we encountered.
 This matrix enables us to build an alignment across all Wikipedia languages.

\section{Evaluation}
We measure the performance of the low rank approximation using
two metrics: mean average precision mate retrieval and
correlation between monolingual similarity profiles between the
query document and its nearest neighbour in the common
representation space. For each evaluation, we randomly select a
training set and test set from the data.

The first evaluation criteria we is the \emph{mean average
  precision mate retrieval score} (AMPMR). This measures the
similarity between the documents and their translations in the
common vector space induced by the latent model. Good models map
the documents close to their translations - indicating that some
language independent (semantic) information was captured.  We
evaluate each latent model (given by projection operators $P_1$
and $P_2$) by considering a pair of aligned test sets $T_1$ and
$T_2$ in languages $L_1$ and $L_2$. We select a query document
$q_1 \in T_1$ and denote the corresponding translated document
$q_2\in T_2$.  We then compute the projections $P_1 q$ and $P_2
T_2$ and rank the elements of $P_2 T_2$ by their similarity to
$P_1 q$ in the projection space (measured by cosine similarity). The mean average precision mate retrieval score is the
inverse of the rank of $P_2 q_2$.

This score does not give us a complete picture. The low rank of a
mate document does not necessarily indicate poor performance if
the documents which outranked it share similar content to the
query. Therefore, we compute an alternative performance measure:
\emph{correlation between monolingual similarity profiles}
(CMSP).
 As before, we choose a query document, target test
corpus, and project them to a common vector space. From the
target corpus, we select the closest document $r \in L_2$ to the
query in the projection space.  We then compute two similarity
profile vectors: $v_1$ contains the monolingual cosine similarity
between $q_1$ and all the documents in $T_1$ and similarly $v_2$
contains the cosine similarities between $r$ and $T_2$. The score
 is the correlation coefficient between $v_1$ and $v_2$.
%Typically that leads to up to $0.05$ lower AMPMR.
Due to space constraints we display only the first evaluation results.

The results for each of the two measures for several fitting
parameters are in Tables~\ref{tb:xlike}, \ref{tb:wiki}, and \ref{tb:wikiK}. We give the result for the pairwise
bilingual retrieval
over all queries in each test set, all pairs of languages and
over ten repetitions of the choice of training and test data.
The Wiki top $9$ experiment was done on $65000$ training documents and retrieval was tested on remaining
$5288$ documents in the following languages: English, German, French, Dutch, Italian, Polish, Spanish, Russian and Japanese. The second set of languages considered was: English, German, Spanish, Chinese, Slovenian, Catalan and Croatian. The second set contains $13000$ aligned documents where $10.000$ were used for training. The second set contains the official languages of the X-LIKE European project and will be denoted by Xlike.  Figures \ref{figure:nbasis} and \ref{figure:ntrain} illustrate how the size of the basis and number of train documents
influences the retrieval.
% The
%experiments are indexed by the number of training samples
%(Ntrain), the number of languages we consider (NViews) and the
%dimensionality of the latent space(Ndims).
%The result is in the final row of Table~\ref{tb:mean}.
\begin{table}
\caption{\label{tb:xlike} {\bf {\sffamily \small{Pairwise retrieval CL-LSI (Xlike)}} }}
\begin{center}
\begin{tabular}{ccccccc}
en & de & es & zh & sl & ca & hr\\
 0 & 0.89 & 0.87 & 0.7 & 0.75 & 0.79 & 0.71\\
 0.89 & 0 & 0.81 & 0.66 & 0.73 & 0.76 & 0.7\\
 0.88 & 0.84 & 0 & 0.65 & 0.7 & 0.79 & 0.69\\
 0.75 & 0.7 & 0.69 & 0 & 0.61 & 0.66 & 0.59\\
 0.76 & 0.75 & 0.72 & 0.59 & 0 & 0.69 & 0.68\\
 0.81 & 0.78 & 0.81 & 0.63 & 0.69 & 0 & 0.67\\
 0.75 & 0.72 & 0.71 & 0.58 & 0.67 & 0.68 & 0
\end{tabular}
\end{center}
\end{table}

\begin{table}
\caption{\label{tb:wiki} {\bf {\sffamily \small{Pairwise retrieval CL-LSI (Wiki)}} }}
\begin{center}
{\scriptsize
%\renewcommand{\arraystretch}{1.8}
\renewcommand\tabcolsep{4pt}
\begin{tabular}{ccccccccc}
en & de & fr & nl & it & pl & es  & ru &ja \\
 0 & 0.91 & 0.9 & 0.88 & 0.89 & 0.86 & 0.9 & 0.86 & 0.88\\
 0.92 & 0 & 0.89 & 0.88 & 0.87 & 0.86 & 0.88 & 0.86 & 0.85\\
 0.9 & 0.9 & 0 & 0.87 & 0.87 & 0.85 & 0.87 & 0.84 & 0.84\\
 0.9 & 0.88 & 0.86 & 0 & 0.84 & 0.84 & 0.86 & 0.82 & 0.83\\
 0.89 & 0.87 & 0.87 & 0.84 & 0 & 0.83 & 0.87 & 0.83 & 0.83\\
 0.87 & 0.87 & 0.85 & 0.84 & 0.83 & 0 & 0.85 & 0.83 & 0.83\\
 0.91 & 0.89 & 0.87 & 0.85 & 0.87 & 0.85 & 0 & 0.85 & 0.85\\
 0.88 & 0.87 & 0.84 & 0.82 & 0.84 & 0.84 & 0.85 & 0 & 0.83\\
 0.89 & 0.87 & 0.85 & 0.83 & 0.83 & 0.83 & 0.85 & 0.84 & 0
\end{tabular}
}
\end{center}
\end{table}

\begin{table}
\caption{\label{tb:wikiK}  {\bf {\sffamily \small{Pairwise retrieval K-means (Wiki)}} }}
\begin{center}
{\scriptsize
%\renewcommand{\arraystretch}{1.8}
\renewcommand\tabcolsep{4pt}
\begin{tabular}{ccccccccc}
0 & 0.75 & 0.74 & 0.68 & 0.71 & 0.65 & 0.75 & 0.65 & 0.66\\
0.73 & 0 & 0.69 & 0.68 & 0.67 & 0.63 & 0.67 & 0.63 & 0.62\\
0.73 & 0.7 & 0 & 0.64 & 0.69 & 0.63 & 0.7 & 0.61 & 0.6\\
0.68 & 0.69 & 0.63 & 0 & 0.61 & 0.59 & 0.63 & 0.58 & 0.57\\
0.73 & 0.7 & 0.7 & 0.63 & 0 & 0.63 & 0.7 & 0.61 & 0.6\\
0.65 & 0.65 & 0.62 & 0.61 & 0.61 & 0 & 0.62 & 0.6 & 0.56\\
0.76 & 0.69 & 0.71 & 0.66 & 0.69 & 0.64 & 0 & 0.62 & 0.61\\
0.65 & 0.65 & 0.62 & 0.6 & 0.59 & 0.61 & 0.61 & 0 & 0.55\\
0.7 & 0.65 & 0.62 & 0.59 & 0.6 & 0.6 & 0.64 & 0.57 & 0
\end{tabular}
}
\end{center}
\end{table}

\begin{figure}
\includegraphics[scale=0.5]{nbasis.png}
\caption{\label{figure:nbasis}{\bf {\sffamily {\small AMPMR and number of base documents (Wiki)}}}}
\end{figure}

\begin{figure}
\includegraphics[scale=0.5]{ntrain.png}
\caption{\label{figure:ntrain}{\bf {\sffamily {\small AMPMR and number of train documents (Xlike)}}}}
\end{figure}

\section{Discussion}

The results illustrate that LSI outperforms
$k$-means in terms of our evaluation criteria. We believe that
this is due to LSI capturing the word co-occurrence
patterns. The performance of LSI increases with number of training documents and size of the basis.
The weighting done with singular eigenvalues typically increases the performance up to $5\%$.
%Normalization of the aligned basis documents decreases applys  additional weighting that destroys the fact that we obtained $k$-rank blockwise approximation and should not be done as it drastically decreases the performance.
The AMPRM score rises with the number of available train documents and size of the basis as
illustrated in Figures \ref{figure:nbasis} and \ref{figure:ntrain}. When forming an aligned corpus, the normalization of document language should be done. Otherwise we always get the rank $k$ approximation that is biased towards the language with documents that have larger number of words. This is undesirable and decreases the quality of the basis.

We also tested CS-LSI on EuroParl corpus and obtained $0.99$ AMPMR score.
That also indicates that performance of CS-LSI increases with the quality of the corpus.
However, additional experimentation is required before we can draw conclusions.

Finally, we note that the results show there is still room for an improvement.
We get decent results for small Wikipedias that could be possibly improved by using hub languages.
For example two Wikipedias can have almost empty intersection but nevertheless sufficiently large intersection
with English Wikipedia. Another option is to allow only partially aligned corpus with empty documents.
Preliminary experiments indicate that this leads to decreased performance.
Future work includes comparing our method with bilingual dictionary based phrase translations for language pairs with such resources.
%A further investigation into low-rank approximations should be done.

%{\color{red}
%Primerjava nacinov izracuna projecirane resitve, ...
%Primerjava tvorbe korpusa,
%je se ves kot za dve strani materiala.
%Tole je samo draft. Posebna zadnja sekcija (Evaluation) se bo se precej spremenila,
%posodobljeni rezultati.
%Najbrz bom dodal se dodaten komentar izracuna projiciranih resitev.
%Ce zrihtamo se MCCA, da dela primerjlivo hitro, lahko dodamo se tudi to.
%Se kaksne dodatne grafe.
%}

%\ \\
% \textbf{\large 1.\ Introduction}\\
%
% Wikipedia is a large source of multilingual data that is specially important for
% the languages that have no good translation tools, multilingual dictionaries as Eurovoc, or strongly aligned multilingual
% corpuses as Europarl available. Documents in different languages are related with so called 'interlanguage' links that can be found on the left of
% the Wikipedia page.
% Wikipedia uses special user-friendly markup language that is very easy to write but very hard to parse. Simplicity of language can cause disambiguation. For example, markup character can have completely meanings in different contexts. % |, [, {
% Moreover parser is not well documented and it does not always obey Wikipedia standards to be able to handle some very common user typos and errors.
% That leads to some unexpected outputs of syntactically incorrect text.%User written?
%
% Wikipedia raw Xml dumps of all currently $270$ active editions were downloaded from \href{http://dumps.wikimedia.org}{Wikipedia dump page}.
% We used only pages-articles.xml files containing current version of all article pages, templates, and other pages, excluding discussion an user home pages.
% Wikipedia articles lay in different namespaces, Main, Media, Special, Talk, File, ...
% In the Xml schema of the dump, each namespace has its own unique key that is independent of the language. This keys are used to indentify links, categories and more across different languages.
% We are interested only in the articles that lay in the main namespaces as they contain interlanguage links.
% The Xml files are too large to be parsed with DOM like parser that needs to store the whole xml tree, instead we implemented Sax like parser
% that tries to simulate behavior of Wikipedia official parser and is as simple, fast and error prone as possible.
% We parse all Wikipedia markup but do not extend the templates.
% Each Wikipedia page is embedded in the page tag. First we check if the title of the page consists of any special namespace and do not process such
% pages, i.e. File:Sun.jng, regular page names do not include colon :. Then we check if this is a redirection page, subtag Redirect exists and we store the redirect link as interlanguage links can point to redirection link also. If nothing of the above applys we extract the text and parse the Wikipedia markup. Currently all the markup is removed and result is the raw text but there is a option to enrich the text with some html tags.
% The largest challenges of parsing are links, tables, and html tags in Wikipedia markup language.
%
% When parsing nested tables we have to be careful about the seperator | that can be nested or non nested.
% Example of nested seperator is
% [[File:Sun.jpg|sun]].
% The table starts with {| and ends with |}, table caption starts with \textbackslash n|+ and ends with newline, table row starts with \textbackslash n|-, table or row heading are seperated by !! and start with \textbackslash n! and scope (row or column) definition. The seperator | indicates the start of
% the new cell, but if there is second non nested |, that indicates parameters, i.e.| bgcolor='red' | Test. Table cells are separated with ||.
% In the code checking for nested | is implemented the cell data is parsed as a normal Wikipedia markup.
% Some heuristic is also implemented to correct some more common user errors, i.e. using || instead of !!, missing closing brackets, prettytable quirks, ...
%
% Another parsing challenge are links. Normal http links are the easiest to parse. They are of the form [http ... ], where optional parameters are separated by | but users also use [[http  .... ]] that is undocumented.
% Internal Wikipedia links are the main focus of our interest.
%
% All such links are of the form illustrated by next two examples, [[File:Sun.jpg| ... | This is the sun]], or [[en:Slovenia]].
% First one is an example of linking to some Wikipedia namespace (File).
% This links are parsed, the link is always before first vertical bar and possible description is after last vertical bar.
% Links can be also nested and are parsed iteratively until there is no more nesting and we can just parse as a  normal text.
% One very common user error is that they forget to close the links, then some heuristic is employed so that we lose as little text as possible.
% For example we detect that we were in the line of the list and stop parsing after the end of the line. The safe assumption is that user forgets to close the first [[ with ending ]] most of the time. We also store this links but they are currently not used.
% One immediate usage of that would be Category links to boost up when categorizing Wikipedia pages across different languages. Or we could use the multimedia links to enrich our structure.
%
%
% The second on is an interlanguage Wikipedia link starting with language code and colon that is followed by the name of the Wikipedia page. The language codes can be found on
% \href{http://meta.wikimedia.org/wiki/List\textunderscore of\textunderscore Wikipedias}{internet}.
% The interlanguage links are stored and then used to build weakly aligned multi-lingual corpus.
%
% We also parse ordinary html tags and some Wikipedia tags in the Wikipedia markup,
% where <, > are written as \&lt and \&gt. For this task we use simple hmtl parser available in glib with added tag checking.
% Almost all valid tags can be found on \href{http://svn.wikimedia.org/viewvc/mediawiki/trunk/phase3/includes/Sanitizer.php?view=markup}{wikimedia.org
% svn}. There are some special actions defined for tags  source, math, ref, gallery, nowiki, pre, the content of such tags is skipped or treated as ordinary text in
% the case of source, nowiki, pre tags.  The Wikipedia markup is constantly changing, the recommended replacement for source tag is syntaxhighlight.
%
% The remaining markup of paragraphs, lines, lists, etc. is not important for us, therefore we skip it.
%
% We know have documents that will be presented as bag of words. We get interlanguage link matrix using previously stored redirection links and interlanguage links.
% If interlanguage link points to the redirection we replace it with the redirection target link. It turns out that we obtain the matrix $M$ that is not symmetric, consequently the graph is not symmetric.
% That means that existence of the interlanguage link in one way (i.e. English to German) does not guarantee that there is an interlanguage link other way around (German to English).
% To correct this we transform this matrix to symmetric $M+M^T$ and obtain undirected graph. In the rare case that we have multiple link pointing to the document, we pick the first one that we encounter.
% This matrix enables us to build an alignment across all Wikipedia languages.
%
% Bag of words processing
% due to the parsing and not using any lemmatisers or multigram models, no list of stopwords, ...
% bag of word matrics are to big to use LSI or K-means or any other projection method.
% Therefore we apply preprocessing to reduce the number of words (features).
%
% Terms in the dictionary are generally not equally important in
%determining similarity between documents, so we must preprocess
%the documents. We compare two options
%
%
%We first use term frequency ($TF$), to prune away
%infrequent terms. Rather than take a fixed number of top terms in
%each document we use an adaptive measure. Let $f(n)$ be a map
%which returns numbers of terms appearing at least $n$ times.  We
%find $n$ such that
%\begin{equation}
%\frac{|f(n+1) - f(n)|}{\mbox{original \# of terms}} < 0.001
%\end{equation}
%and used these as the terms for the document. Once this pruning
%step is complete, we further re-weight the remaining. A term
%weight should correspond to the importance of the term for the
%given corpus.  The common weighting scheme is called Term
%Frequency Inverse Document Frequency (TFIDF) weighting. A
%Inverse Document Frequency (IDF) weight for dictionary term $j$
%is defined as $w_j = \log( DF_j )$ where $DF_j$ is the number of
%documents from the corpora which contain term $j$.  A document
%TFIDF vector is its original vector multiplied element-wise by the
%weights. The $j$-th element of a document vector is given by $
%TF_j \log( DF_j )$. Finally, we re-normalize each vector to have
%Euclidean norm equal to 1.
%
%
% \ \\
% \indent
%
%
%\ \\
% \textbf{\large 2.\ Preliminaries}\\
% \ \\
% \indent
%

%\ \\
%\textbf{\large 6.\ Conclusion}\\
% \ \\
% \indent

\section{Conclusion}

Our work focuses on finding language independent representations of documents written in different languages.
The representations are realized as sets of multi-lingual topics that can be used as proxies to compare documents. We experimented with two methods to compute the representations: CL-LSI and k-means. Our experiments indicate that CL-LSI outperforms $k$-means on information retrieval tasks based on the resulting similarity function. Future work includes incorporating bilingual dictionaries and experimenting with probabilistic approaches to topic modelling.

\section{Learned Concept Vectors}
For the illustrative purposes we include the top 20 vectors we learn, each dimension five most significant words for each of the multilingual topics. We only include 3 top Wikipedia languages here.\\\\
\texttt{
bar text fontsize:xs 8,5 till
he his b film her
cup league club season team
album film band she award
film album d b she
calendar onlyinclude emperor julian year
american b d actor player
bar text fontsize:xs 8,5 till
he his her she duke
album band comune albums province
bar text fontsize:xs 8,5 till
county comune album band italy
bar text fontsize:xs 8,5 till
1r ret 2r she her
1r bc 2r bar round
flag species party she her
flag ret species university calendar
ret bc prix formula racing
flag bc messier star stars
asti turin piedmont alessandria bc
}
\\
\\
\texttt{
bar gemeinde text km2 provinz
er deutscher the us of
fc saison nationalmannschaft er league
the film album oder of
deutscher the film amerikanischer us
kalender jahreswechsel a¨ra zeitrechnung chr
deutscher amerikanischer us politiker komponist
deutscher bar amerikanischer kalender text
er ko¨nig maria war prinzessin
album band stadt single live
county album bar band prozent
county prozent album deutscher maria
deutscher stadt insel fc ko¨nig
open hartplatz atp turnier sand
chr open hartplatz qualifiziert atp
flagge arten weibchen partei ma¨nnchen
bar the flagge text of
formel chr dnf rennen v
flagge chr v flaggen sterne
bar text till shift from
}
\\
\\
\texttt{
hameaux limitrophes administration communes province
the les il a` des
coupe club fc vainqueur buts
the film tv of album
the tv film of album
roi pie´mont naissances julien ii
franc¸ais pie´mont ame´ricain e´crivain pre´sident
lombardie ou hameaux limitrophes italie
roi duc marie prince fils
album groupe l'album ville guitare
comte´ album groupe l'album parti
comte´ album duc roi groupe
ville club roi l'ile fc
tour open finale ext tournoi
tour open finale tournoi ext
drapeau parti espe`ces espe`ce famille
tv the drapeau of e´toiles
formule championnat ferrari gp racing
drapeau e´toiles constellation e´toile jupiter
d'asti monferrato pie´mont milan campanie
}

\section{Acknowledgements}
The authors gratefully acknowledge that the funding for this work was provided by the projecs X-LIKE (ICT-257790-STREP)\cite{xlike},  MultilingualWeb (PSP-2009.5.2 Agr.\# 250500)\cite{mlweb}, PlanetData (ICT-257641-NoE)\cite{pdata} and META-NET (ICT-249119-NoE)\cite{metanet}.

\ \\
 \textbf{\large References} \vspace*{1.3ex} \newcounter{mycount} \begin{list}{[\arabic{mycount}]}
  {\usecounter{mycount}\setlength{\leftmargin}{0.5cm}
   \setlength{\parsep}{0pt}\setlength{\itemsep}{2pt}
   \setlength{\labelsep}{0.1cm}}
   \bibitem{SVDrand} Nathan H., Per-Gunnar M., Joel A.~ T,
    Finding structure with randomness: Probabilistic algorithms for constructing approximate matrix decompositions,
    SIAM Rev. 53, 2 (May 2011), 217-288.
\bibitem{xLDA}
{\em ACL 2010, Proceedings of the 48th Annual Meeting of the Association for
  Computational Linguistics, July 11-16, 2010, Uppsala, Sweden}. The
  Association for Computer Linguistics, 2010.

\bibitem{h-ca-75}
J.~A. Hartigan.
\newblock {\em Clustering Algorithms (Probability $\&$ Mathematical Statistics)}.
\newblock John Wiley \& Sons Inc., 1975.

\bibitem{cl_lsi}
S.T. Dumais, T.A. Letsche, M.L. Littman, and T.K Landauer.
\newblock Automatic cross-language retrieval using latent semantic indexing.
\newblock In {\em AAAI'97 Spring Symposium Series: CrossLanguage Text and
  Speech Retrieval}, pages 18--224, 1997.

\bibitem{Kettenring}
J.~R. Kettenring.
\newblock Canonical analysis of several sets of variables.
\newblock {\em Biometrika}, 58:433--45, 1971.

\bibitem{lsi}
S.~Deerwester, S.~T. Dumais, T.~K. Landauer, G.~W. Furnas, and R.~A. Harshman.
\newblock Indexing by latent semantic analysis.
\newblock {\em Journal of the Society for Information Science}, 41(6):391--407,
  1990.

\bibitem{matrix_comp}
C.F Van~Loan G.H.~Golub.
\newblock {\em Matrix Computations}.
\newblock Johns Hopkins University Press, 1996.

\bibitem{euro_parl_web}
\url{http://www.statmt.org/europarl/index.html}.

\bibitem{xlike}
\url{http://www.xlike.org/}

\bibitem{mlweb}
\url{http://www.multilingualweb.eu/}

\bibitem{pdata}
\url{http://www.planet-data.eu/}

\bibitem{metanet}
\url{http://www.meta-net.eu/}

\bibitem{full_version}
Primoz~\v Skraba, Jan~Rupnik, Andrej~Muhi\v c.
\newblock Low-rank approximations for large, multi-lingual data.
\newblock Avaliable at
  \url{http://ailab.ijs.si/primoz_skraba/papers/nips_full.pdf}.

\end{list}
\end{document}

### 2015-01-08 material/nips_xlite.tex
materials

\documentclass{article} % For LaTeX2e
\usepackage{nips12submit_e,times,url,comment,amsmath,amsfonts}
%\documentstyle[nips10submit_09,times,art10]{article} % For LaTeX 2.09

\usepackage{amsmath}
%\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amssymb}
\newcommand{\RR}{\mathbb{R}}

\title{Cross-Lingual Document Retrieval through Hub Languages}

\begin{comment}
\author{
Jan Rupnik \\
A.I. Laboratory\\
Jo\v zef Stefan Institute\\
Ljubljana, Slovenia\\
\texttt{jan.rupnik@ijs.si} \\
\And
Andrej Muhi\v c\\
A.I. Laboratory\\
Jo\v zef Stefan Institute\\
Ljubljana, Slovenia\\
\texttt{andrej.muhic@ijs.si} \\
\AND
Primo\v z \v Skraba \\
A.I. Laboratory\\
Jo\v zef Stefan Institute\\
Ljubljana, Slovenia\\
\texttt{primoz.skraba@ijs.si} \\
}
\end{comment}
\author{
Jan Rupnik, Andrej Muhi\v c, Primo\v z \v Skraba \\
A.I. Laboratory\\
Jo\v zef Stefan Institute\\
Ljubljana, Slovenia\\
\texttt{(jan.rupnik),(andrej.muhic),(primoz.skraba)@ijs.si} \\
}
% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to \LaTeX{} to determine where to break
% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
% puts 3 of 4 authors names on the first line, and the last on the second
% line, try using \AND instead of \And before the third author name.

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}
\newcommand{\X}{\mathbb{X}}
\newcommand{\Y}{\mathbb{Y}}

\nipsfinalcopy % Uncomment for camera-ready version

\begin{document}

\maketitle

\begin{abstract}
We address the problem of learning similarities between documents written in different languages for language pairs where little or no direct supervision (in the form of a comparable or parallel corpus) is available. To make up for the lack of direct supervision, our approach takes advantage of the fact that they may be linked indirectly by a hub language. That is, correspondences exist between each of the languages and a third,  hub language.
The main goal of our paper is to explore the viability of cross-lingual learning under such conditions.
We propose a method that extracts a set of multilingual topics that facilitate a common representation of documents in different languages. The method is suitable for a comparable multilingual corpus with missing documents.  We evaluate the approach in a truly multi-lingual setting, performing document retrieval across eight Wikipedia languages.
\end{abstract}

\section{Introduction}
%
%Motivation: limited linguistic resources for several pairs, lack of tools to support machine learning tasks.
%Wikipedia comparable corpus can be used to leverage several machine learning tasks. Problem occurs for language pairs when direct document alignments are rare.

Document retrieval is a well-established problem in data
mining. There have been a large number of different approaches
put forward in the literature. In this paper we concentrate on a
specific setting: multi-lingual corpora. As the availability of
multi-lingual documents has exploded in the last few years, there is a pressing
need for automatic cross-lingual processing tools.
%
The prime example is Wikipedia - in 2001 the majority of pages
were written in English, while in 2012, the percentage of English
articles has dropped to 14\%. In this context, we look at how to
find similar documents across languages. In particular, we do not
assume the availability of machine translators, but rather try to
frame the problem such that we can use well-established machine
learning tools designed for monolingual text-mining tasks.

This work represents the continuation of previous work \cite{nips, iti, nips2, sikdd}
where we explored representations of documents which were valid
over multiple languages.  The representations could be interpreted
as multi-lingual topics, which were then used as proxies to compute
cross-lingual similarities between documents.
%
We look at a specific aspect of this problem: the distribution of
articles across languages in Wikipedia is not uniform. While the
percentage of English articles has fallen, English is still the
largest language and one of \emph{hub} languages which not only
have an order of magnitude more articles than other languages,
but also many comparable articles with most of the other languages.

When doing document retrieval, we encounter two quite different situations. If for example, we are looking for a
German article comparable to an English article, there is a large
alignment between the document corpora (given by the intersection
in articles in the two languages) making the problem
well-posed. If, however, we look for a relevant
Slovenian article given a Hindi article, the intersection is small,
making the problem much harder.
%However,
Since almost all languages have a large intersection in
articles with \emph{hub} languages, the question we ask in
this paper is: can we exploit hub languages to perform better
document retrieval between non-hub languages? A positive answer would vastly improve cross-lingual analysis between
less represented languages. In the following section, we
introduce our representation followed by our experiments, which
also shed light on the structural properties of the multilingual
Wikipedia corpus.

\section{Approach}
We begin by introducing some notation. A \emph{multilingual document} $d = (u_1,\ldots u_m)$ is a collection of $m$ documents on the same topic (comparable), where $u_i$ is the document in language $i$ which can be an empty document (missing resource) and $d$ must contain at least two nonempty documents. A comparable corpus $D = {d_1, \ldots, d_N}$ is a collection of multilingual documents. By using the standard vector space model (bag of words) we can represent $D$ as a set of $m$ matrices $X_1,\ldots,X_m$, where $X_i \in \RR^{n_i \times N}$ is the corpus matrix corresponding to the $i$-th language, where $n_i$ is the vocabulary size. Let $X_i^{\ell}$ denote the $\ell$-th column of matrix $X_i$. A \emph{hub language} is a language with a high proportion of non-empty documents in $D$, which in case of Wikipedia is English. Let $a(i,j)$ denote the index set of multilingual documents $d$ that contain non-empty $u_i$ and $u_j$ and $a(i)$ denote the index set of multilingual documents that contain $u_i$.

The goal of our approach is to find a language independent representation of documents by finding a set of mappings $P_1: \RR^{n_1} \rightarrow \RR^k ,\ldots, P_m: \RR^{n_m} \rightarrow \RR^k$ that map documents to a common $k$-dimensional vector space, where similarity between $P_i(x)$ and $P_j(y)$ reflects language independent similarity between $x$ and $y$.

The first step in our method is to project $X_1, \ldots, X_m$ to lower dimensional spaces without destroying the cross-lingual structure. Treating the columns of $X_i$ as observation vectors sampled from an underlying distribution $\mathcal{X}_i \in V_i = \RR^{n_i}$, we can analyze the empirical cross-covariance matrices: $C_{i,j} = \frac{1}{|a(i,j)|-1 }\sum_{\ell \in a(i,j)} (X_i^{\ell} - c_i)\cdot (X_j^{\ell} - c_j)^T$, where $c_i = \frac{1}{a_i} \sum_{\ell \in a(i)}X_i^{\ell}$. By finding low rank approximations of $C_{i,j}$ we can identify the subspaces of $V_i$ and $V_j$ that are relevant for extracting linear patterns between $\mathcal{X}_i$ and $\mathcal{X}_j$. Let $X_1$ represent the hub language corpus matrix. The standard approach to finding the subspaces is to perform the singular value decomposition on the full $N \times N$ covariance matrix composed of blocks $C_{i,j}$. If $|a(i,j)|$ is small for many language pairs (as it is in the case of Wikipedia), then many empirical estimates $C_{i,j}$ are unreliable, which can result in overfitting. For this reason we perform the truncated singular value decomposition on the matrix $C = [C_{1,2}  \cdots  C_{1,m}] \approx U S V^T$, where where $U \in \RR^{n_1 \times k}, S \in \RR^{k \times k}, V \in \RR^{(\sum_{i=2}^m n_i) \times k}$. We split the matrix $V$ vertically in blocks with $n_2, \ldots, n_M$ rows to obtain a set of matrices $V_i$: $V = [V_2^T  \cdots  V_m^T]^T$. Note that columns of $U$ are orthogonal but columns in each $V_i$ are not (columns of V are orthogonal). Let $V_1 := U$. We proceed by reducing the dimensionality of each $X_i$ by setting: $Y_i = V_i^T \cdot X_i$, where $Y_i \in \RR^{k\times N}$. The step is similar to Cross-lingual Latent Semantic Indexing (CL-LSI) \cite{lsi}\cite{cl_lsi} which is less suitable due to a large amount of missing documents. Since singular value decomposition with missing values is a challenging problem and the alternative of replacing missing documents with zero vectors can result in degradation of performance.

The second step involves solving a generalized version of canonical correlation analysis on the matrices $Y_i$ in order to find the mappings $P_i$. The approach is based on the sum of squares of correlations formulation by Kettenring \cite{Kettenring}, where we consider only correlations between pairs $(Y_1, Y_i), i >1$ due to the hub language problem characteristic.
 Let $D_{i,i} \in \RR^{k \times k}$ denote the empirical covariance and $D_{i,j}$ denote the empirical cross-covariance computed based on $Y_i$ and $Y_j$. We solve the following optimization problem:
\begin{equation*}%\label{eq:projZkdimapproxdualqcqp}
\begin{aligned}
& \underset{w_i \in \RR^{k}}{\text{maximize}}
& & \sum_{i = 2}^m  \left(w_1^T D_{1,i} w_i \right)^2
& \text{subject to}
& & w_i^T D_{i,i} w_i = 1, \quad\forall i = 1,\ldots, m.
\end{aligned}
\end{equation*}
By using Lagrangian multiplier techniques (we omit the derivation due to space constraints), the problem can be reformulated as the eigenvalue problem:$\left(\sum_{i=2}^m G_i G_i^T\right) \cdot V = \Lambda \cdot V,$
%\begin{equation*}
% \left(\sum_{i=2}^m G_i G_i^T\right) \cdot V = \Lambda \cdot V,
%\end{equation*}
 where $G_i = H_1^T D_{1,i} H_i$ and $H_i = Chol(D_{i,i})^{-1}$ where $Chol(\cdot)$ is the Cholesky decomposition: $X = Chol(X)^T Chol(X)$. The vectors $w_i$ can be reconstructed from the dominant eigenvector $v$, as: $w_1 = H_1 v$  and $w_i \propto H_i  G_i^T  v$ for $i >1$ ($w_i$ need to be appropriately normalized). Higher dimensional mappings $W_i$ are obtained in a similar way by setting the constraint $W_i^T D_{i,i} W_i = I$, where $I$ is the identity matrix. The constraint forces the columns of $W_i$ to be uncorrelated (orthogonal with respect to covariance matrix). $W_i$ can be extracted from a set of dominant eigenvectors of matrix $V$ according to eigenvalues $\Lambda$.
The technique is related to a generalization of canonical correlation analysis (GCCA) by Carroll\cite{Carroll}, where an unknown group configuration variable is defined and objective is to maximize the sum of squared correlation between the group variable and the others. The problem can be reformulated as an eigenvalue problem. The difference lies in the fact that we set the unknown group configuration variable as the hub language, which simplifies the solution. The complexity of our method is $O(k^3)$, whereas solving the GCCA method scales as $O(N^3)$ where $N$ is the number of samples (see \cite{gifi}). Another issue with GCCA is that it cannot be directly applied to the case of missing documents.

To summarize, we first reduced the dimensionality of our data to $k$-dimensional features and then found a new representation (via linear transformation) that maximizes directions of linear dependence between the languages. The final projections that enable mappings to a common space are defined as: $P_i(x) = W_i^T V_i^T x.$

\section{Experiments}
In this section we evaluate our proposed approach to learn the common document representations. The quality of the learned representation is evaluated on the task of cross-lingual document retrieval on Wikipedia. We will first describe the data and then present the evaluation.
%\subsection{Data}

To investigate the empirical performance of our algorithm %we will test it on a large-scale,
%real-world multilingual dataset that we extracted from Wikipedia  by  using so called 'inter-language' links as an alignment.
%Wikipedia is a large source of multilingual data that enables us to compare documents across the language barrier even for the minority languages that have
% no translation tools, multilingual dictionaries as Eurovoc, or strongly aligned multilingual
% corpora as Europarl available. %'en'    'es'    'ru'    'sl'    'hi'    'war'    'ht'    'pms'
we select a subset of Wikipedia languages containing three major languages, English--\emph{en} (hub language), Spanish--\emph{es}, Russian--\emph{ru}, and five minority (in the sense of Wikipedia sizes) languages, Slovenian--\emph{sl}, Piedmontese--\emph{pms}, Waray-Waray--\emph{war} %(all with about 2 million native speakers)
, Creole--\emph{ht}% (8 million native speakers)
, and Hindi--\emph{hi}.  %(180 million native speakers).
For preprocessing we remove the documents that contain less than 20 different words (stubs) and remove words occur in less than 50 documents as well as the top 100 most frequent words (in each language separately). We represent the documents as normalized TFIDF\cite{Salton88term-weightingapproaches} weighted vectors.
%  These are languages with Wikipedia article sizes comparable to the minority languages.  (since the number of speakers does not directly correlate with the number of Wikipedia articles).
%Some of selected Wikipedia's have  little alignment information available and are therefore good candidates for testing our approach to retrieval using hubs for when no direct alignment information is available. The prime hub candidate is the English language which is well aligned with all other Wikipedia languages, although the alignment quality varies quite a bit.
%Furthermore, we remove the documents that contain less than 20 different words and remove words that are too infrequent as well as the top 100 most frequent words in the vocabularies.
%Furthermore, we call the document consisting of less than 20 different words, a stub. This documents are typically garbage, the titles of the columns in the table, remains of the parsing process, or Wikipedia articles with very little or no information contained in one or two sentences.
% by using the standard vector space model where each term corresponds to a word or a phrase in a fixed vocabulary. An aligned corpus is represented by a set of matrices $D^J$ defined as $D^J := [d_1^J , \ldots , d_{m}^J ] \in \RR^{n_J \times m}$ where $m$ is the number of documents in the aligned corpus and $n_J$ is the dictionary size. Let $N = \sum_{J=1}^M n_J$, where $M$ is the number of languages. The multi-lingual document $k$ of the aligned corpus is formed as a block vector
%$d_k = [ {d_k^1}^T \dots {d_k^M}^T]^T \in \RR^{N}$, and documents are indexed so that $d_k^{J_r}$ and $d_k^{J_s}$ is an aligned document pair for all $k, J_r, J_s,$
%where $1\leq r \leq M,$ $1\leq s \leq M,$ and $r \neq s.$ If the alignment information for the multi-lingual document $k$ in language $j$ is missing, then $d_k^j = 0.$
%additionally identified by its corresponding language.
%Each language has also its own dictionary independent of other languages.
%The individual
%term-document matrices are grouped according to language giving $
%\{ D_1, D_2, \ldots, D_\ell\}$.
%Our main interest are similarities between documents. We can measure the similarities between documents of the same language
% by using the cosine similarity. We note that given a
%transformation function between dictionaries, we could compute
%similarities between documents in different languages. However,
%bilingual-dictionaries are not available for many language pairs in the Wikipedia corpus
%and we will rely on document correspondences to compute document similarities across
%languages in the present paper. Note that this implies a correspondence between
%columns in each $D^J$ for all $J$. Translations between
%dictionaries would give us row-based correspondences, but we will
%investigate this approach in future work.
%As the first step of the processing, we improve the quality of the alignment by removing the stub documents.
%This greatly improves the chance of the successful cross-lingual information retrieval.
%As terms in the dictionary are not equally important for
%determining similarity between documents, we preprocess
%the documents.
%First, we remove the infrequent words and the top 100 most frequent words in every language.
%We form pairs ($\ell$,~$n$) such that a word is infrequent in language $\ell$ if it appears in too few documents $n,$ (\emph{en},~50), (\emph{es},~40), (\emph{ru},~40), (\emph{sl},~10), (\emph{hi},~5), (\emph{war},~5), (\emph{ht},~5), (\emph{pms},~5).
%The criteria for the infrequency was chosen depending on the size of the corpora and in such a way that enough vocabulary was retained for smaller sized Wikipedias.
%At the end we additonally  use term frequency ($TF$), to prune away
%infrequent terms.
%Rather than take a fixed number of top terms in
%each document we use an adaptive measure.
%Let $f(n)$ be a map
%which returns numbers of terms appearing at least $n$ times.  For each language $k$
%we find the maximal $n$ such that
%$\max(10^5, N_J)\leq f(n) \leq 6\cdot 10^5,$ where $N_J$ is number of documents in complete
%dictionary of language $J.$ As we do not apply any stemming and lemmatization this ensures that we get well balanced corpus and retain at least of
%$10^5$ words for languages with rather small Wikipedias but potentially rich vocabularies.
%find $n$ such that
%\begin{equation}
%\frac{|f(n+1) - f(n)|}{\mbox{original \# of terms}} < 0.001
%\end{equation}
%and used these as the terms for the document.
%Once this pruning step is complete, we represent the documents as normalized TFIDF\cite{Salton88term-weightingapproaches} weighted vectors.
%step is complete, we further re-weight the remaining terms.
%Our data is a collection of documents in multiple languages along
%with cross-lingual alignment information. Each document is represented as a bag of words vector consisting of term frequencies (TF).
%A term
%weight should correspond to the importance of the term for the
%given corpus.  The common weighting scheme is called Term
%Frequency Inverse Document Frequency (TFIDF) weighting. An
%Inverse Document Frequency (IDF) weight for the dictionary term $j$
%is defined as $w_j = \log( M /DF_j )$, where $DF_j$ is the number of
%documents in the $M$ document corpus which contain term $j$.  A document
%TFIDF vector is its original vector multiplied element-wise by the
%weights. The $j$-th element of a document vector is given by $
%TF_j \log( M/DF_j )$.
%
%Furthermore, we re-normalize each vector to have
%Euclidean norm equal to 1. This improves the balance in the aligned data and helps to partially resolve the problem when one very short document in one language is
%aligned with large document in other language.
%Finally, we prepare the data for the experiments. We will test retrieval using English language as a hub. Therefore, we keep only the documents that appear in at least one English--(other language) pair.
%metanet
%\subsection{Evaluation}

The evaluation is based on splitting the data into training and test sets (described later). On the training set, we perform the two step procedure to obtain the common document representation as a set of mappings $P_i$. A test set for each language pair, $test_{i,j} = \{(x_\ell,y_\ell) | \ell = 1:n(i,j)\} $, consists of comparable document pairs (linked Wikipedia pages), where $n(i,j)$ is the test set size. We evaluate the representation by measuring mate retrieval quality on the test sets: for each $\ell$, we rank the projected documents $P_j(y_1),\ldots, P_j(y_{n(i,j)})$ according to their similarity with $P_i(x_\ell)$ and compute the rank of the mate document $r(\ell) = rank(P_j(y_\ell))$. The final retrieval score (between -100 and 100) is computed as: $\frac{100}{n(i,j)} \cdot \sum_{\ell = 1}^{n(i,j)} \left( \frac{n(i,j) - r(\ell)}{n(i,j) -1} -0.5\right)$. A score that is less than 0 means that the method performs worse than random retrieval and a score of 100 indicates perfect mate retrieval. The mate retrieval results are included in Table \ref{table:retrieval}.

We observe that the method performs well on all pairs between languages: \emph{en}, \emph{es}, \emph{ru}, \emph{sl}, where at least 50,000 training documents are available. We notice that taking $k = 500$ or $k = 1000$ multilingual topics usually results in similar performance, with some notable exceptions: in the case of (\emph{ht}, \emph{war}) the additional topics result in an increase in performance, as opposed to (\emph{ht},\emph{pms}) where performance drops, which suggests overfitting. The languages where the method performs poorly are \emph{ht} and \emph{war}, which can be explained by the quality of data (see Table \ref{table:rank} and explanation that follows). In case of \emph{pms}, we demonstrate that solid performance can be achieved for language pairs (\emph{pms}, \emph{sl}) and (\emph{pms}, \emph{hi}), where only 2000 training documents are shared between \emph{pms} and \emph{sl} and no training documents are available between \emph{pms} and \emph{hi}. Also observe that in the case of (\emph{pms}, \emph{ht}) the method still obtains a score of 62, even though training set intersection is zero and \emph{ht} data is corrupted, which we will show in the next paragraph.
{
\renewcommand\tabcolsep{3pt}
\begin{table}[h!]
\caption{Pairwise retrieval, 500 topics;1000 topics}\label{table:retrieval}
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
\cline{1-9}
&	en&	es&	ru&	sl&	hi&	war&	ht&	pms\\\cline{1-9}
en&	    &	98~-~98&	95~-~97&	97~-~98&	82~-~84&	76~-~74&	53~-~55&	96~-~97\\
\cline{1-9}
es&	97~-~98&	&	94~-~96&	97~-~98&	85~-~84&	76~-~77&	56~-~57&	96~-~96\\
\cline{1-9}
ru&	96~-~97&	94~-~95&	&	97~-~97&	81~-~82&	73~-~74&	55~-~56&	96~-~96\\
\cline{1-9}
sl&	96~-~97&	95~-~95&	95~-~95&	&	91~-~91&	68~-~68&	59~-~69&	93~-~93\\
\cline{1-9}
hi&	81~-~82&	82~-~81&	80~-~80&	91~-~91&	&	68~-~67&	50~-~55&	87~-~86\\
\cline{1-9}
war&	68~-~63&	71~-~68&	72~-~71&	68~-~68&	66~-~62&	&	28~-~48&	24~-~21\\
\cline{1-9}
ht&	52~-~58&	63~-~66&	66~-~62&	61~-~71&	44~-~55&	16~-~50&	&	62~-~49\\
\cline{1-9}
pms&	95~-~96&	96~-~96&	94~-~94&	93~-~93&	85~-~85&	23~-~26&	66~-~54&	\\
\cline{1-9}
\end{tabular}
\end{center}
\end{table}
}
We now describe the selection of train and test sets. We select the test set documents as all multi-lingual documents with at least one nonempty alignment from the list: (\emph{hi}, \emph{ht}), (\emph{hi}, \emph{pms}), (\emph{war}, \emph{ht}), (\emph{war}, \emph{pms}). This guarantees that we cover all the languages. Moreover this test set is suitable for testing the retrieval thorough the hub as
the chosen pairs have empty alignments.
The remaining documents are used for training.
In Table \ref{table:train_test}, we display the corresponding sizes of training and test documents for each language pair.
The first row represents the size of the training sets used to construct the mappings in low dimensional language independent space using the English--\emph{en} as a hub.
The diagonal elements represent number of the unique training documents and test documents in each language.
{
%\renewcommand{\arraystretch}{0.5}
\renewcommand\tabcolsep{3pt}
\begin{table}[h!]
\caption{Pairwise training:test sizes (in thousands)}
\label{table:train_test}
{
\small
\begin{tabular}{c|c|c|c|c|c|c|c|c|}
&	en&	es&	ru&	sl&	hi&	war&	ht&	pms\\\cline{1-9}
en&	671~-~4.64&	463~-~4.29&	369~-~3.19&	50.3~-~2&	14.4~-~2.76&	8.58~-~2.41&	17~-~2.32&	16.6~-~2.67\\
\cline{2-9}
es&	\multicolumn{1}{c|}{}	&	463~-~4.29&	187~-~2.94&	28.2~-~1.96&	8.72~-~2.48&	6.88~-~2.4&	13.2~-~2&	 13.8~-~2.58\\
\cline{3-9}
ru&	\multicolumn{2}{c|}{}	&	369~-~3.19&	29.6~-~1.92&	9.16~-~2.68&	2.92~-~1.1&	3.23~-~2.2&	10.2~-~1.29\\
\cline{4-9}
sl&	\multicolumn{3}{c|}{}	&	50.3~-~2&	3.83~-~1.65&	1.23~-~0.986&	0.949~-~1.23&	1.85~-~0.988\\
\cline{5-9}
hi&	\multicolumn{4}{c|}{}	&	14.4~-~2.76&	0.579~-~0.76&	0.0~-~2.08&	0.0~-~0.796\\
\cline{6-9}
war&	\multicolumn{5}{c|}{}	&	8.58~-~2.41&	0.043~-~0.534&	0.0~-~1.97\\
\cline{7-9}
ht&	\multicolumn{6}{c|}{}	&	17~-~2.32&	0.0~-~0.355\\
\cline{8-9}
pms&	\multicolumn{7}{c|}{}	&	16.6~-~2.67\\
\cline{9-9}
\end{tabular}
}
\end{table}
}

We further inspect the properties of the training sets by roughly estimating the fraction \texttt{rank(A)/min(size(A))} for each training English matrix and its corresponding mate matrix.
Ideally, these two fractions are approximately the same so  both aligned spaces should have reasonably similar
dimensionality.
We display these numbers as pairs in Table \ref{table:rank}.
\begin{table}[h]
\caption{Dimensionality drift}
\label{table:rank}
\begin{tabular}{|c|c|c|c|c|c|c|}
\cline{1-7}
(en, de)     &   (en, ru)     &   (en, sl)       &     (en, hi)&   (en, war)      &      (en, ht) &   (en, pms)\\
\cline{1-7}
(0.81, 0.89)   &  (0.8, 0.89)  &   (0.98, 0.96)    &    (1, 1)  &   (0.74, 0.56)  &      (1, 0.22)  &   (0.89, 0.38)\\
\cline{1-7}
\end{tabular}
\end{table}
It is clear that in the case of Creole language only at most $22\%$ documents are unique and suitable for the training.
Though we removed the stub documents, many of remaining documents are almost the same, as the quality of some minor Wikipedias is low. This was confirmed for Creole, Waray-Waray, and Piedmontese language
by manual inspection. The low quality documents correspond to templates about the year, person, town, etc. and contain very few unique words.

We also have a problem with the quality of the test data. For example, if we look at test pair (\emph{war}, \emph{ht}) only 386/534 Waray-Waray test documents are unique but on other side
almost all Creole test documents (523/534) are unique. This indicates a poor alignment which leads to poor performance.

\section{Conclusions}
We proposed a method that enables finding common representations for documents in different languages that is tailored
to minority language pairs with limited direct linguistic resources (rare or no comparable document pairs, no dictionary information). We demonstrated that the discovery of cross-lingual mappings is possible even if a pair of languages has no shared linguistic resources, provided that they both share some document correspondences with a hub language.

\section{Acknowledgements}
The authors gratefully acknowledge that the funding for this work was provided by the projects X-LIKE (ICT-257790-STREP)\cite{xlike} and META-NET (ICT-249119-NoE)\cite{metanet}.
%The authors gratefully acknowledge that the funding for this work was provided by the projects X-LIKE (ICT-257790-STREP)\cite{xlike},  MultilingualWeb (PSP-2009.5.2 Agr.\# 250500)\cite{mlweb}, TransLectures (FP7-ICT-2011-7)\cite{translectures}, PlanetData (ICT-257641-NoE)\cite{pdata}, RENDER (ICT-257790-STREP)\cite{render}, and META-NET (ICT-249119-NoE)\cite{metanet}.

\bibliographystyle{unsrt} \bibliography{nips_xlite}
%\bibliographystyle{plain}\bibliography{nips_xlite}

\end{document}

### 2015-01-08 material/pascalCL.tex
materials

\documentclass{article} % For LaTeX2e
\usepackage{nips12submit_e,times,url,comment,amsmath,amsfonts}
%\documentstyle[nips10submit_09,times,art10]{article} % For LaTeX 2.09

%\usepackage{amsthm}
\usepackage{amssymb}
\newcommand{\RR}{\mathbb{R}}

\usepackage{enumitem}
\usepackage{sectsty}
\usepackage{tikz,comment}
\usepackage{stmaryrd}
\usetikzlibrary{%
  matrix,%
  calc,%
  arrows%
}

\title{Cross-Lingual Document Analysis}

\begin{comment}
\author{
Jan Rupnik \\
A.I. Laboratory\\
Jo\v zef Stefan Institute\\
Ljubljana, Slovenia\\
\texttt{jan.rupnik@ijs.si} \\
\And
Andrej Muhi\v c\\
A.I. Laboratory\\
Jo\v zef Stefan Institute\\
Ljubljana, Slovenia\\
\texttt{andrej.muhic@ijs.si} \\
\AND
Primo\v z \v Skraba \\
A.I. Laboratory\\
Jo\v zef Stefan Institute\\
Ljubljana, Slovenia\\
\texttt{primoz.skraba@ijs.si} \\
}
\end{comment}
\author{
Jan Rupnik, Andrej Muhi\v c, Primo\v z \v Skraba \\
A.I. Laboratory\\
Jo\v zef Stefan Institute\\
Ljubljana, Slovenia\\
\texttt{(jan.rupnik),(andrej.muhic),(primoz.skraba)@ijs.si} \\
}

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}
\newcommand{\X}{\mathbb{X}}
\newcommand{\Y}{\mathbb{Y}}

\nipsfinalcopy % Uncomment for camera-ready version

\begin{document}

\maketitle

\begin{abstract}
We present a summary of our work on cross-lingual (CL) document analysis. We focused on learning CL similarity functions and learning language independent document representations. Solutions to either of the two tasks enable us to solve CL text mining tasks (e.g. CL classification, CL retrieval, CL clustering) by using the tools of monolingual text mining. We approached the problems using non-probabilistic methods (typically based on numerical linear algebra) with a special emphasis on scalability. Special attention was devoted to language pairs for which direct training data (translation pairs or comparable documents) was scarce or nonexistent. We showed how to exploit indirect training data through a major (hub) language, such as English.

\end{abstract}

\section{Motivation}
As the availability of multi-lingual content on the web has exploded in the last few years, the need for automatic cross-lingual processing tools has become apparent. The prime example is Wikipedia - in 2001 the majority of pages were written in English, while in 2012, the percentage of English articles has dropped to 14 \%. A standard approach to dealing with multilingual data is by using machine translation techniques. Building good machine translation systems depends on the availability of training data (aligned translations), which is scarce for certain language pairs. However, certain CL tasks admit simpler solutions. In the case of monolingual document retrieval and classification, the widely used vector space model for document representation performs well in practice, even though it discards word order information. Our work relies on the vector space models and is suitable for CL document retrieval, tracking, classification and clustering.

An important aspect of a CL solution is how restrictive its assumptions about the data are. We focus on methods that rely on comparable data (such as the Wikipedia) as opposed to requiring perfect translation pairs and/or bilingual dictionaries. Another important aspect of a CL solution is the computational complexity of training and testing. We focus on methods that can be realized efficiently on a high-end desktop PC.

%applications for a scalable similarity function: tracking information, recommendation
\section{Data representation}

To represent a corpus of documents written in the same language we use the vector space model with tf-idf weights \cite{Salton88term-weightingapproaches}. An aligned multilingual collection of documents is represented as a collection of matrices $X_i \in \RR^{n_i \times s}$ where $i$ ranges over languages, $n_i$ is the size of the vocabulary of the $i$-th language and $\ell$ is the number of documents. The data is aligned over columns ($k$-th column of $X_i$ and $k$-th column of $X_j$ represent comparable documents written in languages $i$ and $j$). Missing documents are represented as zero vectors.

We interpret the collection of aligned documents $\left\{ \left( X_1\left(:,k\right), \ldots, X_m\left(:,k\right) \right) \right\}_{k=1}^{\ell}$ as iid samples obtained from an unknown multivariate distribution over $\RR^{\sum_i{n_i}}.$ Therefore our work is based on analyzing empirical covariance matrices $C_{i,j} = \frac{1}{a_{i,j}}X_i X_j^T$, where $a_{i,j}$ is the number of documents that are nonzero simultaneously in languages $i$ and $j$, where we assume that the data is centered to simplify the presentation.

\section{Learning similarity}
We will now present our formalism for learning a cross-lingual similarity function \cite{nips2}. Using the vector space model to represent two documents in languages $x\in \RR^{n_i}$ and $y \in \RR^{n_j}$, the goal is to define a similarity function for the language pair $(i,j)$: $$s_{i,j}: \RR^{n_i}\times \RR^{n_j} \rightarrow \RR.$$
In our work we focus on finding bilinear similarity functions, i.e. $$s_{i,j}\left(x,y\right) = x^T B_{i,j} y,$$ where $B_{i,j} \in \RR^{n_i \times n_j}$. We considered two main approaches to finding $B$: a regression based approach and a latent space approach.

The \textbf{latent space approach} is defined in terms of two linear operators $P_{i} \in \RR^{k \times n_i}$ and $P_{j} \in \RR^{k \times n_j}$ which together with the standard inner product on the \emph{latent space} $Z = \RR^k$ define the similarity as:
$$s_{i,j}(x,y) := \langle P_i x, P_j y \rangle = x^T P_i^T P_j y.$$
We considered three choices of latent spaces:
\begin{itemize}
\item baseline: $P_i = X_i^T$, $P_j = X_j^T$
\item latent semantic analysis \cite{lsi}
\item canonical correlation analysis \cite{HardoonCCA}
\item in \cite{nips3} we included a generalization of CCA to more than two languages
\item in \cite{iti},\cite{nips} we included $k$-means clustering
\end{itemize}

The \textbf{regression based approach} consists of two ingredients: a linear regression mapping $f : \RR^{n_i} \rightarrow \RR^{n_j}$,$$ f(x) := F\cdot x,$$ where $F \in \RR^{n_j \times n_j}$ and an inner product on the target space $\langle \cdot, \cdot \rangle_D$, which are used to define:
$$s_{i,j}(x,y) := \langle f(x), y \rangle_D = f(x)^T D y = x^T F^T D y.$$ Choices of $D$ included the standard inner product $I$, Mahalanobis inner product $C_{j,j}^{-1}$, which corresponds to whitening, and the covariance matrix $C_{j,j}$ (related to a co-occurrence). We can consider two regression mappings:
\begin{itemize}
\item least squares regression (LSQ), where $F = C_{j,i}C_{i,i}^{-1}$
\item canonical correlation regression\cite{ccar} (CCAR), where $F = C_{j,j}^{-1}C{j,i}$.
\end{itemize}

In \cite{nips2} we elucidated some relationships between the two approaches, for example, CCA is equivalent to the regression based approach with target inner product $D$ as the Mahalanobis inner product. We also conducted an experimental study that compared mono-lingual similarity measures with the proposed cross-lingual similarity measures.

\section{Hub languages}

Continuing the work started in \cite{nips},\cite{sikdd}, we looked at a specific aspect of learning CL similarity functions. The distribution of articles across languages in Wikipedia is not uniform. While the percentage English articles make up as a whole has fallen, in terms of absolute numbers, English is still the largest language. Indeed, there are a number of hub languages which have an order of magnitude more articles than other languages. A large collection (more than $500,000$ articles) of aligned Wikipedia articles is available for the German-English pair, making the learning problem well-posed. If however we consider learning a similarity function for the Slovenian-Hindi pair, there are only a couple of thousand Wikipedia pairs available for learning, which makes the learning problem much harder. However, almost all languages have a large intersection with certain well represented languages, such as English. We refer to English as the hub language. The question we asked in the papers cited is: Can we exploit hub languages to perform better document retrieval between non-hub languages?

In \cite{sikdd} we focused on three languages: Slovenian, Hindi and English (hub). We explored both regression based approaches and common representation based on LSI under various choices of training document collections. We explored several regression scenarios: choosing between LSQ, CCAR and a regression induced by LSI as well as choosing regression sequences (target space and optionally an intermediate space). Here there are five essentially different choices:
\begin{enumerate}
\item $sl\mapsto {\bf hi},$
\item $hi\mapsto {\bf sl}.$
\item $sl\mapsto {\bf en = hub = en} \mapsfrom hi,$
\item $sl\mapsto en \mapsto {\bf hi}$
\item $hi\mapsto en \mapsto {\bf sl}.$
\end{enumerate}
Bold denotes the space where retrieval is done. The first two represent a direct mapping $sl \leftrightarrow hi$, while the remaining methods map to a common hub space (in this case English), with retrieval occurring either in the target language or the hub language. We showed that direct mappings (e.g. $sl\mapsto {\bf hi}$) are more suitable for retrieval when enough data is available, whereas hubs are necessary in the low direct alignment regime. We also noticed that regression based approaches introduced asymmetry in the quality of results as opposed to an approach based on a common representation.

Building on results from \cite{sikdd} we tried incorporating our observations about hub languages in Wikipedia into our methods. We noticed that aligned document sets that do not include English are rare. Taking only cross-covariances that were related to the English language when building common representation models still retains all the vital information. This observation enabled us to formulate a generalization of canonical correlation analysis \cite{Kettenring}, namely the sum of squared correlations, as an eigenvalue problem which can be solved efficiently. We will now present an overview of the approach.

The goal is to find a language independent representation of documents by finding a set of mappings $P_1 ,\ldots, P_m$ that map documents to a common $k$-dimensional vector space. The first step in our method is to project $X_1, \ldots, X_m$ to lower dimensional spaces without destroying the cross-lingual structure. Let $X_1$ denote the hub language. By using the hub language assumption we compute a singular value decomposition of the stacked cross-covariance matrices, related to the hub language, as:
$$[C_{1,2} \cdots C_{1,m}] = U S V^T.$$
We then split the matrix $V$ vertically in blocks with $n_2, \ldots, n_m$ rows to obtain a set of matrices $V_i$: $V = [V_2^T  \cdots  V_m^T]^T$. Note that columns of $U$ are orthogonal but columns in each $V_i$ are not (columns of V are orthogonal). Let $V_1 := U$. We proceed by reducing the dimensionality of each $X_i$ by setting: $Y_i = V_i^T \cdot X_i$, where $Y_i \in \RR^{k\times N}$. The step is similar to Cross-lingual Latent Semantic Indexing (CL-LSI) \cite{lsi},\cite{cl_lsi}, which is less suitable due to a large amount of missing documents. The second step involves solving a generalized version of canonical correlation analysis on the matrices $Y_i$ in order to find the mappings $P_i$. The approach is based on the sum of squares of correlations formulation by Kettenring \cite{Kettenring}, where we consider only correlations between pairs $(Y_1, Y_i), i >1$ due to the hub language problem characteristic.  Let $D_{i,i} \in \RR^{k \times k}$ denote the empirical covariance and $D_{i,j}$ denote the empirical cross-covariance computed based on $Y_i$ and $Y_j$. We solve the following optimization problem:
\begin{equation*}
\begin{aligned}
& \underset{w_i \in \RR^{k}}{\text{maximize}}
& & \sum_{i = 2}^m  \left(w_1^T D_{1,i} w_i \right)^2
& \text{subject to}
& & w_i^T D_{i,i} w_i = 1, \quad\forall i = 1,\ldots, m.
\end{aligned}
\end{equation*}
By using Lagrangian multiplier techniques (we omit the derivation due to space constraints), the problem can be reformulated as an eigenvalue problem.

To summarize, we first reduced the dimensionality of our data to $k$-dimensional features and then found a new representation (via linear transformation) that maximizes directions of linear dependence between the languages. To investigate the empirical performance of the proposed method we selected a subset of Wikipedia languages containing three major languages: English (hub language), Spanish, Russian, and five minority (in the sense of Wikipedia sizes) languages: Slovenian, Piedmontese, Waray-Waray, Creole, and Hindi. We evaluated the method on all pairwise retrieval tasks and found that the results were promising; for example, fair retrieval quality was observed for the Hindi-Piedmontesee pair that contained zero direct alignments.

\textbf{Note on implementation}\\
The preprocessing step requires us to calculate an SVD of a large dense cross-covariance matrix with special structure (fast matrix vector multiplication due to sparse factors). We could use an iterative method like the Lanczos algorithm\cite{golub} with reorthogonalization to find the left singular vectors corresponding to the largest singular values. It turns out that the Lanczos method converges slowly as the ratio of leading singular eigenvalues is close to one. Moreover the Lanczos method is hard to parallelize. Instead we use randomized version of the singular value decomposition (SVD) described in \cite{tropp} than can be viewed as a block Lanczos method. That enables us to use parallelization and can speed up the computation considerably when multiprocessing is available.

\section{Conclusions}

We presented our work \cite{nips},\cite{nips2},\cite{sikdd},\cite{nips3} on cross-lingual similarity measures and language independent embeddings. We presented two main approaches to learn document similarities and conducted several experiments on the task of CL information retrieval.

Finding good embeddings in the presence of missing data presents a challenge - for a learning resource, such as the Wikipedia, the topic coverage may vary greatly across languages.  This means that certain topics might be difficult to represent in all languages. Our work indicates that learning is possible even for language pairs with no direct  alignment information, by leveraging their alignment information with a hub language, such as English.

\section{Acknowledgements}
The authors gratefully acknowledge that the funding for this work was provided by the projects X-LIKE (ICT-257790-STREP)\cite{xlike},  MultilingualWeb (PSP-2009.5.2 Agr.\# 250500)\cite{mlweb}, TransLectures (FP7-ICT-2011-7)\cite{translectures}, PlanetData (ICT-257641-NoE)\cite{pdata}, RENDER (ICT-257790-STREP)\cite{render}, and META-NET (ICT-249119-NoE)\cite{metanet}.

\bibliographystyle{unsrt}
\bibliography{pascalCL}

\end{document}

### 2015-01-08 material/si_kdd2.tex
materials

\documentclass[twocolumn, draft]{article}
\usepackage{enumitem}
\setlist{nolistsep} % or \setlist{noitemsep}
\usepackage{fullpage,times,sectsty,amssymb,amsmath,amsthm}
\usepackage{tikz,comment}
\usepackage{stmaryrd}
\usetikzlibrary{%
  matrix,%
  calc,%
  arrows%
}

\newcommand{\im}{\mathrm{im}}
\title{Multilingual Document Retrieval through Hub Languages}

\author{{ \it Jan Rupnik, Andrej Muhi\v c, Primo\v z \v Skraba}\\
Jo\v zef Stefan Institute\\
Jamova 39\\
Ljubljana, Slovenia\\
}

\date{}
\setlength{\dbltextfloatsep}{-5pt}
\setlength{\dblfloatsep}{-5pt}

\begin{document}
\maketitle
\sectionfont{\sc\large\centering\textbf}
\begin{abstract}
\textbf{In this paper we extend previous work on document retrieval
across multilingual corpora. In this setting it is often assumed
that we have a certain alignment given based on which we can
learn mapping between spaces. In true multilingual corpora
however, we often do not have alignments between all
languages. There are \emph{hub} languages which
have alignments with many other languages. We look at the
effectiveness of leveraging these alignments to learn maps which
may have small or no alignments given. We test several methods
and investigate the performance of various approaches on the
Wikipedia dataset.}

\end{abstract}
\sectionfont{\large\textbf\sc}
\vspace{-0.2cm}
\section{INTRODUCTION}
\vspace{-0.2cm}
Document retrieval is a well-established problem in data
mining. There have been a large number of different approaches
put forward in the literature. In this paper we concentrate on a
specific setting: multi-lingual corpora. As the availability of
multi-lingual documents has exploded in the last few years, the
need for automatic cross-lingual processing tools has become apparent.
%
The prime example is Wikipedia - in 2001 the majority of pages
were written in English, while by 2012 the percentage of English
articles has dropped to 14\%. In this context, we look at how to
find similar documents across languages. In particular, we do not
assume the availability of machine translators, but rather try to
frame the problem such that we can use well-established machine
learning tools designed for monolingual text-mining tasks.

This work represents the continuation of previous work \cite{nips, iti}
where we explored representations of documents which were valid
over multiple languages.  The representations could be interpreted
as multi-lingual topics, which were used as proxies to compute
cross-lingual similarities between documents.
%
We look at a specific aspect of this problem. The
distribution of articles across languages in Wikipedia is not
uniform. While the percentage English articles make up as a whole
has fallen, in terms of absolute numbers, English is still the
largest language. Indeed, there are a number of \emph{hub}
languages which have an order of magnitude more articles than
other languages.

For document retrieval, if for example, we are looking for a
German article comparable to an English article, there is a large
alignment between the document corpora (given by the intersection
in articles in the two languages) making the problem
well-posed. If however we look for a relevant
Slovenian article given a Hindi article, the intersection is small,
making the problem much harder.
However, almost all languages have a large intersection in
articles with the \emph{hub} languages, so the question we ask in
this paper is: can we exploit hub languages to perform better
document retrieval between non-hub languages?

A positive answer would  improve cross-lingual analysis in
particular between less represented languages. In the following
section, we introduce our representation followed by our
experiments, which also shed light on the structural properties
of the multilingual Wikipedia corpus.

\begin{table*}[t]
\begin{center}
\caption{Slovenian-Hindi MAPR retrieval using different maps}
\label{regress_lsq}
{
%\renewcommand{\arraystretch}{1.8}
%\renewcommand\tabcolsep{4pt}
\begin{tabular}{|c|c|c|c|c|c|}\hline &sl$\mapsto${\bf hi}& hi$\mapsto${\bf sl}& sl$\mapsto${\bf en}$\mapsfrom$hi &sl$\mapsto$en$\mapsto${\bf hi} &   hi$\mapsto$en$\mapsto${\bf sl}\\\hline
                  LSQ all     &0.42 & {\bf 0.49} & 0.38 & 0.35 & 0.43\\\hline
                  RCCA all    &{\bf 0.55} & 0.45 & {\bf 0.38} & 0.29 & 0.29\\\hline
                  LSQ common     &0.42 & {\bf 0.48} & 0.47 & 0.42 & {\bf 0.49}\\\hline
                  RCCA common     &{\bf 0.55} & {\bf 0.46} & 0.39 & 0.35 & 0.38\\\hline
                  LSQ empty     &  N/A   & N/A     & 0.27 & 0.28 & {\bf 0.35}\\\hline
                  RCCA empty     &  N/A   & N/A     & {\bf 0.32} & 0.22 & 0.21\\\hline
\end{tabular}\\
\vspace{0.1cm}
\begin{tabular}{|c|c|c|c|c|}\hline
&{\bf sl}$\leftrightarrow${\bf hi} & sl$\mapsto${\bf en}$\mapsfrom$ hi & sl$\mapsto$en$\mapsto${\bf hi} &   hi$\mapsto$en$\mapsto${\bf sl}\\\hline
CL-LSI all&{\bf 0.585} & 0.35 & {\bf 0.56} & 0.54\\\hline
CL-LSI common&{\bf 0.58} & 0.47 & {\bf 0.61} & {\bf 0.61}\\\hline
CL-LSI empty &0 & 0.24 & {\bf 0.48} & 0.46 \\\hline
\end{tabular}
}
\end{center}
\end{table*}
\vspace{-0.2cm}
\section{DATA MODEL}
\vspace{-0.2cm}
The key ingredient to our method is a language independent
representation upon which we can compute similarities. To model
documents we use the standard bag-of-words representation with
TF-IDF (term frequency-inverse document frequency) weighting.
This representation turns each language into a vector space and
cosine similarity induces a metric. From this point on, we
operate on languages as metric spaces denoted generically by $L$.
%metric is a bilinear operator on each vector space
%\begin{equation}
%d: \mathcal{M}_\ell \times \mathcal{M}_\ell \rightarrow \mathbb{R}^{+}
%\end{equation}
Formally, each document is represented as a point in the metric
space. Document comparison can therefore be done by applying the
metric on two point $p,q \in L$.
%Note that although we initially have a very high dimensional
%embedding of each language, the dimensionality of the
%representation must be reduced in order to make computation
%tractable (both from a complexity standpoint and a numerical
%stability standpoint).
%Covariance matrices
This is the starting point of monolingual document retrieval. We
now extend this to the multilingual setting. The benefit of this
linear representation is that maps between languages (metric
spaces) are also linear and we can aim to find a simple (low
dimensional/rank) map between the corpora. As input, we are given
a partial map in the form of point-to-point correspondences and
we must learn the map (usually through some regression).
There are several different formulations of this problem
depending on how we \emph{learn} the map addressed in Section 4.2.

In the multilingual setting, for any two languages the number of
correspondences may be too small to learn effectively. Therefore,
we must go through the hub language (where the number of
correspondence with each language may be large) and effectively
compose the maps. There are numerous ways to do this which we
discuss in the following section.

\begin{comment}
The larger the set of correspondences given as input are, the
easier the learning problem is. In the case of two non-hub
languages, the number of correspondences may be small, so we
frame the problem as a question as to how best to use the
additional information present in the hub languages
\begin{center}
\begin{tikzpicture}
\node (M1) at (0,0) {$\mathcal{M}_\ell$};
\node (M2) at (3,0) {$\mathcal{M}_k$};
\node (Mhub) at (1.5,1.5) {$\mathcal{M}_{hub}$};
\draw[->] (M1) -- node[label={[label distance=0.1cm]120:$f$}] {} (Mhub);
\draw[->] (M2) -- node[label={[label distance=0.1cm]30:$g$}] {} (Mhub);
\draw[<->,dashed] (M1) -- node[label={[label distance=0.1cm]270:$h$}] {} (M2);
\end{tikzpicture}
\end{center}

There are several possiblities on how to find the map
$h$. Without loss of generality, we assume that
$h:\mathcal{M}_\ell \rightarrow \mathcal{M}_k$. Let
$\hat{f},\hat{g}$ and $\hat{h}$ denote the point-to-point
correspondences between the respective spaces. Note that this
implies that $h$ should be an extension of $\hat{h}$ (likewise
for $f$ and $g$).

In the following section we outline some methods for computing
$h$ using information  about $f$ and $g$.
\end{comment}
\begin{comment}
\section{APPROACHES
The most straightforward approach is to augment $\hat{h}$ with
relevant correspondences in $\hat{f}$ and $\hat{g}$. That is we
can define the new set of correspondences by
\begin{equation}
\hat{h} \coprod  (\hat{g}^{-1}\cdot \hat{f})
\end{equation}
Note that $\hat{g}^{-1}$ is not a true inverse but rather is
defined on as the co-image of $\hat{g}$.  That is, the disjoint
union of the initial set of correspondences augments with the
correspondences which can be composed through the hub
language. This implies that we only use part of the hub language,
namely $(\im \hat{f} \cap \im \hat{g})$. This requires all the
correspondences to be consistent (that the above diagram
commutes), which in the case of Wikipedia they are. It remains an
interesting open problem how to deal with inconsistencies.

An alternative approach is to learn $f$ and $g$ independently and
as above compose the linear maps. Therefore define $h = g^{-1}
\cdot f$. Since the maps are generally full rank, it can be
assumed that the inverse exists. Note that in this approach,
composition is simply multiplication (due to the linearity of the
operators).

Anything else?
\end{comment}

%Pairwise retrieval all
\begin{table*}[htbp]
\begin{center}
\caption{CL-LSI MAPR retrieval in common semantic space}
\label{lsi_common}
\begin{tabular}{|c|c|c|c|}\hline
&{\bf sl}& {\bf en}   & {\bf hi}\\\hline
{\bf sl} &0 & 0.77 & {\bf 0.45}\\\hline
{\bf en} &0.73 & 0 & 0.64\\\hline
{\bf hi} & {\bf 0.38} & 0.67 & 0\\\hline
\end{tabular}
\hspace{0.5cm}
\begin{tabular}{|c|c|c|c|}\hline &{\bf sl}& {\bf en}   & {\bf hi}\\\hline{\bf sl} & 0 & 0.81 & {\bf 0.6}\\\hline {\bf en} & 0.77 & 0 & 0.71\\\hline {\bf hi} &{\bf 0.61} & 0.76 & 0 \\\hline\end{tabular}
\hspace{0.5cm}
\begin{tabular}{|c|c|c|c|}\hline&{\bf sl}& {\bf en}   & {\bf hi}\\\hline {\bf sl} & 0 & 0.37 & {\bf 0.22}\\\hline {\bf en} & 0.49 & 0 & 0.36\\\hline {\bf hi} & {\bf 0.11} & 0.29 & 0 \\\hline\end{tabular}\\
\vspace{0.1cm}
\hspace{1cm}All \hspace{3.7cm} Common\hspace{3.7cm} Empty \hspace{0.7cm} \ \\
\end{center}
\end{table*}
%Retrieval using the whole information available
%All
\vspace{-0.2cm}
\section{EXPERIMENTS}
\vspace{-0.2cm}

Experiments were performed using an alignment obtained by
Wikipedia on several languages. Specifically we used Slovenian
(sl, 91272 words), English (en, 344517 words) and Hindi (hi,
72063 words). The markup in brackets denotes language and number
of words in each dictionary.  As a preprocessing step, all stub
documents with fewer than $20$ different words were dropped to
improve the quality of the data.  The remaining alignment
consists of $44426$ Slovenian-English correspondences, $4034$
Slovenian-Hindi correspondences, $14121$ English-Hindi
correspondences and $4017$ joint Slovenian-English-Hindi
correspondences.  After stub removal we keep $614$ of the initial
$1000$ test documents and remove them from the training data.

Retrieval can be done in five essentially different ways using
our metric space approach:
\begin{enumerate}
\item $sl\mapsto {\bf hi},$
\item $hi\mapsto {\bf sl}.$
\item $sl\mapsto {\bf en = hub = en} \mapsfrom
hi,$
\item $sl\mapsto
en \mapsto {\bf hi}$
\item $hi\mapsto en \mapsto {\bf sl}.$
\end{enumerate}
Bold denotes the space where retrieval is done. The first two
represent a direct mapping $sl \leftrightarrow hi$, while the
remaining methods map to a common hub space (in this case
English), with retrieval occurring either in the target language
or the hub language.

To see the amount of information present in the hub languages, we
performed tests on three substantially different datasets
\begin{itemize}
\item \emph{all} -- we use all alignment information available,
\item \emph{common} -- we use only alignment information consistent
through all three languages
\item  \emph{empty} -- we remove all common
alignment to simulate the case where we are forced to use hubs.
\end{itemize}

The evaluation criteria we use is the \emph{mean average
  precision mate retrieval score} (MAPR). This enables us to
compute a similarity between the documents and their translations
in the common vector space induced by the latent model or mapping
in the common space. Good models will map documents close to their
translations -- this indicates that some language independent
(semantic) information was captured.

Let the individual language we are considering be denoted by
$L_1$ and $L_2$.  Each (latent) model is given by projection
operators $P_1$ and $P_2,$ where one can be identity. We evaluate
each model by considering a pair of aligned test sets $X$ and
$Y$ in $L_1$ and $L_2$.
%
We select a query document $x \in X$ and denote the
corresponding translated document $y\in Y$.  We then compute
the projections $P_1 x$ and $P_2 Y$ and rank the elements of
$P_2 Y$ by their similarity to $P_1 x$ in the projection
space (measured by cosine similarity). The mean average precision
mate retrieval score is the inverse of the rank of $P_2 y$. If
only one score is displayed, then this is the average of the
inverse of the rank of $P_1 x$ and the inverse of the rank of
$P_2 y$.
\vspace{-0.2cm}
\subsection{Methods used}
\vspace{-0.2cm}
\begin{table*}[!htbp]
\begin{center}
\caption{\label{lsi_all}CL-LSI MAPR retrieval, full pairwise space}
\begin{tabular}{|c|c|c|c|}\hline & sl & en & hi\\\hline sl & 0 & 0.82 & {\bf 0.57}\\\hline en & 0.77 & 0 & 0.73\\\hline hi & {\bf 0.6} & 0.78 & 0\\\hline\end{tabular}
%Common
\hspace{0.75cm}
\begin{tabular}{|c|c|c|c|} \hline& sl & en & hi \\\hline sl & 0 & 0.81 & {\bf 0.58}\\\hline en & 0.77 & 0 & 0.71\\\hline hi & {\bf 0.58} & 0.77 & 0\\\hline\end{tabular}
%Empty
\hspace{0.75cm}
\begin{tabular}{|c|c|c|c|}\hline & sl & en & hi \\\hline sl & 0 & 0.78 & 0\\\hline en & 0.7 & 0 & 0.71\\\hline hi & 0 & 0.77 & 0 \\\hline\end{tabular}\\
\vspace{0.1cm}
\hspace{1cm}All \hspace{3.7cm} Common\hspace{3.7cm} Empty \hspace{0.7cm} \ \\
\end{center}
\end{table*}

In addition to studying the difference in performance depending
on which space we perform the retrieval in, there is also the
question of how we find the maps.

One approach is to learn the map from the aligned sets $X$ and
$Y$ to use a least squares low rank approach.  That is, we find W
of rank $k$ with which minimizes $\min || W X - Y||_F$ where $k$
is an input parameter. The solution can be obtained using a
truncated SVD of the input $X,$ $W = Y X^+,$ $X = U \Sigma V^*,$
$X_k^+ = V_k \Sigma_k^{-1} U_k,$ where $+$ denotes the pseudo
inverse of matrix $X.$ To speed up the computation, a low rank
approximation of matrix $Y$ is used.  We always use truncated
SVDs of size $1000.$ This approach is denoted as LSQ.

Another method that can be used to relate two aligned sets is
regression canonical correlation analysis (RCCA) that is
described in \cite{ccar}.  Essentially, this results
in the map $q \mapsto (XX')^{-1} X Y' q \approx U_k \Sigma_k^{-1}
V_k^* Y' q$. Note that this must be used on (implicitly) centered
data. Centering explicitly however, is not feasible due to the
large number of words which would result in prohibative RAM
requirements.  Again we use low rank approximations of implicitly
centered $Y$'s to reduce the time complexity and space
complexity.

The third method we use is CL-LSI, latent semantic indexing. It
is described in \cite{cl-lsi}.  This method enables us to compare
documents in the common semantic space.  For the sake of clarity,
we described this method only for two or three aligned document
sets $X_1, X_2, X_3.$ First, we do the SVD decomposition of the
glued aligned documents, then we decouple the basis and map in
the common subspace.
\[
\begin{bmatrix}
X_1\\
X_2
\end{bmatrix} =
\begin{bmatrix}
U_1\\
U_2
\end{bmatrix} \Sigma V^*, \quad
\begin{bmatrix}
X_1\\
Y_1\\
Z_1
\end{bmatrix} =
\begin{bmatrix}
U_1\\
U_2\\
U_3\\
\end{bmatrix} \Sigma V^*,
\]
The map to the common semantic space can be described as $x_i
\mapsto V \Sigma^+ U_i^+ x_i$ for $i=1, 2, 3,$ where we overload
symbols $\Sigma$ and $V$.

To map to English (hub) using the full alignment, we first map
Slovenian $x_1$ to the English word space as $x_1 \mapsto
U^{12}_2 (U^{12}_1)^+ x_1$ and Hindi $x_3$ to English word space
as $x_3 \mapsto U^{23}_1 (U^{23}_2)^+ x_3$. This can be done
efficiently.  Similary we map Slovenian $x_1$ to the
Hindi-English semantic space through English as $x_1 \mapsto
U^{12}_2 (U^{12}_1)^+ x_1 = y_1 \mapsto (U^{23}_1)^+ y_1$ and
Hindi $x_3$ to Hindi-English semantic space $x_3 \mapsto
(U^{23}_2)^+ x_3.$

Mapping through the hub in CL-LSI case enables us to compare
documents in the semantic space which as we will see seems to
boost performance.  In the \emph{all} and
\emph{empty} datasets. we glue documents together all three
languages despite the lack of an alignment to see how performance
degrades in comparison with using the hub.

%Opiši MAPR in povdari, da preslikave z regresijo in LSQ niso simetrične.
%Komentiraj rezultate.
%Zaključek hub lahko izkoristimo, vendar so potrebne še nadaljnje raziskave.
\vspace{-0.2cm}
\subsection{Results}
\vspace{-0.2cm}
As expected, the retrieval is dependent on the mapping used.  It
is important to note the lack of symmetry in retrieval for the
computed RCCA and LSQ mappings.  This is to be expected as we
only use the information about the target(or alternatively the
source) and no common information (as covariance). To illustrate
this, RCCA mapping on the \emph{all} dataset, $hi \mapsto {\bf
  sl},$ results in a retrieval score of $0.55$, but RCCA mapping
other way around on the same dataset, $sl \mapsto {\bf hi},$
results in a retrieval score of $0.45$.

Better performance could be obtained by using canonical
correlation analysis (CCA) although this is a more difficult
computationally and has not yet been tested.  A similar (dual) situation
holds for LSQ mapping where we use only the information about the
source space (rather than the target space).

From Table \ref{regress_lsq} it is not immediately clear which
option of using the hub is the best.  Using the hub does not
improve performance even if we use the whole alignment
information available.  But is clearly the only option if there
is no alignment information through all languages or this
alignment is too small.

The CL-LSI method behaves more consistently and outperforms other
methods.  In this case, the better option is to go through the
hub as we can then compare documents in the semantic space at the
end.  Further tests are needed to better understand this behavior.

In Tables \ref{lsi_common} and \ref{lsi_all} we additionally
display retrieval using mapping in the common semantic space and
using full pairwise alignment, respectively. This gives us an idea
about the quality of each mapping.

\vspace{-0.2cm}
\subsection{Ideal retrieval under  misalignment}
\vspace{-0.2cm}
Consider the \emph{empty} scenario described above: we wish to
compare documents between languages $L_1$ and $L_2$, but we only
have aligned sets for the two languages with a third language
$L_{\mathrm{hub}}$. Our aligned sets $T_1$ and $T_2$
correspond to $ L_1\mapsto L_{\mathrm{hub}}$ and $ L_2\mapsto
L_{\mathrm{hub}}$ respectively.

%: $(A_h, H_a)$ and $b$ and
%language $h$: $(B_h, H_b)$, where $H_A \cap H_B = \emptyset$ (no
%document is shared), where

%$A_h \in \mathbb{R}^{f_a \times
%  n_{a\cap b}}, H_a \in \mathbb{R}^{f_h \times n_{a\cap h}}, H_b
%\in \mathbb{R}^{f_h \times n_{b\cap h}}, B_h \in \mathbb{R}^{f_b
%  \times n_{b\cap h}}$.
We assume that no document is shared between $T_1$ and
$T_2$. Since the alignments are disjoint it may follow that
$rank( T_1 \oplus T_2) = rank( T_1) + rank( T_2)$.
In such cases no nonzero document can be exactly
represented in both bases.
%
%
Let $f_{1}: L_1 \rightarrow L_{\mathrm{hub}}$ and $f_{2}: L_2
\rightarrow L_{\mathrm{hub}}$ represent the regression maps
constructed using the alignment. By using the maps we can cast
the information retrieval problem between documents in languages
$L_1$ and $L_2$ as a monolingual information retrieval problem
$L_{\mathrm{hub}}$. Since $im(f_{1}) \subset span(T_1)$ and
$im(f_{2}) \subset span(T_2)$, all inter-lingual similarities
will be reduced due to the mis-alignment of the spaces
$span(T_1)$ and $span(T_2)$ (rather than all of
$L_{\mathrm{hub}}$). Since the quality of retrieval typically
degrades when no direct alignments are available(see Section
4.2), we investigate what is the best possible retrieval under
the mis-aligned spaces. That is, what is the highest possible
retrieval score on the test set, provided that the images of
$f_{1}, f_{2}$ are restricted to $span(T_1)$ and $span(T_2)$
respectively.
%
As in the previous section, the experiment is based on IR between
Wikipedia pages written in Slovenian ($sl$), Hindi($hi$) and English
($en$). The English language represents the hub language with the
following document matrices: $T_1 \in \mathbb{R}^{406,044
  \times 41,529}, T_2 \in \mathbb{R}^{406,044 \times
  10,331}$, and ${T}_{\mathrm{test}} \in \mathbb{R}^{406,044 \times
  604}$.

$T_{\mathrm{test}}$ is aligned to $A_{\mathrm{test}}$ and $B_{\mathrm{test}}$ and
  $rank([T_1 T_2]) = rank(T_1) + rank(T_2)$. In the ideal case
  $A_{\mathrm{test}}$ and $B_{\mathrm{test}}$ would be mapped to $T_{\mathrm{test}}$ under
  $f_{1}$ and $f_{2}$.

 Let $P_X(\cdot)$ denote the orthogonal projection map to the
 column space of the matrix $X$. Since the images of $f_{1}$ and
 $f_{2}$ are spanned by $T_1$ and $T_2$, the test sets would
 ideally be mapped to $P_{T_1}(T_{\mathrm{test}})$ (ideally
 projected Slovene test documents) and
 $P_{T_2}(T_{\mathrm{test}})$ (ideally projected Hindi test
 documents).

The mean average precision mate retrieval scores we obtain are:
$0.995$ when $A_{test} = \text{query}$, $B_{test} =
\text{target}$ and $0.969$ when $A_{test} = \text{target}$,
$B_{test} = \text{query}$. High retrieval scores indicate that
the space of possible maps admits good quality solutions. This
result shows potential for improving the retrieval quality.
\vspace{-0.2cm}
\section{CONCLUSION}
\vspace{-0.2cm}
The experiments we ran serve two main purposes: the first is an
investigation of the performance of using a hub language to
enable us to compare languages where alignments may not exist
using various different maps and approaches to learning the
maps. The second is a structural study, which illustrates how
much information is present in the maps. In principle this second
part, illustrates that it is possible to find a
linear representation in the hub space which yields very high
retrieval score, even with no overlap in the alignment. This
means that it should be possible to learn maps with very high
retrieval rates.

The first set of experiments show that with the approprate
preprocessing, going throuh hub languages work reasonably
well. However, the lack of symmetry (whereas correspondences are
symmetric) in the maps suggests that this may be degrading
performance. A method which takes both structures into
account may perform better at a higher computational
cost. Further, with no alignment there are distribution issues
which must be addressed (each language has a different
distribution of documents in its monolingual metric space),
suggesting that techniques based on transport distance may prove
effective.
\vspace{-0.2cm}
\section{ACKNOWLEDGEMENTS}
\vspace{-0.2cm}
The authors gratefully acknowledge that the funding for this work was provided by the projects X-LIKE (ICT-257790-STREP), PlanetData (ICT-257641-NoE) and META-NET (ICT-249119-NoE).

\ \\
 \textbf{\large References} \vspace*{1.3ex} \newcounter{mycount} \begin{list}{[\arabic{mycount}]}
  {\usecounter{mycount}\setlength{\leftmargin}{0.5cm}
   \setlength{\parsep}{0pt}\setlength{\itemsep}{2pt}
   \setlength{\labelsep}{0.1cm}}

\bibitem{cl-lsi}
S.~T. Dumais, T.~A. Letsche, M.~L. Littman, and T.~K. Landauer.
\newblock Automatic cross-language retrieval using latent semantic indexing.
\newblock In {\em AAAI'97 Spring Symposium Series: CrossLanguage Text and
  Speech Retrieval}, pages 18--224, 1997.

\bibitem{iti}
Primoz~\v Skraba, Jan~Rupnik, Andrej~Muhi\v c.
\newblock Cross-Lingual Document Similarity.
\newblock In {\em ITI 2012 Information Technology Interfaces}.

\bibitem{ccar}
Jan~Rupnik, Bla\v z Fortuna.
\newblock Regression Canonical Correlation Analysis.
\newblock Learning from Multiple Sources, NIPS Workshop, 13 Dec 2008 Whistler Canada.

\bibitem{nips}
Primoz~\v Skraba, Jan~Rupnik, Andrej~Muhi\v c.
\newblock Low-rank approximations for large, multi-lingual data.
\newblock NIPS Workshop on Low Rank Approximation and Sparse Representation, 2011.
\end{list}

\end{document}

### 2015-01-08 material/similiaritylearning.tex
materials

\documentclass{article} % For LaTeX2e
\usepackage{nips11submit_e,times,url,comment,amsmath,amsfonts}
%\documentstyle[nips10submit_09,times,art10]{article} % For LaTeX 2.09

\usepackage{amsmath}
%\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amssymb}
\newcommand{\RR}{\mathbb{R}}

\title{Spanning Spaces: Learning Cross-Lingual Similarities}

\begin{comment}
\author{
Jan Rupnik \\
A.I. Laboratory\\
Jo\v zef Stefan Institute\\
Ljubljana, Slovenia\\
\texttt{jan.rupnik@ijs.si} \\
\And
Andrej Muhi\v c\\
A.I. Laboratory\\
Jo\v zef Stefan Institute\\
Ljubljana, Slovenia\\
\texttt{andrej.muhic@ijs.si} \\
\AND
Primo\v z \v Skraba \\
A.I. Laboratory\\
Jo\v zef Stefan Institute\\
Ljubljana, Slovenia\\
\texttt{primoz.skraba@ijs.si} \\
}
\end{comment}
\author{
Jan Rupnik, Andrej Muhi\v c, Primo\v z \v Skraba \\
A.I. Laboratory\\
Jo\v zef Stefan Institute\\
Ljubljana, Slovenia\\
\texttt{(jan.rupnik),(andrej.muhic),(primoz.skraba)@ijs.si} \\
}
% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to \LaTeX{} to determine where to break
% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
% puts 3 of 4 authors names on the first line, and the last on the second
% line, try using \AND instead of \And before the third author name.

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}
\newcommand{\X}{\mathbb{X}}
\newcommand{\Y}{\mathbb{Y}}

\nipsfinalcopy % Uncomment for camera-ready version

\begin{document}

\maketitle

\begin{abstract}
In analyzing multilingual text corpora, we have the practical
problem of computing similarities between documents in different
languages. Given two documents in different languages, we use
monolingual similarity to an aligned set to compute a similarity
across languages. We derive several algorithms and show their
relationship the choice of similarity function. We also show
experimental results illustrating the approach.
\end{abstract}

\section{Introduction}
Many machine learning tasks such as clustering, classification,
retrieval, etc. rely on computing a metric to compare (data
points). In our motivating example, multilingual corpora, we are
usually faced with an overwhelming imbalance: some languages have
far more documents than others.  Exacerbating the problem is that
the intersection can be small or non-existent. Therefore, we must
use what alignments exist. In this paper, we derive algorithms
based on 2 general approaches. We find that one approach is
approximated by cross-canonical correlation analysis
(CCA)~\cite{Hotelling} and latent semantic indexing (LSI)\cite{lsi} approximates another.

In the next section we derive algorithms based on analyzing the
covariance structure of the aligned data.  We then conclude by
briefly discussing experimental results on the Europarl
dataset~\cite{europarl} and generalizations and future
directions.

\section{Approach}

We begin with a collection of documents in two
languages\footnote{The extension to multiple language is
 discussed in the Discussion section.}. The documents are
represented via bag-of-words as vector spaces $\X$ and
$\Y$. There are several possible choices of monolingual
similarity functions. We consider functions of the form:
$x'C_{\X\X}^i x$ for $i\in \mathbb{Z}$. For $i=0$, we recover the
squared Euclidean distance and for $i=-1$, we get the Mahalanobis
distance.

As input, we take a choice of $i$ (choice of inner product) along
with an alignment of $\X$ and $\Y$. The problem is to find a
mapping which respects both monolingual similarity and measures
cross-lingual similarity.  There are two pragmatic approaches to finding the mapping. In the
first we reduce the problem to monolingual similarity. We
find some mapping $f: \X \rightarrow \Y$. For a pair of elements
$x\in\X$ and $y\in\Y$, we compute their similarity as the
similarity of $f(x)$ and $y$ in $\Y$. The second approach is to
find an aligned joint basis for $\X$ and $\Y$, which can be
interpreted as mapping both elements to a common space and
computing similarity in this common representation.

For the latter approach, we use LSI to find an aligned basis and
for the former we used least-squares regression. We tested these
approaches for 3 choices of similarity: $i=-1,0,1$, corresponding
to Mahalanobis distance, Euclidean distance, and the metric
weighted by the covariance matrix.

The output of all the algorithms is a linear map (a matrix)
denoted $B$, which computes the cross-lingual similarity as $x^T B
y$ for $x\in\X$ and $y \in \Y$. The simplest choice for $B$ is
$C_{\X\Y}$, the cross-covariance matrix.

{\bf LSI:} To compute $B$, we take the first $d$
eigenvalue-eigenvector pairs $\left(\mu_i, [{u_\X^i}^T,
  {u_\Y^i}^T]^T\right)$ of the cross-covariance matrix, we obtain the approximation of the full
covariance matrix:
\begin{equation}
\begin{bmatrix}
  C_{\X \X} & C_{\X \Y} \\
  C_{\Y \X} & C_{\Y \Y}
\end{bmatrix} \approx \sum_{i = 1}^d \mu_i \begin{bmatrix} u_\X^i \\ u_\Y^i \end{bmatrix} \cdot \begin{bmatrix} {u_\X^i}^T & {u_\Y^i}^T \end{bmatrix}, \qquad \qquad B = \sum_{i = 1}^d \mu_i u_\X^i {u_\Y^i}^T\approx C_{\X \Y}
\end{equation}
The definition of $B$ comes from comparing the upper-right
block. This also shows that LSI approximates the simplest choice
of $B$. It is however far more computationally efficient because
it first finds a low-dimensional representation.

{\bf Regression:} To find $B$ using regression, we compute the
optimal regression matrix as $ C_{\Y \X} (C_{\X \X})^{-1}$. This
corresponds to the mapping $f$ mentioned above.  The inner
product on the space $\Y$ induces $B$ for computing the
similarities between elements in $\X$ and $\Y$: for $B$ it holds
that $x^T B y = \langle f (x), y \rangle$. Therefore $B$ depends
on the metric on $\Y$. For the inner product corresponding to
$C_{\Y\Y}$, $B$ can be derived as $C_{\X \X}^{-1}C_{\X
  \Y}C_{\Y \Y}$.

Note that in the case of regression, we have a choice of which
direction we map: $\X\rightarrow\Y$ or $\Y\rightarrow\X$ except
for Mahalanobis metric where the formula for $B$ is
symmetric. Due to space constraints we only note here that CCA approximates the least-squares
regression for this metric just as LSI approximates the
cross-covariance matrix.

\section{Experiments}
To evaluate a cross-similarity matrix performs, we used
monolingual similarity matrices for supervision. All experiments
are based on an aligned bilingual corpus of English and German
documents and all computations were performed by selecting an
aligned subset of $5000$ documents $T_X, T_Y$ for learning $B$
and an aligned set of $500$ documents $R_X, R_Y$
for testing. For the monolingual similarities $\langle \cdot, \cdot
\rangle_X$ and $\langle \cdot, \cdot \rangle_Y$ we compute
monolingual similarity scores: $S_X := \langle R_X, R_X
\rangle_X$ and $S_Y:= \langle R_Y, R_Y \rangle_Y$ and compare the
similarities to $S_{XY} := R_X^T B R_Y$. The measure chosen was
the average correlation coefficient between rows of $S_{XY}$ and
rows of $S_Y$ and the average correlation coefficient between the
columns of $S_{XY}$ and rows of $S_Y$.

We first observe that metrics $C_{X X}^i$ with higher exponent
$i$ are easier to learn (the average correlation coefficients
between the monolingual and cross-lingual similarity profiles are
around $0.9$. We also observe that $i=-1$ is the only approach
that performs relatively well in the inner product space
$C_{XX}^{-1}$ when compared to the other approaches. This is
predicted by the theory since the cross-view similarity matrix
was derived for that objective. We also note that the ad-hoc
approach of $B:= C_{XY}$ and its LSI-based approximation perform
similarly with only $5$ basis vectors used by LSI, which is quite
surprising.

\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
&\multicolumn{2}{|c|}{}&\multicolumn{5}{|c|}{Regression} \\
\hline
&       $C_{xy}$	&LSI & 	$i=-1$& $i=0$(xy)& $i=0$(yx) & $i=1$(xy) & $i=1$(yx)\\
\hline
$C_{xx}^{-1}$ &	0.06&	0.05&	0.59&	0.22&	0.21&	0.04&	0.04\\
\hline
$C_{yy}^{-1}$ &	0.06&	0.05&	0.58&	0.22&	0.21&	0.04&	0.04\\\hline
$I_{xx}$     &	0.29&	0.28&	0.70&	0.45&	0.45&	0.27&	0.27\\\hline
$I_{yy}$     &   0.27&	0.26&	0.66&	0.43&	0.42&	0.25&	0.25\\\hline
$C_{xx}$     &	0.94&	0.93&	0.34&	0.87&	0.88&	0.94&	0.94\\\hline
$C_{yy}$     &	0.94&	0.93&	0.33	&0.88&	0.88&	0.95&	0.94\\\hline
\end{tabular}
\end{center}

\bibliographystyle{unsrt} \bibliography{nips}

\end{document}

From now on we assume a fixed $i$ and denote the Hilbert space
corresponding the the spaces and choice of inner product by $X =
\left(\RR^n, \langle\cdot, \cdot\rangle_X\right)$ and $Y =
\left(\RR^m, \langle\cdot, \cdot\rangle_Y\right)$. Let $d_X :(X,
X) \rightarrow \RR$ and $d_Y : (Y,Y) \rightarrow \RR$ be the
induced metrics on $\RR^n$ and $\RR^m$ respectively. Let $X
\oplus Y$ denote the direct sum of Hilbert spaces with the inner
product: $\langle (x_1, y_1), (x_2, y_2) \rangle_{X \oplus Y} =
\langle x_1,x_2 \rangle_X + \langle y_1, y_2 \rangle_Y$. Let $A
\subset X \oplus Y$ be a Riemannian manifold with metric tensor
$g$. Let $\pi_X: X \oplus Y \rightarrow X$ and $\pi_Y: X \oplus Y
\rightarrow Y$ denote projection operators to spaces $X$ and
$Y$. Let $A_X := \pi_X(A)$ and $A_Y := \pi_X(A)$. Let $\pi_{A_X}
: X \rightarrow A_X$ and $\pi_{A_Y} : Y \rightarrow A_Y$ denote
projection operators.

\section{Algorithms}

\section{Partially Aligned Metric Spaces}

We assume our spaces (document corpora divided by language) are
mapped from some underlying, space(language)-independent latent
space denoted $\L$. These mappings differ depending on
language. For simplicity, we concentrate on the case of two
spaces: $\X$ and $\Y$. Therefore, we can write $\X
\overleftarrow{f_x} \L \overrightarrow{f_y} \Y$.  There are
numerous metrics we can choose within a single space: this
depends on what similiarity measure we choose between documents.

The 3 methods we concentrate on are kernel versions of
regression, LSI and CCA~\cite{shawe-taylor04kernel}.  By treating
the aligned documents as mappings of common samples, we can
conclude that there exists a bilinear operator for the metrics
$M_\X \times M_\Y\rightarrow \RR$ and we base this mapping on the
based on covariance matrices. We show how the different methods
correspond with different choices of metrics on the space.

The general statement of our problem is: Find an near-isometric
mapping in between the similiarity spaces for all vectors $x\in
\X$ and $y\in Y$. By working in the covariance space, this
problem reduces to finding transformations between basis vectors
of the two spaces. In all cases, we us the correspondence
information to find this transformation. We limit ourselves to
matrix operator $B(x,y) =  y' B x $

{\bf Regression:} In this case we have the choice of minimizing
$Bx-y$ such that $x$ and $y$ are in correspondence. We can
optimze its adjoint($\Y\rightarrow\X$). In terms of the
covariance matrices we can find $f_{\X\rightarrow\Y}(x) =
C_{\X\X} C_{\Y\Y}^-1 $.  These are obtained by by standard
multivariate regression analysis where the first gives the
optimal regression matrix based on the standard inner
product. Alternatively, if we assume an inner product weighted by
the covariance matrix, the optimal mapping becomes just the
cross-correlation matrix.

\textcolor{red}{DRAWBACKS}

{\bf CCA:}

{\bf LSI:}

In the case of
aligned samples, we have well-defined covariance matrices
$C_{\X\X}$ and $C_{\Y\Y}$.

Notation

Let $L_1 = \RR^{n_1},\ldots, L_m = \RR^{n_m}$ be $m$ vector
spaces. Let $X_1 \in L_1, \ldots, X_m \in L_m$ be continuous
random vectors with defined covariance matrices $C_{X_i X_i} :=
E\big(X_i - E(X_i)(X_i - E(X_i))^T\big)$. In this paper we
consider several possible inner product spaces for $L_1,\ldots,
L_m$ that are induced by powers of covariance matrices. Let
$H_i^p := (L_i, \langle \cdot, \cdot \rangle_{(C_{X_i X_i})^p})$
denote the Hilbert space corresponding to space $L_i$ equipped
with the inner product induced by the matrix $(C_{X_i X_i})^p$.

Assumption

Random vectors $X_i$ are different representations of some
underlying random vector $Z \in R^\ell$: $X_i := f_i(Z), \ldots,
X_m := f_m(Z)$, where each $f_i :\RR^\ell \rightarrow \RR^{n_i}$
is a mapping. For this reason, we will refer to spaces $L_1,
\ldots, L_m$ as different views of some latent objects.

Notation

A dataset $D: = [D_1^T,\ldots, D_m^T]^T $ is an aligned data set
if each $D_i$ can be expressed in the following way: $D_i \in
\RR^{n_i \times k} = f_i([z_1,\ldots, z_k])$ and $[z_1,\ldots,
  z_k] \in \RR^{\ell\times k}$ is a sample of $k$ observation
vectors drawn independently according to distribution of $Z$.

2-view formulation of metric learning

Let $(S_1,S_2)$ denote an aligned sample of $k$ pairs of vectors
($S_1 \in \RR^{n_1 \times k}, S_2 \in \RR^{n_2 \times
  k}$). Assign to $S_1$ a matrix of similarities $M_1^p := S_1^T
C_{X_1 X_1}^p S_1$ and matrix $M_2^p := S_2^T C_{X_2 X_2}^p
S_2$. These two matrices represent similarities computed in
$H_1^p$ and $H_2^p$. Our goal is to find a matrix $B$ that can be
used for computing similarities between vectors in $L_1$ and
vectors in $L_2$ that is in "concordance" with similarities in
$H_1^p$ and $H_2^p$. Define $B_{1,2}^p := S_1^T B S_2$. A good
matrix $B$ for sample $S$ is the one which makes $B_{1,2}^p$
close to $M_1^p$ and $M_2^p$. There are several possible measures
of matrix closeness, a typical choice is the Frobenious
norm. Obviously this is not always achievable, but if the random
variables $X_1$ and $X_2$ (they are involved in drawing samples
$S_1$ and $S_2$) share enough of mutual information, it is
possible, for example in the extreme case where $X_1 = X_2$. This
leads to the following optimization problem. We will refer to the
bilinear mapping $B$ as the cross-view similarity measure.

Optimization problem:

\begin{equation}\label{eq:opt1}
\begin{aligned}
& \underset{B \in \RR^{n_1 \times n_2}}{\text{minimize}}
& & E_{(S_1,S_2)} ||B_{1,2}^p - M_1^p||_F^2 + ||B_{1,2}^p - M_2^p||_F^2 \\
\end{aligned}
\end{equation}

Remark:

$S_1$ and $S_2$ are random matrices (i.i.d. samples of random
variables $X_1$ and $X_2$) and they are sampled in a way that
they form an aligned data set.

Questions:

Can the problem be solved analytically?

Does to choice of the parameter $k$ (sample set size) influence the optimal matrix $B$.

Remark:

In practice we only have a single aligned sample, perhaps this
simplifies the optimization problem (drop the expectation).

Current candidate solutions

For $H_1^0, H_2^0$ we propose: $B := C_{X_1 X_2} C_{X_2 X_2}^{-1}$ or $B := C_{X_1 X_1}^{-1} C_{X_1 X_2}$.
For $H_1^1, H_2^1$ we propose: $B := C_{X_1 X_2}$.
For $H_1^2, H_2^2$ we propose: $B := C_{X_1 X_2} C_{X_2 X_2}$ or $B := C_{X_1 X_1} C_{X_1 X_2}$.

Connection to least squares regression

The candidate solutions can be derived easily if we consider
multivariate regression analysis. The aligned sample $S_1 \in
\RR^{n_1 \times k}$ and $S_2 \in \RR^{n_2 \times k}$ can be
interpreted as a set of input output vector pairs. Using the
method of multivariate least squares regression, an optimal
regression matrix from $L_1$ to $L_2$ can be computed as:
$\widetilde{W} := S_2 S_1^T (S_1 S_1^T)^{-1}$. When $k
\rightarrow \infty$ the expression for $\widetilde{W}$ converges
to $W:= C_{X_2 X_1} (C_{X_1 X_1})^{-1}$. The inner product on the
space $L_2$ induces the bilinear mapping $B$ for computing the
similarities between elements in $L_1$ and $L_2$: the matrix $B$
is the matrix for which it holds $x^T B y = \langle W x, y
\rangle$. For the space $H_2^0$ ($L_2$ equipped with the standard
inner product) we obtain the candidate $B = C_{X_1 X_1}^{-1}
C_{X_1 X_2}$, whereas $H_2^1$ induces $B = C_{X_1 X_1}^{-1}C_{X_1 X_2}C_{X_2 X_2}$. If the
direction of the regression is reversed (consider $S_2$ as the
input vectors and $S_1$ as the output vectors), then the inner
product structure of $L_1$ is the one that induces $B$. This way
we obtain the candidate corresponding to $H_1^0$: $B := C_{X_1
  X_2} C_{X_2 X_2}^{-1} $. The possible problem with the
regression approach to similarity induction is that only one
inner product structure is involved in computing $B$, whereas the
search for $B$ in \ref{eq:opt1} considers both inner product
spaces.

Three views

Assume that we have three views $X_1,X_2$ and $X_3$. If $C_{X_1 X_3}$ is not available (too few data points to estimate) we propose:
$B := C_{X_1 X_2} C_{X_2 X_3}$ which fits inner product spaces $H_1^2$ and $H_3^2$.

Long chains are inappropriate practically because they correspond to inner product spaces induced by variance matrices exponentiated to high powers.
Rank of such matrices drops quickly (it approaches the dimension of the dominant eigen-space of the variance matrix) with the exponent and they typically become unsuitable for practical applications.

CCA, SVD and covariances

CCA and SVD represent two forms of analysis of data based on the covariance structures. They can both be used to extract basis vectors for each view that can be used for computing the bilinear mapping that enables computations of cross-view similarity mappings.

Assume that $m=2$.

Canonical correlation analysis can be formulated as following problem:

\begin{equation*}
\begin{aligned}
& \underset{w_1 \in L_1, w_2 \in L_2}{\text{maximize}}
& &  \frac{w_1' C_{X_1 X_2} w_2}{\sqrt{w_1' C_{X_1 X_1} w_1} \sqrt{w_2' C_{X_2 X_2} w_2}}.
\end{aligned}
\end{equation*}

This can be reformulated as:
\begin{equation}\label{eq:cca}
\begin{aligned}
& \underset{w_1 \in L_1, w_2 \in L_2}{\text{maximize}}
& &  w_1' C_{X_1 X_2} w_2\\
& \text{subject to}
& &w_1' C_{X_1 X_1} w_1 = 1
& &w_2' C_{X_2 X_2} w_2 = 1.
\end{aligned}
\end{equation}

Using lagrangian multipliers technique yields a generalized eigenvalue problem:
$$
\begin{bmatrix}
  0 & C_{X_1 X_2} \\
  C_{X_2 X_1} & 0 \\
\end{bmatrix} \cdot
\begin{bmatrix}
  w_1 \\
  w_2 \\
\end{bmatrix} =
\lambda
\begin{bmatrix}
  C_{X_1 X_1} &0  \\
  0 & C_{X_2 X_2} \\
\end{bmatrix} \cdot
\begin{bmatrix}
  w_1 \\
  w_2 \\
\end{bmatrix}
$$

By inverting the matrices $C_{X_1 X_1}$ and $C_{X_2 X_2}$ we arrive to a standard eigenvalue problem:

$$
\begin{bmatrix}\label{eq:cca_ev}
  0 & C_{X_1 X_1}^{-1} C_{X_1 X_2} \\
  C_{X_2 X_2}^{-1} C_{X_2 X_1} & 0 \\
\end{bmatrix} \cdot
\begin{bmatrix}
  w_1 \\
  w_2 \\
\end{bmatrix} =
\lambda
\begin{bmatrix}
  w_1 \\
  w_2 \\
\end{bmatrix}
$$

We then compute the first $d$ largest eigenvectors and their eigenvalues $(\lambda_i, w_1^i, w_2^i),~i=1,\ldots,d$, where $w_1^i$ and $w_2^i$ are scaled so that constraints in \ref{eq:cca} are satisfied: ${w_1^i}^T C_{X_1 X_1} w_1^i = 1$ and ${w_2^i}^T C_{X_2 X_2} w_2^i = 1$.

Combining the constraints with $C_{X_1 X_1}^{-1} C_{X_1 X_2} w_2^i = \lambda_i w_1^i$ from \ref{eq:cca_ev} we observe a close connection between multivariate least squares regression and CCA canonical vectors:
$$C_{X_1 X_1}^{-1} C_{X_1 Y_1} = \left(\sum_{i=1}^d \lambda_i w_1^i {w_2^i}^T\right)C_{X_2 X_2}$$
This means that the basis vectors can be used to form a cross-view similarity mapping:
$$B = \sum_{i=1}^d \lambda_i w_1^i {w_2^i}^T= C_{X_1 X_1}^{-1} C_{X_1 Y_1} C_{X_2 X_2}^{-1}.$$

We have motivated the use of bilinear mapping that seems less biased than the LSR regression variants where the direction of the regression matters.

SVD

Singular value decomposition is a popular approach in dimensionality reduction. It can be readily applied to the cross-view similarity learning in the following way. An aligned sample $\left(S_1, S_2\right)$ of $k$ pairs of observation vectors is analyzed in the following way.

Stack the observations into a matrix of dimensions $k$-by$n_1 +n_2$:
$$
S = \begin{bmatrix} S_1 \\ S_2 \end{bmatrix}
$$
Compute the singular value decomposition of $S = U \Sigma V^T$ and retain the first $d$ columns of $U$ that correspond to the $d$ largest singular values in $\Sigma$. The vectors $U$ can be computed by solving following eigenvalue problem:
$$
\begin{bmatrix}
  S_1 S_1^T & S_1 S_2^T \\
  S_2^T S_1 & S_2 S_2^T \\
\end{bmatrix} \cdot
\begin{bmatrix}
  u_1 \\
  u_2 \\
\end{bmatrix} =
\mu
\begin{bmatrix}
  u_1 \\
  u_2 \\
\end{bmatrix}.
$$
Here we decomposed the eigenvectors to fit the block structure imposed by $S_1$ and $S_2$. Dividing both sides of the equation by the sample size and sending the sample size to infinity corresponds to the optimization on the covariance matrices:
$$
\begin{bmatrix}
  C_{X_1 X_1} & C_{X_1 X_2} \\
  C_{X_2 X_1} & C_{X_2 X_2} \\
\end{bmatrix} \cdot
\begin{bmatrix}
  u_1 \\
  u_2 \\
\end{bmatrix} =
\mu
\begin{bmatrix}
  u_1 \\
  u_2 \\
\end{bmatrix}.
$$
Taking the first $d$ eigenvalue-eigenvector pairs $\left(\mu_i, [{u_1^i}^T, {u_2^i}^T]^T\right)$  we obtain the approximation of the full covariance matrix:
$$
\begin{bmatrix}
  C_{X_1 X_1} & C_{X_1 X_2} \\
  C_{X_2 X_1} & C_{X_2 X_2} \\
\end{bmatrix} \approx \sum_{i = 1}^d \mu_i \begin{bmatrix} u_1^i \\ u_2^i \end{bmatrix} \cdot \begin{bmatrix} {u_1^i}^T & {u_2^i}^T \end{bmatrix}
$$
Comparing the upper-right block shows:
$$C_{X_1 X_2} \approx \sum_{i = 1}^d \mu_i u_1^i {u_2^i}^T.$$
We now see that the SVD based approach yields a cross-view similarity mapping $B = \sum_{i = 1}^d \mu_i u_1^i {u_2^i}^T$ that is an approximation of the regression induced bilinear similarity mapping on inner product space $H_1^1$ (or equivalently $H_2^1$).

Kernel CCA, kernel SVD

In order to apply CCA and SVD to Hilbert spaces other than $H_i^0$ we used their kernel counterparts. The modified methods are described in \cite{shawe-taylor04kernel}.

Corrections: leveraging unaligned data

All involved methods were operating on covariance estimates based on aligned samples. In practice we often encounter the situation where some data is aligned but each view contains additional unaligned data. Such data can be used to obtain better estimates of variance matrices $C_{X_i X_i}$. Let $m=2$. Let $\widetilde{S}_1 \in \RR^{n_1 \times s_1}$ represent a sample of the random variable $X_1$ of size $s_1$, $\widetilde{S}_2 \in \RR^{n_2 \times s_2}$ a sample of random variable $X_2$ of size $s_2$ and $\left(\widehat{S}_1 \subset \widetilde{S}_1, \widehat{S}_2 \subset \widetilde{S}_2\right)$ an aligned sample of size $k$. Assume that the random vectors are centered. We can obtain the following empirical estimates: $\widetilde{C}_{X_1 X_1} := \frac{1}{s_1 -1} \widetilde{S}_1 \widetilde{S}_1^T$, $\widetilde{C}_{X_2 X_2} := \frac{1}{s_2 -1} \widetilde{S}_2 \widetilde{S}_2^T$, $\widehat{C}_{X_1 X_1} := \frac{1}{k -1} \widehat{S}_1 \widehat{S}_1^T$, $\widehat{C}_{X_2 X_2} := \frac{1}{k -1} \widehat{S}_2 \widehat{S}_2^T$ and $\widehat{C}_{X_1 X_2} := \frac{1}{k -1} \widehat{S}_1 \widehat{S}_2^T$. All of the mentioned approaches to finding the bilinear form are based on computations involving the empirical estimates. Computations based on replacing $\widehat{C}_{X_1 X_1}$ with $\widetilde{C}_{X_1 X_1}$ and  replacing$\widehat{C}_{X_2 X_2}$ with $\widetilde{C}_{X_2 X_2}$ typically perform worse than results based on $\widehat{C}_{X_1 X_1}$, $\widehat{C}_{X_2 X_2}$ and $\widehat{C}_{X_1 X_2}$ alone.

The approach we propose is to compute the change of basis that transforms $\widehat{C}_{X_1 X_1}$ to $\widetilde{C}_{X_1 X_1}$, $\widehat{C}_{X_2 X_2}$ to $\widetilde{C}_{X_2 X_2}$ and apply it to $\widehat{C}_{X_1 X_2}$. To achieve this we first compute Cholesky decompositions of $\widetilde{C}_{X_1 X_1} = \widetilde{C}_1^T \widetilde{C}_1$, $\widetilde{C}_{X_2 X_2} = \widetilde{C}_2^T \widetilde{C}_2$, $\widehat{C}_{X_1 X_1} = \widehat{C}_1^T \widehat{C}_1$, $\widehat{C}_{X_2 X_2} = \widehat{C}_2^T \widehat{C}_2$. The matrices that perform the required change of basis are: $\widetilde{C}_1 \widehat{C}_1^{-1}$ and $\widetilde{C}_2 \widehat{C}_2^{-1}$. The new input to our methods, the variance and covariance matrices, are: $\widetilde{C}_{X_1 X_1}$, $\widetilde{C}_{X_2 X_2}$ and $\widetilde{C}_1^T \widehat{C}_1^{-T}  \widehat{C}_{X_1 X_2} \widehat{C}_2^{-1} \widetilde{C}_2$.

Alternative problem definition: differential geometric approach

Let $X = \left(\RR^n, \langle\cdot, \cdot\rangle_X\right)$ and $Y
= \left(\RR^m, \langle\cdot, \cdot\rangle_Y\right)$ be two
Hilbert spaces referred to as views and let $d_X :(X, X)
\rightarrow \RR$ and $d_Y : (Y,Y) \rightarrow \RR$ be the induced
metrics on $\RR^n$ and $\RR^m$ respectively. Let $X \oplus Y$
denote the direct sum of Hilbert spaces where the inner product
is computed as $\langle (x_1, y_1), (x_2, y_2) \rangle_{X \oplus
  Y} = \langle x_1,x_2 \rangle_X + \langle y_1, y_2
\rangle_Y$. Let $A \subset X \oplus Y$ be a Riemannian manifold
with metric tensor $g$. Let $\pi_X: X \oplus Y \rightarrow X$ and
$\pi_Y: X \oplus Y \rightarrow Y$ denote projection operators to
spaces $X$ and $Y$. Let $A_X := \pi_X(A)$ and $A_Y :=
\pi_X(A)$. Let $\pi_{A_X} : X \rightarrow A_X$ and $\pi_{A_Y} : Y
\rightarrow A_Y$ denote projection operators.

Idealized assumption: existence of diffeomorphisms $f_X : A_X \rightarrow A$ and $f_Y : A_Y \rightarrow A$.

Define the cross-view measure of distance as the mapping $B_A : (A_X, A_Y) \rightarrow \RR$,
$$B(x,y) := \int_0^1 \sqrt{ g_{\gamma(t)}\left(\dot{\gamma}(t), \dot{\gamma}(t)\right)}, $$ where $x \in A_X$, $y \in A_Y$, $\gamma$ is a geodesic curve lying in $A$ and $\gamma(0) = \left( x, f_X(x)\right)$ and $\gamma(1) = \left( f_Y(y), y\right)$.

The interpretation of the space $A$ is that it encodes the mutual information between views. An aligned sample $S = (S_X, S_Y) \subset X \oplus Y$ represents a collection of points lying in the space $A$.

 Extending the cross-view measure of distance from $A_X \oplus A_Y$ to full $X \oplus Y$ is straightforward: for $x \in X$ and $y \in Y$ define $B(x,y):= B_A\left(\pi_{A_X}(x), \pi_{A_Y}(y)\right)$.

If $A$ is a linear space

\section{Algorithms}

\noindent {\bf Baseline:}

\noindent {\bf Regression:}

\noindent {\bf CCA with correction:}

\noindent {\bf Global Methods:}

\label{sec:alg}
\section{Experimentation}

\section{Discussion}

\bibliographystyle{unsrt} \bibliography{nips}

\end{document}

### 2015-01-13 material/nips2011.tex
added a paper

\documentclass{article} % For LaTeX2e
\usepackage{nips11submit_e,times,url,comment,amsmath}
%\documentstyle[nips10submit_09,times,art10]{article} % For LaTeX 2.09

\title{Low-Rank Approximations for Large, Multi-lingual Data}

\begin{comment}
\author{
Jan Rupnik \\
A.I. Laboratory\\
Jo\v zef Stefan Institute\\
Ljubljana, Slovenia\\
\texttt{jan.rupnik@ijs.si} \\
\And
Andrej Muhi\v c\\
A.I. Laboratory\\
Jo\v zef Stefan Institute\\
Ljubljana, Slovenia\\
\texttt{andrej.muhic@ijs.si} \\
\AND
Primo\v z \v Skraba \\
A.I. Laboratory\\
Jo\v zef Stefan Institute\\
Ljubljana, Slovenia\\
\texttt{primoz.skraba@ijs.si} \\
}
\end{comment}
\author{
Jan Rupnik, Andrej Muhi\v c, Primo\v z \v Skraba \\
A.I. Laboratory\\
Jo\v zef Stefan Institute\\
Ljubljana, Slovenia\\
\texttt{(jan.rupnik,andrej.muhic,primoz.skraba@ijs.si} \\
}
% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to \LaTeX{} to determine where to break
% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
% puts 3 of 4 authors names on the first line, and the last on the second
% line, try using \AND instead of \And before the third author name.

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

\nipsfinalcopy % Uncomment for camera-ready version

\begin{document}

\maketitle

\begin{abstract}
In this paper we compare low rank approximation methods for data
with a particular structure: documents in multiple
languages. Rather than looking at only 2 languages in time, we
examine the structure in up to 21 languages. The algorithms we
choose to compare are $k$-means, cross-lingual latent semantic
indexing(CL-LSI), and multi-view cannonical correlation analysis
(mCCA). We test these methods on the European Parliament
Proceedings Parallel Corpus.
\end{abstract}

\section{Introduction}
When extracting topics from documents in multiple languages, we
would like to find topics that are not only important within a
language but have an corresponding and equivalent representation
in the other languages. This could be considered a
language-independent measure of importance. Further, by finding a
low-rank approximation which is valid over multiple languages, we
can use well-established machine learning tools designed for
monolingual tasks.  Document collections are a typically
represented as high dimensional data sets, making them a prime
example for lower dimensional representations.

Our work focuses on studying non-probabilistic approaches to
multi-lingual dimensionality reduction. The methods we compare
are described in section~\ref{sec:alg}. Recently a probabilistic
approach based on latent Dirichlet allocation was proposed
\cite{xLDA}, however no implementations were readily available. We plan to implement it and include it in subsequent studies.

The paper is organized as follows: we first introduce the setting
of multilingual data, then describe the algorithms and
datasets. We conclude with the evaluation and discussion.
\begin{comment}
Concepts and... are important across countries. However, much of
what we do is still determined by language. While the Internet
originally had English content almost exclusively, now there is a
diversity of different languages well beyond bi-lingual into
truly multi-lingual territory.

The goal when doing dimensionality reduction on data is to
capture the ``important parts'' of an underlying dataset. These
will preserve the properties of the original high dimensional
data, but faster to deal with algorithmically and highlight the
components (or mixtures thereof) which best explain the data
making it useful for interpretation during analysis as well.

metric analysis

The most common representation of the

metric space....

multilingual is important....

Initially distance of kernel similarity matrices are not only
have an approximate low rank structure, but for this study we
look at matrix with additional structure.

We can view our

 Special structure of multi-lingual data
\end{comment}
\section{Multilingual Data}

Our data is a collection of documents in multiple languages along
with an alignment with correspondences across
languages. Individual documents are represented by a vector $d$
indexed by the terms of a dictionary (the $i$-th element is the
term frequency($TF$) in the document). For a corpus, we gather
document vectors into the term-document matrix $D$.  We choose to
index the columns by documents and the rows by terms $D = (d_1 ,
\ldots , d_m )$. Each document is additionally identified by its
corresponding language. Each language has its own dictionary
independent of other languages.
%The individual
%term-document matrices are grouped according to language giving $
%\{ D_1, D_2, \ldots, D_\ell\}$.

Primarily, we deal with similarities between documents. Since
each language has an independent dictionary, we only define
similarity within a language. We note that given a
transformation function between dictionaries, we could compute
similarities between documents in different languages. However,
as translations of individual words is not tailored to the corpus
in question and is often not well defined (we often have
synonyms, approximate equivalents, etc.), we will use document
correspondences to compute document similarities across
languages. Note that this implies a correspondence between
columns in each $D_k$ for all $k$. Translations between
dictionaries would give us row-based correspondences, but we will
investigate this dimension in future work.

Terms in the dictionary are generally not equally important in
determining similarity between documents, so we must preprocess
the documents. We first use term frequency ($TF$), to prune away
infrequent terms. Rather than take a fixed number of top terms in
each document we use an adaptive measure. Let $f(n)$ be a map
which returns numbers of terms appearing at least $n$ times.  We
find $n$ such that
\begin{equation}
\frac{|f(n+1) - f(n)|}{\mbox{original \# of terms}} < 0.001
\end{equation}
and used these as the terms for the document. Once this pruning
step is complete, we further re-weight the remaining. A term
weight should correspond to the importance of the term for the
given corpus.  The common weighting scheme is called Term
Frequency Inverse Document Frequency (TFIDF) weighting. A
Inverse Document Frequency (IDF) weight for dictionary term $j$
is defined as $w_j = \log( DF_j )$ where $DF_j$ is the number of
documents from the corpora which contain term $j$.  A document
TFIDF vector is its original vector multiplied element-wise by the
weights. The $j$-th element of a document vector is given by $
TF_j \log( DF_j )$. Finally, we re-normalize each vector to have
Euclidean norm equal to 1.

\begin{comment}
We will consider linear combinations of terms as ``concepts.'' The
magnitude of the coefficients of a term in a given combination
could be interpreted as the level of membership of that given
term to the concept. For the application point of view, these
could be interpreted as generalized versions of sets of
terms. Geometrically, we will interpret them as directions in the
term-space.
\end{comment}

\section{Algorithms}
\label{sec:alg}
We compute the low rank approximation of the term-document
matrix using three algorithms: $k$-means\cite{h-ca-75},
cross-lingual latent semantic indexing(CL-LSI)\cite{cl_lsi}, and
multi-view canonical correlation analysis (mCCA)\cite{Kettenring}.

$k$-means is perhaps the most well-known and used clustering
algorithm. The make the data compatible with $k$-means, we merge
all the term-document matrices into a single matrix by stacking
the individual term-document matrices.
\begin{equation}
D_{\mbox{Total}} = \begin{bmatrix}D^T_1 ,D^T_2, \cdots,D^T_\ell\
\end{bmatrix}
^T
\end{equation}
such that the columns respect the alignment of the
documents. Therefore, each document is represented by a long
vector indexed by the terms in all languages. It is these vectors
which determine the similarity on which the $k$-means is computed.

The next method is CL-LSI which is a variant of LSI~\cite{lsi}
for more than one view. Each view in this context represents a
language. It merges the individual term-document matrices in the
same way as we did for $k$-means.  LSI computes a singular value
decomposition of $D_{Total}$. Since the matrix can be large we
can use an iterative method like the Lanczos~\cite{matrix_comp}
algorithm to find the left singular vectors corresponding to the
largest singular values.

Finally, we test mCCA is specifically designed to consider data
from multiple sources (in this case languages). Therefore, we do
not merge the individual term-document matrices into one as in
the other two methods. For each language (view), we estimate the
pairwise correlation coefficients, then we try to find vectors
which maximize the sum of all pairwise correlations over all
languages. This can be written as the following optimization
\begin{equation}
\label{eq:opt1}
\max_{w_1,\ldots, w_\ell} \sum_{i=1}^\ell\sum_{j=i+1}^\ell  \frac{w_i^T D_i D_j^T w_j}{\sqrt{ w_i^T D_i D_i^T w_i\vphantom{ D_j^T } }\sqrt{ w_j^T D_j D_j^T w_j}}
\end{equation}
However, this allows only for a 1-dimensional representation
(i.e. one vector per language). Since we would like to allow for
more vectors per language, we denote the double sum in
Equation~\ref{eq:opt1} by $\mbox{mCCA}(w_1,\ldots,w_\ell)$. If
$\mbox{corr}$ denotes the correlation, to find $M$ vectors, the
optimization objective function  becomes
\begin{equation}
\max_{\substack{w_i^{(j)}\;;\;i=1,\ldots,\ell\\j=1,\ldots,M }}\sum_{s=1}^{\mbox{M}} \mbox{mCCA}\left(w_1^{(s)},\ldots,w_\ell^{(s)}\right),
\qquad\mbox{s.t.} \qquad \mbox{corr} \left(w_i^{(s)},w_i^{(t)}\right) = 0
\quad \forall s\neq t
\end{equation}
We require that the set of $M$ vectors we return for each
language are uncorrelated (so that we do not get copies
of a vector) which together maximize the pairwise correlation
between languages.

\section{Data Set}

To investigate the empirical performance of the low rank
approximations we will test the algorithms on a large-scale,
real-world multilingual dataset: the European Parliament
Proceedings Parallel Corpus v6
(EuroParl)\cite{euro_parl_web}, a corpus released by the
EU Parliament. This source offers a large number of comparable
documents in multiple languages.  In particular, EuroParl
(Release v6) provides transcripts of parliamentary session in
almost all the EU languages. This can be considered a \emph{gold
  standard} in terms of multi-lingual data: these documents are
professional translations of each other.   In the
corpus, each day is available as a separate file, although newer
data is available in smaller units. To create the document,
within each file, we create a document for each speech
ID\footnote{In the corpus, this is referred to as the speaker
  ID.}(so each speech is a document). In the end, the document
contains a document ID, a speech ID, and paragraph markup. The
documents were aligned in two ways: by speaker id and by
paragraphs.  The number of paragraphs was used only if the number
was the same for different languages.  In total, it contains
approximately 21000 documents in the 21 languages of the EU
taking up around 2GB od storage.

\begin{comment}
To investigate the empirical performance of the low rank
approximations we will test the algorithms on two large-scale,
real-world multilingual datasets: European Parliament Proceedings
Parallel Corpus v6 (EuroParl)\cite{europarl,euro_parl_web} and
Wikipedia~\cite{wikipedia}. Both sources offer a large number of
comparable documents in multiple languages.

EuroParl provides documents on legislation, regulations,
parliamentary sessions, etc.  in 21 of the EU languages and is a
\emph{gold standard} in terms of multi-lingual data: these
documents are professional translations of each other.  In the
corpus, each day is available as a separate file, although newer
data is available in smaller units. To create the document,
within each file, we create a document for each speech
ID\footnote{In the corpus, this is referred to as the speaker
  ID.}(so each speech is a document). In the end, the document
contains a document ID, a speech ID, and paragraph markup. The
documents were aligned in two ways: by speaker id and by
paragraphs.  The number of paragraphs was used only if the number
was the same for different languages.  In total, it contains
approximately 21000 documents in the 21 languages of the EU
taking up around 2GB od storage.

Wikipedia is the de-facto encyclopedia on the Internet and can be
downloaded freely. It contains 15 million articles in around 200
languages.The distribution of these articles is not uniform but
heavily biased towards a few languages. The top 3 languages
(English, German, French) contain more than 1 million articles;
the next 6 (Spanish, Italian, Dutch, Japanese, Polish, Portuguese
and Russian) contain more than 650,000 articles; down to Karuni
with 1 article. To make the two dataset comparable, we select the
same languages as above with a similar number of
documents\footnote{Of the chosen languages, Latvian had the
  fewest articles with 37,000 articles}.

\end{comment}

\section{Evaluation}
We measure the performance of the low rank approximation using
two metrics: mean average precision mate retrieval and
correlation between monolingual similarity profiles between the
query document and its nearest neighbour in the joint
representation. For each evaluation, we randomly select a
training set and test set from the data.

The first evaluation criteria we is the \emph{mean average
  precision mate retrieval score} (AMPMR). This measures the
similarity between the documents and their translations in the
common vector space induced by the latent model. Good models map
the documents close to their translations - indicating that some
language independent (semantic) information was captured.  We
evaluate each latent model (given by projection operators $P_1$
and $P_2$) by considering a pair of aligned test sets $T_1$ and
$T_2$ in languages $L_1$ and $L_2$. We select a query document
$q_1 \in T_1$ and denote the corresponding translated document
$q_2\in T_2$.  We then compute the projections $P_1 q$ and $P_2
T_2$ and rank the elements of $P_2 T_2$ by their similarity to
$P_1 q$ in the projection space (measured by Euclidean
distance). The mean average precision mate retrieval score is the
inverse of the rank of $P_2 q_2$.

This score does not give us a complete picture. The low rank of a
mate document does not necessarily indicate poor performance if
the documents which outranked it share similar content to the
query. Therefore, we compute an alternative performance measure:
\emph{correlation between monolingual similarity profiles}
(CMSP). As before, we choose a query document, target test
corpus, and project them to a common vector space. From the
target corpus, we select the closest document $r \in L_2$ to the
query in the projection space.  We then compute two similarity
profile vectors: $v_1$ contains the monolingual cosine similarity
between $q_1$ and all the documents in $T_1$ and similarly $v_2$
contains the cosine similarities between $r$ and $T_2$. The score
 is the correlation coefficient between $v_1$ and $v_2$.

The results for each of the two measures for several fitting
parameters are in Tables~\ref{tb:mean}\footnote{Updated results
  can be found in the full version of this
  paper~\cite{full_version}}. We obtain the results by averaging
over all queries in each test set, all pairs of languages and
over ten repetitions of the choice of training and test data. The
experiments are indexed by the number of training samples
(Ntrain), the number of languages we consider (NViews) and the
dimensionality of the latent space(Ndims). We also ran a
preliminary experiment on a data set based on
Wikipedia\footnote{\url{http://www.wikipedia.org}} for the same
languages.  The result is in the final row of
Table~\ref{tb:mean}.

\begin{table}
\caption{\label{tb:mean} Mean Values for Experiments}
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
Parameters&\multicolumn{3}{|c|}{AMPMR} &  \multicolumn{3}{|c|}{CMSP}\\
\cline{2-7}
(Ntrain,NViews,Ndims)& $k$-means & LSI & mCCA &$k$-means & LSI & mCCA \\
\hline
\hline
(100,   3,    5) & 0.1096 & 0.2306 &0.2184& 0.1135& 0.2007 &0.1889 \\
\hline
(100,   3,   20) & 0.3906 & 0.5292 &0.5151& 0.3308& 0.4466 &0.4329 \\
\hline
(100,   10,   5) & 0.08166& 0.1870 &0.2006& 0.0833& 0.1685 &0.1725\\
\hline
(100,   10,  20) & 0.3204 & 0.4546 &0.4623& 0.2686& 0.3901 &0.3911\\
\hline
(100,   21,   5) & 0.09468& 0.2040 &0.1868& 0.1006& 0.1789 &0.1657\\
\hline
(100,  21,  20) & 0.3255 & 0.4773 &0.4577& 0.2730& 0.4076 &0.3878\\
\hline
(1000,  3,    5) & 0.1090 & 0.4059 &0.3572& 0.1162& 0.3268 &0.2837\\
\hline
(1000,  3,   20) & 0.4180 & 0.8518 &0.8748& 0.3465& 0.7725 &0.7967\\
\hline
(1000,  10,   5) & 0.08320& 0.3126 &0.3801& 0.0894& 0.2533 &0.2965\\
\hline
(1000,  10,  20) & 0.3416 & 0.7695 &0.7954& 0.2880& 0.6921 &0.7148\\
\hline
(1000,  21,   5) & 0.08427& 0.3276 &0.3187& 0.0783& 0.2618 &0.2494\\
\hline
(1000,  21,  20) & 0.3702 & 0.8002 &0.8075& 0.3110& 0.7228 &0.7240\\
\hline
\hline
(1000, 10, 10) &0.2155 & 0.2335 & 0.3169& 0.2612 & 0.2779 & 0.3362\\
\hline
\end{tabular}
\end{center}
\end{table}

\section{Discussion}

The results illustrate that both LSI and mCCA outperform
$k$-means in terms of our evaluation criteria. We believe that
this is due to mCCA and LSI capturing the word co-occurrence
patterns, a well-established fact for the LSI method, which also
holds for mCCA (The decomposition is based on the inter-lingual
covariance matrix). The performance of LSI and mCCA is comparable
in all the tested cases. This is unexpected, since LSI discards
the correspondence information between feature (word) and
language. Note that the above tables are coarse measures, as we
average over all pairs of languages. The two algorithms may have
different fail cases which may explain why this correspondence
information is does not seem important. Furthermore, in our
preliminary experiment, mCCA performed significantly better on
the Wikipedia data set. However, additional experimentation is
required before we can draw conclusions. Finally, we note that
the results show there is still room to improve the methods which
leads us to believe that a further investigation into low-rank
approximations of multilingual structure is warranted.

\bibliographystyle{unsrt} \bibliography{nips}

\pagebreak
\appendix

\section{Learned Concept Vectors}
We include the top 20 vectors we learn, each dimension 5 for
illustrative purposes. We only include 4 languages here, see
~\cite{full_version} for more examples.

\scriptsize
\begin{tabular}{p{0.5\textwidth}p{0.5\textwidth}}
\begin{verbatim}
closed written statements rule 149
women item report voted resolution
you your barroso he mr
women gender equality violence men
you young economic crisis strategy
vote you minutes details item
vote minutes details results see
you acp trade agreement european
euro countries greece acp commission
young budget treaty parliament council
acp agricultural budget agreement farmers
turkey women market accession policy
presidency trafficking violence council turkey
trafficking vote agricultural policy farmers
turkey trafficking balkans accession montenegro
item israel strategy internal trafficking
2020 strategy committee eib small
turkey acp presidency market internal
arctic products greece treaty cosmetic
arctic russia next item young
\end{verbatim}
&
\begin{verbatim}
písemná článek rozprava ukončena prohlášení
žen bodem dalším písemně usnesení
jste barroso vám vás můžete
žen násilí ženy pohlaví ženách
jste krize strategie 2020 barroso
hlasování zápis viz bodem jste
hlasování zápis viz údajů pokračujeme
dohody akt jste partnerství komise
akt komise eurozóny země řecko
parlamentu předsednictví evropského rozpočtu 20
akt dohody zemědělské klimatu zemědělství
žen turecko přistoupení turecka trhu
předsednictví násilí obchodování lidmi rady
obchodování hlasování lidmi písemně soudržnosti
makedonie obchodování turecko lidmi přistoupení
bodem strategie obchodování izrael dalším
2020 eib bodem podniky demokratů
akt turecko předsednictví turecka trh
kosmetických strategie eurozóny arktidy smlouvy
rusko bodem dalším zdraví arktidy
\end{verbatim}
\end{tabular}

\begin{tabular}{p{0.5\textwidth}p{0.5\textwidth}}
\begin{verbatim}
clos article écrites 149 débat
femmes appelle voté rapport l'ordre
vous avez barroso votre pouvez
femmes hommes l'égalité violence vous
vous jeunes crise stratégie économique
votes vote vous verbal commission
votes verbal vote procès résultats
vous acp turquie européen l'accord
euro grèce commission acp zone
jeunes budget traité parlement lisbonne
acp budget agricole pac l'accord
turquie femmes marché intérieur politique
présidence violence turquie conseil êtres
traite vote pac êtres agricole
turquie balkans occidentaux monténégro kosovo
stratégie intérieur l'ordre appelle israël
2020 stratégie entreprises bei pme
turquie acp jeunes intérieur présidence
produits cosmétiques traité zone grèce
russie arctique l'arctique jeunes santé
\end{verbatim}&
\begin{verbatim}
schriftliche geschlossen erklärungen aussprache artikel
frauen gestimmt bericht schriftlich erkläre
sie barroso herr antwort selbstverständlich
frauen gleichstellung gewalt männern geschlechter
strategie sie 2020 barroso krise
abstimmung kommission the nächster punkt
abstimmung the protokoll siehe abstimmungsstunde
akp türkei abkommen kommission staaten
eurozone griechenland akp kommission länder
lissabon parlaments parlament arbeitsmarkt vertrag
akp agrarpolitik abkommen versammlung gap
türkei frauen binnenmarkt agrarpolitik gap
gewalt menschenhandel türkei ratsvorsitz präsidentschaft
menschenhandel agrarpolitik abstimmung gap landwirte
türkei menschenhandel mazedonien montenegro kosovo
israel binnenmarkt strategie punkt menschenhandel
2020 strategie eib unternehmen kopenhagen
türkei akp binnenmarkt gewalt versammlung
arktis nanomaterialien eurozone griechenland vertrag
russland arktis arktischen punkt frau
\end{verbatim}
\end{tabular}

\end{document}

\pagebreak
\appendix

\section{Learned Concept Vectors}
We include the top 20 vectors we learn, each dimension 5 for%five most significant words for each of the multilingual topics 
illustrative purposes. We only include 4 languages here.

\scriptsize
\begin{tabular}{p{0.5\textwidth}p{0.5\textwidth}}
\begin{verbatim}
closed written statements rule 149
women item report voted resolution
you your barroso he mr
women gender equality violence men
you young economic crisis strategy
vote you minutes details item
vote minutes details results see
you acp trade agreement european
euro countries greece acp commission
young budget treaty parliament council
acp agricultural budget agreement farmers
turkey women market accession policy
presidency trafficking violence council turkey
trafficking vote agricultural policy farmers
turkey trafficking balkans accession montenegro
item israel strategy internal trafficking
2020 strategy committee eib small
turkey acp presidency market internal
arctic products greece treaty cosmetic
arctic russia next item young
\end{verbatim}
&
\begin{verbatim}
písemná článek rozprava ukončena prohlášení
žen bodem dalším písemně usnesení
jste barroso vám vás můžete
žen násilí ženy pohlaví ženách
jste krize strategie 2020 barroso
hlasování zápis viz bodem jste
hlasování zápis viz údajů pokračujeme
dohody akt jste partnerství komise
akt komise eurozóny země řecko
parlamentu předsednictví evropského rozpočtu 20
akt dohody zemědělské klimatu zemědělství
žen turecko přistoupení turecka trhu
předsednictví násilí obchodování lidmi rady
obchodování hlasování lidmi písemně soudržnosti
makedonie obchodování turecko lidmi přistoupení
bodem strategie obchodování izrael dalším
2020 eib bodem podniky demokratů
akt turecko předsednictví turecka trh
kosmetických strategie eurozóny arktidy smlouvy
rusko bodem dalším zdraví arktidy
\end{verbatim}
\end{tabular}

\begin{tabular}{p{0.5\textwidth}p{0.5\textwidth}}
\begin{tabular}{p{0.5\textwidth}p{0.5\textwidth}}
\begin{verbatim}
clos article écrites 149 débat
femmes appelle voté rapport l'ordre
vous avez barroso votre pouvez
femmes hommes l'égalité violence vous
vous jeunes crise stratégie économique
votes vote vous verbal commission
votes verbal vote procès résultats
vous acp turquie européen l'accord
euro grèce commission acp zone
jeunes budget traité parlement lisbonne
acp budget agricole pac l'accord
turquie femmes marché intérieur politique
présidence violence turquie conseil êtres
traite vote pac êtres agricole
turquie balkans occidentaux monténégro kosovo
stratégie intérieur l'ordre appelle israël
2020 stratégie entreprises bei pme
turquie acp jeunes intérieur présidence
produits cosmétiques traité zone grèce
russie arctique l'arctique jeunes santé
\end{verbatim}&
\begin{verbatim}
schriftliche geschlossen erklärungen aussprache artikel
frauen gestimmt bericht schriftlich erkläre
sie barroso herr antwort selbstverständlich
frauen gleichstellung gewalt männern geschlechter
strategie sie 2020 barroso krise
abstimmung kommission the nächster punkt
abstimmung the protokoll siehe abstimmungsstunde
akp türkei abkommen kommission staaten
eurozone griechenland akp kommission länder
lissabon parlaments parlament arbeitsmarkt vertrag
akp agrarpolitik abkommen versammlung gap
türkei frauen binnenmarkt agrarpolitik gap
gewalt menschenhandel türkei ratsvorsitz präsidentschaft
menschenhandel agrarpolitik abstimmung gap landwirte
türkei menschenhandel mazedonien montenegro kosovo
israel binnenmarkt strategie punkt menschenhandel
2020 strategie eib unternehmen kopenhagen
türkei akp binnenmarkt gewalt versammlung
arktis nanomaterialien eurozone griechenland vertrag
russland arktis arktischen punkt frau
\end{verbatim}
\end{tabular}

\begin{verbatim}
afsluttet forhandlingen skriftlige artikel erklæringer
kvinder punkt stemte vold betænkning
barroso han hr dem davies
kvinder ligestilling mænd vold kvinders
unge økonomiske 2020 barroso euroområdet
afstemningen protokollen oplysninger afstemning resultater
afstemningen protokollen resultater afstemning oplysninger
avs økonomiske tyrkiet politiske rettigheder
euroområdet grækenland lande avs forslag
unge parlamentet lissabontraktaten parlamentets budget
avs landbrugspolitik fælles aftalen ændringsforslag
tyrkiet landbrugspolitik kvinder marked fælles
formandskab menneskehandel vold tyrkiet spanske
menneskehandel landbrugspolitik afstemningen fælles områder
tyrkiet menneskehandel montenegro kosovo serbien
punkt israel menneskehandel indre marked
2020 eib små virksomheder var
tyrkiet avs unge marked indre
arktis euroområdet grækenland kosmetiske oplysninger
rusland arktis næste arktiske unge
\end{verbatim}&

\begin{tabular}{p{0.5\textwidth}p{0.5\textwidth}}
\begin{verbatim}
cierra artículo declaraciones 142 149
mujeres votado informe siguiente resolución
usted barroso le señor qué
mujeres género igualdad violencia hombres
usted crisis estrategia jóvenes barroso
votación acta véase comisión usted
votación véase acta continuamos resultados
acp acuerdo usted turquía comisión
euro países grecia comisión acp
jóvenes presupuesto parlamento tratado lisboa
acp presupuesto agrícola acuerdo pac
turquía mujeres mercado política adhesión
presidencia violencia seres turquía consejo
votación agrícola seres barroso tráfico
turquía balcanes montenegro occidentales kosovo
israel estrategia siguiente interior gaza
2020 estrategia bei pyme nombre
turquía acp mercado presidencia jóvenes
productos ártico cosméticos tratado grecia
rusia ártico ártica gas salud
\end{verbatim}&
\begin{verbatim}
kirjalikud lõppenud avaldused artikkel kodukorra
naiste päevakorrapunkt järgmine kirjalikult hääletasin
te teie olete barroso teil
naiste soolise võrdõiguslikkuse vägivalla naised
te 2020 strateegia noorte barroso
vt protokoll üksikasjad päevakorrapunkt järgmine
vt üksikasjad protokoll tulemused muud
akv türgi lepingu komisjoni te
euroala kreeka akv riikide komisjon
parlamendi lissaboni noorte eelarve oli
akv põllumajanduspoliitika eelarve ühisassamblee parlamentaarse
türgi naiste põllumajanduspoliitika soolise siseturu
nõukogu türgi inimkaubanduse vägivalla eesistujariik
põllumajanduspoliitika inimkaubanduse lissaboni barroso ühtekuuluvuspoliitika
türgi makedoonia jugoslaavia inimkaubanduse montenegro
iisraeli järgmine päevakorrapunkt palestiina inimkaubanduse
2020 eip strateegia suurusega naiste
türgi akv hispaania siseturu lepingu
arktika euroala kosmeetikatoodete lepingu kreeka
arktika venemaa päevakorrapunkt järgmine toiduohutuse
\end{verbatim}
\end{tabular}
\begin{tabular}{p{0.5\textwidth}p{0.5\textwidth}}
\begin{verbatim}
päättynyt työjärjestyksen artikla kirjalliset lausumat
naisten seuraavana esityslistalla kirjallinen laatima
te barroso puheenjohtaja olette teidän
naisten tasa naisiin sukupuolten miesten
2020 te barroso talouden talous
pöytäkirja kulkua äänestysten ks yksityiskohdat
kulkua äänestysten ks yksityiskohdat pöytäkirja
akt sopimuksen turkin euroopan valtioiden
akt kreikan valtioiden komissio kreikka
lissabonin parlamentin talousarvion nuoret neuvosto
akt yhteisen edustajakokouksen cotonoun talousarvion
turkin naisten turkki tasa 2020
puheenjohtajavaltio ihmiskaupan neuvoston väkivallan turkin
ihmiskaupan lissabonin kirjallinen alueiden viljelijöille
ihmiskaupan turkin länsi jugoslavian balkanin
israelin esityslistalla seuraavana laatima ihmiskaupan
2020 venäjän seuraavana naisten laatima
turkin akt turkki puheenjohtajavaltio sisämarkkinoita
tietojen aineiden lissabonin kreikan tärkeää
venäjän arktisen seuraavana esityslistalla venäjä
\end{verbatim}
\end{tabular}
\begin{verbatim}
rásbeli nyilatkozatok lezárom vitát szabályzat
nők rásban szavaztam napirendi pont
ön barroso úr kapni köszönöm
nők nemek férfiak erőszak egyenlőség
gazdasági ön fiatalok 2020 barroso
szavazás lásd bizottság napirendi szavazásra
lásd szavazás szavazásra jegyzőkönyvet jegyzőkönyvben
akcs európai ön közös politikai
országok akcs görögország bizottság fejlődő
fiatalok parlament lisszaboni szerződés tanács
akcs mezőgazdasági közgyűlés közös költségvetés
törökország nők közös piac belső
emberkereskedelem tanács elnökség erőszak spanyol
emberkereskedelem mezőgazdasági barroso nemzeti szavazásra
emberkereskedelem macedónia jugoszláv törökország nyugat
következő emberkereskedelem palesztin napirendi pont
2020 ebb stratégia kis következő
akcs törökország török fiatalok spanyol
kozmetikai sarkvidék szerződés északi termékek
oroszország sarkvidék északi sarkvidéki napirendi
\end{verbatim}
\end{tabular}
\begin{tabular}{p{0.5\textwidth}p{0.5\textwidth}}
\begin{verbatim}
chiusa scritte articolo dichiarazioni svolgerà
donne reca l'ordine votato relazione
lei barroso riceverà avete onorevole
donne violenza uomini genere parità
lei giovani crisi strategia barroso
votazione verbale commissione dettagliati votazioni
votazione dettagliati verbale vedasi votazioni
acp turchia lei sviluppo commissione
paesi grecia commissione donne acp
giovani bilancio trattato parlamento lisbona
acp agricola bilancio paritetica revisione
turchia donne mercato politica interno
presidenza violenza consiglio turchia esseri
votazione esseri barroso agricoltori agricola
turchia balcani macedonia iugoslava occidentali
strategia israele interno reca l'ordine
2020 strategia pmi imprese piccole
turchia acp interno presidenza giovani
prodotti cosmetici trattato grecia artica
russia artica giovani gas reca
\end{verbatim}&
\begin{verbatim}
raštiški pareiškimai straipsnis tvarkos 142
moterų balsavau moteris pranešimas raštu
jūs jūsų barroso j galite
moterų moteris lyčių vyrų lygybės
jūs euro 2020 ekonomikos darbo
protokolą balsavimo žr jūs rezultatai
žr protokolą balsavimo rezultatai išsamūs
akr europos susitarimo turkijos prekybos
euro akr komisija zonos šalių
biudžeto parlamento lisabonos jaunimo užimtumo
akr žemės ūkio klimato energijos
turkijos moterų turkija ūkio stojimo
žmonėmis tarybai moteris pirmininkaujanti turkijos
žmonėmis žemės ūkio jūs sanglaudos
žmonėmis turkijos jugoslavijos vakarų kosovo
žmonėmis miškų vidaus 2020 izraelio
2020 eib klausimas demokratų vardu
akr turkijos turkija jaunimo vidaus
kosmetikos arkties euro produktų lisabonos
arkties rusijos rusija klausimas saugos
\end{verbatim}
\end{tabular}
\begin{tabular}{p{0.5\textwidth}p{0.5\textwidth}}
\begin{verbatim}
slēgtas pants reglamenta rakstiskas deklarācijas
nākamais sieviešu balsoju punkts rakstiski
jūs jums jūsu barroso esat
sieviešu sievietēm dzimumu sievietes vīriešu
jūs 2020 eiro jauniešu darba
balsošanas jūs nākamais balsošanu protokolu
balsošanas protokolu balsošanu detalizēta sk
ākk jūs tirdzniecības eiropas nolīguma
eiro ākk jaunattīstības komisija zonas
budžeta parlamenta jauniešu lisabonas jauniešiem
ākk lauksaimniecības nolīguma klp klimata
sieviešu turcijas lauksaimniecības tirgus turcija
prezidentūras prezidentūra spānijas gada čehijas
lauksaimniecības klp jūs barroso kohēzijas
dienvidslāvijas maķedonijas turcijas rietumbalkānu melnkalne
nākamais tirgus punkts mums palestīnas
2020 eib gadam mvu budžeta
ākk turcijas turcija spānijas turciju
kosmētikas arktikas lisabonas eiro datu
arktikas krievijas nākamais gāzes arktiku
\end{verbatim}&
\begin{verbatim}
schriftelijke verklaringen gesloten artikel 149
vrouwen verslag schriftelijk gestemd verklaar
u uw bent barroso zult
vrouwen geweld mannen gendergelijkheid u
u jongeren crisis strategie economische
stemmingen u notulen commissie stemming
stemmingen notulen stemming uitslagen bijzonderheden
acs overeenkomst u turkije handel
eurozone griekenland landen commissie acs
jongeren verdrag parlement begroting raad
acs overeenkomst begroting landbouwbeleid landbouw
turkije vrouwen markt landbouwbeleid interne
voorzitterschap mensenhandel geweld raad turkije
mensenhandel landbouwbeleid landbouwers barroso gemeenschappelijk
turkije mensenhandel westelijke montenegro kosovo
israël strategie interne mensenhandel markt
2020 kleine strategie middelgrote ondernemingen
turkije acs jongeren voorzitterschap interne
eurozone arctische producten griekenland verdrag
arctische rusland orde jongeren regio
\end{verbatim}
\end{tabular}
\begin{tabular}{p{0.5\textwidth}p{0.5\textwidth}}
\begin{verbatim}
zamykam pisemne regulaminu oświadczenia art
kobiet punktem sprawozdania piśmie przemocy
pan barroso pani pana otrzyma
kobiet płci przemocy mężczyzn pan
pan 2020 euro strategii pracy
głosowanie patrz pan protokół głosowania
głosowanie patrz protokół głosowania wyniki
umowy akp komisji pan europejskiej
euro akp kraje komisja kobiet
parlamentu budżetu młodych traktatu euro
akp umowy budżetu rolnej wpr
kobiet turcji turcja rynku wpr
ludźmi prezydencji przemocy prezydencja turcji
ludźmi głosowanie barroso obszarów pan
ludźmi turcji macedonii turcja bałkanów
ludźmi punktem strategii 2020 gazy
2020 ebi kobiet małych punktem
akp turcji turcja rynku prezydencja
produktów traktatu informacji strefy danych
pani rosji gazu rosja kolejnym
\end{verbatim}&
\begin{verbatim}
encerrado escritas artigo declarações 149
mulheres segue votei relatório reaberta
barroso senhor deputado presidente presidência
mulheres igualdade violência homens género
jovens crise estratégia barroso económica
votação acta pormenorizados comissão segue
votação pormenorizados acta votações resultados
acp acordo turquia comissão comércio
euro países grécia comissão acp
jovens orçamento parlamento tratado lisboa
acp orçamento agrícola alterações paritária
turquia mulheres mercado adesão igualdade
presidência violência turquia tráfico conselho
tráfico agrícola votação seres barroso
turquia balcãs ocidentais tráfico kosovo
israel estratégia segue interno ordem
2020 estratégia pequenas bei empresas
turquia acp jovens interno mercado
árctico produtos cosméticos tratado grécia
árctico rússia jovens segue gás
\end{verbatim}
\end{tabular}
\begin{tabular}{p{0.5\textwidth}p{0.5\textwidth}}
\begin{verbatim}
închisă scrise declaraţii dezbaterea articolul
femeilor votat următorul femei scris
aţi dumneavoastră veţi barroso vă
femeilor femei bărbaţi gen egalitatea
aţi veţi muncă 2020 barroso
the comisiei ordinea următorul zi
the votarea votul vot detalii
acp aţi european turcia comisiei
euro acp ţările grecia comisia
tinerii lisabona tinerilor 20 parlamentului
acp agricolă pac bugetul acordului
turcia femeilor internă aderare piaţa
preşedinţia turcia consiliului preşedinţiei traficul
barroso traficul traficului the pac
turcia kosovo vest balcanii fosta
următorul gaza internă israelul traficul
2020 strategia 2010 bei mijlocii
turcia acp piaţa preşedinţia internă
arctică cosmetice produsele grecia lisabona
arctică rusia regiunea următorul naturale
\end{verbatim}&
\begin{verbatim}
skončila písomné článok vyhlásenia rozprava
žien alším bodom hlasoval písomne
ste barroso vám pán predseda
žien ženy násilia ženách rovnosti
ste 2020 mladých barroso krízy
ste podrobnostiach zápisnicu hlasovanie hlasovania
podrobnostiach zápisnicu pozri hlasovania hlasovanie
akt dohody ste európskej komisie
eurozóny akt krajiny komisia grécko
parlamentu mladých rozpočet zmluvy lisabonskej
akt dohody klímy spp rozpočet
žien turecko turecka trh spp
predsedníctvo ľuďmi rady predsedníctva násilia
ľuďmi hlasovanie spp barroso ste
ľuďmi balkánu macedónsko turecko pokroku
alším bodom ľuďmi izrael potrebujeme
2020 malých eib stredných podnikov
akt turecko turecka predsedníctvo trhu
kozmetických arktídu výrobkov údajov eurozóny
arktídu rusko pani bodom alším
\end{verbatim}
\end{tabular}
\begin{tabular}{p{0.5\textwidth}p{0.5\textwidth}}
\begin{verbatim}
pisne člen zaključena izjave razprava
naslednja žensk pisni obliki poročilo
ste boste barroso gospod vas
žensk ženskami spolov ženske nasilja
ste boste 2020 barroso mlade
glasovanje izide zapisnik glej ste
izide glasovanje zapisnik glej podrobnosti
akp sporazuma ste komisije evropski
akp komisija grčija gozdov države
mlade parlamenta proračun pogodbe proračuna
akp sporazuma proračun proračuna kmetijske
žensk turčija turčije trg politika
predsedstvo ljudmi sveta nasilja proračun
ljudmi glasovanje skp barroso ravni
ljudmi turčija napredku turčije jugoslovanska
točka ljudmi naslednja izrael 2020
2020 srednje eib podjetja žensk
akp turčija turčije predsedstvo mlade
kozmetičnih podatkov izdelkov pogodbe pomembno
naslednja rusija rusijo točka seveda
\end{verbatim}&
\begin{verbatim}
avslutad härmed förklaringar skriftliga debatten
kvinnor punkt röstade män våld
ni er barroso han ert
kvinnor män våld jämställdhet ni
ni ungdomar krisen ekonomiska 2020
omröstningen ni protokollet omröstningsresultat punkt
omröstningen omröstningsresultat protokollet uppgifter rör
avs ni handel avtalet europeiska
euroområdet länderna avs kommissionen grekland
ungdomar lissabonfördraget ungdomarna budget förslaget
avs gemensamma jordbrukspolitiken avtalet församlingen
turkiet kvinnor gemensamma marknaden jordbrukspolitiken
ordförandeskapet våld människohandel turkiet spanska
omröstningen jordbrukspolitiken ni människohandel gemensamma
turkiet republiken balkan f västra
punkt inre israel marknaden åtgärder
2020 små företag medelstora strategin
turkiet avs inre ordförandeskapet ungdomar
arktis produkter euroområdet nanomaterial grekland
arktis ryssland nästa ungdomar ni
\end{verbatim}
\end{tabular}

\begin{table}
\begin{center}
\caption{Standard Deviation}
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
Parameters&\multicolumn{3}{|c|}{AMPMR} &  \multicolumn{3}{|c|}{CMSP}\\
\cline{2-7}
(NViews,Ndims,Ntrain)& $k$-means & LSI & mCCA &$k$-means & LSI & mCCA \\
\hline
\hline
 &    0.0058  &   0.0100  &   0.0087  &   0.0089  &   0.0118  &   0.0098 \\
  \hline
 &    0.0046  &   0.0424  &   0.0338  &   0.0071  &   0.0341  &   0.0300 \\
  \hline
 &    0.0061  &   0.0097  &   0.0147  &   0.0081  &   0.0125  &   0.0187 \\
  \hline
 &   0.0068  &   0.0416  &   0.0381  &   0.0135  &   0.0332  &   0.0285 \\
  \hline
 &   0.0151  &   0.0093  &   0.0101  &   0.0176  &   0.0122  &   0.0137 \\
  \hline
 &    0.0045  &   0.0530  &   0.0463  &   0.0046  &   0.0458  &   0.0397 \\
\hline\hline
\end{tabular}
\end{center}
\end{table}

### 2015-01-13 rupnik2015.tex

\documentclass[twoside,11pt]{article}
\usepackage{jair, theapa, rawfonts}

\jairheading{1}{1993}{1-15}{6/91}{9/91}
\ShortHeadings{Cross-lingual Document Similarity and Event Detection}
{Rupnik, Muhi\v{c}, Leban, Fortuna, \v{S}kraba \& Grobelnik}
\firstpageno{25}

\begin{document}

\title{Cross-lingual Document Similarity and Event Detection}

\author{\name Jan Rupnik \email jan.rupnik@ijs.si \\
       \name Andrej Muhi\v{c} \email andrej.muhic@ijs.si \\
       \name Gregor Leban \email gregor.leban@ijs.si \\
       \name Bla\v{z} Fortuna \email blaz.fortuna@ijs.si \\
       \name Primo\v{z} \v{S}kraba \email primoz.skraba@ijs.si \\
       \name Marko Grobelnik \email marko.grobelnik@ijs.si \\
       \addr Artificial Intelligence Laboratory, Jozef Stefan Institute,\\
       Jamova cesta 39, 1000 Ljubljana, Slovenia}

\maketitle

\begin{abstract}
Tralala hopsasa.
\end{abstract}

\section{Introduction}
\label{Introduction}

\section{Cross-lingual Document Similarity}

\subsection{Problem definition}

\subsection{Wikipedia dataset}

100 languages!

\section{Multilingual Latent Factor Models}

\subsection{Algorithm}

\subsubsection{LSI}

\subsubsection{Refinement of LSI similar to MCCA}

\subsubsection{Hub languages}

\subsection{Scaling}

100 languages!

\subsection{Evaluation}

\subsubsection{Link prediction}

\section{Cross-lingual Event Linking}

\subsection{Problem definition}

\subsection{Algorithm}

\subsection{Evaluation}

\subsubsection{Compare different feature subsets}

\subsubsection{Accuracy across different language pairs}

\section{Related work}

\subsection{Cross-lingual Document Similarity}

\subsection{Multilingual Latent Factor Models}

Ask Wray

\subsection{Cross-lingual Event Linking}

\section{Discussion}

\acks{The authors wish to thank blah blah blah.}

\vskip 0.2in
\bibliography{rupnik2015}
\bibliographystyle{theapa}

\end{document}

### 2015-01-13 rupnik2015.tex
Added material from workshop papers to the manuscript

@@
 
-\jairheading{1}{1993}{1-15}{6/91}{9/91}
+\usepackage{color}
+\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}
+
+\usepackage{hyperref}
+
+\usepackage{amsmath}
+\usepackage{amsfonts}
+\usepackage{amssymb}
+
+\newcommand{\RR}{\mathbb{R}}
+\newcommand{\X}{\mathbb{X}}
+\newcommand{\Y}{\mathbb{Y}}
+
+\jairheading{x}{xxxx}{x-xx}{x/xx}{x/xx}
 \ShortHeadings{Cross-lingual Document Similarity and Event Detection}
 {Rupnik, Muhi\v{c}, Leban, Fortuna, \v{S}kraba \& Grobelnik}
-\firstpageno{25}
+\firstpageno{1}
 
@@
 
-\title{Cross-lingual Document Similarity and Event Detection}
+\title{Cross-lingual Document Similarity and Event Detection \todo{Better title...}}
 
@@
        \name Gregor Leban \email gregor.leban@ijs.si \\
+       \name Primo\v{z} \v{S}kraba \email primoz.skraba@ijs.si \\
        \name Bla\v{z} Fortuna \email blaz.fortuna@ijs.si \\
-       \name Primo\v{z} \v{S}kraba \email primoz.skraba@ijs.si \\
        \name Marko Grobelnik \email marko.grobelnik@ijs.si \\
-       \addr Artificial Intelligence Laboratory, Jozef Stefan Institute,\\
+       \addr Artificial Intelligence Laboratory, Jo\v{z}ef Stefan Institute,\\
        Jamova cesta 39, 1000 Ljubljana, Slovenia}
@@
 \begin{abstract}
-Tralala hopsasa.
+\todo{tralala hopsasa.}
 \end{abstract}
@@
 \section{Introduction}
-\label{Introduction}
+
+% taken from XLite paper
+Document retrieval is a well-established problem in data mining.
+There have been a large number of different approaches put forward in the literature.
+In this paper we concentrate on a specific setting: multi-lingual corpora.
+As the availability of multi-lingual documents has exploded in the last few years, there is a pressing need for automatic cross-lingual processing tools.
+
+The prime example is Wikipedia - in 2001 the majority of pages were written in English, while in 2012, the percentage of English articles has dropped to 14\%.
+In this context, we look at how to find similar documents across languages.
+In particular, we do not assume the availability of machine translators, but rather try to frame the problem such that we can use well-established machine learning tools designed for monolingual text-mining tasks.
+This work represents the continuation of previous work \cite{nips, iti, nips2, sikdd} where we explored representations of documents which were valid over multiple languages.
+The representations could be interpreted as multi-lingual topics, which were then used as proxies to compute cross-lingual similarities between documents.
+We look at a specific aspect of this problem: the distribution of articles across languages in Wikipedia is not uniform.
+While the percentage of English articles has fallen, English is still the largest language and one of \emph{hub} languages which not only have an order of magnitude more articles than other languages, but also many comparable articles with most of the other languages.
+
+When doing document retrieval, we encounter two quite different situations.
+If for example, we are looking for a German article comparable to an English article, there is a large alignment between the document corpora (given by the intersection in articles in the two languages) making the problem well-posed.
+If, however, we look for a relevant Slovenian article given a Hindi article, the intersection is small, making the problem much harder.
+Since almost all languages have a large intersection in articles with \emph{hub} languages, the question we ask in this paper is: can we exploit hub languages to perform better document retrieval between non-hub languages?
+A positive answer would vastly improve cross-lingual analysis between less represented languages.
+In the following section, we introduce our representation followed by our experiments, which also shed light on the structural properties of the multilingual Wikipedia corpus.
 
@@
 
-\subsection{Wikipedia dataset}
+%Rough structure:
+% - Document Representation}
+% - Multi-language representation}
+% - Document Similarity
 
-100 languages!
+% taken from ITI paper
+Our data is a collection of documents in multiple languages along with an alignment with correspondences across languages.
+Individual documents are represented as vectors by using the standard vector space model where each term corresponds to a word or a phrase in a fixed vocabulary.
+An aligned corpus is represented by a set of matrices $D^J$ defined as $D^J := [d_1^J , \ldots , d_{m_j}^J ] \in \RR^{n_J \times m_J}$ where $m_J$ is the number of documents in language $J$ and $n_J$ is the dictionary size.
+Let $N = \sum_{J=1}^M n_J$, where $M$ is the number of languages.
+Each multi-lingual document of the aligned corpus is formed as a block vector $d_k = [ d_k^1; \dots; d_k^M] \in \RR^{N}$, and documents are indexed so that $d_k^{J_1}$ and $d_k^{J_2}$ is an aligned document pair for all $k, J_1, J_2$.
+
+Our main interest are similarities between documents.
+We can measure the similarities between documents of the same language by using the cosine similarity.
+We note that given a transformation function between dictionaries, we could compute similarities between documents in different languages.
+However, bilingual-dictionaries are not available for many language pairs in the Wikipedia corpus and we will rely on document correspondences to compute document similarities across languages in the present paper.
+Note that this implies a correspondence between columns in each $D^K$ for all $K$.
+
+Terms in the dictionary are generally not equally important in determining similarity between documents, so we must preprocess the documents.
+We first use term frequency ($TF$), to prune away infrequent terms.
+Rather than take a fixed number of top terms in each document we use an adaptive measure.
+
+Let $f(n)$ be a map which returns numbers of terms appearing at least $n$ times.
+For each language $k$ we find maximal $n$ such that $\max(10^5, N_k)\leq f(n) \leq 6\cdot 10^5,$ where $N_k$ is number of documents in complete dictionary of language $k.$ As we do not apply any stemming and lemmatization this ensures that we get well balanced corpus and retain at least of $10^5$ words for languages with rather small Wikipedias but potentially rich vocabularies.
+Once this pruning step is complete, we further re-weight the remaining terms.
+A term weight should correspond to the importance of the term for the given corpus.
+The common weighting scheme is called Term Frequency Inverse Document Frequency (TFIDF) weighting.
+An Inverse Document Frequency (IDF) weight for the dictionary term $j$ is defined as $w_j = \log( DF_j )$, where $DF_j$ is the number of documents in the corpus which contain term $j$.
+A document TFIDF vector is its original vector multiplied element-wise by the weights.
+The $j$-th element of a document vector is given by $ TF_j \log( DF_j )$.
+Finally, we re-normalize each vector to have Euclidean norm equal to 1.
+
+% taken from XLite paper
+We begin by introducing some notation.
+A \emph{multilingual document} $d = (u_1,\ldots u_m)$ is a collection of $m$ documents on the same topic (comparable), where $u_i$ is the document in language $i$ which can be an empty document (missing resource) and $d$ must contain at least two nonempty documents.
+A comparable corpus $D = {d_1, \ldots, d_N}$ is a collection of multilingual documents.
+By using the standard vector space model (bag of words) we can represent $D$ as a set of $m$ matrices $X_1,\ldots,X_m$, where $X_i \in \RR^{n_i \times N}$ is the corpus matrix corresponding to the $i$-th language, where $n_i$ is the vocabulary size.
+Let $X_i^{\ell}$ denote the $\ell$-th column of matrix $X_i$.
+A \emph{hub language} is a language with a high proportion of non-empty documents in $D$, which in case of Wikipedia is English.
+Let $a(i,j)$ denote the index set of multilingual documents $d$ that contain non-empty $u_i$ and $u_j$ and $a(i)$ denote the index set of multilingual documents that contain $u_i$.
+
+\subsection{Datasets}
+
+% Taken from ITI paper
+To investigate the empirical performance of the low rank approximations we will test the algorithms on a large-scale, real-world multilingual dataset that we extracted from Wikipedia by using inter-language links as an alignment.
+This  results in a large number of weakly comparable documents in more than $200$ languages.
+Wikipedia is a large source of multilingual data that is especially important for the languages for which no translation tools, multilingual dictionaries as Eurovoc, or strongly aligned multilingual corpora as Europarl are available.
+Documents in different languages are related with so called 'inter-language' links that can be found on the left of the Wikipedia page.
+The Wikipedia is constantly growing.
+There are currently four Wikipedias with more than $10^6$ articles, $40$ with more than $10^5$ articles, $100$ with more than $10^4$ articles, and $216$ with more than $1000$ articles.
+Wikipedia uses special user-friendly markup language that is very easy to write but very hard to parse.
+Simplicity of language can cause ambiguities and moreover it is constantly changing.
+For example, separator | can be used in different contexts.
+
+Wikipedia raw xml dumps of all currently $270$ active editions were downloaded from the Wikipedia dump page\href{http://dumps.wikimedia.org}.
+The xml files are too large to be parsed with DOM like parser that needs to store the whole xml tree in the memory, instead we implemented Sax like parser that tries to simulate behavior of Wikipedia official parser and is as simple, fast and error prone as possible.
+We parse all Wikipedia markup but do not extend the templates.
+Each Wikipedia page is embedded in the page tag.
+First we check if the title of the page consists of any special namespace and do not process such pages.
+Then we check if this is a redirection page and we store the redirect link as inter-language links can point to redirection link also.
+If nothing of the above applies we extract the text and parse the Wikipedia markup.
+Currently all the markup is removed.
+
+We get inter-language link matrix using previously stored redirection links and inter-language links.
+If inter-language link points to the redirection we replace it with the redirection target link.
+It turns out that we obtain the matrix $M$ that is not symmetric, consequently the underlying graph is not symmetric.
+That means that existence of the inter-language link in one way (i.e.
+English to German) does not guarantee that there is an inter-language link in the reverse direction (German to English).
+To correct this we transform this matrix to symmetric by computing $M+M^T$ and obtaining an undirected graph.
+In the rare case that we have multiple links pointing from the document, we pick the first one that we encountered.
+This matrix enables us to build an alignment across all Wikipedia languages.
 
@@
 
-\subsection{Algorithm}
+\subsection{Algorithms}
+
+\subsubsection{$k$-means}
+
+% Taken from ITI paper
+The $k$-means algorithm is perhaps the most well-known and used clustering algorithm.
+In order to apply the algorithm we first merge all the term-document matrices into a single matrix by stacking the individual term-document matrices and discarding nonaligned documents.
+$$D_{\mbox{Total}} = \begin{bmatrix}D^T_1 ,D^T_2, \cdots,D^T_\ell\ \end{bmatrix}^T$$
+such that the columns respect the alignment of the documents.
+Therefore, each document  is represented by a long vector indexed by the terms in all languages.
+These vectors determine the similarity on which the $k$-means is computed.
+The algorithm outputs a set of centroids that are then separated to form an aligned basis.
+As a final step the the projected documents are computed as the least squares solutions.
+
+In the implementation we never form the matrix $D_{\mbox{Total}}$ explicitly but rather implement implicit multiplication as $D_{\mbox{Total}} x = \sum_{i=1}^\ell D_i x$ and multiplication with transpose as $D_{\mbox{Total}}^T y = \sum_{i=1}^\ell D_i^T y_i,$ where $y$ is partitioned accordingly to $D_{\mbox{Total}}.$ That lowers memory requirement and compacts sparse indexing, leading to faster element access.
 
@@
 
+% Taken from ITI paper
+The next method is CL-LSI which is a variant of LSI~\cite{lsi} for more than one language.
+The method is based on computing the singular value decomposition of $D_{Total}$.
+Since the matrix can be large we can use an iterative method like the Lanczos~\cite{matrix_comp} algorithm with reorthogonalization to find the left singular vectors corresponding to the largest singular values.
+It turns out that Lanczos method converges slowly as the ratio of leading singular eigenvalues is close to one.
+Moreover Lanczos method is hard to parallelize.
+Instead we use randomized version of the singular value decomposition (SVD) described in [1] than can be viewed as block Lanczos method.
+That enables us to use parallelization and can speed up the computation considerably when multiprocessing is available.
+
+At the end we obtain low rank approximation $D_{\mbox{Total}} \approx U_k \Sigma_k V_k^T,$ where $U_k$ are basis vectors of interest and $\Sigma_k$ is truncated diagonal matrix of singular eigenvalues.
+Each column $u_i$ of $U_k$ consists of block vectors $u_i = \begin{bmatrix} u^{1^T}_i & \dots & u^{\ell^T}_i \end{bmatrix}^T.$ We do not normalize each block $j$  as this would destroy the low rank approximation.
+Each block $j$ is normalized $u^{j^1}_i = u^{j^1}_i/||u^{j^1}_i||$ to reduce the bias.
+We obtain the aligned reduced basis $U_{\textrm{aligned}} = \begin{bmatrix} {U^1}^T &\dots & {U^\ell}^T \end{bmatrix}^T,$ where $D_j =  U^j \Sigma_k V_k.$ The reduced language free representation for language $j$ and document $d$ is computed as the weighted least square solution $\Sigma_k^{-1} {U^j}^+ d,$ where $+$ denotes the pseudo inverse of a matrix that is is used because columns of $U^j$ do not form an orthogonal basis.
+The numerical implementation of the least squares is done by QR algorithm.
+
 \subsubsection{Refinement of LSI similar to MCCA}
 
-\subsubsection{Hub languages}
+\subsubsection{Regression}
+
+% Taken from Similarity paper
+We begin with a collection of documents in two languages\footnote{The extension to multiple language is  discussed in the Discussion section.}.
+The documents are represented via bag-of-words as vector spaces $\X$ and $\Y$.
+There are several possible choices of monolingual similarity functions.
+We consider functions of the form: $x'C_{\X\X}^i x$ for $i\in \mathbb{Z}$.
+For $i=0$, we recover the squared Euclidean distance and for $i=-1$, we get the Mahalanobis distance.
+
+As input, we take a choice of $i$ (choice of inner product) along with an alignment of $\X$ and $\Y$.
+The problem is to find a mapping which respects both monolingual similarity and measures cross-lingual similarity.
+In the first we reduce the problem to monolingual similarity.
+We find some mapping $f: \X \rightarrow \Y$.
+For a pair of elements $x\in\X$ and $y\in\Y$, we compute their similarity as the similarity of $f(x)$ and $y$ in $\Y$.
+
+The output of all the algorithms is a linear map (a matrix) denoted $B$, which computes the cross-lingual similarity as $x^T B y$ for $x\in\X$ and $y \in \Y$.
+The simplest choice for $B$ is $C_{\X\Y}$, the cross-covariance matrix.
+
+To find $B$ using regression, we compute the optimal regression matrix as $ C_{\Y \X} (C_{\X \X})^{-1}$.
+This corresponds to the mapping $f$ mentioned above.
+The inner product on the space $\Y$ induces $B$ for computing the similarities between elements in $\X$ and $\Y$: for $B$ it holds that $x^T B y = \langle f (x), y \rangle$.
+Therefore $B$ depends on the metric on $\Y$.
+For the inner product corresponding to $C_{\Y\Y}$, $B$ can be derived as $C_{\X \X}^{-1}C_{\X \Y}C_{\Y \Y}$.
+
+Note that in the case of regression, we have a choice of which direction we map: $\X\rightarrow\Y$ or $\Y\rightarrow\X$ except for Mahalanobis metric where the formula for $B$ is symmetric.
+Due to space constraints we only note here that CCA approximates the least-squares regression for this metric just as LSI approximates the cross-covariance matrix.
+
+% Taken from Pascal CL paper
+he \textbf{regression based approach} consists of two ingredients: a linear regression mapping $f : \RR^{n_i} \rightarrow \RR^{n_j}$,$$ f(x) := F\cdot x,$$ where $F \in \RR^{n_j \times n_j}$ and an inner product on the target space $\langle \cdot, \cdot \rangle_D$, which are used to define:
+$$s_{i,j}(x,y) := \langle f(x), y \rangle_D = f(x)^T D y = x^T F^T D y.$$
+Choices of $D$ included the standard inner product $I$, Mahalanobis inner product $C_{j,j}^{-1}$, which corresponds to whitening, and the covariance matrix $C_{j,j}$ (related to a co-occurrence).
+We can consider two regression mappings:
+\begin{itemize}
+\item least squares regression (LSQ), where $F = C_{j,i}C_{i,i}^{-1}$
+\item canonical correlation regression\cite{ccar} (CCAR), where $F = C_{j,j}^{-1}C{j,i}$.
+\end{itemize}
+
+\subsection{Hub languages}
+
+% Taken from XLite
+The first step in our method is to project $X_1, \ldots, X_m$ to lower dimensional spaces without destroying the cross-lingual structure.
+Treating the columns of $X_i$ as observation vectors sampled from an underlying distribution $\mathcal{X}_i \in V_i = \RR^{n_i}$, we can analyze the empirical cross-covariance matrices: $C_{i,j} = \frac{1}{|a(i,j)|-1 }\sum_{\ell \in a(i,j)} (X_i^{\ell} - c_i)\cdot (X_j^{\ell} - c_j)^T$, where $c_i = \frac{1}{a_i} \sum_{\ell \in a(i)}X_i^{\ell}$.
+By finding low rank approximations of $C_{i,j}$ we can identify the subspaces of $V_i$ and $V_j$ that are relevant for extracting linear patterns between $\mathcal{X}_i$ and $\mathcal{X}_j$.
+Let $X_1$ represent the hub language corpus matrix.
+The standard approach to finding the subspaces is to perform the singular value decomposition on the full $N \times N$ covariance matrix composed of blocks $C_{i,j}$.
+If $|a(i,j)|$ is small for many language pairs (as it is in the case of Wikipedia), then many empirical estimates $C_{i,j}$ are unreliable, which can result in overfitting.
+For this reason we perform the truncated singular value decomposition on the matrix $C = [C_{1,2}  \cdots  C_{1,m}] \approx U S V^T$, where where $U \in \RR^{n_1 \times k}, S \in \RR^{k \times k}, V \in \RR^{(\sum_{i=2}^m n_i) \times k}$.
+We split the matrix $V$ vertically in blocks with $n_2, \ldots, n_M$ rows to obtain a set of matrices $V_i$: $V = [V_2^T  \cdots  V_m^T]^T$.
+Note that columns of $U$ are orthogonal but columns in each $V_i$ are not (columns of V are orthogonal).
+Let $V_1 := U$.
+We proceed by reducing the dimensionality of each $X_i$ by setting: $Y_i = V_i^T \cdot X_i$, where $Y_i \in \RR^{k\times N}$.
+The step is similar to Cross-lingual Latent Semantic Indexing (CL-LSI) \cite{lsi}\cite{cl_lsi} which is less suitable due to a large amount of missing documents.
+Since singular value decomposition with missing values is a challenging problem and the alternative of replacing missing documents with zero vectors can result in degradation of performance.
+
+The second step involves solving a generalized version of canonical correlation analysis on the matrices $Y_i$ in order to find the mappings $P_i$.
+The approach is based on the sum of squares of correlations formulation by Kettenring \cite{Kettenring}, where we consider only correlations between pairs $(Y_1, Y_i), i >1$ due to the hub language problem characteristic.
+Let $D_{i,i} \in \RR^{k \times k}$ denote the empirical covariance and $D_{i,j}$ denote the empirical cross-covariance computed based on $Y_i$ and $Y_j$.
+We solve the following optimization problem:
+\begin{equation*}
+  \begin{aligned}
+    & \underset{w_i \in \RR^{k}}{\text{maximize}}
+    & & \sum_{i = 2}^m  \left(w_1^T D_{1,i} w_i \right)^2
+    & \text{subject to}
+    & & w_i^T D_{i,i} w_i = 1, \quad\forall i = 1,\ldots, m.
+\end{aligned}
+\end{equation*}
+By using Lagrangian multiplier techniques (we omit the derivation due to space constraints \todo{Add what was omitted}), the problem can be reformulated as the eigenvalue problem:$\left(\sum_{i=2}^m G_i G_i^T\right) \cdot V = \Lambda \cdot V,$ where $G_i = H_1^T D_{1,i} H_i$ and $H_i = Chol(D_{i,i})^{-1}$ where $Chol(\cdot)$ is the Cholesky decomposition: $X = Chol(X)^T Chol(X)$.
+The vectors $w_i$ can be reconstructed from the dominant eigenvector $v$, as: $w_1 = H_1 v$  and $w_i \propto H_i  G_i^T  v$ for $i >1$ ($w_i$ need to be appropriately normalized).
+Higher dimensional mappings $W_i$ are obtained in a similar way by setting the constraint $W_i^T D_{i,i} W_i = I$, where $I$ is the identity matrix.
+The constraint forces the columns of $W_i$ to be uncorrelated (orthogonal with respect to covariance matrix).
+$W_i$ can be extracted from a set of dominant eigenvectors of matrix $V$ according to eigenvalues $\Lambda$.
+The technique is related to a generalization of canonical correlation analysis (GCCA) by Carroll\cite{Carroll}, where an unknown group configuration variable is defined and objective is to maximize the sum of squared correlation between the group variable and the others.
+The problem can be reformulated as an eigenvalue problem.
+The difference lies in the fact that we set the unknown group configuration variable as the hub language, which simplifies the solution.
+The complexity of our method is $O(k^3)$, whereas solving the GCCA method scales as $O(N^3)$ where $N$ is the number of samples (see \cite{gifi}).
+Another issue with GCCA is that it cannot be directly applied to the case of missing documents.
+
+To summarize, we first reduced the dimensionality of our data to $k$-dimensional features and then found a new representation (via linear transformation) that maximizes directions of linear dependence between the languages.
+The final projections that enable mappings to a common space are defined as: $P_i(x) = W_i^T V_i^T x.$
 
@@
 \subsubsection{Link prediction}
+
+\subsubsection{Hub language}
+
+% Taken from XLite paper
+
+To investigate the empirical performance of our algorithm we will test it on a large-scale, real-world multilingual dataset that we extracted from Wikipedia  by  using so called 'inter-language' links as an alignment.
+We select a subset of Wikipedia languages containing three major languages, English--\emph{en} (hub language), Spanish--\emph{es}, Russian--\emph{ru}, and five minority (in the sense of Wikipedia sizes) languages, Slovenian--\emph{sl}, Piedmontese--\emph{pms}, Waray-Waray--\emph{war} (all with about 2 million native speakers), Creole--\emph{ht} (8 million native speakers), and Hindi--\emph{hi} (180 million native speakers).
+For preprocessing we remove the documents that contain less than 20 different words (stubs) and remove words occur in less than 50 documents as well as the top 100 most frequent words (in each language separately).
+We represent the documents as normalized TFIDF\cite{Salton88term-weightingapproaches} weighted vectors.
+These are languages with Wikipedia article sizes comparable to the minority languages.
+(since the number of speakers does not directly correlate with the number of Wikipedia articles).
+The prime hub candidate is the English language which is well aligned with all other Wikipedia languages, although the alignment quality varies quite a bit.
+Furthermore, we remove the documents that contain less than 20 different words and remove words that are too infrequent as well as the top 100 most frequent words in the vocabularies.
+Furthermore, we call the document consisting of less than 20 different words, a stub.
+This documents are typically garbage, the titles of the columns in the table, remains of the parsing process, or Wikipedia articles with very little or no information contained in one or two sentences.
+
+The evaluation is based on splitting the data into training and test sets (described later).
+On the training set, we perform the two step procedure to obtain the common document representation as a set of mappings $P_i$.
+A test set for each language pair, $test_{i,j} = \{(x_\ell,y_\ell) | \ell = 1:n(i,j)\} $, consists of comparable document pairs (linked Wikipedia pages), where $n(i,j)$ is the test set size.
+We evaluate the representation by measuring mate retrieval quality on the test sets: for each $\ell$, we rank the projected documents $P_j(y_1),\ldots, P_j(y_{n(i,j)})$ according to their similarity with $P_i(x_\ell)$ and compute the rank of the mate document $r(\ell) = rank(P_j(y_\ell))$.
+The final retrieval score (between -100 and 100) is computed as: $\frac{100}{n(i,j)} \cdot \sum_{\ell = 1}^{n(i,j)} \left( \frac{n(i,j) - r(\ell)}{n(i,j) -1} -0.5\right)$.
+A score that is less than 0 means that the method performs worse than random retrieval and a score of 100 indicates perfect mate retrieval.
+The mate retrieval results are included in Table \ref{table:retrieval}.
+
+We observe that the method performs well on all pairs between languages: \emph{en}, \emph{es}, \emph{ru}, \emph{sl}, where at least 50,000 training documents are available.
+We notice that taking $k = 500$ or $k = 1000$ multilingual topics usually results in similar performance, with some notable exceptions: in the case of (\emph{ht}, \emph{war}) the additional topics result in an increase in performance, as opposed to (\emph{ht},\emph{pms}) where performance drops, which suggests overfitting.
+The languages where the method performs poorly are \emph{ht} and \emph{war}, which can be explained by the quality of data (see Table \ref{table:rank} and explanation that follows).
+In case of \emph{pms}, we demonstrate that solid performance can be achieved for language pairs (\emph{pms}, \emph{sl}) and (\emph{pms}, \emph{hi}), where only 2000 training documents are shared between \emph{pms} and \emph{sl} and no training documents are available between \emph{pms} and \emph{hi}.
+Also observe that in the case of (\emph{pms}, \emph{ht}) the method still obtains a score of 62, even though training set intersection is zero and \emph{ht} data is corrupted, which we will show in the next paragraph.
+{
+\renewcommand\tabcolsep{3pt}
+\begin{table}[h!]
+\caption{Pairwise retrieval, 500 topics;1000 topics}\label{table:retrieval}
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
+\cline{1-9}
+&	en&	es&	ru&	sl&	hi&	war&	ht&	pms\\\cline{1-9}
+en&	    &	98~-~98&	95~-~97&	97~-~98&	82~-~84&	76~-~74&	53~-~55&	96~-~97\\
+\cline{1-9}
+es&	97~-~98&	&	94~-~96&	97~-~98&	85~-~84&	76~-~77&	56~-~57&	96~-~96\\
+\cline{1-9}
+ru&	96~-~97&	94~-~95&	&	97~-~97&	81~-~82&	73~-~74&	55~-~56&	96~-~96\\
+\cline{1-9}
+sl&	96~-~97&	95~-~95&	95~-~95&	&	91~-~91&	68~-~68&	59~-~69&	93~-~93\\
+\cline{1-9}
+hi&	81~-~82&	82~-~81&	80~-~80&	91~-~91&	&	68~-~67&	50~-~55&	87~-~86\\
+\cline{1-9}
+war&	68~-~63&	71~-~68&	72~-~71&	68~-~68&	66~-~62&	&	28~-~48&	24~-~21\\
+\cline{1-9}
+ht&	52~-~58&	63~-~66&	66~-~62&	61~-~71&	44~-~55&	16~-~50&	&	62~-~49\\
+\cline{1-9}
+pms&	95~-~96&	96~-~96&	94~-~94&	93~-~93&	85~-~85&	23~-~26&	66~-~54&	\\
+\cline{1-9}
+\end{tabular}
+\end{center}
+\end{table}
+}
+
+We now describe the selection of train and test sets.
+We select the test set documents as all multi-lingual documents with at least one nonempty alignment from the list: (\emph{hi}, \emph{ht}), (\emph{hi}, \emph{pms}), (\emph{war}, \emph{ht}), (\emph{war}, \emph{pms}).
+This guarantees that we cover all the languages.
+Moreover this test set is suitable for testing the retrieval thorough the hub as the chosen pairs have empty alignments.
+The remaining documents are used for training.
+In Table \ref{table:train_test}, we display the corresponding sizes of training and test documents for each language pair.
+The first row represents the size of the training sets used to construct the mappings in low dimensional language independent space using the English--\emph{en} as a hub.
+The diagonal elements represent number of the unique training documents and test documents in each language.
+
+{
+\renewcommand\tabcolsep{3pt}
+\begin{table}[h!]
+\caption{Pairwise training:test sizes (in thousands)}
+\label{table:train_test}
+{
+\small
+\begin{tabular}{c|c|c|c|c|c|c|c|c|}
+&	en&	es&	ru&	sl&	hi&	war&	ht&	pms\\\cline{1-9}
+en&	671~-~4.64&	463~-~4.29&	369~-~3.19&	50.3~-~2&	14.4~-~2.76&	8.58~-~2.41&	17~-~2.32&	16.6~-~2.67\\
+\cline{2-9}
+es&	\multicolumn{1}{c|}{}	&	463~-~4.29&	187~-~2.94&	28.2~-~1.96&	8.72~-~2.48&	6.88~-~2.4&	13.2~-~2&	 13.8~-~2.58\\
+\cline{3-9}
+ru&	\multicolumn{2}{c|}{}	&	369~-~3.19&	29.6~-~1.92&	9.16~-~2.68&	2.92~-~1.1&	3.23~-~2.2&	10.2~-~1.29\\
+\cline{4-9}
+sl&	\multicolumn{3}{c|}{}	&	50.3~-~2&	3.83~-~1.65&	1.23~-~0.986&	0.949~-~1.23&	1.85~-~0.988\\
+\cline{5-9}
+hi&	\multicolumn{4}{c|}{}	&	14.4~-~2.76&	0.579~-~0.76&	0.0~-~2.08&	0.0~-~0.796\\
+\cline{6-9}
+war&	\multicolumn{5}{c|}{}	&	8.58~-~2.41&	0.043~-~0.534&	0.0~-~1.97\\
+\cline{7-9}
+ht&	\multicolumn{6}{c|}{}	&	17~-~2.32&	0.0~-~0.355\\
+\cline{8-9}
+pms&	\multicolumn{7}{c|}{}	&	16.6~-~2.67\\
+\cline{9-9}
+\end{tabular}
+}
+\end{table}
+}
+
+We further inspect the properties of the training sets by roughly estimating the fraction \texttt{rank(A)/min(size(A))} for each training English matrix and its corresponding mate matrix.
+Ideally, these two fractions are approximately the same so  both aligned spaces should have reasonably similar dimensionality.
+We display these numbers as pairs in Table \ref{table:rank}.
+
+\begin{table}[h]
+\caption{Dimensionality drift}
+\label{table:rank}
+\begin{tabular}{|c|c|c|c|c|c|c|}
+\cline{1-7}
+(en, de)     &   (en, ru)     &   (en, sl)       &     (en, hi)&   (en, war)      &      (en, ht) &   (en, pms)\\
+\cline{1-7}
+(0.81, 0.89)   &  (0.8, 0.89)  &   (0.98, 0.96)    &    (1, 1)  &   (0.74, 0.56)  &      (1, 0.22)  &   (0.89, 0.38)\\
+\cline{1-7}
+\end{tabular}
+\end{table}
+
+It is clear that in the case of Creole language only at most $22\%$ documents are unique and suitable for the training.
+Though we removed the stub documents, many of remaining documents are almost the same, as the quality of some minor Wikipedias is low.
+This was confirmed for Creole, Waray-Waray, and Piedmontese language by manual inspection.
+The low quality documents correspond to templates about the year, person, town, etc.
+and contain very few unique words.
+
+We also have a problem with the quality of the test data.
+For example, if we look at test pair (\emph{war}, \emph{ht}) only 386/534 Waray-Waray test documents are unique but on other side almost all Creole test documents (523/534) are unique.
+This indicates a poor alignment which leads to poor performance.
 

### 2015-01-13 rupnik2015.tex
Added mCCA

@@
 \subsubsection{Refinement of LSI similar to MCCA}
+
+Finally, we test mCCA is specifically designed to consider data from multiple sources (in this case languages). Therefore, we do not merge the individual term-document matrices into one as in
+the other two methods.
+For each language (view), we estimate the pairwise correlation coefficients, then we try to find vectors which maximize the sum of all pairwise correlations over all languages.
+This can be written as the following optimization
+\begin{equation}
+\label{eq:opt1}
+\max_{w_1,\ldots, w_\ell} \sum_{i=1}^\ell\sum_{j=i+1}^\ell  \frac{w_i^T D_i D_j^T w_j}{\sqrt{ w_i^T D_i D_i^T w_i\vphantom{ D_j^T } }\sqrt{ w_j^T D_j D_j^T w_j}}
+\end{equation}
+However, this allows only for a 1-dimensional representation (i.e.
+one vector per language).
+Since we would like to allow for more vectors per language, we denote the double sum in Equation~\ref{eq:opt1} by $\mbox{mCCA}(w_1,\ldots,w_\ell)$.
+If $\mbox{corr}$ denotes the correlation, to find $M$ vectors, the optimization objective function  becomes
+\begin{equation}
+\max_{\substack{w_i^{(j)}\;;\;i=1,\ldots,\ell\\j=1,\ldots,M }}\sum_{s=1}^{\mbox{M}} \mbox{mCCA}\left(w_1^{(s)},\ldots,w_\ell^{(s)}\right),
+\qquad\mbox{s.t.} \qquad \mbox{corr} \left(w_i^{(s)},w_i^{(t)}\right) = 0
+\quad \forall s\neq t
+\end{equation}
+We require that the set of $M$ vectors we return for each language are uncorrelated (so that we do not get copies of a vector) which together maximize the pairwise correlation between languages.
 

### 2015-01-13 rupnik2015.tex
Added most of problem definition. Marked clearly what was copied and what is written.

@@
 \newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}
+\newcommand{\oldText}[1]{\textcolor{red}{#1:}\color{blue}}
 
@@
 \jairheading{x}{xxxx}{x-xx}{x/xx}{x/xx}
-\ShortHeadings{Cross-lingual Document Similarity and Event Detection}
+\ShortHeadings{Cross-lingeringual Document Similarity and Event Detection}
 {Rupnik, Muhi\v{c}, Leban, Fortuna, \v{S}kraba \& Grobelnik}
@@
 
-% taken from XLite paper
+{\oldText{Taken from XLite paper}
 Document retrieval is a well-established problem in data mining.
@@
 In the following section, we introduce our representation followed by our experiments, which also shed light on the structural properties of the multilingual Wikipedia corpus.
+}
 
@@
 
+Document similarity is an important component in techniques from text mining and natural language processing.
+Many techniques use the similarity as a black box, i.e., a kernel in Support Vector Machines.
+Comparison of documents (or other types of text snippets) is a well studied problem \todo{some references to prove that}.
+In this section we define document similarity in a cross-lingual setting, where the similarity function receives documents in different languages.
+We conclude the section by an introduction of two datasets which we used in this paper to learn cross-lingual similarity functions.
+
 \subsection{Problem definition}
 
-%Rough structure:
-% - Document Representation}
-% - Multi-language representation}
-% - Document Similarity
+% Document representation
 
-% taken from ITI paper
-Our data is a collection of documents in multiple languages along with an alignment with correspondences across languages.
-Individual documents are represented as vectors by using the standard vector space model where each term corresponds to a word or a phrase in a fixed vocabulary.
-An aligned corpus is represented by a set of matrices $D^J$ defined as $D^J := [d_1^J , \ldots , d_{m_j}^J ] \in \RR^{n_J \times m_J}$ where $m_J$ is the number of documents in language $J$ and $n_J$ is the dictionary size.
-Let $N = \sum_{J=1}^M n_J$, where $M$ is the number of languages.
-Each multi-lingual document of the aligned corpus is formed as a block vector $d_k = [ d_k^1; \dots; d_k^M] \in \RR^{N}$, and documents are indexed so that $d_k^{J_1}$ and $d_k^{J_2}$ is an aligned document pair for all $k, J_1, J_2$.
+Standard vector space model \cite{xxx} represents documents as vectors, where each team corresponds to word or phrase in a fixed vocabulary.
+More formally, document $d$ is represented by a vector $x \in \RR^n$, where $n$ corresponds to the size of the vocabulary, and vector elements $x_j$ correspond to the number of times term $j$ occurred in the document (also called \emph{term frequency} or $TF$).
+A common way of computing similarity between documents is \emph{cosine similarity},
+$$sim_{cos}(d_1, d_2) = \frac{<x_1, x_2>}{\|x_1\| \|x_2\|},$$
+where $x_1$ and $x_2$ correspond to the vector representations of $d_1$ and $d_2$ respectively.
+Cosine similarity, and other related approaches, assumes that the similarity is reflected in the overlap of words, and as such works only when the documents $d_1$ and $d_2$ are written in the same language.
 
-Our main interest are similarities between documents.
-We can measure the similarities between documents of the same language by using the cosine similarity.
-We note that given a transformation function between dictionaries, we could compute similarities between documents in different languages.
-However, bilingual-dictionaries are not available for many language pairs in the Wikipedia corpus and we will rely on document correspondences to compute document similarities across languages in the present paper.
-Note that this implies a correspondence between columns in each $D^K$ for all $K$.
+Terms in the vocabulary are generally not equally important in determining similarity between documents, and this can be addressed by preprocessing the documents.
+First, term frequency can be used to prune away infrequent terms.
+Rather than take a fixed number of top terms in each document we use an adaptive measure, to identify a threshold for a given monolingual corpus of $N$ documents.
+Let $f(n)$ be a map which returns numbers of terms appearing at least $n$ times in the corpus.
+We select the threshold as the maximal $n$ such that $f(n) \leq 6 \cdot 10^5$.
+The reason for this is that we wanted to support the maximal number of languages given available data, for many of which there are no available tools for stemming lemmatization.
+Using adaptive measure ensures that we get well balanced corpus and retain maximal number of words for languages with rather small available corpus but potentially rich vocabularies.
 
-Terms in the dictionary are generally not equally important in determining similarity between documents, so we must preprocess the documents.
-We first use term frequency ($TF$), to prune away infrequent terms.
-Rather than take a fixed number of top terms in each document we use an adaptive measure.
-
-Let $f(n)$ be a map which returns numbers of terms appearing at least $n$ times.
-For each language $k$ we find maximal $n$ such that $\max(10^5, N_k)\leq f(n) \leq 6\cdot 10^5,$ where $N_k$ is number of documents in complete dictionary of language $k.$ As we do not apply any stemming and lemmatization this ensures that we get well balanced corpus and retain at least of $10^5$ words for languages with rather small Wikipedias but potentially rich vocabularies.
 Once this pruning step is complete, we further re-weight the remaining terms.
 A term weight should correspond to the importance of the term for the given corpus.
-The common weighting scheme is called Term Frequency Inverse Document Frequency (TFIDF) weighting.
-An Inverse Document Frequency (IDF) weight for the dictionary term $j$ is defined as $w_j = \log( DF_j )$, where $DF_j$ is the number of documents in the corpus which contain term $j$.
-A document TFIDF vector is its original vector multiplied element-wise by the weights.
-The $j$-th element of a document vector is given by $ TF_j \log( DF_j )$.
+The common weighting scheme is called \emph{Term Frequency Inverse Document Frequency} ($TFIDF$) weighting.
+An \emph{Inverse Document Frequency} ($IDF$) weight for the dictionary term $j$ is defined as $\log( \frac{N}{DF_j} )$, where $DF_j$ is the number of documents in the corpus which contain term $j$.
+A document $TFIDF$ vector is its original vector multiplied element-wise by the weights.
+The $j$-th element of a document vector is given by $x_j = TF_j \log( \frac{N}{DF_j} )$.
 Finally, we re-normalize each vector to have Euclidean norm equal to 1.
 
-% taken from XLite paper
-We begin by introducing some notation.
-A \emph{multilingual document} $d = (u_1,\ldots u_m)$ is a collection of $m$ documents on the same topic (comparable), where $u_i$ is the document in language $i$ which can be an empty document (missing resource) and $d$ must contain at least two nonempty documents.
+% Multi-language dataset
+
+The cross-lingual similarity model presented in this paper is based on comparable corpora.
+That is, a corpus of documents in multiple languages, with alignment between documents that are of the same topic, or even a rough translation of each other.
+Examples of comparable corpora include Wikipedia, where a specific entry can be described in multiple languages (e.g.
+"Berlin" is currently described in 222 languages), or news articles, where the same event can be described by newspapers in several languages.
+
+More formally, \emph{multilingual document} $d = (u_1,\ldots u_m)$ is a tuple of $m$ documents on the same topic (comparable), where $u_i$ is the document written in language $i$.
+Note that individual document $u_i$ can be an empty document (missing resource) and each $d$ must contain at least two nonempty documents.
 A comparable corpus $D = {d_1, \ldots, d_N}$ is a collection of multilingual documents.
-By using the standard vector space model (bag of words) we can represent $D$ as a set of $m$ matrices $X_1,\ldots,X_m$, where $X_i \in \RR^{n_i \times N}$ is the corpus matrix corresponding to the $i$-th language, where $n_i$ is the vocabulary size.
-Let $X_i^{\ell}$ denote the $\ell$-th column of matrix $X_i$.
-A \emph{hub language} is a language with a high proportion of non-empty documents in $D$, which in case of Wikipedia is English.
-Let $a(i,j)$ denote the index set of multilingual documents $d$ that contain non-empty $u_i$ and $u_j$ and $a(i)$ denote the index set of multilingual documents that contain $u_i$.
+By using the vector space model we can represent $D$ as a set of $m$ matrices $X_1,\ldots,X_m$, where $X_i \in \RR^{n_i \times N}$ is the matrix corresponding to the language $i$ and $n_i$ is the vocabulary size of language $i$.
+Furthermore, let $X_i^{\ell}$ denote the $\ell$-th column of matrix $X_i$, let $a(i,j)$ denote the index set of all multilingual documents $d$ that contain non-empty documents $u_i$ and $u_j$, and let $a(i)$ denote the index set of all multilingual documents that contain $u_i$.
+
+% Document similarity
+
+\todo{Describe document similarity in cross-lingual setting}
 
@@
 
-% Taken from ITI paper
+{\oldText{Taken from ITI paper}
 To investigate the empirical performance of the low rank approximations we will test the algorithms on a large-scale, real-world multilingual dataset that we extracted from Wikipedia by using inter-language links as an alignment.
@@
 This matrix enables us to build an alignment across all Wikipedia languages.
+}
 
@@
 
-% Taken from ITI paper
+{\oldText{Taken from ITI paper}
 The $k$-means algorithm is perhaps the most well-known and used clustering algorithm.
@@
 In the implementation we never form the matrix $D_{\mbox{Total}}$ explicitly but rather implement implicit multiplication as $D_{\mbox{Total}} x = \sum_{i=1}^\ell D_i x$ and multiplication with transpose as $D_{\mbox{Total}}^T y = \sum_{i=1}^\ell D_i^T y_i,$ where $y$ is partitioned accordingly to $D_{\mbox{Total}}.$ That lowers memory requirement and compacts sparse indexing, leading to faster element access.
+}
 
@@
 
-% Taken from ITI paper
+{\oldText{Taken from ITI paper}
 The next method is CL-LSI which is a variant of LSI~\cite{lsi} for more than one language.
@@
 The numerical implementation of the least squares is done by QR algorithm.
+}
 
@@
 
+{\oldText{Taken from Low-rank paper}
 Finally, we test mCCA is specifically designed to consider data from multiple sources (in this case languages). Therefore, we do not merge the individual term-document matrices into one as in
@@
 We require that the set of $M$ vectors we return for each language are uncorrelated (so that we do not get copies of a vector) which together maximize the pairwise correlation between languages.
+}
 
@@
 
-% Taken from Similarity paper
+{\oldText{Taken from Similarity paper}
 We begin with a collection of documents in two languages\footnote{The extension to multiple language is  discussed in the Discussion section.}.
@@
 Due to space constraints we only note here that CCA approximates the least-squares regression for this metric just as LSI approximates the cross-covariance matrix.
+}
 
-% Taken from Pascal CL paper
+{\oldText{Taken from Pascal CL paper}
 he \textbf{regression based approach} consists of two ingredients: a linear regression mapping $f : \RR^{n_i} \rightarrow \RR^{n_j}$,$$ f(x) := F\cdot x,$$ where $F \in \RR^{n_j \times n_j}$ and an inner product on the target space $\langle \cdot, \cdot \rangle_D$, which are used to define:
@@
 \end{itemize}
+}
 
@@
 
-% Taken from XLite
+A \emph{hub language} is a language with a high proportion of non-empty documents in $D$, which in case of Wikipedia is English.
+
+{\oldText{Taken from XLite}
 The first step in our method is to project $X_1, \ldots, X_m$ to lower dimensional spaces without destroying the cross-lingual structure.
@@
 The final projections that enable mappings to a common space are defined as: $P_i(x) = W_i^T V_i^T x.$
+}
 
@@
 
-% Taken from XLite paper
-
+{\oldText{Taken from XLite paper}
 To investigate the empirical performance of our algorithm we will test it on a large-scale, real-world multilingual dataset that we extracted from Wikipedia  by  using so called 'inter-language' links as an alignment.
@@
 This indicates a poor alignment which leads to poor performance.
+}
 

### 2015-02-17 rupnik2015.tex
update

@@
 
-% Document representation
+\textbf{Document representation.}
 
-Standard vector space model \cite{xxx} represents documents as vectors, where each team corresponds to word or phrase in a fixed vocabulary.
-More formally, document $d$ is represented by a vector $x \in \RR^n$, where $n$ corresponds to the size of the vocabulary, and vector elements $x_j$ correspond to the number of times term $j$ occurred in the document (also called \emph{term frequency} or $TF$).
+Standard vector space model \cite{xxx} represents documents as vectors, where each term corresponds to word or phrase in a fixed vocabulary.
+More formally, document $d$ is represented by a vector $x \in \RR^n$, where $n$ corresponds to the size of the vocabulary, and vector elements $x_k$ correspond to the number of times term $k$ occurred in the document, also called \emph{term frequency} or $TF_k(d)$.
+
+We also used a term re-weighting scheme that adjusts for the fact that some words occur more frequently in general.
+A term weight should correspond to the importance of the term for the given corpus.
+The common weighting scheme is called \emph{Term Frequency Inverse Document Frequency} ($TFIDF$) weighting.
+An \emph{Inverse Document Frequency} ($IDF$) weight for the dictionary term $k$ is defined as $\log( \frac{N}{DF_k} )$, where $DF_k$ is the number of documents in the corpus which contain term $k$.
+A document $TFIDF$ vector is its original vector multiplied element-wise by the weights.
+
+The $TFIDF$ weighted vector space model document representation corresponds to a map $\phi : \text{text} \rightarrow \RR^n$ defined by:
+$$\phi(d)_k = {TF}_k(d) \log( \frac{N}{{DF}_k}).$$
+
+\textbf {Mono-lingual similarity.}
 A common way of computing similarity between documents is \emph{cosine similarity},
-$$sim_{cos}(d_1, d_2) = \frac{<x_1, x_2>}{\|x_1\| \|x_2\|},$$
-where $x_1$ and $x_2$ correspond to the vector representations of $d_1$ and $d_2$ respectively.
+$$sim(d_1, d_2) = \frac{\langle \phi(d_1), \phi(d_2)\rangle}{\|\phi(d_1)\| \|\phi(d_2)\|},$$
+where $\langle \cdot,\cdot \rangle$ and $\|\cdot\|$ are standard inner product and Euclidean norm.
 Cosine similarity, and other related approaches, assumes that the similarity is reflected in the overlap of words, and as such works only when the documents $d_1$ and $d_2$ are written in the same language.
 
-Terms in the vocabulary are generally not equally important in determining similarity between documents, and this can be addressed by preprocessing the documents.
-First, term frequency can be used to prune away infrequent terms.
-Rather than take a fixed number of top terms in each document we use an adaptive measure, to identify a threshold for a given monolingual corpus of $N$ documents.
-Let $f(n)$ be a map which returns numbers of terms appearing at least $n$ times in the corpus.
-We select the threshold as the maximal $n$ such that $f(n) \leq 6 \cdot 10^5$.
-The reason for this is that we wanted to support the maximal number of languages given available data, for many of which there are no available tools for stemming lemmatization.
-Using adaptive measure ensures that we get well balanced corpus and retain maximal number of words for languages with rather small available corpus but potentially rich vocabularies.
-
-Once this pruning step is complete, we further re-weight the remaining terms.
-A term weight should correspond to the importance of the term for the given corpus.
-The common weighting scheme is called \emph{Term Frequency Inverse Document Frequency} ($TFIDF$) weighting.
-An \emph{Inverse Document Frequency} ($IDF$) weight for the dictionary term $j$ is defined as $\log( \frac{N}{DF_j} )$, where $DF_j$ is the number of documents in the corpus which contain term $j$.
-A document $TFIDF$ vector is its original vector multiplied element-wise by the weights.
-The $j$-th element of a document vector is given by $x_j = TF_j \log( \frac{N}{DF_j} )$.
-Finally, we re-normalize each vector to have Euclidean norm equal to 1.
-
-% Multi-language dataset
-
-The cross-lingual similarity model presented in this paper is based on comparable corpora.
-That is, a corpus of documents in multiple languages, with alignment between documents that are of the same topic, or even a rough translation of each other.
-Examples of comparable corpora include Wikipedia, where a specific entry can be described in multiple languages (e.g.
-"Berlin" is currently described in 222 languages), or news articles, where the same event can be described by newspapers in several languages.
-
-More formally, \emph{multilingual document} $d = (u_1,\ldots u_m)$ is a tuple of $m$ documents on the same topic (comparable), where $u_i$ is the document written in language $i$.
-Note that individual document $u_i$ can be an empty document (missing resource) and each $d$ must contain at least two nonempty documents.
-A comparable corpus $D = {d_1, \ldots, d_N}$ is a collection of multilingual documents.
-By using the vector space model we can represent $D$ as a set of $m$ matrices $X_1,\ldots,X_m$, where $X_i \in \RR^{n_i \times N}$ is the matrix corresponding to the language $i$ and $n_i$ is the vocabulary size of language $i$.
-Furthermore, let $X_i^{\ell}$ denote the $\ell$-th column of matrix $X_i$, let $a(i,j)$ denote the index set of all multilingual documents $d$ that contain non-empty documents $u_i$ and $u_j$, and let $a(i)$ denote the index set of all multilingual documents that contain $u_i$.
-
-% Document similarity
-
-\todo{Describe document similarity in cross-lingual setting}
+\textbf {Cross-lingual similarity.}
+Processing a multilingual dataset results in several vector spaces with varying dimensionality, one for each language.
+The dimensionality of the vector space corresponding to the $i$-th language is denoted by $n_i$ and and the vector space model mapping is denoted by $\phi_i : \text{text} \rightarrow \RR^{n_i}$.
+The similarity between documents in language $i$ and language $j$ is defined as a bilinear operator represented as a matrix $S_{i,j} \in \RR^{n_i \times n_j}$:
+$$sim_{i,j}(d_1, d_2) = \frac{ \langle \phi_i (d_1), S_{i,j} \phi_j (d_2) \rangle }{\|\phi_i(d_1)\| \|\phi_j(d_2)\|},$$
+where $d_1$ and $d_2$ are documents written in the $i$-th and $j$-th language respectively.
+If the the maximal singular value of $S_{i,j}$ is bounded by $1$, then the similarity scores will lie on the interval $[-1, 1]$.
+In section \ref{sec:models} we will describe some approaches to computing $S_{i,j}$ given training data.
 
@@
 
-\section{Multilingual Latent Factor Models}
+\section{Multilingual Latent Factor Models}\label{sec:models}
+
+\subsection{Multilingual dataset}
+
+The cross-lingual similarity models presented in this paper are based on comparable corpora.
+That is, a corpus of documents in multiple languages, with alignment between documents that are of the same topic, or even a rough translation of each other.
+Wikipedia is an example of a comparable corpus, where a specific entry can be described in multiple languages (e.g.
+"Berlin" is currently described in 222 languages).
+News articles represent another example, where the same event can be described by newspapers in several languages.
+
+More formally, \emph{multilingual document} $d = (u_1,\ldots u_m)$ is a tuple of $m$ documents on the same topic (comparable), where $u_i$ is the document written in language $i$.
+Note that individual document $u_i$ can be an empty document (missing resource) and each $d$ must contain at least two nonempty documents.
+A comparable corpus $D = {d_1, \ldots, d_N}$ is a collection of multilingual documents.
+By using the vector space model we can represent $D$ as a set of $m$ matrices $X_1,\ldots,X_m$, where $X_i \in \RR^{n_i \times N}$ is the matrix corresponding to the language $i$ and $n_i$ is the vocabulary size of language $i$.
+Furthermore, let $X_i^{\ell}$ denote the $\ell$-th column of matrix $X_i$.
 
@@
 
-A \emph{hub language} is a language with a high proportion of non-empty documents in $D$, which in case of Wikipedia is English.
+A \emph{hub language} is a language with a high proportion of non-empty documents in $D$, which in case of Wikipedia is English. We use the following notation to define subsets
+of the multilingual comparable corpus: let $a(i,j)$ denote the index set of all multilingual documents with non-missing data for the $i$-th and $j$-th language:  $a(i,j) = \left\{ d = (u_1,...,u_m) | u_i \neq \emptyset, u_j \neq \emptyset \right\}$ , and let $a(i)$ denote the index set of all multilingual documents with non missing data for the $i$-th language.
 
 {\oldText{Taken from XLite}
+
 The first step in our method is to project $X_1, \ldots, X_m$ to lower dimensional spaces without destroying the cross-lingual structure.
@@
 \vskip 0.2in
-\bibliography{rupnik2015}
+%\bibliography{rupnik2015}
 \bibliographystyle{theapa}

### 2015-02-18 rupnik2015.tex
- added content for cross-lingual event linking

--- 2015-01-13 rupnik2015.tex
@@
 
+One of the use cases where article linking information can be applied is also in an online system for detection of world events, called Event Registry\cite{}.
+Event Registry is a repository of events, where events are automatically identified by analyzing news articles.
+The term event is vague and ambiguous, but for the practical purposes we define it as "any significant happening that is being reported about in the media".
+Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015.
+Events such as these are covered by many articles and the question is how to find all the articles in different languages that are describing a single event.
+
+Event Registry works by collecting and analyzing news articles that are published by news outlets in different languages all over the world.
+Collected articles are first semantically annotated by identifying in them mentions of relevant concepts -- either entities or important keywords.
+The disambiguated and entity linking of the concepts is done using Wikipedia as the main knowledge base.
+The key step in Event Registry is then to apply an online clustering algorithm on the articles in order to identify groups of articles that are discussing the same event.
+For each new article, the clustering algorithm determines if the article should be assigned to some existing cluster or into a new cluster.
+The underlying assumption is that articles that are describing the same event are similar enough and will therefore be put into the same cluster.
+
+In short, the article clustering works as follows (details are available in \cite{}).
+First we create a new document by joining the title and the body of the article.
+We then tokenize the document, remove the stop words and stem the words.
+The remaining words are represented as a vector in the vector-space model.
+Each value in the vector is normalized using the TF-IDF approach, defined as $TFIDF(t,d) = TF(t,d) * IDF(t)$.
+$TF(t,d)$ represents the \emph{term frequency} (number of times the term $t$ appears in the document $d$), and $IDF(t) = log(N / DF(t))$ is the \emph{inverse document frequency} of term $t$.
+$IDF(t)$ is computed by comparing the number of documents mentioning the term $t$ ($DF(t)$) with the total number of documents $N$.
+After normalizing the vector, we use it to find the most similar existing cluster.
+To be able to compare the vector with a cluster we first need a vector representation for each of the clusters.
+This is done by computing a centroid vector of the cluster by simply averaging the vectors of the articles in the cluster.
+By computing the cosine similarity of the article vector with each of the centroid vectors we then find the cluster with the highest similarity.
+If the highest similarity is above the selected threshold (0.4 in our experiments), the article is assigned to the corresponding cluster, otherwise a new cluster is created, initially containing only the single article.
+
+Since articles about an event are commonly written only for a short period of time, we remove clusters once the oldest article in the cluster becomes more than 4 days old.
+This housekeeping mechanism prevents the clustering from becoming slow and also ensures that articles are not assigned to obsolete clusters.
+
+In order to treat the cluster as describing an event in Event Registry, the cluster has to grow above a certain number of articles (the number depends on the language).
+This condition ensures that only content that receives a sufficient coverage in the media is treated as an event.
+Once a new event is identified, a new unique ID is assigned to it and the main information about the event is automatically extracted by analyzing the articles assigned to it.
+The extracted information includes properties such as the date of the event, the location, who is involved in it, what the event is about, etc.
+
+The clustering of articles is done for each language separately -- English articles are, for example, clustered separately from Spanish articles.
+When events are identified, they will therefore be identified from clusters in each language separately.
+This has an undesirable consequence for Event Registry where the same actual event will be represented with two or more event IDs, and each event will only contain coverage in a single language.
+To avoid this problem we want to exploit the available information about the similarity of the articles in a cluster with articles in other languages.
+Using this information we hope to be able to connect clusters in different languages and represent them using a single event ID.
+
 \subsection{Problem definition}
+
+Formally, the problem of identifying the clusters describing the same event can be stated as follows. Given the
 
@@
 \subsection{Evaluation}
+
+evaluate accuracy on different language pairs only. compute accuracies using the CCA features only, using concept features only, or using all features.
 

### 2015-02-19 rupnik2015.tex
update

--- 2015-02-17 rupnik2015.tex
@@
 By using the vector space model we can represent $D$ as a set of $m$ matrices $X_1,\ldots,X_m$, where $X_i \in \RR^{n_i \times N}$ is the matrix corresponding to the language $i$ and $n_i$ is the vocabulary size of language $i$.
-Furthermore, let $X_i^{\ell}$ denote the $\ell$-th column of matrix $X_i$.
+Furthermore, let $X_i^{\ell}$ denote the $\ell$-th column of matrix $X_i$ and the matrices respect the document alignment - the vector $X_i^\ell$ corresponds to the TFIDF vector of the $i$-th component of multilingual document $d_\ell$.
+We use $N$ to denote the total row dimension of $X$, i.e.
+$N:= \sum_{i=1}^m n_i$.
 
@@
 
-{\oldText{Taken from ITI paper}
 The $k$-means algorithm is perhaps the most well-known and used clustering algorithm.
-In order to apply the algorithm we first merge all the term-document matrices into a single matrix by stacking the individual term-document matrices and discarding nonaligned documents.
-$$D_{\mbox{Total}} = \begin{bmatrix}D^T_1 ,D^T_2, \cdots,D^T_\ell\ \end{bmatrix}^T$$
-such that the columns respect the alignment of the documents.
+In order to apply the algorithm we first merge all the term-document matrices into a single matrix $X$ by stacking the individual term-document matrices:
+$$X := \begin{bmatrix}X_1^T ,X_2^T, \cdots, X_m^T \end{bmatrix}^T,$$
+such that the columns respect the alignment of the documents (here MATLAB notation for concatenating matrices is used).
 Therefore, each document  is represented by a long vector indexed by the terms in all languages.
-These vectors determine the similarity on which the $k$-means is computed.
-The algorithm outputs a set of centroids that are then separated to form an aligned basis.
-As a final step the the projected documents are computed as the least squares solutions.
 
-In the implementation we never form the matrix $D_{\mbox{Total}}$ explicitly but rather implement implicit multiplication as $D_{\mbox{Total}} x = \sum_{i=1}^\ell D_i x$ and multiplication with transpose as $D_{\mbox{Total}}^T y = \sum_{i=1}^\ell D_i^T y_i,$ where $y$ is partitioned accordingly to $D_{\mbox{Total}}.$ That lowers memory requirement and compacts sparse indexing, leading to faster element access.
-}
+We then run the $k$-means algorithm \cite{kmeans} and obtain a set of $k$ centroid vectors, which form a matrix: $C := \begin{bmatrix}C^1, \ldots, C^k \end{bmatrix}$.
+The centroid matrix can be split vertically into $m$ blocks: $$C = [C_1^T \cdots C_m^T]^T.$$
+
+Each matrix $C_i$ represents a vector space basis and can be used to map points in $\RR^{n_i}$ into a $k$-dimensional space, where the coordinates of a vector $x \in \RR^{n_i}$ are expressed as: $$(C_i^T C_i)^{-1} C_i^T x_i.$$
+
+The resulting matrix (when appropriately scaled to have unit norm) for similarity computation between language $i$ and language $j$ is defined as:
+$$C_i(C_i^T C_i)^{-1} (C_j^T C_j)^{-1} C_j.$$
+
+%% We could skip this par
+In the implementation of $k$-means clustering we never form the matrix $X$ explicitly but rather implement implicit multiplication as $X x = \sum_{i=1}^\ell X_i x$ and multiplication with transpose as $X^T y = \sum_{i=1}^\ell X_i^T y_i,$ where $y$ is partitioned accordingly to $X$.
+That lowers memory requirement and compacts sparse indexing, leading to faster element access.
 
@@
 
-{\oldText{Taken from ITI paper}
 The next method is CL-LSI which is a variant of LSI~\cite{lsi} for more than one language.
-The method is based on computing the singular value decomposition of $D_{Total}$.
-Since the matrix can be large we can use an iterative method like the Lanczos~\cite{matrix_comp} algorithm with reorthogonalization to find the left singular vectors corresponding to the largest singular values.
-It turns out that Lanczos method converges slowly as the ratio of leading singular eigenvalues is close to one.
-Moreover Lanczos method is hard to parallelize.
-Instead we use randomized version of the singular value decomposition (SVD) described in [1] than can be viewed as block Lanczos method.
-That enables us to use parallelization and can speed up the computation considerably when multiprocessing is available.
+The method is based on computing a truncated singular value decomposition of $X \approx U S V^T$.
+Since the matrix can be large we can use an iterative method like the Lanczos~\cite{matrix_comp} algorithm with reorthogonalization to find the left singular vectors (columns of $U$) corresponding to the largest singular values.
+It turns out that the Lanczos method converges slowly as the gap between the leading singular values is small.
+Moreover, the Lanczos method is hard to parallelize.
+Instead we use a randomized version of the singular value decomposition (SVD) described in \ref{randomizedSVD} that can be viewed as a block Lanczos method.
+That enables us to use parallelization and speeds up the computation considerably.
 
-At the end we obtain low rank approximation $D_{\mbox{Total}} \approx U_k \Sigma_k V_k^T,$ where $U_k$ are basis vectors of interest and $\Sigma_k$ is truncated diagonal matrix of singular eigenvalues.
+The cross-lingual similarity functions are based on a rank-$k$ truncated SVD: $X \approx U \Sigma V^T,$ where $U \in \RR^{N \times k}$ are basis vectors of interest and $\Sigma \in \RR^{k \times k}$ is truncated diagonal matrix of singular eigenvalues.
+
+{\color{red}
  Each column $u_i$ of $U_k$ consists of block vectors $u_i = \begin{bmatrix} u^{1^T}_i & \dots & u^{\ell^T}_i \end{bmatrix}^T.$ We do not normalize each block $j$  as this would destroy the low rank approximation.
 Each block $j$ is normalized $u^{j^1}_i = u^{j^1}_i/||u^{j^1}_i||$ to reduce the bias.
+
 We obtain the aligned reduced basis $U_{\textrm{aligned}} = \begin{bmatrix} {U^1}^T &\dots & {U^\ell}^T \end{bmatrix}^T,$ where $D_j =  U^j \Sigma_k V_k.$ The reduced language free representation for language $j$ and document $d$ is computed as the weighted least square solution $\Sigma_k^{-1} {U^j}^+ d,$ where $+$ denotes the pseudo inverse of a matrix that is is used because columns of $U^j$ do not form an orthogonal basis.

### 2015-02-19 rupnik2015.tex
- more content on ER evaluation

--- 2015-02-18 rupnik2015.tex
@@
 \usepackage{amssymb}
+
+\usepackage[]{algorithm2e}
 
@@
 
-Formally, the problem of identifying the clusters describing the same event can be stated as follows. Given the
+Formally, the problem of identifying the clusters describing the same event can be formalized as follows.
+Each article $a_i$ is written in a language $l$, where $l \in \{A,B,...,H\}$.
+For each language $l$, a set of clusters $C_l$ is generated and maintained that only contains articles in language $l$.
+For any cluster $c_i \in C_{l1}$, the task is then to identify other clusters $c_j \in C_{l2}$ (where $l1 \neq l2$) that are describing the same event as $c_i$.
+We will refer to cluster pairs $c_i$ and $c_j$ that are describing the same event as equivalent clusters.
+An additional requirement of the task is that identification of the equivalent clusters should not be done exhaustively -- for a cluster $c_i$ we wish to avoid testing all other clusters $c_j$.
+Performing exhaustive comparison for each cluster $c_i$ would result in $O(N^2)$ tests, where $N$ is the number of all clusters, which is not feasible when the number of clusters is in tens of thousands.
 
 \subsection{Algorithm}
+
+In order to identify clusters $c_j$ that are equivalent to cluster $c_i$ we have developed a procedure that consists of two main tasks.
+The first task is to efficiently identify a limited set of clusters that are potential candidates for being equivalent to $c_i$.
+The second task is then about performing a detailed analysis of the candidates and finding the ones that are equivalent to $c_i$.
+
+The details of the first task are described in Algorithm~\ref{cluster_merge_algo1}.
+The algorithms starts by individually inspecting each article $a_i$ in the cluster $c_i$.
+Using the CCA it then identifies 10 most similar articles to $a_i$ in each language $l \in L$.
+For each identified similar article $a_j$ we then identify cluster $c_j$ to which the article belongs to and add the cluster to the set of candidates that should be more closely inspected.
+The assumption made by this algorithm is that if two articles in different languages are describing the same event, the CCA similarity between the documents will be higher than when comparing two articles about different events.
+Since the CCA similarity is only an estimate, the assumption does not always hold.
+Due to data redundancy, however, we do not consider this to be an issue.
+Since we are testing each article in the cluster and compute 10 most similar articles for each article, we are very likely to identify among the candidates also the clusters that are equivalent to cluster $c_i$.
+
+\begin{algorithm}[t!]
+\SetKwInOut{Input}{input}\SetKwInOut{Output}{output}
+\Input{test cluster $c_i$,\\a set of clusters $C_l$ for each language $l \in L$}
+\Output{a set of clusters $C$ that are potentially equivalent to $c_i$}
+$C = \{\}$\;
+\For{article $a_i \in c_i$} {
+    \For{language $l \in L$} {
+        \tcc{use CCA to find 10 most similar articles to article $a_i$ in language $l$}
+        $SimArt = getCCASimArts(a_i, l)$\;
+        \For{article $a_j \in SimArt$} {
+            \tcc{find cluster $c_j$ to which article $a_j$ is assigned to}
+            $c_j := c ; c \in C_l, a_j \in c$\;
+            \tcc{add cluster $c_j$ to the set of candidates $C$}
+            $C \leftarrow c_j$\;
+        }
+    }
+}
+\caption{Algorithm for identifying candidate clusters $C$ that are potentially equivalent to $c_i$}
+\label{cluster_merge_algo1}
+\end{algorithm}
+
+The second part of the procedure is described in the Algorithm~\ref{cluster_merge_algo2}.
+The goal of this task is to determine which (if any) of the candidate clusters are equivalent to $c_i$.
+To solve the task we represent it as a supervised learning problem.
+For each candidate cluster $c_j \in C'$ we compute a vector of learning features that should be indicative of whether the $c_i$ and $c_j$ are equivalent or not.
+The details about these features will be described in the next section.
+Once the features are computed, we classify the example with a classification model that was trained using a training dataset of hand-labeled examples.
+If the model predicts that tested clusters are equivalent, the $c_j$ is added to the set $C'$.
+
+The learning dataset used to train a classification model consists of cluster pairs $c_i$ and $c_j$ for which a human annotator provided feedback whether the clusters are equivalent or not.
+The dataset contains XYZ examples, of which XYZ are equivalent clusters pairs and XYZ are not.
+To train a model, the vector of features was first extracted for each learning example in the same way as described in Algorithm~\ref{cluster_merge_algo2}.
+A linear SVM method was then used to train the classifier $M$.
+
+\begin{algorithm}[tb!]
+\SetKwInOut{Input}{input}\SetKwInOut{Output}{output}
+\Input{test cluster $c_i$,\\
+a set of candidate clusters $C$ that are potentially equivalent to $c_i$\\
+classification model $M$ trained on the hand-labeled examples}
+\Output{a set of clusters $C'$ that are equivalent to $c_i$}
+
+$C' = \{\}$\;
+\For{cluster $c_j \in C$} {
+    \tcc{extract a vector of features for cluster pair $c_i$ and $c_j$}
+    $\vec{v} = f(c_i, c_j)$\;
+    \tcc{classify the feature vector $\vec{v}$ using the model $M$}
+    $pred = M(\vec{v})$\;
+    \If{$pred$ = true} {
+        $C' \leftarrow c_j$
+    }
+}
+\caption{Algorithm for identifying clusters $C'$ that are equivalent to cluster $c_i$}
+\label{cluster_merge_algo2}
+\end{algorithm}
+
+\subsubsection{Learning features}
 

### 2015-02-21 rupnik2015.tex
- added learning features

@@
 Event Registry is a repository of events, where events are automatically identified by analyzing news articles.
-The term event is vague and ambiguous, but for the practical purposes we define it as "any significant happening that is being reported about in the media".
+The term event is vague and ambiguous, but for the practical purposes we define it as ``any significant happening that is being reported about in the media''.
 Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015.
@@
 
+In order to use a supervised approach to determine if two clusters $c_i$ and $c_j$ are equivalent we need to extract from the cluster pair as many informative features as possible.
+Based on the availability of the data, we have extracted the following groups of features:
+
+\begin{itemize}
+\item {\bf CCA related features}.
+CCA was already used to obtain candidates for the equivalent clusters.
+Additionally, it can also be used to provide valuable learning features.
+There are two main features that we extract using CCA -- \texttt{linkCount} and \texttt{avgSimScore}.
+\texttt{linkCount} is the number of times an article $a_i \in c_i$ has as one of the 10 most similar articles an article $a_j \in c_j$.
+In other words, it is the number of times an article from $c_i$ has a very similar article, which is in $c_j$.
+Beside the number of links between the two clusters we can also take into account the similarity scores of the links.
+The \texttt{avgSimScore} is the feature that represents the average similarity score of the links between the two clusters.
+
+\item {\bf Concept related features}.
+As mentioned before, articles that are imported into Event Registry are first semantically annotated by linking mentioned entities and keywords to the corresponding Wikipedia pages.
+Whenever Barack Obama is, for example, mentioned in the article, the article is annotated with a link to his Wikipedia page (\emph{http://en.wikipedia.org/wiki/Barack\_Obama}).
+In the same way all mentions of people, locations, organizations and even ordinary keywords (e.g.
+bank, tax, ebola, plane, company) are annotated.
+Although the Spanish article about Obama will be annotated with his Spanish version of the Wikipedia page, we can almost in all cases normalize the Wikipedia links to their English versions.
+This can be done since Wikipedia itself provides information regarding which pages in different languages represent the same concept/entity.
+Using this approach, the word ``avi\'on'' in a Spanish article will be annotated with the same concept as the word ``plane'' in an English article.
+Although articles are in different languages, the annotations can therefore provide a shared vocabulary that can be used to compare articles/clusters.
+By analyzing all articles in clusters $c_i$ and $c_j$ we can identify the most relevant entities and keywords for each cluster.
+Additionally we can also assign weights to the concepts based on how frequently they occur in the articles in the cluster.
+From the list of relevant concepts and corresponding weights we can now define a number of features.
+\texttt{entityCosSim} represents the cosine similarity on the vector of entities mentioned in both clusters.
+Similarly, \texttt{keywordCosSim} represents the cosine similarity on the keywords only.
+In case the weights are not relevant, we have also added two features (\texttt{entityJaccardSim} and \texttt{keywordJaccardSim}), where Jaccard similarity is used instead of the cosine similarity.
+
+\item {\bf Miscellaneous features}.
+This group contains three miscellaneous features that seem discriminative but are unrelated to the previous two groups.
+The first feature is related to the event location.
+By analyzing the locations mentioned in the articles in a cluster we are able to estimate where the event occurred.
+The \texttt{hasSameLocation} feature is a boolean variable that is true when the location of the event in both clusters is the same.
+Another important feature is related to the time when the articles in the clusters were published.
+For articles in each cluster we first compute the average date and time when the articles were published.
+The \texttt{timeDiff} is then defined as the absolute difference in hours between the two computed dates.
+The last feature in this group is computed by analyzing the mentions of dates in the articles.
+Using an extensive set of regular expressions we are able to detect mentions of dates in different forms.
+To compute the feature \texttt{sharedDates} we start by identifying the list of dates mentioned in articles in each cluster separately.
+The value of \texttt{sharedDates} is then determined as the ratio of all dates that are detected in both clusters.
+
+\end{itemize}
+
 \subsection{Evaluation}
+
+In order to evaluate the quality of individual features and the overall model we have performed a set of experiments.
 

### 2015-02-21 rupnik2015.tex

@@
 
-\section{Multilingual Latent Factor Models}
+\section{Multilingual Latent Factor Models}\label{sec:models}
+
+\subsection{Multilingual dataset}
+
+The cross-lingual similarity models presented in this paper are based on comparable corpora.
+That is, a corpus of documents in multiple languages, with alignment between documents that are of the same topic, or even a rough translation of each other.
+Wikipedia is an example of a comparable corpus, where a specific entry can be described in multiple languages (e.g.
+"Berlin" is currently described in 222 languages).
+News articles represent another example, where the same event can be described by newspapers in several languages.
+
+More formally, \emph{multilingual document} $d = (u_1,\ldots u_m)$ is a tuple of $m$ documents on the same topic (comparable), where $u_i$ is the document written in language $i$.
+Note that individual document $u_i$ can be an empty document (missing resource) and each $d$ must contain at least two nonempty documents.
+A comparable corpus $D = {d_1, \ldots, d_N}$ is a collection of multilingual documents.
+By using the vector space model we can represent $D$ as a set of $m$ matrices $X_1,\ldots,X_m$, where $X_i \in \RR^{n_i \times N}$ is the matrix corresponding to the language $i$ and $n_i$ is the vocabulary size of language $i$.
+Furthermore, let $X_i^{\ell}$ denote the $\ell$-th column of matrix $X_i$ and the matrices respect the document alignment - the vector $X_i^\ell$ corresponds to the TFIDF vector of the $i$-th component of multilingual document $d_\ell$.
+We use $N$ to denote the total row dimension of $X$, i.e.
+$N:= \sum_{i=1}^m n_i$.
 

### 2015-02-23 code/matlab/alignProcessDocs.m
matlab code

@@ -1,0 +1,68 @@
+function [mats, wrds, docs, invdoc, didx, widx, langnames] = alignProcessDocs(langnames, datafolder)
+%example: alignProcessDocs({'en','es','de'});
+if nargin == 1
+    datadir = 'd:\work\data\jair\';
+end
+
+%%
+cd(mfilename('fullpath'));
+alangs = cell(size(langnames));
+adocnames = cell(size(langnames));
+indices = cell(size(langnames));
+words = cell(size(langnames));
+
+unique_cell = load(strcat(datadir, 'unique_cell.mat'),'-mat'); unique_cell = unique_cell.unique_cell;
+langids = loadwords(strcat(datadir, 'langnames.txt'));
+
+for i = 1:length(langnames)
+   %tempidx = find langname{i} in langids
+   tidx = find(cellfun(@(x)(strcmp(x,langnames{i})),langids));
+   indices{i} = unique_cell{tidx};
+end
+
+% align
+idxc = aligner(indices);
+disp('start')
+for i = 1:length(langnames)
+   temp = load(strcat(datadir, strcat(langnames{i}, '.matlab')), '-mat');
+   alangs{i} = temp.spmatrix(:, idxc{i});
+   temp = loadwords(strcat(datadir, strcat(langnames{i}, '.names')));
+   adocnames{i} = temp(idxc{i});
+   clear temp;
+   words{i} = loadwords(strcat(datadir, strcat(langnames{i}, 'words.txt')));
+  
+end
+disp('loaded and aligned')
+
+%% prune words
+plangs = cell(size(alangs));
+pwords = cell(size(alangs));
+pwordsIdx = cell(size(alangs));
+for i = 1:length(alangs)
+    [plangs{i}, idx] = wordprune(alangs{i}, 0.975);
+    pwordsIdx{i} = idx;
+    pwords{i} = words{i}(idx);
+end
+clear alangs;
+disp('pruned')
+%%
+invdoc = cell(size(plangs));
+tlangs = cell(size(plangs));
+for i = 1:length(plangs)
+   [tlangs{i}, invdoc{i}] = tfidf(plangs{i});
+   tlangs{i} = normalizeSparseColumns(tlangs{i}); 
+end
+clear plangs;
+disp('tfidf done')
+%% save
+% pwords: pruned word list
+% pwordsIdx: indices in the full word
+% plangs: pruned words, aligned languages
[…]

### 2015-02-23 code/matlab/aligner.m

@@ -1,0 +1,11 @@
+function idxc = aligner(cells) 
+
+idx = cells{1};
+for i = 2:length(cells)
+    idx = intersect(idx, cells{i});
+end
+
+idxc = cell(size(cells));
+for i = 1:length(cells)
+   [~, idxc{i}, ~] = intersect(cells{i}, idx); 
+end

### 2015-02-23 code/matlab/dispv.m

@@ -1,0 +1,2 @@
+function [] = dispv(vec, words)
+disp(words(find(vec))');

### 2015-02-23 code/matlab/doctransform.m

@@ -1,0 +1,19 @@
+function Y = doctransform(X, widx, invdoc)
+%X is a sparse matrix, or set of sparse matrices (several languages)
+
+%filters words using widx (rows of X)
+%applies invdoc weights
+%normalizes
+if ~iscell(X)
+    Y = X(widx,:);
+    Y = tfidf(Y, invdoc);
+    Y = normalizeSparseColumns(Y);
+    
+else
+    Y = cell(size(X));
+    for i = 1:length(X)
+        Y{i} = X{i}(widx{i},:);
+        Y{i} = tfidf(Y{i}, invdoc{i});
+        Y{i} = normalizeSparseColumns(Y{i});
+    end
+end

### 2015-02-23 code/matlab/experiments.m

@@ -1,0 +1,17 @@
+loadit = false;
+if ~loadit
+    clear
+    disp('computing');
+    [mats, wrds, docs, invdoc, didx, widx, langnames] = alignProcessDocs({'en','es','de'});
+    disp('saving');
+    quicksave;
+else
+    quickload;
+end
+
+%%
+nrp = 10;
+niter = 5;
+[projectors, centers] = kmeansprojector(mats, nrp, niter);
+
+[projectors2, centers2] = svdprojector(mats, nrp);

### 2015-02-23 code/matlab/kmeansprojector.m

@@ -1,0 +1,29 @@
+function [projectors, centers] = kmeansprojector(langs, nrp, niter)
+% langs: totally aligned data matrices, column examples
+% nrp: number of random projections
+% niter: number of kmeans iterations
+m = length(langs);
+centers = cell(1,m);
+
+ndim = zeros(m,1);
+for i = 1:m
+    centers{i} = zeros(nrp,1);
+    ndim(i) = size(langs{i},1);
+end
+
+if size(langs,2) > size(langs,1)
+    langs = langs';
+end
+
+[idx,C, nor2] = kmeans2(cell2mat(langs), nrp, niter);
+
+
+projectors = mat2cell(C, ndim, nrp);
+% pseudoinverse: use centroids as a genereal nonotrhonormal basis
+% new coefs = express vectors in terms of the basis
+% projectors: take x and just multiply
+for i = 1:m
+   [Q,R] = qr(projectors{i}, 0);   
+   projectors{i} = R \ Q'; %inv(R) * Q'   
+end
+projectors = projectors';

### 2015-02-23 code/matlab/quickload.m

@@ -1,0 +1,14 @@
+if ~exist('datadir')
+    datadir = 'd:\work\data\jair\';
+end
+
+%%
+cd(mfilename('fullpath'));
+
+load(strcat(datadir, 'mats'));
+load(strcat(datadir, 'wrds'));
+load(strcat(datadir, 'docs'));
+load(strcat(datadir, 'widx'));
+load(strcat(datadir, 'didx'));
+load(strcat(datadir, 'invdoc'));
+load(strcat(datadir, 'langnames'));

### 2015-02-23 code/matlab/quicksave.m

@@ -1,0 +1,11 @@
+if ~exist('datadir')
+    datadir = 'd:\work\data\jair\';
+end
+
+save(strcat(datadir, 'mats.mat'), 'mats', '-v7.3');
+save(strcat(datadir, 'wrds.mat'), 'wrds');
+save(strcat(datadir, 'docs.mat'), 'docs');
+save(strcat(datadir, 'widx.mat'), 'widx');
+save(strcat(datadir, 'didx.mat'), 'didx');
+save(strcat(datadir, 'invdoc.mat'), 'invdoc');
+save(strcat(datadir, 'langnames.mat'), 'langnames');

### 2015-02-23 code/matlab/svd_funn.m

@@ -1,0 +1,430 @@
+function [U,S,V] = svd_fun(multA, multAt, Atmult, m, n, real_matrix, k,its,l)
+%Svd  Low-rank approximation of matrix given in the implicit form in SVD form.
+% A(x, 1) should return A*x
+% A(y, 2) should return A'*y
+% A is of dimension m x n
+%
+%
+%   [U,S,V] = svd_fun(A, m, n)  constructs a nearly optimal rank-6 approximation
+%             USV' to A, using 2 full iterations of a block Lanczos method
+%             of block size 6+2=8, started with an n x 8 random matrix,
+%             when A is m x n; the ref. below explains "nearly optimal."
+%             The smallest dimension of A must be >= 6 when A is
+%             the only input to PCA.
+%
+%   [U,S,V] = svd_fun(A, m, n, k)  constructs a nearly optimal rank-k approximation
+%             USV' to A, using 2 full iterations of a block Lanczos method
+%             of block size k+2, started with an n x (k+2) random matrix,
+%             when A is m x n; the ref. below explains "nearly optimal."
+%             k must be a positive integer <= the smallest dimension of A.
+%
+%   [U,S,V] = svd_fun(A, m, n, k,its)  constructs a nearly optimal rank-k approx. USV'
+%             to A, using its full iterations of a block Lanczos method
+%             of block size k+2, started with an n x (k+2) random matrix,
+%             when A is m x n; the ref. below explains "nearly optimal."
+%             k must be a positive integer <= the smallest dimension of A,
+%             and its must be a nonnegative integer.
+%
+%   [U,S,V] = svd_fun(A, m, n, k,its,l)  constructs a nearly optimal rank-k approx.
+%             USV' to A, using its full iterates of a block Lanczos method
+%             of block size l, started with an n x l random matrix,
+%             when A is m x n; the ref. below explains "nearly optimal."
+%             k must be a positive integer <= the smallest dimension of A,
+%             its must be a nonnegative integer,
+%             and l must be a positive integer >= k.
+%
[…]

### 2015-02-23 code/matlab/svdprojector.m

@@ -1,0 +1,25 @@
+function [projectors, centers, S] = svdprojector(langs, nrp)
+
+m = length(langs);
+centers = cell(1,m);
+
+ndim = zeros(m,1);
+for i = 1:m
+    centers{i} = zeros(nrp,1);
+    ndim(i) = size(langs{i},1);
+end
+
+if size(langs,2) > size(langs,1)
+    langs = langs';
+end
+
+
+[U,S,~] = svd_funn(cell2mat(langs),[],[],sum(ndim), size(langs{1},2),true, nrp, 2, nrp+2);
+
+projectors = mat2cell(U, ndim, nrp);
+
+for i = 1:m
+   [Q,R] = qr(projectors{i}, 0);   
+   projectors{i} = R \ Q'; %inv(R) * Q'   
+end
+projectors = projectors';

### 2015-02-23 code/matlab/wordprune.m

@@ -1,0 +1,22 @@
+function [Y, idx] = wordprune(X, frac, skipfrac)
+
+Z = sign(X); x = sum(Z, 2);
+
+[x2,i2] = sort(-x); c = cumsum(x2)/sum(x2);
+
+%index where c(i) > frac, c(i-1) < frac
+
+m = c < frac;
+M = c >= frac;
+cutidx = find(circshift(m,1) & M);
+
+cutidxstart = 1;
+if nargin == 3
+    ms = c < skipfrac;
+    Ms = c >= skipfrac;
+    cutidxstart = find(circshift(ms,1) & Ms);
+end
+idx = i2(cutidxstart:cutidx);
+Y = X(idx, :);
+
+

### 2015-02-23 code/matlab/experiments.m
code update

@@ -1,8 +1,14 @@
+if ~exist('datadir')
+    datadir = 'd:\work\data\jair\';
+end
+
+%%
+cd(mfilename('fullpath'));
 loadit = false;
 if ~loadit
     clear
     disp('computing');
-    [mats, wrds, docs, invdoc, didx, widx, langnames] = alignProcessDocs({'en','es','de'});
+    [mats, wrds, docs, invdoc, didx, widx, langnames] = alignProcessDocs({'en','es','de'},datadir);
     disp('saving');
     quicksave;
 else

### 2015-02-23 code/matlab/alignProcessDocs.m
code update

@@ -1,4 +1,4 @@
-function [mats, wrds, docs, invdoc, didx, widx, langnames] = alignProcessDocs(langnames, datafolder)
+function [mats, wrds, docs, invdoc, didx, widx, langnames] = alignProcessDocs(langnames, datadir)
 %example: alignProcessDocs({'en','es','de'});
 if nargin == 1
     datadir = 'd:\work\data\jair\';

### 2015-02-23 code/matlab/experiments.m
code update

@@ -1,8 +1,12 @@
+if ~exist('datadir')
+    datadir = 'd:\work\data\jair\';
+end
+
 loadit = false;
 if ~loadit
     clear
     disp('computing');
-    [mats, wrds, docs, invdoc, didx, widx, langnames] = alignProcessDocs({'en','es','de'});
+    [mats, wrds, docs, invdoc, didx, widx, langnames] = alignProcessDocs({'en','es','de'}, datadir);
     disp('saving');
     quicksave;
 else

### 2015-02-23 code/matlab/experiments.m
code update

@@ -4,7 +4,6 @@ end
 
 loadit = false;
 if ~loadit
-    clear
     disp('computing');
     [mats, wrds, docs, invdoc, didx, widx, langnames] = alignProcessDocs({'en','es','de'}, datadir);
     disp('saving');

### 2015-02-23 code/matlab/experiments.m
code update

@@ -11,10 +11,46 @@ if ~loadit
 else
     quickload;
 end
+%%
+n = size(mats{1},2);
+ntest = 1000;
+ntrain = n - ntest;
+peri = randperm(n);
+tridx = peri(1:ntrain);
+tsidx = peri(ntrain+1:end);
+trmats = cell(size(mats));
+tsmats = cell(size(mats));
+m = length(mats);
+for i = 1:m
+   trmats{i} = mats{i}(:,tridx);
+   tsmats{i} = mats{i}(:,tsidx);
+end
+clear mats
 
 %%
 nrp = 10;
 niter = 5;
-[projectors, centers] = kmeansprojector(mats, nrp, niter);
+[projectors, centers] = kmeansprojector(trmats, nrp, niter);
 
-[projectors2, centers2] = svdprojector(mats, nrp);
+[projectors2, centers2] = svdprojector(trmats, nrp);
+
+prj = @(P,X,c) (P*X - (P*c)*ones(1, size(X,2)));
+prjCent = @(P,X,c) (P*X - c*ones(1, size(X,2)) );
+
+% use svd to project data first
+m = length(proj);
+ptrmats = cell(size(trmats));
+ptsmats = cell(size(tsmats));
+
+primalCenters = cell(size(trmats));
+projCenters = cell(size(trmats));
+for i = 1:m
+   primalCenters{i} = mean(trmats{i},2);
+   projCenters{i} = projectors2{i}*primalCenters{i};
+   
+   ptrmats{i} = prj(projectors2{i}, trmats{i}, primalCenters{i});
+   ptsmats{i} = prj(projectors2{i}, tsmats{i}, primalCenters{i});
+end
+
+[projectors3, centers3] = starccaprojector(ptrmats, nrp, projectors2, projCenters);
+

### 2015-02-23 code/matlab/kmeansprojector.m

@@ -7,7 +7,6 @@ centers = cell(1,m);
 
 ndim = zeros(m,1);
 for i = 1:m
-    centers{i} = zeros(nrp,1);
     ndim(i) = size(langs{i},1);
 end
 

### 2015-02-23 code/matlab/svdprojector.m

@@ -4,8 +4,7 @@ m = length(langs);
 centers = cell(1,m);
 
 ndim = zeros(m,1);
-for i = 1:m
-    centers{i} = zeros(nrp,1);
+for i = 1:m    
     ndim(i) = size(langs{i},1);
 end
 
@@ -20,6 +19,6 @@ projectors = mat2cell(U, ndim, nrp);
 
 for i = 1:m
    [Q,R] = qr(projectors{i}, 0);   
-   projectors{i} = R \ Q'; %inv(R) * Q'   
+   projectors{i} = R \ Q'; %inv(R) * Q'    
 end
 projectors = projectors';

### 2015-02-23 code/matlab/wordprune.m
code update

@@ -1,4 +1,4 @@
-function [Y, idx] = wordprune(X, frac, skipfrac)
+function [Y, idx] = wordprune(X, frac, skipfrac, maxWords)
 
 Z = sign(X); x = sum(Z, 2);
 
@@ -11,10 +11,13 @@ M = c >= frac;
 cutidx = find(circshift(m,1) & M);
 
 cutidxstart = 1;
-if nargin == 3
-    ms = c < skipfrac;
-    Ms = c >= skipfrac;
-    cutidxstart = find(circshift(ms,1) & Ms);
+% if nargin >= 3
+%     ms = c < skipfrac;
+%     Ms = c >= skipfrac;
+%     cutidxstart = find(circshift(ms,1) & Ms);
+% end
+if nargin == 5
+    cutidx = min(cutidx, maxWords);
 end
 idx = i2(cutidxstart:cutidx);
 Y = X(idx, :);

### 2015-02-23 code/matlab/experiment_kmeans.m
minor

@@ -1,5 +1,5 @@
 cd('D:\work\latex\jair_paper\code\matlab')
-datadir = 'E:\cross-ling-event-registry'
+datadir = 'E:\cross-ling-event-registry\'
 
 selectedLangs = {'en','es','de'};
 [mats, wrds, docs, invdoc, didx, widx, langnames] = alignProcessDocs(selectedLangs, datadir);

### 2015-02-23 code/matlab/experiment_lsi.m
link

@@ -1,0 +1,14 @@
+%cd('D:\Users\rupnik\work\latex\jair_paper\code\matlab')
+datadir = 'D:\Users\rupnik\work\data\cross-ling-event-registry\'
+
+selectedLangs = {'en','es','de'};
+[mats, wrds, docs, invdoc, didx, widx, langnames] = alignProcessDocs(selectedLangs, datadir);
+
+
+%%
+nrp = 500;
+
+disp('data loaded');
+[projectors, centers] = svdprojector(mats, nrp);
+disp('kmeans saved');
+save svdprojectors;

### 2015-02-23 rupnik2015.tex
- notation changes. small text updates

@@
 The remaining words are represented as a vector in the vector-space model.
-Each value in the vector is normalized using the TF-IDF approach, defined as $TFIDF(t,d) = TF(t,d) * IDF(t)$.
-$TF(t,d)$ represents the \emph{term frequency} (number of times the term $t$ appears in the document $d$), and $IDF(t) = log(N / DF(t))$ is the \emph{inverse document frequency} of term $t$.
-$IDF(t)$ is computed by comparing the number of documents mentioning the term $t$ ($DF(t)$) with the total number of documents $N$.
-After normalizing the vector, we use it to find the most similar existing cluster.
+Each value in the vector is additionally normalized using TF-IDF.
+After normalization, we use the vector to find the most similar existing cluster.
 To be able to compare the vector with a cluster we first need a vector representation for each of the clusters.
@@
 By computing the cosine similarity of the article vector with each of the centroid vectors we then find the cluster with the highest similarity.
-If the highest similarity is above the selected threshold (0.4 in our experiments), the article is assigned to the corresponding cluster, otherwise a new cluster is created, initially containing only the single article.
+If the highest similarity is above the selected threshold (we used 0.4 in our experiments), the article is assigned to the corresponding cluster, otherwise a new cluster is created, initially containing only the single article.
 
@@
 This condition ensures that only content that receives a sufficient coverage in the media is treated as an event.
-Once a new event is identified, a new unique ID is assigned to it and the main information about the event is automatically extracted by analyzing the articles assigned to it.
+Once a new event is identified, a new unique ID is assigned to it and the main information about the event is then automatically extracted by analyzing the articles assigned to it.
 The extracted information includes properties such as the date of the event, the location, who is involved in it, what the event is about, etc.
@@
 When events are identified, they will therefore be identified from clusters in each language separately.
-This has an undesirable consequence for Event Registry where the same actual event will be represented with two or more event IDs, and each event will only contain coverage in a single language.
+As a result, the same event will be represented in Event Registry with two or more event IDs, and each event will only contain coverage in a single language.
 To avoid this problem we want to exploit the available information about the similarity of the articles in a cluster with articles in other languages.
@@
 Formally, the problem of identifying the clusters describing the same event can be formalized as follows.
-Each article $a_i$ is written in a language $l$, where $l \in \{A,B,...,H\}$.
+Each article $a_i$ is written in a language $l$, where $l \in L = \{l_1,l_2,...,l_k\}$.
 For each language $l$, a set of clusters $C_l$ is generated and maintained that only contains articles in language $l$.
-For any cluster $c_i \in C_{l1}$, the task is then to identify other clusters $c_j \in C_{l2}$ (where $l1 \neq l2$) that are describing the same event as $c_i$.
+For any cluster $c_i \in C_{l_a}$, the task is then to identify other clusters $c_j \in C_{l_b}$ (where $l_a \neq l_b$) that are describing the same event as $c_i$.
 We will refer to cluster pairs $c_i$ and $c_j$ that are describing the same event as equivalent clusters.

### 2015-02-23 code/matlab/experiment_lsi.m
intructions

@@ -11,4 +11,4 @@ nrp = 500;
 disp('data loaded');
 [projectors, centers] = svdprojector(mats, nrp);
 disp('kmeans saved');
-save svdprojectors;
+save svdprojectors -v7.3;

### 2015-02-23 code/matlab/instructions.m

@@ -1,0 +1,22 @@
+%load projectors 
+%on pankretas: D:\Users\rupnik\jair\kmeansprojectors
+% (same instructions for lsiprojectors and ccaprojectors)
+
+
+% documents will always be columns
+%langnames : language order
+
+%to compute similarities between two TF docs (based on bow files):
+
+%1: prune words and computed TFIDFN
+%X in lanugae i (matrix with column examples, or a column vector)
+%Y in language j (matrix with column examples, or a column vector)
+X = doctransform(X, widx{i}, invdoc{i});
+Y = doctransform(Y, widx{j}, invdoc{j});
+
+%2: project, normalize
+pX = normalizeSparseColumns(projectors{i}*X);
+pY = normalizeSparseColumns(projectors{j}*Y);
+
+%3: similarity
+S = pX'*pY;

### 2015-02-23 code/matlab/experiment_kmeans.m
minor

@@ -10,5 +10,5 @@ nrp = 500;
 niter = 100;
 disp('data loaded');
 [projectors, centers] = kmeansprojector(mats, nrp, niter);
-disp('kmeans saved');
-save kmeansprojectors;
+disp('kmeans computed');
+save kmeansprojectors -v7.3;

### 2015-02-23 code/matlab/experiments_cca.m
cca experiment

@@ -1,0 +1,28 @@
+%cd('D:\Users\rupnik\work\latex\jair_paper\code\matlab')
+datadir = 'D:\Users\rupnik\work\data\cross-ling-event-registry\'
+load 'D:\Users\rupnik\jair\svdprojectors.mat'
+%%
+%nrp = 5;
+%[projectors, centers] = svdprojector(mats, nrp);
+
+
+prj = @(P,X,c) (P*X - (P*c)*ones(1, size(X,2)));
+prjCent = @(P,X,c) (P*X - c*ones(1, size(X,2)) );
+
+% use svd to project data first
+m = length(mats);
+pmats = cell(size(mats));
+
+primalCenters = cell(size(mats));
+projCenters = cell(size(mats));
+for i = 1:m
+   primalCenters{i} = mean(mats{i},2);
+   projCenters{i} = projectors{i}*primalCenters{i};
+   
+   pmats{i} = prjCent(projectors{i}, mats{i}, projCenters{i});
+end
+
+preproj = projectors;
+[projectors, centers] = starccaprojector(pmats, nrp, preproj, projCenters);
+
+save ccaprojectors -v7.3;

### 2015-02-23 code/matlab/instructions.m

@@ -14,9 +14,9 @@
 X = doctransform(X, widx{i}, invdoc{i});
 Y = doctransform(Y, widx{j}, invdoc{j});
 
-%2: project, normalize
-pX = normalizeSparseColumns(projectors{i}*X);
-pY = normalizeSparseColumns(projectors{j}*Y);
+%2: project, center, normalize
+pX = normalizeSparseColumns(projectors{i}*X - centers{i}*ones(1,size(X,2)));
+pY = normalizeSparseColumns(projectors{j}*Y - centers{j}*ones(1,size(Y,2)));
 
 %3: similarity
 S = pX'*pY;

### 2015-02-24 rupnik2015.tex
LSI updated, regression and mcca removed.

@@
 We then run the $k$-means algorithm \cite{kmeans} and obtain a set of $k$ centroid vectors, which form a matrix: $C := \begin{bmatrix}C^1, \ldots, C^k \end{bmatrix}$.
-The centroid matrix can be split vertically into $m$ blocks: $$C = [C_1^T \cdots C_m^T]^T.$$
+The centroid matrix can be split vertically into $m$ blocks: $$C = [C_1^T \cdots C_m^T]^T,$$ according to the number of dimensions of each language, i.e.
+$C_i \in \RR^{n_i}$.
 
@@
 
-{\color{red}
- Each column $u_i$ of $U_k$ consists of block vectors $u_i = \begin{bmatrix} u^{1^T}_i & \dots & u^{\ell^T}_i \end{bmatrix}^T.$ We do not normalize each block $j$  as this would destroy the low rank approximation.
-Each block $j$ is normalized $u^{j^1}_i = u^{j^1}_i/||u^{j^1}_i||$ to reduce the bias.
+An aligned basis is obtained by first splitting $U$ vertically according to the number of dimensions of each language: $U = [U_1^T \cdots U_m^T]^T$.
+Then, the same as with $k$-means clustering, we compute the pseudoinverses$P_i = (U_i^T U_i)^{-1} U_i^T$.
+The matrices $P_i$ are used to change the basis from the standard basis in $\RR^{n_i}$ to the basis of columns of $U_i$.
 
-We obtain the aligned reduced basis $U_{\textrm{aligned}} = \begin{bmatrix} {U^1}^T &\dots & {U^\ell}^T \end{bmatrix}^T,$ where $D_j =  U^j \Sigma_k V_k.$ The reduced language free representation for language $j$ and document $d$ is computed as the weighted least square solution $\Sigma_k^{-1} {U^j}^+ d,$ where $+$ denotes the pseudo inverse of a matrix that is is used because columns of $U^j$ do not form an orthogonal basis.
-The numerical implementation of the least squares is done by QR algorithm.
-}
-
-\subsubsection{Refinement of LSI similar to MCCA}
-
-{\oldText{Taken from Low-rank paper}
-Finally, we test mCCA is specifically designed to consider data from multiple sources (in this case languages). Therefore, we do not merge the individual term-document matrices into one as in
-the other two methods.
-For each language (view), we estimate the pairwise correlation coefficients, then we try to find vectors which maximize the sum of all pairwise correlations over all languages.
-This can be written as the following optimization
-\begin{equation}
-\label{eq:opt1}
-\max_{w_1,\ldots, w_\ell} \sum_{i=1}^\ell\sum_{j=i+1}^\ell  \frac{w_i^T D_i D_j^T w_j}{\sqrt{ w_i^T D_i D_i^T w_i\vphantom{ D_j^T } }\sqrt{ w_j^T D_j D_j^T w_j}}
-\end{equation}
-However, this allows only for a 1-dimensional representation (i.e.
-one vector per language).
-Since we would like to allow for more vectors per language, we denote the double sum in Equation~\ref{eq:opt1} by $\mbox{mCCA}(w_1,\ldots,w_\ell)$.
-If $\mbox{corr}$ denotes the correlation, to find $M$ vectors, the optimization objective function  becomes
-\begin{equation}
-\max_{\substack{w_i^{(j)}\;;\;i=1,\ldots,\ell\\j=1,\ldots,M }}\sum_{s=1}^{\mbox{M}} \mbox{mCCA}\left(w_1^{(s)},\ldots,w_\ell^{(s)}\right),
-\qquad\mbox{s.t.} \qquad \mbox{corr} \left(w_i^{(s)},w_i^{(t)}\right) = 0
-\quad \forall s\neq t
-\end{equation}
-We require that the set of $M$ vectors we return for each language are uncorrelated (so that we do not get copies of a vector) which together maximize the pairwise correlation between languages.
-}
-
-\subsubsection{Regression}
-
-{\oldText{Taken from Similarity paper}
-We begin with a collection of documents in two languages\footnote{The extension to multiple language is  discussed in the Discussion section.}.
-The documents are represented via bag-of-words as vector spaces $\X$ and $\Y$.
-There are several possible choices of monolingual similarity functions.
-We consider functions of the form: $x'C_{\X\X}^i x$ for $i\in \mathbb{Z}$.
-For $i=0$, we recover the squared Euclidean distance and for $i=-1$, we get the Mahalanobis distance.
-
-As input, we take a choice of $i$ (choice of inner product) along with an alignment of $\X$ and $\Y$.
-The problem is to find a mapping which respects both monolingual similarity and measures cross-lingual similarity.
-In the first we reduce the problem to monolingual similarity.
-We find some mapping $f: \X \rightarrow \Y$.
-For a pair of elements $x\in\X$ and $y\in\Y$, we compute their similarity as the similarity of $f(x)$ and $y$ in $\Y$.
-
-The output of all the algorithms is a linear map (a matrix) denoted $B$, which computes the cross-lingual similarity as $x^T B y$ for $x\in\X$ and $y \in \Y$.
-The simplest choice for $B$ is $C_{\X\Y}$, the cross-covariance matrix.
-
-To find $B$ using regression, we compute the optimal regression matrix as $ C_{\Y \X} (C_{\X \X})^{-1}$.
-This corresponds to the mapping $f$ mentioned above.
-The inner product on the space $\Y$ induces $B$ for computing the similarities between elements in $\X$ and $\Y$: for $B$ it holds that $x^T B y = \langle f (x), y \rangle$.
-Therefore $B$ depends on the metric on $\Y$.
-For the inner product corresponding to $C_{\Y\Y}$, $B$ can be derived as $C_{\X \X}^{-1}C_{\X \Y}C_{\Y \Y}$.
-
-Note that in the case of regression, we have a choice of which direction we map: $\X\rightarrow\Y$ or $\Y\rightarrow\X$ except for Mahalanobis metric where the formula for $B$ is symmetric.
-Due to space constraints we only note here that CCA approximates the least-squares regression for this metric just as LSI approximates the cross-covariance matrix.
-}
-
-{\oldText{Taken from Pascal CL paper}
-he \textbf{regression based approach} consists of two ingredients: a linear regression mapping $f : \RR^{n_i} \rightarrow \RR^{n_j}$,$$ f(x) := F\cdot x,$$ where $F \in \RR^{n_j \times n_j}$ and an inner product on the target space $\langle \cdot, \cdot \rangle_D$, which are used to define:
-$$s_{i,j}(x,y) := \langle f(x), y \rangle_D = f(x)^T D y = x^T F^T D y.$$
-Choices of $D$ included the standard inner product $I$, Mahalanobis inner product $C_{j,j}^{-1}$, which corresponds to whitening, and the covariance matrix $C_{j,j}$ (related to a co-occurrence).
-We can consider two regression mappings:
-\begin{itemize}
-\item least squares regression (LSQ), where $F = C_{j,i}C_{i,i}^{-1}$
-\item canonical correlation regression\cite{ccar} (CCAR), where $F = C_{j,j}^{-1}C{j,i}$.
-\end{itemize}
-}
+The numerical implementation of the least squares is done by QR algorithm, by computing factorizing $U_i = Q R$, where $Q^TQ = I$ and $R$ is triangular matrix.
+$P_i$ is then obtained by solving $R P_i = Q$.
 

### 2015-03-03 rupnik2015.tex
outline added

@@
 \end{abstract}
+\section{Overview}
+Structure
+\begin{itemize}
+\item Introduction
+\item Cross lingual document similarity (model)
+\item  Event Detection
+\item  Combing the approaches
+\item Experiments
+\item Conclusions
+\end{itemize}
+The story:
+\begin{itemize}
+\item Description of the problem - we would liek to understand events ans sequences of events, but our observables (e.g. news stories) occur in a variety of languages
+\item We present an approach to tackle this problem using a combination of techniques....
+\item Ultimately we present a tool showing the feasibility of our approach
+\end{itemize}
 

### 2015-03-04 rupnik2015.tex
code added

--- 2015-02-24 rupnik2015.tex
@@
 
-A \emph{hub language} is a language with a high proportion of non-empty documents in $D$, which in case of Wikipedia is English. We use the following notation to define subsets
-of the multilingual comparable corpus: let $a(i,j)$ denote the index set of all multilingual documents with non-missing data for the $i$-th and $j$-th language:  $a(i,j) = \left\{ d = (u_1,...,u_m) | u_i \neq \emptyset, u_j \neq \emptyset \right\}$ , and let $a(i)$ denote the index set of all multilingual documents with non missing data for the $i$-th language.
+A \emph{hub language} is a language with a high proportion of non-empty documents in $D = \left\{d_1,..., d_{\ell}\right\}$, which in case of Wikipedia is English.
+We use the following notation to define subsets
+of the multilingual comparable corpus: let $a(i,j)$ denote the index set of all multilingual documents with non-missing data for the $i$-th and $j$-th language:  $a(i,j) = \left\{k | d_k = (u_1,...,u_m), u_i \neq \emptyset, u_j \neq \emptyset \right\},$ and let $a(i)$ denote the index set of all multilingual documents with non missing data for the $i$-th language.
 

### 2015-03-04 code/cpp/app_in.cpp

@@ -1,0 +1,98 @@
+#include <base.h>
+#include <mine.h>
+
+void MakeVec(const TStr& Text, TUnicodeVSM::TGlibUBow& Bow, TFOut& FOut, const int& Count) {
+    TUStr UText(Text);
+    TIntKdV Vec = Bow.TextToVec(UText);
+    for (int VecN = 0; VecN < Vec.Len(); VecN++) {
+        FOut.PutStrFmtLn("%d %d %d", Count, Vec[VecN].Key.Val + 1, Vec[VecN].Dat.Val);
+    }
+}
+
+int main(int argc, char* argv[]) {
+	TUnicodeDef::Load("./UnicodeDef.Bin");
+	
+	TStr FPath = "/home/blazf/Projects/datasets/cca_exp/";
+	
+//	{
+//		// sort articles
+//		TIntStrKdV IdTextV;
+//		TFIn AFIn(FPath + "Articles.json");
+//		TStr LnStr; int Count = 0;
+//		while (AFIn.GetNextLn(LnStr)) {
+//			if (Count % 1000 == 0) { printf("R%d\r", Count); } Count++;
+//			PJsonVal Val = TJsonVal::GetValFromStr(LnStr);
+//			IdTextV.Add(TIntStrKd(Val->GetObjInt("id"), LnStr));
+//		}
+//		IdTextV.Sort();
+//		
+//		TFOut AFOut(FPath + "Articles.sorted.json");
+//		for (int i = 0; i < IdTextV.Len(); i++) {
+//			if (Count % 1000 == 0) { printf("W%d\r", Count); } Count++;
+//			AFOut.PutStrLn(IdTextV[i].Dat);
+//		}
+//	}
+
+	{
+		TFIn AFIn(FPath + "Articles.sorted.json");
+		TFOut AFOut(FPath + "id_uri.json");
+		TStr LnStr; int Count = 0;
+		while (AFIn.GetNextLn(LnStr)) {
+			if (Count % 1000 == 0) { printf("R%d\r", Count); } Count++;
+			PJsonVal AVal = TJsonVal::GetValFromStr(LnStr);
+			PJsonVal Val = TJsonVal::NewObj();
+			Val->AddToObj("id", AVal->GetObjInt("id"));
+			Val->AddToObj("uri", AVal->GetObjStr("uri"));
+			AFOut.PutStrLn(Val->SaveStr());
+		}
+		
+	}
+	
[…]

### 2015-03-04 code/matlab/save_fltv.m

@@ -1,0 +1,16 @@
+function [] = save_fltv(x, fname, append)
+mode = 'w';
+if nargin == 3
+    if append
+        mode = 'a';
+    end
+end
+
+fid = fopen(fname, mode);
+
+fwrite(fid, length(x), 'int32'); %mxvals
+fwrite(fid, length(x), 'int32'); %vals
+
+fwrite(fid, x , 'double'); %valv
+fclose(fid);
+

### 2015-03-05 code/matlab/experiment_kmeans.m
paths minor change

@@ -1,5 +1,4 @@
-cd('D:\work\latex\jair_paper\code\matlab')
-datadir = 'E:\cross-ling-event-registry\'
+datadir = 'D:\Users\rupnik\work\data\cross-ling-event-registry\'
 
 selectedLangs = {'en','es','de'};
 [mats, wrds, docs, invdoc, didx, widx, langnames] = alignProcessDocs(selectedLangs, datadir);

### 2015-03-05 code/matlab/experiments_cca.m

@@ -1,6 +1,6 @@
 %cd('D:\Users\rupnik\work\latex\jair_paper\code\matlab')
-datadir = 'D:\Users\rupnik\work\data\cross-ling-event-registry\'
-load 'D:\Users\rupnik\jair\svdprojectors.mat'
+%datadir = 'D:\Users\rupnik\work\data\cross-ling-event-registry\'
+%load 'D:\Users\rupnik\jair\svdprojectors.mat'
 %%
 %nrp = 5;
 %[projectors, centers] = svdprojector(mats, nrp);

### 2015-03-06 rupnik2015.tex
- added story linking evaluation

@@
 
+\begin{table}[h]
+\caption{Accuracy of story linking with different sets of features}
+\label{table:linkingEval}
+\begin{center}
+\begin{tabular}{|c|c|c|c|}
+  \hline
+  \cline{1-4}
+  Features & CA & Precision & Recall \\ \cline{1-4}
+  CCA & 73.2\% & 0.717 & 0.763 \\ \cline{1-4}
+  Concepts & 88.5\% & 0.888 & 0.882 \\ \cline{1-4}
+  Misc & 62.7\% & 0.583 & 0.904 \\ \cline{1-4}
+  CCA + concepts & 89.0\% & 0.893 & 0.887 \\ \cline{1-4}
+  All & 88.9\% & 0.893 & 0.884 \\ \cline{1-4}
+  \hline
+\end{tabular}
+\end{center}
+\end{table}
+
 \subsubsection{Compare different feature subsets}

### 2015-03-06 code/matlab/alignProcessDocs.m
code update

@@ -37,7 +37,7 @@ plangs = cell(size(alangs));
 pwords = cell(size(alangs));
 pwordsIdx = cell(size(alangs));
 for i = 1:length(alangs)
-    [plangs{i}, idx] = wordprune(alangs{i}, 0.975, [], 500000);
+    [plangs{i}, idx] = wordprune(alangs{i}, 0.975, [], 1000000);
     pwordsIdx{i} = idx;
     pwords{i} = words{i}(idx);
 end

### 2015-03-06 code/matlab/kmeansprojector.m

@@ -8,6 +8,7 @@ centers = cell(1,m);
 ndim = zeros(m,1);
 for i = 1:m
     ndim(i) = size(langs{i},1);
+    centers{i} = zeros(nrp,1);
 end
 
 if size(langs,2) > size(langs,1)

### 2015-03-06 code/matlab/svdprojector.m

@@ -6,6 +6,7 @@ centers = cell(1,m);
 ndim = zeros(m,1);
 for i = 1:m    
     ndim(i) = size(langs{i},1);
+    centers{i} = zeros(nrp,1);
 end
 
 if size(langs,2) > size(langs,1)

### 2015-03-06 code/matlab/wordprune.m

@@ -16,7 +16,7 @@ cutidxstart = 1;
 %     Ms = c >= skipfrac;
 %     cutidxstart = find(circshift(ms,1) & Ms);
 % end
-if nargin == 5
+if nargin == 4
     cutidx = min(cutidx, maxWords);
 end
 idx = i2(cutidxstart:cutidx);

### 2015-03-10 rupnik2015.tex
- added evaluation using different models

--- 2015-03-05 rupnik2015.tex
@@
 
+\begin{table}[h]
+\caption{Accuracy of story linking with features obtained from different cross-lingual similarity models}
+\label{table:linkingEvalModels}
+\begin{center}
+\begin{tabular}{|c|c|c|c|}
+  \hline
+  \cline{1-4}
+  Models & CA & Precision & Recall \\ \cline{1-4}
+  CCA & 77.6\% & 0.783 & 0.811 \\ \cline{1-4}
+  $k$-means & 73.9\% & 0.695 & 0.846 \\ \cline{1-4}
+  LSI & 76.1\% & 0.755 & 0.769 \\ \cline{1-4}
+  \hline
+\end{tabular}
+\end{center}
+\end{table}
+
+\begin{table}[h]
+\caption{Accuracy of story linking with different sets of features}
+\label{table:linkingEval}
+\begin{center}
+\begin{tabular}{|c|c|c|c|}
+  \hline
+  \cline{1-4}
+  Features & CA & Precision & Recall \\ \cline{1-4}
+  CCA & 77.6\% & 0.783 & 0.811 \\ \cline{1-4}
+  Concepts & 88.8\% & 0.884 & 0.891 \\ \cline{1-4}
+  Misc & 65.3\% & 0.718 & 0.500 \\ \cline{1-4}
+  CCA + concepts & 89.2\% & 0.891 & 0.893 \\ \cline{1-4}
+  All & 89.6\% & 0.895 & 0.895 \\ \cline{1-4}
+  \hline
+\end{tabular}
+\end{center}
+\end{table}
+
 \subsubsection{Compare different feature subsets}

### 2015-03-11 rupnik2015.tex
- added text describing the experiments

@@
 \subsection{Algorithms}
+\label{sec:algorithms}
 
@@
 
-The learning dataset used to train a classification model consists of cluster pairs $c_i$ and $c_j$ for which a human annotator provided feedback whether the clusters are equivalent or not.
-The dataset contains XYZ examples, of which XYZ are equivalent clusters pairs and XYZ are not.
-To train a model, the vector of features was first extracted for each learning example in the same way as described in Algorithm~\ref{cluster_merge_algo2}.
-A linear SVM method was then used to train the classifier $M$.
-
 \begin{algorithm}[tb!]
@@
 
-\subsubsection{Learning features}
+\subsection{Training a classification model}
 
-In order to use a supervised approach to determine if two clusters $c_i$ and $c_j$ are equivalent we need to extract from the cluster pair as many informative features as possible.
-Based on the availability of the data, we have extracted the following groups of features:
+In order to train a classification model for predicting whether two clusters are equivalent we first need labeled training data.
+To obtain the data we used two human annotators.
+They were provided with a user interface that was displaying information about the articles in two clusters and their task was to determine if the clusters are equivalent or not.
+To obtain a pair of clusters $(c_i, c_j)$ to annotate, we first randomly chose a cluster $c_i$, used the Algorithm~\ref{cluster_merge_algo1} to compute a set of potentially equivalent clusters $C$ and randomly chose a cluster $c_j \in C$.
+
+The dataset provided by the annotators contains 808 examples, of which 402 are equivalent clusters pairs and 406 are not.
+Clusters in each learning example are either in English, Spanish or German language.
+Although Event Registry imports articles also in other languages, we restricted our experiments to these three languages since they have the highest volume of news articles which makes the training problem similarly hard.
+
+To build a classification model we first needed to extract from the labeled examples informative features. Based on the availability of the data, we have extracted the following groups of features:
 
 \begin{itemize}
-\item {\bf CCA related features}.
-CCA was already used to obtain candidates for the equivalent clusters.
+\item {\bf Cross-lingual article linking features}.
+Cross-lingual article linking information was already used to obtain candidates for the equivalent clusters.
 Additionally, it can also be used to provide valuable learning features.
-There are two main features that we extract using CCA -- \texttt{linkCount} and \texttt{avgSimScore}.
+There are two main features that we extract from article links between clusters -- \texttt{linkCount} and \texttt{avgSimScore}.
 \texttt{linkCount} is the number of times an article $a_i \in c_i$ has as one of the 10 most similar articles an article $a_j \in c_j$.
@@
 
+Using the described features and the user-provided label we can train a classification model $M$ that can be used in Algorithm~\ref{cluster_merge_algo2} to classify a new cluster pair as equivalent or not.
+The classification algorithm that we used to train a model was a linear SVM method.
+
 \subsection{Evaluation}
 
-In order to evaluate the quality of individual features and the overall model we have performed a set of experiments.
+In order to determine how accurately we can predict cluster equivalence we have performed two experiments.
+In the first experiment we wanted to test how well do the individual approaches for cross-lingual article linking perform when used for linking the clusters about the same event.
+In the second experiment we tested how accurate is the prediction model when trained on different subsets of learning features.
+To evaluate the prediction accuracy for a given dataset we used 10-fold cross validation.
+In this technique, the dataset is partitioned into 10 folds, where 9 folds are used for training a model and the left-out fold is then used to test the accuracy of the model.
+The learning and testing process is repeated 10 times, each time with a different left-out fold.
 
-evaluate accuracy on different language pairs only. compute accuracies using the CCA features only, using concept features only, or using all features.
+In Section~\ref{sec:algorithms} we have described three main algorithms for identifying similar articles in different languages.
+These algorithms were $k$-means, LSI and CCA.
+To test which of these algorithms performs best we performed the following test.
+Using each algorithm, we analyzed all articles in Event Registry to find the most similar articles in other languages.
+The values of the cross-lingual article linking features (\texttt{linkCount} and \texttt{avgSimScore}) were then recomputed for all annotated learning examples.
+A dataset with only these two features and the label was constructed and used to test how accurate prediction models can be induced using the data.
+The results of the experiment are shown in Table~\ref{table:linkingEvalAlgos}.
+The table shows for each of the algorithms the obtained classification accuracy, precision and recall.
+We can see that the most informative features for cluster linking are obtained using the CCA algorithm, while $k$-means performs the worse.
+{\oldText{TODO:} add some explanation why this might be so.}
+
+In the second experiment we wanted to evaluate how relevant individual groups of features are to correctly determine cluster equivalence.
+For this purpose we tested the accuracies obtained using individual groups of features, as well as using different combination of groups.
+Since we found the CCA to perform best of the three algorithms, we used it to compute the values of the cross-lingual article linking features.
+Results of the evaluation are shown in Table\~ref{table:linkingEval}.
+We can see that using a single group of features, the highest prediction accuracy can be achieved using the concept related features.
+The classification accuracy in this case is 88.8\%.
+By additionally including also the cross-lingual article linking features, the classification accuracy rises slightly to 89.2\%.
+Using all three groups of features, the achieved accuracy is 89.6\%.
+Based on the results we can make the following conclusions.
+The cross-lingual similarity algorithms provide valuable information that can be used to identify clusters that describe the same event in different languages.
+The computed features are however significantly less informative compared to the features computed on the annotated concepts.
+Nevertheless, the cross-lingual article similarity features are very important for two reasons.
+The first reason is that they allow us to identify for a given cluster a limited set of candidate clusters that are potentially equivalent.
+This is a very important feature since it reduces the search space for several orders of magnitude.
+The second reason for the importance of these features is that concept annotations are not available for all articles.
+Annotation of news articles is computationally intensive and can only be done for a subset of collected articles.
 
 \begin{table}[h]
-\caption{Accuracy of story linking with features obtained from different cross-lingual similarity models}
-\label{table:linkingEvalModels}
+\caption{Accuracy of cluster linking with features obtained from different cross-lingual similarity algorithms}
+\label{table:linkingEvalAlgos}
 \begin{center}
@@
 
-\subsubsection{Compare different feature subsets}
-
-\subsubsection{Accuracy across different language pairs}
-
 \section{Related work}

### 2015-03-16 rupnik2015.tex
updated cca, hub languages and GCCA

@@
 \newcommand{\Y}{\mathbb{Y}}
+
+\newcommand{\algInput}{\textbf{input:}}
 
@@
 
-The resulting matrix (when appropriately scaled to have unit norm) for similarity computation between language $i$ and language $j$ is defined as:
+The resulting matrix (when appropriately scaled to have unit norm) for similarity computation between language $i$ and language $j$ is defined as (see the reasoning below):
 $$C_i(C_i^T C_i)^{-1} (C_j^T C_j)^{-1} C_j.$$
 
-%% We could skip this par
-In the implementation of $k$-means clustering we never form the matrix $X$ explicitly but rather implement implicit multiplication as $X x = \sum_{i=1}^\ell X_i x$ and multiplication with transpose as $X^T y = \sum_{i=1}^\ell X_i^T y_i,$ where $y$ is partitioned accordingly to $X$.
-That lowers memory requirement and compacts sparse indexing, leading to faster element access.
+The matrix is a result of mapping documents in a language independent space using pseudo-inverses of the centroid matrices $P_i = (C_i^T C_i)^{-1} C_i$ and then comparing them using the standard inner product, which results in the matrix $P_i^T P_j$.
+For the sake of presentation, we assumed that the centroid vectors are linearly independent (an independent subspace could be obtained using an additional Gram-Schmidt step on the matrix $C$, if this was not the case).
 
-\subsubsection{LSI}
+\subsubsection{Cross-Lingual Latent Semantic Indexing}
 
-The next method is CL-LSI which is a variant of LSI~\cite{lsi} for more than one language.
+The next method is Cross-Lingual Latent Semantic Indexing (CL-LSI)\cite{cl_lsi} which is a variant of LSI~\cite{lsi} for more than one language.
 The method is based on computing a truncated singular value decomposition of $X \approx U S V^T$.
@@
 
+\subsection{Canonical Correlation Analysis}
+ For completeness we now introduce a statistical technique to analyise data from two sources - after which we will describe our main contribution to computing cross-lingual similarities.
+
+ Canonical Correlation Analysis (CCA)~\cite{Hotelling} is a dimensionality reduction technique similar to Principal Component Analysis (PCA)~\cite{Pearson1901On}, with an additional assumption that the data consists of feature vectors that arose from two sources (two views) that share some information (examples: bilingual document collection~\cite{mrpqr}, collection of images and captions~\cite{HardoonSS04}).
+Instead of looking for linear combinations of features that maximize the variance (PCA) we look for a linear combination of feature vectors from the first view and a linear combination for the second view, that are maximally correlated.
+
+Interpreting the nonzero columns of $X_i$ as observation vectors sampled from an underlying distribution $\mathcal{X}_i \in \RR^{n_i}$, the idea is to find two weight vectors $w_i \in \RR^{n_i}$ and $w_j \in \RR^{n_j}$ so that the random variables $w_i^T \cdot \mathcal{X_i}$ and $w_j^T \cdot \mathcal{X_j}$ are maximally correlated ($w_i$ and $w_j$ are used to map the random vectors to random variables, by computing weighted sums of vector components).
+By using the sample matrix notation $X_i$ and $X_j$ (assuming no data is missing for clearer presentation) this problem can be formulated as the following optimization problem:
+
+\begin{equation*}
+\begin{aligned}
+& \underset{w_i \in \RR^{n_i}, w_j \in \RR^{n_j}}{\text{maximize}}
+& & \frac{w_i^T C_{i,j} w_j}{\sqrt{w_i^T C_{i,i} w_i} \sqrt{w_j^T C_{j,j} w_j}},
+\end{aligned}
+\end{equation*}
+where $C_{i,i}$ and $C_{j,j}$ are empirical estimates of variances of $\mathcal{X}_i$ and $\mathcal{X}_j$ respectively and $C_{i,j}$ is an estimate for the covariance matrix.
+Assuming that the observation vectors are centered (only for the purposes of presentation), the matrices are computed in the following way: $C_{i,j} = \frac{1}{n-1}X_i X_j^T$, and similarly for $C_{i,i}$ and $C_{j,j}$.
+The optimization problem can be reduced to an eigenvalue problem and includes inverting the variance matrices $C_{i,i}$ and $C_{j,j}$.
+If the matrices are not invertible, one can use a regularization technique by replacing $C_{i,i}$ with $(1- \kappa)C_{i,i} + \kappa I$, where $\kappa \in [0,1]$ is the regularization coefficient and $I$ is the identity matrix (the same can be applied to $C_{j,j}$.
+A single canonical variable is usually inadequate in representing the original random vector and typically one looks for $k$ projection pairs $(w_i^1, w_j^1),\ldots,(w_i^k, w_j^k)$, so that $w_i^{uT} \mathcal{X}_i$ and $w_j^{uT} \mathcal{X}_j$ are highly correlated and $w_i^{uT} \mathcal{X}_i$ is uncorrelated with $w_i^{vT} \mathcal{X}_i$  for $u \neq v$ and analogously for $w_j^u$ vectors.
+
+Note that the method in its original form is only applicable to two languages where an aligned set of observations is available.
+
 \subsection{Hub languages}
 
+In this section we describe a method that is based on linear embeddings as well, but is more applicable to a large number of languages. The main difficulty with applying the
+LSI and $k$-means approaches lies in the fact, that when one considers a large number of languages, for example, the top 100 Wikipedia languages (ranked by number of articles), the set of completely
+aligned documents can be small or empty. Even if only two languages are considered, the set of aligned documents can be small (for example, Piedmontese and Hindi
+Wikipedias had no interlanguage links). In Wikipedia we observed that even though the training resources are scarce between certain language pairs, often there exists indirect training data
+through a hub language.
+
 A \emph{hub language} is a language with a high proportion of non-empty documents in $D = \left\{d_1,..., d_{\ell}\right\}$, which in case of Wikipedia is English.
-We use the following notation to define subsets
-of the multilingual comparable corpus: let $a(i,j)$ denote the index set of all multilingual documents with non-missing data for the $i$-th and $j$-th language:  $a(i,j) = \left\{k | d_k = (u_1,...,u_m), u_i \neq \emptyset, u_j \neq \emptyset \right\},$ and let $a(i)$ denote the index set of all multilingual documents with non missing data for the $i$-th language.
+We use the following notation to define subsets of the multilingual comparable corpus: let $a(i,j)$ denote the index set of all multilingual documents with non-missing data for the $i$-th and $j$-th language:  $a(i,j) = \left\{k | d_k = (u_1,...,u_m), u_i \neq \emptyset, u_j \neq \emptyset \right\},$ and let $a(i)$ denote the index set of all multilingual documents with non missing data for the $i$-th language.
 
-{\oldText{Taken from XLite}
+We now describe a two step approach to building a cross-lingual similarity matrix.
+The first part is related to LSI and reduces the dimensionality of the data.
+The second step refines the linear mappings and optimizes linear dependence between data.
 
 The first step in our method is to project $X_1, \ldots, X_m$ to lower dimensional spaces without destroying the cross-lingual structure.
-Treating the columns of $X_i$ as observation vectors sampled from an underlying distribution $\mathcal{X}_i \in V_i = \RR^{n_i}$, we can analyze the empirical cross-covariance matrices: $C_{i,j} = \frac{1}{|a(i,j)|-1 }\sum_{\ell \in a(i,j)} (X_i^{\ell} - c_i)\cdot (X_j^{\ell} - c_j)^T$, where $c_i = \frac{1}{a_i} \sum_{\ell \in a(i)}X_i^{\ell}$.
+Treating the nonzero columns of $X_i$ as observation vectors sampled from an underlying distribution $\mathcal{X}_i \in V_i = \RR^{n_i}$, we can analyze the empirical cross-covariance matrices: $C_{i,j} = \frac{1}{|a(i,j)|-1 }\sum_{\ell \in a(i,j)} (X_i^{\ell} - c_i)\cdot (X_j^{\ell} - c_j)^T$, where $c_i = \frac{1}{a_i} \sum_{\ell \in a(i)}X_i^{\ell}$.
 By finding low rank approximations of $C_{i,j}$ we can identify the subspaces of $V_i$ and $V_j$ that are relevant for extracting linear patterns between $\mathcal{X}_i$ and $\mathcal{X}_j$.
 Let $X_1$ represent the hub language corpus matrix.
-The standard approach to finding the subspaces is to perform the singular value decomposition on the full $N \times N$ covariance matrix composed of blocks $C_{i,j}$.
+The LSI approach to finding the subspaces is to perform the singular value decomposition on the full $N \times N$ covariance matrix composed of blocks $C_{i,j}$.
 If $|a(i,j)|$ is small for many language pairs (as it is in the case of Wikipedia), then many empirical estimates $C_{i,j}$ are unreliable, which can result in overfitting.
@@
 We proceed by reducing the dimensionality of each $X_i$ by setting: $Y_i = V_i^T \cdot X_i$, where $Y_i \in \RR^{k\times N}$.
-The step is similar to Cross-lingual Latent Semantic Indexing (CL-LSI) \cite{lsi}\cite{cl_lsi} which is less suitable due to a large amount of missing documents.
-Since singular value decomposition with missing values is a challenging problem and the alternative of replacing missing documents with zero vectors can result in degradation of performance.
+To summarize, the first step reduces the dimensionality of the data and is based on CL-LSI, but optimizes only the hub-language related cross-covariance blocks.
 
@@
 The approach is based on the sum of squares of correlations formulation by Kettenring \cite{Kettenring}, where we consider only correlations between pairs $(Y_1, Y_i), i >1$ due to the hub language problem characteristic.
-Let $D_{i,i} \in \RR^{k \times k}$ denote the empirical covariance and $D_{i,j}$ denote the empirical cross-covariance computed based on $Y_i$ and $Y_j$.
+Let $D_{i,i} \in \RR^{k \times k}$ denote the empirical covariance of $\mathcal{Y}_i$ and $D_{i,j}$ denote the empirical cross-covariance computed based on $\mathcal{Y}_i$ and $\mathcal{Y}_j$.
 We solve the following optimization problem:
@@
 The final projections that enable mappings to a common space are defined as: $P_i(x) = W_i^T V_i^T x.$
-}
 
@@
 100 languages!
+Describe the newsfeed volume, number of events and the number of similarities that need to be computed. Compare the memory requirements of our current implementation to the rest.
+Comment on the caching gains in our implementation.
 
@@
 \begin{algorithm}[t!]
-\SetKwInOut{Input}{input}\SetKwInOut{Output}{output}
-\Input{test cluster $c_i$,\\a set of clusters $C_l$ for each language $l \in L$}
-\Output{a set of clusters $C$ that are potentially equivalent to $c_i$}
+\SetKwInput{KwInput}{input}\SetKwInput{KwOutput}{output}
+\KwInput{test cluster $c_i$,\\a set of clusters $C_l$ for each language $l \in L$}
+\KwOutput{a set of clusters $C$ that are potentially equivalent to $c_i$}
 $C = \{\}$\;
@@
 \begin{algorithm}[tb!]
-\SetKwInOut{Input}{input}\SetKwInOut{Output}{output}
-\Input{test cluster $c_i$,\\
+\SetKwInput{KwInput}{input}\SetKwInput{KwOutput}{output}
+\KwInput{test cluster $c_i$,\\
 a set of candidate clusters $C$ that are potentially equivalent to $c_i$\\
 classification model $M$ trained on the hand-labeled examples}
-\Output{a set of clusters $C'$ that are equivalent to $c_i$}
+\KwOutput{a set of clusters $C'$ that are equivalent to $c_i$}
 

### 2015-03-17 rupnik2015.tex
xling section

@@
 In this section we define document similarity in a cross-lingual setting, where the similarity function receives documents in different languages.
-We conclude the section by an introduction of two datasets which we used in this paper to learn cross-lingual similarity functions.
+We conclude the section by an introduction of \todo{two datasets} which we used in this paper to learn cross-lingual similarity functions.
 
@@
 
-We then run the $k$-means algorithm \cite{kmeans} and obtain a set of $k$ centroid vectors, which form a matrix: $C := \begin{bmatrix}C^1, \ldots, C^k \end{bmatrix}$.
+We then run the $k$-means algorithm \cite{kmeans} and obtain a centroid matrix $C \in \RR^{N \times k}$, where the $k$ columns represent centroid vectors.
 The centroid matrix can be split vertically into $m$ blocks: $$C = [C_1^T \cdots C_m^T]^T,$$ according to the number of dimensions of each language, i.e.
-$C_i \in \RR^{n_i}$.
+$C_i \in \RR^{n_i \times k}$.
 
@@
 \subsection{Canonical Correlation Analysis}
- For completeness we now introduce a statistical technique to analyise data from two sources - after which we will describe our main contribution to computing cross-lingual similarities.
+ For completeness we now introduce a statistical technique to analyze data from two sources - after which we will describe our main contribution to computing cross-lingual similarities.
 
@@
 
-Interpreting the nonzero columns of $X_i$ as observation vectors sampled from an underlying distribution $\mathcal{X}_i \in \RR^{n_i}$, the idea is to find two weight vectors $w_i \in \RR^{n_i}$ and $w_j \in \RR^{n_j}$ so that the random variables $w_i^T \cdot \mathcal{X_i}$ and $w_j^T \cdot \mathcal{X_j}$ are maximally correlated ($w_i$ and $w_j$ are used to map the random vectors to random variables, by computing weighted sums of vector components).
+Interpreting the columns of $X_i$ as observation vectors sampled from an underlying distribution $\mathcal{X}_i \in \RR^{n_i}$, the idea is to find two weight vectors $w_i \in \RR^{n_i}$ and $w_j \in \RR^{n_j}$ so that the random variables $w_i^T \cdot \mathcal{X}_i$ and $w_j^T \cdot \mathcal{X}_j$ are maximally correlated ($w_i$ and $w_j$ are used to map the random vectors to random variables, by computing weighted sums of vector components).
+Let $\rho(x,y)$ denote the sample based correlation coefficient between two vectors of observations $x$ and $y$.
 By using the sample matrix notation $X_i$ and $X_j$ (assuming no data is missing for clearer presentation) this problem can be formulated as the following optimization problem:
@@
 & \underset{w_i \in \RR^{n_i}, w_j \in \RR^{n_j}}{\text{maximize}}
-& & \frac{w_i^T C_{i,j} w_j}{\sqrt{w_i^T C_{i,i} w_i} \sqrt{w_j^T C_{j,j} w_j}},
+& & \rho(w_i^T X_i , w_j^T X_j) = \frac{w_i^T C_{i,j} w_j}{\sqrt{w_i^T C_{i,i} w_i} \sqrt{w_j^T C_{j,j} w_j}},
 \end{aligned}
 \end{equation*}
+
 where $C_{i,i}$ and $C_{j,j}$ are empirical estimates of variances of $\mathcal{X}_i$ and $\mathcal{X}_j$ respectively and $C_{i,j}$ is an estimate for the covariance matrix.
@@
 For this reason we perform the truncated singular value decomposition on the matrix $C = [C_{1,2}  \cdots  C_{1,m}] \approx U S V^T$, where where $U \in \RR^{n_1 \times k}, S \in \RR^{k \times k}, V \in \RR^{(\sum_{i=2}^m n_i) \times k}$.
-We split the matrix $V$ vertically in blocks with $n_2, \ldots, n_M$ rows to obtain a set of matrices $V_i$: $V = [V_2^T  \cdots  V_m^T]^T$.
+We split the matrix $V$ vertically in blocks with $n_2, \ldots, n_m$ rows: $V = [V_2^T  \cdots  V_m^T]^T$.
 Note that columns of $U$ are orthogonal but columns in each $V_i$ are not (columns of V are orthogonal).
@@
 The approach is based on the sum of squares of correlations formulation by Kettenring \cite{Kettenring}, where we consider only correlations between pairs $(Y_1, Y_i), i >1$ due to the hub language problem characteristic.
+The original sum of squared correlation is formulated as:
+\begin{equation*}
+  \begin{aligned}
+    & \underset{w_i \in \RR^{k}}{\text{maximize}}
+    & & \sum_{i < j}^m  \rho(w_i^T Y_i, w_j^T Y_j)^2.
+\end{aligned}
+\end{equation*}
+We solve a similar problem by restricting $i=1$ and omit optimizing over non-hub language pairs.
 Let $D_{i,i} \in \RR^{k \times k}$ denote the empirical covariance of $\mathcal{Y}_i$ and $D_{i,j}$ denote the empirical cross-covariance computed based on $\mathcal{Y}_i$ and $\mathcal{Y}_j$.

### 2015-03-18 rupnik2015.tex
update

@@
 \begin{abstract}
-\todo{tralala hopsasa.}
+Main points:
+\begin{itemize}
+\item Problem of tracking events in a large multilingual stream - we reduce it to event linking
+\item Two contributions. We propose a method that enables cross-lingaul document similarity an approach to linking multilingual data and an approach that uses is to link events.
+\item Evaluation of cross-lingual approaches on Wikipedia and the event linking task.
+\end{itemize}
+
 \end{abstract}
@@
 
+We will present three approaches: a simple approach based on $k$-means clustering, a standard approach based on singular value decomposition and a new method that is applicable in more general
+settings. All approaches are based on linear maps and we do not focus on alternatives, such as machine translation and probabilistic models. Applicability of machine translation relies
+on the availability of training resources (strongly aligned translation corpora), which are not available for many world languages, but there exist comparable corpora. Probabilistic
+methods \todo{provide reasons for not using them: e.g. difficulty with parameter estimation}.
+
 \subsubsection{$k$-means}
@@
 The approach is based on the sum of squares of correlations formulation by Kettenring \cite{Kettenring}, where we consider only correlations between pairs $(Y_1, Y_i), i >1$ due to the hub language problem characteristic.
-The original sum of squared correlation is formulated as:
+We will present the original unconstrained optimization problem, then a constrained formulation based on the hub language problem characteristic. Then we will simplify the constraints and reformulate
+the problem as an eigenvalue problem by using the method of Lagrange multipliers.
+
+The original sum of squared correlation is formulated as an unconstrained problem:
 \begin{equation*}
@@
 Let $D_{i,i} \in \RR^{k \times k}$ denote the empirical covariance of $\mathcal{Y}_i$ and $D_{i,j}$ denote the empirical cross-covariance computed based on $\mathcal{Y}_i$ and $\mathcal{Y}_j$.
-We solve the following optimization problem:
+We solve the following constrained (unit variance constraints) optimization problem:
 \begin{equation*}
@@
 \end{equation*}
-By using Lagrangian multiplier techniques (we omit the derivation due to space constraints \todo{Add what was omitted}), the problem can be reformulated as the eigenvalue problem:$\left(\sum_{i=2}^m G_i G_i^T\right) \cdot V = \Lambda \cdot V,$ where $G_i = H_1^T D_{1,i} H_i$ and $H_i = Chol(D_{i,i})^{-1}$ where $Chol(\cdot)$ is the Cholesky decomposition: $X = Chol(X)^T Chol(X)$.
+The constraints $w_i^T D_{i,i} w_i$ can be simplified by using the Cholesky decomposition $D_{i,i} = K_i^T \cdot K_i$ and substitution: $y_i := K_i w_i$.
+By inverting the $K_i$ matrices and defining  $G_i := K_i^{-T} D_{1,i} K_i^{-1}$, the problem can be reformulated:
+\begin{equation*}
+  \begin{aligned}
+    & \underset{y_i \in \RR^{k}}{\text{maximize}}
+    & & \sum_{i = 2}^m  \left(y_1^T G_{i} y_i \right)^2
+    & \text{subject to}
+    & & y_i^T y_i = 1, \quad\forall i = 1,\ldots, m.
+\end{aligned}
+\end{equation*}
+
+\todo{Finish derivation: Lagrange multipliers, reformulation, backsubstitution, revisit commented text below}
+ %The problem can be reformulated as the eigenvalue problem:$\left(\sum_{i=2}^m G_i G_i^T\right) \cdot V = \Lambda \cdot V,$ where $G_i = H_1^T D_{1,i} H_i$ and $H_i = Chol(D_{i,i})^{-1}$ where $Chol(\cdot)$ is the Cholesky decomposition: $X = Chol(X)^T Chol(X)$.
 The vectors $w_i$ can be reconstructed from the dominant eigenvector $v$, as: $w_1 = H_1 v$  and $w_i \propto H_i  G_i^T  v$ for $i >1$ ($w_i$ need to be appropriately normalized).
@@
 $W_i$ can be extracted from a set of dominant eigenvectors of matrix $V$ according to eigenvalues $\Lambda$.
+
  The technique is related to a generalization of canonical correlation analysis (GCCA) by Carroll\cite{Carroll}, where an unknown group configuration variable is defined and objective is to maximize the sum of squared correlation between the group variable and the others.

### 2015-03-18 rupnik2015.tex
new title and intro

@@
 \jairheading{x}{xxxx}{x-xx}{x/xx}{x/xx}
-\ShortHeadings{Cross-lingeringual Document Similarity and Event Detection}
+\ShortHeadings{Cross-Lingual Document Similarity and Event Tracking}
 {Rupnik, Muhi\v{c}, Leban, Fortuna, \v{S}kraba \& Grobelnik}
@@
 
-\title{Cross-lingual Document Similarity and Event Detection \todo{Better title...}}
+\title{Event Tracking Across Languages - 
+Cross-Lngual Document Similarity and Event Tracking \todo{Better title...}}
 
@@
 \end{itemize}
+
+\section{Introduction}
+
+Content on the internet is becoming increasingly multi-lingual.
+A prime example is Wikipedia - in 2001 the majority of pages were written in English, while in 2012, the percentage of English articles has dropped to 14\%.
+Machine translation remains relatively rudimentary - allowing people to understand simple phrases on we pages, but remain inadequate for more advanced understanding of text.
+At the same time, online news has begun to dominate reporting of current events.
+In this paper we consider the intersection of these developments: how to track events which are reported about in multiple languages.
+
+The term event is vague and ambiguous, but for the practical purposes we define it as ``any significant happening that is being reported about in the media''.
+Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015.
+Events such as these are covered by many articles and the question is how to find all the articles in different languages that are describing a single event.
+
+As input we consider a stream of articles in different languages and a list of events.
+Our goal is to assign articles to their corresponding events.
+A priori we do not know the coverage of the articles, that is not all the events may be covered and we do not know that all the articles necessarily fit into one of the events.
+To achieve our goal, we divide the task into two parts: clustering the articles according to topic (regardless of language) and matching the clusters to events.
+
+To cluster the articles, rather than rely on machine translations, we adopt well-established monolingual text mining tools to the multilingual setting.
+This represents the continuation of previous work \cite{nips, iti, nips2, sikdd} where we explored representations of documents which were valid over multiple languages.
+The representations could be interpreted as multi-lingual topics, which were then used as proxies to compute cross-lingual similarities between documents.
+To learn the representations, we use Wikipedia as a training corpus.
+SIgnificantly, we do not only consider the major or \emph{hub} languages such as English, German, French, etc.
+which have significant overlap in article coverage, but also smaller languages (in terms of number of Wikipedia articles) such as Slovenian and Hindi, which may have a negligible overlap in article coverage.
+We can then define a similarity between any two articles regardless of language,  which allows us to cluster the articles according to topic.
+The underlying assumption being that articles describing the same event are similar  and will therefore be put into the same cluster.
+
+The second step is to match our computed clusters to events. Since our similarity measure and clustering 
+may not be perfect, we need to find the ``best" correspondence between clusters and events.
+In this paper, we test several different algorithms for both steps for perfomance in an integrated system.
+In particular, we test our approach within an online system for detection of world events, called Event Registry\cite{}.
+Event Registry is a repository of events, where events are automatically identified by analyzing news articles.
+
+Event Registry works by collecting and analyzing news articles that are published by news outlets in different languages all over the world.
+Collected articles are first semantically annotated by identifying in them mentions of relevant concepts -- either entities or important keywords.
+The disambiguated and entity linking of the concepts is done using Wikipedia as the main knowledge base.
+   The key step in Event Registry is then to apply an online clustering algorithm on the articles in order to identify groups of articles that are discussing the same event.
+
+   Within this system, we test our multilingual similarity measures as well as our algorithms for matching clusters to events.
+The result is that we can test total performance in a multilingual, real-world setting.
+
+The paper is organized as follows: we first provide an overview of the system as a whole (Section ), followd by details on each of the two steps described above.
+We first introduce several approaches to measures of cross-lingual similarity (Section ) and then the algorithms for matching clusters to events.
+To improve readability, we describe relevant related work for each module in its corresponding Section.
+Finally, we present and interpret the experimental results and finally give conclusions and describe promising future directions.
 

### 2015-03-19 rupnik2015.tex
derivation of the cca based method

--- 2015-03-18 rupnik2015.tex
@@
 We solve the following constrained (unit variance constraints) optimization problem:
-\begin{equation*}
+\begin{equation}\label{squaredCorHubOriginal}
   \begin{aligned}
@@
 \end{aligned}
-\end{equation*}
+\end{equation}
 The constraints $w_i^T D_{i,i} w_i$ can be simplified by using the Cholesky decomposition $D_{i,i} = K_i^T \cdot K_i$ and substitution: $y_i := K_i w_i$.
 By inverting the $K_i$ matrices and defining  $G_i := K_i^{-T} D_{1,i} K_i^{-1}$, the problem can be reformulated:
-\begin{equation*}
+\begin{equation}\label{squaredCorHub}
   \begin{aligned}
@@
 \end{aligned}
-\end{equation*}
+\end{equation}
+A necessary condition for optimality is that the derivatives of the Lagrangian vanish. The Lagrangian of (\ref{squaredCorHub}) is expressed as:
+$$  L(y_1, \ldots, y_m, \lambda_1, \ldots, \lambda_m) = \sum_{i = 2}^m  \left(y_1^T G_{i} y_i \right)^2 + \sum_{i=1}^m \lambda_i \left(y_i^T y_i - 1\right).$$
+Stationarity conditions give us:
+\begin{equation}\label{dLdx1}
+ \frac{\partial}{\partial x_1} L = 0 \Rightarrow \sum_{i = 2}^m  \left(y_1^T G_{i} y_i \right) G_i y_i + \lambda_1 y_1,
+\end{equation}
+\begin{equation}\label{dLdxi}
+\frac{\partial}{\partial x_i} L = 0 \Rightarrow \left(y_1^T G_{i} y_i \right) G_{i}^T y_1 + \lambda_i y_i = 0,~i > 1.
+\end{equation}
+Multiplying the equations (\ref{dLdxi}) with $y_i^T$ and applying the constraints, we can eliminate $\lambda_i$ which gives us:
+\begin{equation}\label{eqy1yi}
+G_{i}^T y_1 = \left(y_1^T G_{i} y_i \right) y_i,~i > 1.
+\end{equation}
+Plugging this into (\ref{dLdx1}), we obtain an eigenvalue problem:
+$$\left( \sum_{i = 2}^m G_i G_{i}^T \right) y_1 + \lambda_1 y_1 = 0.$$
+The eigenvectors of $\left( \sum_{i = 2}^m G_i G_{i}^T \right)$ solve the problem for the first language.
+The solutions for $y_i$ are obtained from (\ref{eqy1yi}): $y_i := \frac{G_{i}^T y_1}{\| G_{i}^T y_1 \|}$.
+Note that the solution (\ref{squaredCorHubOriginal}) to can be recovered by: $w_i := K_i^{-1} y_i$. The linear transformation of the $w$ variables are thus expressed as:
+$$W_1 = $$
+$$W_i = $$
 
-\todo{Finish derivation: Lagrange multipliers, reformulation, backsubstitution, revisit commented text below}
- %The problem can be reformulated as the eigenvalue problem:$\left(\sum_{i=2}^m G_i G_i^T\right) \cdot V = \Lambda \cdot V,$ where $G_i = H_1^T D_{1,i} H_i$ and $H_i = Chol(D_{i,i})^{-1}$ where $Chol(\cdot)$ is the Cholesky decomposition: $X = Chol(X)^T Chol(X)$.
-The vectors $w_i$ can be reconstructed from the dominant eigenvector $v$, as: $w_1 = H_1 v$  and $w_i \propto H_i  G_i^T  v$ for $i >1$ ($w_i$ need to be appropriately normalized).
-Higher dimensional mappings $W_i$ are obtained in a similar way by setting the constraint $W_i^T D_{i,i} W_i = I$, where $I$ is the identity matrix.
-The constraint forces the columns of $W_i$ to be uncorrelated (orthogonal with respect to covariance matrix).
-$W_i$ can be extracted from a set of dominant eigenvectors of matrix $V$ according to eigenvalues $\Lambda$.
-
- The technique is related to a generalization of canonical correlation analysis (GCCA) by Carroll\cite{Carroll}, where an unknown group configuration variable is defined and objective is to maximize the sum of squared correlation between the group variable and the others.
+ The technique is related to the Generalization of Canonical Correlation Analysis (GCCA) by Carroll\cite{Carroll}, where an unknown group configuration variable is defined and objective is to maximize the sum of squared correlation between the group variable and the others.
 The problem can be reformulated as an eigenvalue problem.

### 2015-03-20 rupnik2015.tex
restructured

@@
 \item Introduction
-\item Cross lingual document similarity (model)
-\item  Event Detection
-\item  Combing the approaches
+\item System overview
+\item Cross lingual document similarity (algorithms)
+\item Event linking
 \item Experiments
@@
 \begin{itemize}
-\item Description of the problem - we would liek to understand events ans sequences of events, but our observables (e.g. news stories) occur in a variety of languages
+\item Description of the problem - we would like to understand global news as sequences of events, but our observables (e.g. news stories) occur in a variety of languages
 \item We present an approach to tackle this problem using a combination of techniques....
@@
 To learn the representations, we use Wikipedia as a training corpus.
-SIgnificantly, we do not only consider the major or \emph{hub} languages such as English, German, French, etc.
+Significantly, we do not only consider the major or \emph{hub} languages such as English, German, French, etc.
 which have significant overlap in article coverage, but also smaller languages (in terms of number of Wikipedia articles) such as Slovenian and Hindi, which may have a negligible overlap in article coverage.
@@
 
-\section{Introduction}
+\section{System Overview}
 
-{\oldText{Taken from XLite paper}
-Document retrieval is a well-established problem in data mining.
-There have been a large number of different approaches put forward in the literature.
-In this paper we concentrate on a specific setting: multi-lingual corpora.
-As the availability of multi-lingual documents has exploded in the last few years, there is a pressing need for automatic cross-lingual processing tools.
-
-The prime example is Wikipedia - in 2001 the majority of pages were written in English, while in 2012, the percentage of English articles has dropped to 14\%.
-In this context, we look at how to find similar documents across languages.
-In particular, we do not assume the availability of machine translators, but rather try to frame the problem such that we can use well-established machine learning tools designed for monolingual text-mining tasks.
-This work represents the continuation of previous work \cite{nips, iti, nips2, sikdd} where we explored representations of documents which were valid over multiple languages.
-The representations could be interpreted as multi-lingual topics, which were then used as proxies to compute cross-lingual similarities between documents.
-We look at a specific aspect of this problem: the distribution of articles across languages in Wikipedia is not uniform.
-While the percentage of English articles has fallen, English is still the largest language and one of \emph{hub} languages which not only have an order of magnitude more articles than other languages, but also many comparable articles with most of the other languages.
-
-When doing document retrieval, we encounter two quite different situations.
-If for example, we are looking for a German article comparable to an English article, there is a large alignment between the document corpora (given by the intersection in articles in the two languages) making the problem well-posed.
-If, however, we look for a relevant Slovenian article given a Hindi article, the intersection is small, making the problem much harder.
-Since almost all languages have a large intersection in articles with \emph{hub} languages, the question we ask in this paper is: can we exploit hub languages to perform better document retrieval between non-hub languages?
-A positive answer would vastly improve cross-lingual analysis between less represented languages.
-In the following section, we introduce our representation followed by our experiments, which also shed light on the structural properties of the multilingual Wikipedia corpus.
-}
+Event registry -> Monolingual events
+Events -> features (semantic + CCA) -> linked events
 
@@
 We conclude the section by an introduction of \todo{two datasets} which we used in this paper to learn cross-lingual similarity functions.
+
+\subsection{Related work}
+
+TODO
 
@@
 
-\subsection{Datasets}
-
-{\oldText{Taken from ITI paper}
-To investigate the empirical performance of the low rank approximations we will test the algorithms on a large-scale, real-world multilingual dataset that we extracted from Wikipedia by using inter-language links as an alignment.
-This  results in a large number of weakly comparable documents in more than $200$ languages.
-Wikipedia is a large source of multilingual data that is especially important for the languages for which no translation tools, multilingual dictionaries as Eurovoc, or strongly aligned multilingual corpora as Europarl are available.
-Documents in different languages are related with so called 'inter-language' links that can be found on the left of the Wikipedia page.
-The Wikipedia is constantly growing.
-There are currently four Wikipedias with more than $10^6$ articles, $40$ with more than $10^5$ articles, $100$ with more than $10^4$ articles, and $216$ with more than $1000$ articles.
-Wikipedia uses special user-friendly markup language that is very easy to write but very hard to parse.
-Simplicity of language can cause ambiguities and moreover it is constantly changing.
-For example, separator | can be used in different contexts.
-
-Wikipedia raw xml dumps of all currently $270$ active editions were downloaded from the Wikipedia dump page\href{http://dumps.wikimedia.org}.
-The xml files are too large to be parsed with DOM like parser that needs to store the whole xml tree in the memory, instead we implemented Sax like parser that tries to simulate behavior of Wikipedia official parser and is as simple, fast and error prone as possible.
-We parse all Wikipedia markup but do not extend the templates.
-Each Wikipedia page is embedded in the page tag.
-First we check if the title of the page consists of any special namespace and do not process such pages.
-Then we check if this is a redirection page and we store the redirect link as inter-language links can point to redirection link also.
-If nothing of the above applies we extract the text and parse the Wikipedia markup.
-Currently all the markup is removed.
-
-We get inter-language link matrix using previously stored redirection links and inter-language links.
-If inter-language link points to the redirection we replace it with the redirection target link.
-It turns out that we obtain the matrix $M$ that is not symmetric, consequently the underlying graph is not symmetric.
-That means that existence of the inter-language link in one way (i.e.
-English to German) does not guarantee that there is an inter-language link in the reverse direction (German to English).
-To correct this we transform this matrix to symmetric by computing $M+M^T$ and obtaining an undirected graph.
-In the rare case that we have multiple links pointing from the document, we pick the first one that we encountered.
-This matrix enables us to build an alignment across all Wikipedia languages.
-}
-
 \section{Multilingual Latent Factor Models}\label{sec:models}
 
-\subsection{Multilingual dataset}
+We will fist introduce some notation regarding the training data, which we will use in the next subsections.
+
+We will then present three approaches: a simple approach based on $k$-means clustering, a standard approach based on singular value decomposition and a new method that is applicable in more general
+settings. All approaches are based on linear maps and we do not focus on alternatives, such as machine translation and probabilistic models. Applicability of machine translation relies
+on the availability of training resources (strongly aligned translation corpora), which are not available for many world languages, but there exist comparable corpora. Probabilistic
+methods \todo{provide reasons for not using them: e.g. difficulty with parameter estimation}.
+
+For completeness we will also present the method of Canonical Correlation Analysis, since it is related to our proposed method.
+
+\subsection{Notation}
 
@@
 
-\subsection{Algorithms}
-\label{sec:algorithms}
+We now present the first approach.
 
-We will present three approaches: a simple approach based on $k$-means clustering, a standard approach based on singular value decomposition and a new method that is applicable in more general
-settings. All approaches are based on linear maps and we do not focus on alternatives, such as machine translation and probabilistic models. Applicability of machine translation relies
-on the availability of training resources (strongly aligned translation corpora), which are not available for many world languages, but there exist comparable corpora. Probabilistic
-methods \todo{provide reasons for not using them: e.g. difficulty with parameter estimation}.
-
-\subsubsection{$k$-means}
+\subsection{$k$-means}
 
@@
 
-\subsubsection{Cross-Lingual Latent Semantic Indexing}
+\subsection{Cross-Lingual Latent Semantic Indexing}
 
@@
 The constraints $w_i^T D_{i,i} w_i$ can be simplified by using the Cholesky decomposition $D_{i,i} = K_i^T \cdot K_i$ and substitution: $y_i := K_i w_i$.
-By inverting the $K_i$ matrices and defining  $G_i := K_i^{-T} D_{1,i} K_i^{-1}$, the problem can be reformulated:
+By inverting the $K_i$ matrices and defining  $G_i := K_1^{-T} D_{1,i} K_i^{-1}$, the problem can be reformulated:
 \begin{equation}\label{squaredCorHub}
@@
 Note that the solution (\ref{squaredCorHubOriginal}) to can be recovered by: $w_i := K_i^{-1} y_i$. The linear transformation of the $w$ variables are thus expressed as:
-$$W_1 = $$
-$$W_i = $$
+$$ Y_1 := \text{eigenvectors of} \sum_{i = 2}^m G_i G_{i}^T, $$
+$$ W_1 = K_1^{-1} Y_1 $$
+$$ W_i = K_i^{-1} G_{i}^T Y_1 N,$$
+where $N$ is a diagonal matrix that normalizes $G_{i}^T Y_1$, with $N(j,j) := \frac{1}{\|G(_{i} Y_1(:,j)\|}$.
 
@@
 The final projections that enable mappings to a common space are defined as: $P_i(x) = W_i^T V_i^T x.$
-
-\subsection{Scaling}
-
-100 languages!
-Describe the newsfeed volume, number of events and the number of similarities that need to be computed. Compare the memory requirements of our current implementation to the rest.
-Comment on the caching gains in our implementation.
-
-\subsection{Evaluation}
-
-\subsubsection{Link prediction}
-
-\subsubsection{Hub language}
-
-{\oldText{Taken from XLite paper}
-To investigate the empirical performance of our algorithm we will test it on a large-scale, real-world multilingual dataset that we extracted from Wikipedia  by  using so called 'inter-language' links as an alignment.
-We select a subset of Wikipedia languages containing three major languages, English--\emph{en} (hub language), Spanish--\emph{es}, Russian--\emph{ru}, and five minority (in the sense of Wikipedia sizes) languages, Slovenian--\emph{sl}, Piedmontese--\emph{pms}, Waray-Waray--\emph{war} (all with about 2 million native speakers), Creole--\emph{ht} (8 million native speakers), and Hindi--\emph{hi} (180 million native speakers).
-For preprocessing we remove the documents that contain less than 20 different words (stubs) and remove words occur in less than 50 documents as well as the top 100 most frequent words (in each language separately).
-We represent the documents as normalized TFIDF\cite{Salton88term-weightingapproaches} weighted vectors.
-These are languages with Wikipedia article sizes comparable to the minority languages.
-(since the number of speakers does not directly correlate with the number of Wikipedia articles).
-The prime hub candidate is the English language which is well aligned with all other Wikipedia languages, although the alignment quality varies quite a bit.
-Furthermore, we remove the documents that contain less than 20 different words and remove words that are too infrequent as well as the top 100 most frequent words in the vocabularies.
-Furthermore, we call the document consisting of less than 20 different words, a stub.
-This documents are typically garbage, the titles of the columns in the table, remains of the parsing process, or Wikipedia articles with very little or no information contained in one or two sentences.
-
-The evaluation is based on splitting the data into training and test sets (described later).
-On the training set, we perform the two step procedure to obtain the common document representation as a set of mappings $P_i$.
-A test set for each language pair, $test_{i,j} = \{(x_\ell,y_\ell) | \ell = 1:n(i,j)\} $, consists of comparable document pairs (linked Wikipedia pages), where $n(i,j)$ is the test set size.
-We evaluate the representation by measuring mate retrieval quality on the test sets: for each $\ell$, we rank the projected documents $P_j(y_1),\ldots, P_j(y_{n(i,j)})$ according to their similarity with $P_i(x_\ell)$ and compute the rank of the mate document $r(\ell) = rank(P_j(y_\ell))$.
-The final retrieval score (between -100 and 100) is computed as: $\frac{100}{n(i,j)} \cdot \sum_{\ell = 1}^{n(i,j)} \left( \frac{n(i,j) - r(\ell)}{n(i,j) -1} -0.5\right)$.
-A score that is less than 0 means that the method performs worse than random retrieval and a score of 100 indicates perfect mate retrieval.
-The mate retrieval results are included in Table \ref{table:retrieval}.
-
-We observe that the method performs well on all pairs between languages: \emph{en}, \emph{es}, \emph{ru}, \emph{sl}, where at least 50,000 training documents are available.
-We notice that taking $k = 500$ or $k = 1000$ multilingual topics usually results in similar performance, with some notable exceptions: in the case of (\emph{ht}, \emph{war}) the additional topics result in an increase in performance, as opposed to (\emph{ht},\emph{pms}) where performance drops, which suggests overfitting.
-The languages where the method performs poorly are \emph{ht} and \emph{war}, which can be explained by the quality of data (see Table \ref{table:rank} and explanation that follows).
-In case of \emph{pms}, we demonstrate that solid performance can be achieved for language pairs (\emph{pms}, \emph{sl}) and (\emph{pms}, \emph{hi}), where only 2000 training documents are shared between \emph{pms} and \emph{sl} and no training documents are available between \emph{pms} and \emph{hi}.
-Also observe that in the case of (\emph{pms}, \emph{ht}) the method still obtains a score of 62, even though training set intersection is zero and \emph{ht} data is corrupted, which we will show in the next paragraph.
-{
-\renewcommand\tabcolsep{3pt}
-\begin{table}[h!]
-\caption{Pairwise retrieval, 500 topics;1000 topics}\label{table:retrieval}
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
-\cline{1-9}
-&	en&	es&	ru&	sl&	hi&	war&	ht&	pms\\\cline{1-9}
-en&	    &	98~-~98&	95~-~97&	97~-~98&	82~-~84&	76~-~74&	53~-~55&	 96~-~97\\
-\cline{1-9}
-es&	97~-~98&	&	94~-~96&	97~-~98&	85~-~84&	76~-~77&	56~-~57&	96~-~96\\
-\cline{1-9}
-ru&	96~-~97&	94~-~95&	&	97~-~97&	81~-~82&	73~-~74&	55~-~56&	96~-~96\\
-\cline{1-9}
-sl&	96~-~97&	95~-~95&	95~-~95&	&	91~-~91&	68~-~68&	59~-~69&	93~-~93\\
-\cline{1-9}
-hi&	81~-~82&	82~-~81&	80~-~80&	91~-~91&	&	68~-~67&	50~-~55&	87~-~86\\
-\cline{1-9}
-war&	68~-~63&	71~-~68&	72~-~71&	68~-~68&	66~-~62&	&	28~-~48&	 24~-~21\\
-\cline{1-9}
-ht&	52~-~58&	63~-~66&	66~-~62&	61~-~71&	44~-~55&	16~-~50&	&	62~-~49\\
-\cline{1-9}
-pms&	95~-~96&	96~-~96&	94~-~94&	93~-~93&	85~-~85&	23~-~26&	66~-~54&	 \\
-\cline{1-9}
-\end{tabular}
-\end{center}
-\end{table}
-}
-
-We now describe the selection of train and test sets.
-We select the test set documents as all multi-lingual documents with at least one nonempty alignment from the list: (\emph{hi}, \emph{ht}), (\emph{hi}, \emph{pms}), (\emph{war}, \emph{ht}), (\emph{war}, \emph{pms}).
-This guarantees that we cover all the languages.
-Moreover this test set is suitable for testing the retrieval thorough the hub as the chosen pairs have empty alignments.
-The remaining documents are used for training.
-In Table \ref{table:train_test}, we display the corresponding sizes of training and test documents for each language pair.
-The first row represents the size of the training sets used to construct the mappings in low dimensional language independent space using the English--\emph{en} as a hub.
-The diagonal elements represent number of the unique training documents and test documents in each language.
-
-{
-\renewcommand\tabcolsep{3pt}
-\begin{table}[h!]
-\caption{Pairwise training:test sizes (in thousands)}
-\label{table:train_test}
-{
-\small
-\begin{tabular}{c|c|c|c|c|c|c|c|c|}
-&	en&	es&	ru&	sl&	hi&	war&	ht&	pms\\\cline{1-9}
-en&	671~-~4.64&	463~-~4.29&	369~-~3.19&	50.3~-~2&	14.4~-~2.76&	8.58~-~2.41&	 17~-~2.32&	16.6~-~2.67\\
-\cline{2-9}
-es&	\multicolumn{1}{c|}{}	&	463~-~4.29&	187~-~2.94&	28.2~-~1.96&	8.72~-~2.48&	 6.88~-~2.4&	13.2~-~2&	 13.8~-~2.58\\
-\cline{3-9}
-ru&	\multicolumn{2}{c|}{}	&	369~-~3.19&	29.6~-~1.92&	9.16~-~2.68&	2.92~-~1.1&	 3.23~-~2.2&	10.2~-~1.29\\
-\cline{4-9}
-sl&	\multicolumn{3}{c|}{}	&	50.3~-~2&	3.83~-~1.65&	1.23~-~0.986&	0.949~-~1.23&	 1.85~-~0.988\\
-\cline{5-9}
-hi&	\multicolumn{4}{c|}{}	&	14.4~-~2.76&	0.579~-~0.76&	0.0~-~2.08&	0.0~-~0.796\\
-\cline{6-9}
-war&	\multicolumn{5}{c|}{}	&	8.58~-~2.41&	0.043~-~0.534&	0.0~-~1.97\\
-\cline{7-9}
-ht&	\multicolumn{6}{c|}{}	&	17~-~2.32&	0.0~-~0.355\\
-\cline{8-9}
-pms&	\multicolumn{7}{c|}{}	&	16.6~-~2.67\\
-\cline{9-9}
-\end{tabular}
-}
-\end{table}
-}
-
-We further inspect the properties of the training sets by roughly estimating the fraction \texttt{rank(A)/min(size(A))} for each training English matrix and its corresponding mate matrix.
-Ideally, these two fractions are approximately the same so  both aligned spaces should have reasonably similar dimensionality.
-We display these numbers as pairs in Table \ref{table:rank}.
-
-\begin{table}[h]
-\caption{Dimensionality drift}
-\label{table:rank}
-\begin{tabular}{|c|c|c|c|c|c|c|}
-\cline{1-7}
-(en, de)     &   (en, ru)     &   (en, sl)       &     (en, hi)&   (en, war)      &      (en, ht) &   (en, pms)\\
-\cline{1-7}
-(0.81, 0.89)   &  (0.8, 0.89)  &   (0.98, 0.96)    &    (1, 1)  &   (0.74, 0.56)  &      (1, 0.22)  &   (0.89, 0.38)\\
-\cline{1-7}
-\end{tabular}
-\end{table}
-
-It is clear that in the case of Creole language only at most $22\%$ documents are unique and suitable for the training.
-Though we removed the stub documents, many of remaining documents are almost the same, as the quality of some minor Wikipedias is low.
-This was confirmed for Creole, Waray-Waray, and Piedmontese language by manual inspection.
-The low quality documents correspond to templates about the year, person, town, etc.
-and contain very few unique words.
-
-We also have a problem with the quality of the test data.
-For example, if we look at test pair (\emph{war}, \emph{ht}) only 386/534 Waray-Waray test documents are unique but on other side almost all Creole test documents (523/534) are unique.
-This indicates a poor alignment which leads to poor performance.
-}
 
@@
 Using this information we hope to be able to connect clusters in different languages and represent them using a single event ID.
+
+\subsection{Related work}
+
+TODO
 
@@
 
-\subsection{Evaluation}
+\section{Evaluation}
+
+We describe the datasets involved in the experiments and two evaluations: comparing the crosslingual approaches on an information retrieval task, and evaluation of cross-lingual event linking.
+
+\subsection{Datasets}
+
+{\oldText{Taken from ITI paper}
+To investigate the empirical performance of the low rank approximations we will test the algorithms on a large-scale, real-world multilingual dataset that we extracted from Wikipedia by using inter-language links as an alignment.
+This  results in a large number of weakly comparable documents in more than $200$ languages.
+Wikipedia is a large source of multilingual data that is especially important for the languages for which no translation tools, multilingual dictionaries as Eurovoc, or strongly aligned multilingual corpora as Europarl are available.
+Documents in different languages are related with so called 'inter-language' links that can be found on the left of the Wikipedia page.
+The Wikipedia is constantly growing.
+There are currently four Wikipedias with more than $10^6$ articles, $40$ with more than $10^5$ articles, $100$ with more than $10^4$ articles, and $216$ with more than $1000$ articles.
+Wikipedia uses special user-friendly markup language that is very easy to write but very hard to parse.
+Simplicity of language can cause ambiguities and moreover it is constantly changing.
+For example, separator | can be used in different contexts.
+
+Wikipedia raw xml dumps of all currently $270$ active editions were downloaded from the Wikipedia dump page\href{http://dumps.wikimedia.org}.
+The xml files are too large to be parsed with DOM like parser that needs to store the whole xml tree in the memory, instead we implemented Sax like parser that tries to simulate behavior of Wikipedia official parser and is as simple, fast and error prone as possible.
+We parse all Wikipedia markup but do not extend the templates.
+Each Wikipedia page is embedded in the page tag.
+First we check if the title of the page consists of any special namespace and do not process such pages.
+Then we check if this is a redirection page and we store the redirect link as inter-language links can point to redirection link also.
+If nothing of the above applies we extract the text and parse the Wikipedia markup.
+Currently all the markup is removed.
+
+We get inter-language link matrix using previously stored redirection links and inter-language links.
+If inter-language link points to the redirection we replace it with the redirection target link.
+It turns out that we obtain the matrix $M$ that is not symmetric, consequently the underlying graph is not symmetric.
+That means that existence of the inter-language link in one way (i.e.
+English to German) does not guarantee that there is an inter-language link in the reverse direction (German to English).
+To correct this we transform this matrix to symmetric by computing $M+M^T$ and obtaining an undirected graph.
+In the rare case that we have multiple links pointing from the document, we pick the first one that we encountered.
+This matrix enables us to build an alignment across all Wikipedia languages.
+}
+
+\subsection{Wikipedia Information Retrieval}
+
+Mate retrieval results
+
+\subsection{Experiments With Missing Alignment Data}
+
+{\oldText{Taken from XLite paper}
+To investigate the empirical performance of our algorithm we will test it on a large-scale, real-world multilingual dataset that we extracted from Wikipedia  by  using so called 'inter-language' links as an alignment.
+We select a subset of Wikipedia languages containing three major languages, English--\emph{en} (hub language), Spanish--\emph{es}, Russian--\emph{ru}, and five minority (in the sense of Wikipedia sizes) languages, Slovenian--\emph{sl}, Piedmontese--\emph{pms}, Waray-Waray--\emph{war} (all with about 2 million native speakers), Creole--\emph{ht} (8 million native speakers), and Hindi--\emph{hi} (180 million native speakers).
+For preprocessing we remove the documents that contain less than 20 different words (stubs) and remove words occur in less than 50 documents as well as the top 100 most frequent words (in each language separately).
+We represent the documents as normalized TFIDF\cite{Salton88term-weightingapproaches} weighted vectors.
+These are languages with Wikipedia article sizes comparable to the minority languages.
+(since the number of speakers does not directly correlate with the number of Wikipedia articles).
+The prime hub candidate is the English language which is well aligned with all other Wikipedia languages, although the alignment quality varies quite a bit.
+Furthermore, we remove the documents that contain less than 20 different words and remove words that are too infrequent as well as the top 100 most frequent words in the vocabularies.
+Furthermore, we call the document consisting of less than 20 different words, a stub.
+This documents are typically garbage, the titles of the columns in the table, remains of the parsing process, or Wikipedia articles with very little or no information contained in one or two sentences.
+
+The evaluation is based on splitting the data into training and test sets (described later).
+On the training set, we perform the two step procedure to obtain the common document representation as a set of mappings $P_i$.
+A test set for each language pair, $test_{i,j} = \{(x_\ell,y_\ell) | \ell = 1:n(i,j)\} $, consists of comparable document pairs (linked Wikipedia pages), where $n(i,j)$ is the test set size.
+We evaluate the representation by measuring mate retrieval quality on the test sets: for each $\ell$, we rank the projected documents $P_j(y_1),\ldots, P_j(y_{n(i,j)})$ according to their similarity with $P_i(x_\ell)$ and compute the rank of the mate document $r(\ell) = rank(P_j(y_\ell))$.
+The final retrieval score (between -100 and 100) is computed as: $\frac{100}{n(i,j)} \cdot \sum_{\ell = 1}^{n(i,j)} \left( \frac{n(i,j) - r(\ell)}{n(i,j) -1} -0.5\right)$.
+A score that is less than 0 means that the method performs worse than random retrieval and a score of 100 indicates perfect mate retrieval.
+The mate retrieval results are included in Table \ref{table:retrieval}.
+
+We observe that the method performs well on all pairs between languages: \emph{en}, \emph{es}, \emph{ru}, \emph{sl}, where at least 50,000 training documents are available.
+We notice that taking $k = 500$ or $k = 1000$ multilingual topics usually results in similar performance, with some notable exceptions: in the case of (\emph{ht}, \emph{war}) the additional topics result in an increase in performance, as opposed to (\emph{ht},\emph{pms}) where performance drops, which suggests overfitting.
+The languages where the method performs poorly are \emph{ht} and \emph{war}, which can be explained by the quality of data (see Table \ref{table:rank} and explanation that follows).
+In case of \emph{pms}, we demonstrate that solid performance can be achieved for language pairs (\emph{pms}, \emph{sl}) and (\emph{pms}, \emph{hi}), where only 2000 training documents are shared between \emph{pms} and \emph{sl} and no training documents are available between \emph{pms} and \emph{hi}.
+Also observe that in the case of (\emph{pms}, \emph{ht}) the method still obtains a score of 62, even though training set intersection is zero and \emph{ht} data is corrupted, which we will show in the next paragraph.
+{
+\renewcommand\tabcolsep{3pt}
+\begin{table}[h!]
+\caption{Pairwise retrieval, 500 topics;1000 topics}\label{table:retrieval}
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
+\cline{1-9}
+&	en&	es&	ru&	sl&	hi&	war&	ht&	pms\\\cline{1-9}
+en&	    &	98~-~98&	95~-~97&	97~-~98&	82~-~84&	76~-~74&	53~-~55&	 96~-~97\\
+\cline{1-9}
+es&	97~-~98&	&	94~-~96&	97~-~98&	85~-~84&	76~-~77&	56~-~57&	96~-~96\\
+\cline{1-9}
+ru&	96~-~97&	94~-~95&	&	97~-~97&	81~-~82&	73~-~74&	55~-~56&	96~-~96\\
+\cline{1-9}
+sl&	96~-~97&	95~-~95&	95~-~95&	&	91~-~91&	68~-~68&	59~-~69&	93~-~93\\
+\cline{1-9}
+hi&	81~-~82&	82~-~81&	80~-~80&	91~-~91&	&	68~-~67&	50~-~55&	87~-~86\\
+\cline{1-9}
+war&	68~-~63&	71~-~68&	72~-~71&	68~-~68&	66~-~62&	&	28~-~48&	 24~-~21\\
+\cline{1-9}
+ht&	52~-~58&	63~-~66&	66~-~62&	61~-~71&	44~-~55&	16~-~50&	&	62~-~49\\
+\cline{1-9}
+pms&	95~-~96&	96~-~96&	94~-~94&	93~-~93&	85~-~85&	23~-~26&	66~-~54&	 \\
+\cline{1-9}
+\end{tabular}
+\end{center}
+\end{table}
+}
+
+We now describe the selection of train and test sets.
+We select the test set documents as all multi-lingual documents with at least one nonempty alignment from the list: (\emph{hi}, \emph{ht}), (\emph{hi}, \emph{pms}), (\emph{war}, \emph{ht}), (\emph{war}, \emph{pms}).
+This guarantees that we cover all the languages.
+Moreover this test set is suitable for testing the retrieval thorough the hub as the chosen pairs have empty alignments.
+The remaining documents are used for training.
+In Table \ref{table:train_test}, we display the corresponding sizes of training and test documents for each language pair.
+The first row represents the size of the training sets used to construct the mappings in low dimensional language independent space using the English--\emph{en} as a hub.
+The diagonal elements represent number of the unique training documents and test documents in each language.
+
+{
+\renewcommand\tabcolsep{3pt}
+\begin{table}[h!]
+\caption{Pairwise training:test sizes (in thousands)}
+\label{table:train_test}
+{
+\small
+\begin{tabular}{c|c|c|c|c|c|c|c|c|}
+&	en&	es&	ru&	sl&	hi&	war&	ht&	pms\\\cline{1-9}
+en&	671~-~4.64&	463~-~4.29&	369~-~3.19&	50.3~-~2&	14.4~-~2.76&	8.58~-~2.41&	 17~-~2.32&	16.6~-~2.67\\
+\cline{2-9}
+es&	\multicolumn{1}{c|}{}	&	463~-~4.29&	187~-~2.94&	28.2~-~1.96&	8.72~-~2.48&	 6.88~-~2.4&	13.2~-~2&	 13.8~-~2.58\\
+\cline{3-9}
+ru&	\multicolumn{2}{c|}{}	&	369~-~3.19&	29.6~-~1.92&	9.16~-~2.68&	2.92~-~1.1&	 3.23~-~2.2&	10.2~-~1.29\\
+\cline{4-9}
+sl&	\multicolumn{3}{c|}{}	&	50.3~-~2&	3.83~-~1.65&	1.23~-~0.986&	0.949~-~1.23&	 1.85~-~0.988\\
+\cline{5-9}
+hi&	\multicolumn{4}{c|}{}	&	14.4~-~2.76&	0.579~-~0.76&	0.0~-~2.08&	0.0~-~0.796\\
+\cline{6-9}
+war&	\multicolumn{5}{c|}{}	&	8.58~-~2.41&	0.043~-~0.534&	0.0~-~1.97\\
+\cline{7-9}
+ht&	\multicolumn{6}{c|}{}	&	17~-~2.32&	0.0~-~0.355\\
+\cline{8-9}
+pms&	\multicolumn{7}{c|}{}	&	16.6~-~2.67\\
+\cline{9-9}
+\end{tabular}
+}
+\end{table}
+}
+
+We further inspect the properties of the training sets by roughly estimating the fraction \texttt{rank(A)/min(size(A))} for each training English matrix and its corresponding mate matrix.
+Ideally, these two fractions are approximately the same so  both aligned spaces should have reasonably similar dimensionality.
+We display these numbers as pairs in Table \ref{table:rank}.
+
+\begin{table}[h]
+\caption{Dimensionality drift}
+\label{table:rank}
+\begin{tabular}{|c|c|c|c|c|c|c|}
+\cline{1-7}
+(en, de)     &   (en, ru)     &   (en, sl)       &     (en, hi)&   (en, war)      &      (en, ht) &   (en, pms)\\
+\cline{1-7}
+(0.81, 0.89)   &  (0.8, 0.89)  &   (0.98, 0.96)    &    (1, 1)  &   (0.74, 0.56)  &      (1, 0.22)  &   (0.89, 0.38)\\
+\cline{1-7}
+\end{tabular}
+\end{table}
+
+It is clear that in the case of Creole language only at most $22\%$ documents are unique and suitable for the training.
+Though we removed the stub documents, many of remaining documents are almost the same, as the quality of some minor Wikipedias is low.
+This was confirmed for Creole, Waray-Waray, and Piedmontese language by manual inspection.
+The low quality documents correspond to templates about the year, person, town, etc.
+and contain very few unique words.
+
+We also have a problem with the quality of the test data.
+For example, if we look at test pair (\emph{war}, \emph{ht}) only 386/534 Waray-Waray test documents are unique but on other side almost all Creole test documents (523/534) are unique.
+This indicates a poor alignment which leads to poor performance.
+}
+
+\subsection{Cross-Lingual Event Linking}
 
@@
 
-\section{Related work}
+\section{Discussion}
 
-\subsection{Cross-lingual Document Similarity}
+\subsection{Scaling}
 
-\subsection{Multilingual Latent Factor Models}
-
-Ask Wray
-
-\subsection{Cross-lingual Event Linking}
-
-\section{Discussion}
+100 languages!
+Describe the newsfeed volume, number of events and the number of similarities that need to be computed. Compare the memory requirements of our current implementation to the rest.
+Comment on the caching gains in our implementation.
 

### 2015-03-20 rupnik2015.tex
minor
small change
Adds some material. Rough working outline of assigned work to Andrej.

@@
 \usepackage{jair, theapa, rawfonts}
+\usepackage{graphicx}
 
@@
 \subsection{Datasets}
+\emph{Andrej started shuffling it, will do it till Monday :) }
 
@@
 \subsection{Experiments With Missing Alignment Data}
-
+\emph{Andrej started working on this, done by Monday}
 {\oldText{Taken from XLite paper}
@@
 {\oldText{TODO:} add some explanation why this might be so.}
+\emph{Andrej adds some tests, new experiment description graphs, and explanations for performance, done by Monday}
+%15000 test documents, around 300 000 sth training documents
+%Results for feature spaces of dimensions 500, 800, 1000  overfitting, 800 minor improvement
+%Timex complexity of similarity computation scales linearly with dimension space. We double numbers of articles and decrease memory requirements by 1/2.
+%CCA       CA          AUC         Prec        Recall
+%CCA =     [0.7882      0.8504      0.7692	   0.8209
+%           0.7957      0.8624      0.7788      0.8234
+%           0.8031      0.8654      0.8015      0.8035];
+%LSIS=     [0.7994      0.8627      0.7778      0.8358
+%           0.7956      0.8634      0.7828      0.8159
+%           0.7994      0.8642      0.8030      0.7910];
+%LSIW=     [0.7894      0.8606      0.7673      0.8284;
+%           0.7870      0.8632      0.7662      0.8234;
+%           0.8018      0.8663      0.7854      0.8284];
+
+%AMPR as function of the number of feature vectors
+%500 is good, 800 better, 1000 overfitting, noise!
+\includegraphics{slika.png}
+%Why the hub approach?
+%Optimize for language pairs in the system that have the largest number of articles in the system!
+%Is star CCA really the best???
+%Maybe omit this table?
 
@@
 \subsection{Scaling}
-
+\emph{Andrej asks who is the rest? I think that this section should be short.}
 100 languages!
+%Newsfeed volume
 Describe the newsfeed volume, number of events and the number of similarities that need to be computed. Compare the memory requirements of our current implementation to the rest.

### 2015-03-21 rupnik2015.tex
small

--- 2015-03-19 rupnik2015.tex
@@
 \end{abstract}
-\section{Overview}
-Structure
-\begin{itemize}
-\item Introduction
-\item Cross lingual document similarity (model)
-\item  Event Detection
-\item  Combing the approaches
-\item Experiments
-\item Conclusions
-\end{itemize}
-The story:
-\begin{itemize}
-\item Description of the problem - we would liek to understand events ans sequences of events, but our observables (e.g. news stories) occur in a variety of languages
-\item We present an approach to tackle this problem using a combination of techniques....
-\item Ultimately we present a tool showing the feasibility of our approach
-\end{itemize}
+%\section{Overview}
+%Structure
+%\begin{itemize}
+%\item Introduction
+%\item Cross lingual document similarity (model)
+%\item  Event Detection
+%\item  Combing the approaches
+%\item Experiments
+%\item Conclusions
+%\end{itemize}
+%The story:
+%\begin{itemize}
+%\item Description of the problem - we would liek to understand events ans sequences of events, but our observables (e.g. news stories) occur in a variety of languages
+%\item We present an approach to tackle this problem using a combination of techniques....
+%\item Ultimately we present a tool showing the feasibility of our approach
+%\end{itemize}
 
@@
 The term event is vague and ambiguous, but for the practical purposes we define it as ``any significant happening that is being reported about in the media''.
-Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015.
+Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015 (Figure \ref{}).
 Events such as these are covered by many articles and the question is how to find all the articles in different languages that are describing a single event.
@@
 Finally, we present and interpret the experimental results and finally give conclusions and describe promising future directions. 
-
-\section{Introduction}
-
-{\oldText{Taken from XLite paper}
-Document retrieval is a well-established problem in data mining.
+%
+%\section{Introduction}
+%
+%
+%{\oldText{Taken from XLite paper}
+%Document retrieval is a well-established problem in data mining.
 There have been a large number of different approaches put forward in the literature.
@@
 As the availability of multi-lingual documents has exploded in the last few years, there is a pressing need for automatic cross-lingual processing tools.
-
-The prime example is Wikipedia - in 2001 the majority of pages were written in English, while in 2012, the percentage of English articles has dropped to 14\%.
+%
+%The prime example is Wikipedia - in 2001 the majority of pages were written in English, while in 2012, the percentage of English articles has dropped to 14\%.
 In this context, we look at how to find similar documents across languages.
@@
 While the percentage of English articles has fallen, English is still the largest language and one of \emph{hub} languages which not only have an order of magnitude more articles than other languages, but also many comparable articles with most of the other languages.
-
-When doing document retrieval, we encounter two quite different situations.
+%
+%When doing document retrieval, we encounter two quite different situations.
 If for example, we are looking for a German article comparable to an English article, there is a large alignment between the document corpora (given by the intersection in articles in the two languages) making the problem well-posed.
@@
 In the following section, we introduce our representation followed by our experiments, which also shed light on the structural properties of the multilingual Wikipedia corpus.
-}
+%}
+
+\section{System Description}
 

### 2015-03-22 rupnik2015.tex
everything except evaluation

@@
 \usepackage{amssymb}
+\usepackage{comment}
 
@@
 
-\title{Event Tracking Across Languages - 
-Cross-Lngual Document Similarity and Event Tracking \todo{Better title...}}
+\title{News Across Languages -
+Cross-Lingual Document Similarity and Event Tracking}
 
@@
 %\item Ultimately we present a tool showing the feasibility of our approach
+%%\end{itemize}
+%
+%\section{Overview}
+%Structure
+%\begin{itemize}
+%\item Introduction
+%\item System overview
+%\item Cross lingual document similarity (algorithms)
+%\item Event linking
+%\item Experiments
+%\item Conclusions
 %\end{itemize}
+%The story:
+%\begin{itemize}
+%\item Description of the problem - we would like to understand global news as sequences of events, but our observables (e.g. news stories) occur in a variety of languages
+%\item We present an approach to tackle this problem using a combination of techniques....
+%\item Ultimately we present a tool showing the feasibility of our approach
+%\end{itemize}
+%
 
@@
 
-\section{System Description}
+\section{System Overview}
+
+Event registry $\Rightarrow$ Monolingual events $\Rightarrow$ features (semantic + CCA) $\Rightarrow$ linked events (crosslingual)
 
@@
 The common weighting scheme is called \emph{Term Frequency Inverse Document Frequency} ($TFIDF$) weighting.
-An \emph{Inverse Document Frequency} ($IDF$) weight for the dictionary term $k$ is defined as $\log( \frac{N}{DF_k} )$, where $DF_k$ is the number of documents in the corpus which contain term $k$.
+An \emph{Inverse Document Frequency} ($IDF$) weight for the dictionary term $k$ is defined as $\log\left( \frac{N}{DF_k} \right)$, where $DF_k$ is the number of documents in the corpus which contain term $k$.
 A document $TFIDF$ vector is its original vector multiplied element-wise by the weights.
@@
 The $TFIDF$ weighted vector space model document representation corresponds to a map $\phi : \text{text} \rightarrow \RR^n$ defined by:
-$$\phi(d)_k = {TF}_k(d) \log( \frac{N}{{DF}_k}).$$
+$$\phi(d)_k = {TF}_k(d) \log\left( \frac{N}{{DF}_k}\right).$$
 
@@
 
-\subsection{Datasets}
+\section{Multilingual Latent Factor Models}\label{sec:models}
+In this section, we describe several approaches to computing multilingual simliarities described in the previous section. We present three approaches
+\begin{enumerate}
+\item  a simple approach based on $k$-means clustering
+\item  a standard approach based on singular value decomposition 
+\item a new method that is applicable in more general settings.
+\end{enumerate}
+ We concentrate on approaches that are based on linear maps rather than alternatives, such as machine translation and probabilistic models. Applicability of machine translation relies
+on the availability of training resources (strongly aligned translation corpora), which are not available for many world languages, but there exist comparable corpora. Probabilistic
+methods \todo{provide reasons for not using them: e.g. difficulty with parameter estimation, maybe just say future work?}.
 
-{\oldText{Taken from ITI paper}
-To investigate the empirical performance of the low rank approximations we will test the algorithms on a large-scale, real-world multilingual dataset that we extracted from Wikipedia by using inter-language links as an alignment.
-This  results in a large number of weakly comparable documents in more than $200$ languages.
-Wikipedia is a large source of multilingual data that is especially important for the languages for which no translation tools, multilingual dictionaries as Eurovoc, or strongly aligned multilingual corpora as Europarl are available.
-Documents in different languages are related with so called 'inter-language' links that can be found on the left of the Wikipedia page.
-The Wikipedia is constantly growing.
-There are currently four Wikipedias with more than $10^6$ articles, $40$ with more than $10^5$ articles, $100$ with more than $10^4$ articles, and $216$ with more than $1000$ articles.
-Wikipedia uses special user-friendly markup language that is very easy to write but very hard to parse.
-Simplicity of language can cause ambiguities and moreover it is constantly changing.
-For example, separator | can be used in different contexts.
+For completeness we will also present the method of Canonical Correlation Analysis, since it is related to our proposed method.
 
-Wikipedia raw xml dumps of all currently $270$ active editions were downloaded from the Wikipedia dump page\href{http://dumps.wikimedia.org}.
-The xml files are too large to be parsed with DOM like parser that needs to store the whole xml tree in the memory, instead we implemented Sax like parser that tries to simulate behavior of Wikipedia official parser and is as simple, fast and error prone as possible.
-We parse all Wikipedia markup but do not extend the templates.
-Each Wikipedia page is embedded in the page tag.
-First we check if the title of the page consists of any special namespace and do not process such pages.
-Then we check if this is a redirection page and we store the redirect link as inter-language links can point to redirection link also.
-If nothing of the above applies we extract the text and parse the Wikipedia markup.
-Currently all the markup is removed.
-
-We get inter-language link matrix using previously stored redirection links and inter-language links.
-If inter-language link points to the redirection we replace it with the redirection target link.
-It turns out that we obtain the matrix $M$ that is not symmetric, consequently the underlying graph is not symmetric.
-That means that existence of the inter-language link in one way (i.e.
-English to German) does not guarantee that there is an inter-language link in the reverse direction (German to English).
-To correct this we transform this matrix to symmetric by computing $M+M^T$ and obtaining an undirected graph.
-In the rare case that we have multiple links pointing from the document, we pick the first one that we encountered.
-This matrix enables us to build an alignment across all Wikipedia languages.
-}
-
-\section{Multilingual Latent Factor Models}\label{sec:models}
-
-\subsection{Multilingual dataset}
+\subsection{Notation}
 
@@
 
-\subsection{Algorithms}
-\label{sec:algorithms}
+\subsection{$k$-means}
 
-We will present three approaches: a simple approach based on $k$-means clustering, a standard approach based on singular value decomposition and a new method that is applicable in more general
-settings. All approaches are based on linear maps and we do not focus on alternatives, such as machine translation and probabilistic models. Applicability of machine translation relies
-on the availability of training resources (strongly aligned translation corpora), which are not available for many world languages, but there exist comparable corpora. Probabilistic
-methods \todo{provide reasons for not using them: e.g. difficulty with parameter estimation}.
-
-\subsubsection{$k$-means}
-
-The $k$-means algorithm is perhaps the most well-known and used clustering algorithm.
-In order to apply the algorithm we first merge all the term-document matrices into a single matrix $X$ by stacking the individual term-document matrices:
+The $k$-means algorithm is perhaps the most well-known and widely-used clustering algorithm.
+In order to apply the algorithm, we first merge all the term-document matrices into a single matrix $X$ by stacking the individual term-document matrices:
 $$X := \begin{bmatrix}X_1^T ,X_2^T, \cdots, X_m^T \end{bmatrix}^T,$$
@@
 \subsection{Canonical Correlation Analysis}
- For completeness we now introduce a statistical technique to analyze data from two sources - after which we will describe our main contribution to computing cross-lingual similarities.
+ The final approach is a statistical technique to analyze data from two sources - after which we will describe our main contribution to computing cross-lingual similarities.
 
@@
 
-In this section we describe a method that is based on linear embeddings as well, but is more applicable to a large number of languages. The main difficulty with applying the
+In this section, we describe an extension to CCA, but is more applicable to a large number of languages. The main difficulty with applying the
 LSI and $k$-means approaches lies in the fact, that when one considers a large number of languages, for example, the top 100 Wikipedia languages (ranked by number of articles), the set of completely
-aligned documents can be small or empty. Even if only two languages are considered, the set of aligned documents can be small (for example, Piedmontese and Hindi
-Wikipedias had no interlanguage links). In Wikipedia we observed that even though the training resources are scarce between certain language pairs, often there exists indirect training data
-through a hub language.
+aligned documents is often small or empty. Even if only two languages are considered, the set of aligned documents can be small (for example, Piedmontese and Hindi
+Wikipedias had no interlanguage links). In Wikipedia, we observed that even though the training resources are scarce between certain language pairs, there often exists indirect training data
+through what we refer to as a hub language.
 
-A \emph{hub language} is a language with a high proportion of non-empty documents in $D = \left\{d_1,..., d_{\ell}\right\}$, which in case of Wikipedia is English.
+A \emph{hub language} is a language with a high proportion of non-empty documents in $D = \left\{d_1,..., d_{\ell}\right\}$.
+The prototypical  exampling  in the case of Wikipedia is English.
 We use the following notation to define subsets of the multilingual comparable corpus: let $a(i,j)$ denote the index set of all multilingual documents with non-missing data for the $i$-th and $j$-th language:  $a(i,j) = \left\{k | d_k = (u_1,...,u_m), u_i \neq \emptyset, u_j \neq \emptyset \right\},$ and let $a(i)$ denote the index set of all multilingual documents with non missing data for the $i$-th language.
@@
 
-\subsection{Scaling}
+(figure?)
 
-100 languages!
-Describe the newsfeed volume, number of events and the number of similarities that need to be computed. Compare the memory requirements of our current implementation to the rest.
-Comment on the caching gains in our implementation.
+\section{Cross-lingual Event Linking}
+The main application we test the above simliarity on  is cross-lingual event linking.
+In online media streams - particularly new articles - there is often duplication of reporting, different viewpoints or opinions, all centering around a single event.
+Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015.
+(Sports example)
 
-\subsection{Evaluation}
+Informally, events are covered by many articles and the question we address is how to find all the articles in different languages that are describing a single event.
+We base our evaluation on an online system for detection of world events, called Event Registry\cite{}.
+We do not address the problem of detection of events in this paper but rather consider the problem that given a series of events, how may we best ``tag" or match articles in different languages to these events.
+The events are represented by clusters of articles and so ultimately our problem reduces to find suitable matchings between clusters with articles in different languages.
 
-\subsubsection{Link prediction}
+\begin{comment}
+One of the use cases where article linking information can be applied is also in an online system for detection of world events, called Event Registry\cite{}.
+Event Registry is a repository of events, where events are automatically identified by analyzing news articles.
+The term event is vague and ambiguous, but for the practical purposes we define it as ``any significant happening that is being reported about in the media''.
 
-\subsubsection{Hub language}
+Event Registry works by collecting and analyzing news articles that are published by news outlets in different languages all over the world.
+Collected articles are first semantically annotated by identifying in them mentions of relevant concepts -- either entities or important keywords.
+The disambiguated and entity linking of the concepts is done using Wikipedia as the main knowledge base.
+The key step in Event Registry is then to apply an online clustering algorithm on the articles in order to identify groups of articles that are discussing the same event.
+For each new article, the clustering algorithm determines if the article should be assigned to some existing cluster or into a new cluster.
+The underlying assumption is that articles that are describing the same event are similar enough and will therefore be put into the same cluster.
 
+In short, the article clustering works as follows (details are available in \cite{}).
+First we create a new document by joining the title and the body of the article.
+We then tokenize the document, remove the stop words and stem the words.
+The remaining words are represented as a vector in the vector-space model.
+Each value in the vector is additionally normalized using TF-IDF.
+After normalization, we use the vector to find the most similar existing cluster.
+To be able to compare the vector with a cluster we first need a vector representation for each of the clusters.
+This is done by computing a centroid vector of the cluster by simply averaging the vectors of the articles in the cluster.
+By computing the cosine similarity of the article vector with each of the centroid vectors we then find the cluster with the highest similarity.
+If the highest similarity is above the selected threshold (we used 0.4 in our experiments), the article is assigned to the corresponding cluster, otherwise a new cluster is created, initially containing only the single article.
+
+Since articles about an event are commonly written only for a short period of time, we remove clusters once the oldest article in the cluster becomes more than 4 days old.
+This housekeeping mechanism prevents the clustering from becoming slow and also ensures that articles are not assigned to obsolete clusters.
+
+In order to treat the cluster as describing an event in Event Registry, the cluster has to grow above a certain number of articles (the number depends on the language).
+This condition ensures that only content that receives a sufficient coverage in the media is treated as an event.
+Once a new event is identified, a new unique ID is assigned to it and the main information about the event is then automatically extracted by analyzing the articles assigned to it.
+The extracted information includes properties such as the date of the event, the location, who is involved in it, what the event is about, etc.
+
+The clustering of articles is done for each language separately -- English articles are, for example, clustered separately from Spanish articles.
+When events are identified, they will therefore be identified from clusters in each language separately.
+As a result, the same event will be represented in Event Registry with two or more event IDs, and each event will only contain coverage in a single language.
+To avoid this problem we want to exploit the available information about the similarity of the articles in a cluster with articles in other languages.
+Using this information we hope to be able to connect clusters in different languages and represent them using a single event ID.
+\end{comment}
+
+\subsection{Problem definition}
+
+The problem in the second phase of the pipeline is to identify the clusters (given?) describing the same events.
+Each article $a_i$ is written in a language $l$, where $l \in L = \{l_1,l_2,...,l_k\}$.
+For each language $l$, we generate and maintain a set of language-specific clusters $C_l$.
+That is, all articles are in language $l$.
+For any cluster $c_i \in C_{l_a}$, we would like to identify clusters $c_j \in C_{l_b}$ (where $l_a \neq l_b$) that describe the same event as $c_i$.
+We define \emph{equivalent clusters} as  pairs $(c_i,c_j)$ such that both clusters in the pair describe the same event in languages $C_{l_a}$ and $C_{l_b}$ respectively.
+
+We impose an additional requirement.
+Identification of the equivalent clusters cannot be done exhaustively -- for a cluster $c_i$ we wish to avoid testing all other clusters for all the other languages.
+Performing exhaustive comparison for each cluster $c_i$ would result in $O(N^2)$ tests, where $N$ is the number of all clusters (over all languages), which is not feasible for when the number of clusters is on the order of tens of thousands.
+
+\subsection{Related work}
+
+TODO
+
+\subsection{Algorithm}
+
+In order to identify clusters $c_j$ that are equivalent to cluster $c_i$, we have developed a two-stage algoirhtm.
+First, for a cluster $c_i$, we efficiently identify a small set of candidate clusters and then  find the clusters from among the candidates which are equivalent to $c_i$.
+
+The details of the first step are described in Algorithm~\ref{cluster_merge_algo1}.
+The algorithm begins by individually inspecting each article $a_i$ in the cluster $c_i$.
+Using CCA  (see Section~\ref{}),  it then identifies the 10 most similar articles to $a_i$ in each language $l \in L$.
+For each  similar article $a_j$, we identify its corresponding  cluster $c_j$ and add it to the set of candidates.
+\todo{we should say a few words why this is more efficient - since now we are comparing each article to all other articles - while we said it is not feadible for clusters}.
+While the CCA simliarity is approximate, our  assumption  is that articles in different languages describing the same event, will generally have a higher similarity than articles about different events.
+While this assumption does not always hold, redundancy in the data mitigates these false positives.
+Since we compute the 10 most similar articles for each article in $c_i$, we are likely to identify all the relevant candidates for cluster $c_i$.
+
+\begin{algorithm}[t!]
+\SetKwInput{KwInput}{input}\SetKwInput{KwOutput}{output}
+\KwInput{test cluster $c_i$,\\a set of clusters $C_l$ for each language $l \in L$}
+\KwOutput{a set of clusters $C$ that are potentially equivalent to $c_i$}
+$C = \{\}$\;
+\For{article $a_i \in c_i$} {
+    \For{language $l \in L$} {
+        \tcc{use CCA to find 10 most similar articles to article $a_i$ in language $l$}
+        $SimArt = getCCASimArts(a_i, l)$\;
+        \For{article $a_j \in SimArt$} {
+            \tcc{find cluster $c_j$ to which article $a_j$ is assigned to}
+            $c_j := c ; c \in C_l, a_j \in c$\;
+            \tcc{add cluster $c_j$ to the set of candidates $C$}
+            $C \leftarrow c_j$\;
+        }
+    }
+}
+\caption{Algorithm for identifying candidate clusters $C$ that are potentially equivalent to $c_i$}
+\label{cluster_merge_algo1}
+\end{algorithm}
+
+The second stage of the algorithm (described in the Algorithm~\ref{cluster_merge_algo2}.) determines which (if any) of the candidate clusters are equivalent to $c_i$.
+We approach this as a supervised learning problem.
+For each candidate cluster $c_j \in C'$, we compute a vector of learning features that should be indicative of whether the $c_i$ and $c_j$ are equivalent or not.
+The details of our chosen features will be described in the following Section.
+Once these features are computed, we apply a classification model which was trained using a  dataset  of hand-labeled examples.
+If the model predicts that tested clusters are equivalent, the $c_j$ is added to the set $C'$.
+
+\begin{algorithm}[tb!]
+\SetKwInput{KwInput}{input}\SetKwInput{KwOutput}{output}
+\KwInput{test cluster $c_i$,\\
+a set of candidate clusters $C$ that are potentially equivalent to $c_i$\\
+classification model $M$ trained on the hand-labeled examples}
+\KwOutput{a set of clusters $C'$ that are equivalent to $c_i$}
+
+$C' = \{\}$\;
+\For{cluster $c_j \in C$} {
+    \tcc{extract a vector of features for cluster pair $c_i$ and $c_j$}
+    $\vec{v} = f(c_i, c_j)$\;
+    \tcc{classify the feature vector $\vec{v}$ using the model $M$}
+    $pred = M(\vec{v})$\;
+    \If{$pred$ = true} {
+        $C' \leftarrow c_j$
+    }
+}
+\caption{Algorithm for identifying clusters $C'$ that are equivalent to cluster $c_i$}
+\label{cluster_merge_algo2}
+\end{algorithm}
+
+\subsection{Training a classification model}
+
+In order to train a classification model for predicting whether two clusters are equivalent, we must first have labeled data.
+To obtain the data, we used two human annotators.
+The annotators were provided with an interface displaying information about the articles in two clusters and their task was to determine if the clusters were equivalent or not.
+To obtain a pair of clusters $(c_i, c_j)$ to annotate, we first randomly chose a cluster $c_i$, used the Algorithm~\ref{cluster_merge_algo1} to compute a set of potentially equivalent clusters $C$ and randomly chose a cluster $c_j \in C$.
+
+The dataset provided by the annotators contains 808 examples, of which 402 are equivalent clusters pairs and 406 are not.
+Clusters in each learning example are either in English, Spanish or German.
+Although Event Registry imports articles in other languages as well, we restricted our experiments to these three languages, since they have the highest volume of articles which makes the training problem similarly hard \todo{doesnt this make it easier?}.
+
+To build a classification model we first extract  informative features from the labeled examples. Based on the availability of the data, we extracted the following groups of features:
+
+\begin{itemize}
+\item {\bf Cross-lingual article linking features}.
+Cross-lingual article linking information was already used to obtain candidates for the equivalent clusters.
+Additionally, it can also be used to provide valuable  features for learning.
+There are two main features that we extract from article links between clusters -- \texttt{linkCount} and \texttt{avgSimScore}.
+\texttt{linkCount} is the number of times an article $a_i \in c_i$ has as one of the 10 most similar articles an article $a_j \in c_j$.
+In other words, it is the number of times an article from $c_i$ has a very similar article, which is in $c_j$.
+Beside the number of links between the two clusters we can also take into account the similarity scores of the links.
+The \texttt{avgSimScore} is the feature that represents the average similarity score of the links between the two clusters.
+
+\item {\bf Concept related features}.
+As mentioned before, articles that are imported into Event Registry are first semantically annotated by linking mentioned entities and keywords to the corresponding Wikipedia pages.
+Whenever Barack Obama is, for example, mentioned in the article, the article is annotated with a link to his Wikipedia page (\emph{http://en.wikipedia.org/wiki/Barack\_Obama}).
+In the same way all mentions of people, locations, organizations and even ordinary keywords (e.g.
+bank, tax, ebola, plane, company) are annotated.
+Although the Spanish article about Obama will be annotated with his Spanish version of the Wikipedia page, in many cases we can link the Wikipedia pages to their English versions.
+This can be done since Wikipedia itself provides information regarding which pages in different languages represent the same concept/entity.
+Using this approach, the word ``avi\'on'' in a Spanish article will be annotated with the same concept as the word ``plane'' in an English article.
+Although the articles are in different languages, the annotations can therefore provide a shared vocabulary that can be used to compare articles/clusters.
+By analyzing all the articles in clusters $c_i$ and $c_j$, we can identify the most relevant entities and keywords for each cluster.
+Additionally, we can also assign weights to the concepts based on how frequently they occur in the articles in the cluster.
+From the list of relevant concepts and corresponding weights, we consider the following features: \texttt{entityCosSim} which represents the cosine similarity on the vector of entities mentioned in both clusters; \texttt{keywordCosSim} which represents the cosine similarity on the keywords only; for cases where the weights are not relevant, we have also added two additional features - (\texttt{entityJaccardSim} and \texttt{keywordJaccardSim}), where Jaccard similarity is used instead of the cosine similarity.
+
+\item {\bf Miscellaneous features}.
+This group contains three miscellaneous features that seem discriminative but are unrelated to the previous two groups.
+The first feature is related to the event location.
+By considering the locations mentioned in the articles in a cluster, we are able to estimate where an event occurred.
+The \texttt{hasSameLocation} feature is a boolean variable that is true when the location of the event in both clusters is the same.
+Another important feature is related to the time when the articles in the clusters were published.
+For each cluster, we first compute the average publication time and date of the articles.
+The \texttt{timeDiff} is then defined as the absolute difference in hours between the two computed dates.
+The final feature we consider is computed by analyzing the mentions of dates in the articles.
+Using an extensive set of regular expressions, we are able to detect mentions of dates in different forms.
+To compute the feature \texttt{sharedDates} we start by identifying the list of dates mentioned in articles in each cluster separately.
+The value of \texttt{sharedDates} is then determined as the ratio of all dates that are detected in both clusters.\todo{this should be made more explicit/}
+
+\end{itemize}
+
+Using the  features described above and the user-provided labels, we train a classification model $M$ which is used in Algorithm~\ref{cluster_merge_algo2} to classify a new cluster pair as equivalent or not.
+The classification algorithm that we used to train a model was a linear SVM method.
+
+\section{Evaluation}
+
+We describe the datasets involved in the experiments and two evaluations: comparing the crosslingual approaches on an information retrieval task, and evaluation of cross-lingual event linking.
+
+\subsection{Datasets}
+\emph{Andrej started shuffling it, will do it till Monday :) }
+
+{\oldText{Taken from ITI paper}
+To investigate the empirical performance of the low rank approximations we will test the algorithms on a large-scale, real-world multilingual dataset that we extracted from Wikipedia by using inter-language links as an alignment.
+This  results in a large number of weakly comparable documents in more than $200$ languages.
+Wikipedia is a large source of multilingual data that is especially important for the languages for which no translation tools, multilingual dictionaries as Eurovoc, or strongly aligned multilingual corpora as Europarl are available.
+Documents in different languages are related with so called 'inter-language' links that can be found on the left of the Wikipedia page.
+The Wikipedia is constantly growing.
+There are currently four Wikipedias with more than $10^6$ articles, $40$ with more than $10^5$ articles, $100$ with more than $10^4$ articles, and $216$ with more than $1000$ articles.
+Wikipedia uses special user-friendly markup language that is very easy to write but very hard to parse.
+Simplicity of language can cause ambiguities and moreover it is constantly changing.
+For example, separator | can be used in different contexts.
+
+Wikipedia raw xml dumps of all currently $270$ active editions were downloaded from the Wikipedia dump page\href{http://dumps.wikimedia.org}.
+The xml files are too large to be parsed with DOM like parser that needs to store the whole xml tree in the memory, instead we implemented Sax like parser that tries to simulate behavior of Wikipedia official parser and is as simple, fast and error prone as possible.
+We parse all Wikipedia markup but do not extend the templates.
+Each Wikipedia page is embedded in the page tag.
+First we check if the title of the page consists of any special namespace and do not process such pages.
+Then we check if this is a redirection page and we store the redirect link as inter-language links can point to redirection link also.
+If nothing of the above applies we extract the text and parse the Wikipedia markup.
+Currently all the markup is removed.
+
+We get inter-language link matrix using previously stored redirection links and inter-language links.
+If inter-language link points to the redirection we replace it with the redirection target link.
+It turns out that we obtain the matrix $M$ that is not symmetric, consequently the underlying graph is not symmetric.
+That means that existence of the inter-language link in one way (i.e.
+English to German) does not guarantee that there is an inter-language link in the reverse direction (German to English).
+To correct this we transform this matrix to symmetric by computing $M+M^T$ and obtaining an undirected graph.
+In the rare case that we have multiple links pointing from the document, we pick the first one that we encountered.
+This matrix enables us to build an alignment across all Wikipedia languages.
+}
+
+\subsection{Wikipedia Information Retrieval}
+
+Mate retrieval results
+
+\subsection{Experiments With Missing Alignment Data}
+\emph{Andrej started working on this, done by Monday}
 {\oldText{Taken from XLite paper}
@@
 
-\section{Cross-lingual Event Linking}
-
-One of the use cases where article linking information can be applied is also in an online system for detection of world events, called Event Registry\cite{}.
-Event Registry is a repository of events, where events are automatically identified by analyzing news articles.
-The term event is vague and ambiguous, but for the practical purposes we define it as ``any significant happening that is being reported about in the media''.
-Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015.
-Events such as these are covered by many articles and the question is how to find all the articles in different languages that are describing a single event.
-
-Event Registry works by collecting and analyzing news articles that are published by news outlets in different languages all over the world.
-Collected articles are first semantically annotated by identifying in them mentions of relevant concepts -- either entities or important keywords.
-The disambiguated and entity linking of the concepts is done using Wikipedia as the main knowledge base.
-The key step in Event Registry is then to apply an online clustering algorithm on the articles in order to identify groups of articles that are discussing the same event.
-For each new article, the clustering algorithm determines if the article should be assigned to some existing cluster or into a new cluster.
-The underlying assumption is that articles that are describing the same event are similar enough and will therefore be put into the same cluster.
-
-In short, the article clustering works as follows (details are available in \cite{}).
-First we create a new document by joining the title and the body of the article.
-We then tokenize the document, remove the stop words and stem the words.
-The remaining words are represented as a vector in the vector-space model.
-Each value in the vector is additionally normalized using TF-IDF.
-After normalization, we use the vector to find the most similar existing cluster.
-To be able to compare the vector with a cluster we first need a vector representation for each of the clusters.
-This is done by computing a centroid vector of the cluster by simply averaging the vectors of the articles in the cluster.
-By computing the cosine similarity of the article vector with each of the centroid vectors we then find the cluster with the highest similarity.
-If the highest similarity is above the selected threshold (we used 0.4 in our experiments), the article is assigned to the corresponding cluster, otherwise a new cluster is created, initially containing only the single article.
-
-Since articles about an event are commonly written only for a short period of time, we remove clusters once the oldest article in the cluster becomes more than 4 days old.
-This housekeeping mechanism prevents the clustering from becoming slow and also ensures that articles are not assigned to obsolete clusters.
-
-In order to treat the cluster as describing an event in Event Registry, the cluster has to grow above a certain number of articles (the number depends on the language).
-This condition ensures that only content that receives a sufficient coverage in the media is treated as an event.
-Once a new event is identified, a new unique ID is assigned to it and the main information about the event is then automatically extracted by analyzing the articles assigned to it.
-The extracted information includes properties such as the date of the event, the location, who is involved in it, what the event is about, etc.
-
-The clustering of articles is done for each language separately -- English articles are, for example, clustered separately from Spanish articles.
-When events are identified, they will therefore be identified from clusters in each language separately.
-As a result, the same event will be represented in Event Registry with two or more event IDs, and each event will only contain coverage in a single language.
-To avoid this problem we want to exploit the available information about the similarity of the articles in a cluster with articles in other languages.
-Using this information we hope to be able to connect clusters in different languages and represent them using a single event ID.
-
-\subsection{Problem definition}
-
-Formally, the problem of identifying the clusters describing the same event can be formalized as follows.
-Each article $a_i$ is written in a language $l$, where $l \in L = \{l_1,l_2,...,l_k\}$.
-For each language $l$, a set of clusters $C_l$ is generated and maintained that only contains articles in language $l$.
-For any cluster $c_i \in C_{l_a}$, the task is then to identify other clusters $c_j \in C_{l_b}$ (where $l_a \neq l_b$) that are describing the same event as $c_i$.
-We will refer to cluster pairs $c_i$ and $c_j$ that are describing the same event as equivalent clusters.
-An additional requirement of the task is that identification of the equivalent clusters should not be done exhaustively -- for a cluster $c_i$ we wish to avoid testing all other clusters $c_j$.
-Performing exhaustive comparison for each cluster $c_i$ would result in $O(N^2)$ tests, where $N$ is the number of all clusters, which is not feasible when the number of clusters is in tens of thousands.
-
-\subsection{Algorithm}
-
-In order to identify clusters $c_j$ that are equivalent to cluster $c_i$ we have developed a procedure that consists of two main tasks.
-The first task is to efficiently identify a limited set of clusters that are potential candidates for being equivalent to $c_i$.
-The second task is then about performing a detailed analysis of the candidates and finding the ones that are equivalent to $c_i$.
-
-The details of the first task are described in Algorithm~\ref{cluster_merge_algo1}.
-The algorithms starts by individually inspecting each article $a_i$ in the cluster $c_i$.
-Using the CCA it then identifies 10 most similar articles to $a_i$ in each language $l \in L$.
-For each identified similar article $a_j$ we then identify cluster $c_j$ to which the article belongs to and add the cluster to the set of candidates that should be more closely inspected.
-The assumption made by this algorithm is that if two articles in different languages are describing the same event, the CCA similarity between the documents will be higher than when comparing two articles about different events.
-Since the CCA similarity is only an estimate, the assumption does not always hold.
-Due to data redundancy, however, we do not consider this to be an issue.
-Since we are testing each article in the cluster and compute 10 most similar articles for each article, we are very likely to identify among the candidates also the clusters that are equivalent to cluster $c_i$.
-
-\begin{algorithm}[t!]
-\SetKwInput{KwInput}{input}\SetKwInput{KwOutput}{output}
-\KwInput{test cluster $c_i$,\\a set of clusters $C_l$ for each language $l \in L$}
-\KwOutput{a set of clusters $C$ that are potentially equivalent to $c_i$}
-$C = \{\}$\;
-\For{article $a_i \in c_i$} {
-    \For{language $l \in L$} {
-        \tcc{use CCA to find 10 most similar articles to article $a_i$ in language $l$}
-        $SimArt = getCCASimArts(a_i, l)$\;
-        \For{article $a_j \in SimArt$} {
-            \tcc{find cluster $c_j$ to which article $a_j$ is assigned to}
-            $c_j := c ; c \in C_l, a_j \in c$\;
-            \tcc{add cluster $c_j$ to the set of candidates $C$}
-            $C \leftarrow c_j$\;
-        }
-    }
-}
-\caption{Algorithm for identifying candidate clusters $C$ that are potentially equivalent to $c_i$}
-\label{cluster_merge_algo1}
-\end{algorithm}
-
-The second part of the procedure is described in the Algorithm~\ref{cluster_merge_algo2}.
-The goal of this task is to determine which (if any) of the candidate clusters are equivalent to $c_i$.
-To solve the task we represent it as a supervised learning problem.
-For each candidate cluster $c_j \in C'$ we compute a vector of learning features that should be indicative of whether the $c_i$ and $c_j$ are equivalent or not.
-The details about these features will be described in the next section.
-Once the features are computed, we classify the example with a classification model that was trained using a training dataset of hand-labeled examples.
-If the model predicts that tested clusters are equivalent, the $c_j$ is added to the set $C'$.
-
-\begin{algorithm}[tb!]
-\SetKwInput{KwInput}{input}\SetKwInput{KwOutput}{output}
-\KwInput{test cluster $c_i$,\\
-a set of candidate clusters $C$ that are potentially equivalent to $c_i$\\
-classification model $M$ trained on the hand-labeled examples}
-\KwOutput{a set of clusters $C'$ that are equivalent to $c_i$}
-
-$C' = \{\}$\;
-\For{cluster $c_j \in C$} {
-    \tcc{extract a vector of features for cluster pair $c_i$ and $c_j$}
-    $\vec{v} = f(c_i, c_j)$\;
-    \tcc{classify the feature vector $\vec{v}$ using the model $M$}
-    $pred = M(\vec{v})$\;
-    \If{$pred$ = true} {
-        $C' \leftarrow c_j$
-    }
-}
-\caption{Algorithm for identifying clusters $C'$ that are equivalent to cluster $c_i$}
-\label{cluster_merge_algo2}
-\end{algorithm}
-
-\subsection{Training a classification model}
-
-In order to train a classification model for predicting whether two clusters are equivalent we first need labeled training data.
-To obtain the data we used two human annotators.
-They were provided with a user interface that was displaying information about the articles in two clusters and their task was to determine if the clusters are equivalent or not.
-To obtain a pair of clusters $(c_i, c_j)$ to annotate, we first randomly chose a cluster $c_i$, used the Algorithm~\ref{cluster_merge_algo1} to compute a set of potentially equivalent clusters $C$ and randomly chose a cluster $c_j \in C$.
-
-The dataset provided by the annotators contains 808 examples, of which 402 are equivalent clusters pairs and 406 are not.
-Clusters in each learning example are either in English, Spanish or German language.
-Although Event Registry imports articles also in other languages, we restricted our experiments to these three languages since they have the highest volume of news articles which makes the training problem similarly hard.
-
-To build a classification model we first needed to extract from the labeled examples informative features. Based on the availability of the data, we have extracted the following groups of features:
-
-\begin{itemize}
-\item {\bf Cross-lingual article linking features}.
-Cross-lingual article linking information was already used to obtain candidates for the equivalent clusters.
-Additionally, it can also be used to provide valuable learning features.
-There are two main features that we extract from article links between clusters -- \texttt{linkCount} and \texttt{avgSimScore}.
-\texttt{linkCount} is the number of times an article $a_i \in c_i$ has as one of the 10 most similar articles an article $a_j \in c_j$.
-In other words, it is the number of times an article from $c_i$ has a very similar article, which is in $c_j$.
-Beside the number of links between the two clusters we can also take into account the similarity scores of the links.
-The \texttt{avgSimScore} is the feature that represents the average similarity score of the links between the two clusters.
-
-\item {\bf Concept related features}.
-As mentioned before, articles that are imported into Event Registry are first semantically annotated by linking mentioned entities and keywords to the corresponding Wikipedia pages.
-Whenever Barack Obama is, for example, mentioned in the article, the article is annotated with a link to his Wikipedia page (\emph{http://en.wikipedia.org/wiki/Barack\_Obama}).
-In the same way all mentions of people, locations, organizations and even ordinary keywords (e.g.
-bank, tax, ebola, plane, company) are annotated.
-Although the Spanish article about Obama will be annotated with his Spanish version of the Wikipedia page, we can almost in all cases normalize the Wikipedia links to their English versions.
-This can be done since Wikipedia itself provides information regarding which pages in different languages represent the same concept/entity.
-Using this approach, the word ``avi\'on'' in a Spanish article will be annotated with the same concept as the word ``plane'' in an English article.
-Although articles are in different languages, the annotations can therefore provide a shared vocabulary that can be used to compare articles/clusters.
-By analyzing all articles in clusters $c_i$ and $c_j$ we can identify the most relevant entities and keywords for each cluster.
-Additionally we can also assign weights to the concepts based on how frequently they occur in the articles in the cluster.
-From the list of relevant concepts and corresponding weights we can now define a number of features.
-\texttt{entityCosSim} represents the cosine similarity on the vector of entities mentioned in both clusters.
-Similarly, \texttt{keywordCosSim} represents the cosine similarity on the keywords only.
-In case the weights are not relevant, we have also added two features (\texttt{entityJaccardSim} and \texttt{keywordJaccardSim}), where Jaccard similarity is used instead of the cosine similarity.
-
-\item {\bf Miscellaneous features}.
-This group contains three miscellaneous features that seem discriminative but are unrelated to the previous two groups.
-The first feature is related to the event location.
-By analyzing the locations mentioned in the articles in a cluster we are able to estimate where the event occurred.
-The \texttt{hasSameLocation} feature is a boolean variable that is true when the location of the event in both clusters is the same.
-Another important feature is related to the time when the articles in the clusters were published.
-For articles in each cluster we first compute the average date and time when the articles were published.
-The \texttt{timeDiff} is then defined as the absolute difference in hours between the two computed dates.
-The last feature in this group is computed by analyzing the mentions of dates in the articles.
-Using an extensive set of regular expressions we are able to detect mentions of dates in different forms.
-To compute the feature \texttt{sharedDates} we start by identifying the list of dates mentioned in articles in each cluster separately.
-The value of \texttt{sharedDates} is then determined as the ratio of all dates that are detected in both clusters.
-
-\end{itemize}
-
-Using the described features and the user-provided label we can train a classification model $M$ that can be used in Algorithm~\ref{cluster_merge_algo2} to classify a new cluster pair as equivalent or not.
-The classification algorithm that we used to train a model was a linear SVM method.
-
-\subsection{Evaluation}
+\subsection{Evaluation Of Cross-Lingual Event Linking}
 
@@
 
-\section{Related work}
+\section{Discussion}
 
-\subsection{Cross-lingual Document Similarity}
-
-\subsection{Multilingual Latent Factor Models}
-
-Ask Wray
-
-\subsection{Cross-lingual Event Linking}
-
-\section{Discussion}
+\subsection{Scaling}
+\emph{Andrej asks who is the rest? I think that this section should be short.}
+100 languages!
+%Newsfeed volume
+Describe the newsfeed volume, number of events and the number of similarities that need to be computed. Compare the memory requirements of our current implementation to the rest.
+Comment on the caching gains in our implementation.
 

### 2015-03-22 rupnik2015.tex
related work

--- 2015-03-20 rupnik2015.tex
@@
 
-\usepackage{hyperref}
+\usepackage[implicit=false]{hyperref}
 
@@
 
-Event registry -> Monolingual events
-Events -> features (semantic + CCA) -> linked events
+Event registry $\Rightarrow$ Monolingual events $\Rightarrow$ features (semantic + CCA) $\Rightarrow$ linked events (crosslingual)
 
@@
 
-TODO
+In this section we will relate the method we use to other alternatives in the literature.
+We will first state some requirements for our system, which motivate our choice of focusing on a particular subset of approaches in the rest of the paper.
+
+The goal is to build a system that monitors global media and analyzes how events are being reported on.
+In our approach this boils down to two steps: tracking events separately in each language (based on language detection and an online clustering approach) and then connecting them.
+The pipeline needs to process millions of articles per day and perform billions of similarity computations each day.
+The system should support as many languages as possible.
+We focus on implementations that run on a single shared memory machine, as opposed to clusters of machines.
+This severely simplifies the software implementation and the system maintenance.
+To summarize, the following properties are desireable:
+\begin{itemize}
+\item{training} The building should scale to many languages and should be robust to the quality of training resources.
+The system should be able to take advantage of comparable corpus (as opposed to parallel translation based corpora), with missing data.
+\item{operation efficiency} The similarity computation should be fast.
+\item{operation cost} The system should run on a strong shared machine server. 
+\item{implementation} The system is simple to implement, with few parameters to tune.
+\end{itemize}
+We believe that a cross-lingual similarity component that meets such requirements is very desirable in a commercial setting, where several different costs have to be taken into consideration.
+
+There are three main families of approaches to cross-lingual similarity.
+\begin{itemize}
+\item{translation} The most obvious way to compare documents written in different languages is to use machine translation and perform monolingual similarity.
+One can use free tools such as Moses \cite{Hoang07moses:open} or translation services, such as Google Translate\footnote{https://translate.google.com/}.
+\item{probabilistic topic models}
+\item{matrix factorization}
+
+\end{itemize}
+
+Based on the discussion above, we chose to focus on methods based on vector space models and linear embeddings.
+We propose a method that is more efficient than a simple clustering approach as well as a popular approach (LSI), but is simple to optimize and use.
 
@@
 
-Standard vector space model \cite{xxx} represents documents as vectors, where each term corresponds to word or phrase in a fixed vocabulary.
+Standard vector space model \cite{Salton88term-weightingapproaches} represents documents as vectors, where each term corresponds to word or phrase in a fixed vocabulary.
 More formally, document $d$ is represented by a vector $x \in \RR^n$, where $n$ corresponds to the size of the vocabulary, and vector elements $x_k$ correspond to the number of times term $k$ occurred in the document, also called \emph{term frequency} or $TF_k(d)$.
@@
 The method is based on computing a truncated singular value decomposition of $X \approx U S V^T$.
-Since the matrix can be large we can use an iterative method like the Lanczos~\cite{matrix_comp} algorithm with reorthogonalization to find the left singular vectors (columns of $U$) corresponding to the largest singular values.
+Since the matrix can be large we can use an iterative method like the Lanczos~\cite{golub} algorithm with reorthogonalization to find the left singular vectors (columns of $U$) corresponding to the largest singular values.
 It turns out that the Lanczos method converges slowly as the gap between the leading singular values is small.
 Moreover, the Lanczos method is hard to parallelize.
-Instead we use a randomized version of the singular value decomposition (SVD) described in \ref{randomizedSVD} that can be viewed as a block Lanczos method.
+Instead we use a randomized version of the singular value decomposition (SVD) described in \cite{tropp} that can be viewed as a block Lanczos method.
 That enables us to use parallelization and speeds up the computation considerably.
@@
 
- Canonical Correlation Analysis (CCA)~\cite{Hotelling} is a dimensionality reduction technique similar to Principal Component Analysis (PCA)~\cite{Pearson1901On}, with an additional assumption that the data consists of feature vectors that arose from two sources (two views) that share some information (examples: bilingual document collection~\cite{mrpqr}, collection of images and captions~\cite{HardoonSS04}).
+ Canonical Correlation Analysis (CCA)~\cite{Hotelling} is a dimensionality reduction technique similar to Principal Component Analysis (PCA)~\cite{Pearson1901On}, with an additional assumption that the data consists of feature vectors that arose from two sources (two views) that share some information (examples: bilingual document collection~\cite{mrpqr}, collection of images and captions~\cite{Hardoon_usingimage}).
 Instead of looking for linear combinations of features that maximize the variance (PCA) we look for a linear combination of feature vectors from the first view and a linear combination for the second view, that are maximally correlated.
@@
 
-\subsection{Cross-Lingual Event Linking}
+\subsection{Evaluation Of Cross-Lingual Event Linking}
 
@@
 \vskip 0.2in
-%\bibliography{rupnik2015}
+\bibliography{jair}
 \bibliographystyle{theapa}

### 2015-03-23 rupnik2015.tex
Polish by Andrej

--- 2015-03-22 rupnik2015.tex
@@
 
-\usepackage[implicit=false]{hyperref}
+\usepackage{hyperref}
 
@@
 
-In this section we will relate the method we use to other alternatives in the literature.
-We will first state some requirements for our system, which motivate our choice of focusing on a particular subset of approaches in the rest of the paper.
-
-The goal is to build a system that monitors global media and analyzes how events are being reported on.
-In our approach this boils down to two steps: tracking events separately in each language (based on language detection and an online clustering approach) and then connecting them.
-The pipeline needs to process millions of articles per day and perform billions of similarity computations each day.
-The system should support as many languages as possible.
-We focus on implementations that run on a single shared memory machine, as opposed to clusters of machines.
-This severely simplifies the software implementation and the system maintenance.
-To summarize, the following properties are desireable:
-\begin{itemize}
-\item{training} The building should scale to many languages and should be robust to the quality of training resources.
-The system should be able to take advantage of comparable corpus (as opposed to parallel translation based corpora), with missing data.
-\item{operation efficiency} The similarity computation should be fast.
-\item{operation cost} The system should run on a strong shared machine server. 
-\item{implementation} The system is simple to implement, with few parameters to tune.
-\end{itemize}
-We believe that a cross-lingual similarity component that meets such requirements is very desirable in a commercial setting, where several different costs have to be taken into consideration.
-
-There are three main families of approaches to cross-lingual similarity.
-\begin{itemize}
-\item{translation} The most obvious way to compare documents written in different languages is to use machine translation and perform monolingual similarity.
-One can use free tools such as Moses \cite{Hoang07moses:open} or translation services, such as Google Translate\footnote{https://translate.google.com/}.
-\item{probabilistic topic models}
-\item{matrix factorization}
-
-\end{itemize}
-
-Based on the discussion above, we chose to focus on methods based on vector space models and linear embeddings.
-We propose a method that is more efficient than a simple clustering approach as well as a popular approach (LSI), but is simple to optimize and use.
+TODO
 
@@
 
-Standard vector space model \cite{Salton88term-weightingapproaches} represents documents as vectors, where each term corresponds to word or phrase in a fixed vocabulary.
+Standard vector space model \cite{xxx} represents documents as vectors, where each term corresponds to word or phrase in a fixed vocabulary.
 More formally, document $d$ is represented by a vector $x \in \RR^n$, where $n$ corresponds to the size of the vocabulary, and vector elements $x_k$ correspond to the number of times term $k$ occurred in the document, also called \emph{term frequency} or $TF_k(d)$.
@@
 The method is based on computing a truncated singular value decomposition of $X \approx U S V^T$.
-Since the matrix can be large we can use an iterative method like the Lanczos~\cite{golub} algorithm with reorthogonalization to find the left singular vectors (columns of $U$) corresponding to the largest singular values.
+Since the matrix can be large we can use an iterative method like the Lanczos~\cite{matrix_comp} algorithm with reorthogonalization to find the left singular vectors (columns of $U$) corresponding to the largest singular values.
 It turns out that the Lanczos method converges slowly as the gap between the leading singular values is small.
 Moreover, the Lanczos method is hard to parallelize.
-Instead we use a randomized version of the singular value decomposition (SVD) described in \cite{tropp} that can be viewed as a block Lanczos method.
+Instead we use a randomized version of the singular value decomposition (SVD) described in \ref{randomizedSVD} that can be viewed as a block Lanczos method.
 That enables us to use parallelization and speeds up the computation considerably.
@@
 
- Canonical Correlation Analysis (CCA)~\cite{Hotelling} is a dimensionality reduction technique similar to Principal Component Analysis (PCA)~\cite{Pearson1901On}, with an additional assumption that the data consists of feature vectors that arose from two sources (two views) that share some information (examples: bilingual document collection~\cite{mrpqr}, collection of images and captions~\cite{Hardoon_usingimage}).
+ Canonical Correlation Analysis (CCA)~\cite{Hotelling} is a dimensionality reduction technique similar to Principal Component Analysis (PCA)~\cite{Pearson1901On}, with an additional assumption that the data consists of feature vectors that arose from two sources (two views) that share some information (examples: bilingual document collection~\cite{mrpqr}, collection of images and captions~\cite{HardoonSS04}).
 Instead of looking for linear combinations of features that maximize the variance (PCA) we look for a linear combination of feature vectors from the first view and a linear combination for the second view, that are maximally correlated.
@@
 \subsection{Datasets}
-\emph{Andrej started shuffling it, will do it till Monday :) }
-
-{\oldText{Taken from ITI paper}
+%\emph{Andrej started shuffling it, will do it till Monday :) }
+%{\oldText{Taken from ITI paper}
 To investigate the empirical performance of the low rank approximations we will test the algorithms on a large-scale, real-world multilingual dataset that we extracted from Wikipedia by using inter-language links as an alignment.
@@
 The Wikipedia is constantly growing.
-There are currently four Wikipedias with more than $10^6$ articles, $40$ with more than $10^5$ articles, $100$ with more than $10^4$ articles, and $216$ with more than $1000$ articles.
-Wikipedia uses special user-friendly markup language that is very easy to write but very hard to parse.
+There are currently 12 Wikipedias with more than $10^6$ articles, $52$ with more than $10^5$ articles, $129$ with more than $10^4$ articles, and $236$ with more than $1000$ articles.
+%Wikipedia uses special user-friendly markup language that is very easy to write but very hard to parse.
 Simplicity of language can cause ambiguities and moreover it is constantly changing.
-For example, separator | can be used in different contexts.
+For example, separator | can be used in %different contexts.
 
-Wikipedia raw xml dumps of all currently $270$ active editions were downloaded from the Wikipedia dump page\href{http://dumps.wikimedia.org}.
+%Wikipedia raw xml dumps of all currently $288$ active editions were downloaded from the Wikipedia dump page\href{http://dumps.wikimedia.org}.
 The xml files are too large to be parsed with DOM like parser that needs to store the whole xml tree in the memory, instead we implemented Sax like parser that tries to simulate behavior of Wikipedia official parser and is as simple, fast and error prone as possible.
-We parse all Wikipedia markup but do not extend the templates.
+We parse all Wikipedia markup but do not expand the templates.
 Each Wikipedia page is embedded in the page tag.
@@
 This matrix enables us to build an alignment across all Wikipedia languages.
-}
+%}
 
@@
 \subsection{Experiments With Missing Alignment Data}
-\emph{Andrej started working on this, done by Monday}
-{\oldText{Taken from XLite paper}
-To investigate the empirical performance of our algorithm we will test it on a large-scale, real-world multilingual dataset that we extracted from Wikipedia  by  using so called 'inter-language' links as an alignment.
-We select a subset of Wikipedia languages containing three major languages, English--\emph{en} (hub language), Spanish--\emph{es}, Russian--\emph{ru}, and five minority (in the sense of Wikipedia sizes) languages, Slovenian--\emph{sl}, Piedmontese--\emph{pms}, Waray-Waray--\emph{war} (all with about 2 million native speakers), Creole--\emph{ht} (8 million native speakers), and Hindi--\emph{hi} (180 million native speakers).
-For preprocessing we remove the documents that contain less than 20 different words (stubs) and remove words occur in less than 50 documents as well as the top 100 most frequent words (in each language separately).
+%\emph{Andrej started working on this, done by Monday}
+%{\oldText{Taken from XLite paper}
+%To investigate the empirical performance of our algorithm we will test it on a large-scale, real-world multilingual dataset that we extracted from Wikipedia  by  using so called 'inter-language' links as an alignment.
+ In this subsection we will investigate the empirical performance of cca hub approach.
+We will demonstrate that this approach can be successfully applied even in the case of missing alignment information.
+ To this purpose we select a subset of Wikipedia languages containing three major languages, English--\emph{en} (hub language), Spanish--\emph{es}, Russian--\emph{ru}, and five minority (in the sense of Wikipedia sizes) languages, Slovenian--\emph{sl}, Piedmontese--\emph{pms}, Waray-Waray--\emph{war} (all with about 2 million native speakers), Creole--\emph{ht} (8 million native speakers), and Hindi--\emph{hi} (180 million native speakers).
+For preprocessing we remove the documents that contain less than 20 different words (stubs) and remove words occurring in less than 50 documents as well as the top 100 most frequent words (in each language separately).
 We represent the documents as normalized TFIDF\cite{Salton88term-weightingapproaches} weighted vectors.
-These are languages with Wikipedia article sizes comparable to the minority languages.
+ %These are languages with Wikipedia article sizes comparable to the minority languages.
 (since the number of speakers does not directly correlate with the number of Wikipedia articles).
-The prime hub candidate is the English language which is well aligned with all other Wikipedia languages, although the alignment quality varies quite a bit.
-Furthermore, we remove the documents that contain less than 20 different words and remove words that are too infrequent as well as the top 100 most frequent words in the vocabularies.
+The prime hub candidate is the English %language which is well aligned with all other Wikipedia languages, although the alignment quality varies quite a bit.
+Furthermore, we remove the documents that contain less than 20 different words and remove words that are too %infrequent as well as the top 100 most frequent words in the vocabularies.
+ Although the English language is well aligned with all Wikipedia languages, we must note that quality of alignment varies quite a bit, especially in the case of small Wikipedias.
  Furthermore, we call the document consisting of less than 20 different words, a stub.
@@
 This indicates a poor alignment which leads to poor performance.
-}
+%}
 
@@
 
-In Section~\ref{sec:algorithms} we have described three main algorithms for identifying similar articles in different languages.
+In Section~\ref{sec:models} we have described three main algorithms for identifying similar articles in different languages.
 These algorithms were $k$-means, LSI and CCA.
@@
 We can see that the most informative features for cluster linking are obtained using the CCA algorithm, while $k$-means performs the worse.
-{\oldText{TODO:} add some explanation why this might be so.}
-\emph{Andrej adds some tests, new experiment description graphs, and explanations for performance, done by Monday}
+Firstly we verified the AMPR performance of approaches by holding out 15000 aligned test documents and using 300000 aligned documents
+ as the training set. Picture \cite{pc:AMPR} shows AMPR score as the function of the number of feature vectors. It is clear that CCA outperforms LSI approach and K-means lags far behind.
+Then we verified how number of features influences the accuracy of cluster linking.
+As we can see from Table~\ref{table:linkingEvalAlgosMulti} choosing number of features vectors larger than 500 barely effects linking performance.
+That is probably due to overfitting.
+
+%{\oldText{TODO:} add some explanation why this might be so.}
+%\emph{Andrej adds some tests, new experiment description graphs, and explanations for performance, done by Monday}
 %15000 test documents, around 300 000 sth training documents
@@
 %500 is good, 800 better, 1000 overfitting, noise!
+\begin{figure}
+\label{pic:AMPR}
 \includegraphics{slika.png}
+\end{figure}
 %Why the hub approach?
@@
 %Maybe omit this table?
+\begin{table}[h]
+\caption{Accuracy of cluster linking with 500 - 800 - 1000 features obtained from different cross-lingual similarity algorithms}
+\label{table:linkingEvalAlgosMulti}
+\begin{center}
+\begin{tabular}{|c|c|c|c|}
+  \hline
+  \cline{1-4}
+  Models & CA & Precision & Recall \\ \cline{1-4}
+  CCA & 78.8 - 79.6 - 80.3 \% & 76.9 - 77.9 - 80.2 \%  & 82.1 - 82.3 - 80.4 \%  \\ \cline{1-4}
+  LSI & 78.9 - 78.7 - 80.2\% & 76.7 - 76.6 - 78.5 \%  & 82.8 - 82.3 - 82.8 \%  \\ \cline{1-4}
+  \hline
+\end{tabular}
+\end{center}
+\end{table}
 
@@
 For this purpose we tested the accuracies obtained using individual groups of features, as well as using different combination of groups.
-Since we found the CCA to perform best of the three algorithms, we used it to compute the values of the cross-lingual article linking features.
-Results of the evaluation are shown in Table\~ref{table:linkingEval}.
+Since we found the CCA to perform best of the three algorithms, we recomputed it and used it to compute the values of the cross-lingual article linking features.
+Results of the evaluation are shown in Table~\ref{table:linkingEval}.
 We can see that using a single group of features, the highest prediction accuracy can be achieved using the concept related features.
@@
   Models & CA & Precision & Recall \\ \cline{1-4}
-  CCA & 77.6\% & 0.783 & 0.811 \\ \cline{1-4}
-  $k$-means & 73.9\% & 0.695 & 0.846 \\ \cline{1-4}
-  LSI & 76.1\% & 0.755 & 0.769 \\ \cline{1-4}
+  CCA & 77.6\% & 78.3 \%  & 81.1\%  \\ \cline{1-4}
+  $k$-means & 73.9\% & 69.5\%  & 84.6\%  \\ \cline{1-4}
+  LSI & 76.1\% & 75.5\%  & 76.9\%  \\ \cline{1-4}
   \hline
@@
 100 languages!
-%Newsfeed volume
+%Memory 100 GB per day
+%Newsfeed volume 430k articles per day, 210k english, 36k Spanish, 27k German, 17k French, ....
+%Number of similarities, 430k similarities every 200ms, currently could handle 1M articles,
+%(n \over 2) 500X500 proxy matrices, 300000x500 Projector matrices, 430000x500 data
+%Premultiply with proxy matrices, compare blockwise!
+
 Describe the newsfeed volume, number of events and the number of similarities that need to be computed. Compare the memory requirements of our current implementation to the rest.
@@
 \vskip 0.2in
-\bibliography{jair}
+%\bibliography{rupnik2015}
 \bibliographystyle{theapa}

### 2015-03-23 rupnik2015.tex
related work

--- 2015-03-22 rupnik2015.tex
@@
 Finally, we present and interpret the experimental results and finally give conclusions and describe promising future directions.
-%
-%\section{Introduction}
-%
-%
-%{\oldText{Taken from XLite paper}
-%Document retrieval is a well-established problem in data mining.
-There have been a large number of different approaches put forward in the literature.
-In this paper we concentrate on a specific setting: multi-lingual corpora.
-As the availability of multi-lingual documents has exploded in the last few years, there is a pressing need for automatic cross-lingual processing tools.
-%
-%The prime example is Wikipedia - in 2001 the majority of pages were written in English, while in 2012, the percentage of English articles has dropped to 14\%.
-In this context, we look at how to find similar documents across languages.
-In particular, we do not assume the availability of machine translators, but rather try to frame the problem such that we can use well-established machine learning tools designed for monolingual text-mining tasks.
-This work represents the continuation of previous work \cite{nips, iti, nips2, sikdd} where we explored representations of documents which were valid over multiple languages.
-The representations could be interpreted as multi-lingual topics, which were then used as proxies to compute cross-lingual similarities between documents.
-We look at a specific aspect of this problem: the distribution of articles across languages in Wikipedia is not uniform.
-While the percentage of English articles has fallen, English is still the largest language and one of \emph{hub} languages which not only have an order of magnitude more articles than other languages, but also many comparable articles with most of the other languages.
-%
-%When doing document retrieval, we encounter two quite different situations.
-If for example, we are looking for a German article comparable to an English article, there is a large alignment between the document corpora (given by the intersection in articles in the two languages) making the problem well-posed.
-If, however, we look for a relevant Slovenian article given a Hindi article, the intersection is small, making the problem much harder.
-Since almost all languages have a large intersection in articles with \emph{hub} languages, the question we ask in this paper is: can we exploit hub languages to perform better document retrieval between non-hub languages?
-A positive answer would vastly improve cross-lingual analysis between less represented languages.
-In the following section, we introduce our representation followed by our experiments, which also shed light on the structural properties of the multilingual Wikipedia corpus.
-%}
 
@@
 This severely simplifies the software implementation and the system maintenance.
-To summarize, the following properties are desireable:
+To summarize, these are the relevant design aspects:
 \begin{itemize}
-\item{training} The building should scale to many languages and should be robust to the quality of training resources.
+\item \textbf{model building} - The building should scale to many languages and should be robust to the quality of training resources.
 The system should be able to take advantage of comparable corpus (as opposed to parallel translation based corpora), with missing data.
-\item{operation efficiency} The similarity computation should be fast.
-\item{operation cost} The system should run on a strong shared machine server. 
-\item{implementation} The system is simple to implement, with few parameters to tune.
+Supporting a new language should be simple.
+\item \textbf{operation efficiency} - The similarity computation should be fast - the system needs to handle billions of similarity computations per day.
+\item \textbf{operation cost} - The system should run on a strong shared machine server and not rely on paid services.
+\item \textbf{implementation} - The system is simple to implement, with few parameters to tune.
 \end{itemize}
@@
 \begin{itemize}
-\item{translation} The most obvious way to compare documents written in different languages is to use machine translation and perform monolingual similarity.
-One can use free tools such as Moses \cite{Hoang07moses:open} or translation services, such as Google Translate\footnote{https://translate.google.com/}.
-\item{probabilistic topic models}
-\item{matrix factorization}
-
+\item \textbf{translation} - The most obvious way to compare documents written in different languages is to use machine translation and perform monolingual similarity.
+One can use free tools such as Moses \cite{moses} or translation services, such as Google Translate\footnote{https://translate.google.com/}.
+There are two issues with such approaches: they solve a harder problem than needs to be solved and they are less robust to training resource quality - large sets of translated sentences are typically needed.
+Training Moses for languages with scarce linguistic resources is thus problematic.
+The issue with using online services such as Google Translate is that the APIs are limited and not free.
+The operation efficiency and cost requirements make translation based approaches less suited for our system.
+\item \textbf{probabilistic topic models} - not nice with term weighting, hyperparameters setting, burn-in and other black magic. Representing documents as latent vectors = inference problem.
+\item \textbf{matrix factorization} - nonnegative \cite{huertatowards}, svd based \cite{cl_lsi}, cca \cite{Hotelling}, opca \cite{platt2010translingual} - opca makes the assumption that the vocabulary sizes are the same ?
+that's lame.
 \end{itemize}
@@
 Based on the discussion above, we chose to focus on methods based on vector space models and linear embeddings.
-We propose a method that is more efficient than a simple clustering approach as well as a popular approach (LSI), but is simple to optimize and use.
+We propose a method that is more efficient than popular alternatives (a clustering based approach and latent semantic indexing), but is still simple to optimize and use.
 

### 2015-03-23 rupnik2015.tex
Comments added.
- added related work related to cluster linking

--- 2015-03-23 rupnik2015.tex
@@
 may not be perfect, we need to find the ``best" correspondence between clusters and events.
-In this paper, we test several different algorithms for both steps for perfomance in an integrated system.
-In particular, we test our approach within an online system for detection of world events, called Event Registry\cite{}.
+In this paper, we test several different algorithms for both steps for performance in an integrated system.
+In particular, we test our approach within an online system for detection of world events, called Event Registry\cite{Leban2014W, Leban2014I}.
 Event Registry is a repository of events, where events are automatically identified by analyzing news articles.
@@
 Collected articles are first semantically annotated by identifying in them mentions of relevant concepts -- either entities or important keywords.
-The disambiguated and entity linking of the concepts is done using Wikipedia as the main knowledge base.
+The disambiguation and entity linking of the concepts is done using Wikipedia as the main knowledge base.
    The key step in Event Registry is then to apply an online clustering algorithm on the articles in order to identify groups of articles that are discussing the same event.
@@
 
-The paper is organized as follows: we first provide an overview of the system as a whole (Section ), followd by details on each of the two steps described above.
+The paper is organized as follows: we first provide an overview of the system as a whole (Section ), followed by details on each of the two steps described above.
 We first introduce several approaches to measures of cross-lingual similarity (Section ) and then the algorithms for matching clusters to events.
@@
 \section{Cross-lingual Event Linking}
-The main application we test the above simliarity on  is cross-lingual event linking.
-In online media streams - particularly new articles - there is often duplication of reporting, different viewpoints or opinions, all centering around a single event.
-Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015.
-(Sports example)
 
-Informally, events are covered by many articles and the question we address is how to find all the articles in different languages that are describing a single event.
-We base our evaluation on an online system for detection of world events, called Event Registry\cite{}.
+The main application on which we test the above similiarity is cross-lingual event linking.
+In online media streams -- particularly new articles -- there is often duplication of reporting, different viewpoints or opinions, all centering around a single event.
+Same events are covered by many articles and the question we address is how to find all the articles in different languages that are describing a single event.
+We base our evaluation on an online system for detection of world events, called Event Registry.
 We do not address the problem of detection of events in this paper but rather consider the problem that given a series of events, how may we best ``tag" or match articles in different languages to these events.
@@
 
-In short, the article clustering works as follows (details are available in \cite{}).
+In short, the article clustering works as follows (details are available in \cite{Brank2014}).
 First we create a new document by joining the title and the body of the article.
@@
 
-TODO
+Although there are a number of services that aggregate news by identifying clusters of similar articles, there are very few service that provide linking of clusters over different languages.
+Google News as well as Yahoo!
+News are able to identify clusters of articles about same event, but they offer no linking of clusters across languages.
+The only service that provides cross-lingual cluster linking is the European Media Monitor (EMM)~\cite{Steinberger2008}.
+EMM clusters articles in 60 languages and then tries to determine which clusters of articles in different languages describe the same event.
+To achieve cluster linking, EMM uses three different language independent vector representations for each cluster.
+First vector contains the weighted list of references to countries mentioned in the articles, while the second vector contains the weighted list of mentioned people and organizations.
+The last vector contains the weighted list of Eurovoc subject domain descriptors.
+These descriptors are topics, such as \emph{air transport}, \emph{EC agreement}, \emph{competition} and \emph{pollution control} into which articles are automatically categorized~\cite{Pouliquen2006}.
+Similarity between clusters is then computed using a linear combination of the cosine similarities computed on the lists.
+
+A system that is significantly different, but worth mentioning is the GDELT project~\cite{Leetaru2013Gdelt}.
+In GDELT, events are also extracted from articles, but in their case, an event is specified in a form of a triple containing two actors and a relation.
+The project contains an extensive vocabulary of possible relations, mostly related to political events.
+In order to identify events, GDELT collects articles in more than 65 languages and uses machine translation to translate them to English.
+All information extraction is then done on the translated article.
 

### 2015-03-23 rupnik2015.tex
system overview

@@
 may not be perfect, we need to find the ``best" correspondence between clusters and events.
-In this paper, we test several different algorithms for both steps for performance in an integrated system.
-In particular, we test our approach within an online system for detection of world events, called Event Registry\cite{Leban2014W, Leban2014I}.
+In this paper, we test several different algorithms for both steps for perfomance in an integrated system.
+In particular, we test our approach within an online system for detection of world events, called Event Registry\cite{}.
 Event Registry is a repository of events, where events are automatically identified by analyzing news articles.
@@
 Collected articles are first semantically annotated by identifying in them mentions of relevant concepts -- either entities or important keywords.
-The disambiguation and entity linking of the concepts is done using Wikipedia as the main knowledge base.
+The disambiguated and entity linking of the concepts is done using Wikipedia as the main knowledge base.
    The key step in Event Registry is then to apply an online clustering algorithm on the articles in order to identify groups of articles that are discussing the same event.
@@
 
-The paper is organized as follows: we first provide an overview of the system as a whole (Section ), followed by details on each of the two steps described above.
+The paper is organized as follows: we first provide an overview of the system as a whole (Section ), followd by details on each of the two steps described above.
 We first introduce several approaches to measures of cross-lingual similarity (Section ) and then the algorithms for matching clusters to events.
@@
 
-\section{System Overview}
+\section{Pipeline}
 
-Event registry $\Rightarrow$ Monolingual events $\Rightarrow$ features (semantic + CCA) $\Rightarrow$ linked events (crosslingual)
+We base our techniques of cross-lingual event linking on an online system for detection of world events, called Event Registry\cite{}.
+Event Registry is a repository of events, where events are automatically identified by analyzing news articles.
+Event Registry works by collecting and analyzing news articles that are published by news outlets in different languages all over the world.
+Collected articles are first semantically annotated by identifying in them mentions of relevant concepts -- either entities or important keywords.
+The disambiguated and entity linking of the concepts is done using Wikipedia as the main knowledge base.
+An online clustering algorithm is then applied on the articles in order to identify groups of articles that are discussing the same event.
+For each new article, the clustering algorithm determines if the article should be assigned to some existing cluster or into a new cluster.
+The underlying assumption is that articles that are describing the same event are similar enough and will therefore be put into the same cluster.
+
+The processing is done independently for each language and the output is a clustering of articles in each language, with each cluster ideally representing an event, which are defined by topically, geographically and temporally.
+The goal of this paper is present approaches to  connecting the clusters/events across languages.
+
+Our approach is based on first defining a similarity function between articles in different languages (Section~\ref{}) and then we present several efficient approaches to linking the clusters based on features (Section~\ref{}).
+First, however we describe the pre-processing on the input articles which we then take as input.
+
+The system is designed to handle streams and so we describe how a document which enters the stream is processed.   A new document is first tokenized, stop words are removed and words are stemmed. 
+ What remains is represented in a vector-space model and normalized using $TF-IDF$ (see Section~\ref{} for the definition).
+Since each article is tagged with its language, cosine similiarity  is used to find the modst similar existing cluster, by comparing it to the centroid vector of each cluster.
+A selected threshold is used to determine if the article is not similar to any existing clusters (0.4 was used in our experiments).
+The article is then assigned to the corresponding cluster, otherwise a new cluster is created, initially containing only the single article.
+
+Since articles about an event are commonly written only for a short period of time, we remove clusters once the oldest article in the cluster becomes more than 4 days old.
+This housekeeping mechanism prevents the clustering from becoming slow and also ensures that articles are not assigned to obsolete clusters.
+
+The clusters, we consider must have a sufficient number of articles (which is a language dependent paramater).
+This is to ensure that we consider events.
+Once an event is identified(i.e.
+a cluster of sufficient size), a new unique ID is assigned to it and the main information about the event is then automatically extracted by analyzing the articles assigned to it.
+The extracted information includes properties such as the date of the event, the location, who is involved in it, what the event is about, etc.
+
+ Further details about the system can be found in \cite{}, while our full pipeline is shown in Figure~\ref{}.
+
+\begin{comment}
+In short, the article clustering works as follows (details are available in \cite{}).
+First we create a new document by joining the title and the body of the article.
+We then tokenize the document, remove the stop words and stem the words.
+The remaining words are represented as a vector in the vector-space model.
+Each value in the vector is additionally normalized using TF-IDF.
+After normalization, we use the vector to find the most similar existing cluster.
+To be able to compare the vector with a cluster we first need a vector representation for each of the clusters.
+This is done by computing a centroid vector of the cluster by simply averaging the vectors of the articles in the cluster.
+By computing the cosine similarity of the article vector with each of the centroid vectors we then find the cluster with the highest similarity.
+If the highest similarity is above the selected threshold (we used 0.4 in our experiments),
+
+In order to treat the cluster as describing an event in Event Registry,
+
+The clustering of articles is done for each language separately -- English articles are, for example, clustered separately from Spanish articles.
+When events are identified, they will therefore be identified from clusters in each language separately.
+As a result, the same event will be represented in Event Registry with two or more event IDs, and each event will only contain coverage in a single language.
+To avoid this problem we want to exploit the available information about the similarity of the articles in a cluster with articles in other languages.
+Using this information we hope to be able to connect clusters in different languages and represent them using a single event ID.
+\end{comment}
+%Event registry $\Rightarrow$ Monolingual events $\Rightarrow$ features (semantic + CCA) $\Rightarrow$ linked events (crosslingual)
 
@@
 \section{Cross-lingual Event Linking}
+The main application we test the above simliarity on  is cross-lingual event linking.
+In online media streams - particularly new articles - there is often duplication of reporting, different viewpoints or opinions, all centering around a single event.
+Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015.
+(Sports example)
 
-The main application on which we test the above similiarity is cross-lingual event linking.
-In online media streams -- particularly new articles -- there is often duplication of reporting, different viewpoints or opinions, all centering around a single event.
-Same events are covered by many articles and the question we address is how to find all the articles in different languages that are describing a single event.
-We base our evaluation on an online system for detection of world events, called Event Registry.
+Informally, events are covered by many articles and the question we address is how to find all the articles in different languages that are describing a single event.
+We base our evaluation on an online system for detection of world events, called Event Registry\cite{}.
 We do not address the problem of detection of events in this paper but rather consider the problem that given a series of events, how may we best ``tag" or match articles in different languages to these events.
@@
 
-\begin{comment}
-One of the use cases where article linking information can be applied is also in an online system for detection of world events, called Event Registry\cite{}.
-Event Registry is a repository of events, where events are automatically identified by analyzing news articles.
-The term event is vague and ambiguous, but for the practical purposes we define it as ``any significant happening that is being reported about in the media''.
-
-Event Registry works by collecting and analyzing news articles that are published by news outlets in different languages all over the world.
-Collected articles are first semantically annotated by identifying in them mentions of relevant concepts -- either entities or important keywords.
-The disambiguated and entity linking of the concepts is done using Wikipedia as the main knowledge base.
-The key step in Event Registry is then to apply an online clustering algorithm on the articles in order to identify groups of articles that are discussing the same event.
-For each new article, the clustering algorithm determines if the article should be assigned to some existing cluster or into a new cluster.
-The underlying assumption is that articles that are describing the same event are similar enough and will therefore be put into the same cluster.
-
-In short, the article clustering works as follows (details are available in \cite{Brank2014}).
-First we create a new document by joining the title and the body of the article.
-We then tokenize the document, remove the stop words and stem the words.
-The remaining words are represented as a vector in the vector-space model.
-Each value in the vector is additionally normalized using TF-IDF.
-After normalization, we use the vector to find the most similar existing cluster.
-To be able to compare the vector with a cluster we first need a vector representation for each of the clusters.
-This is done by computing a centroid vector of the cluster by simply averaging the vectors of the articles in the cluster.
-By computing the cosine similarity of the article vector with each of the centroid vectors we then find the cluster with the highest similarity.
-If the highest similarity is above the selected threshold (we used 0.4 in our experiments), the article is assigned to the corresponding cluster, otherwise a new cluster is created, initially containing only the single article.
-
-Since articles about an event are commonly written only for a short period of time, we remove clusters once the oldest article in the cluster becomes more than 4 days old.
-This housekeeping mechanism prevents the clustering from becoming slow and also ensures that articles are not assigned to obsolete clusters.
-
-In order to treat the cluster as describing an event in Event Registry, the cluster has to grow above a certain number of articles (the number depends on the language).
-This condition ensures that only content that receives a sufficient coverage in the media is treated as an event.
-Once a new event is identified, a new unique ID is assigned to it and the main information about the event is then automatically extracted by analyzing the articles assigned to it.
-The extracted information includes properties such as the date of the event, the location, who is involved in it, what the event is about, etc.
-
-The clustering of articles is done for each language separately -- English articles are, for example, clustered separately from Spanish articles.
-When events are identified, they will therefore be identified from clusters in each language separately.
-As a result, the same event will be represented in Event Registry with two or more event IDs, and each event will only contain coverage in a single language.
-To avoid this problem we want to exploit the available information about the similarity of the articles in a cluster with articles in other languages.
-Using this information we hope to be able to connect clusters in different languages and represent them using a single event ID.
-\end{comment}
-
 \subsection{Problem definition}
 
-The problem in the second phase of the pipeline is to identify the clusters (given?) describing the same events.
-Each article $a_i$ is written in a language $l$, where $l \in L = \{l_1,l_2,...,l_k\}$.
-For each language $l$, we generate and maintain a set of language-specific clusters $C_l$.
+The problem in the second phase of the pipeline is to identify the clusters describing the same events.
+Each article $a_i$ is written in a language $l$, where $\ell \in L = \{\ell_1,\ell_2,...,\ell_k\}$.
+For each language $\ell$, we generate and maintain a set of language-specific clusters $C_l$.
 That is, all articles are in language $l$.
-For any cluster $c_i \in C_{l_a}$, we would like to identify clusters $c_j \in C_{l_b}$ (where $l_a \neq l_b$) that describe the same event as $c_i$.
-We define \emph{equivalent clusters} as  pairs $(c_i,c_j)$ such that both clusters in the pair describe the same event in languages $C_{l_a}$ and $C_{l_b}$ respectively.
+For any cluster $c_i \in C_{\ell_a}$, we would like to identify clusters $c_j \in C_{\ell_b}$ (where $\ell_a \neq \ell_b$) that describe the same event as $c_i$.
+We define \emph{equivalent clusters} as  pairs $(c_i,c_j)$ such that both clusters in the pair describe the same event in languages $C_{\ell_a}$ and $C_{\ell_b}$ respectively.
+Furthermore, we iterate over all the languages we consider.
+That is, given a cluster in language $\ell_i$, we look for matching clusters in language $\ell_j$ for $i\neq j$  in a pairwise fashion.
+
+Furthermore, the matching of clusters is a \emph{generalized matching}.
+We cannot assume that there is only one cluster per language per event, nor can we assume complete coverage - that there exists at least one cluster per event in every language.
+This implies that we cannot make any assumptions on the matching, e.g.
+one-to-one, complete, etc.
+This excludes the use of standard weighted bipartite matching type of algortihms for this problem.
 
@@
 
-Although there are a number of services that aggregate news by identifying clusters of similar articles, there are very few service that provide linking of clusters over different languages.
-Google News as well as Yahoo!
-News are able to identify clusters of articles about same event, but they offer no linking of clusters across languages.
-The only service that provides cross-lingual cluster linking is the European Media Monitor (EMM)~\cite{Steinberger2008}.
-EMM clusters articles in 60 languages and then tries to determine which clusters of articles in different languages describe the same event.
-To achieve cluster linking, EMM uses three different language independent vector representations for each cluster.
-First vector contains the weighted list of references to countries mentioned in the articles, while the second vector contains the weighted list of mentioned people and organizations.
-The last vector contains the weighted list of Eurovoc subject domain descriptors.
-These descriptors are topics, such as \emph{air transport}, \emph{EC agreement}, \emph{competition} and \emph{pollution control} into which articles are automatically categorized~\cite{Pouliquen2006}.
-Similarity between clusters is then computed using a linear combination of the cosine similarities computed on the lists.
-
-A system that is significantly different, but worth mentioning is the GDELT project~\cite{Leetaru2013Gdelt}.
-In GDELT, events are also extracted from articles, but in their case, an event is specified in a form of a triple containing two actors and a relation.
-The project contains an extensive vocabulary of possible relations, mostly related to political events.
-In order to identify events, GDELT collects articles in more than 65 languages and uses machine translation to translate them to English.
-All information extraction is then done on the translated article.
+TODO
 

### 2015-03-23 rupnik2015.tex

--- 2015-03-23 rupnik2015.tex
@@
 
+\subsection{Problem definition}
+
+The problem in the second phase of the pipeline is to identify the clusters describing the same events.
+Each article $a_i$ is written in a language $l$, where $\ell \in L = \{\ell_1,\ell_2,...,\ell_k\}$.
+For each language $\ell$, we generate and maintain a set of language-specific clusters $C_l$.
+That is, all articles are in language $l$.
+For any cluster $c_i \in C_{\ell_a}$, we would like to identify clusters $c_j \in C_{\ell_b}$ (where $\ell_a \neq \ell_b$) that describe the same event as $c_i$.
+We define \emph{equivalent clusters} as  pairs $(c_i,c_j)$ such that both clusters in the pair describe the same event in languages $C_{\ell_a}$ and $C_{\ell_b}$ respectively.
+Furthermore, we iterate over all the languages we consider.
+That is, given a cluster in language $\ell_i$, we look for matching clusters in language $\ell_j$ for $i\neq j$  in a pairwise fashion.
+
+Furthermore, the matching of clusters is a \emph{generalized matching}.
+We cannot assume that there is only one cluster per language per event, nor can we assume complete coverage - that there exists at least one cluster per event in every language.
+This implies that we cannot make any assumptions on the matching, e.g.
+one-to-one, complete, etc.
+This excludes the use of standard weighted bipartite matching type of algortihms for this problem.
+
+We impose an additional requirement.
+Identification of the equivalent clusters cannot be done exhaustively -- for a cluster $c_i$ we wish to avoid testing all other clusters for all the other languages.
+Performing exhaustive comparison for each cluster $c_i$ would result in $O(N^2)$ tests, where $N$ is the number of all clusters (over all languages), which is not feasible for when the number of clusters is on the order of tens of thousands.
+
 \begin{comment}
@@
 Using this information we hope to be able to connect clusters in different languages and represent them using a single event ID.
-\end{comment}
 
@@
 
-We impose an additional requirement.
-Identification of the equivalent clusters cannot be done exhaustively -- for a cluster $c_i$ we wish to avoid testing all other clusters for all the other languages.
-Performing exhaustive comparison for each cluster $c_i$ would result in $O(N^2)$ tests, where $N$ is the number of all clusters (over all languages), which is not feasible for when the number of clusters is on the order of tens of thousands.
+\end{comment}
 

### 2015-03-23 rupnik2015.tex
abstract

@@
 \begin{abstract}
+Today, we follow news  which is distributed globally and reported on in many different languages.
+When a significant event occurs, it is often reported on in many different sources.
+In this work, our goal is to consider the problem tracking events in a large multilingual stream.
+In this paper, we consider a particular aspect of this problem, namely how to link collections of articles in different languages which refer to the same event.
+Given a multi-lingual stream,  articles are first clustered according in each language we consider separately.
+Given these clusters, we first propose a
+method to enables cross-lingual document similarity based on Wikipedia, which
+ enabling us to compute the similarity of any two articles regardless of language.
+Using this simliarity, we then propose an approach to find and match clusters across ;anguages which represent the same event.
+We provide an extensive evaluation of the performance of the system as a while as well as the quality of the similarity measure and linking algorithm.
+ \begin{comment}
 Main points:
@@
 \end{itemize}
-
+\end{comment}
 \end{abstract}

### 2015-03-24 rupnik2015.tex
All but Scale fixed

--- 2015-03-23 rupnik2015.tex
@@
 \usepackage{jair, theapa, rawfonts}
-\usepackage{graphicx}
+\usepackage{pgfplots} 
+%\usepackage{graphicx}
 
@@
 
-\section{Pipeline}
+\section{System Overview}
 
-We base our techniques of cross-lingual event linking on an online system for detection of world events, called Event Registry\cite{}.
-Event Registry is a repository of events, where events are automatically identified by analyzing news articles.
-Event Registry works by collecting and analyzing news articles that are published by news outlets in different languages all over the world.
-Collected articles are first semantically annotated by identifying in them mentions of relevant concepts -- either entities or important keywords.
-The disambiguated and entity linking of the concepts is done using Wikipedia as the main knowledge base.
-An online clustering algorithm is then applied on the articles in order to identify groups of articles that are discussing the same event.
-For each new article, the clustering algorithm determines if the article should be assigned to some existing cluster or into a new cluster.
-The underlying assumption is that articles that are describing the same event are similar enough and will therefore be put into the same cluster.
-
-The processing is done independently for each language and the output is a clustering of articles in each language, with each cluster ideally representing an event, which are defined by topically, geographically and temporally.
-The goal of this paper is present approaches to  connecting the clusters/events across languages.
-
-Our approach is based on first defining a similarity function between articles in different languages (Section~\ref{}) and then we present several efficient approaches to linking the clusters based on features (Section~\ref{}).
-First, however we describe the pre-processing on the input articles which we then take as input.
-
-The system is designed to handle streams and so we describe how a document which enters the stream is processed.   A new document is first tokenized, stop words are removed and words are stemmed. 
- What remains is represented in a vector-space model and normalized using $TF-IDF$ (see Section~\ref{} for the definition).
-Since each article is tagged with its language, cosine similiarity  is used to find the modst similar existing cluster, by comparing it to the centroid vector of each cluster.
-A selected threshold is used to determine if the article is not similar to any existing clusters (0.4 was used in our experiments).
-The article is then assigned to the corresponding cluster, otherwise a new cluster is created, initially containing only the single article.
-
-Since articles about an event are commonly written only for a short period of time, we remove clusters once the oldest article in the cluster becomes more than 4 days old.
-This housekeeping mechanism prevents the clustering from becoming slow and also ensures that articles are not assigned to obsolete clusters.
-
-The clusters, we consider must have a sufficient number of articles (which is a language dependent paramater).
-This is to ensure that we consider events.
-Once an event is identified(i.e.
-a cluster of sufficient size), a new unique ID is assigned to it and the main information about the event is then automatically extracted by analyzing the articles assigned to it.
-The extracted information includes properties such as the date of the event, the location, who is involved in it, what the event is about, etc.
-
- Further details about the system can be found in \cite{}, while our full pipeline is shown in Figure~\ref{}.
-
-\begin{comment}
-In short, the article clustering works as follows (details are available in \cite{}).
-First we create a new document by joining the title and the body of the article.
-We then tokenize the document, remove the stop words and stem the words.
-The remaining words are represented as a vector in the vector-space model.
-Each value in the vector is additionally normalized using TF-IDF.
-After normalization, we use the vector to find the most similar existing cluster.
-To be able to compare the vector with a cluster we first need a vector representation for each of the clusters.
-This is done by computing a centroid vector of the cluster by simply averaging the vectors of the articles in the cluster.
-By computing the cosine similarity of the article vector with each of the centroid vectors we then find the cluster with the highest similarity.
-If the highest similarity is above the selected threshold (we used 0.4 in our experiments),
-
-In order to treat the cluster as describing an event in Event Registry,
-
-The clustering of articles is done for each language separately -- English articles are, for example, clustered separately from Spanish articles.
-When events are identified, they will therefore be identified from clusters in each language separately.
-As a result, the same event will be represented in Event Registry with two or more event IDs, and each event will only contain coverage in a single language.
-To avoid this problem we want to exploit the available information about the similarity of the articles in a cluster with articles in other languages.
-Using this information we hope to be able to connect clusters in different languages and represent them using a single event ID.
-\end{comment}
-%Event registry $\Rightarrow$ Monolingual events $\Rightarrow$ features (semantic + CCA) $\Rightarrow$ linked events (crosslingual)
+Event registry $\Rightarrow$ Monolingual events $\Rightarrow$ features (semantic + CCA) $\Rightarrow$ linked events (crosslingual)
 
@@
 An aligned basis is obtained by first splitting $U$ vertically according to the number of dimensions of each language: $U = [U_1^T \cdots U_m^T]^T$.
-Then, the same as with $k$-means clustering, we compute the pseudoinverses$P_i = (U_i^T U_i)^{-1} U_i^T$.
+Then, the same as with $k$-means clustering, we compute the pseudoinverses $P_i = (U_i^T U_i)^{-1} U_i^T$.
 The matrices $P_i$ are used to change the basis from the standard basis in $\RR^{n_i}$ to the basis of columns of $U_i$.
@@
 By using the sample matrix notation $X_i$ and $X_j$ (assuming no data is missing for clearer presentation) this problem can be formulated as the following optimization problem:
-
 \begin{equation*}
@@
 \end{equation*}
-
 where $C_{i,i}$ and $C_{j,j}$ are empirical estimates of variances of $\mathcal{X}_i$ and $\mathcal{X}_j$ respectively and $C_{i,j}$ is an estimate for the covariance matrix.
@@
 If the matrices are not invertible, one can use a regularization technique by replacing $C_{i,i}$ with $(1- \kappa)C_{i,i} + \kappa I$, where $\kappa \in [0,1]$ is the regularization coefficient and $I$ is the identity matrix (the same can be applied to $C_{j,j}$.
-A single canonical variable is usually inadequate in representing the original random vector and typically one looks for $k$ projection pairs $(w_i^1, w_j^1),\ldots,(w_i^k, w_j^k)$, so that $w_i^{uT} \mathcal{X}_i$ and $w_j^{uT} \mathcal{X}_j$ are highly correlated and $w_i^{uT} \mathcal{X}_i$ is uncorrelated with $w_i^{vT} \mathcal{X}_i$  for $u \neq v$ and analogously for $w_j^u$ vectors.
+A single canonical variable is usually inadequate in representing the original random vector and typically one looks for $k$ projection pairs $(w_i^1, w_j^1),\ldots,(w_i^k, w_j^k)$, so that $(w_i^{u})^T \mathcal{X}_i$ and $(w_j^{u})^T \mathcal{X}_j$ are highly correlated and $(w_i^{u})^T \mathcal{X}_i$ is uncorrelated with $(w_i^{v})^T \mathcal{X}_i$  for $u \neq v$ and analogously for $w_j^u$ vectors.
 
@@
 
+\begin{comment}
+One of the use cases where article linking information can be applied is also in an online system for detection of world events, called Event Registry\cite{}.
+Event Registry is a repository of events, where events are automatically identified by analyzing news articles.
+The term event is vague and ambiguous, but for the practical purposes we define it as ``any significant happening that is being reported about in the media''.
+
+Event Registry works by collecting and analyzing news articles that are published by news outlets in different languages all over the world.
+Collected articles are first semantically annotated by identifying in them mentions of relevant concepts -- either entities or important keywords.
+The disambiguated and entity linking of the concepts is done using Wikipedia as the main knowledge base.
+The key step in Event Registry is then to apply an online clustering algorithm on the articles in order to identify groups of articles that are discussing the same event.
+For each new article, the clustering algorithm determines if the article should be assigned to some existing cluster or into a new cluster.
+The underlying assumption is that articles that are describing the same event are similar enough and will therefore be put into the same cluster.
+
+In short, the article clustering works as follows (details are available in \cite{}).
+First we create a new document by joining the title and the body of the article.
+We then tokenize the document, remove the stop words and stem the words.
+The remaining words are represented as a vector in the vector-space model.
+Each value in the vector is additionally normalized using TF-IDF.
+After normalization, we use the vector to find the most similar existing cluster.
+To be able to compare the vector with a cluster we first need a vector representation for each of the clusters.
+This is done by computing a centroid vector of the cluster by simply averaging the vectors of the articles in the cluster.
+By computing the cosine similarity of the article vector with each of the centroid vectors we then find the cluster with the highest similarity.
+If the highest similarity is above the selected threshold (we used 0.4 in our experiments), the article is assigned to the corresponding cluster, otherwise a new cluster is created, initially containing only the single article.
+
+Since articles about an event are commonly written only for a short period of time, we remove clusters once the oldest article in the cluster becomes more than 4 days old.
+This housekeeping mechanism prevents the clustering from becoming slow and also ensures that articles are not assigned to obsolete clusters.
+
+In order to treat the cluster as describing an event in Event Registry, the cluster has to grow above a certain number of articles (the number depends on the language).
+This condition ensures that only content that receives a sufficient coverage in the media is treated as an event.
+Once a new event is identified, a new unique ID is assigned to it and the main information about the event is then automatically extracted by analyzing the articles assigned to it.
+The extracted information includes properties such as the date of the event, the location, who is involved in it, what the event is about, etc.
+
+The clustering of articles is done for each language separately -- English articles are, for example, clustered separately from Spanish articles.
+When events are identified, they will therefore be identified from clusters in each language separately.
+As a result, the same event will be represented in Event Registry with two or more event IDs, and each event will only contain coverage in a single language.
+To avoid this problem we want to exploit the available information about the similarity of the articles in a cluster with articles in other languages.
+Using this information we hope to be able to connect clusters in different languages and represent them using a single event ID.
+\end{comment}
+
 \subsection{Problem definition}
 
-The problem in the second phase of the pipeline is to identify the clusters describing the same events.
-Each article $a_i$ is written in a language $l$, where $\ell \in L = \{\ell_1,\ell_2,...,\ell_k\}$.
-For each language $\ell$, we generate and maintain a set of language-specific clusters $C_l$.
+The problem in the second phase of the pipeline is to identify the clusters (given?) describing the same events.
+Each article $a_i$ is written in a language $l$, where $l \in L = \{l_1,l_2,...,l_k\}$.
+For each language $l$, we generate and maintain a set of language-specific clusters $C_l$.
 That is, all articles are in language $l$.
-For any cluster $c_i \in C_{\ell_a}$, we would like to identify clusters $c_j \in C_{\ell_b}$ (where $\ell_a \neq \ell_b$) that describe the same event as $c_i$.
-We define \emph{equivalent clusters} as  pairs $(c_i,c_j)$ such that both clusters in the pair describe the same event in languages $C_{\ell_a}$ and $C_{\ell_b}$ respectively.
-Furthermore, we iterate over all the languages we consider.
-That is, given a cluster in language $\ell_i$, we look for matching clusters in language $\ell_j$ for $i\neq j$  in a pairwise fashion.
-
-Furthermore, the matching of clusters is a \emph{generalized matching}.
-We cannot assume that there is only one cluster per language per event, nor can we assume complete coverage - that there exists at least one cluster per event in every language.
-This implies that we cannot make any assumptions on the matching, e.g.
-one-to-one, complete, etc.
-This excludes the use of standard weighted bipartite matching type of algortihms for this problem.
+For any cluster $c_i \in C_{l_a}$, we would like to identify clusters $c_j \in C_{l_b}$ (where $l_a \neq l_b$) that describe the same event as $c_i$.
+We define \emph{equivalent clusters} as  pairs $(c_i,c_j)$ such that both clusters in the pair describe the same event in languages $C_{l_a}$ and $C_{l_b}$ respectively.
 
@@
 We can see that the most informative features for cluster linking are obtained using the CCA algorithm, while $k$-means performs the worse.
-Firstly we verified the AMPR performance of approaches by holding out 15000 aligned test documents and using 300000 aligned documents
- as the training set. Picture \cite{pc:AMPR} shows AMPR score as the function of the number of feature vectors. It is clear that CCA outperforms LSI approach and K-means lags far behind.
-Then we verified how number of features influences the accuracy of cluster linking.
-As we can see from Table~\ref{table:linkingEvalAlgosMulti} choosing number of features vectors larger than 500 barely effects linking performance.
+Firstly we verified the AMRR (average of mean reciprocal ranks over all language pairs) performance of approaches by holding out $15000$ aligned test documents and using $300000$ aligned documents
+ as the training set. Picture \ref{pc:AMRR} shows AMRR score as the function of the number of feature vectors. It is clear that CCA outperforms LSI approach and K-means lags far behind.
+Furthermore we inspected how number of features influences the accuracy of cluster linking.
+As we can see from Table~\ref{table:linkingEvalAlgos} choosing number of features vectors larger than $500$ barely effects linking performance.
 That is probably due to overfitting.
+Moreover we tested the performance of detecting large/small cluster linking.
+As expected we obtained that it is harder to link small clusters and we need better retrieval.
+Hub CCA outperform other methods in this sense also, probably as it has the best retrieval performance, see Figure~\ref{pic:AMRR}.
 
@@
 \begin{figure}
-\label{pic:AMPR}
-\includegraphics{slika.png}
+\caption{Average of mean reciprocal ranks}
+\label{pic:AMRR}
+%\includegraphics{slika.png}
+\input{retrieval}
 \end{figure}
@@
 \caption{Accuracy of cluster linking with 500 - 800 - 1000 features obtained from different cross-lingual similarity algorithms}
-\label{table:linkingEvalAlgosMulti}
+\label{table:linkingEvalAlgos}
 \begin{center}
@@
   Models & CA & Precision & Recall \\ \cline{1-4}
-  CCA & 78.8 - 79.6 - 80.3 \% & 76.9 - 77.9 - 80.2 \%  & 82.1 - 82.3 - 80.4 \%  \\ \cline{1-4}
-  LSI & 78.9 - 78.7 - 80.2\% & 76.7 - 76.6 - 78.5 \%  & 82.8 - 82.3 - 82.8 \%  \\ \cline{1-4}
+  CCA      & 78.2 - 79.6 - 80.3 \% & 76.3 - 78.0 - 80.5 \%  & 81.6 - 82.1 - 79.9 \%  \\ \cline{1-4}
+  LSI      & 78.9 - 78.7 - 80.6 \%  & 76.8 - 77.0 - 78.7 \%  & 83.3 - 80.6 - 83.6 \%  \\ \cline{1-4}
+ $k$-means & 73.9\phantom{ - 78.7 - 80.6 }\% & 69.5\phantom{ - 78.7 - 80.6 }\%  & 84.6\phantom{ - 78.7 - 80.6 }\%  \\ \cline{1-4}
+\end{tabular}
+\end{center}
+\end{table}
+
+\begin{table}[h]
+\caption{Accuracy of large/small cluster linking with $500$ features obtained from different cross-lingual similarity algorithms}
+\label{table:linkingEvalAlgosLargeSmall}
+\begin{center}
+\begin{tabular}{|c|c|c|c|}
   \hline
+  \cline{1-4}
+  Models & CA & Precision & Recall \\ \cline{1-4}
+  CCA      & 81.2 - 77.8 \% & 80.5 - 74.5 \%  & 91.3 - 57.5 \%  \\ \cline{1-4}
+  LSI      & 82.8 - 76.4 \% & 81.3 - 70.9 \%  & 93.1 - 57.5 \%  \\ \cline{1-4}
+ $k$-means & 75.5 - 71.2 \% & 72.8 - 70.8 \%  & 95.3 - 36.2 \%  \\ \cline{1-4}
 \end{tabular}
@@
 For this purpose we tested the accuracies obtained using individual groups of features, as well as using different combination of groups.
-Since we found the CCA to perform best of the three algorithms, we recomputed it and used it to compute the values of the cross-lingual article linking features.
+Since we found the CCA to perform best of the three algorithms, we recomputed $500$ feature vectors with better accuracy and used it to compute the values of the cross-lingual article linking features.
 Results of the evaluation are shown in Table~\ref{table:linkingEval}.
@@
 
-\begin{table}[h]
-\caption{Accuracy of cluster linking with features obtained from different cross-lingual similarity algorithms}
-\label{table:linkingEvalAlgos}
-\begin{center}
-\begin{tabular}{|c|c|c|c|}
-  \hline
-  \cline{1-4}
-  Models & CA & Precision & Recall \\ \cline{1-4}
-  CCA & 77.6\% & 78.3 \%  & 81.1\%  \\ \cline{1-4}
-  $k$-means & 73.9\% & 69.5\%  & 84.6\%  \\ \cline{1-4}
-  LSI & 76.1\% & 75.5\%  & 76.9\%  \\ \cline{1-4}
-  \hline
-\end{tabular}
-\end{center}
-\end{table}
+%\begin{table}[h]
+%\caption{Accuracy of cluster linking with features obtained from different cross-lingual similarity algorithms}
+%\label{table:linkingEvalAlgos}
+%\begin{center}
+%\begin{tabular}{|c|c|c|c|}
+%  \hline
+%  \cline{1-4}
+%  Models & CA & Precision & Recall \\ \cline{1-4}
+%  CCA & 77.6\% & 78.3 \%  & 81.1\%  \\ \cline{1-4}
+%  $k$-means & 73.9\% & 69.5\%  & 84.6\%  \\ \cline{1-4}
+%  LSI & 79.2\% & 78.9\%  & 80.6\%  \\ \cline{1-4}
+%  \hline
+%\end{tabular}
+%\end{center}
+%\end{table}
 
@@
 \subsection{Scaling}
-\emph{Andrej asks who is the rest? I think that this section should be short.}
+\emph{I think that this section should be short.}
 100 languages!

### 2015-03-24 retrieval/plot_retrieval.m

@@ -30,14 +30,14 @@ load RetrievalKMEANS.mat
 %   resultKMEANS                 1x10             1200  cell      
 hold on
 %Star CCA the best
-plot(dim, [resultCCA{:}], 'red');
+plot(dim, [resultCCA{:}], '-red');
 %Closely followed by LSI with S
-plot(dim, [resultLSI_WITH_S{:}], 'green');
+%plot(dim, [resultLSI_WITH_S{:}], 'green');
 %Followed by LSI without S
-plot(dim, [resultLSI_WITHOUT_S{:}], 'blue');
+plot(dim, [resultLSI_WITHOUT_S{:}], '-.blue');
 %Last place, slow Kmeans
-plot(dim, [resultKMEANS{:}], 'black');
-legend('Hub CCA', 'LSI with S', 'LSI without S', 'Kmeans');
+plot(dim, [resultKMEANS{:}], 'black.-.');
+legend('Hub CCA', 'LSI', 'k-means');
 xlabel('Number of feature vectors');
-ylabel('AMPR');
+ylabel('AMRR');
 

### 2015-03-24 rupnik2015.tex

--- 2015-03-23 rupnik2015.tex
@@
   Models & CA & Precision & Recall \\ \cline{1-4}
-  CCA & 78.8 - 79.6 - 80.3 \% & 76.9 - 77.9 - 80.2 \%  & 82.1 - 82.3 - 80.4 \%  \\ \cline{1-4}
-  LSI & 78.9 - 78.7 - 80.2\% & 76.7 - 76.6 - 78.5 \%  & 82.8 - 82.3 - 82.8 \%  \\ \cline{1-4}
-  \hline
+  CCA      & 78.2 - 79.6 - 80.3 \% & 76.3 - 78.0 - 80.5 \%  & 81.6 - 82.1 - 79.9 \%  \\ \cline{1-4}
+  LSI      & 78.9 - 78.7 - 80.6 \%  & 76.8 - 77.0 - 78.7 \%  & 83.3 - 80.6 - 83.6 \%  \\ \cline{1-4}
+ $k$-means & 73.9\phantom{ - 78.7 - 80.6 }\% & 69.5\phantom{ - 78.7 - 80.6 }\%  & 84.6\phantom{ - 78.7 - 80.6 }\%  \\ \cline{1-4}
 \end{tabular}
@@
 
+\begin{table}[h]
+\caption{Accuracy of large/small cluster linking with $500$ features obtained from different cross-lingual similarity algorithms}
+\label{table:linkingEvalAlgosLargeSmall}
+\begin{center}
+\begin{tabular}{|c|c|c|c|}
+  \hline
+  \cline{1-4}
+  Models & CA & Precision & Recall \\ \cline{1-4}
+  CCA      & 81.2 - 77.8 \% & 80.5 - 74.5 \%  & 91.3 - 57.5 \%  \\ \cline{1-4}
+  LSI      & 82.8 - 76.4 \% & 81.3 - 70.9 \%  & 93.1 - 57.5 \%  \\ \cline{1-4}
+ $k$-means & 75.5 - 71.2 \% & 72.8 - 70.8 \%  & 95.3 - 36.2 \%  \\ \cline{1-4}
+\end{tabular}
+\end{center}
+\end{table}
+
 In the second experiment we wanted to evaluate how relevant individual groups of features are to correctly determine cluster equivalence.
 For this purpose we tested the accuracies obtained using individual groups of features, as well as using different combination of groups.
-Since we found the CCA to perform best of the three algorithms, we recomputed it and used it to compute the values of the cross-lingual article linking features.
+Since we found the CCA to perform best of the three algorithms, we recomputed $500$ feature vectors with better accuracy and used it to compute the values of the cross-lingual article linking features.
 Results of the evaluation are shown in Table~\ref{table:linkingEval}.

### 2015-03-24 rupnik2015.tex
Scale second draft. Conclusion missing.

@@
 Firstly we verified the AMRR (average of mean reciprocal ranks over all language pairs) performance of approaches by holding out $15000$ aligned test documents and using $300000$ aligned documents
- as the training set. Picture \ref{pc:AMRR} shows AMRR score as the function of the number of feature vectors. It is clear that CCA outperforms LSI approach and K-means lags far behind.
+ as the training set. Figure~\ref{pc:AMRR} shows AMRR score as the function of the number of feature vectors. It is clear that CCA outperforms LSI approach and K-means lags far behind.
 Furthermore we inspected how number of features influences the accuracy of cluster linking.
@@
 That is probably due to overfitting.
-Moreover we tested the performance of detecting large/small cluster linking.
+Moreover we tested the performance of detecting large/small cluster linking, see Table~\ref{table:linkingEvalAlgosLargeSmall}.
 As expected we obtained that it is harder to link small clusters and we need better retrieval.
-Hub CCA outperform other methods in this sense also, probably as it has the best retrieval performance, see Figure~\ref{pic:AMRR}.
+Hub CCA outperforms other methods in this sense also, probably as it has the best retrieval performance, see Figure~\ref{pic:AMRR}.
 
@@
 %Premultiply with proxy matrices, compare blockwise!
-
-Describe the newsfeed volume, number of events and the number of similarities that need to be computed. Compare the memory requirements of our current implementation to the rest.
-Comment on the caching gains in our implementation.
+One of the main advantages of our approach is that scales really well.
+It is fast, very robust to quality of training data, easily extendable, simple to implement and with low machine requirements.
+Ut currently runs on two Intel Xeon E5-2667 v2, 3.30GHz processor machine with 256GB of RAM.
+That is enough to do similarity computation over 150 languages if needed.
+It currently uses Wikipedia as freely available knowledge base.
+
+Currently we compute similarities over $24$ languages with tags: eng, spa, deu, zho, ita, fra, rus, swe, nld, tur, jpn, por, ara, fin, ron, kor, hrv, tam, hun, slv, pol, srp, cat, ukr but we support up to $100$ top Wikipedis languages.
+Our data stream is Newsfeed(http://newsfeed.ijs.si/) that produces $430k$ unique articles per day.
+Our system currently computes 400k similiarites per 200ms, that means that we compute $16*10^10$ similarities per day.
+We
+store one day buffer for every language, for that we need 1.5GB of memory, we store documents as 500 dimensional vectors. For each article we compute top $10$ most similar ones in every language.
+For all linear algebra matrix and vector operations we use high performance numerical libraries as BLAS, OPENBLAS and Intel MKL. This currently allows us to process more than million articles per day.
+In our current implementation we use variation of hub approach.
+Our projector matrices are of size $500\times 300000,$ so every projector takes about $1.1$ GB of RAM.
+Moreover we need proxy matrices of size $500x500$ for every language pair.
+That is 0.5 GB for $24$ languages and $9.2$ GB for $100$ languages.
+All together we need around 135 GB of RAM for the system with 100 languages.
+Using proxy matrices we can project all the input documents in the common space.
+That enables us to do blockwise similarity computations and
+additional improvement in our system efficiency.
+Our code can be therefore easily parallelized using matrix multiplication instead of doing more matrix - vector multiplications.
+This speeds up our code by factor around $4.$ In this way we obtain some caching gains and ability to use vectorization.
+Our system is also easily extendable. If we want to add a language we only need to compute a projector matrix and proxy matrices with all other already available languages.
+
+Comment on the caching gains in our implementation.¸¸Add the conclusion if the story.
 

### 2015-03-24 rupnik2015.tex
related work

--- 2015-03-24 rupnik2015.tex
@@
 \usepackage{jair, theapa, rawfonts}
-\usepackage{pgfplots} 
-%\usepackage{graphicx}
+\usepackage{graphicx}
 
@@
 This severely simplifies the software implementation and the system maintenance.
-To summarize, the following properties are desireable:
+To summarize, these are the relevant design aspects:
 \begin{itemize}
-\item{training} The building should scale to many languages and should be robust to the quality of training resources.
+\item \textbf{model building} - The building should scale to many languages and should be robust to the quality of training resources.
 The system should be able to take advantage of comparable corpus (as opposed to parallel translation based corpora), with missing data.
-\item{operation efficiency} The similarity computation should be fast.
-\item{operation cost} The system should run on a strong shared machine server.
-\item{implementation} The system is simple to implement, with few parameters to tune.
+Supporting a new language should be simple.
+\item \textbf{operation efficiency} - The similarity computation should be fast - the system needs to handle billions of similarity computations per day.
+\item \textbf{operation cost} - The system should run on a strong shared machine server and not rely on paid services.
+\item \textbf{implementation} - The system is simple to implement, with few parameters to tune.
 \end{itemize}
@@
 \begin{itemize}
-\item{translation} The most obvious way to compare documents written in different languages is to use machine translation and perform monolingual similarity.
-One can use free tools such as Moses \cite{Hoang07moses:open} or translation services, such as Google Translate\footnote{https://translate.google.com/}.
-\item{probabilistic topic models}
-\item{matrix factorization}
+\item \textbf{translation} - The most obvious way to compare documents written in different languages is to use machine translation and perform monolingual similarity.
+One can use free tools such as Moses \cite{moses} or translation services, such as Google Translate\footnote{https://translate.google.com/}.
+There are two issues with such approaches: they solve a harder problem than needs to be solved and they are less robust to training resource quality - large sets of translated sentences are typically needed.
+Training Moses for languages with scarce linguistic resources is thus problematic.
+The issue with using online services such as Google Translate is that the APIs are limited and not free.
+The operation efficiency and cost requirements make translation based approaches less suited for our system.
+\item \textbf{probabilistic topic models} -  There exist many variants to modelling documents in a language independent way by using probabilistic graphical models.
+The models include:  Joint Probabilistic Latent Semantic Analysis (JPLSA) \cite{platt2010translingual}, Coupled Probabilistic LSA (CPLSA) \cite{platt2010translingual}, Probabilistic Cross-Lingual LSA\cite{PCL_LSA} and Polylingual Topic Models (PLTM) which is a Bayesian version of PCL_LSA \cite{polyLDA}.
+The methods (except for CPLSA) describe the multilingual document collections as samples from generative probabilistic models, with variations on the assumptions on the model structure.
+The topics represent latent variables that are used to generate observed variables (words), a process specific to each language.
+The parameter estimation is posed as an inference problem which is typically intractable and one usually solves it using approximate techniques.
+Most variants of solutions are based on Gibbs sampling or Variational Inference, which are not trivial to implement and may require an experienced practitioner to apply.
+Furthermore, representing a new document as a mixture of topics is another potentially hard inference problem which needs to be solved.
+\item \textbf{classification problem related} - In approaches such as \cite{cotrain} and \cite{svm2k} the authors find linear embeddings of the data that support good classification results, based on a given classification problem training data.
+We do not consider this family of approaches, since for general language pairs classifier training sets may not be available.
+\item \textbf{matrix factorization} - Non-negative matrix factorization based \cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CLLSI\cite{cl_lsi}, Canonical Correlation Analysis (CCA) \cite{Hotelling}, Oriented Principal Component Analysis \cite{platt2010translingual}.
+The quadratic time and space dependency of the OPCA method makes it impractical for large scale purposes.
+In addition, OPCA forces the vocabulary sizes for all languages to be the same, which is less intuitive.
+For our setting, the method in \cite{nonnegfactor_lsi} has a prohibitively high computational cost when building models (it uses dense matrices of dimensions that is the product of training set size and vocabulary size).
+Our proposed approach combines CCA and CLLSI, which will be presented in more detail.
 
@@
 An aligned basis is obtained by first splitting $U$ vertically according to the number of dimensions of each language: $U = [U_1^T \cdots U_m^T]^T$.
-Then, the same as with $k$-means clustering, we compute the pseudoinverses $P_i = (U_i^T U_i)^{-1} U_i^T$.
+Then, the same as with $k$-means clustering, we compute the pseudoinverses$P_i = (U_i^T U_i)^{-1} U_i^T$.
 The matrices $P_i$ are used to change the basis from the standard basis in $\RR^{n_i}$ to the basis of columns of $U_i$.
@@
 By using the sample matrix notation $X_i$ and $X_j$ (assuming no data is missing for clearer presentation) this problem can be formulated as the following optimization problem:
+
 \begin{equation*}
@@
 \end{equation*}
+
 where $C_{i,i}$ and $C_{j,j}$ are empirical estimates of variances of $\mathcal{X}_i$ and $\mathcal{X}_j$ respectively and $C_{i,j}$ is an estimate for the covariance matrix.
@@
 If the matrices are not invertible, one can use a regularization technique by replacing $C_{i,i}$ with $(1- \kappa)C_{i,i} + \kappa I$, where $\kappa \in [0,1]$ is the regularization coefficient and $I$ is the identity matrix (the same can be applied to $C_{j,j}$.
-A single canonical variable is usually inadequate in representing the original random vector and typically one looks for $k$ projection pairs $(w_i^1, w_j^1),\ldots,(w_i^k, w_j^k)$, so that $(w_i^{u})^T \mathcal{X}_i$ and $(w_j^{u})^T \mathcal{X}_j$ are highly correlated and $(w_i^{u})^T \mathcal{X}_i$ is uncorrelated with $(w_i^{v})^T \mathcal{X}_i$  for $u \neq v$ and analogously for $w_j^u$ vectors.
+A single canonical variable is usually inadequate in representing the original random vector and typically one looks for $k$ projection pairs $(w_i^1, w_j^1),\ldots,(w_i^k, w_j^k)$, so that $w_i^{uT} \mathcal{X}_i$ and $w_j^{uT} \mathcal{X}_j$ are highly correlated and $w_i^{uT} \mathcal{X}_i$ is uncorrelated with $w_i^{vT} \mathcal{X}_i$  for $u \neq v$ and analogously for $w_j^u$ vectors.
 
@@
 The prototypical  exampling  in the case of Wikipedia is English.
-We use the following notation to define subsets of the multilingual comparable corpus: let $a(i,j)$ denote the index set of all multilingual documents with non-missing data for the $i$-th and $j$-th language:  $a(i,j) = \left\{k | d_k = (u_1,...,u_m), u_i \neq \emptyset, u_j \neq \emptyset \right\},$ and let $a(i)$ denote the index set of all multilingual documents with non missing data for the $i$-th language.
+We use the following notation to define subsets of the multilingual comparable corpus: let $a(i,j)$ denote the index set of all multilingual documents with non-missing data for the $i$-th and $j$-th language:  $$a(i,j) = \left\{k~ |~ d_k = (u_1,...,u_m), u_i \neq \emptyset, u_j \neq \emptyset \right\},$$ and let $a(i)$ denote the index set of all multilingual documents with non missing data for the $i$-th language.
 
@@
 The first step in our method is to project $X_1, \ldots, X_m$ to lower dimensional spaces without destroying the cross-lingual structure.
-Treating the nonzero columns of $X_i$ as observation vectors sampled from an underlying distribution $\mathcal{X}_i \in V_i = \RR^{n_i}$, we can analyze the empirical cross-covariance matrices: $C_{i,j} = \frac{1}{|a(i,j)|-1 }\sum_{\ell \in a(i,j)} (X_i^{\ell} - c_i)\cdot (X_j^{\ell} - c_j)^T$, where $c_i = \frac{1}{a_i} \sum_{\ell \in a(i)}X_i^{\ell}$.
+Treating the nonzero columns of $X_i$ as observation vectors sampled from an underlying distribution $\mathcal{X}_i \in V_i = \RR^{n_i}$, we can analyze the empirical cross-covariance matrices:
+$$C_{i,j} = \frac{1}{|a(i,j)|-1 }\sum_{\ell \in a(i,j)} (X_i^{\ell} - c_i)\cdot (X_j^{\ell} - c_j)^T,$$
+ where $c_i = \frac{1}{a_i} \sum_{\ell \in a(i)}X_i^{\ell}$.
 By finding low rank approximations of $C_{i,j}$ we can identify the subspaces of $V_i$ and $V_j$ that are relevant for extracting linear patterns between $\mathcal{X}_i$ and $\mathcal{X}_j$.
@@
 We can see that the most informative features for cluster linking are obtained using the CCA algorithm, while $k$-means performs the worse.
-Firstly we verified the AMRR (average of mean reciprocal ranks over all language pairs) performance of approaches by holding out $15000$ aligned test documents and using $300000$ aligned documents
- as the training set. Picture \ref{pc:AMRR} shows AMRR score as the function of the number of feature vectors. It is clear that CCA outperforms LSI approach and K-means lags far behind.
-Furthermore we inspected how number of features influences the accuracy of cluster linking.
-As we can see from Table~\ref{table:linkingEvalAlgos} choosing number of features vectors larger than $500$ barely effects linking performance.
+Firstly we verified the AMPR performance of approaches by holding out 15000 aligned test documents and using 300000 aligned documents
+ as the training set. Picture \cite{pc:AMPR} shows AMPR score as the function of the number of feature vectors. It is clear that CCA outperforms LSI approach and K-means lags far behind.
+Then we verified how number of features influences the accuracy of cluster linking.
+As we can see from Table~\ref{table:linkingEvalAlgosMulti} choosing number of features vectors larger than 500 barely effects linking performance.
 That is probably due to overfitting.
-Moreover we tested the performance of detecting large/small cluster linking.
-As expected we obtained that it is harder to link small clusters and we need better retrieval.
-Hub CCA outperform other methods in this sense also, probably as it has the best retrieval performance, see Figure~\ref{pic:AMRR}.
 
@@
 \begin{figure}
-\caption{Average of mean reciprocal ranks}
-\label{pic:AMRR}
-%\includegraphics{slika.png}
-\input{retrieval}
+\label{pic:AMPR}
+\includegraphics{slika.png}
 \end{figure}
@@
 \caption{Accuracy of cluster linking with 500 - 800 - 1000 features obtained from different cross-lingual similarity algorithms}
-\label{table:linkingEvalAlgos}
+\label{table:linkingEvalAlgosMulti}
 \begin{center}
@@
   Models & CA & Precision & Recall \\ \cline{1-4}
-  CCA      & 78.2 - 79.6 - 80.3 \% & 76.3 - 78.0 - 80.5 \%  & 81.6 - 82.1 - 79.9 \%  \\ \cline{1-4}
-  LSI      & 78.9 - 78.7 - 80.6 \%  & 76.8 - 77.0 - 78.7 \%  & 83.3 - 80.6 - 83.6 \%  \\ \cline{1-4}
- $k$-means & 73.9\phantom{ - 78.7 - 80.6 }\% & 69.5\phantom{ - 78.7 - 80.6 }\%  & 84.6\phantom{ - 78.7 - 80.6 }\%  \\ \cline{1-4}
-\end{tabular}
-\end{center}
-\end{table}
-
-\begin{table}[h]
-\caption{Accuracy of large/small cluster linking with $500$ features obtained from different cross-lingual similarity algorithms}
-\label{table:linkingEvalAlgosLargeSmall}
-\begin{center}
-\begin{tabular}{|c|c|c|c|}
+  CCA & 78.8 - 79.6 - 80.3 \% & 76.9 - 77.9 - 80.2 \%  & 82.1 - 82.3 - 80.4 \%  \\ \cline{1-4}
+  LSI & 78.9 - 78.7 - 80.2\% & 76.7 - 76.6 - 78.5 \%  & 82.8 - 82.3 - 82.8 \%  \\ \cline{1-4}
   \hline
-  \cline{1-4}
-  Models & CA & Precision & Recall \\ \cline{1-4}
-  CCA      & 81.2 - 77.8 \% & 80.5 - 74.5 \%  & 91.3 - 57.5 \%  \\ \cline{1-4}
-  LSI      & 82.8 - 76.4 \% & 81.3 - 70.9 \%  & 93.1 - 57.5 \%  \\ \cline{1-4}
- $k$-means & 75.5 - 71.2 \% & 72.8 - 70.8 \%  & 95.3 - 36.2 \%  \\ \cline{1-4}
 \end{tabular}
@@
 For this purpose we tested the accuracies obtained using individual groups of features, as well as using different combination of groups.
-Since we found the CCA to perform best of the three algorithms, we recomputed $500$ feature vectors with better accuracy and used it to compute the values of the cross-lingual article linking features.
+Since we found the CCA to perform best of the three algorithms, we recomputed it and used it to compute the values of the cross-lingual article linking features.
 Results of the evaluation are shown in Table~\ref{table:linkingEval}.
@@
 
-%\begin{table}[h]
-%\caption{Accuracy of cluster linking with features obtained from different cross-lingual similarity algorithms}
-%\label{table:linkingEvalAlgos}
-%\begin{center}
-%\begin{tabular}{|c|c|c|c|}
-%  \hline
-%  \cline{1-4}
-%  Models & CA & Precision & Recall \\ \cline{1-4}
-%  CCA & 77.6\% & 78.3 \%  & 81.1\%  \\ \cline{1-4}
-%  $k$-means & 73.9\% & 69.5\%  & 84.6\%  \\ \cline{1-4}
-%  LSI & 79.2\% & 78.9\%  & 80.6\%  \\ \cline{1-4}
-%  \hline
-%\end{tabular}
-%\end{center}
-%\end{table}
+\begin{table}[h]
+\caption{Accuracy of cluster linking with features obtained from different cross-lingual similarity algorithms}
+\label{table:linkingEvalAlgos}
+\begin{center}
+\begin{tabular}{|c|c|c|c|}
+  \hline
+  \cline{1-4}
+  Models & CA & Precision & Recall \\ \cline{1-4}
+  CCA & 77.6\% & 78.3 \%  & 81.1\%  \\ \cline{1-4}
+  $k$-means & 73.9\% & 69.5\%  & 84.6\%  \\ \cline{1-4}
+  LSI & 76.1\% & 75.5\%  & 76.9\%  \\ \cline{1-4}
+  \hline
+\end{tabular}
+\end{center}
+\end{table}
 
@@
 \subsection{Scaling}
-\emph{I think that this section should be short.}
+\emph{Andrej asks who is the rest? I think that this section should be short.}
 100 languages!

### 2015-03-24 rupnik2015.tex

--- 2015-03-24 rupnik2015.tex
@@
 \begin{itemize}
-\item{training} The building should scale to many languages and should be robust to the quality of training resources.
+\item \textbf{training} - The training (building corss-lingual models) should scale to many languages and should be robust to the quality of training resources.
 The system should be able to take advantage of comparable corpus (as opposed to parallel translation based corpora), with missing data.
-\item{operation efficiency} The similarity computation should be fast.
-\item{operation cost} The system should run on a strong shared machine server.
-\item{implementation} The system is simple to implement, with few parameters to tune.
+Supporting a new language should be simple.
+\item \textbf{operation efficiency} - The similarity computation should be fast - the system needs to handle billions of similarity computations per day.
+\item \textbf{operation cost} - The system should run on a strong shared machine server and not rely on paid services.
+\item \textbf{implementation} - The system is simple to implement, with few parameters to tune.
 \end{itemize}

### 2015-03-24 rupnik2015.tex
fix

@@
 \item \textbf{probabilistic topic models} -  There exist many variants to modelling documents in a language independent way by using probabilistic graphical models.
-The models include:  Joint Probabilistic Latent Semantic Analysis (JPLSA) \cite{platt2010translingual}, Coupled Probabilistic LSA (CPLSA) \cite{platt2010translingual}, Probabilistic Cross-Lingual LSA\cite{PCL_LSA} and Polylingual Topic Models (PLTM) which is a Bayesian version of PCL_LSA \cite{polyLDA}.
+The models include:  Joint Probabilistic Latent Semantic Analysis (JPLSA) \cite{platt2010translingual}, Coupled Probabilistic LSA (CPLSA) \cite{platt2010translingual}, Probabilistic Cross-Lingual LSA (PCLLSA)\cite{PCL_LSA} and Polylingual Topic Models (PLTM) which is a Bayesian version of PCLLSA \cite{polyLDA}.
 The methods (except for CPLSA) describe the multilingual document collections as samples from generative probabilistic models, with variations on the assumptions on the model structure.

### 2015-03-24 rupnik2015.tex
citation fixes

--- 2015-03-24 rupnik2015.tex
@@
 To cluster the articles, rather than rely on machine translations, we adopt well-established monolingual text mining tools to the multilingual setting.
-This represents the continuation of previous work \cite{nips, iti, nips2, sikdd} where we explored representations of documents which were valid over multiple languages.
+This represents the continuation of previous work \shortcite{nips,sikdd,nips2,iti} where we explored representations of documents which were valid over multiple languages.
 The representations could be interpreted as multi-lingual topics, which were then used as proxies to compute cross-lingual similarities between documents.
@@
 In this paper, we test several different algorithms for both steps for performance in an integrated system.
-In particular, we test our approach within an online system for detection of world events, called Event Registry\cite{Leban2014W, Leban2014I}.
+In particular, we test our approach within an online system for detection of world events, called Event Registry \shortcite{Leban2014W,Leban2014I}.
 Event Registry is a repository of events, where events are automatically identified by analyzing news articles.
@@
 
-We base our techniques of cross-lingual event linking on an online system for detection of world events, called Event Registry\cite{}.
+We base our techniques of cross-lingual event linking on an online system for detection of world events, called Event Registry \shortcite{Leban2014W,Leban2014I}.
 Event Registry is a repository of events, where events are automatically identified by analyzing news articles.
@@
 \item \textbf{translation} - The most obvious way to compare documents written in different languages is to use machine translation and perform monolingual similarity.
-One can use free tools such as Moses \cite{moses} or translation services, such as Google Translate\footnote{https://translate.google.com/}.
+One can use free tools such as Moses \shortcite{moses} or translation services, such as Google Translate\footnote{https://translate.google.com/}.
 There are two issues with such approaches: they solve a harder problem than needs to be solved and they are less robust to training resource quality - large sets of translated sentences are typically needed.
@@
 \item \textbf{probabilistic topic models} -  There exist many variants to modelling documents in a language independent way by using probabilistic graphical models.
-The models include:  Joint Probabilistic Latent Semantic Analysis (JPLSA) \cite{platt2010translingual}, Coupled Probabilistic LSA (CPLSA) \cite{platt2010translingual}, Probabilistic Cross-Lingual LSA\cite{PCL_LSA} and Polylingual Topic Models (PLTM) which is a Bayesian version of PCL_LSA \cite{polyLDA}.
+The models include:  Joint Probabilistic Latent Semantic Analysis (JPLSA) \shortcite{platt2010translingual}, Coupled Probabilistic LSA (CPLSA) \cite{platt2010translingual}, Probabilistic Cross-Lingual LSA (PCLLSA)\cite{PCL_LSA} and Polylingual Topic Models (PLTM) \shortcite{polyLDA} which is a Bayesian version of PCLLSA.
 The methods (except for CPLSA) describe the multilingual document collections as samples from generative probabilistic models, with variations on the assumptions on the model structure.
@@
 Furthermore, representing a new document as a mixture of topics is another potentially hard inference problem which needs to be solved.
-\item \textbf{classification problem related} - In approaches such as \cite{cotrain} and \cite{svm2k} the authors find linear embeddings of the data that support good classification results, based on a given classification problem training data.
+\item \textbf{classification problem related} - In approaches such as \cite{cotrain} and \shortcite{svm2k} the authors find linear embeddings of the data that support good classification results, based on a given classification problem training data.
 We do not consider this family of approaches, since for general language pairs classifier training sets may not be available.
-\item \textbf{matrix factorization} - Non-negative matrix factorization based \cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CLLSI\cite{cl_lsi}, Canonical Correlation Analysis (CCA) \cite{Hotelling}, Oriented Principal Component Analysis \cite{platt2010translingual}.
+\item \textbf{matrix factorization} - Non-negative matrix factorization based \cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CLLSI\shortcite{cl_lsi}, Canonical Correlation Analysis (CCA) \cite{Hotelling}, Oriented Principal Component Analysis \cite{platt2010translingual}.
 The quadratic time and space dependency of the OPCA method makes it impractical for large scale purposes.
@@
 
-The next method is Cross-Lingual Latent Semantic Indexing (CL-LSI)\cite{cl_lsi} which is a variant of LSI~\cite{lsi} for more than one language.
+The next method is Cross-Lingual Latent Semantic Indexing (CL-LSI)\cite{cl_lsi} which is a variant of LSI~\shortcite{lsi} for more than one language.
 The method is based on computing a truncated singular value decomposition of $X \approx U S V^T$.
@@
 Moreover, the Lanczos method is hard to parallelize.
-Instead we use a randomized version of the singular value decomposition (SVD) described in \cite{tropp} that can be viewed as a block Lanczos method.
+Instead we use a randomized version of the singular value decomposition (SVD) described in \shortcite{tropp} that can be viewed as a block Lanczos method.
 That enables us to use parallelization and speeds up the computation considerably.
@@
 
- Canonical Correlation Analysis (CCA)~\cite{Hotelling} is a dimensionality reduction technique similar to Principal Component Analysis (PCA)~\cite{Pearson1901On}, with an additional assumption that the data consists of feature vectors that arose from two sources (two views) that share some information (examples: bilingual document collection~\cite{mrpqr}, collection of images and captions~\cite{Hardoon_usingimage}).
+ Canonical Correlation Analysis (CCA)~\cite{Hotelling} is a dimensionality reduction technique similar to Principal Component Analysis (PCA)~\cite{Pearson1901On}, with an additional assumption that the data consists of feature vectors that arose from two sources (two views) that share some information (examples: bilingual document collection~\shortcite{mrpqr}, collection of images and captions~\shortcite{Hardoon_usingimage}).
 Instead of looking for linear combinations of features that maximize the variance (PCA) we look for a linear combination of feature vectors from the first view and a linear combination for the second view, that are maximally correlated.
@@
 
- The technique is related to the Generalization of Canonical Correlation Analysis (GCCA) by Carroll\cite{Carroll}, where an unknown group configuration variable is defined and objective is to maximize the sum of squared correlation between the group variable and the others.
+ The technique is related to the Generalization of Canonical Correlation Analysis (GCCA) by Carroll \citeyear{Carroll}, where an unknown group configuration variable is defined and objective is to maximize the sum of squared correlation between the group variable and the others.
 The problem can be reformulated as an eigenvalue problem.
@@
 The last vector contains the weighted list of Eurovoc subject domain descriptors.
-These descriptors are topics, such as \emph{air transport}, \emph{EC agreement}, \emph{competition} and \emph{pollution control} into which articles are automatically categorized~\cite{Pouliquen2006}.
+These descriptors are topics, such as \emph{air transport}, \emph{EC agreement}, \emph{competition} and \emph{pollution control} into which articles are automatically categorized~\shortcite{Pouliquen2006}.
 Similarity between clusters is then computed using a linear combination of the cosine similarities computed on the lists.

### 2015-03-24 rupnik2015.tex
abstract revision

@@
 method to enables cross-lingual document similarity based on Wikipedia, which
- enabling us to compute the similarity of any two articles regardless of language.
-Using this simliarity, we then propose an approach to find and match clusters across ;anguages which represent the same event.
-We provide an extensive evaluation of the performance of the system as a while as well as the quality of the similarity measure and linking algorithm.
+ enables us to compute the similarity of any two articles regardless of language.
+Using this similarity, we then propose an approach to find and match clusters across languages which represent the same event.
+We provide an extensive evaluation of the system as a whole, as well as an evaluation of the quality of the similarity measure and linking algorithm.
  \begin{comment}

### 2015-03-24 rupnik2015.tex
Scale part update. We need the wrap up of the story.

--- 2015-03-24 rupnik2015.tex
@@
 Firstly we verified the AMRR (average of mean reciprocal ranks over all language pairs) performance of approaches by holding out $15000$ aligned test documents and using $300000$ aligned documents
- as the training set. Figure~\ref{pc:AMRR} shows AMRR score as the function of the number of feature vectors. It is clear that CCA outperforms LSI approach and K-means lags far behind.
+ as the training set. Figure~\ref{pic:AMRR} shows AMRR score as the function of the number of feature vectors. It is clear that CCA outperforms LSI approach and K-means lags far behind.
 Furthermore we inspected how number of features influences the accuracy of cluster linking.
@@
 Moreover we tested the performance of detecting large/small cluster linking, see Table~\ref{table:linkingEvalAlgosLargeSmall}.
-As expected we obtained that it is harder to link small clusters and we need better retrieval.
-Hub CCA outperforms other methods in this sense also, probably as it has the best retrieval performance, see Figure~\ref{pic:AMRR}.
+As expected we obtained that it is harder to link small clusters and we need better retrieval accuracy in such cases.
+CCA outperforms other methods in this sense also, probably as it has the best retrieval performance, see Figure~\ref{pic:AMRR}.
+This can be also used
+to get representatives of clusters more efficiently.
 
@@
 It is fast, very robust to quality of training data, easily extendable, simple to implement and with low machine requirements.
-Ut currently runs on two Intel Xeon E5-2667 v2, 3.30GHz processor machine with 256GB of RAM.
+The similarity pipeline is the most demanding part and currently runs on two Intel Xeon E5-2667 v2, 3.30GHz processor machine with 256GB of RAM.
 That is enough to do similarity computation over 150 languages if needed.
 It currently uses Wikipedia as freely available knowledge base.
+Similarity pipeline dramastically reduces the search space when linking clusters.
 
 Currently we compute similarities over $24$ languages with tags: eng, spa, deu, zho, ita, fra, rus, swe, nld, tur, jpn, por, ara, fin, ron, kor, hrv, tam, hun, slv, pol, srp, cat, ukr but we support up to $100$ top Wikipedis languages.
-Our data stream is Newsfeed(http://newsfeed.ijs.si/) that produces $430k$ unique articles per day.
+Our data stream is Newsfeed (http://newsfeed.ijs.si/) that produces $430k$ unique articles per day.
 Our system currently computes 400k similiarites per 200ms, that means that we compute $16*10^10$ similarities per day.
 We
-store one day buffer for every language, for that we need 1.5GB of memory, we store documents as 500 dimensional vectors. For each article we compute top $10$ most similar ones in every language.
+store one day buffer for every language, for that we need 1.5GB of memory, we store documents as 500 dimensional vectors.
+We must note that time complexity of similarity computations scales linearly with dimension of the feature space and does not  depend of number of languages.
+For each article we compute top $10$  most similar ones in every other language.
 For all linear algebra matrix and vector operations we use high performance numerical libraries as BLAS, OPENBLAS and Intel MKL. This currently allows us to process more than million articles per day.
@@
 This speeds up our code by factor around $4.$ In this way we obtain some caching gains and ability to use vectorization.
-Our system is also easily extendable. If we want to add a language we only need to compute a projector matrix and proxy matrices with all other already available languages.
-
-Comment on the caching gains in our implementation.¸¸Add the conclusion if the story.
+Our system is also easily extendable. If we want to add new language we only need to compute a projector matrix and proxy matrices with all other already available languages.
+
+Add the conclusion of the story and the wrap up, Andrej needs help with this.
 

### 2015-03-24 rupnik2015.tex
introduction (ongoing)

--- 2015-03-24 rupnik2015.tex
@@
 Content on the internet is becoming increasingly multi-lingual.
-A prime example is Wikipedia - in 2001 the majority of pages were written in English, while in 2012, the percentage of English articles has dropped to 14\%.
+A prime example is Wikipedia - in 2001 the majority of pages were written in English, while in 2015, the percentage of English articles has dropped to 14\%.
 Machine translation remains relatively rudimentary - allowing people to understand simple phrases on we pages, but remain inadequate for more advanced understanding of text.
@@
 
-As input we consider a stream of articles in different languages and a list of events.
+As input we consider a high volume of articles in different languages. By using a language detector the stream is split into separate monolingual streams. 
+The documents are semantically annotated and linked across languages by using a cross-lingual similarity measure. Within each monolingual stream, an online
+clustering approach is employed, where tracked clusters correspond to events. Finally, using
+
+  and using the cross-lingual techniques  as well as processed using standard text mining techniquesBy using an online clustering approach
+
+%As input we consider a stream of articles in different languages and a list of events.
 Our goal is to assign articles to their corresponding events.
@@
 To achieve our goal, we divide the task into two parts: clustering the articles according to topic (regardless of language) and matching the clusters to events.
-
-To cluster the articles, rather than rely on machine translations, we adopt well-established monolingual text mining tools to the multilingual setting.
+%
+%To cluster the articles, rather than rely on machine translations, we adopt well-established monolingual text mining tools to the multilingual setting.
 This represents the continuation of previous work \shortcite{nips,sikdd,nips2,iti} where we explored representations of documents which were valid over multiple languages.
@@
 The underlying assumption being that articles describing the same event are similar  and will therefore be put into the same cluster.
-
-The second step is to match our computed clusters to events. Since our similarity measure and clustering
-may not be perfect, we need to find the ``best" correspondence between clusters and events.
+%
+%The second step is to match our computed clusters to events. Since our similarity measure and clustering
+%may not be perfect, we need to find the ``best" correspondence between clusters and events.
 In this paper, we test several different algorithms for both steps for performance in an integrated system.
@@
 
-   Within this system, we test our multilingual similarity measures as well as our algorithms for matching clusters to events.
+   Within this system, we test our multilingual similarity measures as well as our algorithms for matching clusters across languages.
 The result is that we can test total performance in a multilingual, real-world setting.

### 2015-03-24 rupnik2015.tex
images

--- 2015-03-24 rupnik2015.tex
@@
 The term event is vague and ambiguous, but for the practical purposes we define it as ``any significant happening that is being reported about in the media''.
-Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015 (Figure \ref{}).
+Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015 (Figure \ref{fig:event2}).
 Events such as these are covered by many articles and the question is how to find all the articles in different languages that are describing a single event.
 
+\begin{figure}
+\end{figure}
 As input we consider a stream of articles in different languages and a list of events.
@@
 Firstly we verified the AMRR (average of mean reciprocal ranks over all language pairs) performance of approaches by holding out $15000$ aligned test documents and using $300000$ aligned documents
- as the training set. Figure~\ref{pic:AMRR} shows AMRR score as the function of the number of feature vectors. It is clear that CCA outperforms LSI approach and K-means lags far behind.
+ as the training set. Figure~\ref{pc:AMRR} shows AMRR score as the function of the number of feature vectors. It is clear that CCA outperforms LSI approach and K-means lags far behind.
 Furthermore we inspected how number of features influences the accuracy of cluster linking.
@@
 Moreover we tested the performance of detecting large/small cluster linking, see Table~\ref{table:linkingEvalAlgosLargeSmall}.
-As expected we obtained that it is harder to link small clusters and we need better retrieval accuracy in such cases.
-CCA outperforms other methods in this sense also, probably as it has the best retrieval performance, see Figure~\ref{pic:AMRR}.
-This can be also used
-to get representatives of clusters more efficiently.
+As expected we obtained that it is harder to link small clusters and we need better retrieval.
+Hub CCA outperforms other methods in this sense also, probably as it has the best retrieval performance, see Figure~\ref{pic:AMRR}.
 
@@
 It is fast, very robust to quality of training data, easily extendable, simple to implement and with low machine requirements.
-The similarity pipeline is the most demanding part and currently runs on two Intel Xeon E5-2667 v2, 3.30GHz processor machine with 256GB of RAM.
+Ut currently runs on two Intel Xeon E5-2667 v2, 3.30GHz processor machine with 256GB of RAM.
 That is enough to do similarity computation over 150 languages if needed.
 It currently uses Wikipedia as freely available knowledge base.
-Similarity pipeline dramastically reduces the search space when linking clusters.
 
 Currently we compute similarities over $24$ languages with tags: eng, spa, deu, zho, ita, fra, rus, swe, nld, tur, jpn, por, ara, fin, ron, kor, hrv, tam, hun, slv, pol, srp, cat, ukr but we support up to $100$ top Wikipedis languages.
-Our data stream is Newsfeed (http://newsfeed.ijs.si/) that produces $430k$ unique articles per day.
+Our data stream is Newsfeed(http://newsfeed.ijs.si/) that produces $430k$ unique articles per day.
 Our system currently computes 400k similiarites per 200ms, that means that we compute $16*10^10$ similarities per day.
 We
-store one day buffer for every language, for that we need 1.5GB of memory, we store documents as 500 dimensional vectors.
-We must note that time complexity of similarity computations scales linearly with dimension of the feature space and does not  depend of number of languages.
-For each article we compute top $10$  most similar ones in every other language.
+store one day buffer for every language, for that we need 1.5GB of memory, we store documents as 500 dimensional vectors. For each article we compute top $10$ most similar ones in every language.
 For all linear algebra matrix and vector operations we use high performance numerical libraries as BLAS, OPENBLAS and Intel MKL. This currently allows us to process more than million articles per day.
@@
 This speeds up our code by factor around $4.$ In this way we obtain some caching gains and ability to use vectorization.
-Our system is also easily extendable. If we want to add new language we only need to compute a projector matrix and proxy matrices with all other already available languages.
-
-Add the conclusion of the story and the wrap up, Andrej needs help with this.
+Our system is also easily extendable. If we want to add a language we only need to compute a projector matrix and proxy matrices with all other already available languages.
+
+Comment on the caching gains in our implementation.¸¸Add the conclusion if the story.
 

### 2015-03-24 rupnik2015.tex

--- 2015-03-24 rupnik2015.tex
@@
 The term event is vague and ambiguous, but for the practical purposes we define it as ``any significant happening that is being reported about in the media''.
-Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015 (Figure \ref{}).
+Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015 (Figure \ref{fig:event2}).
 Events such as these are covered by many articles and the question is how to find all the articles in different languages that are describing a single event.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.75\textwidth]{events2.png}
+\caption{\label{fig:event2} Events are represented by collections of articles about an event, in this case the Malaysian airliner which was shot down over the Ukraine.}
+\end{figure}
+
+As input we consider a stream of articles in different languages and a list of events.
+Our goal is to assign articles to their corresponding events.
+A priori we do not know the coverage of the articles, that is not all the events may be covered and we do not know that all the articles necessarily fit into one of the events.
+To achieve our goal, we divide the task into two parts: clustering the articles according to topic (regardless of language) and matching the clusters to events.
 
@@
 
-The paper is organized as follows: we first provide an overview of the system as a whole (Section ), followed by details on each of the two steps described above.
-We first introduce several approaches to measures of cross-lingual similarity (Section ) and then the algorithms for matching clusters to events.
+The paper is organized as follows: we first provide an overview of the system as a whole (Section ~\ref{sec:pipeline}), followed by details on each of the two steps described above.
+We first introduce several approaches to measures of cross-lingual similarity (Section~\ref{sec:crosslingual}) and then the algorithms for matching clusters iin different languages(Section~\ref{sec:linking}).
 To improve readability, we describe relevant related work for each module in its corresponding Section.
@@
 
-\section{Pipeline}
+\section{Pipeline}\label{sec:pipeline}
+
+\begin{figure}[tb]
+\centering
+\includegraphics[width=0.75\textwidth]{pipeline1}
+\caption{\label{fig:pipeline} Our pipeline - from the Event Registry we obtain a stream of multi-lingual documents and monolingual clusters.
+Our contribution is shown on the right, where using cross-lingual similarity, we link the clusters across languages which correspond to the same event.}
+\end{figure}
 
@@
 
-Our approach is based on first defining a similarity function between articles in different languages (Section~\ref{}) and then we present several efficient approaches to linking the clusters based on features (Section~\ref{}).
-First, however we describe the pre-processing on the input articles which we then take as input.
-
-The system is designed to handle streams and so we describe how a document which enters the stream is processed.   A new document is first tokenized, stop words are removed and words are stemmed.
- What remains is represented in a vector-space model and normalized using $TF-IDF$ (see Section~\ref{} for the definition).
+Our approach is based on first defining a similarity function between articles in different languages (Section~\ref{sec:crosslingual}) and then we present several efficient approaches to linking the clusters based on features (Section~\ref{sec:linking}).
+The pipeline is shown in Figure~\ref{fig:pipeline} First, however we describe the pre-processing on the input articles which we then take as input.
+
+The system is designed to handle streams and so we describe how a document which enters the stream is processed.
+The sequence of operations is shown in  Figure~\ref{fig:erpipeline}.
+A new document is first tokenized, stop words are removed and words are stemmed.
+ What remains is represented in a vector-space model and normalized using $TF-IDF$ (see Section~\ref{sec:tfidf} for the definition).
 Since each article is tagged with its language, cosine similiarity  is used to find the modst similar existing cluster, by comparing it to the centroid vector of each cluster.
@@
 
- Further details about the system can be found in \cite{}, while our full pipeline is shown in Figure~\ref{}.
+ Further details about the system can be found in \cite{}.
+
+\begin{figure}[tbp]
+\centering
+\includegraphics[width=\textwidth]{ev_pipeline}
+\caption{\label{fig:erpipeline}  The Event Registry pipeline, after obtaining the raw stream of multi-lingual news articles, it first cleans and processes articles individually.
+It then aggregates and clusters the articles into similar events (within each language) and finally provides an interface allowing us to query for articles.}
+\end{figure}
 
@@
 \section{Cross-lingual Document Similarity}
-
+\label{sec:crosslingual}
 Document similarity is an important component in techniques from text mining and natural language processing.
@@
 \subsection{Problem definition}
-
+\label{sec:tfidf}
 \textbf{Document representation.}
@@
 
-\subsection{Canonical Correlation Analysis}
+\subsection{Canonical Correlation Analysis}\label{sec:CCA}
  The final approach is a statistical technique to analyze data from two sources - after which we will describe our main contribution to computing cross-lingual similarities.
@@
 Same events are covered by many articles and the question we address is how to find all the articles in different languages that are describing a single event.
-We base our evaluation on an online system for detection of world events, called Event Registry.
+We base our evaluation on an online system for detection of world events, Event Registry.
 We do not address the problem of detection of events in this paper but rather consider the problem that given a series of events, how may we best ``tag" or match articles in different languages to these events.
@@
 
+\begin{figure}[tb]
+\centering
+\includegraphics[width=0.6\textwidth]{clusters}
+\caption{\label{fig:clusters}  Given clusters of articles in different languages, the goal of  linking is to match clusters which have similar content across languages.
+The challenge in linking is to overcome the errors in the clustering and differing content as well as make the approach scalable.
+Therefore, we first identify candidate clusters.
+}
+\end{figure}
+
 Furthermore, the matching of clusters is a \emph{generalized matching}.
@@
 This excludes the use of standard weighted bipartite matching type of algortihms for this problem.
+An example is shown in Figure~\ref{fig:clusters}, where a cluster may contain articles which are closely matched with many clusters in a different language.
 
@@
 We define \emph{equivalent clusters} as  pairs $(c_i,c_j)$ such that both clusters in the pair describe the same event in languages $C_{l_a}$ and $C_{l_b}$ respectively.
+For example in Figure~\ref{fig:clusters}
 
@@
 First, for a cluster $c_i$, we efficiently identify a small set of candidate clusters and then  find the clusters from among the candidates which are equivalent to $c_i$.
+An example is  shown in  Figure~\ref{fig:clusters}.
 
@@
 The algorithm begins by individually inspecting each article $a_i$ in the cluster $c_i$.
-Using CCA  (see Section~\ref{}),  it then identifies the 10 most similar articles to $a_i$ in each language $l \in L$.
+Using CCA  (see Section~\ref{sec:CCA}),  it then identifies the 10 most similar articles to $a_i$ in each language $l \in L$.
 For each  similar article $a_j$, we identify its corresponding  cluster $c_j$ and add it to the set of candidates.

### 2015-03-24 rupnik2015.tex
update

--- 2015-03-24 rupnik2015.tex
@@
 As input we consider a high volume of articles in different languages. By using a language detector the stream is split into separate monolingual streams. 
-The documents are semantically annotated and linked across languages by using a cross-lingual similarity measure. Within each monolingual stream, an online
-clustering approach is employed, where tracked clusters correspond to events. Finally, using
-
-  and using the cross-lingual techniques  as well as processed using standard text mining techniquesBy using an online clustering approach
+Within each monolingual stream, an online clustering approach is employed, where tracked clusters correspond to our definition of events - this is based on the Event Registry system \shortcite{Leban2014W,Leban2014I}.
+Our main goal is to connect such clusters (representations of events) across languages, that is detect that a set of documents in language 1 reports on the same event as a set of documents in language 2.
+
+Our approach to combines two ingredients: a cross-lingual document similarity measure, which can be interpreted as a language independent topic model, and
+semantic annotation of documents, which enables an alternative way to comparing documents by comparing the lists of detected semantic entities.
+
+The first approach represents a continuation of previous work \shortcite{nips,sikdd,nips2,iti} where we explored representations of documents which were valid over multiple languages.
+The representations could be interpreted as multi-lingual topics, which were then used as proxies to compute cross-lingual similarities between documents.
+To learn the representations, we use Wikipedia as a training corpus.
+Significantly, we do not only consider the major or \emph{hub} languages such as English, German, French, etc.
+which have significant overlap in article coverage, but also smaller languages (in terms of number of Wikipedia articles) such as Slovenian and Hindi, which may have a negligible overlap in article coverage.
+We can then define a similarity between any two articles regardless of language,  which allows us to cluster the articles according to topic.
+The underlying assumption being that articles describing the same event are similar  and will therefore be put into the same cluster.
 
@@
 
-Event Registry works by collecting and analyzing news articles that are published by news outlets in different languages all over the world.
+%
+%Event Registry works by collecting and analyzing news articles that are published by news outlets in different languages all over the world.
 Collected articles are first semantically annotated by identifying in them mentions of relevant concepts -- either entities or important keywords.
 The disambiguation and entity linking of the concepts is done using Wikipedia as the main knowledge base.
-   The key step in Event Registry is then to apply an online clustering algorithm on the articles in order to identify groups of articles that are discussing the same event.
-
-   Within this system, we test our multilingual similarity measures as well as our algorithms for matching clusters across languages.
+%   The key step in Event Registry is then to apply an online clustering algorithm on the articles in order to identify groups of articles that are discussing the same event.
+%
+%   Within this system, we test our multilingual similarity measures as well as our algorithms for matching clusters across languages.
 The result is that we can test total performance in a multilingual, real-world setting.

### 2015-03-24 rupnik2015.tex
introduction update

@@
 A priori we do not know the coverage of the articles, that is not all the events may be covered and we do not know that all the articles necessarily fit into one of the events.
-To achieve our goal, we divide the task into two parts: clustering the articles according to topic (regardless of language) and matching the clusters to events.
+The task is divided into two parts: detecting events within each language and then linking events across languages.
+In this paper we address the second step.
 
@@
 
-Our approach to combines two ingredients: a cross-lingual document similarity measure, which can be interpreted as a language independent topic model, and
-semantic annotation of documents, which enables an alternative way to comparing documents by comparing the lists of detected semantic entities.
+Our approach combines two ingredients: a cross-lingual document similarity measure, which can be interpreted as a language independent topic model, and
+semantic annotation of documents, which enables an alternative way to comparing documents.
 
@@
 The underlying assumption being that articles describing the same event are similar  and will therefore be put into the same cluster.
+
+Based on the similarity function we propose a novel algorithm for linking events/clusters across languages.
+The approach is based on learning a classification model from labelled data based on several sets of features.
+In addition to these features, cross-lingual similarity is also used to quickly identify a small list of potential linking candidates for each cluster.
+This greatly increases the scalability of the system.
 

### 2015-03-24 rupnik2015.tex
- updated text in Cross-lingual Event Linking

--- 2015-03-24 rupnik2015.tex
@@
 As input we consider a high volume of articles in different languages. By using a language detector the stream is split into separate monolingual streams. 
-Within each monolingual stream, an online clustering approach is employed, where tracked clusters correspond to our definition of events - this is based on the Event Registry system \shortcite{Leban2014W,Leban2014I}.
-Our main goal is to connect such clusters (representations of events) across languages, that is detect that a set of documents in language 1 reports on the same event as a set of documents in language 2.
-
-Our approach to combines two ingredients: a cross-lingual document similarity measure, which can be interpreted as a language independent topic model, and
-semantic annotation of documents, which enables an alternative way to comparing documents by comparing the lists of detected semantic entities.
-
-The first approach represents a continuation of previous work \shortcite{nips,sikdd,nips2,iti} where we explored representations of documents which were valid over multiple languages.
-The representations could be interpreted as multi-lingual topics, which were then used as proxies to compute cross-lingual similarities between documents.
-To learn the representations, we use Wikipedia as a training corpus.
-Significantly, we do not only consider the major or \emph{hub} languages such as English, German, French, etc.
-which have significant overlap in article coverage, but also smaller languages (in terms of number of Wikipedia articles) such as Slovenian and Hindi, which may have a negligible overlap in article coverage.
-We can then define a similarity between any two articles regardless of language,  which allows us to cluster the articles according to topic.
-The underlying assumption being that articles describing the same event are similar  and will therefore be put into the same cluster.
+The documents are semantically annotated and linked across languages by using a cross-lingual similarity measure. Within each monolingual stream, an online
+clustering approach is employed, where tracked clusters correspond to events. Finally, using
+
+  and using the cross-lingual techniques  as well as processed using standard text mining techniquesBy using an online clustering approach
 
@@
 
-%
-%Event Registry works by collecting and analyzing news articles that are published by news outlets in different languages all over the world.
+Event Registry works by collecting and analyzing news articles that are published by news outlets in different languages all over the world.
 Collected articles are first semantically annotated by identifying in them mentions of relevant concepts -- either entities or important keywords.
 The disambiguation and entity linking of the concepts is done using Wikipedia as the main knowledge base.
-%   The key step in Event Registry is then to apply an online clustering algorithm on the articles in order to identify groups of articles that are discussing the same event.
-%
-%   Within this system, we test our multilingual similarity measures as well as our algorithms for matching clusters across languages.
+   The key step in Event Registry is then to apply an online clustering algorithm on the articles in order to identify groups of articles that are discussing the same event.
+
+   Within this system, we test our multilingual similarity measures as well as our algorithms for matching clusters across languages.
 The result is that we can test total performance in a multilingual, real-world setting.
@@
 
-The main application on which we test the above similiarity is cross-lingual event linking.
+The main application on which we test the above similarity is cross-lingual event linking.
 In online media streams -- particularly new articles -- there is often duplication of reporting, different viewpoints or opinions, all centering around a single event.
@@
 
-The problem in the second phase of the pipeline is to identify the clusters describing the same events.
+The problem of cross-lingual event linking is to identify the clusters describing the same events.
 Each article $a_i$ is written in a language $l$, where $\ell \in L = \{\ell_1,\ell_2,...,\ell_k\}$.
@@
 Furthermore, the matching of clusters is a \emph{generalized matching}.
-We cannot assume that there is only one cluster per language per event, nor can we assume complete coverage - that there exists at least one cluster per event in every language.
+We cannot assume that there is only one cluster per language per event, nor can we assume complete coverage -- that there exists at least one cluster per event in every language.
 This implies that we cannot make any assumptions on the matching, e.g.
@@
 Identification of the equivalent clusters cannot be done exhaustively -- for a cluster $c_i$ we wish to avoid testing all other clusters for all the other languages.
-Performing exhaustive comparison for each cluster $c_i$ would result in $O(N^2)$ tests, where $N$ is the number of all clusters (over all languages), which is not feasible for when the number of clusters is on the order of tens of thousands.
+Performing exhaustive comparison for each cluster $c_i$ would result in $O(N^2)$ tests, where $N$ is the number of all clusters (over all languages), which is not feasible when the number of clusters is on the order of tens of thousands.
 
@@
 
-Although there are a number of services that aggregate news by identifying clusters of similar articles, there are very few service that provide linking of clusters over different languages.
+Although there are a number of services that aggregate news by identifying clusters of similar articles, there are almost no services that provide linking of clusters over different languages.
 Google News as well as Yahoo!
 News are able to identify clusters of articles about same event, but they offer no linking of clusters across languages.
-The only service that provides cross-lingual cluster linking is the European Media Monitor (EMM)~\cite{Steinberger2008}.
+The only service that we found, which provides cross-lingual cluster linking, is the European Media Monitor (EMM)~\cite{Steinberger2008}.
 EMM clusters articles in 60 languages and then tries to determine which clusters of articles in different languages describe the same event.
@@
 Similarity between clusters is then computed using a linear combination of the cosine similarities computed on the lists.
+If the similarity is above the threshold, the clusters are linked.
 
@@
 In order to identify clusters $c_j$ that are equivalent to cluster $c_i$, we have developed a two-stage algoirhtm.
-First, for a cluster $c_i$, we efficiently identify a small set of candidate clusters and then  find the clusters from among the candidates which are equivalent to $c_i$.
+For a cluster $c_i$, we first efficiently identify a small set of candidate clusters and then find those clusters among the candidates, which are equivalent to $c_i$.
 
@@
 The algorithm begins by individually inspecting each article $a_i$ in the cluster $c_i$.
-Using CCA  (see Section~\ref{}),  it then identifies the 10 most similar articles to $a_i$ in each language $l \in L$.
+Using a cross-lingual document similarity measure (see Section~\ref{}) it identifies the 10 most similar articles to $a_i$ in each language $l \in L$.
 For each  similar article $a_j$, we identify its corresponding  cluster $c_j$ and add it to the set of candidates.
-\todo{we should say a few words why this is more efficient - since now we are comparing each article to all other articles - while we said it is not feadible for clusters}.
-While the CCA simliarity is approximate, our  assumption  is that articles in different languages describing the same event, will generally have a higher similarity than articles about different events.
+The set of candidate clusters obtained in this way is several orders of magnitude smaller than the number of all clusters.
+Although computed document similarities are approximate, our  assumption  is that articles in different languages describing the same event, will generally have a higher similarity than articles about different events.
 While this assumption does not always hold, redundancy in the data mitigates these false positives.

### 2015-03-24 rupnik2015.tex
Updated abstract

--- 2015-03-24 rupnik2015.tex
@@
 \begin{abstract}
-Today, we follow news  which is distributed globally and reported on in many different languages.
-When a significant event occurs, it is often reported on in many different sources.
-In this work, our goal is to consider the problem tracking events in a large multilingual stream.
-In this paper, we consider a particular aspect of this problem, namely how to link collections of articles in different languages which refer to the same event.
-Given a multi-lingual stream,  articles are first clustered according in each language we consider separately.
-Given these clusters, we first propose a
-method to enables cross-lingual document similarity based on Wikipedia, which
- enables us to compute the similarity of any two articles regardless of language.
-Using this similarity, we then propose an approach to find and match clusters across languages which represent the same event.
-We provide an extensive evaluation of the system as a whole, as well as an evaluation of the quality of the similarity measure and linking algorithm.
+Today, we follow news which is distributed globally.
+Significant events are reported by different sources and in different languages.
+In this work, we address the problem of tracking and events in a large multilingual stream.
+We consider a particular aspect of this problem, namely how to link collections of articles in different languages which refer to the same event.
+Given a multi-lingual stream and clusters of articles from each language, we first propose a method for cross-lingual document similarity based on Wikipedia, which enables us to compute the similarity of any two articles regardless of language.
+The proposed method is can scale to 100 languages and can match articles from languages with little or no direct overlap in the training data.
+Using this similarity, we then propose an approach to link clusters of articles across languages which represent the same event.
+We provide an extensive evaluation of the system as a whole, as well as an evaluation of the quality and robustness of the similarity measure and the linking algorithm.
 \begin{comment}

### 2015-03-24 rupnik2015.tex

--- 2015-03-24 rupnik2015.tex
@@
 
-The main application on which we test the above similiarity is cross-lingual event linking.
+The main application on which we test the above similarity is cross-lingual event linking.
 In online media streams -- particularly new articles -- there is often duplication of reporting, different viewpoints or opinions, all centering around a single event.
 Same events are covered by many articles and the question we address is how to find all the articles in different languages that are describing a single event.
-We base our evaluation on an online system for detection of world events, Event Registry.
+We base our evaluation on an online system for detection of world events, called Event Registry.
 We do not address the problem of detection of events in this paper but rather consider the problem that given a series of events, how may we best ``tag" or match articles in different languages to these events.
@@
 We define \emph{equivalent clusters} as  pairs $(c_i,c_j)$ such that both clusters in the pair describe the same event in languages $C_{l_a}$ and $C_{l_b}$ respectively.
-For example in Figure~\ref{fig:clusters}
 
@@
 In order to identify clusters $c_j$ that are equivalent to cluster $c_i$, we have developed a two-stage algoirhtm.
-First, for a cluster $c_i$, we efficiently identify a small set of candidate clusters and then  find the clusters from among the candidates which are equivalent to $c_i$.
+For a cluster $c_i$, we first efficiently identify a small set of candidate clusters and then find those clusters among the candidates, which are equivalent to $c_i$.
 An example is  shown in  Figure~\ref{fig:clusters}.
@@
 The algorithm begins by individually inspecting each article $a_i$ in the cluster $c_i$.
-Using CCA  (see Section~\ref{sec:CCA}),  it then identifies the 10 most similar articles to $a_i$ in each language $l \in L$.
+Using a cross-lingual document similarity measure (see Section~\ref{}), it identifies the 10 most similar articles to $a_i$ in each language $l \in L$.
 For each  similar article $a_j$, we identify its corresponding  cluster $c_j$ and add it to the set of candidates.
-\todo{we should say a few words why this is more efficient - since now we are comparing each article to all other articles - while we said it is not feadible for clusters}.
-While the CCA simliarity is approximate, our  assumption  is that articles in different languages describing the same event, will generally have a higher similarity than articles about different events.
+The set of candidate clusters obtained in this way is several orders of magnitude smaller than the number of all clusters.
+Although computed document similarities are approximate, our  assumption  is that articles in different languages describing the same event, will generally have a higher similarity than articles about different events.
 While this assumption does not always hold, redundancy in the data mitigates these false positives.

### 2015-03-24 rupnik2015.tex
Few fixes in introduction.

--- 2015-03-24 rupnik2015.tex
@@
 A prime example is Wikipedia - in 2001 the majority of pages were written in English, while in 2015, the percentage of English articles has dropped to 14\%.
-Machine translation remains relatively rudimentary - allowing people to understand simple phrases on we pages, but remain inadequate for more advanced understanding of text.
+Machine translation remains relatively rudimentary - allowing people to understand simple phrases on web pages, but remain inadequate for more advanced understanding of text.
 At the same time, online news has begun to dominate reporting of current events.
@@
 
-As input we consider a high volume of articles in different languages. By using a language detector the stream is split into separate monolingual streams.
+As input we consider a high volume of articles in different languages.
+By using a language detector the stream is split into separate monolingual streams.
 Within each monolingual stream, an online clustering approach is employed, where tracked clusters correspond to our definition of events - this is based on the Event Registry system \shortcite{Leban2014W,Leban2014I}.
-Our main goal is to connect such clusters (representations of events) across languages, that is detect that a set of documents in language 1 reports on the same event as a set of documents in language 2.
-
-Our approach combines two ingredients: a cross-lingual document similarity measure, which can be interpreted as a language independent topic model, and
-semantic annotation of documents, which enables an alternative way to comparing documents.
+Our main goal in this paper is to connect such clusters (representations of events) across languages, that is detect that a set of articles in language $A$ reports on the same event as a set of articles in language $B$.
+
+Our approach to linking clusters across languages combines two ingredients: a cross-lingual document similarity measure, which can be interpreted as a language independent topic model, and semantic annotation of documents, which enables an alternative way to comparing documents.
 
@@
 The paper is organized as follows: we first provide an overview of the system as a whole (Section ~\ref{sec:pipeline}), followed by details on each of the two steps described above.
-We first introduce several approaches to measures of cross-lingual similarity (Section~\ref{sec:crosslingual}) and then the algorithms for matching clusters iin different languages(Section~\ref{sec:linking}).
+We first introduce several approaches for measuring cross-lingual similarity (Section~\ref{sec:crosslingual}) and then the algorithms for matching clusters in different languages (Section~\ref{sec:linking}).
 To improve readability, we describe relevant related work for each module in its corresponding Section.
-Finally, we present and interpret the experimental results and finally give conclusions and describe promising future directions.
+Finally, we present and interpret the experimental results, give conclusions and describe promising future directions.
 

### 2015-03-24 rupnik2015.tex
- small fix

--- 2015-03-24 rupnik2015.tex
@@
 The algorithm begins by individually inspecting each article $a_i$ in the cluster $c_i$.
-Using a cross-lingual document similarity measure (see Section~\ref{}), it identifies the 10 most similar articles to $a_i$ in each language $l \in L$.
+Using a chosen method for computing cross-lingual document similarity (see Section~\ref{sec:models}), it identifies the 10 most similar articles to $a_i$ in each language $l \in L$.
 For each  similar article $a_j$, we identify its corresponding  cluster $c_j$ and add it to the set of candidates.

### 2015-03-24 rupnik2015.tex
discussion and future work

--- 2015-03-24 rupnik2015.tex
@@
 \begin{abstract}
-Today, we follow news which is distributed globally.
-Significant events are reported by different sources and in different languages.
-In this work, we address the problem of tracking and events in a large multilingual stream.
-We consider a particular aspect of this problem, namely how to link collections of articles in different languages which refer to the same event.
-Given a multi-lingual stream and clusters of articles from each language, we first propose a method for cross-lingual document similarity based on Wikipedia, which enables us to compute the similarity of any two articles regardless of language.
-The proposed method is can scale to 100 languages and can match articles from languages with little or no direct overlap in the training data.
-Using this similarity, we then propose an approach to link clusters of articles across languages which represent the same event.
-We provide an extensive evaluation of the system as a whole, as well as an evaluation of the quality and robustness of the similarity measure and the linking algorithm.
+Today, we follow news  which is distributed globally and reported on in many different languages.
+When a significant event occurs, it is often reported on in many different sources.
+In this work, our goal is to consider the problem tracking events in a large multilingual stream.
+In this paper, we consider a particular aspect of this problem, namely how to link collections of articles in different languages which refer to the same event.
+Given a multi-lingual stream,  articles are first clustered according in each language we consider separately.
+Given these clusters, we first propose a
+method to enables cross-lingual document similarity based on Wikipedia, which
+ enables us to compute the similarity of any two articles regardless of language.
+Using this similarity, we then propose an approach to find and match clusters across languages which represent the same event.
+We provide an extensive evaluation of the system as a whole, as well as an evaluation of the quality of the similarity measure and linking algorithm.
  \begin{comment}
@@
 
-The main application on which we test the above similarity is cross-lingual event linking.
+The main application on which we test the above similiarity is cross-lingual event linking.
 In online media streams -- particularly new articles -- there is often duplication of reporting, different viewpoints or opinions, all centering around a single event.
 Same events are covered by many articles and the question we address is how to find all the articles in different languages that are describing a single event.
-We base our evaluation on an online system for detection of world events, called Event Registry.
+We base our evaluation on an online system for detection of world events, Event Registry.
 We do not address the problem of detection of events in this paper but rather consider the problem that given a series of events, how may we best ``tag" or match articles in different languages to these events.
@@
 
-The problem of cross-lingual event linking is to identify the clusters describing the same events.
+The problem in the second phase of the pipeline is to identify the clusters describing the same events.
 Each article $a_i$ is written in a language $l$, where $\ell \in L = \{\ell_1,\ell_2,...,\ell_k\}$.
@@
 Furthermore, the matching of clusters is a \emph{generalized matching}.
-We cannot assume that there is only one cluster per language per event, nor can we assume complete coverage -- that there exists at least one cluster per event in every language.
+We cannot assume that there is only one cluster per language per event, nor can we assume complete coverage - that there exists at least one cluster per event in every language.
 This implies that we cannot make any assumptions on the matching, e.g.
@@
 Identification of the equivalent clusters cannot be done exhaustively -- for a cluster $c_i$ we wish to avoid testing all other clusters for all the other languages.
-Performing exhaustive comparison for each cluster $c_i$ would result in $O(N^2)$ tests, where $N$ is the number of all clusters (over all languages), which is not feasible when the number of clusters is on the order of tens of thousands.
+Performing exhaustive comparison for each cluster $c_i$ would result in $O(N^2)$ tests, where $N$ is the number of all clusters (over all languages), which is not feasible for when the number of clusters is on the order of tens of thousands.
 
@@
 
-Although there are a number of services that aggregate news by identifying clusters of similar articles, there are almost no services that provide linking of clusters over different languages.
+Although there are a number of services that aggregate news by identifying clusters of similar articles, there are very few service that provide linking of clusters over different languages.
 Google News as well as Yahoo!
 News are able to identify clusters of articles about same event, but they offer no linking of clusters across languages.
-The only service that we found, which provides cross-lingual cluster linking, is the European Media Monitor (EMM)~\cite{Steinberger2008}.
+The only service that provides cross-lingual cluster linking is the European Media Monitor (EMM)~\cite{Steinberger2008}.
 EMM clusters articles in 60 languages and then tries to determine which clusters of articles in different languages describe the same event.
@@
 Similarity between clusters is then computed using a linear combination of the cosine similarities computed on the lists.
-If the similarity is above the threshold, the clusters are linked.
 
@@
 In order to identify clusters $c_j$ that are equivalent to cluster $c_i$, we have developed a two-stage algoirhtm.
-For a cluster $c_i$, we first efficiently identify a small set of candidate clusters and then find those clusters among the candidates, which are equivalent to $c_i$.
+First, for a cluster $c_i$, we efficiently identify a small set of candidate clusters and then  find the clusters from among the candidates which are equivalent to $c_i$.
 An example is  shown in  Figure~\ref{fig:clusters}.
@@
 The algorithm begins by individually inspecting each article $a_i$ in the cluster $c_i$.
-Using a chosen method for computing cross-lingual document similarity (see Section~\ref{sec:models}), it identifies the 10 most similar articles to $a_i$ in each language $l \in L$.
+Using CCA  (see Section~\ref{sec:CCA}),  it then identifies the 10 most similar articles to $a_i$ in each language $l \in L$.
 For each  similar article $a_j$, we identify its corresponding  cluster $c_j$ and add it to the set of candidates.
-The set of candidate clusters obtained in this way is several orders of magnitude smaller than the number of all clusters.
-Although computed document similarities are approximate, our  assumption  is that articles in different languages describing the same event, will generally have a higher similarity than articles about different events.
+\todo{we should say a few words why this is more efficient - since now we are comparing each article to all other articles - while we said it is not feadible for clusters}.
+While the CCA simliarity is approximate, our  assumption  is that articles in different languages describing the same event, will generally have a higher similarity than articles about different events.
 While this assumption does not always hold, redundancy in the data mitigates these false positives.
@@
 
-\section{Discussion}
+\section{Discussion and Future Work}
+
+We now discuss some aspects of the methods and the system as a whole.
+\begin{itemize}
+\item \textbf{LSI and hub CCA} - We first comment on the empirical validation of LSI and hub CCA.
+On the task of mate retrieval we observe that refining the LSI based projections with hub CCA leads to improved retrieval precision, but the methods perform comparably on the task of event linking.
+Further inspection showed that the CCA based approach reached a higher precision on smaller clusters.
+The interpretation is that the linking features are highly aggregated for large clusters, which compensates the lower per-document percision of LSI.
+Another possible reason is, that the advantage that we show on Wikipedia is lost on the news domain - which indicates that domain adaptation should be explored.
+
+\item \textbf{Features} - The experiments show that the hub CCA based features present good baseline, which can greatly benefit from additional semantic extraction based features.
+Even though the CCA based features do not greatly improve the system performance when the semantic features are used, there are two main benefits in the approach: the linking process can be sped up (smaller candidate cluster sets), and in the case when semantic extraction is not possible.
+
+\item \textbf{Future work} - Currently the system is loosely coupled - the language component is built independently of the rest of the system, in particular the linking component.
+It is possible that better embeddings can be obtained by methods that jointly optimize a classification task and the embedding.
+
+    Another point of interest is to evaluate the system on languages with scarce linguistic resources, where semantic annotation might not be available.
+For this reason, the labelled dataset of linked clusters should to be extended first.
+The mate retrieval evaluation showed that even for language pairs with no training set overlap, the hub cca recovers some signal.
+
+\end{itemize}
 

### 2015-03-24 rupnik2015.tex

--- 2015-03-24 rupnik2015.tex
@@
 The algorithm begins by individually inspecting each article $a_i$ in the cluster $c_i$.
-Using a chosen method for computing cross-lingual document similarity (see Section~\ref{sec:models}), it identifies the 10 most similar articles to $a_i$ in each language $l \in L$.
+Using a cross-lingual document similarity measure (see Section~\ref{}), it identifies the 10 most similar articles to $a_i$ in each language $l \in L$.
 For each  similar article $a_j$, we identify its corresponding  cluster $c_j$ and add it to the set of candidates.

### 2015-03-24 rupnik2015.tex
related work changes

--- 2015-03-24 rupnik2015.tex
@@
 If the similarity is above the threshold, the clusters are linked.
+Compared to EMM, our approach uses document similarities to obtain a small set of potentially equivalent clusters.
+Additionally, we do not decide if two clusters are equivalent based on a hand-set threshold on a similarity value -- instead we use a classification model that uses a larger set of features related to the tested clusters.
 

### 2015-03-24 rupnik2015.tex
Scale polished. Jan preberi in dodaj odvzemi po želji.

--- 2015-03-24 rupnik2015.tex
@@
 Content on the internet is becoming increasingly multi-lingual.
-A prime example is Wikipedia - in 2001 the majority of pages were written in English, while in 2015, the percentage of English articles has dropped to 14\%.
-Machine translation remains relatively rudimentary - allowing people to understand simple phrases on web pages, but remain inadequate for more advanced understanding of text.
+A prime example is Wikipedia - in 2001 the majority of pages were written in English, while in 2012, the percentage of English articles has dropped to 14\%.
+Machine translation remains relatively rudimentary - allowing people to understand simple phrases on we pages, but remain inadequate for more advanced understanding of text.
 At the same time, online news has begun to dominate reporting of current events.
@@
 The term event is vague and ambiguous, but for the practical purposes we define it as ``any significant happening that is being reported about in the media''.
-Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015 (Figure \ref{fig:event2}).
+Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015 (Figure \ref{}).
 Events such as these are covered by many articles and the question is how to find all the articles in different languages that are describing a single event.
-
-\begin{figure}
-\centering
-\includegraphics[width=0.75\textwidth]{events2.png}
-\caption{\label{fig:event2} Events are represented by collections of articles about an event, in this case the Malaysian airliner which was shot down over the Ukraine.}
-\end{figure}
 
@@
 A priori we do not know the coverage of the articles, that is not all the events may be covered and we do not know that all the articles necessarily fit into one of the events.
-The task is divided into two parts: detecting events within each language and then linking events across languages.
-In this paper we address the second step.
-
-As input we consider a high volume of articles in different languages.
-By using a language detector the stream is split into separate monolingual streams.
-Within each monolingual stream, an online clustering approach is employed, where tracked clusters correspond to our definition of events - this is based on the Event Registry system \shortcite{Leban2014W,Leban2014I}.
-Our main goal in this paper is to connect such clusters (representations of events) across languages, that is detect that a set of articles in language $A$ reports on the same event as a set of articles in language $B$.
-
-Our approach to linking clusters across languages combines two ingredients: a cross-lingual document similarity measure, which can be interpreted as a language independent topic model, and semantic annotation of documents, which enables an alternative way to comparing documents.
-
-The first approach represents a continuation of previous work \shortcite{nips,sikdd,nips2,iti} where we explored representations of documents which were valid over multiple languages.
-The representations could be interpreted as multi-lingual topics, which were then used as proxies to compute cross-lingual similarities between documents.
-To learn the representations, we use Wikipedia as a training corpus.
-Significantly, we do not only consider the major or \emph{hub} languages such as English, German, French, etc.
-which have significant overlap in article coverage, but also smaller languages (in terms of number of Wikipedia articles) such as Slovenian and Hindi, which may have a negligible overlap in article coverage.
-We can then define a similarity between any two articles regardless of language,  which allows us to cluster the articles according to topic.
-The underlying assumption being that articles describing the same event are similar  and will therefore be put into the same cluster.
-
-Based on the similarity function we propose a novel algorithm for linking events/clusters across languages.
-The approach is based on learning a classification model from labelled data based on several sets of features.
-In addition to these features, cross-lingual similarity is also used to quickly identify a small list of potential linking candidates for each cluster.
-This greatly increases the scalability of the system.
-
-%As input we consider a stream of articles in different languages and a list of events.
-Our goal is to assign articles to their corresponding events.
-A priori we do not know the coverage of the articles, that is not all the events may be covered and we do not know that all the articles necessarily fit into one of the events.
 To achieve our goal, we divide the task into two parts: clustering the articles according to topic (regardless of language) and matching the clusters to events.
-%
-%To cluster the articles, rather than rely on machine translations, we adopt well-established monolingual text mining tools to the multilingual setting.
+
+To cluster the articles, rather than rely on machine translations, we adopt well-established monolingual text mining tools to the multilingual setting.
 This represents the continuation of previous work \shortcite{nips,sikdd,nips2,iti} where we explored representations of documents which were valid over multiple languages.
@@
 The underlying assumption being that articles describing the same event are similar  and will therefore be put into the same cluster.
-%
-%The second step is to match our computed clusters to events. Since our similarity measure and clustering
-%may not be perfect, we need to find the ``best" correspondence between clusters and events.
+
+The second step is to match our computed clusters to events. Since our similarity measure and clustering
+may not be perfect, we need to find the ``best" correspondence between clusters and events.
 In this paper, we test several different algorithms for both steps for performance in an integrated system.
@@
 
-%
-%Event Registry works by collecting and analyzing news articles that are published by news outlets in different languages all over the world.
+Event Registry works by collecting and analyzing news articles that are published by news outlets in different languages all over the world.
 Collected articles are first semantically annotated by identifying in them mentions of relevant concepts -- either entities or important keywords.
 The disambiguation and entity linking of the concepts is done using Wikipedia as the main knowledge base.
-%   The key step in Event Registry is then to apply an online clustering algorithm on the articles in order to identify groups of articles that are discussing the same event.
-%
-%   Within this system, we test our multilingual similarity measures as well as our algorithms for matching clusters across languages.
+   The key step in Event Registry is then to apply an online clustering algorithm on the articles in order to identify groups of articles that are discussing the same event.
+
+   Within this system, we test our multilingual similarity measures as well as our algorithms for matching clusters to events.
 The result is that we can test total performance in a multilingual, real-world setting.
 
-The paper is organized as follows: we first provide an overview of the system as a whole (Section ~\ref{sec:pipeline}), followed by details on each of the two steps described above.
-We first introduce several approaches for measuring cross-lingual similarity (Section~\ref{sec:crosslingual}) and then the algorithms for matching clusters in different languages (Section~\ref{sec:linking}).
+The paper is organized as follows: we first provide an overview of the system as a whole (Section ), followed by details on each of the two steps described above.
+We first introduce several approaches to measures of cross-lingual similarity (Section ) and then the algorithms for matching clusters to events.
 To improve readability, we describe relevant related work for each module in its corresponding Section.
-Finally, we present and interpret the experimental results, give conclusions and describe promising future directions.
-
-\section{Pipeline}\label{sec:pipeline}
-
-\begin{figure}[tb]
-\centering
-\includegraphics[width=0.75\textwidth]{pipeline1}
-\caption{\label{fig:pipeline} Our pipeline - from the Event Registry we obtain a stream of multi-lingual documents and monolingual clusters.
-Our contribution is shown on the right, where using cross-lingual similarity, we link the clusters across languages which correspond to the same event.}
-\end{figure}
+Finally, we present and interpret the experimental results and finally give conclusions and describe promising future directions.
+
+\section{Pipeline}
 
@@
 
-Our approach is based on first defining a similarity function between articles in different languages (Section~\ref{sec:crosslingual}) and then we present several efficient approaches to linking the clusters based on features (Section~\ref{sec:linking}).
-The pipeline is shown in Figure~\ref{fig:pipeline} First, however we describe the pre-processing on the input articles which we then take as input.
-
-The system is designed to handle streams and so we describe how a document which enters the stream is processed.
-The sequence of operations is shown in  Figure~\ref{fig:erpipeline}.
-A new document is first tokenized, stop words are removed and words are stemmed.
- What remains is represented in a vector-space model and normalized using $TF-IDF$ (see Section~\ref{sec:tfidf} for the definition).
+Our approach is based on first defining a similarity function between articles in different languages (Section~\ref{}) and then we present several efficient approaches to linking the clusters based on features (Section~\ref{}).
+First, however we describe the pre-processing on the input articles which we then take as input.
+
+The system is designed to handle streams and so we describe how a document which enters the stream is processed.   A new document is first tokenized, stop words are removed and words are stemmed.
+ What remains is represented in a vector-space model and normalized using $TF-IDF$ (see Section~\ref{} for the definition).
 Since each article is tagged with its language, cosine similiarity  is used to find the modst similar existing cluster, by comparing it to the centroid vector of each cluster.
@@
 
- Further details about the system can be found in \cite{}.
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=\textwidth]{ev_pipeline}
-\caption{\label{fig:erpipeline}  The Event Registry pipeline, after obtaining the raw stream of multi-lingual news articles, it first cleans and processes articles individually.
-It then aggregates and clusters the articles into similar events (within each language) and finally provides an interface allowing us to query for articles.}
-\end{figure}
+ Further details about the system can be found in \cite{}, while our full pipeline is shown in Figure~\ref{}.
 
@@
 \section{Cross-lingual Document Similarity}
-\label{sec:crosslingual}
+
 Document similarity is an important component in techniques from text mining and natural language processing.
@@
 \subsection{Problem definition}
-\label{sec:tfidf}
+
 \textbf{Document representation.}
@@
 
-\subsection{Canonical Correlation Analysis}\label{sec:CCA}
+\subsection{Canonical Correlation Analysis}
  The final approach is a statistical technique to analyze data from two sources - after which we will describe our main contribution to computing cross-lingual similarities.
@@
 
-The main application on which we test the above similarity is cross-lingual event linking.
+The main application on which we test the above similiarity is cross-lingual event linking.
 In online media streams -- particularly new articles -- there is often duplication of reporting, different viewpoints or opinions, all centering around a single event.
@@
 
-\begin{figure}[tb]
-\centering
-\includegraphics[width=0.6\textwidth]{clusters}
-\caption{\label{fig:clusters}  Given clusters of articles in different languages, the goal of  linking is to match clusters which have similar content across languages.
-The challenge in linking is to overcome the errors in the clustering and differing content as well as make the approach scalable.
-Therefore, we first identify candidate clusters.
-}
-\end{figure}
-
 Furthermore, the matching of clusters is a \emph{generalized matching}.
-We cannot assume that there is only one cluster per language per event, nor can we assume complete coverage -- that there exists at least one cluster per event in every language.
+We cannot assume that there is only one cluster per language per event, nor can we assume complete coverage - that there exists at least one cluster per event in every language.
 This implies that we cannot make any assumptions on the matching, e.g.
@@
 This excludes the use of standard weighted bipartite matching type of algortihms for this problem.
-An example is shown in Figure~\ref{fig:clusters}, where a cluster may contain articles which are closely matched with many clusters in a different language.
 
@@
 Similarity between clusters is then computed using a linear combination of the cosine similarities computed on the lists.
-If the similarity is above the threshold, the clusters are linked.
-Compared to EMM, our approach uses document similarities to obtain a small set of potentially equivalent clusters.
-Additionally, we do not decide if two clusters are equivalent based on a hand-set threshold on a similarity value -- instead we use a classification model that uses a larger set of features related to the tested clusters.
 
@@
 In order to identify clusters $c_j$ that are equivalent to cluster $c_i$, we have developed a two-stage algoirhtm.
-For a cluster $c_i$, we first efficiently identify a small set of candidate clusters and then find those clusters among the candidates, which are equivalent to $c_i$.
-An example is  shown in  Figure~\ref{fig:clusters}.
+First, for a cluster $c_i$, we efficiently identify a small set of candidate clusters and then  find the clusters from among the candidates which are equivalent to $c_i$.
 
@@
 The algorithm begins by individually inspecting each article $a_i$ in the cluster $c_i$.
-Using a chosen method for computing cross-lingual document similarity (see Section~\ref{sec:models}), it identifies the 10 most similar articles to $a_i$ in each language $l \in L$.
+Using CCA  (see Section~\ref{}),  it then identifies the 10 most similar articles to $a_i$ in each language $l \in L$.
 For each  similar article $a_j$, we identify its corresponding  cluster $c_j$ and add it to the set of candidates.
-The set of candidate clusters obtained in this way is several orders of magnitude smaller than the number of all clusters.
-Although computed document similarities are approximate, our  assumption  is that articles in different languages describing the same event, will generally have a higher similarity than articles about different events.
+\todo{we should say a few words why this is more efficient - since now we are comparing each article to all other articles - while we said it is not feadible for clusters}.
+While the CCA simliarity is approximate, our  assumption  is that articles in different languages describing the same event, will generally have a higher similarity than articles about different events.
 While this assumption does not always hold, redundancy in the data mitigates these false positives.
@@
 
-In order to determine how accurately we can predict cluster equivalence we have performed two experiments.
+In order to determine how accurately we can predict cluster equivalence we have performed two experiments in multi-lingual setting using English, German and Spanish language for which we had labeled data to evaluate the linking performance.
 In the first experiment we wanted to test how well do the individual approaches for cross-lingual article linking perform when used for linking the clusters about the same event.
@@
 To test which of these algorithms performs best we performed the following test.
+As a training set we used common Wikipedia alignment for all three languages.
 Using each algorithm, we analyzed all articles in Event Registry to find the most similar articles in other languages.
@@
 The table shows for each of the algorithms the obtained classification accuracy, precision and recall.
-We can see that the most informative features for cluster linking are obtained using the CCA algorithm, while $k$-means performs the worse.
-Firstly we verified the AMRR (average of mean reciprocal ranks over all language pairs) performance of approaches by holding out $15000$ aligned test documents and using $300000$ aligned documents
- as the training set. Figure~\ref{pc:AMRR} shows AMRR score as the function of the number of feature vectors. It is clear that CCA outperforms LSI approach and K-means lags far behind.
-Furthermore we inspected how number of features influences the accuracy of cluster linking.
+We can see that the most informative features for cluster linking are obtained using the CCA or the LSI algorithm, while $k$-means performs the worse.
+To decide which method to use we made some further experiments.
+Firstly we verified the AMRR (average of mean reciprocal ranks over all language pairs) performance of approaches on Wikipedia data by holding out $15000$ aligned test documents and using $300000$ aligned documents
+ as the training set.
+Figure~\ref{pic:AMRR} shows AMRR score as the function of the number of feature vectors.
+It is clear that CCA outperforms LSI approach and K-means lags far behind when testing on Wikipedia data.
+Similar performance of the LSI and CCA for cluster linking can be explained by domain missmatch, we train and test our model  AMRR performance on Wikipedia data but at the end we apply it to the news domain.
+Sadly we do not have any readily available data for this domain so we were not able to test properly.
+Furthermore we inspect how number of features influences the accuracy of cluster linking.
 As we can see from Table~\ref{table:linkingEvalAlgos} choosing number of features vectors larger than $500$ barely effects linking performance.
@@
 Moreover we tested the performance of detecting large/small cluster linking, see Table~\ref{table:linkingEvalAlgosLargeSmall}.
-As expected we obtained that it is harder to link small clusters and we need better retrieval.
-Hub CCA outperforms other methods in this sense also, probably as it has the best retrieval performance, see Figure~\ref{pic:AMRR}.
+Large clusters aggregate a lot of information so the accuracy is not so important as long as we assign to correct clusters.
+In the case of small clusters accuracy should matter more.
+As expected we obtained that it is harder to link small clusters and we need better retrieval accuracy in such cases.
+CCA outperforms other methods in this sense, probably as it has the best retrieval performance, see Figure~\ref{pic:AMRR}.
+Added accuracy is also beneficial when we want
+to pick good representatives of clusters. Therefore we decided to use CCA for cluster linking.
 
@@
 For this purpose we tested the accuracies obtained using individual groups of features, as well as using different combination of groups.
-Since we found the CCA to perform best of the three algorithms, we recomputed $500$ feature vectors with better accuracy and used it to compute the values of the cross-lingual article linking features.
+Since we found the CCA to perform best of the three algorithms, we used it to compute the values of the cross-lingual article linking features.
 Results of the evaluation are shown in Table~\ref{table:linkingEval}.
@@
   Features & CA & Precision & Recall \\ \cline{1-4}
-  CCA & 77.6\% & 0.783 & 0.811 \\ \cline{1-4}
+  CCA & 78.2\% & 0.763 & 0.816 \\ \cline{1-4}
   Concepts & 88.8\% & 0.884 & 0.891 \\ \cline{1-4}
@@
 It is fast, very robust to quality of training data, easily extendable, simple to implement and with low machine requirements.
-Ut currently runs on two Intel Xeon E5-2667 v2, 3.30GHz processor machine with 256GB of RAM.
+The similarity pipeline is the most demanding part and currently runs on two Intel Xeon E5-2667 v2, 3.30GHz processor machine with 256GB of RAM.
 That is enough to do similarity computation over 150 languages if needed.
 It currently uses Wikipedia as freely available knowledge base.
+Similarity pipeline dramastically reduces the search space when linking clusters.
 
 Currently we compute similarities over $24$ languages with tags: eng, spa, deu, zho, ita, fra, rus, swe, nld, tur, jpn, por, ara, fin, ron, kor, hrv, tam, hun, slv, pol, srp, cat, ukr but we support up to $100$ top Wikipedis languages.
-Our data stream is Newsfeed(http://newsfeed.ijs.si/) that produces $430k$ unique articles per day.
-Our system currently computes 400k similiarites per 200ms, that means that we compute $16*10^10$ similarities per day.
+Our data stream is Newsfeed (http://newsfeed.ijs.si/) that produces $430k$ unique articles per day.
+Our system currently computes 400k similiarites per 200ms, that means that we compute $16 \cdot 10^{10}$ similarities per day.
 We
-store one day buffer for every language, for that we need 1.5GB of memory, we store documents as 500 dimensional vectors. For each article we compute top $10$ most similar ones in every language.
-For all linear algebra matrix and vector operations we use high performance numerical libraries as BLAS, OPENBLAS and Intel MKL. This currently allows us to process more than million articles per day.
+store one day buffer for every language, for that we need 1.5 GB of memory, we store documents as 500 dimensional vectors.
+We must note that time complexity of similarity computations scales linearly with dimension of the feature space and does not  depend of number of languages.
+For each article we compute top $10$  most similar ones in every other language.
+For all linear algebra matrix and vector operations we use high performance numerical linear algebra libraries as BLAS, OPENBLAS and Intel MKL.
+This currently allows us to process more than million articles per day.
 In our current implementation we use variation of hub approach.
 Our projector matrices are of size $500\times 300000,$ so every projector takes about $1.1$ GB of RAM.
-Moreover we need proxy matrices of size $500x500$ for every language pair.
+Moreover we need proxy matrices of size $500\times500$ for every language pair.
 That is 0.5 GB for $24$ languages and $9.2$ GB for $100$ languages.
 All together we need around 135 GB of RAM for the system with 100 languages.
-Using proxy matrices we can project all the input documents in the common space.
-That enables us to do blockwise similarity computations and
+ Usage of proxy matrices enables projection of all input documents in the common space and handling language pairs with missing or low alignment.
+That enables us to do block-wise similarity computations and
 additional improvement in our system efficiency.
@@
 This speeds up our code by factor around $4.$ In this way we obtain some caching gains and ability to use vectorization.
-Our system is also easily extendable. If we want to add a language we only need to compute a projector matrix and proxy matrices with all other already available languages.
-
-Comment on the caching gains in our implementation.¸¸Add the conclusion if the story.
+Our system is also easily extendable. If we want to add new language we only need to compute a projector matrix and proxy matrices with all other already available languages.
+
+Add the conclusion of the story and the wrap up, Andrej needs help with this.
+We presented the system ...
 

### 2015-03-24 rupnik2015.tex
- changes througout the document

--- 2015-03-24 rupnik2015.tex
@@
 Compared to EMM, our approach uses document similarities to obtain a small set of potentially equivalent clusters.
-Additionally, we do not decide if two clusters are equivalent based on a hand-set threshold on a similarity value -- instead we use a classification model that uses a larger set of features related to the tested clusters.
+Additionally, we do not decide if two clusters are equivalent based on a hand-set threshold on a similarity value -- instead we use a classification model that uses a larger set of features related to the tested pair of clusters.
 
@@
 The second stage of the algorithm (described in the Algorithm~\ref{cluster_merge_algo2}.) determines which (if any) of the candidate clusters are equivalent to $c_i$.
-We approach this as a supervised learning problem.
+We treat this task as a supervised learning problem.
 For each candidate cluster $c_j \in C'$, we compute a vector of learning features that should be indicative of whether the $c_i$ and $c_j$ are equivalent or not.
-The details of our chosen features will be described in the following Section.
+The details of our chosen features will be described in the following section.
 Once these features are computed, we apply a classification model which was trained using a  dataset  of hand-labeled examples.
@@
 Clusters in each learning example are either in English, Spanish or German.
-Although Event Registry imports articles in other languages as well, we restricted our experiments to these three languages, since they have the highest volume of articles which makes the training problem similarly hard \todo{doesnt this make it easier?}.
+Although Event Registry imports articles in other languages as well, we restricted our experiments to these three languages.
+We chose only these three languages since they have very large number of articles and clusters per day which makes the cluster linking problem hard due to large number of possible links.
 
@@
 \item {\bf Concept related features}.
-As mentioned before, articles that are imported into Event Registry are first semantically annotated by linking mentioned entities and keywords to the corresponding Wikipedia pages.
+Articles that are imported into Event Registry are first semantically annotated by linking mentioned entities and keywords to the corresponding Wikipedia pages.
 Whenever Barack Obama is, for example, mentioned in the article, the article is annotated with a link to his Wikipedia page (\emph{http://en.wikipedia.org/wiki/Barack\_Obama}).
@@
 Using this approach, the word ``avi\'on'' in a Spanish article will be annotated with the same concept as the word ``plane'' in an English article.
-Although the articles are in different languages, the annotations can therefore provide a shared vocabulary that can be used to compare articles/clusters.
+Although the articles are in different languages, the annotations can therefore provide a language-independent vocabulary that can be used to compare articles/clusters.
 By analyzing all the articles in clusters $c_i$ and $c_j$, we can identify the most relevant entities and keywords for each cluster.
@@
 To compute the feature \texttt{sharedDates} we start by identifying the list of dates mentioned in articles in each cluster separately.
-The value of \texttt{sharedDates} is then determined as the ratio of all dates that are detected in both clusters.\todo{this should be made more explicit/}
+The value of \texttt{sharedDates} is then determined as the Jaccard similarity on the two lists of dates.
 
@@
 %To investigate the empirical performance of our algorithm we will test it on a large-scale, real-world multilingual dataset that we extracted from Wikipedia  by  using so called 'inter-language' links as an alignment.
- In this subsection we will investigate the empirical performance of cca hub approach.
+ In this subsection we will investigate the empirical performance of CCA hub approach.
 We will demonstrate that this approach can be successfully applied even in the case of missing alignment information.
@@
 Using each algorithm, we analyzed all articles in Event Registry to find the most similar articles in other languages.
-The values of the cross-lingual article linking features (\texttt{linkCount} and \texttt{avgSimScore}) were then recomputed for all annotated learning examples.
+The values of the cross-lingual article linking features (\texttt{linkCount} and \texttt{avgSimScore}) were therefore recomputed for all annotated learning examples using all three methods.
 A dataset with only these two features and the label was constructed and used to test how accurate prediction models can be induced using the data.
@@
 We can see that the most informative features for cluster linking are obtained using the CCA algorithm, while $k$-means performs the worse.
-Firstly we verified the AMRR (average of mean reciprocal ranks over all language pairs) performance of approaches by holding out $15000$ aligned test documents and using $300000$ aligned documents
- as the training set. Figure~\ref{pc:AMRR} shows AMRR score as the function of the number of feature vectors. It is clear that CCA outperforms LSI approach and K-means lags far behind.
+Firstly we verified the AMRR (average of mean reciprocal ranks over all language pairs) performance of approaches by holding out $15000$ aligned test documents and using $300000$ aligned documents as the training set.
+Figure~\ref{pc:AMRR} shows AMRR score as the function of the number of feature vectors.
+It is clear that CCA outperforms LSI approach and K-means lags far behind.\todo{the sentences after "Firstly we verified" are not clear - seem out of context}
+
 Furthermore we inspected how number of features influences the accuracy of cluster linking.
@@
 That is probably due to overfitting.
-Moreover we tested the performance of detecting large/small cluster linking, see Table~\ref{table:linkingEvalAlgosLargeSmall}.
-As expected we obtained that it is harder to link small clusters and we need better retrieval.
-Hub CCA outperforms other methods in this sense also, probably as it has the best retrieval performance, see Figure~\ref{pic:AMRR}.
+
+Additionally we also tested how accurately can we link clusters depending on the cluster size.
+For this experiment we separated the learning examples into two datasets -- one dataset contained clusters where the sum of articles in both clusters is below 20 articles (356 learning examples) while the other dataset contained the rest of the examples (452 examples).
+10 fold cross validation was then done separately for each of the two datasets.
+Results are shown in Table~\ref{table:linkingEvalAlgosLargeSmall}.
+As expected, we can see that it is harder to link small clusters than large clusters since we have less articles that provide the necessary information.
+We can also see that CCA and LSI outperform the $k$-means approach, which is likely due to their high retrieval performance (see Figure~\ref{pic:AMRR}).
 
@@
 \begin{table}[h]
-\caption{Accuracy of large/small cluster linking with $500$ features obtained from different cross-lingual similarity algorithms}
+\caption{Accuracy of cluster linking using $500$ features on two datasets containing large (left number) and small (right number) clusters}
 \label{table:linkingEvalAlgosLargeSmall}
@@
 \item \textbf{Features} - The experiments show that the hub CCA based features present good baseline, which can greatly benefit from additional semantic extraction based features.
-Even though the CCA based features do not greatly improve the system performance when the semantic features are used, there are two main benefits in the approach: the linking process can be sped up (smaller candidate cluster sets), and in the case when semantic extraction is not possible.
-
-\item \textbf{Future work} - Currently the system is loosely coupled - the language component is built independently of the rest of the system, in particular the linking component.
+Even though the CCA based features do not greatly improve the system performance when the semantic features are used, there are two main benefits of the approach.
+By obtaining a small set of candidate clusters, the cluster linking task can be performed significantly compared to exhaustive evaluation of all cluster pairs.
+Additionally, for many languages, the named entity and keyword extraction is simply not available which prevents us from using them as learning features.
+
+\item \textbf{Future work} - Currently the system is loosely coupled -- the language component is built independently of the rest of the system, in particular the linking component.
 It is possible that better embeddings can be obtained by methods that jointly optimize a classification task and the embedding.
@@
     Another point of interest is to evaluate the system on languages with scarce linguistic resources, where semantic annotation might not be available.
-For this reason, the labelled dataset of linked clusters should to be extended first.
-The mate retrieval evaluation showed that even for language pairs with no training set overlap, the hub cca recovers some signal.
+For this task, the labeled dataset of linked clusters should be extended first.
+The mate retrieval evaluation showed that even for language pairs with no training set overlap, the hub CCA recovers some signal.
 

### 2015-03-24 rupnik2015.tex
Update

--- 2015-03-24 rupnik2015.tex
@@
 \subsection{Evaluation Of Cross-Lingual Event Linking}
-
 In order to determine how accurately we can predict cluster equivalence we have performed two experiments in multi-lingual setting using English, German and Spanish language for which we had labeled data to evaluate the linking performance.
@@
 These algorithms were $k$-means, LSI and CCA.
+As a training set we used common Wikipedia alignment for all three languages.
 To test which of these algorithms performs best we performed the following test.
-As a training set we used common Wikipedia alignment for all three languages.
 Using each algorithm, we analyzed all articles in Event Registry to find the most similar articles in other languages.
@@
 The table shows for each of the algorithms the obtained classification accuracy, precision and recall.
-We can see that the most informative features for cluster linking are obtained using the CCA or the LSI algorithm, while $k$-means performs the worse.
+We can see that the most informative features for cluster linking are obtained using the CCA algorithm, while $k$-means performs the worse.
+Firstly we verified the AMRR (average of mean reciprocal ranks over all language pairs) performance of approaches by holding out $15000$ aligned test documents and using $300000$ aligned documents
+Furthermore we inspected how number of features influences the accuracy of cluster linking.
+As we can see from Table~\ref{table:linkingEvalAlgos} choosing number of features vectors larger than $500$ barely effects linking performance.
+That is probably due to overfitting.
+Moreover we tested the performance of detecting large/small cluster linking, see Table~\ref{table:linkingEvalAlgosLargeSmall}.
+As expected we obtained that it is harder to link small clusters and we need better retrieval.
+Hub CCA outperforms other methods in this sense also, probably as it has the best retrieval performance, see Figure~\ref{pic:AMRR}.
+
 To decide which method to use we made some further experiments.
@@
 
-\item \textbf{Features} - The experiments show that the hub CCA based features present good baseline, which can greatly benefit from additional semantic extraction based features.
-Even though the CCA based features do not greatly improve the system performance when the semantic features are used, there are two main benefits in the approach: the linking process can be sped up (smaller candidate cluster sets), and in the case when semantic extraction is not possible.
-
-\item \textbf{Future work} - Currently the system is loosely coupled - the language component is built independently of the rest of the system, in particular the linking component.
-It is possible that better embeddings can be obtained by methods that jointly optimize a classification task and the embedding.
-
-    Another point of interest is to evaluate the system on languages with scarce linguistic resources, where semantic annotation might not be available.
-For this reason, the labelled dataset of linked clusters should to be extended first.
-The mate retrieval evaluation showed that even for language pairs with no training set overlap, the hub cca recovers some signal.
-
 \end{itemize}
@@
 Currently we compute similarities over $24$ languages with tags: eng, spa, deu, zho, ita, fra, rus, swe, nld, tur, jpn, por, ara, fin, ron, kor, hrv, tam, hun, slv, pol, srp, cat, ukr but we support up to $100$ top Wikipedis languages.
-Our data stream is Newsfeed (http://newsfeed.ijs.si/) that produces $430k$ unique articles per day.
+Our data stream is Newsfeed (http://newsfeed.ijs.si/) that provides $430k$ unique articles per day.
 Our system currently computes 400k similiarites per 200ms, that means that we compute $16 \cdot 10^{10}$ similarities per day.
@@
 Our system is also easily extendable. If we want to add new language we only need to compute a projector matrix and proxy matrices with all other already available languages.
-
-Add the conclusion of the story and the wrap up, Andrej needs help with this.
-We presented the system ...
+Add the conclusion of the story and the wrap up, Andrej needs help with this. We presented the system ...
 

### 2015-03-24 rupnik2015.tex
typos

@@
 
-Currently we compute similarities over $24$ languages with tags: eng, spa, deu, zho, ita, fra, rus, swe, nld, tur, jpn, por, ara, fin, ron, kor, hrv, tam, hun, slv, pol, srp, cat, ukr but we support up to $100$ top Wikipedis languages.
+Currently we compute similarities over $24$ languages with tags: eng, spa, deu, zho, ita, fra, rus, swe, nld, tur, jpn, por, ara, fin, ron, kor, hrv, tam, hun, slv, pol, srp, cat, ukr but we support up to $100$ top Wikipedia languages.
 Our data stream is Newsfeed (http://newsfeed.ijs.si/) that provides $430k$ unique articles per day.

### 2015-03-24 rupnik2015.tex
update to section 3

--- 2015-03-24 rupnik2015.tex
@@
 The paper is organized as follows: we first provide an overview of the system as a whole (Section ~\ref{sec:pipeline}), followed by details on each of the two steps described above.
-We first introduce several approaches for measuring cross-lingual similarity (Section~\ref{sec:crosslingual}) and then the algorithms for matching clusters in different languages (Section~\ref{sec:linking}).
+We first introduce the problem of cross-lingual similarity computation (Section~\ref{sec:crosslingual}), which is followed by several approaches for measuring cross-lingual similarity (Section~\ref{sec:models}) and then the algorithms for matching clusters in different languages (Section~\ref{sec:linking}).
 To improve readability, we describe relevant related work for each module in its corresponding Section.
@@
 
-The processing is done independently for each language and the output is a clustering of articles in each language, with each cluster ideally representing an event, which are defined by topically, geographically and temporally.
-The goal of this paper is present approaches to  connecting the clusters/events across languages.
-
-Our approach is based on first defining a similarity function between articles in different languages (Section~\ref{sec:crosslingual}) and then we present several efficient approaches to linking the clusters based on features (Section~\ref{sec:linking}).
-The pipeline is shown in Figure~\ref{fig:pipeline} First, however we describe the pre-processing on the input articles which we then take as input.
+The processing is done independently for each language and the output is a clustering of articles in each language, with each cluster ideally representing an event, which are defined  topically, geographically and temporally.
+The goal of this paper is to present approaches to connecting the clusters/events across languages.
+
+Our approach is based on first defining a similarity function between articles in different languages (Section~\ref{sec:models}) and then we present several efficient approaches to linking the clusters based on features (Section~\ref{sec:linking}).
+The pipeline is shown in Figure~\ref{fig:pipeline}.
+First, however we describe the pre-processing on the input articles which we then take as input.
 
@@
  What remains is represented in a vector-space model and normalized using $TF-IDF$ (see Section~\ref{sec:tfidf} for the definition).
-Since each article is tagged with its language, cosine similiarity  is used to find the modst similar existing cluster, by comparing it to the centroid vector of each cluster.
+Since each article is tagged with its language, cosine similarity  is used to find the most similar existing cluster, by comparing it to the centroid vector of each cluster.
 A selected threshold is used to determine if the article is not similar to any existing clusters (0.4 was used in our experiments).
@@
 
-The clusters, we consider must have a sufficient number of articles (which is a language dependent paramater).
-This is to ensure that we consider events.
-Once an event is identified(i.e.
+The clusters, we consider must have a sufficient number of articles (which is a language dependent parameter), which is in line with our notion of an event.
+Once an event is identified (i.e.
 a cluster of sufficient size), a new unique ID is assigned to it and the main information about the event is then automatically extracted by analyzing the articles assigned to it.
@@
 
- Further details about the system can be found in \cite{}.
+ The system is described in more detail elsewhere \shortcite{Leban2014W,Leban2014I}.
 
@@
 
-\section{Cross-lingual Document Similarity}
+\section{Cross-lingual Document Similarity}\label{sec:crosslingual}
 \label{sec:crosslingual}
@@
 Many techniques use the similarity as a black box, i.e., a kernel in Support Vector Machines.
-Comparison of documents (or other types of text snippets) is a well studied problem \todo{some references to prove that}.
-In this section we define document similarity in a cross-lingual setting, where the similarity function receives documents in different languages.
-We conclude the section by an introduction of \todo{two datasets} which we used in this paper to learn cross-lingual similarity functions.
+Comparison of documents (or other types of text snippets) in a monolingual setting is a well studied problem in the field of information retrieval \cite{Salton88term-weightingapproaches}.
+In this section we will first cover the related work on cross-lingual similarity models and then formally introduce the problem.
 
@@
 
-\section{Multilingual Latent Factor Models}\label{sec:models}
+\section{Cross-Lingual Models}\label{sec:models}
 In this section, we describe several approaches to computing multilingual simliarities described in the previous section. We present three approaches
@@
 
-\section{Cross-lingual Event Linking}
+\section{Cross-lingual Event Linking}\label{sec:linking}
 

### 2015-03-24 rupnik2015.tex
full pass

@@
 \begin{abstract}
-Today, we follow news which is distributed globally.
+Today we follow news which is distributed globally.
 Significant events are reported by different sources and in different languages.
@@
 Given a multi-lingual stream and clusters of articles from each language, we first propose a method for cross-lingual document similarity based on Wikipedia, which enables us to compute the similarity of any two articles regardless of language.
-The proposed method is can scale to 100 languages and can match articles from languages with little or no direct overlap in the training data.
+The proposed method can scale to 100 languages and can match articles from languages with little or no direct overlap in the training data.
 Using this similarity, we then propose an approach to link clusters of articles across languages which represent the same event.
@@
 
-The term event is vague and ambiguous, but for the practical purposes we define it as ``any significant happening that is being reported about in the media''.
-Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015 (Figure \ref{fig:event2}).
+The term event is vague and ambiguous, but for the practical purposes, we define it as ``any significant happening that is being reported about in the media.'' Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015 (Figure \ref{fig:event2}).
 Events such as these are covered by many articles and the question is how to find all the articles in different languages that are describing a single event.
@@
 
-As input we consider a stream of articles in different languages and a list of events.
+As input, we consider a stream of articles in different languages and a list of events.
 Our goal is to assign articles to their corresponding events.
-A priori we do not know the coverage of the articles, that is not all the events may be covered and we do not know that all the articles necessarily fit into one of the events.
+A priori, we do not know the coverage of the articles, that is not all the events may be covered and we do not know that all the articles necessarily fit into one of the events.
 The task is divided into two parts: detecting events within each language and then linking events across languages.
@@
 We focus on implementations that run on a single shared memory machine, as opposed to clusters of machines.
-This severely simplifies the software implementation and the system maintenance.
+This severely simplifies the software implementation and system maintenance.
 To summarize, the following properties are desireable:
@@
 The system should be able to take advantage of comparable corpus (as opposed to parallel translation based corpora), with missing data.
-Supporting a new language should be simple.
-\item \textbf{operation efficiency} - The similarity computation should be fast - the system needs to handle billions of similarity computations per day.
+Supporting a new language should be straightforward.
+\item \textbf{operation efficiency} - The similarity computation should be fast - the system must be able to handle billions of similarity computations per day.
 \item \textbf{operation cost} - The system should run on a strong shared machine server and not rely on paid services.
@@
 The parameter estimation is posed as an inference problem which is typically intractable and one usually solves it using approximate techniques.
-Most variants of solutions are based on Gibbs sampling or Variational Inference, which are not trivial to implement and may require an experienced practitioner to apply.
-Furthermore, representing a new document as a mixture of topics is another potentially hard inference problem which needs to be solved.
+Most variants of solutions are based on Gibbs sampling or Variational Inference, which are nontrivial to implement and may require an experienced practitioner to apply.
+Furthermore, representing a new document as a mixture of topics is another potentially hard inference problem which must be solved.
 \item \textbf{classification problem related} - In approaches such as \cite{cotrain} and \shortcite{svm2k} the authors find linear embeddings of the data that support good classification results, based on a given classification problem training data.
@@
 
-(figure?)
-
 \section{Cross-lingual Event Linking}\label{sec:linking}
@@
 
-In order to train a classification model for predicting whether two clusters are equivalent, we must first have labeled data.
+In order to train a classification model for predicting whether two clusters are equivalent, we must first have labelled data.
 To obtain the data, we used two human annotators.
@@
 Articles that are imported into Event Registry are first semantically annotated by linking mentioned entities and keywords to the corresponding Wikipedia pages.
-Whenever Barack Obama is, for example, mentioned in the article, the article is annotated with a link to his Wikipedia page (\emph{http://en.wikipedia.org/wiki/Barack\_Obama}).
+Whenever Barack Obama is, for example, mentioned in the article, the article is annotated with a link to his Wikipedia page\footnote{\emph{http://en.wikipedia.org/wiki/Barack\_Obama}}.
 In the same way all mentions of people, locations, organizations and even ordinary keywords (e.g.
@@
 
-We describe the datasets involved in the experiments and two evaluations: comparing the crosslingual approaches on an information retrieval task, and evaluation of cross-lingual event linking.
+We describe the datasets involved in the experiments and two evaluations: comparing the cross-lingual approaches on an information retrieval task, and an evaluation of cross-lingual event linking.
 
@@
 %{\oldText{Taken from ITI paper}
-To investigate the empirical performance of the low rank approximations we will test the algorithms on a large-scale, real-world multilingual dataset that we extracted from Wikipedia by using inter-language links as an alignment.
+To investigate the empirical performance of the low rank approximations we will test the algorithms on a large-scale, real-world multi-lingual dataset that we extracted from Wikipedia by using inter-language links for alignment.
 This  results in a large number of weakly comparable documents in more than $200$ languages.
-Wikipedia is a large source of multilingual data that is especially important for the languages for which no translation tools, multilingual dictionaries as Eurovoc, or strongly aligned multilingual corpora as Europarl are available.
+Wikipedia is a large source of multi-lingual data that is especially important for the languages for which no translation tools, multi-lingual dictionaries as Eurovoc, or strongly aligned multi-lingual corpora as Europarl are available.
 Documents in different languages are related with so called 'inter-language' links that can be found on the left of the Wikipedia page.
 The Wikipedia is constantly growing.
-There are currently 12 Wikipedias with more than $10^6$ articles, $52$ with more than $10^5$ articles, $129$ with more than $10^4$ articles, and $236$ with more than $1000$ articles.
+There are currently 12 Wikipedias with more than 1 million%$10^6$
+ articles, $52$ with more than 100k %$10^5$ 
+ articles, $129$ with more than 10 k articles, and $236$ with more than $1000$ articles.
 %Wikipedia uses special user-friendly markup language that is very easy to write but very hard to parse.
@@
 Each Wikipedia page is embedded in the page tag.
-First we check if the title of the page consists of any special namespace and do not process such pages.
-Then we check if this is a redirection page and we store the redirect link as inter-language links can point to redirection link also.
-If nothing of the above applies we extract the text and parse the Wikipedia markup.
-Currently all the markup is removed.
+First, we check if the title of the page consists of any special namespace and do not process such pages.
+Then, we check if this is a redirection page and we store the redirect link as inter-language links can point to redirection link also.
+If none of the above applies, we extract the text and parse the Wikipedia markup.
+Currently, all the markup is removed.
 
@@
 %To investigate the empirical performance of our algorithm we will test it on a large-scale, real-world multilingual dataset that we extracted from Wikipedia  by  using so called 'inter-language' links as an alignment.
- In this subsection we will investigate the empirical performance of CCA hub approach.
+ In this subsection, we will investigate the empirical performance of CCA hub approach.
 We will demonstrate that this approach can be successfully applied even in the case of missing alignment information.
- To this purpose we select a subset of Wikipedia languages containing three major languages, English--\emph{en} (hub language), Spanish--\emph{es}, Russian--\emph{ru}, and five minority (in the sense of Wikipedia sizes) languages, Slovenian--\emph{sl}, Piedmontese--\emph{pms}, Waray-Waray--\emph{war} (all with about 2 million native speakers), Creole--\emph{ht} (8 million native speakers), and Hindi--\emph{hi} (180 million native speakers).
-For preprocessing we remove the documents that contain less than 20 different words (stubs) and remove words occurring in less than 50 documents as well as the top 100 most frequent words (in each language separately).
+ To this purpose, we select a subset of Wikipedia languages containing three major languages, English--\emph{en} (hub language), Spanish--\emph{es}, Russian--\emph{ru}, and five minority (in the sense of Wikipedia sizes) languages, Slovenian--\emph{sl}, Piedmontese--\emph{pms}, Waray-Waray--\emph{war} (all with about 2 million native speakers), Creole--\emph{ht} (8 million native speakers), and Hindi--\emph{hi} (180 million native speakers).
+For preprocessing, we remove the documents that contain less than 20 different words (stubs) and remove words occurring in less than 50 documents as well as the top 100 most frequent words (in each language separately).
 We represent the documents as normalized TFIDF\cite{Salton88term-weightingapproaches} weighted vectors.
@@
 
-The evaluation is based on splitting the data into training and test sets (described later).
+The evaluation is based on splitting the data into training and test sets (which are described later).
 On the training set, we perform the two step procedure to obtain the common document representation as a set of mappings $P_i$.
@@
 It is clear that in the case of Creole language only at most $22\%$ documents are unique and suitable for the training.
-Though we removed the stub documents, many of remaining documents are almost the same, as the quality of some minor Wikipedias is low.
+Though we removed the stub documents, many of remaining documents are nearly the same, as the quality of some minor Wikipedias is low.
 This was confirmed for Creole, Waray-Waray, and Piedmontese language by manual inspection.
@@
 
-We also have a problem with the quality of the test data.
+There is also have a problem with the quality of the test data.
 For example, if we look at test pair (\emph{war}, \emph{ht}) only 386/534 Waray-Waray test documents are unique but on other side almost all Creole test documents (523/534) are unique.
@@
 \subsection{Evaluation Of Cross-Lingual Event Linking}
-In order to determine how accurately we can predict cluster equivalence we have performed two experiments in multi-lingual setting using English, German and Spanish language for which we had labeled data to evaluate the linking performance.
-In the first experiment we wanted to test how well do the individual approaches for cross-lingual article linking perform when used for linking the clusters about the same event.
+In order to determine how accurately we can predict cluster equivalence, we performed two experiments in multi-lingual setting using English, German and Spanish language for which we had labelled data to evaluate the linking performance.
+In the first experiment, we tested how well  the individual approaches for cross-lingual article linking perform when used for linking the clusters about the same event.
 In the second experiment we tested how accurate is the prediction model when trained on different subsets of learning features.
@@
 
-In Section~\ref{sec:models} we have described three main algorithms for identifying similar articles in different languages.
+In Section~\ref{sec:models}, we  described three main algorithms for identifying similar articles in different languages.
 These algorithms were $k$-means, LSI and CCA.
-As a training set we used common Wikipedia alignment for all three languages.
-To test which of these algorithms performs best we performed the following test.
+As a training set, we used common Wikipedia alignment for all three languages.
+To test which of these algorithms performed best, we performed the following test.
 Using each algorithm, we analyzed all articles in Event Registry to find the most similar articles in other languages.
@@
 Firstly we verified the AMRR (average of mean reciprocal ranks over all language pairs) performance of approaches by holding out $15000$ aligned test documents and using $300000$ aligned documents
-Furthermore we inspected how number of features influences the accuracy of cluster linking.
-As we can see from Table~\ref{table:linkingEvalAlgos} choosing number of features vectors larger than $500$ barely effects linking performance.
-That is probably due to overfitting.
-Moreover we tested the performance of detecting large/small cluster linking, see Table~\ref{table:linkingEvalAlgosLargeSmall}.
+Furthermore, we inspected how number of features influences the accuracy of cluster linking.
+As we can see from Table~\ref{table:linkingEvalAlgos} choosing number of features vectors larger than $500$ barely effects linking performance, which is most likely due to overfitting.
+Moreover, we tested the performance of detecting large/small cluster linking, see Table~\ref{table:linkingEvalAlgosLargeSmall}.
 As expected we obtained that it is harder to link small clusters and we need better retrieval.
@@
 
-To decide which method to use we made some further experiments.
-Firstly we verified the AMRR (average of mean reciprocal ranks over all language pairs) performance of approaches on Wikipedia data by holding out $15000$ aligned test documents and using $300000$ aligned documents
+To decide which method to use, we perfomed some further experiments.
+Firstly, we verified the AMRR (average of mean reciprocal ranks over all language pairs) performance of the different approaches on the  Wikipedia data by holding out $15000$ aligned test documents and using $300000$ aligned documents
  as the training set.
@@
 It is clear that CCA outperforms LSI approach and K-means lags far behind when testing on Wikipedia data.
-Similar performance of the LSI and CCA for cluster linking can be explained by domain missmatch, we train and test our model  AMRR performance on Wikipedia data but at the end we apply it to the news domain.
-Sadly we do not have any readily available data for this domain so we were not able to test properly.
-Furthermore we inspect how number of features influences the accuracy of cluster linking.
-As we can see from Table~\ref{table:linkingEvalAlgos} choosing number of features vectors larger than $500$ barely effects linking performance.
+The similar performance of the LSI and CCA for cluster linking can be explained by domain mismatch.
+We train and test our model  AMRR performance on the Wikipedia data, but in the end apply it to the news domain.
+Unfortunately, we do not have any readily available data for this domain so we were not able to test properly.
+Furthermore, we inspect how number of features influences the accuracy of cluster linking.
+As we can see from Table~\ref{table:linkingEvalAlgos} choosing number of features vectors larger than $500$ barely affects linking performance.
 That is probably due to overfitting.
 Moreover we tested the performance of detecting large/small cluster linking, see Table~\ref{table:linkingEvalAlgosLargeSmall}.
-Large clusters aggregate a lot of information so the accuracy is not so important as long as we assign to correct clusters.
-In the case of small clusters accuracy should matter more.
-As expected we obtained that it is harder to link small clusters and we need better retrieval accuracy in such cases.
-CCA outperforms other methods in this sense, probably as it has the best retrieval performance, see Figure~\ref{pic:AMRR}.
-Added accuracy is also beneficial when we want
-to pick good representatives of clusters. Therefore we decided to use CCA for cluster linking.
+Large clusters aggregate a lot of information, so the accuracy is not as important so long as we  assign articles to correct clusters.
+In the case of small clusters, we expect the  accuracy to matter more.
+As expected, we obtained that it is harder to link small clusters and so require better retrieval accuracy.
+CCA outperforms other methods in this sense, mos likely since  it has the best retrieval performance, see Figure~\ref{pic:AMRR}.
+It is worth noting that added accuracy is also beneficial when we wish
+to choose good representatives of clusters. Therefore, we decided to use CCA for cluster linking.
 
@@
 
-In the second experiment we wanted to evaluate how relevant individual groups of features are to correctly determine cluster equivalence.
-For this purpose we tested the accuracies obtained using individual groups of features, as well as using different combination of groups.
-Since we found the CCA to perform best of the three algorithms, we used it to compute the values of the cross-lingual article linking features.
-Results of the evaluation are shown in Table~\ref{table:linkingEval}.
-We can see that using a single group of features, the highest prediction accuracy can be achieved using the concept related features.
+In the second experiment, we evaluate how relevant individual groups of features are to correctly determine cluster equivalence.
+For this purpose, we tested the accuracies obtained using individual groups of features, as well as using different combination of groups.
+Since  CCA had the best  performance of the three algorithms, we used it to compute the values of the cross-lingual article linking features.
+The results of the evaluation are shown in Table~\ref{table:linkingEval}.
+We can see that using a single group of features, the highest prediction accuracy can be achieved using  concept related features.
 The classification accuracy in this case is 88.8\%.
@@
 Using all three groups of features, the achieved accuracy is 89.6\%.
-Based on the results we can make the following conclusions.
+
+Based on the results, we can make the following conclusions.
 The cross-lingual similarity algorithms provide valuable information that can be used to identify clusters that describe the same event in different languages.
@@
 Nevertheless, the cross-lingual article similarity features are very important for two reasons.
-The first reason is that they allow us to identify for a given cluster a limited set of candidate clusters that are potentially equivalent.
-This is a very important feature since it reduces the search space for several orders of magnitude.
-The second reason for the importance of these features is that concept annotations are not available for all articles.
-Annotation of news articles is computationally intensive and can only be done for a subset of collected articles.
+The first  is that they allow us to identify for a given cluster a limited set of candidate clusters that are potentially equivalent.
+This is a very important feature since it reduces the search space by several orders of magnitude.
+The second reason these features are important is that concept annotations are not available for all articles as the annotation of news articles is computationally intensive and can only be done for a subset of collected articles.
 
@@
 The interpretation is that the linking features are highly aggregated for large clusters, which compensates the lower per-document percision of LSI.
-Another possible reason is, that the advantage that we show on Wikipedia is lost on the news domain - which indicates that domain adaptation should be explored.
+Another possible reason is that the advantage that we show on Wikipedia is lost on the news domain - which suggest that domain adaptation should be explored.
 
@@
 %Premultiply with proxy matrices, compare blockwise!
-One of the main advantages of our approach is that scales really well.
-It is fast, very robust to quality of training data, easily extendable, simple to implement and with low machine requirements.
-The similarity pipeline is the most demanding part and currently runs on two Intel Xeon E5-2667 v2, 3.30GHz processor machine with 256GB of RAM.
-That is enough to do similarity computation over 150 languages if needed.
-It currently uses Wikipedia as freely available knowledge base.
-Similarity pipeline dramastically reduces the search space when linking clusters.
+One of the main advantages of our approach is highly scalable.
+It is fast, very robust to quality of training data, easily extendable, simple to implement and has relatively small hardware requirements.
+The similarity pipeline is the most computationally intensive part and currently runs on two Intel Xeon E5-2667 v2, 3.30GHz processor machine with 256GB of RAM.
+This is sufficient to do similarity computation over 150 languages if needed.
+It currently uses Wikipedia as a freely available knowledge base and experiments show that the similarity pipeline dramatically reduces the search space when linking clusters.
 
 Currently we compute similarities over $24$ languages with tags: eng, spa, deu, zho, ita, fra, rus, swe, nld, tur, jpn, por, ara, fin, ron, kor, hrv, tam, hun, slv, pol, srp, cat, ukr but we support up to $100$ top Wikipedia languages.
-Our data stream is Newsfeed (http://newsfeed.ijs.si/) that provides $430k$ unique articles per day.
+Our data stream is Newsfeed (http://newsfeed.ijs.si/) which provides $430k$ unique articles per day.
 Our system currently computes 400k similiarites per 200ms, that means that we compute $16 \cdot 10^{10}$ similarities per day.
 We
-store one day buffer for every language, for that we need 1.5 GB of memory, we store documents as 500 dimensional vectors.
-We must note that time complexity of similarity computations scales linearly with dimension of the feature space and does not  depend of number of languages.
-For each article we compute top $10$  most similar ones in every other language.
-For all linear algebra matrix and vector operations we use high performance numerical linear algebra libraries as BLAS, OPENBLAS and Intel MKL.
-This currently allows us to process more than million articles per day.
-In our current implementation we use variation of hub approach.
+store one day buffer for each language which requires 1.5 GB of memory with documents   stored as 500 dimensional vectors.
+We  note that the time complexity of the similarity computations scales linearly with dimension of the feature space and does not  depend of number of languages.
+For each article, we compute the top $10$  most similar ones in every other language.
+
+For all linear algebra matrix and vector operations, we use high performance numerical linear algebra libraries as BLAS, OPENBLAS and Intel MKL, which currently allows us to process more than million articles per day.
+In our current implementation, we use the variation of hub approach.
 Our projector matrices are of size $500\times 300000,$ so every projector takes about $1.1$ GB of RAM.
-Moreover we need proxy matrices of size $500\times500$ for every language pair.
+Moreover, we need proxy matrices of size $500\times500$ for every language pair.
 That is 0.5 GB for $24$ languages and $9.2$ GB for $100$ languages.
@@
  Usage of proxy matrices enables projection of all input documents in the common space and handling language pairs with missing or low alignment.
-That enables us to do block-wise similarity computations and
-additional improvement in our system efficiency.
-Our code can be therefore easily parallelized using matrix multiplication instead of doing more matrix - vector multiplications.
-This speeds up our code by factor around $4.$ In this way we obtain some caching gains and ability to use vectorization.
-Our system is also easily extendable. If we want to add new language we only need to compute a projector matrix and proxy matrices with all other already available languages.
-Add the conclusion of the story and the wrap up, Andrej needs help with this. We presented the system ...
+That enables us to do block-wise similarity computations further improving system efficiency.
+Our code can therefore be easily parallelized using matrix multiplication irather than performing more matrix - vector multiplications.
+This speeds up our code by a factor around $4.$ In this way, we obtain some caching gains and ability to use vectorization.
+Our system is also easily extendable. Adding a new language requires the  computation of  a projector matrix and proxy matrices with all other already available languages.
+%Add the conclusion of the story and the wrap up, Andrej needs help with this. We presented the system ...
+\subsection{Future Work}
 

### 2015-03-24 rupnik2015.tex
added stuff that got removed

--- 2015-03-24 rupnik2015.tex
@@
 
+\item \textbf{Features} - The experiments show that the hub CCA based features present good baseline, which can greatly benefit from additional semantic extraction based features.
+Even though the CCA based features do not greatly improve the system performance when the semantic features are used, there are two main benefits in the approach: the linking process can be sped up (smaller candidate cluster sets), and in the case when semantic extraction is not possible.
+
+\item \textbf{Future work} - Currently the system is loosely coupled - the language component is built independently of the rest of the system, in particular the linking component.
+It is possible that better embeddings can be obtained by methods that jointly optimize a classification task and the embedding.
+
+    Another point of interest is to evaluate the system on languages with scarce linguistic resources, where semantic annotation might not be available.
+For this reason, the labelled dataset of linked clusters should to be extended first.
+The mate retrieval evaluation showed that even for language pairs with no training set overlap, the hub cca recovers some signal.
+
 \end{itemize}

### 2015-03-24 rupnik2015.tex
small update

@@
 \begin{itemize}
-\item \textbf{training} - The training (building corss-lingual models) should scale to many languages and should be robust to the quality of training resources.
+\item \textbf{Training} - The training (building corss-lingual models) should scale to many languages and should be robust to the quality of training resources.
 The system should be able to take advantage of comparable corpus (as opposed to parallel translation based corpora), with missing data.
 Supporting a new language should be straightforward.
-\item \textbf{operation efficiency} - The similarity computation should be fast - the system must be able to handle billions of similarity computations per day.
-\item \textbf{operation cost} - The system should run on a strong shared machine server and not rely on paid services.
-\item \textbf{implementation} - The system is simple to implement, with few parameters to tune.
+\item \textbf{Operation efficiency} - The similarity computation should be fast - the system must be able to handle billions of similarity computations per day.
+\item \textbf{Operation cost} - The system should run on a strong shared machine server and not rely on paid services.
+\item \textbf{Implementation} - The system is simple to implement, with few parameters to tune.
 \end{itemize}
@@
 \begin{itemize}
-\item \textbf{translation} - The most obvious way to compare documents written in different languages is to use machine translation and perform monolingual similarity.
+\item \textbf{Translation} - The most obvious way to compare documents written in different languages is to use machine translation and perform monolingual similarity.
 One can use free tools such as Moses \shortcite{moses} or translation services, such as Google Translate\footnote{https://translate.google.com/}.
@@
 The operation efficiency and cost requirements make translation based approaches less suited for our system.
-\item \textbf{probabilistic topic models} -  There exist many variants to modelling documents in a language independent way by using probabilistic graphical models.
+\item \textbf{Probabilistic topic models} -  There exist many variants to modelling documents in a language independent way by using probabilistic graphical models.
 The models include:  Joint Probabilistic Latent Semantic Analysis (JPLSA) \shortcite{platt2010translingual}, Coupled Probabilistic LSA (CPLSA) \cite{platt2010translingual}, Probabilistic Cross-Lingual LSA (PCLLSA)\cite{PCL_LSA} and Polylingual Topic Models (PLTM) \shortcite{polyLDA} which is a Bayesian version of PCLLSA.
@@
 Furthermore, representing a new document as a mixture of topics is another potentially hard inference problem which must be solved.
-\item \textbf{classification problem related} - In approaches such as \cite{cotrain} and \shortcite{svm2k} the authors find linear embeddings of the data that support good classification results, based on a given classification problem training data.
+\item \textbf{Classification problem related} - In approaches such as \cite{cotrain} and \shortcite{svm2k} the authors find linear embeddings of the data that support good classification results, based on a given classification problem training data.
 We do not consider this family of approaches, since for general language pairs classifier training sets may not be available.
-\item \textbf{matrix factorization} - Non-negative matrix factorization based \cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CLLSI\shortcite{cl_lsi}, Canonical Correlation Analysis (CCA) \cite{Hotelling}, Oriented Principal Component Analysis \cite{platt2010translingual}.
+\item \textbf{Matrix factorization} - Non-negative matrix factorization based \cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CLLSI\shortcite{cl_lsi}, Canonical Correlation Analysis (CCA) \cite{Hotelling}, Oriented Principal Component Analysis \cite{platt2010translingual}.
 The quadratic time and space dependency of the OPCA method makes it impractical for large scale purposes.
@@
 
-\subsection{Wikipedia Information Retrieval}
-
-Mate retrieval results
-
 \subsection{Experiments With Missing Alignment Data}

### 2015-03-24 rupnik2015.tex
discussion

--- 2015-03-24 rupnik2015.tex
@@
 
-\subsection{Wikipedia Information Retrieval}
-
-Mate retrieval results
+%\subsection{Wikipedia Information Retrieval}
+
+%Mate retrieval results
 
@@
 
-\section{Discussion and Future Work}
-
-We now discuss some aspects of the methods and the system as a whole.
-\begin{itemize}
-\item \textbf{LSI and hub CCA} - We first comment on the empirical validation of LSI and hub CCA.
-On the task of mate retrieval we observe that refining the LSI based projections with hub CCA leads to improved retrieval precision, but the methods perform comparably on the task of event linking.
-Further inspection showed that the CCA based approach reached a higher precision on smaller clusters.
-The interpretation is that the linking features are highly aggregated for large clusters, which compensates the lower per-document percision of LSI.
-Another possible reason is that the advantage that we show on Wikipedia is lost on the news domain - which suggest that domain adaptation should be explored.
-
-\item \textbf{Features} - The experiments show that the hub CCA based features present good baseline, which can greatly benefit from additional semantic extraction based features.
-Even though the CCA based features do not greatly improve the system performance when the semantic features are used, there are two main benefits in the approach: the linking process can be sped up (smaller candidate cluster sets), and in the case when semantic extraction is not possible.
-
-\item \textbf{Future work} - Currently the system is loosely coupled - the language component is built independently of the rest of the system, in particular the linking component.
-It is possible that better embeddings can be obtained by methods that jointly optimize a classification task and the embedding.
-
-    Another point of interest is to evaluate the system on languages with scarce linguistic resources, where semantic annotation might not be available.
-For this reason, the labelled dataset of linked clusters should to be extended first.
-The mate retrieval evaluation showed that even for language pairs with no training set overlap, the hub cca recovers some signal.
-
-\end{itemize}
-
 \subsection{Scaling}
@@
 %Add the conclusion of the story and the wrap up, Andrej needs help with this. We presented the system ...
-\subsection{Future Work}
+\section{Discussion and Future Work}
+In this paper we have presented a cross-lingual system for linking events in different languages.
+Building on an existing system, Event Registry, we present and evaluate several approaches to computing a  cross-lingual simialirty function.
+We also present an approach to linking events and evaluate effectiveness of various features.
+The  final pipeline is scalable both in terms of number of articles and number of languages, while accurately linking events.
+
+Finally, we discuss some aspects of the methods and the system as a whole.
+\begin{itemize}
+\item \textbf{LSI and hub CCA} - We first comment on the empirical validation of LSI and hub CCA.
+On the task of mate retrieval we observe that refining the LSI based projections with hub CCA leads to improved retrieval precision, but the methods perform comparably on the task of event linking.
+Further inspection showed that the CCA based approach reached a higher precision on smaller clusters.
+The interpretation is that the linking features are highly aggregated for large clusters, which compensates the lower per-document percision of LSI.
+Another possible reason is that the advantage that we show on Wikipedia is lost on the news domain - which suggest that domain adaptation should be explored.
+
+\item \textbf{Features} - The experiments show that the hub CCA based features present good baseline, which can greatly benefit from additional semantic extraction based features.
+Even though the CCA based features do not greatly improve the system performance when the semantic features are used, there are two main benefits in the approach: the linking process can be sped up (smaller candidate cluster sets), and in the case when semantic extraction is not possible.
+
+\item \textbf{Future work} - Currently the system is loosely coupled - the language component is built independently of the rest of the system, in particular the linking component.
+It is possible that better embeddings can be obtained by methods that jointly optimize a classification task and the embedding.
+
+    Another point of interest is to evaluate the system on languages with scarce linguistic resources, where semantic annotation might not be available.
+For this reason, the labelled dataset of linked clusters should to be extended first.
+The mate retrieval evaluation showed that even for language pairs with no training set overlap, the hub cca recovers some signal.
+
+\end{itemize}
+
+%\subsection{Future Work}
 

### 2015-03-24 rupnik2015.tex

--- 2015-03-24 rupnik2015.tex
@@
 
+%\subsection{Wikipedia Information Retrieval}
+
+%Mate retrieval results
+
 \subsection{Experiments With Missing Alignment Data}

### 2015-03-24 rupnik2015.tex
minor
likanje

@@
 
-\subsection{Problem definition}
-\label{sec:tfidf}
+\subsection{Problem definition}\label{sec:tfidf}
+We will start by describing how documents are represented as vectors and how to compare in a mono-lingual setting.
+We will then define a way to measure cross-lingual similarity that is natural for the models we consider.
+
 \textbf{Document representation.} 
-
 Standard vector space model \cite{Salton88term-weightingapproaches} represents documents as vectors, where each term corresponds to word or phrase in a fixed vocabulary.
@@
 \end{enumerate}
- We concentrate on approaches that are based on linear maps rather than alternatives, such as machine translation and probabilistic models. Applicability of machine translation relies
-on the availability of training resources (strongly aligned translation corpora), which are not available for many world languages, but there exist comparable corpora. Probabilistic
-methods \todo{provide reasons for not using them: e.g. difficulty with parameter estimation, maybe just say future work?}.
-
+ We concentrate on approaches that are based on linear maps rather than alternatives, such as machine translation and probabilistic models, as discussed in the related work section. 
 For completeness we will also present the method of Canonical Correlation Analysis, since it is related to our proposed method.
@@
 \subsection{Scaling}
-\emph{I think that this section should be short.}
-100 languages!
+%\emph{I think that this section should be short.}
+%100 languages!
 %Memory 100 GB per day
@@
 
-\acks{The authors wish to thank blah blah blah.}
+%\acks{The authors wish to thank blah blah blah.}
 

### 2015-03-24 rupnik2015.tex
likanje

@@
 \subsection{Problem definition}\label{sec:tfidf}
-We will start by describing how documents are represented as vectors and how to compare in a mono-lingual setting.
+We will first describe how documents are represented as vectors and how to compare in a mono-lingual setting.
 We will then define a way to measure cross-lingual similarity that is natural for the models we consider.
@@
  We concentrate on approaches that are based on linear maps rather than alternatives, such as machine translation and probabilistic models, as discussed in the related work section. 
-For completeness we will also present the method of Canonical Correlation Analysis, since it is related to our proposed method.
+For completeness we will also present the method of Canonical Correlation Analysis, since it is related to our proposed method. We will start by introducing some notation.
 
@@
 
- Canonical Correlation Analysis (CCA)~\cite{Hotelling} is a dimensionality reduction technique similar to Principal Component Analysis (PCA)~\cite{Pearson1901On}, with an additional assumption that the data consists of feature vectors that arose from two sources (two views) that share some information (examples: bilingual document collection~\shortcite{mrpqr}, collection of images and captions~\shortcite{Hardoon_usingimage}).
+ Canonical Correlation Analysis (CCA)~\cite{Hotelling} is a dimensionality reduction technique similar to Principal Component Analysis (PCA)~\cite{Pearson1901On}, with an additional assumption that the data consists of feature vectors that arose from two sources (two views) that share some information.
+Examples include: bilingual document collection ~\shortcite{mrpqr} and collection of images and captions~\shortcite{Hardoon_usingimage}.
 Instead of looking for linear combinations of features that maximize the variance (PCA) we look for a linear combination of feature vectors from the first view and a linear combination for the second view, that are maximally correlated.
@@
     \For{language $l \in L$} {
-        \tcc{use CCA to find 10 most similar articles to article $a_i$ in language $l$}
+        \tcc{use hub CCA to find 10 most similar articles to article $a_i$ in language $l$}
         $SimArt = getCCASimArts(a_i, l)$\;
@@
 Using the  features described above and the user-provided labels, we train a classification model $M$ which is used in Algorithm~\ref{cluster_merge_algo2} to classify a new cluster pair as equivalent or not.
-The classification algorithm that we used to train a model was a linear SVM method.
+The classification algorithm that we used to train a model was a linear Support Vector Machine (SVM) method \shortcite{shawe-taylor04kernel}.
 
@@
 
-We describe the datasets involved in the experiments and two evaluations: comparing the cross-lingual approaches on an information retrieval task, and an evaluation of cross-lingual event linking.
-
-\subsection{Datasets}
+We will describe the main dataset for building cross-lingual models which is based on Wikipedia. After that we will present three sets of experiments. The first set of experiments
+will establish that the hub based approach can deal with language pairs where little or no training data is available. The second set of experiments will compare the main approaches
+the we presented on the task of mate retrieval. We will then explore how different choices of features influence the performance of event linking.
+
+\subsection{Wikipedia Comparable Corpus}
 %\emph{Andrej started shuffling it, will do it till Monday :) }
@@
  In this subsection, we will investigate the empirical performance of CCA hub approach.
-We will demonstrate that this approach can be successfully applied even in the case of missing alignment information.
+We will demonstrate that this approach can be successfully applied even in the case of fully missing alignment information.
  To this purpose, we select a subset of Wikipedia languages containing three major languages, English--\emph{en} (hub language), Spanish--\emph{es}, Russian--\emph{ru}, and five minority (in the sense of Wikipedia sizes) languages, Slovenian--\emph{sl}, Piedmontese--\emph{pms}, Waray-Waray--\emph{war} (all with about 2 million native speakers), Creole--\emph{ht} (8 million native speakers), and Hindi--\emph{hi} (180 million native speakers).
@@
 In Section~\ref{sec:models}, we  described three main algorithms for identifying similar articles in different languages.
-These algorithms were $k$-means, LSI and CCA.
+These algorithms were $k$-means, LSI and hub CCA.
 As a training set, we used common Wikipedia alignment for all three languages.
@@
 The table shows for each of the algorithms the obtained classification accuracy, precision and recall.
-We can see that the most informative features for cluster linking are obtained using the CCA algorithm, while $k$-means performs the worse.
-Firstly we verified the AMRR (average of mean reciprocal ranks over all language pairs) performance of approaches by holding out $15000$ aligned test documents and using $300000$ aligned documents
+We can see that the most informative features for cluster linking are obtained using the hub CCA algorithm, while $k$-means performs the worse.
+Firstly we verified the Average (over language pairs) Mean Reciprocal Ranks (AMRR) performance of approaches by holding out $15000$ aligned test documents and using $300000$ aligned documents
 Furthermore, we inspected how number of features influences the accuracy of cluster linking.
@@
 
-To decide which method to use, we perfomed some further experiments.
+To decide which method to use, we performed some further experiments.
 Firstly, we verified the AMRR (average of mean reciprocal ranks over all language pairs) performance of the different approaches on the  Wikipedia data by holding out $15000$ aligned test documents and using $300000$ aligned documents
@@
 Figure~\ref{pic:AMRR} shows AMRR score as the function of the number of feature vectors.
-It is clear that CCA outperforms LSI approach and K-means lags far behind when testing on Wikipedia data.
-The similar performance of the LSI and CCA for cluster linking can be explained by domain mismatch.
+It is clear that hub CCA outperforms LSI approach and K-means lags far behind when testing on Wikipedia data.
+The similar performance of the LSI and hub CCA for cluster linking can be explained by domain mismatch.
 We train and test our model  AMRR performance on the Wikipedia data, but in the end apply it to the news domain.
@@
 As expected, we obtained that it is harder to link small clusters and so require better retrieval accuracy.
-CCA outperforms other methods in this sense, mos likely since  it has the best retrieval performance, see Figure~\ref{pic:AMRR}.
+Hub CCA outperforms other methods in this sense, most likely since  it has the best retrieval performance, see Figure~\ref{pic:AMRR}.
 It is worth noting that added accuracy is also beneficial when we wish
-to choose good representatives of clusters. Therefore, we decided to use CCA for cluster linking.
+to choose good representatives of clusters. Therefore, we decided to use hub CCA for cluster linking.
 
@@
 For this purpose, we tested the accuracies obtained using individual groups of features, as well as using different combination of groups.
-Since  CCA had the best  performance of the three algorithms, we used it to compute the values of the cross-lingual article linking features.
+Since hub CCA had the best  performance of the three algorithms, we used it to compute the values of the cross-lingual article linking features.
 The results of the evaluation are shown in Table~\ref{table:linkingEval}.
@@
 Our data stream is Newsfeed (http://newsfeed.ijs.si/) which provides $430k$ unique articles per day.
-Our system currently computes 400k similiarites per 200ms, that means that we compute $16 \cdot 10^{10}$ similarities per day.
+Our system currently computes 400k similarities per 200ms, that means that we compute $16 \cdot 10^{10}$ similarities per day.
 We
@@
 That enables us to do block-wise similarity computations further improving system efficiency.
-Our code can therefore be easily parallelized using matrix multiplication irather than performing more matrix - vector multiplications.
+Our code can therefore be easily parallelized using matrix multiplication rather than performing more matrix - vector multiplications.
 This speeds up our code by a factor around $4.$ In this way, we obtain some caching gains and ability to use vectorization.
 Our system is also easily extendable. Adding a new language requires the  computation of  a projector matrix and proxy matrices with all other already available languages.
-%Add the conclusion of the story and the wrap up, Andrej needs help with this. We presented the system ...
+
 \section{Discussion and Future Work}
 In this paper we have presented a cross-lingual system for linking events in different languages.
-Building on an existing system, Event Registry, we present and evaluate several approaches to computing a  cross-lingual simialirty function.
+Building on an existing system, Event Registry, we present and evaluate several approaches to computing a  cross-lingual similarity function.
 We also present an approach to linking events and evaluate effectiveness of various features.
@@
 Further inspection showed that the CCA based approach reached a higher precision on smaller clusters.
-The interpretation is that the linking features are highly aggregated for large clusters, which compensates the lower per-document percision of LSI.
+The interpretation is that the linking features are highly aggregated for large clusters, which compensates the lower per-document precision of LSI.
 Another possible reason is that the advantage that we show on Wikipedia is lost on the news domain - which suggest that domain adaptation should be explored.
@@
 For this reason, the labelled dataset of linked clusters should to be extended first.
-The mate retrieval evaluation showed that even for language pairs with no training set overlap, the hub cca recovers some signal.
+The mate retrieval evaluation showed that even for language pairs with no training set overlap, the hub CCA recovers some signal.
 

### 2015-03-24 rupnik2015.tex
likanje

@@
 will establish that the hub based approach can deal with language pairs where little or no training data is available. The second set of experiments will compare the main approaches
-the we presented on the task of mate retrieval. We will then explore how different choices of features influence the performance of event linking.
+the we presented on the task of mate retrieval and the task of event linking. We will also examine how different choices of features impact the event linking performance.
 
 \subsection{Wikipedia Comparable Corpus}
-%\emph{Andrej started shuffling it, will do it till Monday :) }
-%{\oldText{Taken from ITI paper}
+
 To investigate the empirical performance of the low rank approximations we will test the algorithms on a large-scale, real-world multi-lingual dataset that we extracted from Wikipedia by using inter-language links for alignment.
@@
  articles, $129$ with more than 10 k articles, and $236$ with more than $1000$ articles.
-%Wikipedia uses special user-friendly markup language that is very easy to write but very hard to parse.
-Simplicity of language can cause ambiguities and moreover it is constantly changing.
-For example, separator | can be used in %different contexts.
-
-%Wikipedia raw xml dumps of all currently $288$ active editions were downloaded from the Wikipedia dump page\href{http://dumps.wikimedia.org}.
-The xml files are too large to be parsed with DOM like parser that needs to store the whole xml tree in the memory, instead we implemented Sax like parser that tries to simulate behavior of Wikipedia official parser and is as simple, fast and error prone as possible.
-We parse all Wikipedia markup but do not expand the templates.
+
 Each Wikipedia page is embedded in the page tag.
@@
 This matrix enables us to build an alignment across all Wikipedia languages.
-%}
-
-%\subsection{Wikipedia Information Retrieval}
-
-%Mate retrieval results
 
 \subsection{Experiments With Missing Alignment Data}
-%\emph{Andrej started working on this, done by Monday}
-%{\oldText{Taken from XLite paper}
-%To investigate the empirical performance of our algorithm we will test it on a large-scale, real-world multilingual dataset that we extracted from Wikipedia  by  using so called 'inter-language' links as an alignment.
+
  In this subsection, we will investigate the empirical performance of CCA hub approach.

### 2015-03-24 rupnik2015.tex
updated

@@
 
-\subsection{Experiments With Missing Alignment Data}
+\subsection{Experiments With Missing Alignment Data}\label{experiments:hubcca}
 
@@
 
+% which method?
 In Section~\ref{sec:models}, we  described three main algorithms for identifying similar articles in different languages.
@@
 We can see that the most informative features for cluster linking are obtained using the hub CCA algorithm, while $k$-means performs the worse.
-Firstly we verified the Average (over language pairs) Mean Reciprocal Ranks (AMRR) performance of approaches by holding out $15000$ aligned test documents and using $300000$ aligned documents
-Furthermore, we inspected how number of features influences the accuracy of cluster linking.
-As we can see from Table~\ref{table:linkingEvalAlgos} choosing number of features vectors larger than $500$ barely effects linking performance, which is most likely due to overfitting.
-Moreover, we tested the performance of detecting large/small cluster linking, see Table~\ref{table:linkingEvalAlgosLargeSmall}.
-As expected we obtained that it is harder to link small clusters and we need better retrieval.
-Hub CCA outperforms other methods in this sense also, probably as it has the best retrieval performance, see Figure~\ref{pic:AMRR}.
-
-To decide which method to use, we performed some further experiments.
-Firstly, we verified the AMRR (average of mean reciprocal ranks over all language pairs) performance of the different approaches on the  Wikipedia data by holding out $15000$ aligned test documents and using $300000$ aligned documents
- as the training set.
-Figure~\ref{pic:AMRR} shows AMRR score as the function of the number of feature vectors.
+
+% AMMR experiments
+We also compared the proposed approaches on the task of Wikipedia mate retrieval (the same task as in Section~\ref{experiments:hubcca}).
+We computed the Average (over language pairs) Mean Reciprocal Rank (AMRR) \shortcite{voorhees1999trec}  performance of the different approaches on the  Wikipedia data by holding out $15000$ aligned test documents and using $300000$ aligned documents as the training set.
+The Figure~\ref{pic:AMRR} shows AMRR score as the function of the number of feature vectors.
 It is clear that hub CCA outperforms LSI approach and K-means lags far behind when testing on Wikipedia data.
-The similar performance of the LSI and hub CCA for cluster linking can be explained by domain mismatch.
-We train and test our model  AMRR performance on the Wikipedia data, but in the end apply it to the news domain.
-Unfortunately, we do not have any readily available data for this domain so we were not able to test properly.
-Furthermore, we inspect how number of features influences the accuracy of cluster linking.
-As we can see from Table~\ref{table:linkingEvalAlgos} choosing number of features vectors larger than $500$ barely affects linking performance.
-That is probably due to overfitting.
-Moreover we tested the performance of detecting large/small cluster linking, see Table~\ref{table:linkingEvalAlgosLargeSmall}.
-Large clusters aggregate a lot of information, so the accuracy is not as important so long as we  assign articles to correct clusters.
+The hub CCA approach manages with $500$ features performam comparably to $LSI$ based approach with $1000$-features, which shows that the $CCA$ method can improve both model memory footprint as well as similarity computation time.
+
+% number of topics?
+Furthermore, we inspected how the number of features (topics) influences the accuracy of cluster linking.
+As we can see from Table~\ref{table:linkingEvalAlgos} choosing number of features vectors larger than $500$ barely affects linking performance, which is in contrast with the fact that additional features helped improve AMMR, see Figure~\ref{pic:AMRR}.
+Such differences may have arisen due to different domains of training and testing (Wikipedia pages versus news articles).
+
+% cluster size?
+We also analyzed how cluster size influences the linking performance.
+The intuition is that clusters with large number of documents are easier to link, which can be seen in Table~\ref{table:linkingEvalAlgosLargeSmall}.
+Large clusters aggregate a lot of information, so the performance of the similarity model is not as important so long as we assign articles to correct clusters.
 In the case of small clusters, we expect the accuracy to matter more.
-As expected, we obtained that it is harder to link small clusters and so require better retrieval accuracy.
-Hub CCA outperforms other methods in this sense, most likely since  it has the best retrieval performance, see Figure~\ref{pic:AMRR}.
-It is worth noting that added accuracy is also beneficial when we wish
-to choose good representatives of clusters. Therefore, we decided to use hub CCA for cluster linking.
+Our experiments indicate that this is indeed the case: it is harder to link small clusters and better retrieval accuracy may matter more.
+Hub CCA outperforms other methods in this sense in terms of precision, which is consistent with its AMRR performance, see Figure~\ref{pic:AMRR}.
+It is worth noting that added accuracy is also beneficial when we wish to choose good representatives of clusters. Therefore, we decided to use hub CCA in the cluster linking component.
 

### 2015-03-24 rupnik2015.tex
small fixes

@@
 \begin{figure}
-\caption{Average of mean reciprocal ranks}
 \label{pic:AMRR}
 %\includegraphics{slika.png}
+\centering
 \input{retrieval}
+\caption{Average of mean reciprocal ranks}
 \end{figure}
@@
 %Premultiply with proxy matrices, compare blockwise!
-One of the main advantages of our approach is highly scalable.
+One of the main advantages of our approach is that it is highly scalable.
 It is fast, very robust to quality of training data, easily extendable, simple to implement and has relatively small hardware requirements.
@@
 
-Currently we compute similarities over $24$ languages with tags: eng, spa, deu, zho, ita, fra, rus, swe, nld, tur, jpn, por, ara, fin, ron, kor, hrv, tam, hun, slv, pol, srp, cat, ukr but we support up to $100$ top Wikipedia languages.
+Currently we compute similarities over $24$ languages with tags: eng, spa, deu, zho, ita, fra, rus, swe, nld, tur, jpn, por, ara, fin, ron, kor, hrv, tam, hun, slv, pol, srp, cat, ukr but we support any language from the top $100$ Wikipedia languages.
 Our data stream is Newsfeed (http://newsfeed.ijs.si/) which provides $430k$ unique articles per day.

### 2015-03-24 rupnik2015.tex
Small fixes of typos.

@@
 A new document is first tokenized, stop words are removed and words are stemmed.
- What remains is represented in a vector-space model and normalized using $TF-IDF$ (see Section~\ref{sec:tfidf} for the definition).
+ What remains is represented in a vector-space model and normalized using $TFIDF$ (see Section~\ref{sec:tfidf} for the definition).
 Since each article is tagged with its language, cosine similarity  is used to find the most similar existing cluster, by comparing it to the centroid vector of each cluster.
@@
 We do not address the problem of detection of events in this paper but rather consider the problem that given a series of events, how may we best ``tag" or match articles in different languages to these events.
-The events are represented by clusters of articles and so ultimately our problem reduces to find suitable matchings between clusters with articles in different languages.
+The events are represented by clusters of articles and so ultimately our problem reduces to finding suitable matchings between clusters with articles in different languages.
 
@@
 The Wikipedia is constantly growing.
-There are currently 12 Wikipedias with more than 1 million%$10^6$
+There are currently 12 Wikipedias with more than 1 million %$10^6$
  articles, $52$ with more than 100k %$10^5$
- articles, $129$ with more than 10 k articles, and $236$ with more than $1000$ articles.
+ articles, $129$ with more than 10k articles, and $236$ with more than $1000$ articles.
 
@@
 
- In this subsection, we will investigate the empirical performance of CCA hub approach.
+ In this subsection, we will investigate the empirical performance of hub CCA approach.
 We will demonstrate that this approach can be successfully applied even in the case of fully missing alignment information.
@@
 \subsection{Evaluation Of Cross-Lingual Event Linking}
-In order to determine how accurately we can predict cluster equivalence, we performed two experiments in multi-lingual setting using English, German and Spanish language for which we had labelled data to evaluate the linking performance.
+In order to determine how accurately we can predict cluster equivalence, we performed two experiments in multi-lingual setting, English, German and Spanish language for which we had labelled data to evaluate the linking performance.
 In the first experiment, we tested how well  the individual approaches for cross-lingual article linking perform when used for linking the clusters about the same event.
@@
 It is clear that hub CCA outperforms LSI approach and K-means lags far behind when testing on Wikipedia data.
-The hub CCA approach manages with $500$ features performam comparably to $LSI$ based approach with $1000$-features, which shows that the $CCA$ method can improve both model memory footprint as well as similarity computation time.
+The hub CCA approach with $500$ features manages to perform comparably to $LSI$ based approach with $1000$-features, which shows that the $CCA$ method can improve both model memory footprint as well as similarity computation time.
 
@@
 store one day buffer for each language which requires 1.5 GB of memory with documents   stored as 500 dimensional vectors.
-We  note that the time complexity of the similarity computations scales linearly with dimension of the feature space and does not  depend of number of languages.
+We  note that the time complexity of the similarity computations scales linearly with dimension of the feature space and does not  depend on number of languages.
 For each article, we compute the top $10$  most similar ones in every other language.
@@
 \item \textbf{Features} - The experiments show that the hub CCA based features present good baseline, which can greatly benefit from additional semantic extraction based features.
-Even though the CCA based features do not greatly improve the system performance when the semantic features are used, there are two main benefits in the approach: the linking process can be sped up (smaller candidate cluster sets), and in the case when semantic extraction is not possible.
+Even though the CCA based features do not greatly improve the system performance when the semantic features are used, there are two main benefits in the approach: the linking process can be sped up (smaller candidate cluster sets), and usage in the case when semantic extraction is not possible.
 

### 2015-07-10 rupnik2015.tex
first pass

@@
 \newcommand{\algInput}{\textbf{input:}}
+
+\newcommand{\primoz}[#1]{\textcolor{red}{#1}}
 
@@
 Comparison of documents (or other types of text snippets) in a monolingual setting is a well studied problem in the field of information retrieval \cite{Salton88term-weightingapproaches}.
-In this section we will first cover the related work on cross-lingual similarity models and then formally introduce the problem.
+We first recount some system requirements, then review  related approaches to cross-lingual similarity models and then formally introduce the problem as well as our approach.
+
+\primoz{we may want to move this out to the pipeline section}
+\subsection{System Requirements}
+
+Our goal is to build a system that monitors global media and analyzes how events are being reported on.
+Our approach consists of two steps: tracking events separately in each language (based on language detection and an online clustering approach) and then connecting them.
+The pipeline must be able to process millions of articles per day and perform billions of similarity computations each day - implying that similarity computation should be highly scalable.
+%The system should support as many languages as possible.
+ We focus on implementations that run on a single shared memory machine, as opposed to clusters of machines. This  simplifies implementation and system maintenance.
+To summarize, the following properties are desireable:
+\begin{itemize}
+\item \textbf{Training} - The training (building corss-lingual models) should scale to many languages and should be robust to the quality of training resources.
+The system should be able to take advantage of comparable corpus (as opposed to parallel translation based corpora), with missing data.
+Supporting a new language should be straightforward.
+\item \textbf{Operation efficiency} - The similarity computation should be fast - the system must be able to handle billions of similarity computations per day.
+\item \textbf{Operation cost} - The system should run on a strong shared machine server and not rely on paid services.
+\item \textbf{Implementation} - The system is straightforward to implement, with few parameters to tune.
+\end{itemize}
+We believe that a cross-lingual similarity component that meets such requirements is very desirable in a commercial setting, where several different costs have to be taken into consideration.
 
@@
 
-In this section we will relate the method we use to other alternatives in the literature.
+%In this section we will relate the method we use to other alternatives in the literature.
 We will first state some requirements for our system, which motivate our choice of focusing on a particular subset of approaches in the rest of the paper.
@@
 The system should be able to take advantage of comparable corpus (as opposed to parallel translation based corpora), with missing data.
-Supporting a new language should be straightforward.
-\item \textbf{Operation efficiency} - The similarity computation should be fast - the system must be able to handle billions of similarity computations per day.
-\item \textbf{Operation cost} - The system should run on a strong shared machine server and not rely on paid services.
+\primoz{We could say something more subject to moving this section}%Supporting a new language should be straightforward.
+\item \textbf{Efficiency} - The similarity computation should be fast - the system must be able to handle billions of similarity computations per day.
+\item \textbf{Cost} - The system should run on a strong shared machine server and not rely on paid services.
 \item \textbf{Implementation} - The system is simple to implement, with few parameters to tune.
@@
 
-There are three main families of approaches to cross-lingual similarity.
+There are four main families of approaches to cross-lingual similarity.
 \begin{itemize}
@@
 
-Based on the discussion above, we chose to focus on methods based on vector space models and linear embeddings.
+Based on our requirements in Section~\ref{} \primoz{System requirements}, we chose to focus on methods based on vector space models and linear embeddings.
 We propose a method that is more efficient than popular alternatives (a clustering based approach and latent semantic indexing), but is still simple to optimize and use.
@@
 \subsection{Problem definition}\label{sec:tfidf}
-We will first describe how documents are represented as vectors and how to compare in a mono-lingual setting.
+We will first describe how documents are represented as vectors and how to compare douments in a mono-lingual setting.
 We will then define a way to measure cross-lingual similarity that is natural for the models we consider.
@@
 \textbf{Document representation.}
-Standard vector space model \cite{Salton88term-weightingapproaches} represents documents as vectors, where each term corresponds to word or phrase in a fixed vocabulary.
-More formally, document $d$ is represented by a vector $x \in \RR^n$, where $n$ corresponds to the size of the vocabulary, and vector elements $x_k$ correspond to the number of times term $k$ occurred in the document, also called \emph{term frequency} or $TF_k(d)$.
+The standard vector space model \cite{Salton88term-weightingapproaches} represents documents as vectors, where each term corresponds to word or phrase in a fixed vocabulary.
+Formally, document $d$ is represented by a vector $x \in \RR^n$, where $n$ corresponds to the size of the vocabulary, and vector elements $x_k$ correspond to the number of times term $k$ occurred in the document, also called \emph{term frequency} or $TF_k(d)$.
 
@@
 An \emph{Inverse Document Frequency} ($IDF$) weight for the dictionary term $k$ is defined as $\log\left( \frac{N}{DF_k} \right)$, where $DF_k$ is the number of documents in the corpus which contain term $k$.
-A document $TFIDF$ vector is its original vector multiplied element-wise by the weights.
-
+Therefore we can definte a document's $TFIDF$ as
+$$ x_{ij}  := \frac{\mbox{term frequency in document } i}{\mbox{inverse document frequency of term } j}
+%A document $TFIDF$ vector is its original vector multiplied element-wise by the weights.
+%
 The $TFIDF$ weighted vector space model document representation corresponds to a map $\phi : \text{text} \rightarrow \RR^n$ defined by:
@@
 where $\langle \cdot,\cdot \rangle$ and $\|\cdot\|$ are standard inner product and Euclidean norm.
-Cosine similarity, and other related approaches, assumes that the similarity is reflected in the overlap of words, and as such works only when the documents $d_1$ and $d_2$ are written in the same language.
+Cosine similarity, and other related approaches, assumes that the similarity is reflected in the overlap of words.
+This metric is only meaningful when the documents $d_1$ and $d_2$ are defined on the same space (e.g.
+have a comparable vocabulary or are written in the same language).
 

### 2015-07-15 rupnik2015.tex
check
revisions in section 4. Still left TODO: 4.5 and additional figures

@@
 
-\newcommand{\primoz}[#1]{\textcolor{red}{#1}}
+\newcommand{\primoz}[1]{\textcolor{red}{#1}}
 
@@
 Therefore we can definte a document's $TFIDF$ as
-$$ x_{ij}  := \frac{\mbox{term frequency in document } i}{\mbox{inverse document frequency of term } j}
-%A document $TFIDF$ vector is its original vector multiplied element-wise by the weights.
-%
+$$ x_{ij}  := \frac{\mbox{term frequency in document } i}{\mbox{inverse document frequency of term } j}.$$
 The $TFIDF$ weighted vector space model document representation corresponds to a map $\phi : \text{text} \rightarrow \RR^n$ defined by:
@@
 \section{Cross-Lingual Models}\label{sec:models}
-In this section, we describe several approaches to computing multilingual simliarities described in the previous section. We present three approaches
-\begin{enumerate}
-\item  a simple approach based on $k$-means clustering
-\item  a standard approach based on singular value decomposition
-\item a new method that is applicable in more general settings.
-\end{enumerate}
+In this section, we describe several approaches to the problem of computing multilingual simliarities introduced in \ref{sec:tfidf}. We present four approaches:
+a simple approach based on $k$-means clustering in \ref{sec:kmeans}, a standard approach based on singular value decomposition in \ref{sec:LSI}, a related
+approach called Canonical Correlation Analysis (CCA) in \ref{sec:CCA} and finally a new, more generally applicable method, which is an extension of CCA to more than two languages in \ref{sec:hublang}.
+
+CCA can be used to find
+correlated patterns for a pair of languages, whereas the extended method optimizes a
+sum of squared correlations between several language pairs.
+The method takes advantage of the hub language assumption in the training dataset (the English Wikipedia corresponds to a hub) as well as a preprocessing step based on LSI that acts as regularization and improves scalability.
+
 We concentrate on approaches that are based on linear maps rather than alternatives, such as machine translation and probabilistic models, as discussed in the related work section.
-For completeness we will also present the method of Canonical Correlation Analysis, since it is related to our proposed method. We will start by introducing some notation.
+We will start by introducing some notation.
 
@@
 
-\subsection{$k$-means}
-
-The $k$-means algorithm is perhaps the most well-known and widely-used clustering algorithm.
+\subsection{$k$-means}\label{sec:kmeans}
+
+The $k$-means algorithm is perhaps the most well-known and widely-used clustering algorithm. We now present its application 
+to computing cross-lingual similarities.
+The idea is based on concatenating the corpus matrices, running standard $k$-means clustering to obtain the matrix of centroids, "reversing" the concatenation step to obtain a set of aligned bases, which are finally used to compute cross-lingual similarities.
+
  In order to apply the algorithm, we first merge all the term-document matrices into a single matrix $X$ by stacking the individual term-document matrices:
@@
 
+The matrices $C_i$ are obtained by analyzing the multilingual corpus matrix $X$ (obtained from Wikipedia for example) and we now show how to use them to compute cross-lingual document similarities on new documents (news documents, for example).
+
 Each matrix $C_i$ represents a vector space basis and can be used to map points in $\RR^{n_i}$ into a $k$-dimensional space, where the coordinates of a vector $x \in \RR^{n_i}$ are expressed as: $$(C_i^T C_i)^{-1} C_i^T x_i.$$
@@
 The matrix is a result of mapping documents in a language independent space using pseudo-inverses of the centroid matrices $P_i = (C_i^T C_i)^{-1} C_i$ and then comparing them using the standard inner product, which results in the matrix $P_i^T P_j$.
-For the sake of presentation, we assumed that the centroid vectors are linearly independent (an independent subspace could be obtained using an additional Gram-Schmidt step on the matrix $C$, if this was not the case).
-
-\subsection{Cross-Lingual Latent Semantic Indexing}
+For the sake of presentation, we assumed that the centroid vectors are linearly independent (an independent subspace could be obtained using an additional Gram-Schmidt step~\cite{golub} on the matrix $C$, if this was not the case).
+
+\subsection{Cross-Lingual Latent Semantic Indexing}\label{sec:LSI}
 
@@
 
-\subsection{Hub languages}
-
-In this section, we describe an extension to CCA, but is more applicable to a large number of languages. The main difficulty with applying the
+\subsection{Hub languages}\label{sec:hublang}
+
+In this section, we describe an extension to CCA, which is more applicable to a large number of languages. The main difficulty with applying the
 LSI and $k$-means approaches lies in the fact, that when one considers a large number of languages, for example, the top 100 Wikipedia languages (ranked by number of articles), the set of completely

### 2015-07-15 rupnik2015.tex
Restructured: Description of the pipeline comes after introduction, requirements for the system as a whole (needs to cover event linking) related work (cross-lingual + event linking) comes after

@@
 
-\section{Cross-lingual Document Similarity}\label{sec:crosslingual}
-\label{sec:crosslingual}
-Document similarity is an important component in techniques from text mining and natural language processing.
-Many techniques use the similarity as a black box, i.e., a kernel in Support Vector Machines.
-Comparison of documents (or other types of text snippets) in a monolingual setting is a well studied problem in the field of information retrieval \cite{Salton88term-weightingapproaches}.
-We first recount some system requirements, then review  related approaches to cross-lingual similarity models and then formally introduce the problem as well as our approach.
-
 \primoz{we may want to move this out to the pipeline section}
@@
  We focus on implementations that run on a single shared memory machine, as opposed to clusters of machines. This  simplifies implementation and system maintenance.
-To summarize, the following properties are desireable:
+To summarize, the following properties are desirable:
 \begin{itemize}
-\item \textbf{Training} - The training (building corss-lingual models) should scale to many languages and should be robust to the quality of training resources.
+\item \textbf{Training} - The training (building cross-lingual models) should scale to many languages and should be robust to the quality of training resources.
 The system should be able to take advantage of comparable corpus (as opposed to parallel translation based corpora), with missing data.
@@
 
+\subsection{Related work}
+
+Although there are a number of services that aggregate news by identifying clusters of similar articles, there are almost no services that provide linking of clusters over different languages.
+Google News as well as Yahoo!
+News are able to identify clusters of articles about same event, but they offer no linking of clusters across languages.
+The only service that we found, which provides cross-lingual cluster linking, is the European Media Monitor (EMM)~\cite{Steinberger2008}.
+EMM clusters articles in 60 languages and then tries to determine which clusters of articles in different languages describe the same event.
+To achieve cluster linking, EMM uses three different language independent vector representations for each cluster.
+First vector contains the weighted list of references to countries mentioned in the articles, while the second vector contains the weighted list of mentioned people and organizations.
+The last vector contains the weighted list of Eurovoc subject domain descriptors.
+These descriptors are topics, such as \emph{air transport}, \emph{EC agreement}, \emph{competition} and \emph{pollution control} into which articles are automatically categorized~\shortcite{Pouliquen2006}.
+Similarity between clusters is then computed using a linear combination of the cosine similarities computed on the lists.
+If the similarity is above the threshold, the clusters are linked.
+Compared to EMM, our approach uses document similarities to obtain a small set of potentially equivalent clusters.
+Additionally, we do not decide if two clusters are equivalent based on a hand-set threshold on a similarity value -- instead we use a classification model that uses a larger set of features related to the tested pair of clusters.
+
+A system that is significantly different, but worth mentioning is the GDELT project~\cite{Leetaru2013Gdelt}.
+In GDELT, events are also extracted from articles, but in their case, an event is specified in a form of a triple containing two actors and a relation.
+The project contains an extensive vocabulary of possible relations, mostly related to political events.
+In order to identify events, GDELT collects articles in more than 65 languages and uses machine translation to translate them to English.
+All information extraction is then done on the translated article.
+
+\section{Cross-lingual Document Similarity}\label{sec:crosslingual}
+\label{sec:crosslingual}
+Document similarity is an important component in techniques from text mining and natural language processing.
+Many techniques use the similarity as a black box, i.e., a kernel in Support Vector Machines.
+Comparison of documents (or other types of text snippets) in a monolingual setting is a well studied problem in the field of information retrieval \cite{Salton88term-weightingapproaches}.
+We first recount some system requirements, then review  related approaches to cross-lingual similarity models and then formally introduce the problem as well as our approach.
+
 \subsection{Problem definition}\label{sec:tfidf}
@@
 
-\section{Cross-Lingual Models}\label{sec:models}
+\subsection{Cross-Lingual Models}\label{sec:models}
 In this section, we describe several approaches to the problem of computing multilingual simliarities introduced in \ref{sec:tfidf}. We present four approaches:
@@
 
-\subsection{Hub languages}\label{sec:hublang}
-
-In this section, we describe an extension to CCA, which is more applicable to a large number of languages. The main difficulty with applying the
+\subsection{Hub language based CCA Extension}\label{sec:hublang}
+Building cross-lingual similarity models based on comparable corpora is challenging for two main reasons.
+The problem of missing alignment data: when a number of languages is large, the dataset of documents that cover all languages is small (or may even be empty).
+Even if only two languages are considered, the set of aligned documents can be small (an extreme example is given by the Piedmontese and Hindi Wikipedias where no inter-language links are available).
+The second problem is that the number of documents and the involved
+1. missing data
+2. LSI step
+3. Sum of squared correlations
+
+In this section, we describe an extension of CCA to more than two languages which can be trained on large comparable corpora. The main difficulty with applying the
 LSI and $k$-means approaches lies in the fact, that when one considers a large number of languages, for example, the top 100 Wikipedia languages (ranked by number of articles), the set of completely
@@
 
-\subsection{Related work}
-
-Although there are a number of services that aggregate news by identifying clusters of similar articles, there are almost no services that provide linking of clusters over different languages.
-Google News as well as Yahoo!
-News are able to identify clusters of articles about same event, but they offer no linking of clusters across languages.
-The only service that we found, which provides cross-lingual cluster linking, is the European Media Monitor (EMM)~\cite{Steinberger2008}.
-EMM clusters articles in 60 languages and then tries to determine which clusters of articles in different languages describe the same event.
-To achieve cluster linking, EMM uses three different language independent vector representations for each cluster.
-First vector contains the weighted list of references to countries mentioned in the articles, while the second vector contains the weighted list of mentioned people and organizations.
-The last vector contains the weighted list of Eurovoc subject domain descriptors.
-These descriptors are topics, such as \emph{air transport}, \emph{EC agreement}, \emph{competition} and \emph{pollution control} into which articles are automatically categorized~\shortcite{Pouliquen2006}.
-Similarity between clusters is then computed using a linear combination of the cosine similarities computed on the lists.
-If the similarity is above the threshold, the clusters are linked.
-Compared to EMM, our approach uses document similarities to obtain a small set of potentially equivalent clusters.
-Additionally, we do not decide if two clusters are equivalent based on a hand-set threshold on a similarity value -- instead we use a classification model that uses a larger set of features related to the tested pair of clusters.
-
-A system that is significantly different, but worth mentioning is the GDELT project~\cite{Leetaru2013Gdelt}.
-In GDELT, events are also extracted from articles, but in their case, an event is specified in a form of a triple containing two actors and a relation.
-The project contains an extensive vocabulary of possible relations, mostly related to political events.
-In order to identify events, GDELT collects articles in more than 65 languages and uses machine translation to translate them to English.
-All information extraction is then done on the translated article.
-
 \subsection{Algorithm}

### 2015-07-16 rupnik2015.tex
Bib and main paper fixes

--- 2015-07-10 rupnik2015.tex
@@
 Therefore we can definte a document's $TFIDF$ as
-$$ x_{ij}  := \frac{\mbox{term frequency in document } i}{\mbox{inverse document frequency of term } j}
+$$ x_{ij}  := \frac{\mbox{term frequency in document } i}{\mbox{inverse document frequency of term } j}$$
 %A document $TFIDF$ vector is its original vector multiplied element-wise by the weights.
@@
 
-We will describe the main dataset for building cross-lingual models which is based on Wikipedia. After that we will present three sets of experiments. The first set of experiments
+We will describe the main dataset for building cross-lingual models which is based on Wikipedia. After that we will present two sets of experiments. The first set of experiments
 will establish that the hub based approach can deal with language pairs where little or no training data is available. The second set of experiments will compare the main approaches
@@
 This  results in a large number of weakly comparable documents in more than $200$ languages.
-Wikipedia is a large source of multi-lingual data that is especially important for the languages for which no translation tools, multi-lingual dictionaries as Eurovoc, or strongly aligned multi-lingual corpora as Europarl are available.
+Wikipedia is a large source of multi-lingual data that is especially important for the languages for which no translation tools, multi-lingual dictionaries as Eurovoc \cite{eurovoc}, or strongly aligned multi-lingual corpora as Europarl \cite{europarl} are available.
 Documents in different languages are related with so called 'inter-language' links that can be found on the left of the Wikipedia page.
@@
 In the rare case that we have multiple links pointing from the document, we pick the first one that we encountered.
-This matrix enables us to build an alignment across all Wikipedia languages.
+\todo{Make this more sharp} This matrix enables us to build an alignment across all Wikipedia languages.
 
@@
 This documents are typically garbage, the titles of the columns in the table, remains of the parsing process, or Wikipedia articles with very little or no information contained in one or two sentences.
+\todo{Move this to footnote for stubs}
 
@@
 \begin{figure}
-\label{pic:AMRR}
 %\includegraphics{slika.png}
@@
 \caption{Average of mean reciprocal ranks}
+\label{pic:AMRR}
 \end{figure}

### 2015-07-17 rupnik2015.tex
Section 6 changes. Gregor can you help with the stuff marked with orange.

@@
 will establish that the hub based approach can deal with language pairs where little or no training data is available. The second set of experiments will compare the main approaches
-the we presented on the task of mate retrieval and the task of event linking. We will also examine how different choices of features impact the event linking performance.
+that we presented on the task of mate retrieval and the task of event linking. We will also examine how different choices of features impact the event linking performance.
 
@@
 To correct this we transform this matrix to symmetric by computing $M+M^T$ and obtaining an undirected graph.
-In the rare case that we have multiple links pointing from the document, we pick the first one that we encountered.
-\todo{Make this more sharp} This matrix enables us to build an alignment across all Wikipedia languages.
+In the rare case that after symmetrization we have multiple links pointing from the document, we pick the first one that we encountered.
+This matrix enables us to build an alignment across all Wikipedia \footnote{https://www.wikipedia.org/ dumps available in 2013} languages.
 
@@
 We will demonstrate that this approach can be successfully applied even in the case of fully missing alignment information.
- To this purpose, we select a subset of Wikipedia languages containing three major languages, English--\emph{en} (hub language), Spanish--\emph{es}, Russian--\emph{ru}, and five minority (in the sense of Wikipedia sizes) languages, Slovenian--\emph{sl}, Piedmontese--\emph{pms}, Waray-Waray--\emph{war} (all with about 2 million native speakers), Creole--\emph{ht} (8 million native speakers), and Hindi--\emph{hi} (180 million native speakers).
-For preprocessing, we remove the documents that contain less than 20 different words (stubs) and remove words occurring in less than 50 documents as well as the top 100 most frequent words (in each language separately).
+ To this purpose, we select a subset of Wikipedia languages containing three major languages, English (4,212k articles)--\emph{en} (hub language), Spanish (9,686k articles)--\emph{es}, Russian (9,662k articles)--\emph{ru}, and five minority (in the sense of Wikipedia sizes) languages, Slovenian (136k articles)--\emph{sl}, Piedmontese (59k articles)--\emph{pms}, Waray-Waray (112k articles)--\emph{war} (all with about 2 million native speakers), Creole (54k articles)--\emph{ht} (8 million native speakers), and Hindi (97k articles)--\emph{hi} (180 million native speakers).
+For preprocessing, we remove the documents that contain less than 20 different words (stubs \footnote{This documents are typically garbage, the titles of the columns in the table, remains of the parsing process, or Wikipedia articles with very little or no information contained in one or two sentences.}) and remove words occurring in less than 50 documents as well as the top 100 most frequent words (in each language separately).
 We represent the documents as normalized TFIDF\cite{Salton88term-weightingapproaches} weighted vectors.
@@
 Furthermore, we remove the documents that contain less than 20 different words and remove words that are too %infrequent as well as the top 100 most frequent words in the vocabularies.
- Although the English language is well aligned with all Wikipedia languages, we must note that quality of alignment varies quite a bit, especially in the case of small Wikipedias.
- Furthermore, we call the document consisting of less than 20 different words, a stub.
-This documents are typically garbage, the titles of the columns in the table, remains of the parsing process, or Wikipedia articles with very little or no information contained in one or two sentences.
-\todo{Move this to footnote for stubs}
-
-The evaluation is based on splitting the data into training and test sets (which are described later).
+ Although the English language is well aligned, number of articles highly correlates with the
+    number of articles connected to English, with all Wikipedia languages, we must note that quality of alignment varies quite a bit, especially in the case of small Wikipedias.
+% \todo{Move this to footnote for stubs}
+
+The evaluation is based on splitting the data into training and test sets. %(which are described later). 
+We select the test set documents as all multi-lingual documents with at least one nonempty alignment from the list: (\emph{hi}, \emph{ht}), (\emph{hi}, \emph{pms}), (\emph{war}, \emph{ht}), (\emph{war}, \emph{pms}).
+This guarantees that we cover all the languages.
+Moreover this test set is suitable for testing the retrieval thorough the hub as the chosen pairs have empty alignments.
+The remaining documents are used for training.
+In Table \ref{table:train_test}, we display the corresponding sizes of training and test documents for each language pair.
+
 On the training set, we perform the two step procedure to obtain the common document representation as a set of mappings $P_i$.
@@
 \begin{table}[h!]
-\caption{Pairwise retrieval, 500 topics;1000 topics}\label{table:retrieval}
+\caption{Pairwise retrieval, 500 topics on the left -- 1000 topics on the right}\label{table:retrieval}
 \begin{center}
@@
 }
-
-We now describe the selection of train and test sets.
+%
+%We now describe the selection of train and test sets.
 We select the test set documents as all multi-lingual documents with at least one nonempty alignment from the list: (\emph{hi}, \emph{ht}), (\emph{hi}, \emph{pms}), (\emph{war}, \emph{ht}), (\emph{war}, \emph{pms}).
@@
 \begin{table}[h!]
-\caption{Pairwise training:test sizes (in thousands)}
+\centering
+\caption{Pairwise training -- test sizes (in thousands) 
+The first row represents the size of the training sets used to construct the mappings in low dimensional language independent space using the English--\emph{en} as a hub.
+The diagonal elements represent number of the unique training documents and test documents in each language.
+}
 \label{table:train_test}
@@
 
-There is also have a problem with the quality of the test data.
+There is also a problem with the quality of the test data.
 For example, if we look at test pair (\emph{war}, \emph{ht}) only 386/534 Waray-Waray test documents are unique but on other side almost all Creole test documents (523/534) are unique.
@@
 To evaluate the prediction accuracy for a given dataset we used 10-fold cross validation.
-In this technique, the dataset is partitioned into 10 folds, where 9 folds are used for training a model and the left-out fold is then used to test the accuracy of the model.
-The learning and testing process is repeated 10 times, each time with a different left-out fold.
+%In this technique, the dataset is partitioned into 10 folds, where 9 folds are used for training a model and the left-out fold is then used to test the accuracy of the model.
+The learning %and testing process is repeated 10 times, each time with a different left-out fold.
 
@@
 The results of the experiment are shown in Table~\ref{table:linkingEvalAlgos}.
-The table shows for each of the algorithms the obtained classification accuracy, precision and recall.
+%The table shows for each of the algorithms the obtained classification accuracy, precision and recall.
  We can see that the most informative features for cluster linking are obtained using the hub CCA algorithm, while $k$-means performs the worse.
@@
 We also compared the proposed approaches on the task of Wikipedia mate retrieval (the same task as in Section~\ref{experiments:hubcca}).
-We computed the Average (over language pairs) Mean Reciprocal Rank (AMRR) \shortcite{voorhees1999trec}  performance of the different approaches on the  Wikipedia data by holding out $15000$ aligned test documents and using $300000$ aligned documents as the training set.
+We computed the Average (over language pairs) Mean Reciprocal Rank (AMRR) \shortcite{voorhees1999trec}  performance of the different approaches on the  Wikipedia data by holding out $15,000$ aligned test documents and using $300,000$ aligned documents as the training set.
 The Figure~\ref{pic:AMRR} shows AMRR score as the function of the number of feature vectors.
@@
 Furthermore, we inspected how the number of features (topics) influences the accuracy of cluster linking.
-As we can see from Table~\ref{table:linkingEvalAlgos} choosing number of features vectors larger than $500$ barely affects linking performance, which is in contrast with the fact that additional features helped improve AMMR, see Figure~\ref{pic:AMRR}.
+As we can see from Table~\ref{table:linkingEvalAlgos} choosing number of feature vectors larger than $500$ barely affects linking performance, which is in contrast with the fact that additional features helped improve AMMR, see Figure~\ref{pic:AMRR}.
 Such differences may have arisen due to different domains of training and testing (Wikipedia pages versus news articles).
@@
 \begin{table}[h]
-\caption{Accuracy of cluster linking with 500 - 800 - 1000 features obtained from different cross-lingual similarity algorithms}
+\caption{Accuracy of cluster linking with 500 - 800 - 1000 topic vectors obtained from different cross-lingual similarity algorithms.
+The table shows for each of the algorithms the obtained classification accuracy, precision and recall.}
 \label{table:linkingEvalAlgos}
@@
   \cline{1-4}
-  Models & CA & Precision & Recall \\ \cline{1-4}
+  Models & A & Precision & Recall \\ \cline{1-4}
   CCA      & 78.2 - 79.6 - 80.3 \% & 76.3 - 78.0 - 80.5 \%  & 81.6 - 82.1 - 79.9 \%  \\ \cline{1-4}
@@
 The similarity pipeline is the most computationally intensive part and currently runs on two Intel Xeon E5-2667 v2, 3.30GHz processor machine with 256GB of RAM.
-This is sufficient to do similarity computation over 150 languages if needed.
+This is sufficient to do similarity computation over 150 languages if needed, see \url{http://xling.ijs.si} for live demo over 100 languages.
 It currently uses Wikipedia as a freely available knowledge base and experiments show that the similarity pipeline dramatically reduces the search space when linking clusters.
@@
 Our data stream is Newsfeed (http://newsfeed.ijs.si/) which provides $430k$ unique articles per day.
-Our system currently computes 400k similarities per 200ms, that means that we compute $16 \cdot 10^{10}$ similarities per day.
+Our system currently computes 2 million similarities per second, that means that we compute $16 \cdot 10^{10}$ similarities per day.
 We
@@
 
-For all linear algebra matrix and vector operations, we use high performance numerical linear algebra libraries as BLAS, OPENBLAS and Intel MKL, which currently allows us to process more than million articles per day.
+For all linear algebra matrix and vector operations, we use high performance numerical linear algebra libraries as BLAS, OPENBLAS and Intel MKL, which currently allows us to process more than one million articles per day.
 In our current implementation, we use the variation of hub approach.
-Our projector matrices are of size $500\times 300000,$ so every projector takes about $1.1$ GB of RAM.
+Our projector matrices are of size $500\times 300,000,$ so every projector takes about $1.1$ GB of RAM.
 Moreover, we need proxy matrices of size $500\times500$ for every language pair.

### 2015-07-20 rupnik2015.tex
updated the pipeline section. added a better image and removed two existing ones. added 3 references

@@
 
-\begin{figure}[tb]
+We base our techniques of cross-lingual event linking on an online system for detection of world events, called Event Registry \shortcite{Leban2014W,Leban2014I}.
+Event Registry is a repository of events, where events are automatically identified by analyzing news articles that are collected from numerous news outlets all over the world.
+The important components in the pipeline of the Event Registry are shown in Figure~\ref{fig:erpipeline} and will now be briefly described.
+
+\begin{figure}[tbp]
 \centering
-\includegraphics[width=0.75\textwidth]{pipeline1}
-\caption{\label{fig:pipeline} Our pipeline - from the Event Registry we obtain a stream of multi-lingual documents and monolingual clusters.
-Our contribution is shown on the right, where using cross-lingual similarity, we link the clusters across languages which correspond to the same event.}
+\includegraphics[width=\textwidth]{pipeline}
+\caption{\label{fig:erpipeline}  The Event Registry pipeline.
+After new articles are collected, they are first analyzed individually (Article-level processing).
+In the next step, groups of articles about the same event are identified and relevant information about the event is extracted (Event construction phase).
+Although the pipeline contains several components, we will in this paper focus only on the two highlighted in the image.}
 \end{figure}
 
-We base our techniques of cross-lingual event linking on an online system for detection of world events, called Event Registry \shortcite{Leban2014W,Leban2014I}.
-Event Registry is a repository of events, where events are automatically identified by analyzing news articles.
-Event Registry works by collecting and analyzing news articles that are published by news outlets in different languages all over the world.
+The collection of the news articles is performed using the Newsfeed service~\cite{Trampus2012}.
+The service monitors RSS feeds of around 100.000 mainstream news outlets available globally.
+Whenever a new article is detected in the RSS feed, the service downloads all available information about the article and sends the article through the pipeline.
+Newsfeed downloads daily on average around 200.000 news articles in various languages, where English, Spanish and German are the most common.
+
 Collected articles are first semantically annotated by identifying in them mentions of relevant concepts -- either entities or important keywords.
-The disambiguated and entity linking of the concepts is done using Wikipedia as the main knowledge base.
-An online clustering algorithm is then applied on the articles in order to identify groups of articles that are discussing the same event.
+The disambiguation and entity linking of the concepts is done using Wikipedia as the main knowledge base.
+The algorithm for semantic annotation is based on~\cite{zhang2014saaacamactat}.
+As a part of the semantic annotation we also analyze the dateline of the article to identify the location of the described event as well as extract dates mentioned in the article using a set of regular expressions.
+Since articles are frequently revised we also detect if the collected article is just a revision of a previous one so that we can use this information in next phases of the pipeline.
+The last important processing step on the document level is to efficiently identify which articles in other available languages are most similar to this article.
+The methodology for this task is one of the main contributions of this paper and is described in details in Section~\ref{sec:crosslingual}).
+
+As a next step, an online clustering algorithm is applied on the articles in order to identify groups of articles that are discussing the same event.
 For each new article, the clustering algorithm determines if the article should be assigned to some existing cluster or into a new cluster.
 The underlying assumption is that articles that are describing the same event are similar enough and will therefore be put into the same cluster.
-
-The processing is done independently for each language and the output is a clustering of articles in each language, with each cluster ideally representing an event, which are defined  topically, geographically and temporally.
-The goal of this paper is to present approaches to connecting the clusters/events across languages.
-
-Our approach is based on first defining a similarity function between articles in different languages (Section~\ref{sec:models}) and then we present several efficient approaches to linking the clusters based on features (Section~\ref{sec:linking}).
-The pipeline is shown in Figure~\ref{fig:pipeline}.
-First, however we describe the pre-processing on the input articles which we then take as input.
-
-The system is designed to handle streams and so we describe how a document which enters the stream is processed.
-The sequence of operations is shown in  Figure~\ref{fig:erpipeline}.
-A new document is first tokenized, stop words are removed and words are stemmed.
- What remains is represented in a vector-space model and normalized using $TFIDF$ (see Section~\ref{sec:tfidf} for the definition).
-Since each article is tagged with its language, cosine similarity  is used to find the most similar existing cluster, by comparing it to the centroid vector of each cluster.
-A selected threshold is used to determine if the article is not similar to any existing clusters (0.4 was used in our experiments).
-The article is then assigned to the corresponding cluster, otherwise a new cluster is created, initially containing only the single article.
-
+For clustering, each new article is first tokenized, stop words are removed and words are stemmed.
+The remaining tokens are represented in a vector-space model and normalized using TF-IDF (see Section~\ref{sec:tfidf} for the definition).
+The cosine similarity is used to find the most similar existing cluster, by comparing the document's vector to the centroid vector of each cluster.
+A user-defined threshold is used to determine if the article is not similar enough to any existing clusters (0.4 was used in our experiments).
+If the highest similarity is above the threshold, the article is assigned to the corresponding cluster, otherwise a new cluster is created, initially containing only the single article.
 Since articles about an event are commonly written only for a short period of time, we remove clusters once the oldest article in the cluster becomes more than 4 days old.
 This housekeeping mechanism prevents the clustering from becoming slow and also ensures that articles are not assigned to obsolete clusters.
-
-The clusters, we consider must have a sufficient number of articles (which is a language dependent parameter), which is in line with our notion of an event.
-Once an event is identified (i.e.
-a cluster of sufficient size), a new unique ID is assigned to it and the main information about the event is then automatically extracted by analyzing the articles assigned to it.
-The extracted information includes properties such as the date of the event, the location, who is involved in it, what the event is about, etc.
-
- The system is described in more detail elsewhere \shortcite{Leban2014W,Leban2014I}.
-
-\begin{figure}[tbp]
-\centering
-\includegraphics[width=\textwidth]{ev_pipeline}
-\caption{\label{fig:erpipeline}  The Event Registry pipeline, after obtaining the raw stream of multi-lingual news articles, it first cleans and processes articles individually.
-It then aggregates and clusters the articles into similar events (within each language) and finally provides an interface allowing us to query for articles.}
-\end{figure}
-
-\begin{comment}
-In short, the article clustering works as follows (details are available in \cite{}).
-First we create a new document by joining the title and the body of the article.
-We then tokenize the document, remove the stop words and stem the words.
-The remaining words are represented as a vector in the vector-space model.
-Each value in the vector is additionally normalized using TF-IDF.
-After normalization, we use the vector to find the most similar existing cluster.
-To be able to compare the vector with a cluster we first need a vector representation for each of the clusters.
-This is done by computing a centroid vector of the cluster by simply averaging the vectors of the articles in the cluster.
-By computing the cosine similarity of the article vector with each of the centroid vectors we then find the cluster with the highest similarity.
-If the highest similarity is above the selected threshold (we used 0.4 in our experiments),
-
-In order to treat the cluster as describing an event in Event Registry,
-
-The clustering of articles is done for each language separately -- English articles are, for example, clustered separately from Spanish articles.
-When events are identified, they will therefore be identified from clusters in each language separately.
-As a result, the same event will be represented in Event Registry with two or more event IDs, and each event will only contain coverage in a single language.
-To avoid this problem we want to exploit the available information about the similarity of the articles in a cluster with articles in other languages.
-Using this information we hope to be able to connect clusters in different languages and represent them using a single event ID.
-\end{comment}
-%Event registry $\Rightarrow$ Monolingual events $\Rightarrow$ features (semantic + CCA) $\Rightarrow$ linked events (crosslingual)
+Details of the clustering approach are described in~\cite{brank2014}.
+
+Once the number of articles in a cluster reaches a threshold (which is a language dependent parameter), we assume that the articles in the cluster are describing an event.
+At that point, a new event with a unique ID is created in Event Registry, and the cluster with the articles is assigned to it.
+By analyzing the articles we extract the main information about the event, such as the event location, date, most relevant entities and keywords, etc.
+
+Since articles in a cluster are in a single language, we also want to identify any other existing clusters that are reporting about the same event in another language and join these clusters into the same event.
+This task is performed using a classification approach that is the second major contribution of this paper.
+It is described in details in Section~\ref{sec:linking}.
+
+When a cluster is identified and information about the event is extracted, all available data is stored in a custom-built database system.
+The data is then accessible through the API or a web interface\footnote{http://eventregistry.org/}, which provide numerous search and visualization options.
 
@@
 
-\section{Cross-lingual Document Similarity}\label{sec:crosslingual}
+\section{Cross-lingual Document Similarity}
 \label{sec:crosslingual}

### 2015-07-21 rupnik2015.tex
fixes to the discussion and future work

@@
 If the highest similarity is above the threshold, the article is assigned to the corresponding cluster, otherwise a new cluster is created, initially containing only the single article.
+Whenever an article is assigned to a cluster, the cluster's centroid vector is also updated.
 Since articles about an event are commonly written only for a short period of time, we remove clusters once the oldest article in the cluster becomes more than 4 days old.
@@
 \section{Discussion and Future Work}
+
 In this paper we have presented a cross-lingual system for linking events in different languages.
@@
 
-Finally, we discuss some aspects of the methods and the system as a whole.
-\begin{itemize}
-\item \textbf{LSI and hub CCA} - We first comment on the empirical validation of LSI and hub CCA.
 On the task of mate retrieval we observe that refining the LSI based projections with hub CCA leads to improved retrieval precision, but the methods perform comparably on the task of event linking.
@@
 The interpretation is that the linking features are highly aggregated for large clusters, which compensates the lower per-document precision of LSI.
-Another possible reason is that the advantage that we show on Wikipedia is lost on the news domain - which suggest that domain adaptation should be explored.
-
-\item \textbf{Features} - The experiments show that the hub CCA based features present good baseline, which can greatly benefit from additional semantic extraction based features.
-Even though the CCA based features do not greatly improve the system performance when the semantic features are used, there are two main benefits in the approach: the linking process can be sped up (smaller candidate cluster sets), and usage in the case when semantic extraction is not possible.
-
-\item \textbf{Future work} - Currently the system is loosely coupled - the language component is built independently of the rest of the system, in particular the linking component.
+Another possible reason is that the advantage that we show on Wikipedia is lost on the news domain.
+This hypothesis could be validated by testing the approach on documents from a different domain.
+
+The experiments show that the hub CCA based features present good baseline, which can greatly benefit from additional semantic-based features.
+Even though the CCA-based features do not greatly improve the system performance when the semantic features are used, there are two main benefits in the approach: the linking process can be sped up (smaller set of candidate clusters), and usage in the case when semantic extraction is not possible.
+
+\subsection{Future Work}
+
+Currently the system is loosely coupled -- the language component is built independently of the rest of the system, in particular the linking component.
 It is possible that better embeddings can be obtained by methods that jointly optimize a classification task and the embedding.
@@
 Another point of interest is to evaluate the system on languages with scarce linguistic resources, where semantic annotation might not be available.
-For this reason, the labelled dataset of linked clusters should to be extended first.
+For this purpose, the labelled dataset of linked clusters should be extended first.
 The mate retrieval evaluation showed that even for language pairs with no training set overlap, the hub CCA recovers some signal.
 
-\end{itemize}
-
-%\subsection{Future Work}
+In order to further improve the performance of the classifier for cluster linking, additional features should also be extracted from articles and clusters and checked if they can increase the accuracy of the classification.
+Since the amount of linguistic resources vary significantly from language to language it would also make sense to build a separate classifier for each language pair.
+Intuitively, this should improve performance since weights of individual learning features could be adapted to the tested pair of languages.
 

### 2015-07-22 rupnik2015.tex
added table for accuracies for different language pairs + paragraph with the description added F1 score added standard deviations added some references

@@
 
-\primoz{we may want to move this out to the pipeline section}
 \subsection{System Requirements}
+\label{sec:sysreq}
 
@@
 
-\subsection{Related work}
-
-%In this section we will relate the method we use to other alternatives in the literature.
-We will first state some requirements for our system, which motivate our choice of focusing on a particular subset of approaches in the rest of the paper.
-
-The goal is to build a system that monitors global media and analyzes how events are being reported on.
-In our approach this boils down to two steps: tracking events separately in each language (based on language detection and an online clustering approach) and then connecting them.
-The pipeline needs to process millions of articles per day and perform billions of similarity computations each day.
-The system should support as many languages as possible.
-We focus on implementations that run on a single shared memory machine, as opposed to clusters of machines.
-This severely simplifies the software implementation and system maintenance.
-To summarize, the following properties are desireable:
-\begin{itemize}
-\item \textbf{Training} - The training (building corss-lingual models) should scale to many languages and should be robust to the quality of training resources.
-The system should be able to take advantage of comparable corpus (as opposed to parallel translation based corpora), with missing data.
-\primoz{We could say something more subject to moving this section}%Supporting a new language should be straightforward.
-\item \textbf{Efficiency} - The similarity computation should be fast - the system must be able to handle billions of similarity computations per day.
-\item \textbf{Cost} - The system should run on a strong shared machine server and not rely on paid services.
-\item \textbf{Implementation} - The system is simple to implement, with few parameters to tune.
-\end{itemize}
-We believe that a cross-lingual similarity component that meets such requirements is very desirable in a commercial setting, where several different costs have to be taken into consideration.
+\section{Related work}
+
+In this section we will relate the methods we use to other alternatives in the literature.
+Since there are two distinctive tasks that we tackle in this paper (computing cross-lingual document similarity and cross-lingual cluster linking) we have separated the related work into two parts.
+
+\subsection{Cross-lingual document similarity}
 
 There are four main families of approaches to cross-lingual similarity.
-\begin{itemize}
-\item \textbf{Translation} - The most obvious way to compare documents written in different languages is to use machine translation and perform monolingual similarity.
+
+\textbf{Translation}.
+The most obvious way to compare documents written in different languages is to use machine translation and perform monolingual similarity.
 One can use free tools such as Moses \shortcite{moses} or translation services, such as Google Translate\footnote{https://translate.google.com/}.
@@
 The operation efficiency and cost requirements make translation based approaches less suited for our system.
-\item \textbf{Probabilistic topic models} -  There exist many variants to modelling documents in a language independent way by using probabilistic graphical models.
+
+\textbf{Probabilistic topic models}.
+There exist many variants to modelling documents in a language independent way by using probabilistic graphical models.
 The models include:  Joint Probabilistic Latent Semantic Analysis (JPLSA) \shortcite{platt2010translingual}, Coupled Probabilistic LSA (CPLSA) \cite{platt2010translingual}, Probabilistic Cross-Lingual LSA (PCLLSA)\cite{PCL_LSA} and Polylingual Topic Models (PLTM) \shortcite{polyLDA} which is a Bayesian version of PCLLSA.
@@
 Furthermore, representing a new document as a mixture of topics is another potentially hard inference problem which must be solved.
-\item \textbf{Classification problem related} - In approaches such as \cite{cotrain} and \shortcite{svm2k} the authors find linear embeddings of the data that support good classification results, based on a given classification problem training data.
+
+\textbf{Classification problem related}.
+In approaches such as \cite{cotrain} and \shortcite{svm2k} the authors find linear embeddings of the data that support good classification results, based on a given classification problem training data.
 We do not consider this family of approaches, since for general language pairs classifier training sets may not be available.
-\item \textbf{Matrix factorization} - Non-negative matrix factorization based \cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CLLSI\shortcite{cl_lsi}, Canonical Correlation Analysis (CCA) \cite{Hotelling}, Oriented Principal Component Analysis \cite{platt2010translingual}.
+
+\textbf{Matrix factorization}.
+Non-negative matrix factorization based \cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CLLSI\shortcite{cl_lsi}, Canonical Correlation Analysis (CCA) \cite{Hotelling}, Oriented Principal Component Analysis \cite{platt2010translingual}.
 The quadratic time and space dependency of the OPCA method makes it impractical for large scale purposes.
@@
 
-\end{itemize}
-
-Based on our requirements in Section~\ref{} \primoz{System requirements}, we chose to focus on methods based on vector space models and linear embeddings.
+Based on our requirements in Section~\ref{sec:sysreq}, we chose to focus on methods based on vector space models and linear embeddings.
 We propose a method that is more efficient than popular alternatives (a clustering based approach and latent semantic indexing), but is still simple to optimize and use.
 
-\subsection{Related work}
+\subsection{Cross-lingual cluster linking}
 
@@
 News are able to identify clusters of articles about same event, but they offer no linking of clusters across languages.
-The only service that we found, which provides cross-lingual cluster linking, is the European Media Monitor (EMM)~\cite{Steinberger2008}.
+The only service that we found, which provides cross-lingual cluster linking, is the European Media Monitor (EMM)~\cite{pouliquen2008story,Steinberger2008}.
 EMM clusters articles in 60 languages and then tries to determine which clusters of articles in different languages describe the same event.
@@
 If the the maximal singular value of $S_{i,j}$ is bounded by $1$, then the similarity scores will lie on the interval $[-1, 1]$.
-In section \ref{sec:models} we will describe some approaches to computing $S_{i,j}$ given training data.
+In Section \ref{sec:models} we will describe some approaches to computing $S_{i,j}$ given training data.
 
@@
 \subsection{Training a classification model}
+\label{trainClassifier}
 
@@
 
-\item {\bf Concept related features}.
+\item {\bf Concept-related features}.
 Articles that are imported into Event Registry are first semantically annotated by linking mentioned entities and keywords to the corresponding Wikipedia pages.
@@
 As a training set, we used common Wikipedia alignment for all three languages.
-To test which of these algorithms performed best, we performed the following test.
-Using each algorithm, we analyzed all articles in Event Registry to find the most similar articles in other languages.
-The values of the cross-lingual article linking features (\texttt{linkCount} and \texttt{avgSimScore}) were then recomputed for all annotated learning examples.
-A dataset with only these two features and the label was constructed and used to test how accurate prediction models can be induced using the data.
-The results of the experiment are shown in Table~\ref{table:linkingEvalAlgos}.
+To test which of these algorithms performed best, we made the following test.
+For each of the three algorithms, we analyzed all articles in Event Registry and for each article computed the most similar articles in other languages.
+To test how informative the identified similar articles are for cluster linking we then trained three classifiers as described in Section~\ref{trainClassifier} -- one for each algorithm.
+Each classifier was allowed to use as learning features only the cross-lingual article linking features for which values are determined based on the selected algorithm ($k$-means, LSI and hub CCA).
+The results of the trained models are shown in Table~\ref{table:linkingEvalAlgos}.
 %The table shows for each of the algorithms the obtained classification accuracy, precision and recall.
@@
 % cluster size?
-We also analyzed how cluster size influences the linking performance.
-The intuition is that clusters with large number of documents are easier to link, which can be seen in Table~\ref{table:linkingEvalAlgosLargeSmall}.
-Large clusters aggregate a lot of information, so the performance of the similarity model is not as important so long as we assign articles to correct clusters.
-In the case of small clusters, we expect the accuracy to matter more.
-Our experiments indicate that this is indeed the case: it is harder to link small clusters and better retrieval accuracy may matter more.
-Hub CCA outperforms other methods in this sense in terms of precision, which is consistent with its AMRR performance, see Figure~\ref{pic:AMRR}.
+We also analyzed how cluster size influences the accuracy of cluster linking.
+We would expect that if the tested pair of clusters has a larger number of articles then the classifier should be able to more accurately predict whether the clusters should be linked or not.
+The reasoning is that the large clusters would provide more document linking information (more articles mean more links to other similar articles) as well as more accurately aggregated semantic information.
+In the case of smaller clusters, the errors of the similarity models have greater impact which should decrease the performance of the classifier, too.
+To validate this hypothesis we have split the learning examples into two datasets -- one containing cluster pairs where the combined number of articles from both clusters is below 20 and one dataset where the combined number is 20 or more.
+The results of the experiment can be seen in Table~\ref{table:linkingEvalAlgosLargeSmall}.
+As it can be seen, the results confirm our expectations: for smaller clusters it is indeed harder to correctly predict if the cluster pair should be merged or not.
+\todo{Andrej: A lahko poeditiras naslednje stavke - niso zelo jasno napisani.} Hub CCA outperforms other methods in this sense in terms of precision, which is consistent with its AMRR performance, see Figure~\ref{pic:AMRR}.
 It is worth noting that added accuracy is also beneficial when we wish to choose good representatives of clusters. Therefore, we decided to use hub CCA in the cluster linking component.
@@
 \begin{table}[h]
-\caption{Accuracy of cluster linking using $500$ features on two datasets containing large (left number) and small (right number) clusters}
+\caption{Accuracy of cluster linking using $500$ topic vectors on two datasets containing large (left number) and small (right number) clusters.
+The dataset with small clusters contained the subset of learning examples in which the combined number of articles from both clusters of the cluster pair were below 20.
+The remaining learning examples were put into the dataset of large clusters.}
 \label{table:linkingEvalAlgosLargeSmall}
@@
   \cline{1-4}
-  Models & CA & Precision & Recall \\ \cline{1-4}
+  Models & A & Precision & Recall \\ \cline{1-4}
   CCA      & 81.2 - 77.8 \% & 80.5 - 74.5 \%  & 91.3 - 57.5 \%  \\ \cline{1-4}
@@
 The results of the evaluation are shown in Table~\ref{table:linkingEval}.
-We can see that using a single group of features, the highest prediction accuracy can be achieved using  concept related features.
-The classification accuracy in this case is 88.8\%.
-By additionally including also the cross-lingual article linking features, the classification accuracy rises slightly to 89.2\%.
-Using all three groups of features, the achieved accuracy is 89.6\%.
-
-Based on the results, we can make the following conclusions.
+We can see that using a single group of features, the highest prediction accuracy can be achieved using  concept-related features.
+The classification accuracy in this case is 88.5\%.
+By additionally including also the cross-lingual article linking features, the classification accuracy rises slightly to 89.4\%.
+Using all three groups of features, the achieved accuracy is 89.2\%.
+
+To test if the accuracy of the predictions is language dependent we have also performed the evaluations separately on individual language pairs.
+For this experiment we have split the annotated learning examples into three datasets, where each dataset contained only examples for one language pair.
+When training the classifier all three groups of features were available.
+The results are shown in Table~\ref{langPairEval}.
+We can see that the highest accuracy can be obtained when we need to decide if an English and German cluster represent the same event.
+A possible interpretation for these results is that this language pair has the most language resources (the highest number of Wikipedia pages) which results in the most accurate cross-lingual article similarity measure as well as the most extensive semantic annotation of the articles.
+
+Based on the performed experiments, we can make the following conclusions.
 The cross-lingual similarity algorithms provide valuable information that can be used to identify clusters that describe the same event in different languages.
-The computed features are however significantly less informative compared to the features computed on the annotated concepts.
+For the task of cluster linking, the cross-lingual article linking features are however significantly less informative compared to the concept-related features that are extracted from the semantic annotations.
 Nevertheless, the cross-lingual article similarity features are very important for two reasons.
@@
 The second reason these features are important is that concept annotations are not available for all articles as the annotation of news articles is computationally intensive and can only be done for a subset of collected articles.
+The prediction accuracies for individual language pairs are comparable although it seems that the achievable accuracy correlates with the amount of available language resources.
 
@@
 \begin{table}[h]
-\caption{Accuracy of story linking with different sets of features}
+\caption{The accuracy of the classifier for story linking using different sets of learning features.}
 \label{table:linkingEval}
 \begin{center}
-\begin{tabular}{|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|c|}
   \hline
-  \cline{1-4}
-  Features & CA & Precision & Recall \\ \cline{1-4}
-  CCA & 78.2\% & 0.763 & 0.816 \\ \cline{1-4}
-  Concepts & 88.8\% & 0.884 & 0.891 \\ \cline{1-4}
-  Misc & 65.3\% & 0.718 & 0.500 \\ \cline{1-4}
-  CCA + concepts & 89.2\% & 0.891 & 0.893 \\ \cline{1-4}
-  All & 89.6\% & 0.895 & 0.895 \\ \cline{1-4}
+  \cline{1-5}
+  Features & A & Precision & Recall & F1 \\ \cline{1-5}
+  CCA & $78.3\% \pm 5.9\%$ & $0.782 \pm 0.070$ & $0.789 \pm 0.052$ & $0.784 \pm 0.055$ \\ \cline{1-5}
+  Concepts & $88.5\% \pm 2.7\%$ & $0.886 \pm 0.048$ & $0.886 \pm 0.022$ & $0.885 \pm 0.024$ \\ \cline{1-5}
+  Misc & $54.8\% \pm 6.7\%$ & $0.618 \pm 0.165$ & $0.582 \pm 0.302$ & $0.524 \pm 0.130$ \\ \cline{1-5}
+  CCA + Concepts & $89.4\% \pm 2.5\%$ & $0.894 \pm 0.046$ & $0.896 \pm 0.024$ & $0.894 \pm 0.023$ \\ \cline{1-5}
+  CCA + Misc & $78.8\% \pm 5.0\%$ & $0.789 \pm 0.071$ & $0.794 \pm 0.046$ & $0.790 \pm 0.045$ \\ \cline{1-5}
+  Concepts + Misc & $88.7\% \pm 2.6\%$ & $0.888 \pm 0.046$ & $0.888 \pm 0.022$ & $0.887 \pm 0.023$ \\ \cline{1-5}
+  All & $89.2\% \pm 2.6\%$ & $0.888 \pm 0.049$ & $0.901 \pm 0.019$ & $0.893 \pm 0.023$ \\ \cline{1-5}
+  \hline
+\end{tabular}
+\end{center}
+\end{table}
+
+\begin{table}[h]
+\caption{The accuracy of the classifier for story linking on training data for each language pair separately using all learning features.}
+\label{table:langPairEval}
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|}
+  \hline
+  \cline{1-5}
+  Language pair & A & Precision & Recall & F1 \\ \cline{1-5}
+  en, de & $91.8\% \pm 5.5\%$ & $0.917 \pm 0.063$ & $0.937 \pm 0.063$ & $0.925 \pm 0.051$ \\ \cline{1-5}
+  en, es & $87.7\% \pm 5.4\%$ & $0.877 \pm 0.074$ & $0.885 \pm 0.098$ & $0.876 \pm 0.059$ \\ \cline{1-5}
+  es, de & $88.6\% \pm 4.3\%$ & $0.897 \pm 0.091$ & $0.843 \pm 0.119$ & $0.859 \pm 0.060$ \\ \cline{1-5}
   \hline
@@
 
-Currently the system is loosely coupled -- the language component is built independently of the rest of the system, in particular the linking component.
+Currently the system is loosely-coupled -- the language component is built independently of the rest of the system, in particular the linking component.
 It is possible that better embeddings can be obtained by methods that jointly optimize a classification task and the embedding.

### 2015-07-22 results/accuracyForLanguagePairs.py
added scripts and results reported in the paper

@@ -1,0 +1,88 @@
+"""
+evaluate accuracy of the story linking module. the similar articles are computed using CCA with 500 feature vector
+evaluations are done using all learning features but separately for each language pair for languages eng, deu and spa
+for each test we print the accuracy, precision, recall, and f1
+for the reviewers we've also added computation of standard deviation of the scores
+"""
+import os, sys, json, Orange, numpy, orngStat
+
+fname = r"e:\Data\JAIR-Data\2015-03-23\LrnStoryLinkingFtrVec - sim_cca_with_500.tab"
+
+data = Orange.data.Table(fname)
+
+print "original data size: %d" % len(data)
+
+def validExample(ex):
+    langs = ex["Language pair"]
+    pair = langs.value.split("-")
+    if pair[0] == pair[1]:
+        return False
+    if pair[0] in ["eng", "deu", "spa"] and pair[1] in ["eng", "deu", "spa"]:
+        return True
+    return False
+
+def validLangPair(ex, lang1, lang2):
+    langs = ex["Language pair"]
+    pair = langs.value.split("-")
+    if (pair[0] == lang1 and pair[1] == lang2) or (pair[1] == lang1 and pair[0] == lang2):
+        return True
+    return False
+
+# remove instances where the language pair is between the same language
+data = Orange.data.Table([d for d in data if validExample(d)])
+
+print "data after removing same langs: %d" % len(data)
+
+cca = ['CCAMatchCount', 'CCAArtMatchRatio', 'CCAAvgMatchWgt', 'CCAAvgCosSim']
+semantic = ["ConceptCosSim", "ConceptJaccardSim", "EntityCosSim", "EntityJaccardSim", "KeywordJaccardSim", "KeywordCosSim"]
+misc = ["TimeDiff", "HasSameLocation", "MatchingDateRatio", "MatchingDateCount", "ExtractedDateJaccardSim", "CoreTimeDiff"]
+
+attrs = cca + semantic + misc
+attributes = [attr for attr in data.domain.features if attr.name in attrs]
+new_domain = Orange.data.Domain(attributes, data.domain.class_var)
+
+classIndex = list(data.domain.classVar.values).index("yes")
+
+tests = [
+    (("eng", "deu"), "eng-deu"),
+    (("eng", "spa"), "eng-spa"),
+    (("spa", "deu"), "spa-deu")]
+
+for (pair, label) in tests:    
[…]

### 2015-07-22 results/accuracyUsingDifferentFeatures.py

@@ -1,0 +1,78 @@
+"""
+evaluate accuracy of the story linking module. the similar articles are computed using CCA with 500 feature vector
+evaluations are done using different sets of learning features
+for each combination of features we print the accuracy, precision, recall, and f1
+for the reviewers we've also added computation of standard deviation of the scores
+"""
+import os, sys, json, Orange, numpy, orngStat
+
+fname = r"e:\Data\JAIR-Data\2015-03-23\LrnStoryLinkingFtrVec - sim_cca_with_500.tab"
+
+data = Orange.data.Table(fname)
+
+print "original data size: %d" % len(data)
+
+def validExample(ex):
+    langs = ex["Language pair"]
+    pair = langs.value.split("-")
+    if pair[0] == pair[1]:
+        return False
+    if pair[0] in ["eng", "deu", "spa"] and pair[1] in ["eng", "deu", "spa"]:
+        return True
+    return False
+
+# remove instances where the language pair is between the same language
+data = Orange.data.Table([d for d in data if validExample(d)])
+
+print "data after removing same langs: %d" % len(data)
+
+cca = ['CCAMatchCount', 'CCAArtMatchRatio', 'CCAAvgMatchWgt', 'CCAAvgCosSim']
+semantic = ["ConceptCosSim", "ConceptJaccardSim", "EntityCosSim", "EntityJaccardSim", "KeywordJaccardSim", "KeywordCosSim"]
+misc = ["TimeDiff", "HasSameLocation", "MatchingDateRatio", "MatchingDateCount", "ExtractedDateJaccardSim", "CoreTimeDiff"]
+
+tests = [
+    (cca, "cca"), (semantic, "semantic"), (misc, "misc"),
+    (cca + semantic, "cca + semantic"), (cca + misc, "cca + misc"), (semantic + misc, "semantic + misc"),
+    (cca + semantic + misc, "cca + semantic + misc")]
+
+for test in tests:
+    attributes = [attr for attr in data.domain.features if attr.name in test[0]]
+    new_domain = Orange.data.Domain(attributes, data.domain.class_var)
+    new_data = Orange.data.Table(new_domain, data)
+    classIndex = list(new_data.domain.classVar.values).index("yes")
+
+    svm = Orange.classification.svm.SVMLearner()
+    res = Orange.evaluation.testing.cross_validation([svm], new_data, folds=10)
[…]

### 2015-08-06 rupnik2015.tex
complete pass

@@
 \begin{abstract}
-Today we follow news which is distributed globally.
+
+In today's world, we follow news which is distributed globally.
 Significant events are reported by different sources and in different languages.
 In this work, we address the problem of tracking and events in a large multilingual stream.
-We consider a particular aspect of this problem, namely how to link collections of articles in different languages which refer to the same event.
-Given a multi-lingual stream and clusters of articles from each language, we first propose a method for cross-lingual document similarity based on Wikipedia, which enables us to compute the similarity of any two articles regardless of language.
+Within a recently developed system (EventRegistry  \shortcite{Leban2014W,Leban2014I}), we examine two aspects of this problem: how to compare articles in different languages and how to link collections of articles in different languages which refer to the same event.
+Taking a multi-lingual stream and clusters of articles from each language, we compare different cross-lingual document similarity measures based on Wikipedia.
+This allows us to compute the similarity of any two articles regardless of language.
+Building on previous work, we show there are methods which scale well and can compute a meaningful similarity between articles from languages with little or no direct overlap in the training data.
+Using this capability, we then propose an approach to link clusters of articles across languages which represent the same event.
+We provide an extensive evaluation of the system as a whole, as well as an evaluation of the quality and robustness of the similarity measure and the linking algorithm.
+
+%Given a multi-lingual stream and clusters of articles from each language, we first propose a method for cross-lingual document similarity based on Wikipedia, which enables us to compute the similarity of any two articles regardless of language.
 The proposed method can scale to 100 languages and can match articles from languages with little or no direct overlap in the training data.
@@
 Content on the internet is becoming increasingly multi-lingual.
-A prime example is Wikipedia - in 2001 the majority of pages were written in English, while in 2015, the percentage of English articles has dropped to 14\%.
-Machine translation remains relatively rudimentary - allowing people to understand simple phrases on web pages, but remain inadequate for more advanced understanding of text.
+A prime example is Wikipedia - in 2001, the majority of pages were written in English, while in 2015, the percentage of English articles has dropped to 14\%.
 At the same time, online news has begun to dominate reporting of current events.
+However, machine translation remains relatively rudimentary - allowing people to understand simple phrases on web pages, but remain inadequate for more advanced understanding of text.
 In this paper we consider the intersection of these developments: how to track events which are reported about in multiple languages.
@@
 Event Registry is a repository of events, where events are automatically identified by analyzing news articles that are collected from numerous news outlets all over the world.
-The important components in the pipeline of the Event Registry are shown in Figure~\ref{fig:erpipeline} and will now be briefly described.
+The important components in the pipeline of the Event Registry are shown in Figure~\ref{fig:erpipeline} and are briefly described below.
 
@@
 The collection of the news articles is performed using the Newsfeed service~\cite{Trampus2012}.
-The service monitors RSS feeds of around 100.000 mainstream news outlets available globally.
+The service monitors RSS feeds of around 100,000 mainstream news outlets available globally.
 Whenever a new article is detected in the RSS feed, the service downloads all available information about the article and sends the article through the pipeline.
-Newsfeed downloads daily on average around 200.000 news articles in various languages, where English, Spanish and German are the most common.
+Newsfeed downloads daily on average around 200,000 news articles in various languages, where English, Spanish and German are the most common.
 
@@
 
-As a next step, an online clustering algorithm is applied on the articles in order to identify groups of articles that are discussing the same event.
+As a next step, an online clustering algorithm is applied to the articles in order to identify groups of articles that are discussing the same event.
 For each new article, the clustering algorithm determines if the article should be assigned to some existing cluster or into a new cluster.
 The underlying assumption is that articles that are describing the same event are similar enough and will therefore be put into the same cluster.
-For clustering, each new article is first tokenized, stop words are removed and words are stemmed.
+For clustering, each new article is first tokenized, stop words removed and the remaining words are stemmed.
 The remaining tokens are represented in a vector-space model and normalized using TF-IDF (see Section~\ref{sec:tfidf} for the definition).
-The cosine similarity is used to find the most similar existing cluster, by comparing the document's vector to the centroid vector of each cluster.
+Cosine similarity is used to find the most similar existing cluster, by comparing the document's vector to the centroid vector of each cluster.
 A user-defined threshold is used to determine if the article is not similar enough to any existing clusters (0.4 was used in our experiments).
@@
 At that point, a new event with a unique ID is created in Event Registry, and the cluster with the articles is assigned to it.
-By analyzing the articles we extract the main information about the event, such as the event location, date, most relevant entities and keywords, etc.
-
-Since articles in a cluster are in a single language, we also want to identify any other existing clusters that are reporting about the same event in another language and join these clusters into the same event.
+By analyzing the articles, we extract the main information about the event, such as the event location, date, most relevant entities and keywords, etc.
+
+Since articles in a cluster are in a single language, we also want to identify any other existing clusters that report about the same event in other languages and join these clusters into the same event.
 This task is performed using a classification approach that is the second major contribution of this paper.
-It is described in details in Section~\ref{sec:linking}.
+It is described in detail in Section~\ref{sec:linking}.
 
@@
 Our approach consists of two steps: tracking events separately in each language (based on language detection and an online clustering approach) and then connecting them.
-The pipeline must be able to process millions of articles per day and perform billions of similarity computations each day - implying that similarity computation should be highly scalable.
-%The system should support as many languages as possible.
- We focus on implementations that run on a single shared memory machine, as opposed to clusters of machines. This  simplifies implementation and system maintenance.
+The pipeline must be able to process millions of articles per day and perform billions of similarity computations each day.
+Both step rely heavily on simliarity computation, which implies that this must be highly scalable.
+%
+%while both steps must be scalable, the underlying similarity computation is used 
+%- implying that similarity computation should be highly scalable.  %The system should support as many languages as possible.
+Therefore, we focus on implementations that run on a single shared memory machine, as opposed to clusters of machines. This  simplifies implementation and system maintenance.
 To summarize, the following properties are desirable:
@@
 In approaches such as \cite{cotrain} and \shortcite{svm2k} the authors find linear embeddings of the data that support good classification results, based on a given classification problem training data.
-We do not consider this family of approaches, since for general language pairs classifier training sets may not be available.
+We do not consider this family of approaches, since for general language pairs classifier training sets may not be available.\todo{add more here}
 
@@
 For our setting, the method in \cite{nonnegfactor_lsi} has a prohibitively high computational cost when building models (it uses dense matrices of dimensions that is the product of training set size and vocabulary size).
-Our proposed approach combines CCA and CLLSI, which will be presented in more detail.
+Our proposed approach combines CCA and CLLSI.
+
+\todo{add the related work which was missing}
 
@@
 Google News as well as Yahoo!
-News are able to identify clusters of articles about same event, but they offer no linking of clusters across languages.
+News are able to identify clusters of articles about same event, but offer no linking of clusters across languages.
 The only service that we found, which provides cross-lingual cluster linking, is the European Media Monitor (EMM)~\cite{pouliquen2008story,Steinberger2008}.
@@
 All information extraction is then done on the translated article.
+\todo{add the missing related work}
 
@@
 Document similarity is an important component in techniques from text mining and natural language processing.
-Many techniques use the similarity as a black box, i.e., a kernel in Support Vector Machines.
+Many techniques use the similarity as a black box, e.g., a kernel in Support Vector Machines.
 Comparison of documents (or other types of text snippets) in a monolingual setting is a well studied problem in the field of information retrieval \cite{Salton88term-weightingapproaches}.
-We first recount some system requirements, then review  related approaches to cross-lingual similarity models and then formally introduce the problem as well as our approach.
+We first formally introduce the problem followed by our approach.
 
@@
 We will first describe how documents are represented as vectors and how to compare douments in a mono-lingual setting.
-We will then define a way to measure cross-lingual similarity that is natural for the models we consider.
-
-\textbf{Document representation.}
+We then define a way to measure cross-lingual similarity that is natural for the models we consider.
+
+\noindent\textbf{Document representation.}
 The standard vector space model \cite{Salton88term-weightingapproaches} represents documents as vectors, where each term corresponds to word or phrase in a fixed vocabulary.
@@
 An \emph{Inverse Document Frequency} ($IDF$) weight for the dictionary term $k$ is defined as $\log\left( \frac{N}{DF_k} \right)$, where $DF_k$ is the number of documents in the corpus which contain term $k$.
-Therefore we can definte a document's $TFIDF$ as
+\todo{our corpus is dynamic how do we deal with this?} Therefore we can definte a document's $TFIDF$ as
 $$ x_{ij}  := \frac{\mbox{term frequency in document } i}{\mbox{inverse document frequency of term } j}.$$
@@
 
-\textbf {Mono-lingual similarity.}
+\noindent\textbf {Mono-lingual similarity.}
 A common way of computing similarity between documents is \emph{cosine similarity},
@@
 Cosine similarity, and other related approaches, assumes that the similarity is reflected in the overlap of words.
-This metric is only meaningful when the documents $d_1$ and $d_2$ are defined on the same space (e.g.
-have a comparable vocabulary or are written in the same language).
-
-\textbf {Cross-lingual similarity.}
+While always defined, this metric is only meaningful when the documents $d_1$ and $d_2$ are defined on the same space (e.g.
+have a comparable vocabulary or are written in the same language, since otherwise their supports are disjoint).
+
+\noindent\textbf {Cross-lingual similarity.}
 Processing a multilingual dataset results in several vector spaces with varying dimensionality, one for each language.
@@
 In Section \ref{sec:models} we will describe some approaches to computing $S_{i,j}$ given training data.
+\todo{add some more details/intuition.}
 
@@
 approach called Canonical Correlation Analysis (CCA) in \ref{sec:CCA} and finally a new, more generally applicable method, which is an extension of CCA to more than two languages in \ref{sec:hublang}.
-
+%
 CCA can be used to find
@@
 The method takes advantage of the hub language assumption in the training dataset (the English Wikipedia corresponds to a hub) as well as a preprocessing step based on LSI that acts as regularization and improves scalability.
+\todo{we need to set this up better}
 
@@
 Wikipedia is an example of a comparable corpus, where a specific entry can be described in multiple languages (e.g.
-"Berlin" is currently described in 222 languages).
+``Berlin" is currently described in 222 languages).
 News articles represent another example, where the same event can be described by newspapers in several languages.
@@
 We use $N$ to denote the total row dimension of $X$, i.e.
-$N:= \sum_{i=1}^m n_i$.
+$N:= \sum_{i=1}^m n_i$.\todo{picture - this part isnt that clear}
 
@@
 
-The $k$-means algorithm is perhaps the most well-known and widely-used clustering algorithm. We now present its application
+The $k$-means algorithm is perhaps the most well-known and widely-used clustering algorithm. Here, we present its application
 to computing cross-lingual similarities.
-The idea is based on concatenating the corpus matrices, running standard $k$-means clustering to obtain the matrix of centroids, "reversing" the concatenation step to obtain a set of aligned bases, which are finally used to compute cross-lingual similarities.
+The idea is based on concatenating the corpus matrices, running standard $k$-means clustering to obtain the matrix of centroids, ``reversing" the concatenation step to obtain a set of aligned bases, which are finally used to compute cross-lingual similarities.
 
@@
 Therefore, each document  is represented by a long vector indexed by the terms in all languages.
+\todo{picture here}
 
@@
 $C_i \in \RR^{n_i \times k}$.
-
-The matrices $C_i$ are obtained by analyzing the multilingual corpus matrix $X$ (obtained from Wikipedia for example) and we now show how to use them to compute cross-lingual document similarities on new documents (news documents, for example).
-
-Each matrix $C_i$ represents a vector space basis and can be used to map points in $\RR^{n_i}$ into a $k$-dimensional space, where the coordinates of a vector $x \in \RR^{n_i}$ are expressed as: $$(C_i^T C_i)^{-1} C_i^T x_i.$$
+%
+The matrices $C_i$ are computed using a multilingual corpus matrix $X$ (obtained from Wikipedia for example).
+
+To compute  cross-lingual document similarities on new documents, note that each matrix $C_i$ represents a vector space basis and can be used to map points in $\RR^{n_i}$ into a $k$-dimensional space, where the coordinates of a vector $x \in \RR^{n_i}$ are expressed as: $$(C_i^T C_i)^{-1} C_i^T x_i.$$
 
@@
 
-The next method is Cross-Lingual Latent Semantic Indexing (CL-LSI)\cite{cl_lsi} which is a variant of LSI~\shortcite{lsi} for more than one language.
+The second approach we consider is Cross-Lingual Latent Semantic Indexing (CL-LSI)\cite{cl_lsi} which is a variant of LSI~\shortcite{lsi} for more than one language.
 The method is based on computing a truncated singular value decomposition of $X \approx U S V^T$.
@@
 Moreover, the Lanczos method is hard to parallelize.
-Instead we use a randomized version of the singular value decomposition (SVD) described in \shortcite{tropp} that can be viewed as a block Lanczos method.
+Instead, we use a randomized version of the singular value decomposition (SVD) described in \shortcite{tropp} that can be viewed as a block Lanczos method.
 That enables us to use parallelization and speeds up the computation considerably.
@@
 \subsection{Canonical Correlation Analysis}\label{sec:CCA}
- The final approach is a statistical technique to analyze data from two sources - after which we will describe our main contribution to computing cross-lingual similarities.
-
+ The final approach is a statistical technique to analyze data from two sources - after which we will describe our extension to computing cross-lingual similarities.
+%
  Canonical Correlation Analysis (CCA)~\cite{Hotelling} is a dimensionality reduction technique similar to Principal Component Analysis (PCA)~\cite{Pearson1901On}, with an additional assumption that the data consists of feature vectors that arose from two sources (two views) that share some information.
@@
 Let $\rho(x,y)$ denote the sample based correlation coefficient between two vectors of observations $x$ and $y$.
-By using the sample matrix notation $X_i$ and $X_j$ (assuming no data is missing for clearer presentation) this problem can be formulated as the following optimization problem:
+By using the sample matrix notation $X_i$ and $X_j$ (assuming no data is missing to simplify the exposition), this problem can be formulated as the following optimization problem:
 \begin{equation*}
@@
 Note that the method in its original form is only applicable to two languages where an aligned set of observations is available.
+\todo{each of these sections should have a matrix picture..}
 
@@
 Even if only two languages are considered, the set of aligned documents can be small (an extreme example is given by the Piedmontese and Hindi Wikipedias where no inter-language links are available).
-The second problem is that the number of documents and the involved
+The second problem is the number of documents and involved
 1. missing data
@@
 3. Sum of squared correlations
-
-In this section, we describe an extension of CCA to more than two languages which can be trained on large comparable corpora. The main difficulty with applying the
-LSI and $k$-means approaches lies in the fact, that when one considers a large number of languages, for example, the top 100 Wikipedia languages (ranked by number of articles), the set of completely
-aligned documents is often small or empty. Even if only two languages are considered, the set of aligned documents can be small (for example, Piedmontese and Hindi
+\todo{this list makes no sense}
+
+We describe an extension of CCA to more than two languages, which can be trained on large comparable corpora. The main difficulty with applying the
+LSI and $k$-means approaches is the fact that when one considers a large number of languages, e.g., the top 100 Wikipedia languages (ranked by number of articles), the set of completely
+aligned documents (over all languages) is often small or empty. Even if only two languages are considered, the set of aligned documents can be small (for example, Piedmontese and Hindi
 Wikipedias had no interlanguage links). In Wikipedia, we observed that even though the training resources are scarce between certain language pairs, there often exists indirect training data
@@
 The first part is related to LSI and reduces the dimensionality of the data.
-The second step refines the linear mappings and optimizes linear dependence between data.
+The second step refines the linear mappings and optimizes the linear dependence between data.
 
@@
 If $|a(i,j)|$ is small for many language pairs (as it is in the case of Wikipedia), then many empirical estimates $C_{i,j}$ are unreliable, which can result in overfitting.
-For this reason we perform the truncated singular value decomposition on the matrix $C = [C_{1,2}  \cdots  C_{1,m}] \approx U S V^T$, where where $U \in \RR^{n_1 \times k}, S \in \RR^{k \times k}, V \in \RR^{(\sum_{i=2}^m n_i) \times k}$.
+For this reason, we perform the truncated singular value decomposition on the matrix $C = [C_{1,2}  \cdots  C_{1,m}] \approx U S V^T$, where where $U \in \RR^{n_1 \times k}, S \in \RR^{k \times k}, V \in \RR^{(\sum_{i=2}^m n_i) \times k}$.
 We split the matrix $V$ vertically in blocks with $n_2, \ldots, n_m$ rows: $V = [V_2^T  \cdots  V_m^T]^T$.
@@
 \end{equation*}
-We solve a similar problem by restricting $i=1$ and omit optimizing over non-hub language pairs.
+We solve a similar problem by restricting $i=1$ and omitting the optimization over non-hub language pairs.
 Let $D_{i,i} \in \RR^{k \times k}$ denote the empirical covariance of $\mathcal{Y}_i$ and $D_{i,j}$ denote the empirical cross-covariance computed based on $\mathcal{Y}_i$ and $\mathcal{Y}_j$.
@@
 
- The technique is related to the Generalization of Canonical Correlation Analysis (GCCA) by Carroll \citeyear{Carroll}, where an unknown group configuration variable is defined and objective is to maximize the sum of squared correlation between the group variable and the others.
+ The technique is related to  Generalization of Canonical Correlation Analysis (GCCA) by Carroll \citeyear{Carroll}, where an unknown group configuration variable is defined and objective is to maximize the sum of squared correlation between the group variable and the others.
 The problem can be reformulated as an eigenvalue problem.
 The difference lies in the fact that we set the unknown group configuration variable as the hub language, which simplifies the solution.
-The complexity of our method is $O(k^3)$, whereas solving the GCCA method scales as $O(N^3)$ where $N$ is the number of samples (see \cite{gifi}).
+The complexity of our method is $O(k^3)$, \todo{what is $k$?} whereas solving the GCCA method scales as $O(N^3)$ where $N$ is the number of samples (see \cite{gifi}).
 Another issue with GCCA is that it cannot be directly applied to the case of missing documents.
@@
 In online media streams -- particularly new articles -- there is often duplication of reporting, different viewpoints or opinions, all centering around a single event.
-Same events are covered by many articles and the question we address is how to find all the articles in different languages that are describing a single event.
-We base our evaluation on an online system for detection of world events, called Event Registry.
+The same events are covered by many articles and the question we address is how to find all the articles in different languages that are describing a single event.
+We base our evaluation on an online system for detection of world events, Event Registry.
 We do not address the problem of detection of events in this paper but rather consider the problem that given a series of events, how may we best ``tag" or match articles in different languages to these events.
@@
 \subsection{Problem definition}
+\todo{we may want to synchronize notation}
 
@@
 
-In order to identify clusters $c_j$ that are equivalent to cluster $c_i$, we have developed a two-stage algoirhtm.
+In order to identify clusters $c_j$ that are equivalent to cluster $c_i$, we have developed a two-stage algorithm.
 For a cluster $c_i$, we first efficiently identify a small set of candidate clusters and then find those clusters among the candidates, which are equivalent to $c_i$.
@@
 
-To build a classification model we first extract  informative features from the labeled examples. Based on the availability of the data, we extracted the following groups of features:
+To build a classification model, we first extract  informative features from the labelled examples. Based on the availability of the data, we extracted the following groups of features:
 
@@
 Additionally, we can also assign weights to the concepts based on how frequently they occur in the articles in the cluster.
-From the list of relevant concepts and corresponding weights, we consider the following features: \texttt{entityCosSim} which represents the cosine similarity on the vector of entities mentioned in both clusters; \texttt{keywordCosSim} which represents the cosine similarity on the keywords only; for cases where the weights are not relevant, we have also added two additional features - (\texttt{entityJaccardSim} and \texttt{keywordJaccardSim}), where Jaccard similarity is used instead of the cosine similarity.
+From the list of relevant concepts and corresponding weights, we consider the following features: \texttt{entityCosSim} which represents the cosine similarity on the vector of entities mentioned in both clusters; \texttt{keywordCosSim} which represents the cosine similarity on the keywords only; for cases where the weights are not relevant, we have also added two additional features - (\texttt{entityJaccardSim} and \texttt{keywordJaccardSim}), where Jaccard similarity is used rather than cosine similarity.
 
@@
 
-We will describe the main dataset for building cross-lingual models which is based on Wikipedia. After that we will present two sets of experiments. The first set of experiments
-will establish that the hub based approach can deal with language pairs where little or no training data is available. The second set of experiments will compare the main approaches
-that we presented on the task of mate retrieval and the task of event linking. We will also examine how different choices of features impact the event linking performance.
+We describe the main dataset for building cross-lingual models which is based on Wikipedia and then present two sets of experiments. The first set of experiments
+establishes that the hub based approach can deal with language pairs where little or no training data is available. The second set of experiments compares the main approaches
+that we presented on the task of mate retrieval and the task of event linking. Finally, we examine how different choices of features impact the event linking performance.
 
@@
 We represent the documents as normalized TFIDF\cite{Salton88term-weightingapproaches} weighted vectors.
+ \todo{tfidf over what corpus  - is this static?}
  %These are languages with Wikipedia article sizes comparable to the minority languages.
@@
 Furthermore, we remove the documents that contain less than 20 different words and remove words that are too %infrequent as well as the top 100 most frequent words in the vocabularies.
- Although the English language is well aligned, number of articles highly correlates with the
+ Although the English language is well aligned,\todo{well-aligned with what} number of articles highly correlates with the
     number of articles connected to English, with all Wikipedia languages, we must note that quality of alignment varies quite a bit, especially in the case of small Wikipedias.
@@
 
-We observe that the method performs well on all pairs between languages: \emph{en}, \emph{es}, \emph{ru}, \emph{sl}, where at least 50,000 training documents are available.
-We notice that taking $k = 500$ or $k = 1000$ multilingual topics usually results in similar performance, with some notable exceptions: in the case of (\emph{ht}, \emph{war}) the additional topics result in an increase in performance, as opposed to (\emph{ht},\emph{pms}) where performance drops, which suggests overfitting.
+We observe that the method performs well on all pairs of languages, where at least 50,000 training documents are available(\emph{en}, \emph{es}, \emph{ru}, \emph{sl}).
+We note that taking $k = 500$ or $k = 1000$ multilingual topics usually results in similar performance, with some notable exceptions: in the case of (\emph{ht}, \emph{war}) the additional topics result in an increase in performance, as opposed to (\emph{ht},\emph{pms}) where performance drops, which suggests overfitting.
 The languages where the method performs poorly are \emph{ht} and \emph{war}, which can be explained by the quality of data (see Table \ref{table:rank} and explanation that follows).
@@
 It is clear that in the case of Creole language only at most $22\%$ documents are unique and suitable for the training.
-Though we removed the stub documents, many of remaining documents are nearly the same, as the quality of some minor Wikipedias is low.
-This was confirmed for Creole, Waray-Waray, and Piedmontese language by manual inspection.
+Though we removed the stub documents, many of remaining documents are nearly the same, as the quality of some smaller Wikipedias is low.
+This was confirmed for the Creole, Waray-Waray, and Piedmontese languages by manual inspection.
 The low quality documents correspond to templates about the year, person, town, etc.
@@
 
-Currently we compute similarities over $24$ languages with tags: eng, spa, deu, zho, ita, fra, rus, swe, nld, tur, jpn, por, ara, fin, ron, kor, hrv, tam, hun, slv, pol, srp, cat, ukr but we support any language from the top $100$ Wikipedia languages.
+Currently, we compute similarities over $24$ languages with tags: eng, spa, deu, zho, ita, fra, rus, swe, nld, tur, jpn, por, ara, fin, ron, kor, hrv, tam, hun, slv, pol, srp, cat, ukr but we support any language from the top $100$ Wikipedia languages.
 Our data stream is Newsfeed (http://newsfeed.ijs.si/) which provides $430k$ unique articles per day.
@@
 \section{Discussion and Future Work}
+\todo{We should mention the api, datasetswhatever here - for reprodicability}
 

### 2015-08-27 rupnik2015.tex
placeholder figure and a revision of 4.2, 4.3

@@
 %
-CCA can be used to find
-correlated patterns for a pair of languages, whereas the extended method optimizes a
-sum of squared correlations between several language pairs.
-The method takes advantage of the hub language assumption in the training dataset (the English Wikipedia corresponds to a hub) as well as a preprocessing step based on LSI that acts as regularization and improves scalability.
-\todo{we need to set this up better}
+CCA can be used to find correlated patterns for a pair of languages, whereas the extended method optimizes a
+Sum of Squared Correlations (SSCOR) between several language pairs, which was introduced in \cite{Kettenring}.
+The SSCOR problem is difficult to solve in our setting (hundreds of thousands of features, hundreds of thousands of examples).
+To tackle this, we proposed a method which consists of two ingredients.
+ The first one is based on an observation that certain datasets (such as Wikipedia) are biased towards one language (English), which can be exploited
+ to reformulate a difficult optimization problem as an eigenvector problem. The second ingredient is dimensionality reduction using LSI, which
+ makes the eigenvector problem computationally and numerically tractable. \todo{we need to set this up better. Revised by Jan. Check if OK}
 
@@
 Note that individual document $u_i$ can be an empty document (missing resource) and each $d$ must contain at least two nonempty documents.
-A comparable corpus $D = {d_1, \ldots, d_N}$ is a collection of multilingual documents.
-By using the vector space model we can represent $D$ as a set of $m$ matrices $X_1,\ldots,X_m$, where $X_i \in \RR^{n_i \times N}$ is the matrix corresponding to the language $i$ and $n_i$ is the vocabulary size of language $i$.
+A comparable corpus $D = {d_1, \ldots, d_s}$ is a collection of $s$ multilingual documents.
+By using the vector space model we can represent $D$ as a set of $m$ matrices $X_1,\ldots,X_m$, where $X_i \in \RR^{n_i \times s}$ is the matrix corresponding to the language $i$ and $n_i$ is the vocabulary size of language $i$.
 Furthermore, let $X_i^{\ell}$ denote the $\ell$-th column of matrix $X_i$ and the matrices respect the document alignment - the vector $X_i^\ell$ corresponds to the TFIDF vector of the $i$-th component of multilingual document $d_\ell$.
 We use $N$ to denote the total row dimension of $X$, i.e.
-$N:= \sum_{i=1}^m n_i$.\todo{picture - this part isnt that clear}
+$N:= \sum_{i=1}^m n_i$.
+See the Figure~\ref{fig:stacked_matrices} for an illustration of the introduced notation.
+
+\begin{figure}[tbp]
+\centering
+\includegraphics[width=\textwidth]{stacked_matrices}
+\caption{\label{fig:stacked_matrices} Multilingual corpora and their matrix representations using the vector space model.}
+\end{figure}
 

### 2015-08-31 rupnik2015.tex
added submitted paper

@@
 
- In order to apply the algorithm, we first merge all the term-document matrices into a single matrix $X$ by stacking the individual term-document matrices:
+ In order to apply the algorithm, we first merge all the term-document matrices into a single matrix $X$ by stacking the individual term-document matrices (as seen in Figure~\ref{fig:stacked_matrices}):
 $$X := \begin{bmatrix}X_1^T ,X_2^T, \cdots, X_m^T \end{bmatrix}^T,$$
@@
 Therefore, each document  is represented by a long vector indexed by the terms in all languages.
-\todo{picture here}
+\todo{picture here.
+Updated: added Figure 3}
 
@@
 
-To compute  cross-lingual document similarities on new documents, note that each matrix $C_i$ represents a vector space basis and can be used to map points in $\RR^{n_i}$ into a $k$-dimensional space, where the coordinates of a vector $x \in \RR^{n_i}$ are expressed as: $$(C_i^T C_i)^{-1} C_i^T x_i.$$
+To compute  cross-lingual document similarities on new documents, note that each matrix $C_i$ represents a vector space basis and can be used to map points in $\RR^{n_i}$ into a $k$-dimensional space, where the new coordinates of a vector $x \in \RR^{n_i}$ are expressed as: $$(C_i^T C_i)^{-1} C_i^T x_i.$$
 

### 2015-08-31 rupnik2015.tex
placeholder figures for kmeans and lsi

@@
 The idea is based on concatenating the corpus matrices, running standard $k$-means clustering to obtain the matrix of centroids, ``reversing" the concatenation step to obtain a set of aligned bases, which are finally used to compute cross-lingual similarities.
+See the Figure~\ref{fig:kmeans} for an illustration of process of computing multilingual clusters and using them to change the coordinates to ``topic`` space.
+
+\begin{figure}[tbp]
+\centering
+\includegraphics[width=\textwidth]{kmeans}
+\caption{\label{fig:kmeans} $k$-means algorithm and coordinate change.}
+\end{figure}
 
@@
 The method is based on computing a truncated singular value decomposition of $X \approx U S V^T$.
+See the Figure~\ref{fig:lsi} for the decomposition.
+Representing documents in ``topic`` coordinates is done in the same way as in the $k$-means case (see Figure~\ref{fig:kmeans}).
+
+\begin{figure}[tbp]
+\centering
+\includegraphics[width=\textwidth]{lsi}
+\caption{\label{fig:lsi} LSI multilingual corpus matrix decomposition.}
+\end{figure}
+
  Since the matrix can be large we can use an iterative method like the Lanczos~\cite{golub} algorithm with reorthogonalization to find the left singular vectors (columns of $U$) corresponding to the largest singular values.
@@
 
-The numerical implementation of the least squares is done by QR algorithm, by computing factorizing $U_i = Q R$, where $Q^TQ = I$ and $R$ is triangular matrix.
+The numerical implementation of the least squares is done by QR algorithm~\cite{golub}, by computing factorizing $U_i = Q R$, where $Q^TQ = I$ and $R$ is triangular matrix.
 $P_i$ is then obtained by solving $R P_i = Q$.
@@
 Building cross-lingual similarity models based on comparable corpora is challenging for two main reasons.
-The problem of missing alignment data: when a number of languages is large, the dataset of documents that cover all languages is small (or may even be empty).
+The first problem is related to missing alignment data: when a number of languages is large, the dataset of documents that cover all languages is small (or may even be empty).
 Even if only two languages are considered, the set of aligned documents can be small (an extreme example is given by the Piedmontese and Hindi Wikipedias where no inter-language links are available).
-The second problem is the number of documents and involved
+The second challenge is scale - the data is high-dimensional (hundreds of thousands of features per language, many languages) and the number of multilingual documents may be large (over one million in case of Wikipedia).
+
 1. missing data
@@
 aligned documents (over all languages) is often small or empty. Even if only two languages are considered, the set of aligned documents can be small (for example, Piedmontese and Hindi
-Wikipedias had no interlanguage links). In Wikipedia, we observed that even though the training resources are scarce between certain language pairs, there often exists indirect training data
-through what we refer to as a hub language.
+Wikipedias had no interlanguage links).
+In Wikipedia, we observed that even though the training resources are scarce between certain language pairs, there often exists indirect training data through what we refer to as a hub language.
 

### 2015-09-01 rupnik2015.tex
updated figures

@@
 The SSCOR problem is difficult to solve in our setting (hundreds of thousands of features, hundreds of thousands of examples).
-To tackle this, we proposed a method which consists of two ingredients.
+To tackle this, we propose a method which consists of two ingredients.
  The first one is based on an observation that certain datasets (such as Wikipedia) are biased towards one language (English), which can be exploited
@@
 \centering
-\includegraphics[width=\textwidth]{stacked_matrices}
+\includegraphics[width=\textwidth]{stacked_matrices.pdf}
 \caption{\label{fig:stacked_matrices} Multilingual corpora and their matrix representations using the vector space model.}
@@
 The idea is based on concatenating the corpus matrices, running standard $k$-means clustering to obtain the matrix of centroids, ``reversing" the concatenation step to obtain a set of aligned bases, which are finally used to compute cross-lingual similarities.
-See the Figure~\ref{fig:kmeans} for an illustration of process of computing multilingual clusters and using them to change the coordinates to ``topic`` space.
+See the Figure~\ref{fig:kmeans} for an illustration of process of computing multilingual clusters and using them to change the coordinates to cluster centroid ``topic" space.
 
@@
 \centering
-\includegraphics[width=\textwidth]{kmeans}
+\includegraphics[width=\textwidth]{kmeans.pdf}
 \caption{\label{fig:kmeans} $k$-means algorithm and coordinate change.}
@@
 %
-The matrices $C_i$ are computed using a multilingual corpus matrix $X$ (obtained from Wikipedia for example).
+To reiterate, the matrices $C_i$ are computed using a multilingual corpus matrix $X$ (based on Wikipedia for example).
 
@@
 The second approach we consider is Cross-Lingual Latent Semantic Indexing (CL-LSI)\cite{cl_lsi} which is a variant of LSI~\shortcite{lsi} for more than one language.
-The method is based on computing a truncated singular value decomposition of $X \approx U S V^T$.
+The approach is very similar to $k$-means, where we first concatenate the corpus matrices, compute a decomposition, which in case of CL-LSI is a truncated Singular Value Decomposition (SVD), decouple the
+ column space matrix and use the blocks to compute linear maps to a common vector space, where standard cosine similarity is used to compare documents.
+
+ The method is based on computing a truncated singular value decomposition of the concatenated corpus matrix $X \approx U S V^T$.
 See the Figure~\ref{fig:lsi} for the decomposition.
-Representing documents in ``topic`` coordinates is done in the same way as in the $k$-means case (see Figure~\ref{fig:kmeans}).
+Representing documents in ``topic`` coordinates is done in the same way as in the $k$-means case (see Figure~\ref{fig:kmeans}), which we will
 
@@
 \centering
-\includegraphics[width=\textwidth]{lsi}
+\includegraphics[width=\textwidth]{lsi.pdf}
 \caption{\label{fig:lsi} LSI multilingual corpus matrix decomposition.}
@@
 
- Since the matrix can be large we can use an iterative method like the Lanczos~\cite{golub} algorithm with reorthogonalization to find the left singular vectors (columns of $U$) corresponding to the largest singular values.
+The cross-lingual similarity functions are based on a rank-$k$ truncated SVD: $X \approx U \Sigma V^T,$ where $U \in \RR^{N \times k}$ are basis vectors of interest and $\Sigma \in \RR^{k \times k}$ is truncated diagonal matrix of singular eigenvalues.
+An aligned basis is obtained by first splitting $U$ vertically according to the number of dimensions of each language: $U = [U_1^T \cdots U_m^T]^T$.
+Then, the same as with $k$-means clustering, we compute the pseudoinverses $P_i = (U_i^T U_i)^{-1} U_i^T$.
+The matrices $P_i$ are used to change the basis from the standard basis in $\RR^{n_i}$ to the basis spanned by the columns of $U_i$.
+
+\textbf{A note on implementation}
+
+  Since the matrix $X$ can be large we could use an iterative method like the Lanczos~\cite{golub} algorithm with reorthogonalization to find the left singular vectors (columns of $U$) corresponding to the largest singular values.
 It turns out that the Lanczos method converges slowly as the gap between the leading singular values is small.
@@
 
-The cross-lingual similarity functions are based on a rank-$k$ truncated SVD: $X \approx U \Sigma V^T,$ where $U \in \RR^{N \times k}$ are basis vectors of interest and $\Sigma \in \RR^{k \times k}$ is truncated diagonal matrix of singular eigenvalues.
-
-An aligned basis is obtained by first splitting $U$ vertically according to the number of dimensions of each language: $U = [U_1^T \cdots U_m^T]^T$.
-Then, the same as with $k$-means clustering, we compute the pseudoinverses $P_i = (U_i^T U_i)^{-1} U_i^T$.
-The matrices $P_i$ are used to change the basis from the standard basis in $\RR^{n_i}$ to the basis of columns of $U_i$.
-
-The numerical implementation of the least squares is done by QR algorithm~\cite{golub}, by computing factorizing $U_i = Q R$, where $Q^TQ = I$ and $R$ is triangular matrix.
-$P_i$ is then obtained by solving $R P_i = Q$.
+To compute the matrices $P_i$ we used the QR algorithm~\cite{golub} to factorize $U_i$ as $U_i = Q_i R_i$, where $Q_i^TQ_i = I$ and $R_i$ is triangular matrix.
+$P_i$ is then obtained by solving $R_i P_i = Q_i$.
 
 \subsection{Canonical Correlation Analysis}\label{sec:CCA}
- The final approach is a statistical technique to analyze data from two sources - after which we will describe our extension to computing cross-lingual similarities.
+ We now present a statistical technique to analyze data from two sources, extension of which we will be presented in the next section.
 %

### 2015-09-01 rupnik2015.tex
hub cca section

@@
 \subsection{Canonical Correlation Analysis}\label{sec:CCA}
- We now present a statistical technique to analyze data from two sources, extension of which we will be presented in the next section.
+ We now present a statistical technique to analyze data from two sources, extension of which will be presented in the next section.
 %
@@
 The first problem is related to missing alignment data: when a number of languages is large, the dataset of documents that cover all languages is small (or may even be empty).
-Even if only two languages are considered, the set of aligned documents can be small (an extreme example is given by the Piedmontese and Hindi Wikipedias where no inter-language links are available).
+Even if only two languages are considered, the set of aligned documents can be small (an extreme example is given by the Piedmontese and Hindi Wikipedias where no inter-language links are available), in which case none of the methods presented so far are applicable.
  The second challenge is scale - the data is high-dimensional (hundreds of thousands of features per language, many languages) and the number of multilingual documents may be large (over one million in case of Wikipedia).
-
-1. missing data
-2. LSI step
-3. Sum of squared correlations
-\todo{this list makes no sense}
-
-We describe an extension of CCA to more than two languages, which can be trained on large comparable corpora. The main difficulty with applying the
-LSI and $k$-means approaches is the fact that when one considers a large number of languages, e.g., the top 100 Wikipedia languages (ranked by number of articles), the set of completely
-aligned documents (over all languages) is often small or empty. Even if only two languages are considered, the set of aligned documents can be small (for example, Piedmontese and Hindi
-Wikipedias had no interlanguage links).
+The optimization problem related to CCA is not
+ easy to solve: the covariance matrices themselves (or in a dual formulation, the kernel matrices) are prohibitively large to fit in memory (even storing a 100,000 by 100,000 element matrix requires 80GB of memory) and iterative matrix-multiplication based approaches to
+ solving generalized eigenvalue problems are required (the covariance matrices can be expressed as products of sparse matrices, which means we have fast matrix-vector multiplication).
+
+We now describe an extension of CCA to more than two languages, which can be trained on large comparable corpora.
+ The extension we consider is based on a generalization of CCA to more than two views, introduced in \cite{Kettenring}, namely the Sum of Squared Correlation SSCOR, which we will state formally in the next paragraph.
+
+ \todo{MESSAGE: Difficult problem, but if the data has a hub, it becomes an eigenvalue problem. With dimensionality reduction based on LSI it can easily be solved.}
+
 In Wikipedia, we observed that even though the training resources are scarce between certain language pairs, there often exists indirect training data through what we refer to as a hub language.

### 2015-09-03 rupnik2015.tex
hub languages updates

@@
 a simple approach based on $k$-means clustering in \ref{sec:kmeans}, a standard approach based on singular value decomposition in \ref{sec:LSI}, a related
-approach called Canonical Correlation Analysis (CCA) in \ref{sec:CCA} and finally a new, more generally applicable method, which is an extension of CCA to more than two languages in \ref{sec:hublang}.
+approach called Canonical Correlation Analysis (CCA) in \ref{sec:CCA} and finally a new method, which is an extension of CCA to more than two languages in \ref{sec:hublang}.
 %
@@
 The optimization problem related to CCA is not
- easy to solve: the covariance matrices themselves (or in a dual formulation, the kernel matrices) are prohibitively large to fit in memory (even storing a 100,000 by 100,000 element matrix requires 80GB of memory) and iterative matrix-multiplication based approaches to
- solving generalized eigenvalue problems are required (the covariance matrices can be expressed as products of sparse matrices, which means we have fast matrix-vector multiplication).
-
-We now describe an extension of CCA to more than two languages, which can be trained on large comparable corpora.
- The extension we consider is based on a generalization of CCA to more than two views, introduced in \cite{Kettenring}, namely the Sum of Squared Correlation SSCOR, which we will state formally in the next paragraph.
-
- \todo{MESSAGE: Difficult problem, but if the data has a hub, it becomes an eigenvalue problem. With dimensionality reduction based on LSI it can easily be solved.}
-
+ trivial to solve: the covariance matrices themselves are prohibitively large to fit in memory (even storing a 100,000 by 100,000 element matrix requires 80GB of memory) and iterative matrix-multiplication based approaches to  solving generalized eigenvalue problems are required (the covariance matrices can be expressed as products of sparse matrices, which means we have fast matrix-vector multiplication).
+
+We now describe an extension of CCA to more than two languages, which can be trained on large comparable corpora and can handle missing data.
+ The extension we consider is based on a generalization of CCA to more than two views, introduced in \cite{Kettenring}, namely the Sum of Squared Correlation SSCOR, which we will state formally later in this section.
+Our approach exploits a certain characteristic of the data, namely the \emph{hub language} characteristic (introduced in the next paragraph) in two
+ ways: to reduce the dimensionality of the data and to simplify the optimization problem.
+
+\textbf{Hub language characteristic}
 In Wikipedia, we observed that even though the training resources are scarce between certain language pairs, there often exists indirect training data through what we refer to as a hub language.
@@
 A \emph{hub language} is a language with a high proportion of non-empty documents in $D = \left\{d_1,..., d_{\ell}\right\}$.
-The prototypical  exampling  in the case of Wikipedia is English.
+The prototypical example in the case of Wikipedia is English.
 We use the following notation to define subsets of the multilingual comparable corpus: let $a(i,j)$ denote the index set of all multilingual documents with non-missing data for the $i$-th and $j$-th language:  $$a(i,j) = \left\{k~ |~ d_k = (u_1,...,u_m), u_i \neq \emptyset, u_j \neq \emptyset \right\},$$ and let $a(i)$ denote the index set of all multilingual documents with non missing data for the $i$-th language.
@@
 The second step refines the linear mappings and optimizes the linear dependence between data.
+
+\textbf{Step 1: Hub language based dimensionality reduction}
 
@@
 To summarize, the first step reduces the dimensionality of the data and is based on CL-LSI, but optimizes only the hub-language related cross-covariance blocks.
+
+\textbf{Step 2: Simplifying and solving SSCOR}
 
@@
 
+\textbf{Remark}
  The technique is related to  Generalization of Canonical Correlation Analysis (GCCA) by Carroll \citeyear{Carroll}, where an unknown group configuration variable is defined and objective is to maximize the sum of squared correlation between the group variable and the others.
@@
 The difference lies in the fact that we set the unknown group configuration variable as the hub language, which simplifies the solution.
-The complexity of our method is $O(k^3)$, \todo{what is $k$?} whereas solving the GCCA method scales as $O(N^3)$ where $N$ is the number of samples (see \cite{gifi}).
+The complexity of our method is $O(k^3)$, where $k$ is the reduced dimension from the LSI preprocessing step, whereas solving the GCCA method scales as $O(s^3)$, where $s$ is the number of samples (see \cite{gifi}).
 Another issue with GCCA is that it cannot be directly applied to the case of missing documents.

### 2015-09-03 rupnik2015.tex
added datasets
Went over section 5. Comments are on Google Docs

@@
 These descriptors are topics, such as \emph{air transport}, \emph{EC agreement}, \emph{competition} and \emph{pollution control} into which articles are automatically categorized~\shortcite{Pouliquen2006}.
-Similarity between clusters is then computed using a linear combination of the cosine similarities computed on the lists.
+Similarity between clusters is then computed using a linear combination of the cosine similarities computed on the three vectors.
 If the similarity is above the threshold, the clusters are linked.
@@
 \centering
-\includegraphics[width=\textwidth]{stacked_matrices.pdf}
+\includegraphics[width=9cm]{stacked_matrices.pdf}
 \caption{\label{fig:stacked_matrices} Multilingual corpora and their matrix representations using the vector space model.}
@@
 \centering
-\includegraphics[width=\textwidth]{kmeans.pdf}
+\includegraphics[width=10cm]{kmeans.pdf}
 \caption{\label{fig:kmeans} $k$-means algorithm and coordinate change.}
@@
 \centering
-\includegraphics[width=\textwidth]{lsi.pdf}
+\includegraphics[width=10cm]{lsi.pdf}
 \caption{\label{fig:lsi} LSI multilingual corpus matrix decomposition.}
@@
 
-The main application on which we test the above similarity is cross-lingual event linking.
-In online media streams -- particularly new articles -- there is often duplication of reporting, different viewpoints or opinions, all centering around a single event.
+The main application on which we test the cross-lingual similarity is cross-lingual event linking.
+In online media streams -- particularly news articles -- there is often duplication of reporting, different viewpoints or opinions, all centering around a single event.
 The same events are covered by many articles and the question we address is how to find all the articles in different languages that are describing a single event.
-We base our evaluation on an online system for detection of world events, Event Registry.
-We do not address the problem of detection of events in this paper but rather consider the problem that given a series of events, how may we best ``tag" or match articles in different languages to these events.
+In this paper we consider the problem of matching events from different languages.
+We do not address the problem of detection of events and instead base our evaluation on an online system for detection of world events, Event Registry.
 The events are represented by clusters of articles and so ultimately our problem reduces to finding suitable matchings between clusters with articles in different languages.
@@
 \subsection{Problem definition}
-\todo{we may want to synchronize notation}
-
-The problem of cross-lingual event linking is to identify the clusters describing the same events.
-Each article $a_i$ is written in a language $l$, where $\ell \in L = \{\ell_1,\ell_2,...,\ell_k\}$.
-For each language $\ell$, we generate and maintain a set of language-specific clusters $C_l$.
-That is, all articles are in language $l$.
-For any cluster $c_i \in C_{\ell_a}$, we would like to identify clusters $c_j \in C_{\ell_b}$ (where $\ell_a \neq \ell_b$) that describe the same event as $c_i$.
-We define \emph{equivalent clusters} as  pairs $(c_i,c_j)$ such that both clusters in the pair describe the same event in languages $C_{\ell_a}$ and $C_{\ell_b}$ respectively.
-Furthermore, we iterate over all the languages we consider.
-That is, given a cluster in language $\ell_i$, we look for matching clusters in language $\ell_j$ for $i\neq j$  in a pairwise fashion.
+
+The problem of cross-lingual event linking is to match monolingual clusters of news articles that describing the same event across languages.
+For example, we want to match a cluster of Spanish news articles and a cluster of English news articles that both describe the same earthquake.
+
+Each article $a \in A$ is written in a language $\ell$, where $\ell \in L = \{\ell_1,\ell_2,...,\ell_m\}$.
+For each language $\ell$, we obtain a set of monolingual clusters $C_{\ell}$.
+That is, all articles from each cluster $c \in C_{\ell}$ are in language $\ell$.
+Given a pair of languages $\ell_a \in L$, $\ell_b \in L$ and $\ell_a \not= \ell_b$, we would like to identify all cluster pairs $(c_i, c_j) \in C_{\ell_a} \times C_{\ell_b}$ such that $c_i$ and $c_j$ describe the same event.
 
@@
 \centering
-\includegraphics[width=0.6\textwidth]{clusters}
-\caption{\label{fig:clusters}  Given clusters of articles in different languages, the goal of  linking is to match clusters which have similar content across languages.
-The challenge in linking is to overcome the errors in the clustering and differing content as well as make the approach scalable.
-Therefore, we first identify candidate clusters.
-}
+\includegraphics[width=0.7\textwidth]{clusters.pdf}
+\caption{\label{fig:clusters}  Clusters composed of English and Spanish news articles.
+Arrows link English articles with their Spanish $k$-nearest neighbor matches according to the cross-lingual similarity.}
 \end{figure}
 
-Furthermore, the matching of clusters is a \emph{generalized matching}.
-We cannot assume that there is only one cluster per language per event, nor can we assume complete coverage -- that there exists at least one cluster per event in every language.
-This implies that we cannot make any assumptions on the matching, e.g.
-one-to-one, complete, etc.
-This excludes the use of standard weighted bipartite matching type of algortihms for this problem.
+Matching of clusters is a \emph{generalized matching} problem.
+We cannot assume that there is only one cluster per language per event, nor can we assume complete coverage -- i.e.
+that there exists at least one cluster per event in every language.
+This is partly due to news coverage which might be more granular in some languages, partly due to noise and errors in the event detection process.
+This implies that we cannot make assumptions on the matching (e.g., one-to-one or complete matchin) and excludes the use of standard weighted bipartite matching type of algortihms for this problem.
 An example is shown in Figure~\ref{fig:clusters}, where a cluster may contain articles which are closely matched with many clusters in a different language.
 
-We impose an additional requirement.
-Identification of the equivalent clusters cannot be done exhaustively -- for a cluster $c_i$ we wish to avoid testing all other clusters for all the other languages.
-Performing exhaustive comparison for each cluster $c_i$ would result in $O(N^2)$ tests, where $N$ is the number of all clusters (over all languages), which is not feasible when the number of clusters is on the order of tens of thousands.
-
-\begin{comment}
-One of the use cases where article linking information can be applied is also in an online system for detection of world events, called Event Registry\cite{}.
-Event Registry is a repository of events, where events are automatically identified by analyzing news articles.
-The term event is vague and ambiguous, but for the practical purposes we define it as ``any significant happening that is being reported about in the media''.
-
-Event Registry works by collecting and analyzing news articles that are published by news outlets in different languages all over the world.
-Collected articles are first semantically annotated by identifying in them mentions of relevant concepts -- either entities or important keywords.
-The disambiguated and entity linking of the concepts is done using Wikipedia as the main knowledge base.
-The key step in Event Registry is then to apply an online clustering algorithm on the articles in order to identify groups of articles that are discussing the same event.
-For each new article, the clustering algorithm determines if the article should be assigned to some existing cluster or into a new cluster.
-The underlying assumption is that articles that are describing the same event are similar enough and will therefore be put into the same cluster.
-
-In short, the article clustering works as follows (details are available in \cite{Brank2014}).
-First we create a new document by joining the title and the body of the article.
-We then tokenize the document, remove the stop words and stem the words.
-The remaining words are represented as a vector in the vector-space model.
-Each value in the vector is additionally normalized using TF-IDF.
-After normalization, we use the vector to find the most similar existing cluster.
-To be able to compare the vector with a cluster we first need a vector representation for each of the clusters.
-This is done by computing a centroid vector of the cluster by simply averaging the vectors of the articles in the cluster.
-By computing the cosine similarity of the article vector with each of the centroid vectors we then find the cluster with the highest similarity.
-If the highest similarity is above the selected threshold (we used 0.4 in our experiments), the article is assigned to the corresponding cluster, otherwise a new cluster is created, initially containing only the single article.
-
-Since articles about an event are commonly written only for a short period of time, we remove clusters once the oldest article in the cluster becomes more than 4 days old.
-This housekeeping mechanism prevents the clustering from becoming slow and also ensures that articles are not assigned to obsolete clusters.
-
-In order to treat the cluster as describing an event in Event Registry, the cluster has to grow above a certain number of articles (the number depends on the language).
-This condition ensures that only content that receives a sufficient coverage in the media is treated as an event.
-Once a new event is identified, a new unique ID is assigned to it and the main information about the event is then automatically extracted by analyzing the articles assigned to it.
-The extracted information includes properties such as the date of the event, the location, who is involved in it, what the event is about, etc.
-
-The clustering of articles is done for each language separately -- English articles are, for example, clustered separately from Spanish articles.
-When events are identified, they will therefore be identified from clusters in each language separately.
-As a result, the same event will be represented in Event Registry with two or more event IDs, and each event will only contain coverage in a single language.
-To avoid this problem we want to exploit the available information about the similarity of the articles in a cluster with articles in other languages.
-Using this information we hope to be able to connect clusters in different languages and represent them using a single event ID.
-
-\subsection{Problem definition}
-
-The problem in the second phase of the pipeline is to identify the clusters (given?) describing the same events.
-Each article $a_i$ is written in a language $l$, where $l \in L = \{l_1,l_2,...,l_k\}$.
-For each language $l$, we generate and maintain a set of language-specific clusters $C_l$.
-That is, all articles are in language $l$.
-For any cluster $c_i \in C_{l_a}$, we would like to identify clusters $c_j \in C_{l_b}$ (where $l_a \neq l_b$) that describe the same event as $c_i$.
-We define \emph{equivalent clusters} as  pairs $(c_i,c_j)$ such that both clusters in the pair describe the same event in languages $C_{l_a}$ and $C_{l_b}$ respectively.
-
-\end{comment}
+We are also seeking for an algorithm which does not do exhaustive comparison of all clusters, since that can become prohibitively expensive when working in a real-time setting.
+More specifically, we wish to avoid testing cluster $c_i$ with all the clusters from all the other languages.
+Performing exhaustive comparison would result in $O(|C|^2)$ tests, where $|C|$ is the number of all clusters (over all languages), which is not feasible when the number of clusters is on the order of tens of thousands.
+We address this by testing only clusters that are connected with at least one $k$-nearest neighbor (marked as \emph{candidate clusters} in Figure~\ref{fig:clusters}).
 
@@
 
-In order to identify clusters $c_j$ that are equivalent to cluster $c_i$, we have developed a two-stage algorithm.
+In order to identify clusters that are equivalent to cluster $c_i$, we have developed a two-stage algorithm.
 For a cluster $c_i$, we first efficiently identify a small set of candidate clusters and then find those clusters among the candidates, which are equivalent to $c_i$.
@@
 The details of the first step are described in Algorithm~\ref{cluster_merge_algo1}.
-The algorithm begins by individually inspecting each article $a_i$ in the cluster $c_i$.
-Using a chosen method for computing cross-lingual document similarity (see Section~\ref{sec:models}), it identifies the 10 most similar articles to $a_i$ in each language $l \in L$.
+The algorithm begins by individually inspecting each news article $a_i$ in the cluster $c_i$.
+Using a chosen method for computing cross-lingual document similarity (see Section~\ref{sec:models}), it identifies the 10 most similar news articles to $a_i$ in each language $\ell \in L$.
 For each similar article $a_j$, we identify its corresponding  cluster $c_j$ and add it to the set of candidates.
-The set of candidate clusters obtained in this way is several orders of magnitude smaller than the number of all clusters.
+The set of candidate clusters obtained in this way is several orders of magnitude smaller than the number of all clusters, and at most linear with respect to the number of news articles in cluster $c_i$.
+In practice, clusters contain highly related articles and as such similar articles from other languages mostly fall in only a few candidate clusters.
+
 Although computed document similarities are approximate, our  assumption is that articles in different languages describing the same event, will generally have a higher similarity than articles about different events.
@@
 \SetKwInput{KwInput}{input}\SetKwInput{KwOutput}{output}
-\KwInput{test cluster $c_i$,\\a set of clusters $C_l$ for each language $l \in L$}
+\KwInput{test cluster $c_i$, a set of clusters $C_\ell$ for each language $\ell \in L$}
 \KwOutput{a set of clusters $C$ that are potentially equivalent to $c_i$}
-$C = \{\}$\;
+$C \leftarrow \{\}$\;
 \For{article $a_i \in c_i$} {
-    \For{language $l \in L$} {
-        \tcc{use hub CCA to find 10 most similar articles to article $a_i$ in language $l$}
-        $SimArt = getCCASimArts(a_i, l)$\;
-        \For{article $a_j \in SimArt$} {
+    \For{language $\ell \in L$} {
+        \tcc{use hub CCA to find 10 most similar articles to article $a_i$ in language $\ell$}
+        $SimilarArticles = getCCASimilarArticles(a_i, \ell)$\;
+        \For{article $a_j \in SimilarArticles$} {
             \tcc{find cluster $c_j$ to which article $a_j$ is assigned to}
-            $c_j := c ; c \in C_l, a_j \in c$\;
+            $c_j \leftarrow c$, such that $c \in C_\ell$ and $a_j \in c$\;
             \tcc{add cluster $c_j$ to the set of candidates $C$}
-            $C \leftarrow c_j$\;
+            $C \leftarrow C \cup \{ c_j \}$\;
         }
@@
 
-The second stage of the algorithm (described in the Algorithm~\ref{cluster_merge_algo2}.) determines which (if any) of the candidate clusters are equivalent to $c_i$.
+The second stage of the algorithm determines which (if any) of the candidate clusters are equivalent to $c_i$.
 We treat this task as a supervised learning problem.
-For each candidate cluster $c_j \in C'$, we compute a vector of learning features that should be indicative of whether the $c_i$ and $c_j$ are equivalent or not.
-The details of our chosen features will be described in the following section.
-Once these features are computed, we apply a classification model which was trained using a  dataset  of hand-labeled examples.
-If the model predicts that tested clusters are equivalent, the $c_j$ is added to the set $C'$.
-
-\begin{algorithm}[tb!]
-\SetKwInput{KwInput}{input}\SetKwInput{KwOutput}{output}
-\KwInput{test cluster $c_i$,\\
-a set of candidate clusters $C$ that are potentially equivalent to $c_i$\\
-classification model $M$ trained on the hand-labeled examples}
-\KwOutput{a set of clusters $C'$ that are equivalent to $c_i$}
-
-$C' = \{\}$\;
-\For{cluster $c_j \in C$} {
-    \tcc{extract a vector of features for cluster pair $c_i$ and $c_j$}
-    $\vec{v} = f(c_i, c_j)$\;
-    \tcc{classify the feature vector $\vec{v}$ using the model $M$}
-    $pred = M(\vec{v})$\;
-    \If{$pred$ = true} {
-        $C' \leftarrow c_j$
-    }
-}
-\caption{Algorithm for identifying clusters $C'$ that are equivalent to cluster $c_i$}
-\label{cluster_merge_algo2}
-\end{algorithm}
-
-\subsection{Training a classification model}
-\label{trainClassifier}
-
-In order to train a classification model for predicting whether two clusters are equivalent, we must first have labelled data.
-To obtain the data, we used two human annotators.
-The annotators were provided with an interface displaying information about the articles in two clusters and their task was to determine if the clusters were equivalent or not.
-To obtain a pair of clusters $(c_i, c_j)$ to annotate, we first randomly chose a cluster $c_i$, used the Algorithm~\ref{cluster_merge_algo1} to compute a set of potentially equivalent clusters $C$ and randomly chose a cluster $c_j \in C$.
-
-The dataset provided by the annotators contains 808 examples, of which 402 are equivalent clusters pairs and 406 are not.
-Clusters in each learning example are either in English, Spanish or German.
-Although Event Registry imports articles in other languages as well, we restricted our experiments to these three languages.
-We chose only these three languages since they have very large number of articles and clusters per day which makes the cluster linking problem hard due to large number of possible links.
-
-To build a classification model, we first extract  informative features from the labelled examples. Based on the availability of the data, we extracted the following groups of features:
+For each candidate cluster $c_j \in C$, we compute a vector of learning features that should be indicative of whether the $c_i$ and $c_j$ are equivalent or not and apply a binary classification model that predicts if the clusters are equivalent or not.
+The classification algorithm that we used to train a model was a linear Support Vector Machine (SVM) method \shortcite{shawe-taylor04kernel}.
+
+We use three groups of features to describe cluster pair $(c_i, c_j)$.
+The first group is based on {\bf cross-lingual article links}, which are derived using cross-lingual similarity: each news article $a_i$ is linked with its $10$-nearest neighbors articles from all other languages.
+The group contains the following features:
 
 \begin{itemize}
-\item {\bf Cross-lingual article linking features}.
-Cross-lingual article linking information was already used to obtain candidates for the equivalent clusters.
-Additionally, it can also be used to provide valuable  features for learning.
-There are two main features that we extract from article links between clusters -- \texttt{linkCount} and \texttt{avgSimScore}.
-\texttt{linkCount} is the number of times an article $a_i \in c_i$ has as one of the 10 most similar articles an article $a_j \in c_j$.
-In other words, it is the number of times an article from $c_i$ has a very similar article, which is in $c_j$.
-Beside the number of links between the two clusters we can also take into account the similarity scores of the links.
-The \texttt{avgSimScore} is the feature that represents the average similarity score of the links between the two clusters.
-
-\item {\bf Concept-related features}.
-Articles that are imported into Event Registry are first semantically annotated by linking mentioned entities and keywords to the corresponding Wikipedia pages.
+\item \texttt{linkCount} is the number of times any of the news articles from $c_j$ is among $10$-nearest neighbors for articles from $c_i$.
+In other words, it is the number of times an article from $c_i$ has a very similar article (i.e.
+is among 10 most similar) in $c_j$.
+\item \texttt{avgSimScore} is the average similarity score of the links, as identified for \texttt{linkCount}, between the two clusters.
+\end{itemize}
+
+The second group are {\bf concept-related features}.
+Articles that are imported into Event Registry are annotated by disambiguating mentioned \emph{entities} and \emph{keywords} to the corresponding Wikipedia pages~\cite{zhang2014}.
 Whenever Barack Obama is, for example, mentioned in the article, the article is annotated with a link to his Wikipedia page\footnote{\emph{http://en.wikipedia.org/wiki/Barack\_Obama}}.
-In the same way all mentions of people, locations, organizations and even ordinary keywords (e.g.
+In the same way all mentions of entities (people, locations, organizations) and ordinary keywords (e.g.
 bank, tax, ebola, plane, company) are annotated.
@@
 Additionally, we can also assign weights to the concepts based on how frequently they occur in the articles in the cluster.
-From the list of relevant concepts and corresponding weights, we consider the following features: \texttt{entityCosSim} which represents the cosine similarity on the vector of entities mentioned in both clusters; \texttt{keywordCosSim} which represents the cosine similarity on the keywords only; for cases where the weights are not relevant, we have also added two additional features - (\texttt{entityJaccardSim} and \texttt{keywordJaccardSim}), where Jaccard similarity is used rather than cosine similarity.
-
-\item {\bf Miscellaneous features}.
-This group contains three miscellaneous features that seem discriminative but are unrelated to the previous two groups.
-The first feature is related to the event location.
-By considering the locations mentioned in the articles in a cluster, we are able to estimate where an event occurred.
-The \texttt{hasSameLocation} feature is a boolean variable that is true when the location of the event in both clusters is the same.
-Another important feature is related to the time when the articles in the clusters were published.
-For each cluster, we first compute the average publication time and date of the articles.
-The \texttt{timeDiff} is then defined as the absolute difference in hours between the two computed dates.
-The final feature we consider is computed by analyzing the mentions of dates in the articles.
-Using an extensive set of regular expressions, we are able to detect mentions of dates in different forms.
-To compute the feature \texttt{sharedDates} we start by identifying the list of dates mentioned in articles in each cluster separately.
-The value of \texttt{sharedDates} is then determined as the Jaccard similarity on the two lists of dates.
-
+From the list of relevant concepts and corresponding weights, we consider the following features:
+
+\begin{itemize}
+\item \texttt{entityCosSim} is the cosine similarity between vectors of entities from clusters $c_i$ and $c_j$.
+\item \texttt{keywordCosSim} is the cosine similarity between vectors of keywords from clusters $c_i$ and $c_j$.
+\item \texttt{entityJaccardSim} is  Jaccard similarity coefficient~\cite{levandowsky1971} between sets of entities from clusters $c_i$ and $c_j$.
+\item \texttt{keywordJaccardSim} is  Jaccard similarity coefficient between sets of keywords from clusters $c_i$ and $c_j$.
 \end{itemize}
 
-Using the  features described above and the user-provided labels, we train a classification model $M$ which is used in Algorithm~\ref{cluster_merge_algo2} to classify a new cluster pair as equivalent or not.
-The classification algorithm that we used to train a model was a linear Support Vector Machine (SVM) method \shortcite{shawe-taylor04kernel}.
+The last group of features contains three {\bf miscellaneous features} that seem discriminative but are unrelated to the previous two groups:
+\begin{itemize}
+\item \texttt{hasSameLocation} feature is a boolean variable that is true when the location of the event in both clusters is the same.
+Location of events is estimated by considering the locations mentioned in the articles from a cluster and is provided by Event Registry.
+\item \texttt{timeDiff} is the absolute difference in hours between the two events.
+The publication time and date of the events is computed as the average publication time and date of all the articles and is provided by Event Registry.
+\item \texttt{sharedDates} is determined as the Jaccard similarity coefficient between sets of date mentiones extracted from articles.
+We use extracted mentions of dates provided by Event Registry, which uses an extensive set of regular expressions to detect and normalize mentions of dates in different forms.
+\end{itemize}
 
@@
 
+We created a manually lablelled dataset in order to evaluate cross-lingual event linking using two human annotators.
+The annotators were provided with an interface listing the articles, their content from and top concepts for a pair of clusters and their task was to determine if the clusters were equivalent or not (i.e.
+discuss same event).
+To obtain a pair of clusters $(c_i, c_j)$ to annotate, we first randomly chose a cluster $c_i$, used the Algorithm~\ref{cluster_merge_algo1} to compute a set of potentially equivalent clusters $C$ and randomly chose a cluster $c_j \in C$.
+The dataset provided by the annotators contains 808 examples, of which 402 are equivalent clusters pairs and 406 are not.
+Clusters in each learning example are either in English, Spanish or German.
+Although Event Registry imports articles in other languages as well, we restricted our experiments to these three languages.
+We chose only these three languages since they have very large number of articles and clusters per day which makes the cluster linking problem hard due to large number of possible links.
+
 % which method?

### 2015-09-03 rupnik2015.tex
updates

--- 2015-09-03 rupnik2015.tex
@@
 \centering
-\includegraphics[width=0.75\textwidth]{events2.png}
-\caption{\label{fig:event2} Events are represented by collections of articles about an event, in this case the Malaysian airliner which was shot down over the Ukraine.}
+\includegraphics[width=1\textwidth]{events2.png}
+\caption{\label{fig:event2} Events are represented by collections of articles about an event, in this case the Malaysian airliner which was shot down over the Ukraine.
+The results shown in the figure can be obtained using the query \url{http://eventregistry.org/event/997350\#?lang=eng\&tab=articles}.}
 \end{figure}
@@
 For clustering, each new article is first tokenized, stop words removed and the remaining words are stemmed.
-The remaining tokens are represented in a vector-space model and normalized using TF-IDF (see Section~\ref{sec:tfidf} for the definition).
+The remaining tokens are represented in a vector-space model and normalized using TF-IDF\footnote{The IDF weights are dynamically computed for each new article over all news articles within a 10 day window.} (see Section~\ref{sec:tfidf} for the definition).
 Cosine similarity is used to find the most similar existing cluster, by comparing the document's vector to the centroid vector of each cluster.
@@
 \textbf{Matrix factorization}.
-Non-negative matrix factorization based \cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CLLSI\shortcite{cl_lsi}, Canonical Correlation Analysis (CCA) \cite{Hotelling}, Oriented Principal Component Analysis \cite{platt2010translingual}.
+Non-negative matrix factorization based \cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CL-LSI\shortcite{cl_lsi}\cite{multilingualBook}, Canonical Correlation Analysis (CCA) \cite{Hotelling}, Oriented Principal Component Analysis \cite{platt2010translingual}.
 The quadratic time and space dependency of the OPCA method makes it impractical for large scale purposes.
@@
 For our setting, the method in \cite{nonnegfactor_lsi} has a prohibitively high computational cost when building models (it uses dense matrices of dimensions that is the product of training set size and vocabulary size).
-Our proposed approach combines CCA and CLLSI.
+Our proposed approach combines CCA and CL-LSI.
+Another closely related method is Explicit Semantic Analysis (ESA) \cite{ESA}, which uses Wikipedia (as do we in the current work) to compare documents.
+The authors of ESA compare it to CL-LSI and find that CL-LSI can outperform ESA in an information retrieval, but is costlier to optimize over a large corpus (ESA requires no training).
+We find that the scalability argument does not apply in our case: based on advances in numerical linear algebra we can solve large CL-LSI problems (millions of documents as oposed to 10,000 document limit reported in \cite{ESA}).
 
@@
 
+\subsection{Repeatability of Experiments}
+We have made publicly available at \url{https://github.com/rupnikj/jair_paper.git} both the code and data that were used in the experiments.
+The manually labelled dataset used in the evaluation of event linking is available at in the ``dataset`` subfolder of the github repository.
+The included archive contains two folders: ``positive'' and ``negative'', where the first folder includes examples of cluster pairs in two languages that represent the same event and the second folder contains pairs of clusters in two languages that do not represent different events.
+Each example is a JSON file that contains at the top level information about a pair of clusters (including text of the articles) as well as a set of ``meta'' attributes, that correspond to features described in \ref{trainClassifier}.
+
+The ``code'' folder includes MATLAB scripts for building cross-lingual similarity models introduced in \ref{sec:models}, which can be used with publicly available Wikipedia corpus to reproduce the cross-lingual similarity evaluation.
+We have also made available the similarity computation over 100 languages as a service at \url{xling.ijs.si}.
+
+In addition, the Event Registry system (\url{http://eventregistry.org/}) comes with an API, documented at \url{https://github.com/gregorleban/event-registry-python}, that can be used to download events and articles.
+
 \section{Discussion and Future Work}
-\todo{We should mention the api, datasetswhatever here - for reprodicability}
 

### 2015-09-03 rupnik2015.tex
idf

@@
 An \emph{Inverse Document Frequency} ($IDF$) weight for the dictionary term $k$ is defined as $\log\left( \frac{N}{DF_k} \right)$, where $DF_k$ is the number of documents in the corpus which contain term $k$.
-\todo{our corpus is dynamic how do we deal with this?} Therefore we can definte a document's $TFIDF$ as
+When building cross-lingual models, the IDF scores were computed with respect to the Wikipedia corpus.
+In the other part of our system, we computed TFIDF vectors on streams of news documents in multiple languages.
+There the IDF scores for each language changed dynamically - for each new document we computed the IDF of all news articles within a 10 day window.
+
+Therefore we can define a document's $TFIDF$ as
 $$ x_{ij}  := \frac{\mbox{term frequency in document } i}{\mbox{inverse document frequency of term } j}.$$

### 2015-09-03 code/UBow/app_out.cpp
Added the code

@@ -1,0 +1,273 @@
+#include <base.h>
+#include <mine.h>
+#include <algorithm>
+#include <functional>
+#include <iostream>
+
+TStrV LangV;
+TIntStrH IdUriH;
+THash<TInt, THash<TInt, TFltIntKdV> > SimH;
+
+void PrintVec(TFltV& Vec){
+	int n = Vec.Len();
+	for (int i = 0; i < n; i++){
+		printf("%g\t", Vec[i].Val);
+		printf("\n");
+	}
+}
+
+void my_partial_sort(double* v, double* result, int *idx, const int len, const int& top = 100) {
+	// initialize original index locations
+	// sort indexes based on comparing values in v
+	std::partial_sort(idx, idx + top, idx + len,
+		[&v](size_t i1, size_t i2) {return v[i1] > v[i2]; });
+	for (int i = 0; i < top; i++){
+		result[i] = v[idx[i]];
+	}
+}
+
+//What are the most similar articles for doc in X1 in X2 with language id LangId
+void CalcSim(const TFltVV& X1, const TFltV& IdV1, const TFltVV& X2, 
+        const TFltV& IdV2, const int& LangId, const int& MaxIdDiff, 
+        const int& MaxN) {
+    
+    const int Len1 = X1.GetCols();
+    const int Len2 = X2.GetCols();
+    // placeholder for similarities
+    TFltIntKdV SimIdV; int LastStart = 0;
+    // major loop
+    //for (int i = 10000; i < 10100; i++) {
+    for (int i = 1; i < Len1; i++) {
+        if (i % 100 == 0) { printf("%6d / %6d\n", i, Len1); }
+        const int Id1 = (int)IdV1[i];
+        // move start pointer to the first element within id window
+        for (int j = LastStart; j < Len2; j++) {
+            const int Id2 = (int)IdV2[j];
+            // should stop once within id window
+            if (Id1 - Id2 < MaxIdDiff) { break; }
+            // else move on
+            LastStart++;
+        }
+        // compute similarities
+        for (int j = LastStart; j < Len2; j++) {
[…]

### 2015-09-04 rupnik2015.tex
Grammar and various corrections from Section 7. Changed feature vector to topic vector to avoid confusion.
Some minor changes are still pending. Primoz should take a look.
Only black - are not corrected.

@@
 Significant events are reported by different sources and in different languages.
-In this work, we address the problem of tracking and events in a large multilingual stream.
+In this work, we address the problem of tracking of events in a large multilingual stream.
 Within a recently developed system (EventRegistry  \shortcite{Leban2014W,Leban2014I}), we examine two aspects of this problem: how to compare articles in different languages and how to link collections of articles in different languages which refer to the same event.
-Taking a multi-lingual stream and clusters of articles from each language, we compare different cross-lingual document similarity measures based on Wikipedia.
+Taking a multilingual stream and clusters of articles from each language, we compare different cross-lingual document similarity measures based on Wikipedia.
 This allows us to compute the similarity of any two articles regardless of language.
@@
 
-Content on the internet is becoming increasingly multi-lingual.
+Content on the internet is becoming increasingly multilingual.
 A prime example is Wikipedia - in 2001, the majority of pages were written in English, while in 2015, the percentage of English articles has dropped to 14\%.
@@
 \includegraphics[width=1\textwidth]{events2.png}
-\caption{\label{fig:event2} Events are represented by collections of articles about an event, in this case the Malaysian airliner which was shot down over the Ukraine.
+\caption{\label{fig:event2} Events are represented by collections of articles about an event, in this case the Malaysian airliner which was shot down over Ukraine.
 The results shown in the figure can be obtained using the query \url{http://eventregistry.org/event/997350\#?lang=eng\&tab=articles}.}
@@
 The first approach represents a continuation of previous work \shortcite{nips,sikdd,nips2,iti} where we explored representations of documents which were valid over multiple languages.
-The representations could be interpreted as multi-lingual topics, which were then used as proxies to compute cross-lingual similarities between documents.
+The representations could be interpreted as multilingual topics, which were then used as proxies to compute cross-lingual similarities between documents.
 To learn the representations, we use Wikipedia as a training corpus.
@@
 We can then define a similarity between any two articles regardless of language,  which allows us to cluster the articles according to topic.
-The underlying assumption being that articles describing the same event are similar  and will therefore be put into the same cluster.
+The underlying assumption is that articles describing the same event are similar  and will therefore be put into the same cluster.
 
@@
 We first introduce the problem of cross-lingual similarity computation (Section~\ref{sec:crosslingual}), which is followed by several approaches for measuring cross-lingual similarity (Section~\ref{sec:models}) and then the algorithms for matching clusters in different languages (Section~\ref{sec:linking}).
-To improve readability, we describe relevant related work for each module in its corresponding Section.
+To improve readability, we describe relevant related work for each module in its corresponding section.
 Finally, we present and interpret the experimental results, give conclusions and describe promising future directions.
@@
 In the next step, groups of articles about the same event are identified and relevant information about the event is extracted (Event construction phase).
-Although the pipeline contains several components, we will in this paper focus only on the two highlighted in the image.}
+Although the pipeline contains several components, we focus only on the two highlighted in the image.}
 \end{figure}
@@
 The last important processing step on the document level is to efficiently identify which articles in other available languages are most similar to this article.
-The methodology for this task is one of the main contributions of this paper and is described in details in Section~\ref{sec:crosslingual}).
+The methodology for this task is one of the main contributions of this paper and is explained in details in Section~\ref{sec:crosslingual}).
 
@@
 \item \textbf{Training} - The training (building cross-lingual models) should scale to many languages and should be robust to the quality of training resources.
-The system should be able to take advantage of comparable corpus (as opposed to parallel translation based corpora), with missing data.
+The system should be able to take advantage of comparable corpora (as opposed to parallel translation-based corpora), with missing data.
 Supporting a new language should be straightforward.
@@
 The issue with using online services such as Google Translate is that the APIs are limited and not free.
-The operation efficiency and cost requirements make translation based approaches less suited for our system.
+The operation efficiency and cost requirements make translation-based approaches less suited for our system.
 
@@
 
-\textbf{Classification problem related}.
+\textbf{Classification problem-related}.
 In approaches such as \cite{cotrain} and \shortcite{svm2k} the authors find linear embeddings of the data that support good classification results, based on a given classification problem training data.
@@
 \textbf{Matrix factorization}.
-Non-negative matrix factorization based \cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CL-LSI\shortcite{cl_lsi}\cite{multilingualBook}, Canonical Correlation Analysis (CCA) \cite{Hotelling}, Oriented Principal Component Analysis \cite{platt2010translingual}.
+Non-negative matrix factorization-based \cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CL-LSI\shortcite{cl_lsi}\cite{multilingualBook}, Canonical Correlation Analysis (CCA) \cite{Hotelling}, Oriented Principal Component Analysis \cite{platt2010translingual}.
 The quadratic time and space dependency of the OPCA method makes it impractical for large scale purposes.
@@
 Based on our requirements in Section~\ref{sec:sysreq}, we chose to focus on methods based on vector space models and linear embeddings.
-We propose a method that is more efficient than popular alternatives (a clustering based approach and latent semantic indexing), but is still simple to optimize and use.
+We propose a method that is more efficient than popular alternatives (a clustering-based approach and latent semantic indexing), but is still simple to optimize and use.
 
@@
 Google News as well as Yahoo!
-News are able to identify clusters of articles about same event, but offer no linking of clusters across languages.
+News are able to identify clusters of articles about the same event, but offer no linking of clusters across languages.
 The only service that we found, which provides cross-lingual cluster linking, is the European Media Monitor (EMM)~\cite{pouliquen2008story,Steinberger2008}.
@@
 To achieve cluster linking, EMM uses three different language independent vector representations for each cluster.
-First vector contains the weighted list of references to countries mentioned in the articles, while the second vector contains the weighted list of mentioned people and organizations.
+The first vector contains the weighted list of references to countries mentioned in the articles, while the second vector contains the weighted list of mentioned people and organizations.
 The last vector contains the weighted list of Eurovoc subject domain descriptors.
@@
 Many techniques use the similarity as a black box, e.g., a kernel in Support Vector Machines.
-Comparison of documents (or other types of text snippets) in a monolingual setting is a well studied problem in the field of information retrieval \cite{Salton88term-weightingapproaches}.
+Comparison of documents (or other types of text snippets) in a monolingual setting is a well-studied problem in the field of information retrieval \cite{Salton88term-weightingapproaches}.
 We first formally introduce the problem followed by our approach.
@@
 \noindent\textbf{Document representation.}
-The standard vector space model \cite{Salton88term-weightingapproaches} represents documents as vectors, where each term corresponds to word or phrase in a fixed vocabulary.
+The standard vector space model \cite{Salton88term-weightingapproaches} represents documents as vectors, where each term corresponds to a word or a phrase in a fixed vocabulary.
 Formally, document $d$ is represented by a vector $x \in \RR^n$, where $n$ corresponds to the size of the vocabulary, and vector elements $x_k$ correspond to the number of times term $k$ occurred in the document, also called \emph{term frequency} or $TF_k(d)$.
@@
 where $\langle \cdot,\cdot \rangle$ and $\|\cdot\|$ are standard inner product and Euclidean norm.
-Cosine similarity, and other related approaches, assumes that the similarity is reflected in the overlap of words.
+Cosine similarity, and other related approaches, assume that the similarity is reflected in the overlap of words.
 While always defined, this metric is only meaningful when the documents $d_1$ and $d_2$ are defined on the same space (e.g.
@@
 where $d_1$ and $d_2$ are documents written in the $i$-th and $j$-th language respectively.
-If the the maximal singular value of $S_{i,j}$ is bounded by $1$, then the similarity scores will lie on the interval $[-1, 1]$.
+If the maximal singular value of $S_{i,j}$ is bounded by $1$, then the similarity scores will lie on the interval $[-1, 1]$.
 In Section \ref{sec:models} we will describe some approaches to computing $S_{i,j}$ given training data.
@@
 
-We concentrate on approaches that are based on linear maps rather than alternatives, such as machine translation and probabilistic models, as discussed in the related work section.
+We concentrate on approaches that are based on linear maps rather than alternatives, such as machine translation and probabilistic models, as discussed in the section on related work.
 We will start by introducing some notation.
@@
 
-The cross-lingual similarity functions are based on a rank-$k$ truncated SVD: $X \approx U \Sigma V^T,$ where $U \in \RR^{N \times k}$ are basis vectors of interest and $\Sigma \in \RR^{k \times k}$ is truncated diagonal matrix of singular eigenvalues.
+The cross-lingual similarity functions are based on a rank-$k$ truncated SVD: $X \approx U \Sigma V^T,$ where $U \in \RR^{N \times k}$ are basis vectors of interest and $\Sigma \in \RR^{k \times k}$ is a truncated diagonal matrix of singular eigenvalues.
 An aligned basis is obtained by first splitting $U$ vertically according to the number of dimensions of each language: $U = [U_1^T \cdots U_m^T]^T$.
@@
 %
- Canonical Correlation Analysis (CCA)~\cite{Hotelling} is a dimensionality reduction technique similar to Principal Component Analysis (PCA)~\cite{Pearson1901On}, with an additional assumption that the data consists of feature vectors that arose from two sources (two views) that share some information.
+ Canonical Correlation Analysis (CCA)~\cite{Hotelling} is a dimensionality reduction technique similar to Principal Component Analysis (PCA)~\cite{Pearson1901On}, with the additional assumption that the data consists of feature vectors that arose from two sources (two views) that share some information.
 Examples include: bilingual document collection ~\shortcite{mrpqr} and collection of images and captions~\shortcite{Hardoon_usingimage}.
@@
 Interpreting the columns of $X_i$ as observation vectors sampled from an underlying distribution $\mathcal{X}_i \in \RR^{n_i}$, the idea is to find two weight vectors $w_i \in \RR^{n_i}$ and $w_j \in \RR^{n_j}$ so that the random variables $w_i^T \cdot \mathcal{X}_i$ and $w_j^T \cdot \mathcal{X}_j$ are maximally correlated ($w_i$ and $w_j$ are used to map the random vectors to random variables, by computing weighted sums of vector components).
-Let $\rho(x,y)$ denote the sample based correlation coefficient between two vectors of observations $x$ and $y$.
+Let $\rho(x,y)$ denote the sample-based correlation coefficient between two vectors of observations $x$ and $y$.
 By using the sample matrix notation $X_i$ and $X_j$ (assuming no data is missing to simplify the exposition), this problem can be formulated as the following optimization problem:
@@
 We now describe an extension of CCA to more than two languages, which can be trained on large comparable corpora and can handle missing data.
- The extension we consider is based on a generalization of CCA to more than two views, introduced in \cite{Kettenring}, namely the Sum of Squared Correlation SSCOR, which we will state formally later in this section.
+ The extension we consider is based on a generalization of CCA to more than two views, introduced in \cite{Kettenring}, namely the Sum of Squared Correlations SSCOR, which we will state formally later in this section.
 Our approach exploits a certain characteristic of the data, namely the \emph{hub language} characteristic (introduced in the next paragraph) in two
@@
 
-The first step in our method is to project $X_1, \ldots, X_m$ to lower dimensional spaces without destroying the cross-lingual structure.
+The first step in our method is to project $X_1, \ldots, X_m$ to lower-dimensional spaces without destroying the cross-lingual structure.
 Treating the nonzero columns of $X_i$ as observation vectors sampled from an underlying distribution $\mathcal{X}_i \in V_i = \RR^{n_i}$, we can analyze the empirical cross-covariance matrices:
@@
  where $c_i = \frac{1}{a_i} \sum_{\ell \in a(i)}X_i^{\ell}$.
-By finding low rank approximations of $C_{i,j}$ we can identify the subspaces of $V_i$ and $V_j$ that are relevant for extracting linear patterns between $\mathcal{X}_i$ and $\mathcal{X}_j$.
+By finding low-rank approximations of $C_{i,j}$ we can identify the subspaces of $V_i$ and $V_j$ that are relevant for extracting linear patterns between $\mathcal{X}_i$ and $\mathcal{X}_j$.
 Let $X_1$ represent the hub language corpus matrix.
@@
 If $|a(i,j)|$ is small for many language pairs (as it is in the case of Wikipedia), then many empirical estimates $C_{i,j}$ are unreliable, which can result in overfitting.
-For this reason, we perform the truncated singular value decomposition on the matrix $C = [C_{1,2}  \cdots  C_{1,m}] \approx U S V^T$, where where $U \in \RR^{n_1 \times k}, S \in \RR^{k \times k}, V \in \RR^{(\sum_{i=2}^m n_i) \times k}$.
+For this reason, we perform the truncated singular value decomposition on the matrix $C = [C_{1,2}  \cdots  C_{1,m}] \approx U S V^T$, where $U \in \RR^{n_1 \times k}, S \in \RR^{k \times k}, V \in \RR^{(\sum_{i=2}^m n_i) \times k}$.
 We split the matrix $V$ vertically in blocks with $n_2, \ldots, n_m$ rows: $V = [V_2^T  \cdots  V_m^T]^T$.
@@
 The solutions for $y_i$ are obtained from (\ref{eqy1yi}): $y_i := \frac{G_{i}^T y_1}{\| G_{i}^T y_1 \|}$.
-Note that the solution (\ref{squaredCorHubOriginal}) to can be recovered by: $w_i := K_i^{-1} y_i$. The linear transformation of the $w$ variables are thus expressed as:
+Note that the solution (\ref{squaredCorHubOriginal}) can be recovered by: $w_i := K_i^{-1} y_i$. The linear transformation of the $w$ variables are thus expressed as:
 $$ Y_1 := \text{eigenvectors of} \sum_{i = 2}^m G_i G_{i}^T, $$
@@
 \textbf{Remark}
- The technique is related to  Generalization of Canonical Correlation Analysis (GCCA) by Carroll \citeyear{Carroll}, where an unknown group configuration variable is defined and objective is to maximize the sum of squared correlation between the group variable and the others.
+ The technique is related to  Generalization of Canonical Correlation Analysis (GCCA) by Carroll \citeyear{Carroll}, where an unknown group configuration variable is defined and the objective is to maximize the sum of squared correlation between the group variable and the others.
 The problem can be reformulated as an eigenvalue problem.
@@
 We treat this task as a supervised learning problem.
-For each candidate cluster $c_j \in C$, we compute a vector of learning features that should be indicative of whether the $c_i$ and $c_j$ are equivalent or not and apply a binary classification model that predicts if the clusters are equivalent or not.
+For each candidate cluster $c_j \in C$, we compute a vector of learning features that should be indicative of whether $c_i$ and $c_j$ are equivalent or not and apply a binary classification model that predicts if the clusters are equivalent or not.
 The classification algorithm that we used to train a model was a linear Support Vector Machine (SVM) method \shortcite{shawe-taylor04kernel}.
@@
 
-To investigate the empirical performance of the low rank approximations we will test the algorithms on a large-scale, real-world multi-lingual dataset that we extracted from Wikipedia by using inter-language links for alignment.
+To investigate the empirical performance of the low-rank approximations we will test the algorithms on a large-scale, real-world multilingual dataset that we extracted from Wikipedia by using inter-language links for alignment.
 This  results in a large number of weakly comparable documents in more than $200$ languages.
-Wikipedia is a large source of multi-lingual data that is especially important for the languages for which no translation tools, multi-lingual dictionaries as Eurovoc \cite{eurovoc}, or strongly aligned multi-lingual corpora as Europarl \cite{europarl} are available.
+Wikipedia is a large source of multilingual data that is especially important for the languages for which no translation tools, multilingual dictionaries as Eurovoc \cite{eurovoc}, or strongly aligned multilingual corpora as Europarl \cite{europarl} are available.
 Documents in different languages are related with so called 'inter-language' links that can be found on the left of the Wikipedia page.
@@
 First, we check if the title of the page consists of any special namespace and do not process such pages.
-Then, we check if this is a redirection page and we store the redirect link as inter-language links can point to redirection link also.
+Then, we check if this is a redirection page and we store the redirect link because inter-language links can point to redirection links also.
 If none of the above applies, we extract the text and parse the Wikipedia markup.
@@
 We get inter-language link matrix using previously stored redirection links and inter-language links.
-If inter-language link points to the redirection we replace it with the redirection target link.
+If an inter-language link points to the redirection we replace it with the redirection target link.
 It turns out that we obtain the matrix $M$ that is not symmetric, consequently the underlying graph is not symmetric.
@@
 English to German) does not guarantee that there is an inter-language link in the reverse direction (German to English).
-To correct this we transform this matrix to symmetric by computing $M+M^T$ and obtaining an undirected graph.
+To correct this we transform this matrix to be symmetric by computing $M+M^T$ and obtaining an undirected graph.
 In the rare case that after symmetrization we have multiple links pointing from the document, we pick the first one that we encountered.
@@
  To this purpose, we select a subset of Wikipedia languages containing three major languages, English (4,212k articles)--\emph{en} (hub language), Spanish (9,686k articles)--\emph{es}, Russian (9,662k articles)--\emph{ru}, and five minority (in the sense of Wikipedia sizes) languages, Slovenian (136k articles)--\emph{sl}, Piedmontese (59k articles)--\emph{pms}, Waray-Waray (112k articles)--\emph{war} (all with about 2 million native speakers), Creole (54k articles)--\emph{ht} (8 million native speakers), and Hindi (97k articles)--\emph{hi} (180 million native speakers).
-For preprocessing, we remove the documents that contain less than 20 different words (stubs \footnote{This documents are typically garbage, the titles of the columns in the table, remains of the parsing process, or Wikipedia articles with very little or no information contained in one or two sentences.}) and remove words occurring in less than 50 documents as well as the top 100 most frequent words (in each language separately).
+For preprocessing, we remove the documents that contain less than 20 different words (stubs \footnote{Such documents are typically garbage, the titles of the columns in the table, remains of the parsing process, or Wikipedia articles with very little or no information contained in one or two sentences.}) and remove words occurring in less than 50 documents as well as the top 100 most frequent words (in each language separately).
 We represent the documents as normalized TFIDF\cite{Salton88term-weightingapproaches} weighted vectors.
@@
 The evaluation is based on splitting the data into training and test sets. %(which are described later).
-We select the test set documents as all multi-lingual documents with at least one nonempty alignment from the list: (\emph{hi}, \emph{ht}), (\emph{hi}, \emph{pms}), (\emph{war}, \emph{ht}), (\emph{war}, \emph{pms}).
+We select the test set documents as all multilingual documents with at least one nonempty alignment from the list: (\emph{hi}, \emph{ht}), (\emph{hi}, \emph{pms}), (\emph{war}, \emph{ht}), (\emph{war}, \emph{pms}).
 This guarantees that we cover all the languages.
-Moreover this test set is suitable for testing the retrieval thorough the hub as the chosen pairs have empty alignments.
+Moreover this test set is suitable for testing the retrieval through the hub as the chosen pairs have empty alignments.
 The remaining documents are used for training.
@@
 \caption{Pairwise training -- test sizes (in thousands)
-The first row represents the size of the training sets used to construct the mappings in low dimensional language independent space using the English--\emph{en} as a hub.
-The diagonal elements represent number of the unique training documents and test documents in each language.
+The first row represents the size of the training sets used to construct the mappings in low-dimensional language independent space using the English--\emph{en} as a hub.
+The diagonal elements represent the number of the unique training documents and test documents in each language.
 }
@@
 
-We further inspect the properties of the training sets by roughly estimating the fraction \texttt{rank(A)/min(size(A))} for each training English matrix and its corresponding mate matrix.
+We further inspect the properties of the training sets by roughly estimating the fraction \texttt{rank(A)/min(size(A))} for each English training matrix and its corresponding mate matrix.
 Ideally, these two fractions are approximately the same so  both aligned spaces should have reasonably similar dimensionality.
@@
 
-It is clear that in the case of Creole language only at most $22\%$ documents are unique and suitable for the training.
-Though we removed the stub documents, many of remaining documents are nearly the same, as the quality of some smaller Wikipedias is low.
+It is clear that in the case of the Creole language only at most $22\%$ documents are unique and suitable for the training.
+Though we removed the stub documents, many of the remaining documents are nearly the same, as the quality of some smaller Wikipedias is low.
 This was confirmed for the Creole, Waray-Waray, and Piedmontese languages by manual inspection.
@@
 There is also a problem with the quality of the test data.
-For example, if we look at test pair (\emph{war}, \emph{ht}) only 386/534 Waray-Waray test documents are unique but on other side almost all Creole test documents (523/534) are unique.
+For example, if we look at the test pair (\emph{war}, \emph{ht}) only 386/534 Waray-Waray test documents are unique but on the other side almost all Creole test documents (523/534) are unique.
 This indicates a poor alignment which leads to poor performance.
@@
 \subsection{Evaluation Of Cross-Lingual Event Linking}
-In order to determine how accurately we can predict cluster equivalence, we performed two experiments in multi-lingual setting, English, German and Spanish language for which we had labelled data to evaluate the linking performance.
+In order to determine how accurately we can predict cluster equivalence, we performed two experiments in a multilingual setting using the English, German and Spanish language for which we had labelled data to evaluate the linking performance.
 In the first experiment, we tested how well  the individual approaches for cross-lingual article linking perform when used for linking the clusters about the same event.
-In the second experiment we tested how accurate is the prediction model when trained on different subsets of learning features.
+In the second experiment we tested how accurate the prediction model is when trained on different subsets of learning features.
 To evaluate the prediction accuracy for a given dataset we used 10-fold cross validation.
@@
 
-We created a manually lablelled dataset in order to evaluate cross-lingual event linking using two human annotators.
+We created a manually labelled dataset in order to evaluate cross-lingual event linking using two human annotators.
 The annotators were provided with an interface listing the articles, their content from and top concepts for a pair of clusters and their task was to determine if the clusters were equivalent or not (i.e.
@@
 %The table shows for each of the algorithms the obtained classification accuracy, precision and recall.
-We can see that the most informative features for cluster linking are obtained using the hub CCA algorithm, while $k$-means performs the worse.
+We can see that the most informative features for cluster linking are obtained using the hub CCA algorithm, while $k$-means performs the worst.
 
@@
 It is clear that hub CCA outperforms LSI approach and K-means lags far behind when testing on Wikipedia data.
-The hub CCA approach with $500$ features manages to perform comparably to $LSI$ based approach with $1000$-features, which shows that the $CCA$ method can improve both model memory footprint as well as similarity computation time.
+The hub CCA approach with $500$ topic vectors manages to perform comparably to the $LSI$-based approach with $1000$ topic vectors, which shows that the $CCA$ method can improve both model memory footprint as well as similarity computation time.
 
 % number of topics?
-Furthermore, we inspected how the number of features (topics) influences the accuracy of cluster linking.
-As we can see from Table~\ref{table:linkingEvalAlgos} choosing number of feature vectors larger than $500$ barely affects linking performance, which is in contrast with the fact that additional features helped improve AMMR, see Figure~\ref{pic:AMRR}.
+Furthermore, we inspected how the number of topics influences the accuracy of cluster linking.
+As we can see from Table~\ref{table:linkingEvalAlgos} choosing number of topic vectors larger than $500$ barely affects linking performance, which is in contrast with the fact that additional topics helped to improve AMMR, see Figure~\ref{pic:AMRR}.
 Such differences may have arisen due to different domains of training and testing (Wikipedia pages versus news articles).
@@
 We
-store one day buffer for each language which requires 1.5 GB of memory with documents   stored as 500 dimensional vectors.
+store one day buffer for each language which requires 1.5 GB of memory with documents   stored as 500-dimensional vectors.
 We  note that the time complexity of the similarity computations scales linearly with dimension of the feature space and does not  depend on number of languages.
@@
 For all linear algebra matrix and vector operations, we use high performance numerical linear algebra libraries as BLAS, OPENBLAS and Intel MKL, which currently allows us to process more than one million articles per day.
-In our current implementation, we use the variation of hub approach.
+In our current implementation, we use the variation of the hub approach.
 Our projector matrices are of size $500\times 300,000,$ so every projector takes about $1.1$ GB of RAM.
@@
 All together we need around 135 GB of RAM for the system with 100 languages.
- Usage of proxy matrices enables projection of all input documents in the common space and handling language pairs with missing or low alignment.
+ Usage of proxy matrices enables the projection of all input documents in the common space and handling language pairs with missing or low alignment.
 That enables us to do block-wise similarity computations further improving system efficiency.
 Our code can therefore be easily parallelized using matrix multiplication rather than performing more matrix - vector multiplications.
-This speeds up our code by a factor around $4.$ In this way, we obtain some caching gains and ability to use vectorization.
+This speeds up our code by a factor of around $4.$ In this way, we obtain some caching gains and ability to use vectorization.
 Our system is also easily extendable. Adding a new language requires the  computation of  a projector matrix and proxy matrices with all other already available languages.
@@
 
-On the task of mate retrieval we observe that refining the LSI based projections with hub CCA leads to improved retrieval precision, but the methods perform comparably on the task of event linking.
-Further inspection showed that the CCA based approach reached a higher precision on smaller clusters.
+On the task of mate retrieval we observe that refining the LSI-based projections with hub CCA leads to improved retrieval precision, but the methods perform comparably on the task of event linking.
+Further inspection showed that the CCA-based approach reached a higher precision on smaller clusters.
 The interpretation is that the linking features are highly aggregated for large clusters, which compensates the lower per-document precision of LSI.
@@
 
-The experiments show that the hub CCA based features present good baseline, which can greatly benefit from additional semantic-based features.
+The experiments show that the hub CCA-based features present a good baseline, which can greatly benefit from additional semantic-based features.
 Even though the CCA-based features do not greatly improve the system performance when the semantic features are used, there are two main benefits in the approach: the linking process can be sped up (smaller set of candidate clusters), and usage in the case when semantic extraction is not possible.
@@
 
-Currently the system is loosely-coupled -- the language component is built independently of the rest of the system, in particular the linking component.
+Currently the system is loosely-coupled -- the language component is built independently from the rest of the system, in particular the linking component.
 It is possible that better embeddings can be obtained by methods that jointly optimize a classification task and the embedding.

### 2015-09-04 rupnik2015.tex
related work

--- 2015-09-03 rupnik2015.tex
@@
 The authors of ESA compare it to CL-LSI and find that CL-LSI can outperform ESA in an information retrieval, but is costlier to optimize over a large corpus (ESA requires no training).
-We find that the scalability argument does not apply in our case: based on advances in numerical linear algebra we can solve large CL-LSI problems (millions of documents as oposed to 10,000 document limit reported in \cite{ESA}).
+We find that the scalability argument does not apply in our case: based on advances in numerical linear algebra we can solve large CL-LSI problems (millions of documents as opposed to 10,000 document limit reported in \cite{ESA}).
 
@@
 When building cross-lingual models, the IDF scores were computed with respect to the Wikipedia corpus.
-In the other part of our system, we computed TFIDF vectors on streams of news documents in multiple languages.
+In the other part of our system, we computed TFIDF vectors on streams of news articles in multiple languages.
 There the IDF scores for each language changed dynamically - for each new document we computed the IDF of all news articles within a 10 day window.

### 2015-09-04 rupnik2015.tex

--- 2015-09-04 rupnik2015.tex
@@
 \textbf{Matrix factorization}.
-Non-negative matrix factorization-based \cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CL-LSI\shortcite{cl_lsi}\cite{multilingualBook}, Canonical Correlation Analysis (CCA) \cite{Hotelling}, Oriented Principal Component Analysis \cite{platt2010translingual}.
+Non-negative matrix factorization based \cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CL-LSI\shortcite{cl_lsi}\cite{multilingualBook}, Canonical Correlation Analysis (CCA) \cite{Hotelling}, Oriented Principal Component Analysis \cite{platt2010translingual}.
 The quadratic time and space dependency of the OPCA method makes it impractical for large scale purposes.

### 2015-09-04 rupnik2015.tex
related work

@@
 \caption{\label{fig:event2} Events are represented by collections of articles about an event, in this case the Malaysian airliner which was shot down over Ukraine.
-The results shown in the figure can be obtained using the query \url{http://eventregistry.org/event/997350\#?lang=eng\&tab=articles}.}
+The results shown in the figure can be obtained using the query \url{http://eventregistry.org/event/997350\#?lang=eng\&tab=articles}.
+The content presented is part of the Event Registry system, developed by the authors.}
 \end{figure}
@@
 \item \textbf{Operation efficiency} - The similarity computation should be fast - the system must be able to handle billions of similarity computations per day.
+Computing the similarity between a new document and a set of known documents should be efficient (the main application is linking documents between two monolingual streams).
 \item \textbf{Operation cost} - The system should run on a strong shared machine server and not rely on paid services.
@@
 
-\textbf{Translation}.
-The most obvious way to compare documents written in different languages is to use machine translation and perform monolingual similarity.
+\textbf{Translation and dictionary based}.
+The most obvious way to compare documents written in different languages is to use machine translation and perform monolingual similarity, see  \cite{multilingualBook}\cite{plagiarism} for several variations of translation based approaches.
 One can use free tools such as Moses \shortcite{moses} or translation services, such as Google Translate\footnote{https://translate.google.com/}.
@@
 The operation efficiency and cost requirements make translation-based approaches less suited for our system.
+CL-VSM \cite{plagiarism} and the approach presented in \cite{pouliquen2008story} compare documents by using dictionaries, such as EuroVoc \cite{eurovoc}.
+The generality of such approaches is limited by the quality of available linguistic resources, which may be scarce or non-existent for certain language pairs.
 
@@
 Another closely related method is Explicit Semantic Analysis (ESA) \cite{ESA}, which uses Wikipedia (as do we in the current work) to compare documents.
+It can be interpreted as using the sample covariance matrix between features of two languages to define the dot product that is used to compute similarities.
 The authors of ESA compare it to CL-LSI and find that CL-LSI can outperform ESA in an information retrieval, but is costlier to optimize over a large corpus (ESA requires no training).
 We find that the scalability argument does not apply in our case: based on advances in numerical linear algebra we can solve large CL-LSI problems (millions of documents as opposed to 10,000 document limit reported in \cite{ESA}).
+In addition, ESA is less suited for computing similarities between two large monolingual streams.
+For example, each day we have to compute similarities between 500,000 English and 500,000 German news articles.
+Comparing each German news article with 500,000 English news articles is either prohibitively slow (involves projecting all English articles on Wikipedia) or consumes too much memory (involves storing the projected English articles, which for a Wikipedia of size 1,000,000 is a 500,000 by 1000,0000 non-sparse matrix).
+
+\textbf{Monolingual}.
+Finally, related work includes monolingual approaches that treat document written in different languages in a monolingual fashion.
+The intuition is that named entities (for example, ``Obama'') and cognate words (for example, ``tsunami'') are written in the same or similar fashion in many languages.
+For example, the Cross-Language Character n-Gram Model (CL-CNG) \cite{plagiarism} represents documents as bags of character n-grams.
+Another approach is to use language dependent keyword lists based on cognate words \cite{pouliquen2008story}.
+These approaches may be suitable for comparing documents written in languages that share a writing system, which does not apply to the case of global news tracking.
 

### 2015-09-04 rupnik2015.tex
related work

@@
 
-There are four main families of approaches to cross-lingual similarity.
+There are five main families of approaches to cross-lingual similarity.
 
@@
 The operation efficiency and cost requirements make translation-based approaches less suited for our system.
-CL-VSM \cite{plagiarism} and the approach presented in \cite{pouliquen2008story} compare documents by using dictionaries, such as EuroVoc \cite{eurovoc}.
+Closely related are works Cross-Lingual Vector Space Model (CL-VSM) \cite{plagiarism} and the approach presented in \cite{pouliquen2008story} which both compare documents by using dictionaries, which in both cases are EuroVoc dictionaries \cite{eurovoc}.
 The generality of such approaches is limited by the quality of available linguistic resources, which may be scarce or non-existent for certain language pairs.

### 2015-09-04 rupnik2015.tex
Updated pipeline figure to vector format.
A lot of small fixes. I could not properly answer some TODOS and i also added some.

--- 2015-09-04 rupnik2015.tex
@@
 
-Collected articles are first semantically annotated by identifying in them mentions of relevant concepts -- either entities or important keywords.
+Collected articles are first semantically annotated by identifying mentions of relevant concepts -- either entities or important keywords.
 The disambiguation and entity linking of the concepts is done using Wikipedia as the main knowledge base.
@@
 The underlying assumption is that articles that are describing the same event are similar enough and will therefore be put into the same cluster.
-For clustering, each new article is first tokenized, stop words removed and the remaining words are stemmed.
+For clustering, each new article is first tokenized, stop words are removed and the remaining words are stemmed.
 The remaining tokens are represented in a vector-space model and normalized using TF-IDF\footnote{The IDF weights are dynamically computed for each new article over all news articles within a 10 day window.} (see Section~\ref{sec:tfidf} for the definition).
@@
 The authors of ESA compare it to CL-LSI and find that CL-LSI can outperform ESA in an information retrieval, but is costlier to optimize over a large corpus (ESA requires no training).
-We find that the scalability argument does not apply in our case: based on advances in numerical linear algebra we can solve large CL-LSI problems (millions of documents as opposed to 10,000 document limit reported in \cite{ESA}).
+We find that the scalability argument does not apply in our case: based on advances in numerical linear algebra we can solve large CL-LSI problems (millions of documents as oposed to 10,000 document limit reported in \cite{ESA}).
 
@@
 When building cross-lingual models, the IDF scores were computed with respect to the Wikipedia corpus.
-In the other part of our system, we computed TFIDF vectors on streams of news articles in multiple languages.
+In the other part of our system, we computed TFIDF vectors on streams of news documents in multiple languages.
 There the IDF scores for each language changed dynamically - for each new document we computed the IDF of all news articles within a 10 day window.
@@
 where $\langle \cdot,\cdot \rangle$ and $\|\cdot\|$ are standard inner product and Euclidean norm.
-Cosine similarity, and other related approaches, assume that the similarity is reflected in the overlap of words.
+Cosine similarity and other related approaches assume that the similarity is reflected in the overlap of words.
 While always defined, this metric is only meaningful when the documents $d_1$ and $d_2$ are defined on the same space (e.g.
@@
 \todo{add some more details/intuition.}
+\begin{comment}
+Technical detail.
+Map everything in the english space.
+$S_{1, 2} = Q^{en}_1^T Q^{en}_2,$ where  Q^{en}_1 and Q^{en}_2 are orthogonal matrices.
+All non zero singular values are thus automatically equal to one.
+\end{comment}
 
@@
  to reformulate a difficult optimization problem as an eigenvector problem. The second ingredient is dimensionality reduction using LSI, which
- makes the eigenvector problem computationally and numerically tractable. \todo{we need to set this up better. Revised by Jan. Check if OK}
+ makes the eigenvector problem computationally and numerically tractable. \todo{we need to set this up better. Revised by Jan. Check if OK. Andrej checked}
 
@@
 \caption{\label{fig:lsi} LSI multilingual corpus matrix decomposition.}
+\todo{Missing matrix S of singular values!!!}
 \end{figure}
@@
 \textbf{Hub language characteristic}
-In Wikipedia, we observed that even though the training resources are scarce between certain language pairs, there often exists indirect training data through what we refer to as a hub language.
+In the case of Wikipedia we observed that even though the training resources are scarce for certain language pairs, there often exists indirect training data that we can use as a proxy -- hub language.\todo{Check this, I prettified it.}
 
@@
 \begin{equation}\label{dLdx1}
- \frac{\partial}{\partial x_1} L = 0 \Rightarrow \sum_{i = 2}^m  \left(y_1^T G_{i} y_i \right) G_i y_i + \lambda_1 y_1,
+ \frac{\partial}{\partial x_1} L = 0 \Rightarrow \sum_{i = 2}^m  \left(y_1^T G_{i} y_i \right) G_i y_i + \lambda_1 y_1 = 0,
 \end{equation}
@@
 This is partly due to news coverage which might be more granular in some languages, partly due to noise and errors in the event detection process.
-This implies that we cannot make assumptions on the matching (e.g., one-to-one or complete matchin) and excludes the use of standard weighted bipartite matching type of algortihms for this problem.
+This implies that we cannot make assumptions on the matching (e.g., one-to-one or complete matching) and excludes the use of standard weighted bipartite matching type of algorithms for this problem.
 An example is shown in Figure~\ref{fig:clusters}, where a cluster may contain articles which are closely matched with many clusters in a different language.
@@
 
-Although computed document similarities are approximate, our  assumption is that articles in different languages describing the same event, will generally have a higher similarity than articles about different events.
+Although computed document similarities are approximate, our  assumption is that articles in different languages describing the same event will generally have a higher similarity than articles about different events.
 While this assumption does not always hold, redundancy in the data mitigates these false positives.
@@
 \item \texttt{hasSameLocation} feature is a boolean variable that is true when the location of the event in both clusters is the same.
-Location of events is estimated by considering the locations mentioned in the articles from a cluster and is provided by Event Registry.
+The location of events is estimated by considering the locations mentioned in the articles that form a cluster and is provided by Event Registry.
 \item \texttt{timeDiff} is the absolute difference in hours between the two events.
@@
 We represent the documents as normalized TFIDF\cite{Salton88term-weightingapproaches} weighted vectors.
- \todo{tfidf over what corpus  - is this static?}
+ \todo{tfidf over what corpus  - is this static? Andrej: It is always static in hub based approach, i think based on en - es alignment in this case!}
  %These are languages with Wikipedia article sizes comparable to the minority languages.
@@
 Furthermore, we remove the documents that contain less than 20 different words and remove words that are too %infrequent as well as the top 100 most frequent words in the vocabularies.
- Although the English language is well aligned,\todo{well-aligned with what} number of articles highly correlates with the
+ Although the English language is well aligned with other languages,\todo{well-aligned with what} number of articles highly correlates with the
     number of articles connected to English, with all Wikipedia languages, we must note that quality of alignment varies quite a bit, especially in the case of small Wikipedias.
@@
 For each of the three algorithms, we analyzed all articles in Event Registry and for each article computed the most similar articles in other languages.
-To test how informative the identified similar articles are for cluster linking we then trained three classifiers as described in Section~\ref{trainClassifier} -- one for each algorithm.
+To test how informative the identified similar articles are for cluster linking we then trained three classifiers as described in Section~\ref{trainClassifier}\todo{This section does not exist!} -- one for each algorithm.
 Each classifier was allowed to use as learning features only the cross-lingual article linking features for which values are determined based on the selected algorithm ($k$-means, LSI and hub CCA).
@@
 As it can be seen, the results confirm our expectations: for smaller clusters it is indeed harder to correctly predict if the cluster pair should be merged or not.
-\todo{Andrej: A lahko poeditiras naslednje stavke - niso zelo jasno napisani.} Hub CCA outperforms other methods in this sense in terms of precision, which is consistent with its AMRR performance, see Figure~\ref{pic:AMRR}.
+\todo{Andrej: A lahko poeditiras naslednje stavke - niso zelo jasno napisani.
+Trenutno ne znam bolje :(} The hub CCA outperforms other methods in this sense in terms of precision, which is consistent with its AMRR performance, see Figure~\ref{pic:AMRR}.
 It is worth noting that added accuracy is also beneficial when we wish to choose good representatives of clusters. Therefore, we decided to use hub CCA in the cluster linking component.

### 2015-09-04 rupnik2015.tex

@@
 \centering
-\includegraphics[width=\textwidth]{pipeline}
+\includegraphics[width=\textwidth]{pipeline.pdf}
 \caption{\label{fig:erpipeline}  The Event Registry pipeline.
@@
 
-There are four main families of approaches to cross-lingual similarity.
-
-\textbf{Translation}.
-The most obvious way to compare documents written in different languages is to use machine translation and perform monolingual similarity.
+There are five main families of approaches to cross-lingual similarity.
+
+\textbf{Translation and dictionary based}.
+The most obvious way to compare documents written in different languages is to use machine translation and perform monolingual similarity, see  \cite{multilingualBook}\cite{plagiarism} for several variations of translation based approaches.
 One can use free tools such as Moses \shortcite{moses} or translation services, such as Google Translate\footnote{https://translate.google.com/}.
@@
 The operation efficiency and cost requirements make translation-based approaches less suited for our system.
+Closely related are works Cross-Lingual Vector Space Model (CL-VSM) \cite{plagiarism} and the approach presented in \cite{pouliquen2008story} which both compare documents by using dictionaries, which in both cases are EuroVoc dictionaries \cite{eurovoc}.
+The generality of such approaches is limited by the quality of available linguistic resources, which may be scarce or non-existent for certain language pairs.
 
@@
 Another closely related method is Explicit Semantic Analysis (ESA) \cite{ESA}, which uses Wikipedia (as do we in the current work) to compare documents.
+It can be interpreted as using the sample covariance matrix between features of two languages to define the dot product that is used to compute similarities.
 The authors of ESA compare it to CL-LSI and find that CL-LSI can outperform ESA in an information retrieval, but is costlier to optimize over a large corpus (ESA requires no training).
-We find that the scalability argument does not apply in our case: based on advances in numerical linear algebra we can solve large CL-LSI problems (millions of documents as oposed to 10,000 document limit reported in \cite{ESA}).
+We find that the scalability argument does not apply in our case: based on advances in numerical linear algebra we can solve large CL-LSI problems (millions of documents as opposed to 10,000 document limit reported in \cite{ESA}).
+In addition, ESA is less suited for computing similarities between two large monolingual streams.
+For example, each day we have to compute similarities between 500,000 English and 500,000 German news articles.
+Comparing each German news article with 500,000 English news articles is either prohibitively slow (involves projecting all English articles on Wikipedia) or consumes too much memory (involves storing the projected English articles, which for a Wikipedia of size 1,000,000 is a 500,000 by 1000,0000 non-sparse matrix).
+
+\textbf{Monolingual}.
+Finally, related work includes monolingual approaches that treat document written in different languages in a monolingual fashion.
+The intuition is that named entities (for example, ``Obama'') and cognate words (for example, ``tsunami'') are written in the same or similar fashion in many languages.
+For example, the Cross-Language Character n-Gram Model (CL-CNG) \cite{plagiarism} represents documents as bags of character n-grams.
+Another approach is to use language dependent keyword lists based on cognate words \cite{pouliquen2008story}.
+These approaches may be suitable for comparing documents written in languages that share a writing system, which does not apply to the case of global news tracking.
 

### 2015-09-04 rupnik2015.tex
related work

--- 2015-09-04 rupnik2015.tex
@@
 
-There are five main families of approaches to cross-lingual similarity.
+There are four main families of approaches to cross-lingual similarity.
 
@@
 Furthermore, representing a new document as a mixture of topics is another potentially hard inference problem which must be solved.
-
-\textbf{Classification problem-related}.
-In approaches such as \cite{cotrain} and \shortcite{svm2k} the authors find linear embeddings of the data that support good classification results, based on a given classification problem training data.
-We do not consider this family of approaches, since for general language pairs classifier training sets may not be available.\todo{add more here}
 
@@
 
-\todo{add the related work which was missing}
-
 Based on our requirements in Section~\ref{sec:sysreq}, we chose to focus on methods based on vector space models and linear embeddings.
@@
 All information extraction is then done on the translated article.
-\todo{add the missing related work}
 

### 2015-09-04 rupnik2015.tex
updates

@@
 We provide an extensive evaluation of the system as a whole, as well as an evaluation of the quality and robustness of the similarity measure and the linking algorithm.
-
-%Given a multi-lingual stream and clusters of articles from each language, we first propose a method for cross-lingual document similarity based on Wikipedia, which enables us to compute the similarity of any two articles regardless of language.
-The proposed method can scale to 100 languages and can match articles from languages with little or no direct overlap in the training data.
-Using this similarity, we then propose an approach to link clusters of articles across languages which represent the same event.
-We provide an extensive evaluation of the system as a whole, as well as an evaluation of the quality and robustness of the similarity measure and the linking algorithm.
-\begin{comment}
-Main points:
-\begin{itemize}
-\item Problem of tracking events in a large multilingual stream - we reduce it to event linking
-\item Two contributions. We propose a method that enables cross-lingaul document similarity an approach to linking multilingual data and an approach that uses is to link events.
-\item Evaluation of cross-lingual approaches on Wikipedia and the event linking task.
-\end{itemize}
-\end{comment}
 \end{abstract}
-%\section{Overview}
-%Structure
-%\begin{itemize}
-%\item Introduction
-%\item Cross lingual document similarity (model)
-%\item  Event Detection
-%\item  Combing the approaches
-%\item Experiments
-%\item Conclusions
-%\end{itemize}
-%The story:
-%\begin{itemize}
-%\item Description of the problem - we would liek to understand events ans sequences of events, but our observables (e.g. news stories) occur in a variety of languages
-%\item We present an approach to tackle this problem using a combination of techniques....
-%\item Ultimately we present a tool showing the feasibility of our approach
-%%\end{itemize}
-%
-%\section{Overview}
-%Structure
-%\begin{itemize}
-%\item Introduction
-%\item System overview
-%\item Cross lingual document similarity (algorithms)
-%\item Event linking
-%\item Experiments
-%\item Conclusions
-%\end{itemize}
-%The story:
-%\begin{itemize}
-%\item Description of the problem - we would like to understand global news as sequences of events, but our observables (e.g. news stories) occur in a variety of languages
-%\item We present an approach to tackle this problem using a combination of techniques....
-%\item Ultimately we present a tool showing the feasibility of our approach
-%\end{itemize}
-%
 
@@
 This greatly increases the scalability of the system.
-
-%As input we consider a stream of articles in different languages and a list of events.
-Our goal is to assign articles to their corresponding events.
-A priori we do not know the coverage of the articles, that is not all the events may be covered and we do not know that all the articles necessarily fit into one of the events.
-To achieve our goal, we divide the task into two parts: clustering the articles according to topic (regardless of language) and matching the clusters to events.
-%
-%To cluster the articles, rather than rely on machine translations, we adopt well-established monolingual text mining tools to the multilingual setting.
-This represents the continuation of previous work \shortcite{nips,sikdd,nips2,iti} where we explored representations of documents which were valid over multiple languages.
-The representations could be interpreted as multi-lingual topics, which were then used as proxies to compute cross-lingual similarities between documents.
-To learn the representations, we use Wikipedia as a training corpus.
-Significantly, we do not only consider the major or \emph{hub} languages such as English, German, French, etc.
-which have significant overlap in article coverage, but also smaller languages (in terms of number of Wikipedia articles) such as Slovenian and Hindi, which may have a negligible overlap in article coverage.
-We can then define a similarity between any two articles regardless of language,  which allows us to cluster the articles according to topic.
-The underlying assumption being that articles describing the same event are similar  and will therefore be put into the same cluster.
-%
-%The second step is to match our computed clusters to events. Since our similarity measure and clustering
-%may not be perfect, we need to find the ``best" correspondence between clusters and events.
-In this paper, we test several different algorithms for both steps for performance in an integrated system.
-In particular, we test our approach within an online system for detection of world events, called Event Registry \shortcite{Leban2014W,Leban2014I}.
-Event Registry is a repository of events, where events are automatically identified by analyzing news articles.
-
-%
-%Event Registry works by collecting and analyzing news articles that are published by news outlets in different languages all over the world.
-Collected articles are first semantically annotated by identifying in them mentions of relevant concepts -- either entities or important keywords.
-The disambiguation and entity linking of the concepts is done using Wikipedia as the main knowledge base.
-%   The key step in Event Registry is then to apply an online clustering algorithm on the articles in order to identify groups of articles that are discussing the same event.
-%
-%   Within this system, we test our multilingual similarity measures as well as our algorithms for matching clusters across languages.
-The result is that we can test total performance in a multilingual, real-world setting.
 
@@
 Both step rely heavily on simliarity computation, which implies that this must be highly scalable.
-%
-%while both steps must be scalable, the underlying similarity computation is used
-%- implying that similarity computation should be highly scalable.  %The system should support as many languages as possible.
+
 Therefore, we focus on implementations that run on a single shared memory machine, as opposed to clusters of machines. This  simplifies implementation and system maintenance.
@@
 \todo{add some more details/intuition.}
-\begin{comment}
-Technical detail.
-Map everything in the english space.
-$S_{1, 2} = Q^{en}_1^T Q^{en}_2,$ where  Q^{en}_1 and Q^{en}_2 are orthogonal matrices.
-All non zero singular values are thus automatically equal to one.
-\end{comment}
 
@@
  to reformulate a difficult optimization problem as an eigenvector problem. The second ingredient is dimensionality reduction using LSI, which
- makes the eigenvector problem computationally and numerically tractable. \todo{we need to set this up better. Revised by Jan. Check if OK. Andrej checked}
+ makes the eigenvector problem computationally and numerically tractable.
 
@@
 Therefore, each document  is represented by a long vector indexed by the terms in all languages.
-\todo{picture here.
-Updated: added Figure 3}
 
@@
 Note that the method in its original form is only applicable to two languages where an aligned set of observations is available.
-\todo{each of these sections should have a matrix picture..}
 
@@
 
-\subsection{Algorithm}
+\subsection{Algorithm}\label{algo:features}
 
@@
 We represent the documents as normalized TFIDF\cite{Salton88term-weightingapproaches} weighted vectors.
- \todo{tfidf over what corpus  - is this static? Andrej: It is always static in hub based approach, i think based on en - es alignment in this case!}
- %These are languages with Wikipedia article sizes comparable to the minority languages.
-(since the number of speakers does not directly correlate with the number of Wikipedia articles).
-The prime hub candidate is the English %language which is well aligned with all other Wikipedia languages, although the alignment quality varies quite a bit.
-Furthermore, we remove the documents that contain less than 20 different words and remove words that are too %infrequent as well as the top 100 most frequent words in the vocabularies.
- Although the English language is well aligned with other languages,\todo{well-aligned with what} number of articles highly correlates with the
-    number of articles connected to English, with all Wikipedia languages, we must note that quality of alignment varies quite a bit, especially in the case of small Wikipedias.
-% \todo{Move this to footnote for stubs}
+The IDF scores are computed for each language with the exception of English based on its aligned documents with the English Wikipedia.
+The English language IDF scores are based on all English documents for which aligned Spanish documents exist.
 
@@
 }
-%
-%We now describe the selection of train and test sets.
-We select the test set documents as all multi-lingual documents with at least one nonempty alignment from the list: (\emph{hi}, \emph{ht}), (\emph{hi}, \emph{pms}), (\emph{war}, \emph{ht}), (\emph{war}, \emph{pms}).
-This guarantees that we cover all the languages.
-Moreover this test set is suitable for testing the retrieval thorough the hub as the chosen pairs have empty alignments.
-The remaining documents are used for training.
-In Table \ref{table:train_test}, we display the corresponding sizes of training and test documents for each language pair.
-The first row represents the size of the training sets used to construct the mappings in low dimensional language independent space using the English--\emph{en} as a hub.
-The diagonal elements represent number of the unique training documents and test documents in each language.
 
@@
 To evaluate the prediction accuracy for a given dataset we used 10-fold cross validation.
-%In this technique, the dataset is partitioned into 10 folds, where 9 folds are used for training a model and the left-out fold is then used to test the accuracy of the model.
-The learning %and testing process is repeated 10 times, each time with a different left-out fold.
 
@@
 
-% which method?
 In Section~\ref{sec:models}, we  described three main algorithms for identifying similar articles in different languages.
@@
 For each of the three algorithms, we analyzed all articles in Event Registry and for each article computed the most similar articles in other languages.
-To test how informative the identified similar articles are for cluster linking we then trained three classifiers as described in Section~\ref{trainClassifier}\todo{This section does not exist!} -- one for each algorithm.
+To test how informative the identified similar articles are for cluster linking we then trained three classifiers as described in Section~\ref{algo:features} -- one for each algorithm.
 Each classifier was allowed to use as learning features only the cross-lingual article linking features for which values are determined based on the selected algorithm ($k$-means, LSI and hub CCA).
 The results of the trained models are shown in Table~\ref{table:linkingEvalAlgos}.
-%The table shows for each of the algorithms the obtained classification accuracy, precision and recall.
+
 We can see that the most informative features for cluster linking are obtained using the hub CCA algorithm, while $k$-means performs the worst.
@@
 
-%{\oldText{TODO:} add some explanation why this might be so.}
-%\emph{Andrej adds some tests, new experiment description graphs, and explanations for performance, done by Monday}
-%15000 test documents, around 300 000 sth training documents
-%Results for feature spaces of dimensions 500, 800, 1000  overfitting, 800 minor improvement
-%Timex complexity of similarity computation scales linearly with dimension space. We double numbers of articles and decrease memory requirements by 1/2.
-%CCA       CA          AUC         Prec        Recall
-%CCA =     [0.7882      0.8504      0.7692	   0.8209
-%           0.7957      0.8624      0.7788      0.8234
-%           0.8031      0.8654      0.8015      0.8035];
-%LSIS=     [0.7994      0.8627      0.7778      0.8358
-%           0.7956      0.8634      0.7828      0.8159
-%           0.7994      0.8642      0.8030      0.7910];
-%LSIW=     [0.7894      0.8606      0.7673      0.8284;
-%           0.7870      0.8632      0.7662      0.8234;
-%           0.8018      0.8663      0.7854      0.8284];
-
 %AMPR as function of the number of feature vectors
@@
 \end{figure}
-%Why the hub approach?
-%Optimize for language pairs in the system that have the largest number of articles in the system!
-%Is star CCA really the best???
-%Maybe omit this table?
+
 \begin{table}[h]
@@
 When training the classifier all three groups of features were available.
-The results are shown in Table~\ref{langPairEval}.
+The results are shown in Table~\ref{table:langPairEval}.
 We can see that the highest accuracy can be obtained when we need to decide if an English and German cluster represent the same event.
@@
 The prediction accuracies for individual language pairs are comparable although it seems that the achievable accuracy correlates with the amount of available language resources.
-
-%\begin{table}[h]
-%\caption{Accuracy of cluster linking with features obtained from different cross-lingual similarity algorithms}
-%\label{table:linkingEvalAlgos}
-%\begin{center}
-%\begin{tabular}{|c|c|c|c|}
-%  \hline
-%  \cline{1-4}
-%  Models & CA & Precision & Recall \\ \cline{1-4}
-%  CCA & 77.6\% & 78.3 \%  & 81.1\%  \\ \cline{1-4}
-%  $k$-means & 73.9\% & 69.5\%  & 84.6\%  \\ \cline{1-4}
-%  LSI & 79.2\% & 78.9\%  & 80.6\%  \\ \cline{1-4}
-%  \hline
-%\end{tabular}
-%\end{center}
-%\end{table}
 
@@
 \subsection{Scaling}
-%\emph{I think that this section should be short.}
-%100 languages!
-%Memory 100 GB per day
-%Newsfeed volume 430k articles per day, 210k english, 36k Spanish, 27k German, 17k French, ....
-%Number of similarities, 430k similarities every 200ms, currently could handle 1M articles,
-%(n \over 2) 500X500 proxy matrices, 300000x500 Projector matrices, 430000x500 data
-%Premultiply with proxy matrices, compare blockwise!
+
 One of the main advantages of our approach is that it is highly scalable.
@@
 The included archive contains two folders: ``positive'' and ``negative'', where the first folder includes examples of cluster pairs in two languages that represent the same event and the second folder contains pairs of clusters in two languages that do not represent different events.
-Each example is a JSON file that contains at the top level information about a pair of clusters (including text of the articles) as well as a set of ``meta'' attributes, that correspond to features described in \ref{trainClassifier}.
+Each example is a JSON file that contains at the top level information about a pair of clusters (including text of the articles) as well as a set of ``meta'' attributes, that correspond to features described in Section \ref{algo:features}.
 

### 2015-09-04 rupnik2015.tex
figure updated

@@
 \caption{\label{fig:lsi} LSI multilingual corpus matrix decomposition.}
-\todo{Missing matrix S of singular values!!!}
 \end{figure}
@@
 \textbf{Hub language characteristic}
-In the case of Wikipedia we observed that even though the training resources are scarce for certain language pairs, there often exists indirect training data that we can use as a proxy -- hub language.\todo{Check this, I prettified it.}
+In the case of Wikipedia, we observed that even though the training resources are scarce for certain language pairs, there often exists indirect training data.
+By considering a third language, which has training data with both languages in the pair,  we can use the composition of learned maps as a proxy.
+We refer to this third language as a hub language.
 

### 2015-09-04 rupnik2015.tex
updates

@@
 
-The paper is organized as follows: we first provide an overview of the system as a whole (Section ~\ref{sec:pipeline}), followed by details on each of the two steps described above.
+The paper is organized as follows: we first provide an overview of the system as a whole (Section ~\ref{sec:pipeline}), followed by details on the cross-lingual similarity computation and our approach to linking clusters.
 We first introduce the problem of cross-lingual similarity computation (Section~\ref{sec:crosslingual}), which is followed by several approaches for measuring cross-lingual similarity (Section~\ref{sec:models}) and then the algorithms for matching clusters in different languages (Section~\ref{sec:linking}).
-To improve readability, we describe relevant related work for each module in its corresponding section.
+To improve readability, we will describe relevant related work for each module in its corresponding section.
 Finally, we present and interpret the experimental results, give conclusions and describe promising future directions.
@@
 Event Registry is a repository of events, where events are automatically identified by analyzing news articles that are collected from numerous news outlets all over the world.
-The important components in the pipeline of the Event Registry are shown in Figure~\ref{fig:erpipeline} and are briefly described below.
+The important components in the pipeline of the Event Registry are shown in Figure~\ref{fig:erpipeline}.
+We will now briefly describe the main components.
 
@@
 \subsection{Cross-Lingual Models}\label{sec:models}
-In this section, we describe several approaches to the problem of computing multilingual simliarities introduced in \ref{sec:tfidf}. We present four approaches:
+In this section, we will describe several approaches to the problem of computing multilingual simliarities introduced in \ref{sec:tfidf}. We present four approaches:
 a simple approach based on $k$-means clustering in \ref{sec:kmeans}, a standard approach based on singular value decomposition in \ref{sec:LSI}, a related
@@
 
-The resulting matrix (when appropriately scaled to have unit norm) for similarity computation between language $i$ and language $j$ is defined as (see the reasoning below):
+The resulting matrix (when appropriately scaled to have unit norm) for similarity computation between language $i$ and language $j$ is defined as (see the reasoning in the next paragraph):
 $$C_i(C_i^T C_i)^{-1} (C_j^T C_j)^{-1} C_j.$$
@@
 We proceed by reducing the dimensionality of each $X_i$ by setting: $Y_i = V_i^T \cdot X_i$, where $Y_i \in \RR^{k\times N}$.
-To summarize, the first step reduces the dimensionality of the data and is based on CL-LSI, but optimizes only the hub-language related cross-covariance blocks.
+To summarize, the first step reduces the dimensionality of the data and is based on CL-LSI, but optimizes only the hub language related cross-covariance blocks.
 
@@
 
-We describe the main dataset for building cross-lingual models which is based on Wikipedia and then present two sets of experiments. The first set of experiments
+We will describe the main dataset for building cross-lingual models which is based on Wikipedia and then present two sets of experiments. The first set of experiments
 establishes that the hub based approach can deal with language pairs where little or no training data is available. The second set of experiments compares the main approaches
@@
  articles, $52$ with more than 100k %$10^5$
- articles, $129$ with more than 10k articles, and $236$ with more than $1000$ articles.
+ articles, $129$ with more than 10k articles, and $236$ with more than $1,000$ articles.
 
 Each Wikipedia page is embedded in the page tag.
-First, we check if the title of the page consists of any special namespace and do not process such pages.
+First, we check if the title of the page consists of any special namespace (which includes categories and discussion pages) and do not process such pages.
 Then, we check if this is a redirection page and we store the redirect link because inter-language links can point to redirection links also.
@@
 We observe that the method performs well on all pairs of languages, where at least 50,000 training documents are available(\emph{en}, \emph{es}, \emph{ru}, \emph{sl}).
-We note that taking $k = 500$ or $k = 1000$ multilingual topics usually results in similar performance, with some notable exceptions: in the case of (\emph{ht}, \emph{war}) the additional topics result in an increase in performance, as opposed to (\emph{ht},\emph{pms}) where performance drops, which suggests overfitting.
+We note that taking $k = 500$ or $k = 1,000$ multilingual topics usually results in similar performance, with some notable exceptions: in the case of (\emph{ht}, \emph{war}) the additional topics result in an increase in performance, as opposed to (\emph{ht},\emph{pms}) where performance drops, which suggests overfitting.
 The languages where the method performs poorly are \emph{ht} and \emph{war}, which can be explained by the quality of data (see Table \ref{table:rank} and explanation that follows).
-In case of \emph{pms}, we demonstrate that solid performance can be achieved for language pairs (\emph{pms}, \emph{sl}) and (\emph{pms}, \emph{hi}), where only 2000 training documents are shared between \emph{pms} and \emph{sl} and no training documents are available between \emph{pms} and \emph{hi}.
+In case of \emph{pms}, we demonstrate that solid performance can be achieved for language pairs (\emph{pms}, \emph{sl}) and (\emph{pms}, \emph{hi}), where only 2,000 training documents are shared between \emph{pms} and \emph{sl} and no training documents are available between \emph{pms} and \emph{hi}.
 Also observe that in the case of (\emph{pms}, \emph{ht}) the method still obtains a score of 62, even though training set intersection is zero and \emph{ht} data is corrupted, which we will show in the next paragraph.
@@
 \begin{table}[h!]
-\caption{Pairwise retrieval, 500 topics on the left -- 1000 topics on the right}\label{table:retrieval}
+\caption{Pairwise retrieval, 500 topics on the left -- 1,000 topics on the right}\label{table:retrieval}
 \begin{center}
@@
 It is clear that hub CCA outperforms LSI approach and K-means lags far behind when testing on Wikipedia data.
-The hub CCA approach with $500$ topic vectors manages to perform comparably to the $LSI$-based approach with $1000$ topic vectors, which shows that the $CCA$ method can improve both model memory footprint as well as similarity computation time.
+The hub CCA approach with $500$ topic vectors manages to perform comparably to the $LSI$-based approach with $1,000$ topic vectors, which shows that the $CCA$ method can improve both model memory footprint as well as similarity computation time.
 
@@
 Furthermore, we inspected how the number of topics influences the accuracy of cluster linking.
-As we can see from Table~\ref{table:linkingEvalAlgos} choosing number of topic vectors larger than $500$ barely affects linking performance, which is in contrast with the fact that additional topics helped to improve AMMR, see Figure~\ref{pic:AMRR}.
+As we can see from Table~\ref{table:linkingEvalAlgos} choosing a number of features larger than $500$ barely affects linking performance, which is in contrast with the fact that additional topics helped to improve AMMR, see Figure~\ref{pic:AMRR}.
 Such differences may have arisen due to different domains of training and testing (Wikipedia pages versus news articles).
@@
 %AMPR as function of the number of feature vectors
-%500 is good, 800 better, 1000 overfitting, noise!
 \begin{figure}
@@
 \begin{table}[h]
-\caption{Accuracy of cluster linking with 500 - 800 - 1000 topic vectors obtained from different cross-lingual similarity algorithms.
+\caption{Accuracy of cluster linking with 500 - 800 - 1,000 topic vectors obtained from different cross-lingual similarity algorithms.
 The table shows for each of the algorithms the obtained classification accuracy, precision and recall.}
@@
 It is fast, very robust to quality of training data, easily extendable, simple to implement and has relatively small hardware requirements.
-The similarity pipeline is the most computationally intensive part and currently runs on two Intel Xeon E5-2667 v2, 3.30GHz processor machine with 256GB of RAM.
-This is sufficient to do similarity computation over 150 languages if needed, see \url{http://xling.ijs.si} for live demo over 100 languages.
+The similarity pipeline is the most computationally intensive part and currently runs on a machine with two Intel Xeon E5-2667 v2, 3.30GHz processors with 256GB of RAM.
+This is sufficient to do similarity computation over a large number of languages if needed.
 It currently uses Wikipedia as a freely available knowledge base and experiments show that the similarity pipeline dramatically reduces the search space when linking clusters.

### 2015-09-05 rupnik2015.tex
updates

@@
 Our goal is to assign articles to their corresponding events.
-A priori, we do not know the coverage of the articles, that is not all the events may be covered and we do not know that all the articles necessarily fit into one of the events.
+A priori, we do not know the coverage of the articles, that is, not all the events may be covered and we do not know that all the articles necessarily fit into one of the events.
 The task is divided into two parts: detecting events within each language and then linking events across languages.
@@
 Within each monolingual stream, an online clustering approach is employed, where tracked clusters correspond to our definition of events - this is based on the Event Registry system \shortcite{Leban2014W,Leban2014I}.
-Our main goal in this paper is to connect such clusters (representations of events) across languages, that is detect that a set of articles in language $A$ reports on the same event as a set of articles in language $B$.
+Our main goal in this paper is to connect such clusters (representations of events) across languages, that is, to detect that a set of articles in language $A$ reports on the same event as a set of articles in language $B$.
 
@@
 Since articles in a cluster are in a single language, we also want to identify any other existing clusters that report about the same event in other languages and join these clusters into the same event.
-This task is performed using a classification approach that is the second major contribution of this paper.
+This task is performed using a classification approach which is the second major contribution of this paper.
 It is described in detail in Section~\ref{sec:linking}.
@@
 The pipeline must be able to process millions of articles per day and perform billions of similarity computations each day.
-Both step rely heavily on simliarity computation, which implies that this must be highly scalable.
+Both step rely heavily on similarity computation, which implies that this must be highly scalable.
 
@@
 The system should be able to take advantage of comparable corpora (as opposed to parallel translation-based corpora), with missing data.
-Supporting a new language should be straightforward.
 \item \textbf{Operation efficiency} - The similarity computation should be fast - the system must be able to handle billions of similarity computations per day.
@@
 In addition, OPCA forces the vocabulary sizes for all languages to be the same, which is less intuitive.
-For our setting, the method in \cite{nonnegfactor_lsi} has a prohibitively high computational cost when building models (it uses dense matrices of dimensions that is the product of training set size and vocabulary size).
+For our setting, the method in \cite{nonnegfactor_lsi} has a prohibitively high computational cost when building models (it uses dense matrices whose dimensions are a product of the training set size and the vocabulary size).
 Our proposed approach combines CCA and CL-LSI.
-Another closely related method is Explicit Semantic Analysis (ESA) \cite{ESA}, which uses Wikipedia (as do we in the current work) to compare documents.
-It can be interpreted as using the sample covariance matrix between features of two languages to define the dot product that is used to compute similarities.
-The authors of ESA compare it to CL-LSI and find that CL-LSI can outperform ESA in an information retrieval, but is costlier to optimize over a large corpus (ESA requires no training).
+Another closely related method is Cross-Lingual Explicit Semantic Analysis (CL-ESA) \cite{ESA}, which uses Wikipedia (as do we in the current work) to compare documents.
+It can be interpreted as using the sample covariance matrix between features of two languages to define the dot product which is used to compute similarities.
+The authors of CL-ESA compare it to CL-LSI and find that CL-LSI can outperform CL-ESA in an information retrieval, but is costlier to optimize over a large corpus (CL-ESA requires no training).
 We find that the scalability argument does not apply in our case: based on advances in numerical linear algebra we can solve large CL-LSI problems (millions of documents as opposed to 10,000 document limit reported in \cite{ESA}).
-In addition, ESA is less suited for computing similarities between two large monolingual streams.
+In addition, CL-ESA is less suited for computing similarities between two large monolingual streams.
 For example, each day we have to compute similarities between 500,000 English and 500,000 German news articles.
@@
 
-A system that is significantly different, but worth mentioning is the GDELT project~\cite{Leetaru2013Gdelt}.
+A system, which is significantly different but worth mentioning, is the GDELT project~\cite{Leetaru2013Gdelt}.
 In GDELT, events are also extracted from articles, but in their case, an event is specified in a form of a triple containing two actors and a relation.
@@
 We will first describe how documents are represented as vectors and how to compare douments in a mono-lingual setting.
-We then define a way to measure cross-lingual similarity that is natural for the models we consider.
+We then define a way to measure cross-lingual similarity which is natural for the models we consider.
 
@@
 The cross-lingual similarity models presented in this paper are based on comparable corpora.
-That is, a corpus of documents in multiple languages, with alignment between documents that are of the same topic, or even a rough translation of each other.
+A \emph{comparable corpus} is a collection of documents in multiple languages, with alignment between documents that are of the same topic, or even a rough translation of each other.
 Wikipedia is an example of a comparable corpus, where a specific entry can be described in multiple languages (e.g.
@@
 For each language $\ell$, we obtain a set of monolingual clusters $C_{\ell}$.
-That is, all articles from each cluster $c \in C_{\ell}$ are in language $\ell$.
+More precisely, the articles corresponding to each cluster $c \in C_{\ell}$ are written in the language $\ell$.
 Given a pair of languages $\ell_a \in L$, $\ell_b \in L$ and $\ell_a \not= \ell_b$, we would like to identify all cluster pairs $(c_i, c_j) \in C_{\ell_a} \times C_{\ell_b}$ such that $c_i$ and $c_j$ describe the same event.
@@
 \centering
-\caption{Pairwise training -- test sizes (in thousands)
+\caption{Training -- test sizes (in thousands).
 The first row represents the size of the training sets used to construct the mappings in low-dimensional language independent space using the English--\emph{en} as a hub.
@@
 
-We further inspect the properties of the training sets by roughly estimating the fraction \texttt{rank(A)/min(size(A))} for each English training matrix and its corresponding mate matrix.
-Ideally, these two fractions are approximately the same so  both aligned spaces should have reasonably similar dimensionality.
+We further inspect the properties of the training sets by roughly estimating the fraction \texttt{rank(A)/min(rows(A), cols(A))} for each English training matrix and its corresponding mate matrix, where $rows(A)$ and $cols(A)$ denote the number of rows and columns respectively.
+The denominator represents the theoretically highest possible rank the matrix $A$ could have.
+Ideally, these two fractions should be approximately the same - both aligned spaces should have reasonably similar dimensionality.
 We display these numbers as pairs in Table \ref{table:rank}.
@@
 \begin{table}[h]
-\caption{Dimensionality drift}
+\caption{Dimensionality drift.
+Each column corresponds to a pair of aligned corpus matrices between English and another language.
+The numbers represent the ratio between the numerical rank and the highest possible rank.
+For example, the column $en -- ht$ tells us that for the English-Creole pairwise-aligned corpus matrix pair, the English counterpart has full rank, but the Creole counterpart is far having full rank.}
 \label{table:rank}
@@
 \cline{1-7}
-(en, de)     &   (en, ru)     &   (en, sl)       &     (en, hi)&   (en, war)      &      (en, ht) &   (en, pms)\\
+en -- de     &   en -- ru     &   en -- sl       &     en -- hi &   en -- war      &      en -- ht &   en -- pms\\
 \cline{1-7}
-(0.81, 0.89)   &  (0.8, 0.89)  &   (0.98, 0.96)    &    (1, 1)  &   (0.74, 0.56)  &      (1, 0.22)  &   (0.89, 0.38)\\
+0.81 -- 0.89   &  0.8 -- 0.89  &   0.98 -- 0.96    &    1 -- 1  &  0.74 -- 0.56  &      1 -- 0.22  &   0.89 -- 0.38\\
 \cline{1-7}
@@
 To test how informative the identified similar articles are for cluster linking we then trained three classifiers as described in Section~\ref{algo:features} -- one for each algorithm.
-Each classifier was allowed to use as learning features only the cross-lingual article linking features for which values are determined based on the selected algorithm ($k$-means, LSI and hub CCA).
+Each classifier was allowed to use as learning features \textbf{only} the cross-lingual article linking features for which values are determined based on the selected algorithm ($k$-means, LSI and hub CCA).
 The results of the trained models are shown in Table~\ref{table:linkingEvalAlgos}.
+We also show how the number of topics (the dimensions of the latent space) influences the quality, except in the case of the $k$-means algorithm, where only the performance on 500 topic vectors is reported, due to higher computational cost.
 
@@
 The Figure~\ref{pic:AMRR} shows AMRR score as the function of the number of feature vectors.
-It is clear that hub CCA outperforms LSI approach and K-means lags far behind when testing on Wikipedia data.
+It is clear that hub CCA outperforms LSI approach and $k$-means lags far behind when testing on Wikipedia data.
 The hub CCA approach with $500$ topic vectors manages to perform comparably to the $LSI$-based approach with $1,000$ topic vectors, which shows that the $CCA$ method can improve both model memory footprint as well as similarity computation time.
@@
   \cline{1-4}
-  Models & A & Precision & Recall \\ \cline{1-4}
+  Models & Accuracy & Precision & Recall \\ \cline{1-4}
   CCA      & 78.2 - 79.6 - 80.3 \% & 76.3 - 78.0 - 80.5 \%  & 81.6 - 82.1 - 79.9 \%  \\ \cline{1-4}

### 2015-09-05 rupnik2015.tex
updates

@@
 $$sim(d_1, d_2) = \frac{\langle \phi(d_1), \phi(d_2)\rangle}{\|\phi(d_1)\| \|\phi(d_2)\|},$$
-where $\langle \cdot,\cdot \rangle$ and $\|\cdot\|$ are standard inner product and Euclidean norm.
-Cosine similarity and other related approaches assume that the similarity is reflected in the overlap of words.
-While always defined, this metric is only meaningful when the documents $d_1$ and $d_2$ are defined on the same space (e.g.
-have a comparable vocabulary or are written in the same language, since otherwise their supports are disjoint).
+where $\langle \cdot,\cdot \rangle$ and $\|\cdot\|$ are standard inner product and Euclidean norm. When dealing with two or more languages, one could ignore the language information
+and build a vector space using the union of tokens over the languages.
+A cosine similarity function in such a space can be useful to some extent, for example "internet" or "Obama" may appear both in a Spanish and English texts and the presence of such terms in both an English and a Spanish document would contribute to their similarity.
+In general however, large parts of vocabularies may not intersect, which means that given a language pair, many words in both languages cannot contribute to the similarity score, which makes the similarity function very insensitive.
 
@@
 If the maximal singular value of $S_{i,j}$ is bounded by $1$, then the similarity scores will lie on the interval $[-1, 1]$.
-In Section \ref{sec:models} we will describe some approaches to computing $S_{i,j}$ given training data.
-\todo{add some more details/intuition.}
+We will provide an overview of the models in Section \ref{sec:models} and then introduce additional notation in \ref{sec:notation}.
+Starting with Section \ref{sec:kmeans} and ending with Section \ref{sec:hublang} we will describe some approaches to computing $S_{i,j}$ given training data.
 
@@
 
-\subsection{Notation}
+\subsection{Notation}\label{sec:notation}
 
@@
 More formally, \emph{multilingual document} $d = (u_1,\ldots u_m)$ is a tuple of $m$ documents on the same topic (comparable), where $u_i$ is the document written in language $i$.
-Note that individual document $u_i$ can be an empty document (missing resource) and each $d$ must contain at least two nonempty documents.
+Note that individual document $u_i$ can be an empty document (missing resource) and each $d$ must contain at least \textbf{two nonempty documents}.
+This means that in our analyses we discard strictly monolingual documents for which no cross-lingual information is available.
 A comparable corpus $D = {d_1, \ldots, d_s}$ is a collection of $s$ multilingual documents.
@@
 A \emph{hub language} is a language with a high proportion of non-empty documents in $D = \left\{d_1,..., d_{\ell}\right\}$.
+As we have mentioned, we only focus on multilingual documents that include at least two languages.
 The prototypical example in the case of Wikipedia is English.
+Our notion of the hub language could be interpreted in the following way.
+If a non-English Wikipedia page contains one or more links to variants of the page in other languages, English is very likely to be one of them. That makes English a hub language.
+
 We use the following notation to define subsets of the multilingual comparable corpus: let $a(i,j)$ denote the index set of all multilingual documents with non-missing data for the $i$-th and $j$-th language:  $$a(i,j) = \left\{k~ |~ d_k = (u_1,...,u_m), u_i \neq \emptyset, u_j \neq \emptyset \right\},$$ and let $a(i)$ denote the index set of all multilingual documents with non missing data for the $i$-th language.
@@
 We use three groups of features to describe cluster pair $(c_i, c_j)$.
-The first group is based on {\bf cross-lingual article links}, which are derived using cross-lingual similarity: each news article $a_i$ is linked with its $10$-nearest neighbors articles from all other languages.
+The first group is based on {\bf cross-lingual article links}, which are derived using cross-lingual similarity: each news article $a_i$ is linked with its $10$-nearest neighbors articles from all other languages (10 per each language).
 The group contains the following features:
@@
 
-\subsection{Scaling}
+\subsection{Remarks on the scalability of the implementation}
 
@@
 
-\subsection{Repeatability of Experiments}
+\subsection{Remarks on the repeatability of Experiments}
 We have made publicly available at \url{https://github.com/rupnikj/jair_paper.git} both the code and data that were used in the experiments.

### 2015-09-05 rupnik2015.tex
updates

@@
 In this work, we address the problem of tracking of events in a large multilingual stream.
-Within a recently developed system (EventRegistry  \shortcite{Leban2014W,Leban2014I}), we examine two aspects of this problem: how to compare articles in different languages and how to link collections of articles in different languages which refer to the same event.
+Within a recently developed system EventRegistry  \shortcite{Leban2014W,Leban2014I} we examine two aspects of this problem: how to compare articles in different languages and how to link collections of articles in different languages which refer to the same event.
 Taking a multilingual stream and clusters of articles from each language, we compare different cross-lingual document similarity measures based on Wikipedia.
@@
 As it can be seen, the results confirm our expectations: for smaller clusters it is indeed harder to correctly predict if the cluster pair should be merged or not.
-\todo{Andrej: A lahko poeditiras naslednje stavke - niso zelo jasno napisani.
-Trenutno ne znam bolje :(} The hub CCA outperforms other methods in this sense in terms of precision, which is consistent with its AMRR performance, see Figure~\ref{pic:AMRR}.
+
+ The hub CCA outperforms other methods in this sense in terms of precision, which is consistent with its AMRR performance, see Figure~\ref{pic:AMRR}.
 It is worth noting that added accuracy is also beneficial when we wish to choose good representatives of clusters. Therefore, we decided to use hub CCA in the cluster linking component.
@@
   Models & Accuracy & Precision & Recall \\ \cline{1-4}
-  CCA      & 78.2 - 79.6 - 80.3 \% & 76.3 - 78.0 - 80.5 \%  & 81.6 - 82.1 - 79.9 \%  \\ \cline{1-4}
+  hub CCA      & 78.2 - 79.6 - 80.3 \% & 76.3 - 78.0 - 80.5 \%  & 81.6 - 82.1 - 79.9 \%  \\ \cline{1-4}
   LSI      & 78.9 - 78.7 - 80.6 \%  & 76.8 - 77.0 - 78.7 \%  & 83.3 - 80.6 - 83.6 \%  \\ \cline{1-4}
@@
   \cline{1-4}
-  Models & A & Precision & Recall \\ \cline{1-4}
-  CCA      & 81.2 - 77.8 \% & 80.5 - 74.5 \%  & 91.3 - 57.5 \%  \\ \cline{1-4}
+  Models & Accuracy & Precision & Recall \\ \cline{1-4}
+  hub CCA      & 81.2 - 77.8 \% & 80.5 - 74.5 \%  & 91.3 - 57.5 \%  \\ \cline{1-4}
   LSI      & 82.8 - 76.4 \% & 81.3 - 70.9 \%  & 93.1 - 57.5 \%  \\ \cline{1-4}
@@
   Features & A & Precision & Recall & F1 \\ \cline{1-5}
-  CCA & $78.3\% \pm 5.9\%$ & $0.782 \pm 0.070$ & $0.789 \pm 0.052$ & $0.784 \pm 0.055$ \\ \cline{1-5}
+  hub CCA & $78.3\% \pm 5.9\%$ & $0.782 \pm 0.070$ & $0.789 \pm 0.052$ & $0.784 \pm 0.055$ \\ \cline{1-5}
   Concepts & $88.5\% \pm 2.7\%$ & $0.886 \pm 0.048$ & $0.886 \pm 0.022$ & $0.885 \pm 0.024$ \\ \cline{1-5}
   Misc & $54.8\% \pm 6.7\%$ & $0.618 \pm 0.165$ & $0.582 \pm 0.302$ & $0.524 \pm 0.130$ \\ \cline{1-5}
-  CCA + Concepts & $89.4\% \pm 2.5\%$ & $0.894 \pm 0.046$ & $0.896 \pm 0.024$ & $0.894 \pm 0.023$ \\ \cline{1-5}
-  CCA + Misc & $78.8\% \pm 5.0\%$ & $0.789 \pm 0.071$ & $0.794 \pm 0.046$ & $0.790 \pm 0.045$ \\ \cline{1-5}
+  hub CCA + Concepts & $89.4\% \pm 2.5\%$ & $0.894 \pm 0.046$ & $0.896 \pm 0.024$ & $0.894 \pm 0.023$ \\ \cline{1-5}
+  hub CCA + Misc & $78.8\% \pm 5.0\%$ & $0.789 \pm 0.071$ & $0.794 \pm 0.046$ & $0.790 \pm 0.045$ \\ \cline{1-5}
   Concepts + Misc & $88.7\% \pm 2.6\%$ & $0.888 \pm 0.046$ & $0.888 \pm 0.022$ & $0.887 \pm 0.023$ \\ \cline{1-5}

### 2015-09-05 rupnik2015.tex
paragraph added to intro

@@
 Events such as these are covered by many articles and the question is how to find all the articles in different languages that are describing a single event.
+Generally, events are more specific than general themes as the time component plays an important role -- for example, the two wars in Iraq would be considered as separate events.
 
@@
 
+The goal of this work is to examine how well events can be tracked in a multilingual reporting environment.
+Since this, represents complicated pipeline, we concentrate on two specific elements, namely multilingual similarity of articles and event linking using this similarity.
+Overall, these should be considered from a systems perspective (considering the system as a whole) rather than considering these problems in isolation.
+
 As input, we consider a stream of articles in different languages and a list of events.
@@
 
-As input we consider a high volume of articles in different languages.
+ We consider a high volume of articles in different languages.
 By using a language detector the stream is split into separate monolingual streams.

### 2015-09-05 rupnik2015.tex
intro

--- 2015-09-05 rupnik2015.tex
@@
 
-As input we consider a high volume of articles in different languages.
-By using a language detector the stream is split into separate monolingual streams.
+ We consider a high volume of articles in different languages.
+By using a language detector, the stream is split into separate monolingual streams.
 Within each monolingual stream, an online clustering approach is employed, where tracked clusters correspond to our definition of events - this is based on the Event Registry system \shortcite{Leban2014W,Leban2014I}.
@@
 Our approach to linking clusters across languages combines two ingredients: a cross-lingual document similarity measure, which can be interpreted as a language independent topic model, and semantic annotation of documents, which enables an alternative way to comparing documents.
+Since this work represents a complicated pipeline, we concentrate on these two specific elements.
+Overall, the approach should be considered from a systems' perspective (considering the system as a whole) rather than considering these problems in isolation.
 
@@
 
-Based on the similarity function we propose a novel algorithm for linking events/clusters across languages.
+Based on the similarity function, we propose a novel algorithm for linking events/clusters across languages.
 The approach is based on learning a classification model from labelled data based on several sets of features.

### 2015-09-05 rupnik2015.tex
paper overview revised

@@
 
-The paper is organized as follows: we first provide an overview of the system as a whole (Section ~\ref{sec:pipeline}), followed by details on the cross-lingual similarity computation and our approach to linking clusters.
-We first introduce the problem of cross-lingual similarity computation (Section~\ref{sec:crosslingual}), which is followed by several approaches for measuring cross-lingual similarity (Section~\ref{sec:models}) and then the algorithms for matching clusters in different languages (Section~\ref{sec:linking}).
-To improve readability, we will describe relevant related work for each module in its corresponding section.
-Finally, we present and interpret the experimental results, give conclusions and describe promising future directions.
+The paper is organized as follows: we first provide an overview of the system as a whole (Section~\ref{sec:pipeline}), which includes a subsection that summarizes the main system requirements in Section~\ref{sec:sysreq}.
+We then present related work in Section~\ref{sec:related}.
+The related work covers work on cross-lingual document similarity as well as work on cross-lingual cluster linking.
+In Section~\ref{sec:crosslingual} we introduce the main problem of cross-lingual document similarity computation, which is followed by an overview of the models that we considered in Section~\ref{sec:models}.
+We then establish necessary notation in Section~\ref{sec:notation} and proceed with four subsections that describe the main cross-lingual similarity approaches in detail.
+The approach in Section~\ref{sec:kmeans} serves as a baseline, and the approaches in Section~\ref{sec:CCA} and Section~\ref{sec:LSI} serve as the basis on which we proposed a new approach presented in Section~\ref{sec:hublang}.
+In Section~\ref{sec:linking} we introduce the central problem of cross-lingual linking of clusters of news articles and our approach that combines the cross-lingual similarity functions with knowledge extraction based techniques.
+Finally, we present and interpret the experimental results in Section~\ref{sec:evaluation} and discuss conclusions and point out several promising future directions.
 
@@
 
-\section{Evaluation}
+\section{Evaluation}\label{sec:evaluation}
 

### 2015-09-05 rupnik2015.tex
textbf changed to paragraph

--- 2015-09-05 rupnik2015.tex
@@
 
-In this section we will relate the methods we use to other alternatives in the literature.
-Since there are two distinctive tasks that we tackle in this paper (computing cross-lingual document similarity and cross-lingual cluster linking) we have separated the related work into two parts.
+In this section, we describe previous work described  in the literature.
+Since there are two distinctive tasks that we tackle in this paper (computing cross-lingual document similarity and cross-lingual cluster linking), we have separated the related work into two corresponding parts.
 
@@
 Comparison of documents (or other types of text snippets) in a monolingual setting is a well-studied problem in the field of information retrieval \cite{Salton88term-weightingapproaches}.
-We first formally introduce the problem followed by our approach.
+We first formally introduce the problem followed by a description of  our approach.
 
@@
 
-More formally, \emph{multilingual document} $d = (u_1,\ldots u_m)$ is a tuple of $m$ documents on the same topic (comparable), where $u_i$ is the document written in language $i$.
+More formally, a \emph{multilingual document} $d = (u_1,\ldots u_m)$ is a tuple of $m$ documents on the same topic (comparable), where $u_i$ is the document written in language $i$.
 Note that individual document $u_i$ can be an empty document (missing resource) and each $d$ must contain at least \textbf{two nonempty documents}.
@@
 A comparable corpus $D = {d_1, \ldots, d_s}$ is a collection of $s$ multilingual documents.
-By using the vector space model we can represent $D$ as a set of $m$ matrices $X_1,\ldots,X_m$, where $X_i \in \RR^{n_i \times s}$ is the matrix corresponding to the language $i$ and $n_i$ is the vocabulary size of language $i$.
+By using the vector space model, we can represent $D$ as a set of $m$ matrices $X_1,\ldots,X_m$, where $X_i \in \RR^{n_i \times s}$ is the matrix corresponding to the language $i$ and $n_i$ is the vocabulary size of language $i$.
 Furthermore, let $X_i^{\ell}$ denote the $\ell$-th column of matrix $X_i$ and the matrices respect the document alignment - the vector $X_i^\ell$ corresponds to the TFIDF vector of the $i$-th component of multilingual document $d_\ell$.
@@
 
-\textbf{A note on implementation}
+\paragraph{A note on implementation}
 
@@
 
-\textbf{Hub language characteristic}
+\paragraph{Hub language characteristic.}
 In the case of Wikipedia, we observed that even though the training resources are scarce for certain language pairs, there often exists indirect training data.
@@
 
-\textbf{Step 1: Hub language based dimensionality reduction}
+\paragraph{Step 1: Hub language based dimensionality reduction.}
 
@@
 
-\textbf{Step 2: Simplifying and solving SSCOR}
-
+\paragraph{Step 2: Simplifying and solving SSCOR.}
 The second step involves solving a generalized version of canonical correlation analysis on the matrices $Y_i$ in order to find the mappings $P_i$.
@@
 
-\textbf{Remark}
+\paragraph{Remark}
  The technique is related to  Generalization of Canonical Correlation Analysis (GCCA) by Carroll \citeyear{Carroll}, where an unknown group configuration variable is defined and the objective is to maximize the sum of squared correlation between the group variable and the others.

### 2015-09-05 rupnik2015.tex
small

@@
 \subsection{Remarks on the repeatability of Experiments}
-We have made publicly available at \url{https://github.com/rupnikj/jair_paper.git} both the code and data that were used in the experiments.
+We have made  both the code and data that were used in the experiments publicly available at \url{https://github.com/rupnikj/jair_paper.git}.
 The manually labelled dataset used in the evaluation of event linking is available at in the ``dataset`` subfolder of the github repository.
@@
 
-On the task of mate retrieval we observe that refining the LSI-based projections with hub CCA leads to improved retrieval precision, but the methods perform comparably on the task of event linking.
+On the task of mate retrieval, we observe that refining the LSI-based projections with hub CCA leads to improved retrieval precision, but the methods perform comparably on the task of event linking.
 Further inspection showed that the CCA-based approach reached a higher precision on smaller clusters.

### 2015-09-05 rupnik2015.tex
changes

@@
 
-\paragraph{Remark}
- The technique is related to  Generalization of Canonical Correlation Analysis (GCCA) by Carroll \citeyear{Carroll}, where an unknown group configuration variable is defined and the objective is to maximize the sum of squared correlation between the group variable and the others.
+\paragraph{Remark.} The technique is related to  Generalization of Canonical Correlation Analysis (GCCA) by Carroll \citeyear{Carroll}, where an unknown group configuration variable is defined and the objective is to maximize the sum of squared correlation between the group variable and the others.
 The problem can be reformulated as an eigenvalue problem.

### 2015-09-05 rupnik2015.tex
overview fixed

@@
 
-The paper is organized as follows: we first provide an overview of the system as a whole (Section~\ref{sec:pipeline}), which includes a subsection that summarizes the main system requirements in Section~\ref{sec:sysreq}.
+The paper is organized as follows: we first provide an overview of the system as a whole (Section~\ref{sec:pipeline}), which includes a subsection that summarizes the main system requirements.
 We then present related work in Section~\ref{sec:related}.
 The related work covers work on cross-lingual document similarity as well as work on cross-lingual cluster linking.
-In Section~\ref{sec:crosslingual} we introduce the main problem of cross-lingual document similarity computation, which is followed by an overview of the models that we considered in Section~\ref{sec:models}.
-We then establish necessary notation in Section~\ref{sec:notation} and proceed with four subsections that describe the main cross-lingual similarity approaches in detail.
-The approach in Section~\ref{sec:kmeans} serves as a baseline, and the approaches in Section~\ref{sec:CCA} and Section~\ref{sec:LSI} serve as the basis on which we proposed a new approach presented in Section~\ref{sec:hublang}.
-In Section~\ref{sec:linking} we introduce the central problem of cross-lingual linking of clusters of news articles and our approach that combines the cross-lingual similarity functions with knowledge extraction based techniques.
+In Section~\ref{sec:crosslingual}, we introduce the problem of cross-lingual document similarity computation, followed by an overview  and description of the main cross-lingual similarity approaches we consider in detail, including a new approach based on hub languages.
+%The approach in Section~\ref{sec:kmeans} serves as a baseline, and the approaches in Section~\ref{sec:CCA} and Section~\ref{sec:LSI} serve as the basis on which we proposed a new approach presented in Section~\ref{sec:hublang}
+ In Section~\ref{sec:linking}, we introduce the central problem of cross-lingual linking of clusters of news articles and our approach that combines the cross-lingual similarity functions with knowledge extraction based techniques.
 Finally, we present and interpret the experimental results in Section~\ref{sec:evaluation} and discuss conclusions and point out several promising future directions.
@@
 
-As a next step, an online clustering algorithm is applied to the articles in order to identify groups of articles that are discussing the same event.
+As the next step, an online clustering algorithm is applied to the articles in order to identify groups of articles that are discussing the same event.
 For each new article, the clustering algorithm determines if the article should be assigned to some existing cluster or into a new cluster.
@@
 
-\section{Related work}
+\section{Related work}\label{sec:related}
 

### 2015-09-05 rupnik2015.tex
typos

@@
 \subsection{Problem definition}\label{sec:tfidf}
-We will first describe how documents are represented as vectors and how to compare douments in a mono-lingual setting.
+We will first describe how documents are represented as vectors and how to compare documents in a mono-lingual setting.
 We then define a way to measure cross-lingual similarity which is natural for the models we consider.
@@
 Processing a multilingual dataset results in several vector spaces with varying dimensionality, one for each language.
-The dimensionality of the vector space corresponding to the $i$-th language is denoted by $n_i$ and and the vector space model mapping is denoted by $\phi_i : \text{text} \rightarrow \RR^{n_i}$.
+The dimensionality of the vector space corresponding to the $i$-th language is denoted by $n_i$ and the vector space model mapping is denoted by $\phi_i : \text{text} \rightarrow \RR^{n_i}$.
 The similarity between documents in language $i$ and language $j$ is defined as a bilinear operator represented as a matrix $S_{i,j} \in \RR^{n_i \times n_j}$:
@@
 \subsection{Cross-Lingual Models}\label{sec:models}
-In this section, we will describe several approaches to the problem of computing multilingual simliarities introduced in \ref{sec:tfidf}. We present four approaches:
+In this section, we will describe several approaches to the problem of computing multilingual similarities introduced in \ref{sec:tfidf}. We present four approaches:
 a simple approach based on $k$-means clustering in \ref{sec:kmeans}, a standard approach based on singular value decomposition in \ref{sec:LSI}, a related

### 2015-09-05 response.txt
responses and an update

Reviewer 1

This paper presents a model for article clusters linking across languages. Although it is stressed from the very beginning that this model is intended to link news articles, both news and Wikipedia articles are used for experimentation.

Fair efforts are made to extensively evaluate the proposed model and demonstrate its performance. Nevertheless, I observe a few important issues including:

1. It is not clear what the main point of the paper is: the proposal of a new cross-language similarity model or its application into event linking across languages.

Authors' response: We have rephrased the introduction that the main point of this paper is to analyze the system as a whole while concentrating on two specific parts of the pipeline.

2. The description of the proposed model is not completely clear (and the departing point is a black box which is not properly described, but the reader is addressed to previous research work).

Authors' response: We have reorganized the paper and added descriptions to make the paper more self contained. Furthermore, we have made our test dataset and pipeline (code) publicly available to allow for further comparisons.

3. I miss some important related models in the comparison (e.g., CL-ESA)

Authors' response: The related work section was revised.

4. Reproducing most of the experiments is close to impossible.

Authors' response: We have added a section on reproducibility where we provide access to relevant code and data in the evaluation.

5. The paper does not flow properly. It is verbose when defining standard concepts and scarce when key aspects should be discussed.

Authors' response: We have revised the paper to improve the flow and presentation.

Please, see my detailed comments below:

PER-SECTION COMMENTS

ABSTRACT
As you reduce your experiments to up to eight languages only, I would avoid being so ambitious as for claiming "it can scale to 100 languages".Even if this could be truth, no evidence is presented in this paper to support this claim.

Authors' response: We have revised the abstract and removed the claim (although xling.ijs.si, implemented by the authors, demonstrates that this is possible, the reviewers point that we have not tested it in this context remains valid).

SECTION 1

"Machine translation remains relatively rudimentary -
allowing people to understand simple phrases on web pages, but remain inadequate for more advanced understanding of text." Why is this sentence relevant to justify your current work?

Authors' response: We have found that a common question is why is direct machine translation (e.g. via Google translate) insufficient for extending monolingual techniques - this sentence is meant to address that.

Figure 1: is this snapshot part of your own system? If not, please give proper credit to the source.

Authors' response: The figure is part of our system. We included an elaboration:  "The results shown in the figure can be obtained using the query \url{http://eventregistry.org/event/997350\#?lang=eng\&tab=articles}. The content presented is part of the Event Registry system, developed by the authors."

SECTION 2
Figure 2: This figure does not help to better understand the pipeline as not too many (additional from the text) details are given. Please, stress the modules that represent your current contributions, maybe with some different color.

Figure 3: Once again, this figure does not contribute to better understanding your model. May I suggest you to describe the different steps in a list and drop the figure?

Authors' response: Both images were removed and replaced with an image that provides information about the components in the pipeline. The image is much more informative and helps to clarify the overall system as well as the components that are described in this paper.

"A new document is first tokenized, stop words are removed and words are stemmed." I understand these are "generic" steps that you run on different languages and this causes the tools, algorithms and vocabularies different. I suggest you to add a table summarizing the tools, resources, and the proper associated citations (alternatively, it could also be an appendix)

Authors' response: We have added references to the components and tools used in the system pipeline.

"The clusters, we consider must have..." Please, rewrite.

Authors' response: The text has been rewritten.

"The system is described in more detail elsewhere (Leban et al., 2014b, 2014a)." I miss here the proper details of the clustering strategy. How is the on-line algorithm set up? Is the model deterministic? That is, no matter what documents come first, the final clustering is going to be the same? How do clusters and centroids evolve? I understand most of this represents previous work. Still, try to make this paper as self-contained as possible.

Authors' response: More details about the clustering approach have been added. Additionally, a reference to a paper describing the clustering approach has been also provided.

SECTION 3
"The system should support as many languages as possible." You can safely remove this sentence

Authors' response: The sentence has been removed.

Section 3.1: regarding the desirable properties, I believe they are essential and may not be necessary to mention them in such a detail (they divert the attention of the reader).

Authors' response: We have elaborated further. The requirements now have their own section and we stated them because some of them are specific to our use case - linking large news streams.

"Supporting a new language should be straightforward." It is only straightforward if a comparable corpus is at hand. I understand such resources are found with more ease, but this does not mean that adding new languages is straightforward.

Authors' response: The sentence has been removed.

You say 3 families, but include 4. Beside considering the standard literature on the topic (e.g., [1]), you could consider the analysis of cross-language similarity measures included in [2] to compliment (even re-arrange) yours. One particular model used there ---CL-ESA---, is pretty similar to yours and should be worth including it in the comparison.

Authors' response: We have revised the section and extended the related work, which now covers CL-ESA as well.

The discussion about classification-based models is scarce. The description nor the justification are convincing.

Authors' response: We removed the classification-based references, since we determined they are not suitable for cross-lingual similarity computation (the first reference is based on translations, the second reference only finds one dimensional projections).

Page 6: "Based on the discussion above [...]". I do not see an actual discussion of the models. They are just listed (acronyms and extended version, and a short reasoning to justify why the authors discard it are included). This is the risk when no proper section to review related work is included.

Authors' response: We have rephrased the sentence and restructured and extended the related work section.

Section 3.2: "how to compare" Compare what?

Authors' response: The phrase in question was changed to: "how to compare documents"

"A document TFIDF vector is its original vector multiplied element-wise by the weights." This sentence does not make sense. All in all, your definition of the tf-idf weighting schema is too verbose and even confusing. This is a standard model for weighting the relevance of a term in a document. Please, reduce the verbose description and be more precise; a couple of sentences should be more than enough (you might use the resulting extra space to better discuss the related work).

Authors' response: we have revised the paragraph on TFIDF.

"Cosine similarity, and other related approaches, assumes that the similarity is reflected in the overlap of words, and as such works only when the documents d1 and d2 are written in the same language." This is false. Cosine measures the similarity between two vectors, regardless of what such vectors represent. In this case, they are words, but it could be anything. If one would be able to come out with a vectorial representation valid across languages (i.e., by mapping the terms into a common space), cosine could be used as well. Please, fix this issue.

Authors' response: The paragraph was revised and made more precise.

SECTION 4
In general, I find the definitions of the models a little bit cryptic. In particular, I find the connection between canonical correlation analysis and hub languages obscure.

Authors' response: We have extended the presentation of both CCA and its extension.

When listing the three approaches (which might be included in the paragraph rather than appear itemized), add a short description of the third model as you did with the other two.

Authors' response: We have added additional remarks on the third model and restructured the text.

Section 4.2: It is not clear to me whether these matrix representations/clusters are for the Wikipedia comparable corpus or for the news corpus.

Authors' response: We have included a clarification.

Please, add the proper citation for "Gram-Schmidt step".

Authors' response: We have added a reference to Gram-Schmidt.

SECTION 5
The distribution of this section is poor. For instance, resources are used before they are defined. This is counterintuitive. Please, reorganize as necessary.

Authors' response: The text has been rearranged: we moved the features before the algorithm and the dataset description to evaluation section.

Section 5.1: what are both l_a and l_b? Are they the different languages? You had defined them as l_1, l_2,... l_k just a few lines before.

Authors' response: The issue was fixed with more consistent notation.

Figure 4 and around. The caption fails to describe the figure, making it hard to understand. Please, move this discussion to the main text (if it is actually necessary) and add a caption describing the figure. For instance. what is the meaning of the different-color arrows?

Authors' response: The figure has been redrawn. Figure caption describes only what is shown in the Figure, implications and explanations are moved to the main text.

Moreover, I am confused here. I thought you had said that the computation of similarities was made with respect to clusters' centroids.Was that only true for the monolingual clustering process? Is it true for this cross-language setting?

Authors' response: This is now clarified in text.

Last paragraph of 5.1: How do you deal with this issue exactly? What are the constraints?

Authors' response: We have added an explanation to the last paragraph and improved the Figure.

Section 5.2: I guess with "the lists" you mean the three vectors?

Authors' response: Yes, we have corrected the text accordingly.

Algorithm 2 is not necessary. The ML pipeline in which features are obtained for classifying an object is well known. You already describe everything necessary (what features are used, what learner is applied, how the data was annotated) later.
Authors' response: Algorithm 2 has been removed and the text was updated accordingly.
Still, I miss a few things: what is the information displayed to the annotators? Does it resembles the features? Is it just a collection of texts? 
 Authors' response: More details have been added.
Are the resources (e.g., annotated data) publicly available?
 Authors' response: Yes, we now explain this in the subsection on repeatability of experiments.

Section 5.4: The descriptions in this subsection are confusing. It is not straightforward to map that two articles are linked across languages if one of them appears among the 10 most similar documents of the other one.

Authors' response: We have cleaned up the descriptions.

Up to now, I find myself confused about whether you are talking about Wikipedia articles or news articles. For instance, when you talk about cross-language article linking, the idea of inter-language links in Wikipedia comes to my mind. Even if I accept that this confusion could be the result of my personal background, I encourage you to make the descriptions more clear and sharp to avoid it.

Authors' response: We now use “news article” instead of “article” in most places.

"an interface displaying information" What interface was shown to the annotators? What information they had at hand?

Authors' response: We have added additional information on the annotation process.

Cross-lingual article linking features: As the authors give no details about this process, understanding how this was used (and how it is used now) is difficult. Moreover, if I understood properly, the cluster-generation process is fully monolingual. What exactly is this cross-lingual linking information?

Authors' response: We have added details describing how we get from cross-lingal similarity to cross-lingual links using a 10-nearest neighbour approach.

Concept related features: I do not feel particularly comfortable with your claim that finding a named entity and identifying the corresponding Wikipedia article means semantically annotating a text.

 Authors' response: The text has been rephrased.

"[...] for cases where the weights are not relevant, [...]". In what cases the weight is not relevant when computing a similarity? Please, specify. The measure is called "Jaccard coefficient"; please add the corresponding reference.

 Authors' response: The text has been revised and a reference has been added.

SECTION 6
Section 6.1 should not be part of the evaluation, but of the description of the resources applied in your model. Therefore, you should move it to Section 5.

Authors' response: We have reorganized the paper, so that Section 4 and Section 5 describe the methodology and Section 6 describes the data and the experiments. The approaches in Section 4 and Section 5 are general and independent of the Wikipedia dataset. Wikipedia was used to build and evaluate the models from Section 4 and we feel that the most appropriate place to introduce it is in the Evaluation section.

Include the proper citations for both Eurovoc and Europarl please.

Authors' response: We have included the proper citations.

What special namespace? Do you mean for instance the categories or discussion pages? Please, clarify.

Authors' response: We refered to categories and discussion pages. Clarification has been included in the paper.

Page 18, last paragraph: Even if a naive solution, it seems to make sense. Something that I miss is how you handle those cases in which an article in one language links to a missing one in the other language. 
Authors' response: Missing links are simply omitted in our approach.
I am assuming you are only generating bilingual matrices here. Otherwise, I miss how you handle cases in which {a,b} are articles in language 1 and c is an article in language 2. What happens if a points to c and c points to b? Both a and b become equivalent to c? This is not 100% correct.

Authors' response: This results in multiple documents pointing from the document c, if first one is a, link to b is omitted
  a  b  c
a 0  0  1
b 0  0  1
c 1  1  0
In the rare case that after symmetrization we have multiple links pointing from the document, we pick the first one that we encountered.

Last paragraph in 6.1: If you are actually processing the more than 250 languages in Wikipedia, "all" is correct. If not, please be sharp and mention only the number of editions you are processing (and for which you report results here).

Authors' response: We have added a footnote to clarify: \footnote{https://www.wikipedia.org/ dumps available in 2013}

Section 6.2: Why do you measure these languages in terms of native speakers? Measuring them in terms of Wikipedia articles seems to make more sense in this context.

Authors' response: We have added the number of articles in wikipedia to the text and  highlighted the difference with native speakers. We feel the latter is relevant since the number of news stories does depend on the number of speakers, meaning we may need to compare many articles from languages which have small wikipedias.

"Although the English language is well aligned with all Wikipedia languages". This is false. For instance, see [3,4].

Authors' response: This is true in the sense that number of articles is  highly correlated with the number of articles connected to English. It, however,  does not imply anything about the quality of alignment. The point of [3] is in fact to  shows that quality of alignment is very bad.

"Furthermore, we call the document consisting of less than 20 different words, a stub." Remove this. It is implicit in your preprocessing description. You can include the description "This documents are typically..." in a footnote.

Authors' response: The text has been moved to a footnote.

"The evaluation is based on splitting the data into training and test sets (which are described later)." It would be better if you describe this right now. Otherwise, the reader has to go forward and backward all the time to understand what is going on.

Authors' response: The description of test and train test was moved as proposed.

Table 1: I guess the number on the left (right) corresponds to 500 (1000) topics? Please, describe it explicitly.

Authors' response: We have explicitly described it in the caption.

"The fist row represents the size of the training [...]" Move this to the caption of Table 2. In general I find your captions poor and even misleading (e.g., see Table 4 and Figure 5).

 Authors' response: We have moved the description into the caption.

Table 2: The way in which you build the training-test sets causes the results obtained on the different language pairs incomparable. The distributions of training-test sets is also completely different across languages. Moreover, I guess you do not have training instances for hi-[ht,pms], war-pms and others? How can you still run experiments on these languages (with such a high performance, like in hi-pms)? Am I missing something?

Authors' response: Scores are not directly comparable and meant to be only used pairwise. The test documents are the same.  One of the points of our system is that we use the hub language as a proxy language. By mapping into the proxy language from the target and source language, we can obtain a mapping between source and target. The quality of these is presented in the experiments.

"In this technique, the dataset is partitioned [...]" Remove the description of 10-fold cross validation. This procedure is well known.

Authors' response: We omit the description.

"The table shows for each of the algorithms the obtained classification accuracy, precision and recall." Move it to the caption.

Authors' response: We moved the text into table's caption.

Figure 6.3: do you mean Figure 5?

Authors' response: The reference has been fixed.

Table 4:
- CA -> A
Authors' response: CA has been replaced with Accuracy
- The number of features corresponds to the size of the vocabulary?
Authors' response: The number of features corresponds to the number of "topics" or the dimension of the latent space.
- As you are running a fold validation, these must be average values. Please, add the standard deviation as well.
Authors' response: Standard deviations were added for the tables where the deviations fit - in some tables deviations would make the table too wide and very hard to read.
- Why is k-means applied with 500 features only?
Authors' response: We have provided an explanation in the paper - the reason is high computational cost and low quality results on 500 features.
- It is interesting to observe that LSI performs better with less features. Why? How sensitive are the models to the number of features?
Authors' response: A possible interpretation is that using the truncated SVD decomposition acts as denoising. The progression of 500 - 800 - 1000 aims to demonstrate the diminishing return of adding more latent features. Higher numbers of feature also induce higher similarity computation cost - there are two tradeoffs: the computational complexity and quality of results.
- All in all, the results of CCA and LSI seem quite comparable.
Authors' response: This is not surprising as CCA is just a refinement.
- Can you report the results for each language pair?
Authors' response: The table with these results was added
- Please, add F-measure (also to Table 5).
Authors' response: F measures were added to some of the tables. In some experiments, the table becomes too wide so it was not added. If necessary, the results are already computed in the results folder in the two html files (available at https://github.com/rupnikj/jair_paper/tree/master/results).

Table 5: what is large/small? Maybe this evolution could be better observed in a figure and that would avoid you the necessity of defining large and small.

Authors' response: The details about this are now provided and the text has been rewritten.

I guess "cross-lingual article linking features" means concepts in the table. Linking them is not so straightforward.

Authors' response: The concepts correspond to performing "semantic" analysis on texts, where texts are annotated with entities, and using those annotations sets for cross-lingual similarity computation. The cross-lingual article linking features are derived based on cross-lingual comparisons of documents contained in clusters. We have revised the part that describes the features to make those details clearer.

"The computed features are however significantly less informative compared to the features computed on the annotated concepts." What computed features?

Authors' response: The text has been rewritten to make this point clear.

Table 6: for completeness, Misc+concepts should be added as well.

Authors' response: The missing feature pairs have been added.

"This is sufficient to do similarity computation over 150 languages if needed." Either justify this estimation, show the empirical evidence, or drop this sentence.

Authors' response: We have revised the sentence.

REFERENCES
[1] Peters, C., Braschler, M. and Clough, P. (2012) Multilingual Information Retrieval: From Research to Practice, Springer: Heidelberg, Germany, ISBN 978-3-642-23007-3, 217 pages. Available from Springer.

[2] Potthast, Barrón-Cedeño, Stein, and Rosso. Cross-Language Plagiarism Detection. Language Resources and Evaluation (LRE), 45 : 45-62, March 2011
Typos and grammar issues:

[3] Paramita, Clough, Aker, Gaizauskas: Correlation between Similarity Measures for Inter-Language Linked Wikipedia Articles. Proc. of the 8th Intl. Language Resources and Evaluation (LREC 2012). pp. 790-797. ELRA, Istanbul, Turkey (2012)

[4] http://en.wikipedia.org/wiki/Wikipedia:WikiProject_Interlanguage_Links

TYPOS:
the Ukraine -> Ukraine
corresponding Section -> section
desireable ->
the the
new articles -> news (?)
the c_j -> c_j
millionarticles
also have a

Authors' response: We have corrected the paper accordingly. The authors believe that using the phrase "news articles" makes the presentation clearer.
----------------------------------------------------------------------------------------------------
Reviewer 2

The paper presents research on cross-lingual document similarity using CCA, and the use of the method as the basis of an event-tracking application on multilingual news feeds.

 In my opinion the research described in the paper is original and relevant, and it is an excellent match for the Special Track.

However, I think the the paper needs a significant effort to improve its clarity and organization before being publishable. Details follow.

* Figure 1 is illegible. Please take the screenshot with better resolution.

Authors' response: We have added a higher resolution screenshot as well as a link.

* Figures 2 and 3 are rather obvious and maybe innecessary. On the other hand, some figures would help to understand the construction of the matrices described in section 4.

Authors' response: We have added additional figures.

* Section 4 is a bit messy. The authors state that they present 3 approaches, but then they present 4 (because they introduce CCA "for completeness"). Also, the itemize at the beggining of section 4 could be better correlated with its corresponding subsections (either with more similar names, or just referencing the subsection for each item) to ease the reader's path trough the paper.

Authors' response: We revised the section 4. Cross-Lingual Models  accordingly.

* In section 4.5, the sentence "we describe an extension to CCA, but is more applicable to ...", is unclear.  It should be "..., which is applicable to.." (?)
(btw, I don't think "applicable" is a gradual adjective. It is either applicable or not, but it can't be "more applicable")

Authors' response: We have revised the text according to the recommendation.

* Avoid vague references such as "as described above/below".
 E.g. section 5 "the main application on which we test the above similarity" should be "... we test the similarity described in section 4.5"  (there were 4 similarities "above")

Authors' response: We have corrected the issue.

* Figure 4 is referred in section 5.1 as an example of generalized matching (vs bipartite matching). The same figure is referred in section 5.3 as an example of the two stage algorithm of filtering a reduced set of candidate clusters to then compute their actual similarity. I can see how the Figure shows the later, but I fail to see how it shows the former. If two figures are needed use two, do not reuse, because it causes confusion (though in this case, I'd say that the generalized matching idea does not need a support figure).

Authors' response: We have corrected the issue.

* Section 6 is also a bit messy. The introduction mentions three sets of experiments, but then only two subsections follow. The third "set" of experiments seems to have been clustered into the second set. Please clarify the section introduction and state which subsections describe each set of experiments (ideally one subsection per set).

Authors' response: We have rephrased it to indicate the two sets of experiments and rewritten the section introduction.

* Section 6, introduction "..compare the main approaches the we presented..." should be "...that we presented"

Authors' response: We have corrected the issue.

* Section 6.1, "millionarticles" => "million articles".  Please use a speller.

Authors' response: We have corrected the issue.

* Section 6.2 presents the cross-lingual document similarity results, and briefly discusses them, to go next to the train/test set construction. I think it would be clearer if first the data sets are presented  (maybe even in their own subsubsection) and after, results are reported and discussed.

Authors' response: This has been corrected -- the train/test construction is now described first. We believe that the additional subsection would be more distracting than clarifying.

* In Caption of Table 1,  "500 topics;1000 topics" should be
  "500 topics - 100 topics" (with a dash) to mimic the format
  of the data cells, thus suggesting the right interpretation.
  The same in Table 2:  "training:test" should be "training - test"

Authors' response: We have corrected both issues.

* The description of Table 2 is unclear. The first row and the diagonal are described, but they contain exactly the same values. Is that an error?
If it is not, then, what is the point of having them both?

Authors' response: This is correct. We only select articles aligned with English. We have added a clarification in the figure caption.

* Table 3 should follow the pattern of tables 1 and 2, and use a dash as
  separator, instead of presenting values as pairs (which they are not).
  Also, caption should express what are the left/right values respectively.

Authors' response: We have revised the paper accordingly.

* Last paragraph in section 6.2: "There is also have a problem"

Authors' response: We have corrected the issue.

* Section 6.3 also needs to improve in clarity.
  - The first paragraph ends with a completely unnecessary description of what
    is 10-fold cross validation. I guess JAIR readers do know what it is.
Authors' response: We have addressed this in the paper.
  - The second paragraph is completely confusing. It describes that a dataset
    with only two features (linkCount and avgSimScore) is used, but then
    results are shown in Table 4, with presents results with
    500-800-1,000 features.
Authors' response: We have rewritten the part related to describing the experiment.
  - (BTW, spell figures over 1,000 with appropriate commas:
     i.e. 15,000 and 300,000 instead of 15000 and 300000)
Authors' response: We have corrected the issue.
  - Third paragraph: "manages with 500 features performam comparably to LSI"
    ==> performing (??)
Authors' response: We have corrected the issue.
  - fourth paragraph: "choosing number of features vectors larger than..."
    ==> "choosing a number of features larger than... "   (I guess....)
Authors' response: We have corrected the issue.
* Section 6.4
    - "400k similarities per 200ms" is a weird measure.  Why not just
       "2 million similarities per second" ?
    - "process more than million articles" => "process more than one million articles"

Authors' response: Both issues were corrected.

* Section 7
This section should summarize the presented research, highlight the main contributions, and outline future research lines. It more or less does, but again in a messy style. There is an itemize that shows three items of different types: One contribution, one result, and one further work. If they are different things, do not put them in the same itemize, use different paragraphs.

Authors' response: We have removed the bullets and updated the content in the discussion. The content is now cleaner and should flow nicely.

     Also, some connective text would be nice, introducing each part (instead of an itemized list): when the summary starts, when contributions are higlighted, when future work is described (which should be completely separated from the rest).

Authors' response: This was also taken into account - the future work is now a subsection.

----------------------------------------------------------------------------------------------------
Reviewer 3

The authors describe and evaluate several methods to link clusters of news articles (produced elsewhere) across a wide range of languages. The multilingual feature spaces used for the cross-lingual similarity calculations are produced on the basis of Wikipedia articles and their cross-lingual links, including via hub languages. Canonical Correlation Analysis (Hub-CCA) was identified as being the best-performing method. The methods were tested both on news article clusters and on Wikipedia pages. They were tested not only for main language pairs, but also for language pairs with little coverage.

The task of linking similar documents across languages is rather difficult and also computationally demanding, especially when extending to many different languages and language pairs. Relatively little previous work exists and only a very small number of real-life working systems with such functionality have been developed, while the importance of cross-lingual information access is of increasing importance in our globalised world. The authors have put a lot of work into developing the methods and the system. The evaluation is very comprehensive and convincing. The article is generally well-written and clear (but many typos will need correcting).

Questions, comments and suggestions for improvement, section by section:

1. Introduction

I feel that you need to describe more clearly what the term ‘event’ means for you and what kinds of ‘events’ you intend to link. Are events time-limited? Are the many articles, spread over several weeks, of the recent Ebola outbreak in a whole range of countries one event or many? Similarly for discussions on bird flu, global warming, (civil) war in Ukraine, etc. In such cases, the borderline between events on one hand and topics (subject domains) on the other is not clear. You cannot solve this, but you can take a stand on what you intend to include. You took a practical stand by deciding that you will only look at articles up to four days of age. A possible extension of your system (future work) might be to keep news clusters ‘alive’ (i.e. link to them) as long as new articles get added to them.

Authors' response: The reviewer is correct that the term event is ill-defined. In the end, we used human annotation to define what belongs to an event. We agree that where to draw the line where do define an event is an interesting and highly non-trivial question and well beyond the scope of this paper. An event for us, is a spatio-temporal entity, where articles must be sufficiently close in time, space and content. We have added this explanation to the introduction.

3. Cross-lingual Document Similarity

I feel that section 3.1 should be split into ‘Requirements of our system’ (or the like) and ‘related work’ as the first page of 3.1 describes your needs/your settings rather than the work of others.

Authors' response: The requirements have been moved into the pipeline section. The related work is now a separate section.

You mention the cross-lingual news linking work in NewsExplorer/Europe Media Monitor (EMM) at a later point, but shouldn’t the cross-lingual linking methods presented by Pouliquen et al. be included in your section 3.1? Some of their papers explicitly referring to the cross-lingual linking process are:

-          Pouliquen & Steinberger (2008). Story tracking: linking similar news over time and across languages.
-          Pouliquen et al. (2004). Automatic Linking of Similar Texts Across Languages.
-          Pouliquen et al. (2004). Multilingual and Cross-lingual News Topic Tracking.
-          Pouliquen et al. (2003). Automatic Identification of Document Translations in Large Multilingual Document Collections.

Authors' response: The papers by Pouliquen overlap quite a bit and describe the similar methodologies. We’ve added another reference to the “Story tracking …” paper and extended the cross-lingual similarity related work section.

The latter seems to address an alternative way of evaluating the cross-lingual text linking in the setting where it is unknown whether an equivalent cross-lingual equivalence exists.

Authors' response: We have considered the evaluation reported in Pouliquen et al. (2003), but we concluded that such an evaluation is most relevant in the setting of plagiarism detection, where one wants to detect translated content. In our case we focus primarily on finding comparable documents, the scenario most relevant to the event linking component in the full EventRegistry system.

In 3.2, on the basis of what corpus are the IDF values produced?

Authors' response: We have added a footnote. The IDF is computed dynamically for each new article over all the articles in the feed that are within a 10 day window.

4. Cross-lingual models

What are your criteria to distinguish ‘hub languages’ from the others? How many hub languages do you use? Or maybe I did not understand well and this does not matter?

Authors' response: The hub languages are languages that are present in a large portion of multilingual documents. If a Wikipedia page written in a non-English language has one or more links to variants of the page in other languages, English is very likely to be one of them. That makes it a hub language.

Section 4.5 is three pages long and I wonder whether all the contents fit under the header ‘hub languages’. Please check and amend, if needed.

Authors' response: We restructured and revised the section to provide a clearer presentation.

The large number of steps and methods used in the processes described in section 4.5 make it hard to determine where anything might go wrong or where anything might not be optimal.

Authors' response: We split the presentation into subsections that motivate the method, expose the main assumption in the data and how to exploit it in two steps (dimensionality reduction and optimization problem simplification).

5. Cross-lingual event linking

Under ‘related work’ (5.2), you may want to mention the multilingual NewsTin media monitoring system. To my knowledge, they cover eleven languages. Although they apply multilingual categorisation rather than direct cross-lingual linking of related news, the functionality is similar.

For EMM, the most representative overview publication I could find is:

-          Steinberger et al. (2009). An Introduction to the Europe Media Monitor Family of Applications.

Authors' response: The related work has been moved to different part of the paper and updated according to the recommendations.

In 5.3, you explain that you calculate cross-lingual similarities for all individual news articles, which seems to be a rather heavy process. I understand that you do initially need to do this as you use the results to identify the most similar clusters in the other languages. However, wouldn’t it make sense to then use the output of your ‘Algorithm 1’ to learn how to link the clusters directly?

Authors' response: Article-article similarity is computed only when new articles appear in news stream. Clusters change as new articles are available and that would require re-evaluation of cluster matches after each update.

Do you compare all individual articles across languages on the same day only or on all articles within your four-day window?

Authors' response: We compare articles within a one day window.

Please clarify (on page 17, concept-related features) how you recognise mentions of people, locations, keywords, etc. as it is non-trivial to do this for a large variety of languages. Is there any word sense disambiguation involved? Or do you only consider those ‘concepts’ that are marked with a hyperlink?

Authors' response: We have added a reference to the disambiguation approach.

I have the same remark regarding the ‘miscellaneous features’, where you say that you recognise the event locations. This, again, is a non-trivial task as place names are frequently homographic with person names, with other place names and also with common words of a language. Furthermore, usually more than one location name is mentioned in an article or an article cluster. How do you recognise THE event location? This needs clarifying.

Authors' response: We have included an explanation that this is provided by the EventRegistry system and it is a task on itself.

I believe that when you (repeatedly) refer to Figure 6.3, you want to make reference to Figure 5. Please correct.

Authors' response: The reference has been fixed.

It would be interesting to see some sort of an error analysis on wrongly identified web links.

Authors' response: We agree with the reviewer that a better understanding of the wrongly identified links would be interesting. If such additional structure could be found in the errors, it could certainly improve the results. However, we do not know of an established methodology to perform such an analysis and feel that developing a new methodology for this is out of the scope of this paper.

Good work!

========================
English language corrections:

The paper is rather well written and it is mostly clear. However, there are many small English mistakes. In some sections there are more than in others, so if your best English speaker could go through the final text, this would be useful.

I found two general TYPES of mistakes:

-          Lacking hyphens. I believe that the general rule is that you need a hyphen if you combine two different parts of speech as noun modifiers, e.g. ‘concept-related features’, ‘language-(in)dependent space’, ‘low-rank approximations’, ‘hub language-related blocks’ (but no hyphen in noun-noun compounds!), ‘well-studied problem’, ‘translation-based’, ‘factorization-based’,  ‘clustering-based’, ‘sample-based’, ‘lower-dimensional’, ‘low-dimensional’, ‘concept-related’, ‘500-dimensional’, ‘LSI-based’, ‘CCA-based’, ‘extraction-based’, etc.

-          Missing determiners. This is a particular challenge for people speaking Slavic languages and it is difficult to give some generic advice on when to use the definite, indefinite or no determiner.

-          ‘That is’. I feel that you overuse this expression a bit, but mostly you should add ‘that is’ to the sentence it refers to (add a comma before) and certainly not start a new sentence with ‘That is’.

And here is a non-exhaustive list of proposals for changes:

-          Tracking and events à tracking of events (in the Abstract)
-          The term ‘event’ (usage ‘event’ as meta-language term)
-          ‘The disambiguated and entity linking of the concepts’ à ?
-          Please reformulate the last sentence of the caption of Figure 2. I find it hard to parse.
-          assumption being that à assumption is that
-          Page 3, last sentence of Section 1: you twice use ‘first’, while the second step should probably be called ‘Second’ or ‘Then’?
-          multi-lingual à multilingual
-          Caption of Figure 3: add a colon between ‘articles’ and ‘it first’?
-          Pre-processing on the à of the
-          corss-lingual à cross-lingual
-          take advantage of comparable corpus à corpora
-          classification problem related à classification problem-related
-          Page 7: Standard vector à The standard vector
-          corresponds to word or phrase à words or phrases (Or: a word or a phrase)
-          approaches, assumes à approaches, assume
-          the the maximal à the maximal
-          we describe … described in à reformulate
-          related work section à section on related work (?)
-          is truncated diagonal matrix à is a truncated …
-          by QR algorithm à by the QR algorithm
-          sources – after which à sources. After this (?)
-          with an additional assumption à with the … (?)
-          to CCA, but is à to CCA, which is
-          where where à where
-          to can be recovered à can be …
-          and objective is à and the …
-          sum of squared correlation à correlations
-          particularly new articles à news articles (?)
-          about same event à about the …
-          . First vector à . The first vector
-          algoirthm à algorithm
-          millionarticles à million articles
-          can point to redirection link also à redirection links
-          link as inter-language links à link because inter-language links (confusing otherwise: ‘we store the link as something else’)
-          If inter-language link à If an …
-          transform this matrix to symmetric à to be symmetric
-          of CCA hub à of the CCA hub
-          This documents are à These (Or: Such)
-          We perform the two-step à a two-step (?)
-          Train and test sets à training and test sets
-          Thorough the hub à through
-          represent number of à represent the number of
-          for each training English matrix à for each English training matrix (?) (Or: each training of the English matrix?)
-          of Creole language à of the …
-          many of remaining à many of the remaining
-          confirmed for Creole … language à confirmed for the Creole … languages
-          is also have a à is also a
-          look at test à look at the test
-          on other side à on the other side
-          in multilingual setting using English, … language à in a multilingual setting using the English, … languages
-          how accurate is the prediction model when à how accurate the prediction model is when
-          performs the worse à the worst
-          500 features performain à performing
-          to LS based approach à to the LSI-based approach
-          choosing number of features vectors à choosing the number of feature vectors
-          with large number of documents à large numbers. 
Authors' response: The authors believe that both versions are grammatically correct.
-          processor machine with à machines
-          depend of number à depend on the number
-          more than million à more than one million
-          variation of hub approach à of the hub …
-          enables projection à enables the projection
-          a factor around 4 à a factor of around 4
-          present good baseline à present a good …
-          independently of the à independently from the (?)

Authors' response: We have corrected the English language related problems.

----------------------------------------------------------------------------------------------------

### 2015-09-05 rupnik2015.tex
responses and an update

@@
 Each Wikipedia page is embedded in the page tag.
-First, we check if the title of the page consists of any special namespace (which includes categories and discussion pages) and do not process such pages.
+First, we check if the title of the page starts with a Wikipedia namespace (which includes categories and discussion pages) and do not process the page if it does.
 Then, we check if this is a redirection page and we store the redirect link because inter-language links can point to redirection links also.

### 2015-09-05 rupnik2015.tex
updates

@@
 In this work, we address the problem of tracking of events in a large multilingual stream.
-Within a recently developed system EventRegistry  \shortcite{Leban2014W,Leban2014I} we examine two aspects of this problem: how to compare articles in different languages and how to link collections of articles in different languages which refer to the same event.
+Within a recently developed system Event Registry  \shortcite{Leban2014W,Leban2014I} we examine two aspects of this problem: how to compare articles in different languages and how to link collections of articles in different languages which refer to the same event.
 Taking a multilingual stream and clusters of articles from each language, we compare different cross-lingual document similarity measures based on Wikipedia.
@@
 
-The paper is organized as follows: we first provide an overview of the system as a whole (Section~\ref{sec:pipeline}), which includes a subsection that summarizes the main system requirements.
+The paper is organized as follows: we first provide an overview of the system as a whole in Section~\ref{sec:pipeline}, which includes a subsection that summarizes the main system requirements.
 We then present related work in Section~\ref{sec:related}.
 The related work covers work on cross-lingual document similarity as well as work on cross-lingual cluster linking.
-In Section~\ref{sec:crosslingual}, we introduce the problem of cross-lingual document similarity computation, followed by an overview  and description of the main cross-lingual similarity approaches we consider in detail, including a new approach based on hub languages.
+In Section~\ref{sec:crosslingual}, we introduce the problem of cross-lingual document similarity computation, which is followed by an overview  and description of the main cross-lingual similarity approaches we consider in detail, including a new approach based on hub languages.
 %The approach in Section~\ref{sec:kmeans} serves as a baseline, and the approaches in Section~\ref{sec:CCA} and Section~\ref{sec:LSI} serve as the basis on which we proposed a new approach presented in Section~\ref{sec:hublang}
@@
 and build a vector space using the union of tokens over the languages.
-A cosine similarity function in such a space can be useful to some extent, for example "internet" or "Obama" may appear both in a Spanish and English texts and the presence of such terms in both an English and a Spanish document would contribute to their similarity.
-In general however, large parts of vocabularies may not intersect, which means that given a language pair, many words in both languages cannot contribute to the similarity score, which makes the similarity function very insensitive.
+A cosine similarity function in such a space can be useful to some extent, for example "internet" or "Obama" may appear both in Spanish and English texts and the presence of such terms in both an English and a Spanish document would contribute to their similarity.
+In general however, large parts of vocabularies may not intersect, which means that given a language pair, many words in both languages cannot contribute to the similarity score, which makes the similarity function very insensitive to the data.
 
@@
 \subsection{Cross-Lingual Models}\label{sec:models}
-In this section, we will describe several approaches to the problem of computing multilingual similarities introduced in \ref{sec:tfidf}. We present four approaches:
-a simple approach based on $k$-means clustering in \ref{sec:kmeans}, a standard approach based on singular value decomposition in \ref{sec:LSI}, a related
-approach called Canonical Correlation Analysis (CCA) in \ref{sec:CCA} and finally a new method, which is an extension of CCA to more than two languages in \ref{sec:hublang}.
+In this section, we will describe several approaches to the problem of computing multilingual similarities introduced in Section~\ref{sec:tfidf}. We present four approaches:
+a simple approach based on $k$-means clustering in Section~\ref{sec:kmeans}, a standard approach based on singular value decomposition in Section~\ref{sec:LSI}, a related
+approach called Canonical Correlation Analysis (CCA) in Section~\ref{sec:CCA} and finally a new method, which is an extension of CCA to more than two languages in Section~\ref{sec:hublang}.
 %
@@
 To tackle this, we propose a method which consists of two ingredients.
- The first one is based on an observation that certain datasets (such as Wikipedia) are biased towards one language (English), which can be exploited
- to reformulate a difficult optimization problem as an eigenvector problem. The second ingredient is dimensionality reduction using LSI, which
+ The first one is based on an observation that certain datasets (such as Wikipedia) are biased towards one language (English for Wikipedia), which can be exploited
+ to reformulate a difficult optimization problem as an eigenvector problem. The second ingredient is dimensionality reduction using CL-LSI, which
  makes the eigenvector problem computationally and numerically tractable.
@@
 Note that individual document $u_i$ can be an empty document (missing resource) and each $d$ must contain at least \textbf{two nonempty documents}.
-This means that in our analyses we discard strictly monolingual documents for which no cross-lingual information is available.
+This means that in our analysis we discard strictly monolingual documents for which no cross-lingual information is available.
 A comparable corpus $D = {d_1, \ldots, d_s}$ is a collection of $s$ multilingual documents.
@@
 
+We will now describe four models to cross-lingual similarity computation in the next sections.
 \subsection{$k$-means}\label{sec:kmeans}
@@
 
+In the next section we will describe our second approach.
 \subsection{Cross-Lingual Latent Semantic Indexing}\label{sec:LSI}
@@
 
-  Since the matrix $X$ can be large we could use an iterative method like the Lanczos~\cite{golub} algorithm with reorthogonalization to find the left singular vectors (columns of $U$) corresponding to the largest singular values.
+  Since the matrix $X$ can be large we could use an iterative method like the Lanczos algorithm with reorthogonalization~\cite{golub} to find the left singular vectors (columns of $U$) corresponding to the largest singular values.
 It turns out that the Lanczos method converges slowly as the gap between the leading singular values is small.
 Moreover, the Lanczos method is hard to parallelize.
-Instead, we use a randomized version of the singular value decomposition (SVD) described in \shortcite{tropp} that can be viewed as a block Lanczos method.
+Instead, we use a randomized version of the SVD \shortcite{tropp} that can be viewed as a block Lanczos method.
 That enables us to use parallelization and speeds up the computation considerably.
@@
 
+In the next section we will introduce our third approach.
 \subsection{Canonical Correlation Analysis}\label{sec:CCA}
@@
 Note that the method in its original form is only applicable to two languages where an aligned set of observations is available.
+The next section will describe a scalable an extension of CCA to more than two languages.
 
@@
  To this purpose, we select a subset of Wikipedia languages containing three major languages, English (4,212k articles)--\emph{en} (hub language), Spanish (9,686k articles)--\emph{es}, Russian (9,662k articles)--\emph{ru}, and five minority (in the sense of Wikipedia sizes) languages, Slovenian (136k articles)--\emph{sl}, Piedmontese (59k articles)--\emph{pms}, Waray-Waray (112k articles)--\emph{war} (all with about 2 million native speakers), Creole (54k articles)--\emph{ht} (8 million native speakers), and Hindi (97k articles)--\emph{hi} (180 million native speakers).
-For preprocessing, we remove the documents that contain less than 20 different words (stubs \footnote{Such documents are typically garbage, the titles of the columns in the table, remains of the parsing process, or Wikipedia articles with very little or no information contained in one or two sentences.}) and remove words occurring in less than 50 documents as well as the top 100 most frequent words (in each language separately).
+For preprocessing, we remove the documents that contain less than 20 different words (referred to as stubs\footnote{Such documents are typically of low value as a linguistic resource.
+Examples include the titles of the columns in the table, remains of the parsing process, or Wikipedia articles with very little or no information contained in one or two sentences.}) and remove words occurring in less than 50 documents as well as the top 100 most frequent words (in each language separately).
 We represent the documents as normalized TFIDF\cite{Salton88term-weightingapproaches} weighted vectors.
@@
 
- The hub CCA outperforms other methods in this sense in terms of precision, which is consistent with its AMRR performance, see Figure~\ref{pic:AMRR}.
-It is worth noting that added accuracy is also beneficial when we wish to choose good representatives of clusters. Therefore, we decided to use hub CCA in the cluster linking component.
+The hub CCA attains higher precision and classification accuracy on the task of linking small cluster pairs than the other methods, while LSI is slightly better on linking large cluster pairs.
+The gain in precision of LSI over hub CCA on linking large clusters is much smaller than the gain in precision of hub CCA over LSI on linking small clusters.
+For that reason we decided to use hub CCA as the similarity computation component in our system.
 
@@
 
-\subsection{Remarks on the repeatability of Experiments}
+\subsection{Remarks on the repeatability of experiments}
 We have made  both the code and data that were used in the experiments publicly available at \url{https://github.com/rupnikj/jair_paper.git}.

### 2015-11-10 rupnik2015.tex
minor revision

@@
 \newcommand{\primoz}[1]{\textcolor{red}{#1}}
+
+\makeatletter
+\setlength{\@fptop}{0pt}
+\makeatother
 
@@
 In this work, we address the problem of tracking of events in a large multilingual stream.
-Within a recently developed system Event Registry  \shortcite{Leban2014W,Leban2014I} we examine two aspects of this problem: how to compare articles in different languages and how to link collections of articles in different languages which refer to the same event.
+Within a recently developed system Event Registry~\shortcite{Leban2014W,Leban2014I} we examine two aspects of this problem: how to compare articles in different languages and how to link collections of articles in different languages which refer to the same event.
 Taking a multilingual stream and clusters of articles from each language, we compare different cross-lingual document similarity measures based on Wikipedia.
@@
 
-Content on the internet is becoming increasingly multilingual.
-A prime example is Wikipedia - in 2001, the majority of pages were written in English, while in 2015, the percentage of English articles has dropped to 14\%.
+Content on the Internet is becoming increasingly multilingual.
+A prime example is Wikipedia.
+In 2001, the majority of pages were written in English, while in 2015, the percentage of English articles has dropped to 14\%.
 At the same time, online news has begun to dominate reporting of current events.
-However, machine translation remains relatively rudimentary - allowing people to understand simple phrases on web pages, but remain inadequate for more advanced understanding of text.
+However, machine translation remains relatively rudimentary.
+It allows people to understand simple phrases on web pages, but remains inadequate for more advanced understanding of text.
 In this paper we consider the intersection of these developments: how to track events which are reported about in multiple languages.
 
-The term event is vague and ambiguous, but for the practical purposes, we define it as ``any significant happening that is being reported about in the media.'' Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015 (Figure \ref{fig:event2}).
+The term event is vague and ambiguous, but for the practical purposes, we define it as ``any significant happening that is being reported about in the media.'' Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015 (Figure~\ref{fig:event2}).
 Events such as these are covered by many articles and the question is how to find all the articles in different languages that are describing a single event.
@@
 By using a language detector, the stream is split into separate monolingual streams.
-Within each monolingual stream, an online clustering approach is employed, where tracked clusters correspond to our definition of events - this is based on the Event Registry system \shortcite{Leban2014W,Leban2014I}.
+Within each monolingual stream, an online clustering approach is employed, where tracked clusters correspond to our definition of events - this is based on the Event Registry system~\shortcite{Leban2014W,Leban2014I}.
 Our main goal in this paper is to connect such clusters (representations of events) across languages, that is, to detect that a set of articles in language $A$ reports on the same event as a set of articles in language $B$.
 
-Our approach to linking clusters across languages combines two ingredients: a cross-lingual document similarity measure, which can be interpreted as a language independent topic model, and semantic annotation of documents, which enables an alternative way to comparing documents.
+Our approach to link clusters across languages combines two ingredients: a cross-lingual document similarity measure, which can be interpreted as a language independent topic model, and semantic annotation of documents, which enables an alternative way to comparing documents.
 Since this work represents a complicated pipeline, we concentrate on these two specific elements.
@@
 
-The first approach represents a continuation of previous work \shortcite{nips,sikdd,nips2,iti} where we explored representations of documents which were valid over multiple languages.
+The first ingredient of our approach to link clusters across languages represents a continuation of previous work~\shortcite{nips,sikdd,nips2,iti} where we explored representations of documents which were valid over multiple languages.
 The representations could be interpreted as multilingual topics, which were then used as proxies to compute cross-lingual similarities between documents.
@@
 The related work covers work on cross-lingual document similarity as well as work on cross-lingual cluster linking.
-In Section~\ref{sec:crosslingual}, we introduce the problem of cross-lingual document similarity computation, which is followed by an overview  and description of the main cross-lingual similarity approaches we consider in detail, including a new approach based on hub languages.
+In Section~\ref{sec:crosslingual}, we introduce the problem of cross-lingual document similarity computation and describe several approaches to the problem, most notably a new approach based on hub languages.
 %The approach in Section~\ref{sec:kmeans} serves as a baseline, and the approaches in Section~\ref{sec:CCA} and Section~\ref{sec:LSI} serve as the basis on which we proposed a new approach presented in Section~\ref{sec:hublang}
@@
 
-We base our techniques of cross-lingual event linking on an online system for detection of world events, called Event Registry \shortcite{Leban2014W,Leban2014I}.
+We base our techniques of cross-lingual event linking on an online system for detection of world events, called Event Registry~\shortcite{Leban2014W,Leban2014I}.
 Event Registry is a repository of events, where events are automatically identified by analyzing news articles that are collected from numerous news outlets all over the world.
@@
 The disambiguation and entity linking of the concepts is done using Wikipedia as the main knowledge base.
-The algorithm for semantic annotation is based on~\cite{zhang2014saaacamactat}.
-As a part of the semantic annotation we also analyze the dateline of the article to identify the location of the described event as well as extract dates mentioned in the article using a set of regular expressions.
+The algorithm for semantic annotation uses machine learning to detect significant terms within unstructured text and link them to the appropriate Wikipedia articles.
+The approach models link probability and combines prior word sense distributions with context based sense distributions.
+The details are reported in~\cite{Milne:2008:LLW:1458082.1458150} and ~\cite{zhang2014saaacamactat}.
+As a part of the semantic annotation we also analyze the dateline of the article to identify the location of the described event as well as to extract dates mentioned in the article using a set of regular expressions.
 Since articles are frequently revised we also detect if the collected article is just a revision of a previous one so that we can use this information in next phases of the pipeline.
 The last important processing step on the document level is to efficiently identify which articles in other available languages are most similar to this article.
-The methodology for this task is one of the main contributions of this paper and is explained in details in Section~\ref{sec:crosslingual}).
+The methodology for this task is one of the main contributions of this paper and is explained in details in Section~\ref{sec:crosslingual}.
 
@@
 The pipeline must be able to process millions of articles per day and perform billions of similarity computations each day.
-Both step rely heavily on similarity computation, which implies that this must be highly scalable.
+Both steps rely heavily on similarity computation, which implies that this must be highly scalable.
 
@@
 \textbf{Translation and dictionary based}.
-The most obvious way to compare documents written in different languages is to use machine translation and perform monolingual similarity, see  \cite{multilingualBook}\cite{plagiarism} for several variations of translation based approaches.
-One can use free tools such as Moses \shortcite{moses} or translation services, such as Google Translate\footnote{https://translate.google.com/}.
+The most obvious way to compare documents written in different languages is to use machine translation and perform monolingual similarity, see  ~\cite{multilingualBook,plagiarism} for several variations of translation based approaches.
+One can use free tools such as Moses~\shortcite{moses} or translation services, such as Google Translate\footnote{https://translate.google.com/}.
 There are two issues with such approaches: they solve a harder problem than needs to be solved and they are less robust to training resource quality - large sets of translated sentences are typically needed.
@@
 The operation efficiency and cost requirements make translation-based approaches less suited for our system.
-Closely related are works Cross-Lingual Vector Space Model (CL-VSM) \cite{plagiarism} and the approach presented in \cite{pouliquen2008story} which both compare documents by using dictionaries, which in both cases are EuroVoc dictionaries \cite{eurovoc}.
+Closely related are works Cross-Lingual Vector Space Model (CL-VSM)~\cite{plagiarism} and the approach presented in~\cite{pouliquen2008story} which both compare documents by using dictionaries, which in both cases are EuroVoc dictionaries~\cite{eurovoc}.
 The generality of such approaches is limited by the quality of available linguistic resources, which may be scarce or non-existent for certain language pairs.
@@
 There exist many variants to modelling documents in a language independent way by using probabilistic graphical models.
-The models include:  Joint Probabilistic Latent Semantic Analysis (JPLSA) \shortcite{platt2010translingual}, Coupled Probabilistic LSA (CPLSA) \cite{platt2010translingual}, Probabilistic Cross-Lingual LSA (PCLLSA)\cite{PCL_LSA} and Polylingual Topic Models (PLTM) \shortcite{polyLDA} which is a Bayesian version of PCLLSA.
+The models include:  Joint Probabilistic Latent Semantic Analysis (JPLSA)~\shortcite{platt2010translingual}, Coupled Probabilistic LSA (CPLSA)~\cite{platt2010translingual}, Probabilistic Cross-Lingual LSA (PCLLSA)~\cite{PCL_LSA} and Polylingual Topic Models (PLTM)~\shortcite{polyLDA} which is a Bayesian version of PCLLSA.
 The methods (except for CPLSA) describe the multilingual document collections as samples from generative probabilistic models, with variations on the assumptions on the model structure.
@@
 The parameter estimation is posed as an inference problem which is typically intractable and one usually solves it using approximate techniques.
-Most variants of solutions are based on Gibbs sampling or Variational Inference, which are nontrivial to implement and may require an experienced practitioner to apply.
+Most variants of solutions are based on Gibbs sampling or Variational Inference, which are nontrivial to implement and may require an experienced practitioner to be applied.
 Furthermore, representing a new document as a mixture of topics is another potentially hard inference problem which must be solved.
@@
 \textbf{Matrix factorization}.
-Non-negative matrix factorization based \cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CL-LSI\shortcite{cl_lsi}\cite{multilingualBook}, Canonical Correlation Analysis (CCA) \cite{Hotelling}, Oriented Principal Component Analysis \cite{platt2010translingual}.
+Several matrix factorization based approaches exist in the literature.
+The models include: Non-negative matrix factorization based~\cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CL-LSI~\shortcite{cl_lsi,multilingualBook}, Canonical Correlation Analysis (CCA)~\cite{Hotelling}, Oriented Principal Component Analysis (OPCA)~\cite{platt2010translingual}.
 The quadratic time and space dependency of the OPCA method makes it impractical for large scale purposes.
 In addition, OPCA forces the vocabulary sizes for all languages to be the same, which is less intuitive.
-For our setting, the method in \cite{nonnegfactor_lsi} has a prohibitively high computational cost when building models (it uses dense matrices whose dimensions are a product of the training set size and the vocabulary size).
+For our setting, the method in ~\cite{nonnegfactor_lsi} has a prohibitively high computational cost when building models (it uses dense matrices whose dimensions are a product of the training set size and the vocabulary size).
 Our proposed approach combines CCA and CL-LSI.
-Another closely related method is Cross-Lingual Explicit Semantic Analysis (CL-ESA) \cite{ESA}, which uses Wikipedia (as do we in the current work) to compare documents.
+Another closely related method is Cross-Lingual Explicit Semantic Analysis (CL-ESA)~\cite{ESA}, which uses Wikipedia (as do we in the current work) to compare documents.
 It can be interpreted as using the sample covariance matrix between features of two languages to define the dot product which is used to compute similarities.
 The authors of CL-ESA compare it to CL-LSI and find that CL-LSI can outperform CL-ESA in an information retrieval, but is costlier to optimize over a large corpus (CL-ESA requires no training).
-We find that the scalability argument does not apply in our case: based on advances in numerical linear algebra we can solve large CL-LSI problems (millions of documents as opposed to 10,000 document limit reported in \cite{ESA}).
+We find that the scalability argument does not apply in our case: based on advances in numerical linear algebra we can solve large CL-LSI problems (millions of documents as opposed to the 10,000 document limit reported in~\cite{ESA}).
 In addition, CL-ESA is less suited for computing similarities between two large monolingual streams.
@@
 The intuition is that named entities (for example, ``Obama'') and cognate words (for example, ``tsunami'') are written in the same or similar fashion in many languages.
-For example, the Cross-Language Character n-Gram Model (CL-CNG) \cite{plagiarism} represents documents as bags of character n-grams.
-Another approach is to use language dependent keyword lists based on cognate words \cite{pouliquen2008story}.
+For example, the Cross-Language Character n-Gram Model (CL-CNG)~\cite{plagiarism} represents documents as bags of character $n$-grams.
+Another approach is to use language dependent keyword lists based on cognate words ~\cite{pouliquen2008story}.
 These approaches may be suitable for comparing documents written in languages that share a writing system, which does not apply to the case of global news tracking.
@@
 Many techniques use the similarity as a black box, e.g., a kernel in Support Vector Machines.
-Comparison of documents (or other types of text snippets) in a monolingual setting is a well-studied problem in the field of information retrieval \cite{Salton88term-weightingapproaches}.
+Comparison of documents (or other types of text snippets) in a monolingual setting is a well-studied problem in the field of information retrieval ~\cite{Salton88term-weightingapproaches}.
 We first formally introduce the problem followed by a description of  our approach.
@@
 \noindent\textbf{Document representation.}
-The standard vector space model \cite{Salton88term-weightingapproaches} represents documents as vectors, where each term corresponds to a word or a phrase in a fixed vocabulary.
+The standard vector space model~\cite{Salton88term-weightingapproaches} represents documents as vectors, where each term corresponds to a word or a phrase in a fixed vocabulary.
 Formally, document $d$ is represented by a vector $x \in \RR^n$, where $n$ corresponds to the size of the vocabulary, and vector elements $x_k$ correspond to the number of times term $k$ occurred in the document, also called \emph{term frequency} or $TF_k(d)$.
@@
 and build a vector space using the union of tokens over the languages.
-A cosine similarity function in such a space can be useful to some extent, for example "internet" or "Obama" may appear both in Spanish and English texts and the presence of such terms in both an English and a Spanish document would contribute to their similarity.
-In general however, large parts of vocabularies may not intersect, which means that given a language pair, many words in both languages cannot contribute to the similarity score, which makes the similarity function very insensitive to the data.
+A cosine similarity function in such a space can be useful to some extent, for example ``Internet'' or ``Obama'' may appear both in Spanish and English texts and the presence of such terms in both an English and a Spanish document would contribute to their similarity.
+In general however, large parts of vocabularies may not intersect.
+This means that given a language pair, many words in both languages cannot contribute to the similarity score.
+Such cases can make the similarity function very insensitive to the data.
 
@@
 We will provide an overview of the models in Section \ref{sec:models} and then introduce additional notation in \ref{sec:notation}.
-Starting with Section \ref{sec:kmeans} and ending with Section \ref{sec:hublang} we will describe some approaches to computing $S_{i,j}$ given training data.
+Starting with Section \ref{sec:kmeans} and ending with Section \ref{sec:hublang} we will describe some approaches to compute $S_{i,j}$ given training data.
 
 \subsection{Cross-Lingual Models}\label{sec:models}
-In this section, we will describe several approaches to the problem of computing multilingual similarities introduced in Section~\ref{sec:tfidf}. We present four approaches:
+In this section, we will describe several approaches to the problem of computing the multilingual similarities introduced in Section~\ref{sec:tfidf}. We present four approaches:
 a simple approach based on $k$-means clustering in Section~\ref{sec:kmeans}, a standard approach based on singular value decomposition in Section~\ref{sec:LSI}, a related
@@
 CCA can be used to find correlated patterns for a pair of languages, whereas the extended method optimizes a
-Sum of Squared Correlations (SSCOR) between several language pairs, which was introduced in \cite{Kettenring}.
+Sum of Squared Correlations (SSCOR) between several language pairs, which was introduced in~\cite{Kettenring}.
 The SSCOR problem is difficult to solve in our setting (hundreds of thousands of features, hundreds of thousands of examples).
@@
 A \emph{comparable corpus} is a collection of documents in multiple languages, with alignment between documents that are of the same topic, or even a rough translation of each other.
-Wikipedia is an example of a comparable corpus, where a specific entry can be described in multiple languages (e.g.
-``Berlin" is currently described in 222 languages).
+Wikipedia is an example of a comparable corpus, where a specific entry can be described in multiple languages (e.g., ``Berlin" is currently described in 222 languages).
 News articles represent another example, where the same event can be described by newspapers in several languages.
@@
 More formally, a \emph{multilingual document} $d = (u_1,\ldots u_m)$ is a tuple of $m$ documents on the same topic (comparable), where $u_i$ is the document written in language $i$.
-Note that individual document $u_i$ can be an empty document (missing resource) and each $d$ must contain at least \textbf{two nonempty documents}.
+Note that an individual document $u_i$ can be an empty document (missing resource) and each $d$ must contain at least \textbf{two nonempty documents}.
 This means that in our analysis we discard strictly monolingual documents for which no cross-lingual information is available.
@@
 Furthermore, let $X_i^{\ell}$ denote the $\ell$-th column of matrix $X_i$ and the matrices respect the document alignment - the vector $X_i^\ell$ corresponds to the TFIDF vector of the $i$-th component of multilingual document $d_\ell$.
-We use $N$ to denote the total row dimension of $X$, i.e.
-$N:= \sum_{i=1}^m n_i$.
-See the Figure~\ref{fig:stacked_matrices} for an illustration of the introduced notation.
+We use $N$ to denote the total row dimension of $X$, i.e., $N:= \sum_{i=1}^m n_i$.
+See Figure~\ref{fig:stacked_matrices} for an illustration of the introduced notation.
 
@@
 \centering
-\includegraphics[width=9cm]{stacked_matrices.pdf}
+\includegraphics[width=9cm]{stacked_matrices1-crop.pdf}
 \caption{\label{fig:stacked_matrices} Multilingual corpora and their matrix representations using the vector space model.}
@@
 
-We will now describe four models to cross-lingual similarity computation in the next sections.
+We will now describe four models to cross-lingual similarity computation in the next sub-sections.
 \subsection{$k$-means}\label{sec:kmeans}
@@
 The $k$-means algorithm is perhaps the most well-known and widely-used clustering algorithm. Here, we present its application
-to computing cross-lingual similarities.
+to compute cross-lingual similarities.
 The idea is based on concatenating the corpus matrices, running standard $k$-means clustering to obtain the matrix of centroids, ``reversing" the concatenation step to obtain a set of aligned bases, which are finally used to compute cross-lingual similarities.
-See the Figure~\ref{fig:kmeans} for an illustration of process of computing multilingual clusters and using them to change the coordinates to cluster centroid ``topic" space.
+See Figure~\ref{fig:kmeans} for overview of the procedure.
+The left side of Figure~\ref{fig:kmeans} illustrates the decomposition and the right side summarizes the coordinate change.
 
@@
 
-We then run the $k$-means algorithm \cite{kmeans} and obtain a centroid matrix $C \in \RR^{N \times k}$, where the $k$ columns represent centroid vectors.
-The centroid matrix can be split vertically into $m$ blocks: $$C = [C_1^T \cdots C_m^T]^T,$$ according to the number of dimensions of each language, i.e.
-$C_i \in \RR^{n_i \times k}$.
+We then run the $k$-means algorithm~\cite{kmeans} and obtain a centroid matrix $C \in \RR^{N \times k}$, where the $k$ columns represent centroid vectors.
+The centroid matrix can be split vertically into $m$ blocks: $$C = [C_1^T \cdots C_m^T]^T,$$ according to the number of dimensions of each language, i.e., $C_i \in \RR^{n_i \times k}$.
 %
@@
 
-The resulting matrix (when appropriately scaled to have unit norm) for similarity computation between language $i$ and language $j$ is defined as (see the reasoning in the next paragraph):
+The resulting matrix for similarity computation between language $i$ and language $j$ is defined up to a scaling factor as:
 $$C_i(C_i^T C_i)^{-1} (C_j^T C_j)^{-1} C_j.$$
@@
 The matrix is a result of mapping documents in a language independent space using pseudo-inverses of the centroid matrices $P_i = (C_i^T C_i)^{-1} C_i$ and then comparing them using the standard inner product, which results in the matrix $P_i^T P_j$.
-For the sake of presentation, we assumed that the centroid vectors are linearly independent (an independent subspace could be obtained using an additional Gram-Schmidt step~\cite{golub} on the matrix $C$, if this was not the case).
-
-In the next section we will describe our second approach.
+For the sake of presentation, we assumed that the centroid vectors are linearly independent.
+(An independent subspace could be obtained using an additional Gram-Schmidt step~\cite{golub} on the matrix $C$, if this was not the case.)
+
 \subsection{Cross-Lingual Latent Semantic Indexing}\label{sec:LSI}
 
-The second approach we consider is Cross-Lingual Latent Semantic Indexing (CL-LSI)\cite{cl_lsi} which is a variant of LSI~\shortcite{lsi} for more than one language.
+The second approach we consider is Cross-Lingual Latent Semantic Indexing (CL-LSI)~\cite{cl_lsi} which is a variant of LSI~\shortcite{lsi} for more than one language.
 The approach is very similar to $k$-means, where we first concatenate the corpus matrices, compute a decomposition, which in case of CL-LSI is a truncated Singular Value Decomposition (SVD), decouple the
@@
  The method is based on computing a truncated singular value decomposition of the concatenated corpus matrix $X \approx U S V^T$.
-See the Figure~\ref{fig:lsi} for the decomposition.
-Representing documents in ``topic`` coordinates is done in the same way as in the $k$-means case (see Figure~\ref{fig:kmeans}), which we will
+See Figure~\ref{fig:lsi} for the decomposition.
+Representing documents in ``topic`` coordinates is done in the same way as in the $k$-means case (see Figure~\ref{fig:kmeans}), we will describe how to compute the coordinate change functions.
 
@@
 
-\paragraph{A note on implementation}
+\paragraph{Implementation note}
 
@@
 Moreover, the Lanczos method is hard to parallelize.
-Instead, we use a randomized version of the SVD \shortcite{tropp} that can be viewed as a block Lanczos method.
+Instead, we use a randomized version of the SVD~\shortcite{tropp} that can be viewed as a block Lanczos method.
 That enables us to use parallelization and speeds up the computation considerably.
 
-To compute the matrices $P_i$ we used the QR algorithm~\cite{golub} to factorize $U_i$ as $U_i = Q_i R_i$, where $Q_i^TQ_i = I$ and $R_i$ is triangular matrix.
+To compute the matrices $P_i$ we used the QR algorithm~\cite{golub} to factorize $U_i$ as $U_i = Q_i R_i$, where $Q_i^TQ_i = I$ and $R_i$ is a triangular matrix.
 $P_i$ is then obtained by solving $R_i P_i = Q_i$.
 
-In the next section we will introduce our third approach.
 \subsection{Canonical Correlation Analysis}\label{sec:CCA}
- We now present a statistical technique to analyze data from two sources, extension of which will be presented in the next section.
+ We now present a statistical technique to analyze data from two sources, an extension of which will be presented in the next section.
 %
@@
 The optimization problem can be reduced to an eigenvalue problem and includes inverting the variance matrices $C_{i,i}$ and $C_{j,j}$.
-If the matrices are not invertible, one can use a regularization technique by replacing $C_{i,i}$ with $(1- \kappa)C_{i,i} + \kappa I$, where $\kappa \in [0,1]$ is the regularization coefficient and $I$ is the identity matrix (the same can be applied to $C_{j,j}$.
+If the matrices are not invertible, one can use a regularization technique by replacing $C_{i,i}$ with $(1- \kappa)C_{i,i} + \kappa I$, where $\kappa \in [0,1]$ is the regularization coefficient and $I$ is the identity matrix.
+(The same can be applied to $C_{j,j}$.)
 A single canonical variable is usually inadequate in representing the original random vector and typically one looks for $k$ projection pairs $(w_i^1, w_j^1),\ldots,(w_i^k, w_j^k)$, so that $(w_i^{u})^T \mathcal{X}_i$ and $(w_j^{u})^T \mathcal{X}_j$ are highly correlated and $(w_i^{u})^T \mathcal{X}_i$ is uncorrelated with $(w_i^{v})^T \mathcal{X}_i$  for $u \neq v$ and analogously for $w_j^u$ vectors.
@@
 Note that the method in its original form is only applicable to two languages where an aligned set of observations is available.
-The next section will describe a scalable an extension of CCA to more than two languages.
+The next section will describe a scalable extension of CCA to more than two languages.
 
@@
 Even if only two languages are considered, the set of aligned documents can be small (an extreme example is given by the Piedmontese and Hindi Wikipedias where no inter-language links are available), in which case none of the methods presented so far are applicable.
- The second challenge is scale - the data is high-dimensional (hundreds of thousands of features per language, many languages) and the number of multilingual documents may be large (over one million in case of Wikipedia).
-The optimization problem related to CCA is not
+ The second challenge is scale - the data is high-dimensional (many languages with hundreds of thousands of features per language) and the number of multilingual documents may be large (over one million in case of Wikipedia).
+The optimization problem posed by CCA is not
  trivial to solve: the covariance matrices themselves are prohibitively large to fit in memory (even storing a 100,000 by 100,000 element matrix requires 80GB of memory) and iterative matrix-multiplication based approaches to  solving generalized eigenvalue problems are required (the covariance matrices can be expressed as products of sparse matrices, which means we have fast matrix-vector multiplication).
@@
 We now describe an extension of CCA to more than two languages, which can be trained on large comparable corpora and can handle missing data.
- The extension we consider is based on a generalization of CCA to more than two views, introduced in \cite{Kettenring}, namely the Sum of Squared Correlations SSCOR, which we will state formally later in this section.
-Our approach exploits a certain characteristic of the data, namely the \emph{hub language} characteristic (introduced in the next paragraph) in two
+ The extension we consider is based on a generalization of CCA to more than two views, introduced in~\cite{Kettenring}, namely the Sum of Squared Correlations SSCOR, which we will state formally later in this section.
+Our approach exploits a certain characteristic of the data, namely the \emph{hub language} characteristic (see below) in two
  ways: to reduce the dimensionality of the data and to simplify the optimization problem.
@@
 The second step involves solving a generalized version of canonical correlation analysis on the matrices $Y_i$ in order to find the mappings $P_i$.
-The approach is based on the sum of squares of correlations formulation by Kettenring \cite{Kettenring}, where we consider only correlations between pairs $(Y_1, Y_i), i >1$ due to the hub language problem characteristic.
-We will present the original unconstrained optimization problem, then a constrained formulation based on the hub language problem characteristic. Then we will simplify the constraints and reformulate
-the problem as an eigenvalue problem by using the method of Lagrange multipliers.
-
-The original sum of squared correlation is formulated as an unconstrained problem:
+The approach is based on the sum of squares of correlations formulation by Kettenring~\cite{Kettenring}, where we consider only correlations between pairs $(Y_1, Y_i), i >1$ due to the hub language problem characteristic.
+We will present the original unconstrained optimization problem, then a constrained formulation based on the hub language problem characteristic.
+Then we will simplify the constraints and reformulate the problem as an eigenvalue problem by using Lagrange multipliers.
+
+The original sum of squared correlations is formulated as an unconstrained problem:
 \begin{equation*}
@@
 
-\paragraph{Remark.} The technique is related to  Generalization of Canonical Correlation Analysis (GCCA) by Carroll \citeyear{Carroll}, where an unknown group configuration variable is defined and the objective is to maximize the sum of squared correlation between the group variable and the others.
+\paragraph{Remark.} The technique is related to  Generalization of Canonical Correlation Analysis (GCCA) by Carroll~\citeyear{Carroll}, where an unknown group configuration variable is defined and the objective is to maximize the sum of squared correlations between the group variable and the others.
 The problem can be reformulated as an eigenvalue problem.
 The difference lies in the fact that we set the unknown group configuration variable as the hub language, which simplifies the solution.
-The complexity of our method is $O(k^3)$, where $k$ is the reduced dimension from the LSI preprocessing step, whereas solving the GCCA method scales as $O(s^3)$, where $s$ is the number of samples (see \cite{gifi}).
+The complexity of our method is $O(k^3)$, where $k$ is the reduced dimension from the LSI preprocessing step, whereas solving the GCCA method scales as $O(s^3)$, where $s$ is the number of samples (see~\cite{gifi}).
 Another issue with GCCA is that it cannot be directly applied to the case of missing documents.
@@
 Matching of clusters is a \emph{generalized matching} problem.
-We cannot assume that there is only one cluster per language per event, nor can we assume complete coverage -- i.e.
-that there exists at least one cluster per event in every language.
+We cannot assume that there is only one cluster per language per event, nor can we assume complete coverage -- i.e., that there exists at least one cluster per event in every language.
 This is partly due to news coverage which might be more granular in some languages, partly due to noise and errors in the event detection process.
@@
 For each candidate cluster $c_j \in C$, we compute a vector of learning features that should be indicative of whether $c_i$ and $c_j$ are equivalent or not and apply a binary classification model that predicts if the clusters are equivalent or not.
-The classification algorithm that we used to train a model was a linear Support Vector Machine (SVM) method \shortcite{shawe-taylor04kernel}.
+The classification algorithm that we used to train a model was a linear Support Vector Machine (SVM) method~\shortcite{shawe-taylor04kernel}.
 
@@
 \item \texttt{linkCount} is the number of times any of the news articles from $c_j$ is among $10$-nearest neighbors for articles from $c_i$.
-In other words, it is the number of times an article from $c_i$ has a very similar article (i.e.
-is among 10 most similar) in $c_j$.
+In other words, it is the number of times an article from $c_i$ has a very similar article (i.e., is among 10 most similar) in $c_j$.
 \item \texttt{avgSimScore} is the average similarity score of the links, as identified for \texttt{linkCount}, between the two clusters.
@@
 Articles that are imported into Event Registry are annotated by disambiguating mentioned \emph{entities} and \emph{keywords} to the corresponding Wikipedia pages~\cite{zhang2014}.
-Whenever Barack Obama is, for example, mentioned in the article, the article is annotated with a link to his Wikipedia page\footnote{\emph{http://en.wikipedia.org/wiki/Barack\_Obama}}.
-In the same way all mentions of entities (people, locations, organizations) and ordinary keywords (e.g.
-bank, tax, ebola, plane, company) are annotated.
+Whenever Barack Obama is, for example, mentioned in the article, the article is annotated with a link to his Wikipedia page.
+In the same way, all mentions of entities (people, locations, organizations) and ordinary keywords (e.g., bank, tax, ebola, plane, company) are annotated.
 Although the Spanish article about Obama will be annotated with his Spanish version of the Wikipedia page, in many cases we can link the Wikipedia pages to their English versions.
@@
 This  results in a large number of weakly comparable documents in more than $200$ languages.
-Wikipedia is a large source of multilingual data that is especially important for the languages for which no translation tools, multilingual dictionaries as Eurovoc \cite{eurovoc}, or strongly aligned multilingual corpora as Europarl \cite{europarl} are available.
+Wikipedia is a large source of multilingual data that is especially important for the languages for which no translation tools, multilingual dictionaries as Eurovoc~\cite{eurovoc}, or strongly aligned multilingual corpora as Europarl~\cite{europarl} are available.
 Documents in different languages are related with so called 'inter-language' links that can be found on the left of the Wikipedia page.
@@
 It turns out that we obtain the matrix $M$ that is not symmetric, consequently the underlying graph is not symmetric.
-That means that existence of the inter-language link in one way (i.e.
-English to German) does not guarantee that there is an inter-language link in the reverse direction (German to English).
+That means that existence of the inter-language link in one way (i.e., English to German) does not guarantee that there is an inter-language link in the reverse direction (German to English).
 To correct this we transform this matrix to be symmetric by computing $M+M^T$ and obtaining an undirected graph.
 In the rare case that after symmetrization we have multiple links pointing from the document, we pick the first one that we encountered.
-This matrix enables us to build an alignment across all Wikipedia \footnote{https://www.wikipedia.org/ dumps available in 2013} languages.
+This matrix enables us to build an alignment across all Wikipedia\footnote{https://www.wikipedia.org/ dumps available in 2013} languages.
 
@@
 We will demonstrate that this approach can be successfully applied even in the case of fully missing alignment information.
- To this purpose, we select a subset of Wikipedia languages containing three major languages, English (4,212k articles)--\emph{en} (hub language), Spanish (9,686k articles)--\emph{es}, Russian (9,662k articles)--\emph{ru}, and five minority (in the sense of Wikipedia sizes) languages, Slovenian (136k articles)--\emph{sl}, Piedmontese (59k articles)--\emph{pms}, Waray-Waray (112k articles)--\emph{war} (all with about 2 million native speakers), Creole (54k articles)--\emph{ht} (8 million native speakers), and Hindi (97k articles)--\emph{hi} (180 million native speakers).
+ To this purpose, we select a subset of Wikipedia languages containing three major languages, English (4,212k articles)--\emph{en} (hub language), Spanish (9,686k articles)--\emph{es}, Russian (9,662k articles)--\emph{ru}, and five minority (in terms of Wikipedia sizes) languages, Slovenian (136k articles)--\emph{sl}, Piedmontese (59k articles)--\emph{pms}, Waray-Waray (112k articles)--\emph{war} (all with about 2 million native speakers), Creole (54k articles)--\emph{ht} (8 million native speakers), and Hindi (97k articles)--\emph{hi} (180 million native speakers).
 For preprocessing, we remove the documents that contain less than 20 different words (referred to as stubs\footnote{Such documents are typically of low value as a linguistic resource.
 Examples include the titles of the columns in the table, remains of the parsing process, or Wikipedia articles with very little or no information contained in one or two sentences.}) and remove words occurring in less than 50 documents as well as the top 100 most frequent words (in each language separately).
-We represent the documents as normalized TFIDF\cite{Salton88term-weightingapproaches} weighted vectors.
-The IDF scores are computed for each language with the exception of English based on its aligned documents with the English Wikipedia.
+We represent the documents as normalized TFIDF~\cite{Salton88term-weightingapproaches} weighted vectors.
+The IDF scores are computed for each language based on its aligned documents with the English Wikipedia.
 The English language IDF scores are based on all English documents for which aligned Spanish documents exist.
@@
 We observe that the method performs well on all pairs of languages, where at least 50,000 training documents are available(\emph{en}, \emph{es}, \emph{ru}, \emph{sl}).
-We note that taking $k = 500$ or $k = 1,000$ multilingual topics usually results in similar performance, with some notable exceptions: in the case of (\emph{ht}, \emph{war}) the additional topics result in an increase in performance, as opposed to (\emph{ht},\emph{pms}) where performance drops, which suggests overfitting.
+We note that taking $k = 500$ or $k = 1,000$ multilingual topics usually results in similar performance, with some notable exceptions: in the case of (\emph{ht}, \emph{war}) the additional topics result in an increase in performance, as opposed to (\emph{ht}, \emph{pms}) where performance drops, which suggests overfitting.
 The languages where the method performs poorly are \emph{ht} and \emph{war}, which can be explained by the quality of data (see Table \ref{table:rank} and explanation that follows).
@@
 Also observe that in the case of (\emph{pms}, \emph{ht}) the method still obtains a score of 62, even though training set intersection is zero and \emph{ht} data is corrupted, which we will show in the next paragraph.
+{
+\renewcommand\tabcolsep{3pt}
+\begin{table}[h!]
+\centering
+\caption{Training -- test sizes (in thousands).
+The first row represents the size of the training sets used to construct the mappings in low-dimensional language independent space using \emph{en} as a hub.
+The diagonal elements represent the number of the unique training documents and test documents in each language.
+}
+\label{table:train_test}
+{
+\small
+\begin{tabular}{c|c|c|c|c|c|c|c|c|}
+&	en&	es&	ru&	sl&	hi&	war&	ht&	pms\\\cline{1-9}
+en&	671~-~4.64&	463~-~4.29&	369~-~3.19&	50.3~-~2&	14.4~-~2.76&	8.58~-~2.41&	 17~-~2.32&	16.6~-~2.67\\
+\cline{2-9}
+es&	\multicolumn{1}{c|}{}	&	463~-~4.29&	187~-~2.94&	28.2~-~1.96&	8.72~-~2.48&	 6.88~-~2.4&	13.2~-~2&	 13.8~-~2.58\\
+\cline{3-9}
+ru&	\multicolumn{2}{c|}{}	&	369~-~3.19&	29.6~-~1.92&	9.16~-~2.68&	2.92~-~1.1&	 3.23~-~2.2&	10.2~-~1.29\\
+\cline{4-9}
+sl&	\multicolumn{3}{c|}{}	&	50.3~-~2&	3.83~-~1.65&	1.23~-~0.986&	0.949~-~1.23&	 1.85~-~0.988\\
+\cline{5-9}
+hi&	\multicolumn{4}{c|}{}	&	14.4~-~2.76&	0.579~-~0.76&	0.0~-~2.08&	0.0~-~0.796\\
+\cline{6-9}
+war&	\multicolumn{5}{c|}{}	&	8.58~-~2.41&	0.043~-~0.534&	0.0~-~1.97\\
+\cline{7-9}
+ht&	\multicolumn{6}{c|}{}	&	17~-~2.32&	0.0~-~0.355\\
+\cline{8-9}
+pms&	\multicolumn{7}{c|}{}	&	16.6~-~2.67\\
+\cline{9-9}
+\end{tabular}
+}
+\end{table}
+}
+
 {
@@
 
-{
-\renewcommand\tabcolsep{3pt}
-\begin{table}[h!]
-\centering
-\caption{Training -- test sizes (in thousands).
-The first row represents the size of the training sets used to construct the mappings in low-dimensional language independent space using the English--\emph{en} as a hub.
-The diagonal elements represent the number of the unique training documents and test documents in each language.
-}
-\label{table:train_test}
-{
-\small
-\begin{tabular}{c|c|c|c|c|c|c|c|c|}
-&	en&	es&	ru&	sl&	hi&	war&	ht&	pms\\\cline{1-9}
-en&	671~-~4.64&	463~-~4.29&	369~-~3.19&	50.3~-~2&	14.4~-~2.76&	8.58~-~2.41&	 17~-~2.32&	16.6~-~2.67\\
-\cline{2-9}
-es&	\multicolumn{1}{c|}{}	&	463~-~4.29&	187~-~2.94&	28.2~-~1.96&	8.72~-~2.48&	 6.88~-~2.4&	13.2~-~2&	 13.8~-~2.58\\
-\cline{3-9}
-ru&	\multicolumn{2}{c|}{}	&	369~-~3.19&	29.6~-~1.92&	9.16~-~2.68&	2.92~-~1.1&	 3.23~-~2.2&	10.2~-~1.29\\
-\cline{4-9}
-sl&	\multicolumn{3}{c|}{}	&	50.3~-~2&	3.83~-~1.65&	1.23~-~0.986&	0.949~-~1.23&	 1.85~-~0.988\\
-\cline{5-9}
-hi&	\multicolumn{4}{c|}{}	&	14.4~-~2.76&	0.579~-~0.76&	0.0~-~2.08&	0.0~-~0.796\\
-\cline{6-9}
-war&	\multicolumn{5}{c|}{}	&	8.58~-~2.41&	0.043~-~0.534&	0.0~-~1.97\\
-\cline{7-9}
-ht&	\multicolumn{6}{c|}{}	&	17~-~2.32&	0.0~-~0.355\\
-\cline{8-9}
-pms&	\multicolumn{7}{c|}{}	&	16.6~-~2.67\\
-\cline{9-9}
-\end{tabular}
-}
-\end{table}
-}
-
-We further inspect the properties of the training sets by roughly estimating the fraction \texttt{rank(A)/min(rows(A), cols(A))} for each English training matrix and its corresponding mate matrix, where $rows(A)$ and $cols(A)$ denote the number of rows and columns respectively.
+We further inspect the properties of the training sets by roughly estimating the fraction $\frac{rank(A)}{min\left(rows\left(A\right),~cols\left(A\right)\right)}$ for each English training matrix and its corresponding mate matrix, where $rows(A)$ and $cols(A)$ denote the number of rows and columns respectively.
 The denominator represents the theoretically highest possible rank the matrix $A$ could have.
@@
 \cline{1-7}
-en -- de     &   en -- ru     &   en -- sl       &     en -- hi &   en -- war      &      en -- ht &   en -- pms\\
+en -- es     &   en -- ru     &   en -- sl       &     en -- hi &   en -- war      &      en -- ht &   en -- pms\\
 \cline{1-7}
@@
 \subsection{Evaluation Of Cross-Lingual Event Linking}
-In order to determine how accurately we can predict cluster equivalence, we performed two experiments in a multilingual setting using the English, German and Spanish language for which we had labelled data to evaluate the linking performance.
+In order to determine how accurately we can predict cluster equivalence, we performed two experiments in a multilingual setting using English, German and Spanish languages for which we had labelled data to evaluate the linking performance.
 In the first experiment, we tested how well  the individual approaches for cross-lingual article linking perform when used for linking the clusters about the same event.
@@
 We created a manually labelled dataset in order to evaluate cross-lingual event linking using two human annotators.
-The annotators were provided with an interface listing the articles, their content from and top concepts for a pair of clusters and their task was to determine if the clusters were equivalent or not (i.e.
-discuss same event).
-To obtain a pair of clusters $(c_i, c_j)$ to annotate, we first randomly chose a cluster $c_i$, used the Algorithm~\ref{cluster_merge_algo1} to compute a set of potentially equivalent clusters $C$ and randomly chose a cluster $c_j \in C$.
+The annotators were provided with an interface listing the articles, their content from and top concepts for a pair of clusters and their task was to determine if the clusters were equivalent or not (i.e., discuss same event).
+To obtain a pair of clusters $(c_i, c_j)$ to annotate, we first randomly chose a cluster $c_i$, used Algorithm~\ref{cluster_merge_algo1} to compute a set of potentially equivalent clusters $C$ and randomly chose a cluster $c_j \in C$.
 The dataset provided by the annotators contains 808 examples, of which 402 are equivalent clusters pairs and 406 are not.
@@
 
-We can see that the most informative features for cluster linking are obtained using the hub CCA algorithm, while $k$-means performs the worst.
+We observe that, for the task of cluster linking, LSI and hub CCA perform comparably and both outperform $k$-means.
 
@@
 We also compared the proposed approaches on the task of Wikipedia mate retrieval (the same task as in Section~\ref{experiments:hubcca}).
-We computed the Average (over language pairs) Mean Reciprocal Rank (AMRR) \shortcite{voorhees1999trec}  performance of the different approaches on the  Wikipedia data by holding out $15,000$ aligned test documents and using $300,000$ aligned documents as the training set.
-The Figure~\ref{pic:AMRR} shows AMRR score as the function of the number of feature vectors.
+We computed the Average (over language pairs) Mean Reciprocal Rank (AMRR)~\shortcite{voorhees1999trec}  performance of the different approaches on the  Wikipedia data by holding out $15,000$ aligned test documents and using $300,000$ aligned documents as the training set.
+Figure~\ref{pic:AMRR} shows AMRR score as the function of the number of feature vectors.
 It is clear that hub CCA outperforms LSI approach and $k$-means lags far behind when testing on Wikipedia data.
@@
 \begin{table}[h]
-\caption{Accuracy of cluster linking with 500 - 800 - 1,000 topic vectors obtained from different cross-lingual similarity algorithms.
+\caption{Accuracy of cluster linking with 500/800/1,000 topic vectors obtained from different cross-lingual similarity algorithms.
 The table shows for each of the algorithms the obtained classification accuracy, precision and recall.}
@@
 \begin{center}
-\begin{tabular}{|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|c|}
   \hline
-  \cline{1-4}
-  Models & Accuracy & Precision & Recall \\ \cline{1-4}
-  hub CCA      & 78.2 - 79.6 - 80.3 \% & 76.3 - 78.0 - 80.5 \%  & 81.6 - 82.1 - 79.9 \%  \\ \cline{1-4}
-  LSI      & 78.9 - 78.7 - 80.6 \%  & 76.8 - 77.0 - 78.7 \%  & 83.3 - 80.6 - 83.6 \%  \\ \cline{1-4}
- $k$-means & 73.9\phantom{ - 78.7 - 80.6 }\% & 69.5\phantom{ - 78.7 - 80.6 }\%  & 84.6\phantom{ - 78.7 - 80.6 }\%  \\ \cline{1-4}
+  \cline{1-5}
+  Models & Accuracy \% & Precision \% & Recall \% & $F_1$ \% \\ \cline{1-5}
+  hub CCA  & 78.2/79.6/80.3 & 76.3/78.0/80.5  & 81.6/82.1/79.9 & 78.9/80.0/80.2
+  \\ \cline{1-5}
+  LSI      & 78.9/78.7/80.6  & 76.8/77.0/78.7 & 83.3/80.6/83.6 & 79.9/78.8/81.1  \\ \cline{1-5}
+ $k$-means & 73.9/-/- & 69.5/-/- & 84.6/-/- &  76.3/-/- \\ \cline{1-5}
+ %$k$-means & 73.9/-/-\phantom{/78.7/80.6 } & 69.5/-/-\phantom{/78.7/80.6 }  & 84.6/-/-\phantom{/78.7/80.6 } &  76.3\phantom{/78.7/80.6 }  \\ \cline{1-5}
 \end{tabular}
@@
 \begin{center}
-\begin{tabular}{|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|c|}
   \hline
-  \cline{1-4}
-  Models & Accuracy & Precision & Recall \\ \cline{1-4}
-  hub CCA      & 81.2 - 77.8 \% & 80.5 - 74.5 \%  & 91.3 - 57.5 \%  \\ \cline{1-4}
-  LSI      & 82.8 - 76.4 \% & 81.3 - 70.9 \%  & 93.1 - 57.5 \%  \\ \cline{1-4}
- $k$-means & 75.5 - 71.2 \% & 72.8 - 70.8 \%  & 95.3 - 36.2 \%  \\ \cline{1-4}
+  \cline{1-5}
+  Models & Accuracy \% & Precision \% & Recall \% & $F_1$ \% \\ \cline{1-5}
+  hub CCA  & 81.2 - 77.8 & 80.5 - 74.5 & 91.3 - 57.5 & 85.6 - 64.9 \\ \cline{1-5}
+  LSI      & 82.8 - 76.4 & 81.3 - 70.9 & 93.1 - 57.5 & 86.8 - 63.5 \\ \cline{1-5}
+ $k$-means & 75.5 - 71.2 & 72.8 - 70.8 & 95.3 - 36.2 & 82.5 - 47.9 \\ \cline{1-5}
 \end{tabular}
@@
 In the second experiment, we evaluate how relevant individual groups of features are to correctly determine cluster equivalence.
-For this purpose, we tested the accuracies obtained using individual groups of features, as well as using different combination of groups.
+For this purpose, we computed accuracy using individual groups of features, as well as using different combination of groups.
 Since hub CCA had the best  performance of the three algorithms, we used it to compute the values of the cross-lingual article linking features.
@@
   \cline{1-5}
-  Features & A & Precision & Recall & F1 \\ \cline{1-5}
-  hub CCA & $78.3\% \pm 5.9\%$ & $0.782 \pm 0.070$ & $0.789 \pm 0.052$ & $0.784 \pm 0.055$ \\ \cline{1-5}
-  Concepts & $88.5\% \pm 2.7\%$ & $0.886 \pm 0.048$ & $0.886 \pm 0.022$ & $0.885 \pm 0.024$ \\ \cline{1-5}
-  Misc & $54.8\% \pm 6.7\%$ & $0.618 \pm 0.165$ & $0.582 \pm 0.302$ & $0.524 \pm 0.130$ \\ \cline{1-5}
-  hub CCA + Concepts & $89.4\% \pm 2.5\%$ & $0.894 \pm 0.046$ & $0.896 \pm 0.024$ & $0.894 \pm 0.023$ \\ \cline{1-5}
-  hub CCA + Misc & $78.8\% \pm 5.0\%$ & $0.789 \pm 0.071$ & $0.794 \pm 0.046$ & $0.790 \pm 0.045$ \\ \cline{1-5}
-  Concepts + Misc & $88.7\% \pm 2.6\%$ & $0.888 \pm 0.046$ & $0.888 \pm 0.022$ & $0.887 \pm 0.023$ \\ \cline{1-5}
-  All & $89.2\% \pm 2.6\%$ & $0.888 \pm 0.049$ & $0.901 \pm 0.019$ & $0.893 \pm 0.023$ \\ \cline{1-5}
+  Features & Accuracy \% & Precision \% & Recall \% & $F_1$ \%  \\ \cline{1-5}
+  hub CCA            & $78.3 \pm 5.9$ & $78.2 \pm  7.0$ & $78.9 \pm  5.2$ & $78.4 \pm  5.5$ \\ \cline{1-5}
+  Concepts           & $88.5 \pm 2.7$ & $88.6 \pm  4.8$ & $88.6 \pm  2.2$ & $88.5 \pm  2.4$ \\ \cline{1-5}
+  Misc               & $54.8 \pm 6.7$ & $61.8 \pm 16.5$ & $58.2 \pm 30.2$ & $52.4 \pm 13.0$ \\ \cline{1-5}
+  hub CCA + Concepts & $89.4 \pm 2.5$ & $89.4 \pm  4.6$ & $89.6 \pm  2.4$ & $89.4 \pm  2.3$ \\ \cline{1-5}
+  hub CCA + Misc     & $78.8 \pm 5.0$ & $78.9 \pm  7.1$ & $79.4 \pm  4.6$ & $79.0 \pm  4.5$ \\ \cline{1-5}
+  Concepts + Misc    & $88.7 \pm 2.6$ & $88.8 \pm  4.6$ & $88.8 \pm  2.2$ & $88.7 \pm  2.3$ \\ \cline{1-5}
+  All                & $89.2 \pm 2.6$ & $88.8 \pm  4.9$ & $90.1 \pm  1.9$ & $89.3 \pm  2.3$ \\ \cline{1-5}
   \hline
@@
   \cline{1-5}
-  Language pair & A & Precision & Recall & F1 \\ \cline{1-5}
-  en, de & $91.8\% \pm 5.5\%$ & $0.917 \pm 0.063$ & $0.937 \pm 0.063$ & $0.925 \pm 0.051$ \\ \cline{1-5}
-  en, es & $87.7\% \pm 5.4\%$ & $0.877 \pm 0.074$ & $0.885 \pm 0.098$ & $0.876 \pm 0.059$ \\ \cline{1-5}
-  es, de & $88.6\% \pm 4.3\%$ & $0.897 \pm 0.091$ & $0.843 \pm 0.119$ & $0.859 \pm 0.060$ \\ \cline{1-5}
+  Language pair & Accuracy \% & Precision \% & Recall \% & $F_1$ \% \\ \cline{1-5}
+  en, de & $91.8 \pm 5.5$ & $91.7 \pm  6.3$ & $93.7 \pm  6.3$ & $92.5 \pm  5.1$ \\ \cline{1-5}
+  en, es & $87.7 \pm 5.4$ & $87.7 \pm  7.4$ & $88.5 \pm  9.8$ & $87.6 \pm  5.9$ \\ \cline{1-5}
+  es, de & $88.6 \pm 4.3$ & $89.7 \pm  9.1$ & $84.3 \pm 11.9$ & $85.9 \pm  6.0$ \\ \cline{1-5}
   \hline
@@
 
-Currently, we compute similarities over $24$ languages with tags: eng, spa, deu, zho, ita, fra, rus, swe, nld, tur, jpn, por, ara, fin, ron, kor, hrv, tam, hun, slv, pol, srp, cat, ukr but we support any language from the top $100$ Wikipedia languages.
-Our data stream is Newsfeed (http://newsfeed.ijs.si/) which provides $430k$ unique articles per day.
+Currently, we compute similarities over $24$ languages with tags: \emph{eng}, \emph{spa}, \emph{deu}, \emph{zho}, \emph{ita}, \emph{fra}, \emph{rus}, \emph{swe}, \emph{nld}, \emph{tur}, \emph{jpn}, \emph{por}, \emph{ara}, \emph{fin}, \emph{ron}, \emph{kor}, \emph{hrv}, \emph{tam}, \emph{hun}, \emph{slv}, \emph{pol}, \emph{srp}, \emph{cat}, \emph{ukr} but we support any language from the top $100$ Wikipedia languages.
+Our data stream is Newsfeed (http://newsfeed.ijs.si/) which provides 430k unique articles per day.
 Our system currently computes 2 million similarities per second, that means that we compute $16 \cdot 10^{10}$ similarities per day.
@@
 
-\subsection{Remarks on the repeatability of experiments}
+\subsection{Remarks on the reproducibility of experiments}
 We have made  both the code and data that were used in the experiments publicly available at \url{https://github.com/rupnikj/jair_paper.git}.
@@
 In this paper we have presented a cross-lingual system for linking events in different languages.
-Building on an existing system, Event Registry, we present and evaluate several approaches to computing a  cross-lingual similarity function.
-We also present an approach to linking events and evaluate effectiveness of various features.
+Building on an existing system, Event Registry, we present and evaluate several approaches to compute a  cross-lingual similarity function.
+We also present an approach to link events and evaluate effectiveness of various features.
 The  final pipeline is scalable both in terms of number of articles and number of languages, while accurately linking events.
@@
 The experiments show that the hub CCA-based features present a good baseline, which can greatly benefit from additional semantic-based features.
-Even though the CCA-based features do not greatly improve the system performance when the semantic features are used, there are two main benefits in the approach: the linking process can be sped up (smaller set of candidate clusters), and usage in the case when semantic extraction is not possible.
+Even though in our experiments the addition of CCA-based features to semantic features did not lead to great performance improvements, there are two important benefits in the approach.
+First, the linking process can be sped up by using a smaller set of candidate clusters.
+Second, the approach is robust to languages where semantic extraction is not available, due to scarce linguistic resources.
 

### 2015-11-17 rupnik2015.tex
cleaned up the bibliography

@@
 The results are shown in Table~\ref{table:langPairEval}.
-We can see that the highest accuracy can be obtained when we need to decide if an English and German cluster represent the same event.
-A possible interpretation for these results is that this language pair has the most language resources (the highest number of Wikipedia pages) which results in the most accurate cross-lingual article similarity measure as well as the most extensive semantic annotation of the articles.
+We can see that the performance of cluster linking on the English-German dataset is the highest in terms of accuracy, precision, recall and $F_1$.
+The performance on the English-Spanish dataset is comparable to the performance on the English-German dataset, where the former achieves higher recall (and slightly higher $F_1$ score), while the latter achieves higher precision.
+A possible explanation of these results is that the higher quantity and quality of English-German language resources leads to a more accurate cross-lingual article similarity measure as well as to a more extensive semantic annotation of the articles.
 

### 2015-12-30 rupnik2015.tex
acknowledgments added
ack revised

@@
 \documentclass[twoside,11pt]{article}
-\usepackage{jair, theapa, rawfonts}
+\usepackage{jair, theapa, rawfonts, url}
 \usepackage{pgfplots}
@@
 
-%\acks{The authors wish to thank blah blah blah.}
+\acks{The authors gratefully acknowledge that the funding for this work was provided by the projects X-LIKE (ICT-257790-STREP), MultilingualWeb (PSP-2009.5.2 Agr.\# 250500), TransLectures (FP7-ICT-2011-7), PlanetData (ICT-257641-NoE), RENDER (ICT-257790-STREP), XLime (FP7-ICT-611346), and META-NET (ICT-249119-NoE).}
 

### 2016-01-06 rupnik2015.tex
fixed bib and style changes

@@
 
-\jairheading{x}{xxxx}{x-xx}{x/xx}{x/xx}
+\jairheading{55}{2016}{x-xx}{3/24}{1/16}
 \ShortHeadings{Cross-Lingual Document Similarity and Event Tracking}
@@
 In this work, we address the problem of tracking of events in a large multilingual stream.
-Within a recently developed system Event Registry~\shortcite{Leban2014W,Leban2014I} we examine two aspects of this problem: how to compare articles in different languages and how to link collections of articles in different languages which refer to the same event.
+Within a recently developed system Event Registry we examine two aspects of this problem: how to compare articles in different languages and how to link collections of articles in different languages which refer to the same event.
 Taking a multilingual stream and clusters of articles from each language, we compare different cross-lingual document similarity measures based on Wikipedia.
@@
 Content on the Internet is becoming increasingly multilingual.
-A prime example is Wikipedia.
+A prime example is Wiki-pedia.
 In 2001, the majority of pages were written in English, while in 2015, the percentage of English articles has dropped to 14\%.
@@
 The approach models link probability and combines prior word sense distributions with context based sense distributions.
-The details are reported in~\cite{Milne:2008:LLW:1458082.1458150} and ~\cite{zhang2014saaacamactat}.
+The details are reported by
+~\citeauthor{Milne:2008:LLW:1458082.1458150}~\citeyear{Milne:2008:LLW:1458082.1458150} and~\citeauthor{zhang2014saaacamactat}~\citeyear{zhang2014saaacamactat}.
 As a part of the semantic annotation we also analyze the dateline of the article to identify the location of the described event as well as to extract dates mentioned in the article using a set of regular expressions.
@@
 
-As the next step, an online clustering algorithm is applied to the articles in order to identify groups of articles that are discussing the same event.
+As the next step, an online clustering algorithm~\cite{brank2014} is applied to the articles in order to identify groups of articles that are discussing the same event.
 For each new article, the clustering algorithm determines if the article should be assigned to some existing cluster or into a new cluster.
@@
 This housekeeping mechanism prevents the clustering from becoming slow and also ensures that articles are not assigned to obsolete clusters.
-Details of the clustering approach are described in~\cite{brank2014}.
 
@@
 When a cluster is identified and information about the event is extracted, all available data is stored in a custom-built database system.
-The data is then accessible through the API or a web interface\footnote{http://eventregistry.org/}, which provide numerous search and visualization options.
+The data is then accessible through the API or a web interface (\url{http://eventregistry.org/}), which provide numerous search and visualization options.
 
@@
 
-\section{Related work}\label{sec:related}
+\section{Related Work}\label{sec:related}
 
@@
 
-\subsection{Cross-lingual document similarity}
+\subsection{Cross-Lingual Document Similarity}
 
@@
 
-\textbf{Translation and dictionary based}.
-The most obvious way to compare documents written in different languages is to use machine translation and perform monolingual similarity, see  ~\cite{multilingualBook,plagiarism} for several variations of translation based approaches.
-One can use free tools such as Moses~\shortcite{moses} or translation services, such as Google Translate\footnote{https://translate.google.com/}.
+\subsubsection{Translation and Dictionary Based Approaches} The most obvious way to compare documents written in different languages is to use machine
+translation and perform monolingual similarity \cite<see>{multilingualBook,plagiarism}.
+One can use free tools such as Moses~\shortcite{moses} or translation services, such as Google Translate (\mbox{\url{https://translate.google.com/}}).
 There are two issues with such approaches: they solve a harder problem than needs to be solved and they are less robust to training resource quality - large sets of translated sentences are typically needed.
@@
 The operation efficiency and cost requirements make translation-based approaches less suited for our system.
-Closely related are works Cross-Lingual Vector Space Model (CL-VSM)~\cite{plagiarism} and the approach presented in~\cite{pouliquen2008story} which both compare documents by using dictionaries, which in both cases are EuroVoc dictionaries~\cite{eurovoc}.
+Closely related are works Cross-Lingual Vector Space Model (CL-VSM)~\cite{plagiarism} and the approach presented by~\citeauthor{pouliquen2008story}~\citeyear{pouliquen2008story} which both compare documents by using dictionaries, which in both cases are EuroVoc dictionaries~\cite{eurovoc}.
 The generality of such approaches is limited by the quality of available linguistic resources, which may be scarce or non-existent for certain language pairs.
 
-\textbf{Probabilistic topic models}.
-There exist many variants to modelling documents in a language independent way by using probabilistic graphical models.
+\subsubsection{Probabilistic Topic Model Based Approaches} There exist many variants to modelling documents in a language independent way by using probabilistic graphical models.
 The models include:  Joint Probabilistic Latent Semantic Analysis (JPLSA)~\shortcite{platt2010translingual}, Coupled Probabilistic LSA (CPLSA)~\cite{platt2010translingual}, Probabilistic Cross-Lingual LSA (PCLLSA)~\cite{PCL_LSA} and Polylingual Topic Models (PLTM)~\shortcite{polyLDA} which is a Bayesian version of PCLLSA.
@@
 
-\textbf{Matrix factorization}.
-Several matrix factorization based approaches exist in the literature.
+\subsubsection{Matrix Factorization Base Approaches} Several matrix factorization based approaches exist in the literature.
 The models include: Non-negative matrix factorization based~\cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CL-LSI~\shortcite{cl_lsi,multilingualBook}, Canonical Correlation Analysis (CCA)~\cite{Hotelling}, Oriented Principal Component Analysis (OPCA)~\cite{platt2010translingual}.
@@
 In addition, OPCA forces the vocabulary sizes for all languages to be the same, which is less intuitive.
-For our setting, the method in ~\cite{nonnegfactor_lsi} has a prohibitively high computational cost when building models (it uses dense matrices whose dimensions are a product of the training set size and the vocabulary size).
+For our setting, the method in~\cite{nonnegfactor_lsi} has a prohibitively high computational cost when building models (it uses dense matrices whose dimensions are a product of the training set size and the vocabulary size).
 Our proposed approach combines CCA and CL-LSI.
@@
 The authors of CL-ESA compare it to CL-LSI and find that CL-LSI can outperform CL-ESA in an information retrieval, but is costlier to optimize over a large corpus (CL-ESA requires no training).
-We find that the scalability argument does not apply in our case: based on advances in numerical linear algebra we can solve large CL-LSI problems (millions of documents as opposed to the 10,000 document limit reported in~\cite{ESA}).
+We find that the scalability argument does not apply in our case: based on advances in numerical linear algebra we can solve large CL-LSI problems that involve millions of documents as opposed to the 10,000 document limit reported by~\citeauthor{ESA}~\citeyear{ESA}.
 In addition, CL-ESA is less suited for computing similarities between two large monolingual streams.
 For example, each day we have to compute similarities between 500,000 English and 500,000 German news articles.
-Comparing each German news article with 500,000 English news articles is either prohibitively slow (involves projecting all English articles on Wikipedia) or consumes too much memory (involves storing the projected English articles, which for a Wikipedia of size 1,000,000 is a 500,000 by 1000,0000 non-sparse matrix).
-
-\textbf{Monolingual}.
-Finally, related work includes monolingual approaches that treat document written in different languages in a monolingual fashion.
+Comparing each German news article with 500,000 English news articles is either prohibitively slow (involves projecting all English articles on Wikipedia) or consumes too much memory (involves storing the projected English articles, which for a Wikipedia of size 1,000,000 is a 500,000 by 1,000,000 non-sparse matrix).
+
+\subsubsection{Monolingual Approaches} Finally, related work includes monolingual approaches that treat document written in different languages in a monolingual fashion.
 The intuition is that named entities (for example, ``Obama'') and cognate words (for example, ``tsunami'') are written in the same or similar fashion in many languages.
 For example, the Cross-Language Character n-Gram Model (CL-CNG)~\cite{plagiarism} represents documents as bags of character $n$-grams.
-Another approach is to use language dependent keyword lists based on cognate words ~\cite{pouliquen2008story}.
+Another approach is to use language dependent keyword lists based on cognate words~\cite{pouliquen2008story}.
 These approaches may be suitable for comparing documents written in languages that share a writing system, which does not apply to the case of global news tracking.
@@
 
-\subsection{Cross-lingual cluster linking}
+\subsection{Cross-Lingual Cluster Linking}
 
@@
 
+\begin{sloppypar}
 A system, which is significantly different but worth mentioning, is the GDELT project~\cite{Leetaru2013Gdelt}.
@@
 All information extraction is then done on the translated article.
-
-\section{Cross-lingual Document Similarity}
+\end{sloppypar}
+
+\section{Cross-Lingual Document Similarity}
 \label{sec:crosslingual}
@@
 Many techniques use the similarity as a black box, e.g., a kernel in Support Vector Machines.
-Comparison of documents (or other types of text snippets) in a monolingual setting is a well-studied problem in the field of information retrieval ~\cite{Salton88term-weightingapproaches}.
+Comparison of documents (or other types of text snippets) in a monolingual setting is a well-studied problem in the field of information retrieval~\cite{Salton88term-weightingapproaches}.
 We first formally introduce the problem followed by a description of  our approach.
 
-\subsection{Problem definition}\label{sec:tfidf}
+\subsection{Problem Definition}\label{sec:tfidf}
 We will first describe how documents are represented as vectors and how to compare documents in a mono-lingual setting.
@@
 
-\noindent\textbf{Document representation.}
+\subsubsection{Document Representation}
 The standard vector space model~\cite{Salton88term-weightingapproaches} represents documents as vectors, where each term corresponds to a word or a phrase in a fixed vocabulary.
@@
 
-\noindent\textbf {Mono-lingual similarity.}
+\subsubsection{Mono-Lingual Similarity}
 A common way of computing similarity between documents is \emph{cosine similarity},
@@
 
-\noindent\textbf {Cross-lingual similarity.}
+\subsubsection{Cross-Lingual Similarity}
 Processing a multilingual dataset results in several vector spaces with varying dimensionality, one for each language.
@@
 CCA can be used to find correlated patterns for a pair of languages, whereas the extended method optimizes a
-Sum of Squared Correlations (SSCOR) between several language pairs, which was introduced in~\cite{Kettenring}.
+Sum of Squared Correlations (SSCOR) between several language pairs, which was introduced by~\citeauthor{Kettenring}~\citeyear{Kettenring}.
 The SSCOR problem is difficult to solve in our setting (hundreds of thousands of features, hundreds of thousands of examples).
@@
 We will now describe four models to cross-lingual similarity computation in the next sub-sections.
-\subsection{$k$-means}\label{sec:kmeans}
+\subsection{$k$-Means}\label{sec:kmeans}
 
@@
 
-\paragraph{Implementation note}
+\subsubsection{Implementation Note}
 
@@
  We now present a statistical technique to analyze data from two sources, an extension of which will be presented in the next section.
-%
+\begin{sloppypar}
  Canonical Correlation Analysis (CCA)~\cite{Hotelling} is a dimensionality reduction technique similar to Principal Component Analysis (PCA)~\cite{Pearson1901On}, with the additional assumption that the data consists of feature vectors that arose from two sources (two views) that share some information.
-Examples include: bilingual document collection ~\shortcite{mrpqr} and collection of images and captions~\shortcite{Hardoon_usingimage}.
+Examples include: bilingual document collection~\shortcite{mrpqr} and collection of images and captions~\shortcite{Hardoon_usingimage}.
 Instead of looking for linear combinations of features that maximize the variance (PCA) we look for a linear combination of feature vectors from the first view and a linear combination for the second view, that are maximally correlated.
-
+\end{sloppypar}
 Interpreting the columns of $X_i$ as observation vectors sampled from an underlying distribution $\mathcal{X}_i \in \RR^{n_i}$, the idea is to find two weight vectors $w_i \in \RR^{n_i}$ and $w_j \in \RR^{n_j}$ so that the random variables $w_i^T \cdot \mathcal{X}_i$ and $w_j^T \cdot \mathcal{X}_j$ are maximally correlated ($w_i$ and $w_j$ are used to map the random vectors to random variables, by computing weighted sums of vector components).
@@
 
-\subsection{Hub language based CCA Extension}\label{sec:hublang}
+\subsection{Hub Language Based CCA Extension}\label{sec:hublang}
 Building cross-lingual similarity models based on comparable corpora is challenging for two main reasons.
@@
 We now describe an extension of CCA to more than two languages, which can be trained on large comparable corpora and can handle missing data.
- The extension we consider is based on a generalization of CCA to more than two views, introduced in~\cite{Kettenring}, namely the Sum of Squared Correlations SSCOR, which we will state formally later in this section.
+ The extension we consider is based on a generalization of CCA to more than two views, introduced by~\citeauthor{Kettenring}~\citeyear{Kettenring}, namely the Sum of Squared Correlations SSCOR, which we will state formally later in this section.
 Our approach exploits a certain characteristic of the data, namely the \emph{hub language} characteristic (see below) in two
@@
 
-\paragraph{Hub language characteristic.}
+\subsubsection{Hub Language Characteristic}
 In the case of Wikipedia, we observed that even though the training resources are scarce for certain language pairs, there often exists indirect training data.
@@
 
-\paragraph{Step 1: Hub language based dimensionality reduction.}
+\subsubsection{Step 1: Hub Language Based Dimensionality Reduction}
 
@@
 
-\paragraph{Step 2: Simplifying and solving SSCOR.}
+\subsubsection{Step 2: Simplifying and Solving SSCOR.}
 The second step involves solving a generalized version of canonical correlation analysis on the matrices $Y_i$ in order to find the mappings $P_i$.
@@
 
-\paragraph{Remark.} The technique is related to  Generalization of Canonical Correlation Analysis (GCCA) by Carroll~\citeyear{Carroll}, where an unknown group configuration variable is defined and the objective is to maximize the sum of squared correlations between the group variable and the others.
+\subsubsection{Remark}
+The technique is related to  Generalization of Canonical Correlation Analysis (GCCA) by Carroll~\citeyear{Carroll}, where an unknown group configuration variable is defined and the objective is to maximize the sum of squared correlations between the group variable and the others.
 The problem can be reformulated as an eigenvalue problem.
 The difference lies in the fact that we set the unknown group configuration variable as the hub language, which simplifies the solution.
-The complexity of our method is $O(k^3)$, where $k$ is the reduced dimension from the LSI preprocessing step, whereas solving the GCCA method scales as $O(s^3)$, where $s$ is the number of samples (see~\cite{gifi}).
+The complexity of our method is $O(k^3)$, where $k$ is the reduced dimension from the LSI preprocessing step, whereas solving the GCCA method scales as $O(s^3)$, where $s$ is the number of samples~\cite<see>{gifi}.
 Another issue with GCCA is that it cannot be directly applied to the case of missing documents.
@@
 
-\section{Cross-lingual Event Linking}\label{sec:linking}
+\section{Cross-Lingual Event Linking}\label{sec:linking}
 
@@
 
-\subsection{Problem definition}
+\subsection{Problem Definition}
 
@@
 The publication time and date of the events is computed as the average publication time and date of all the articles and is provided by Event Registry.
-\item \texttt{sharedDates} is determined as the Jaccard similarity coefficient between sets of date mentiones extracted from articles.
+\item \texttt{sharedDates} is determined as the Jaccard similarity coefficient between sets of date mentions extracted from articles.
 We use extracted mentions of dates provided by Event Registry, which uses an extensive set of regular expressions to detect and normalize mentions of dates in different forms.
@@
 In the rare case that after symmetrization we have multiple links pointing from the document, we pick the first one that we encountered.
-This matrix enables us to build an alignment across all Wikipedia\footnote{https://www.wikipedia.org/ dumps available in 2013} languages.
+This matrix enables us to build an alignment across all Wikipedia\footnote{The dataset is based on Wikipedia dumps available in 2013.} languages.
 
@@
 &	en&	es&	ru&	sl&	hi&	war&	ht&	pms\\\cline{1-9}
-en&	671~-~4.64&	463~-~4.29&	369~-~3.19&	50.3~-~2&	14.4~-~2.76&	8.58~-~2.41&	 17~-~2.32&	16.6~-~2.67\\
+en&	671~-~4.6&	463~-~4.3&	369~-~3.2&	50.3~-~2.0&	14.4~-~2.8&	8.58~-~2.4&	 17~-~2.3&	16.6~-~2.7\\
 \cline{2-9}
-es&	\multicolumn{1}{c|}{}	&	463~-~4.29&	187~-~2.94&	28.2~-~1.96&	8.72~-~2.48&	 6.88~-~2.4&	13.2~-~2&	 13.8~-~2.58\\
+es&	\multicolumn{1}{c|}{}	&	463~-~4.3&	187~-~2.9&	28.2~-~2.0&	8.7~-~2.5&	 6.9~-~2.4&	13.2~-~2&	 13.8~-~2.6\\
 \cline{3-9}
-ru&	\multicolumn{2}{c|}{}	&	369~-~3.19&	29.6~-~1.92&	9.16~-~2.68&	2.92~-~1.1&	 3.23~-~2.2&	10.2~-~1.29\\
+ru&	\multicolumn{2}{c|}{}	&	369~-~3.2&	29.6~-~1.9 &	9.2~-~2.7&	2.9~-~1.1&	 3.2~-~2.2&	10.2~-~1.3\\
 \cline{4-9}
-sl&	\multicolumn{3}{c|}{}	&	50.3~-~2&	3.83~-~1.65&	1.23~-~0.986&	0.949~-~1.23&	 1.85~-~0.988\\
+sl&	\multicolumn{3}{c|}{}	&	50.3~-~2&	3.8~-~1.6&	1.2~-~0.99&	0.95~-~1.2&	 1.8~-~1.0\\
 \cline{5-9}
-hi&	\multicolumn{4}{c|}{}	&	14.4~-~2.76&	0.579~-~0.76&	0.0~-~2.08&	0.0~-~0.796\\
+hi&	\multicolumn{4}{c|}{}	&	14.4~-~2.8&	0.58~-~0.8&	0.0~-~2.1&	0.0~-~0.8\\
 \cline{6-9}
-war&	\multicolumn{5}{c|}{}	&	8.58~-~2.41&	0.043~-~0.534&	0.0~-~1.97\\
+war&	\multicolumn{5}{c|}{}	&	8.6~-~2.4&	0.04~-~0.5&	0.0~-~2.0\\
 \cline{7-9}
-ht&	\multicolumn{6}{c|}{}	&	17~-~2.32&	0.0~-~0.355\\
+ht&	\multicolumn{6}{c|}{}	&	17~-~2.3&	0.0~-~0.4\\
 \cline{8-9}
-pms&	\multicolumn{7}{c|}{}	&	16.6~-~2.67\\
+pms&	\multicolumn{7}{c|}{}	&	16.6~-~2.7\\
 \cline{9-9}
@@
 
-\subsection{Evaluation Of Cross-Lingual Event Linking}
+\subsection{Evaluation of Cross-Lingual Event Linking}
 In order to determine how accurately we can predict cluster equivalence, we performed two experiments in a multilingual setting using English, German and Spanish languages for which we had labelled data to evaluate the linking performance.
@@
 
-\subsection{Remarks on the scalability of the implementation}
+\subsection{Remarks on the Scalability of the Implementation}
 
@@
 
-\subsection{Remarks on the reproducibility of experiments}
+\subsection{Remarks on the Reproducibility of Experiments}
 We have made  both the code and data that were used in the experiments publicly available at \url{https://github.com/rupnikj/jair_paper.git}.

### 2016-01-07 rupnik2015.tex
et. al. fix

@@
 By using a language detector, the stream is split into separate monolingual streams.
-Within each monolingual stream, an online clustering approach is employed, where tracked clusters correspond to our definition of events - this is based on the Event Registry system~\shortcite{Leban2014W,Leban2014I}.
+Within each monolingual stream, an online clustering approach is employed, where tracked clusters correspond to our definition of events - this is based on the Event Registry system~\cite{Leban2014W,Leban2014I}.
 Our main goal in this paper is to connect such clusters (representations of events) across languages, that is, to detect that a set of articles in language $A$ reports on the same event as a set of articles in language $B$.
@@
 
-The first ingredient of our approach to link clusters across languages represents a continuation of previous work~\shortcite{nips,sikdd,nips2,iti} where we explored representations of documents which were valid over multiple languages.
+The first ingredient of our approach to link clusters across languages represents a continuation of previous work~\cite{nips,sikdd,nips2,iti} where we explored representations of documents which were valid over multiple languages.
 The representations could be interpreted as multilingual topics, which were then used as proxies to compute cross-lingual similarities between documents.
@@
 
-We base our techniques of cross-lingual event linking on an online system for detection of world events, called Event Registry~\shortcite{Leban2014W,Leban2014I}.
+We base our techniques of cross-lingual event linking on an online system for detection of world events, called Event Registry~\cite{Leban2014W,Leban2014I}.
 Event Registry is a repository of events, where events are automatically identified by analyzing news articles that are collected from numerous news outlets all over the world.
@@
 \subsubsection{Probabilistic Topic Model Based Approaches} There exist many variants to modelling documents in a language independent way by using probabilistic graphical models.
-The models include:  Joint Probabilistic Latent Semantic Analysis (JPLSA)~\shortcite{platt2010translingual}, Coupled Probabilistic LSA (CPLSA)~\cite{platt2010translingual}, Probabilistic Cross-Lingual LSA (PCLLSA)~\cite{PCL_LSA} and Polylingual Topic Models (PLTM)~\shortcite{polyLDA} which is a Bayesian version of PCLLSA.
+The models include:  Joint Probabilistic Latent Semantic Analysis (JPLSA)~\cite{platt2010translingual}, Coupled Probabilistic LSA (CPLSA)~\cite{platt2010translingual}, Probabilistic Cross-Lingual LSA (PCLLSA)~\cite{PCL_LSA} and Polylingual Topic Models (PLTM)~\cite{polyLDA} which is a Bayesian version of PCLLSA.
 The methods (except for CPLSA) describe the multilingual document collections as samples from generative probabilistic models, with variations on the assumptions on the model structure.
@@
 \subsubsection{Matrix Factorization Base Approaches} Several matrix factorization based approaches exist in the literature.
-The models include: Non-negative matrix factorization based~\cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CL-LSI~\shortcite{cl_lsi,multilingualBook}, Canonical Correlation Analysis (CCA)~\cite{Hotelling}, Oriented Principal Component Analysis (OPCA)~\cite{platt2010translingual}.
+The models include: Non-negative matrix factorization based~\cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CL-LSI~\cite{cl_lsi,multilingualBook}, Canonical Correlation Analysis (CCA)~\cite{Hotelling}, Oriented Principal Component Analysis (OPCA)~\cite{platt2010translingual}.
 The quadratic time and space dependency of the OPCA method makes it impractical for large scale purposes.
@@
 The last vector contains the weighted list of Eurovoc subject domain descriptors.
-These descriptors are topics, such as \emph{air transport}, \emph{EC agreement}, \emph{competition} and \emph{pollution control} into which articles are automatically categorized~\shortcite{Pouliquen2006}.
+These descriptors are topics, such as \emph{air transport}, \emph{EC agreement}, \emph{competition} and \emph{pollution control} into which articles are automatically categorized~\cite{Pouliquen2006}.
 Similarity between clusters is then computed using a linear combination of the cosine similarities computed on the three vectors.
@@
 
-The second approach we consider is Cross-Lingual Latent Semantic Indexing (CL-LSI)~\cite{cl_lsi} which is a variant of LSI~\shortcite{lsi} for more than one language.
+The second approach we consider is Cross-Lingual Latent Semantic Indexing (CL-LSI)~\cite{cl_lsi} which is a variant of LSI~\cite{lsi} for more than one language.
 The approach is very similar to $k$-means, where we first concatenate the corpus matrices, compute a decomposition, which in case of CL-LSI is a truncated Singular Value Decomposition (SVD), decouple the
@@
 Moreover, the Lanczos method is hard to parallelize.
-Instead, we use a randomized version of the SVD~\shortcite{tropp} that can be viewed as a block Lanczos method.
+Instead, we use a randomized version of the SVD~\cite{tropp} that can be viewed as a block Lanczos method.
 That enables us to use parallelization and speeds up the computation considerably.
@@
  Canonical Correlation Analysis (CCA)~\cite{Hotelling} is a dimensionality reduction technique similar to Principal Component Analysis (PCA)~\cite{Pearson1901On}, with the additional assumption that the data consists of feature vectors that arose from two sources (two views) that share some information.
-Examples include: bilingual document collection~\shortcite{mrpqr} and collection of images and captions~\shortcite{Hardoon_usingimage}.
+Examples include: bilingual document collection~\cite{mrpqr} and collection of images and captions~\cite{Hardoon_usingimage}.
 Instead of looking for linear combinations of features that maximize the variance (PCA) we look for a linear combination of feature vectors from the first view and a linear combination for the second view, that are maximally correlated.
@@
 For each candidate cluster $c_j \in C$, we compute a vector of learning features that should be indicative of whether $c_i$ and $c_j$ are equivalent or not and apply a binary classification model that predicts if the clusters are equivalent or not.
-The classification algorithm that we used to train a model was a linear Support Vector Machine (SVM) method~\shortcite{shawe-taylor04kernel}.
+The classification algorithm that we used to train a model was a linear Support Vector Machine (SVM) method~\cite{shawe-taylor04kernel}.
 
@@
 We also compared the proposed approaches on the task of Wikipedia mate retrieval (the same task as in Section~\ref{experiments:hubcca}).
-We computed the Average (over language pairs) Mean Reciprocal Rank (AMRR)~\shortcite{voorhees1999trec}  performance of the different approaches on the  Wikipedia data by holding out $15,000$ aligned test documents and using $300,000$ aligned documents as the training set.
+We computed the Average (over language pairs) Mean Reciprocal Rank (AMRR)~\cite{voorhees1999trec}  performance of the different approaches on the  Wikipedia data by holding out $15,000$ aligned test documents and using $300,000$ aligned documents as the training set.
 Figure~\ref{pic:AMRR} shows AMRR score as the function of the number of feature vectors.

### 2016-01-08 rupnik2015.tex
fixes after proofreading

@@
 
-The term event is vague and ambiguous, but for the practical purposes, we define it as ``any significant happening that is being reported about in the media.'' Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015 (Figure~\ref{fig:event2}).
+The term event is vague and ambiguous, but for the practical purposes, we define it as ``any significant happening that is being reported about in the media.'' Examples of events would include shooting down of the Malaysia Airlines plane over Ukraine on July 18th, 2014 (see Figure~\ref{fig:event2}) and HSBC's admittance of aiding their clients in tax evasion on February 9th, 2015.
 Events such as these are covered by many articles and the question is how to find all the articles in different languages that are describing a single event.
@@
 
-Based on the similarity function, we propose a novel algorithm for linking events/clusters across languages.
-The approach is based on learning a classification model from labelled data based on several sets of features.
+Using the similarity function, we propose a novel algorithm for linking events/clusters across languages.
+We pose the task as a classification problem based on several sets of features.
 In addition to these features, cross-lingual similarity is also used to quickly identify a small list of potential linking candidates for each cluster.
@@
 The pipeline must be able to process millions of articles per day and perform billions of similarity computations each day.
-Both steps rely heavily on similarity computation, which implies that this must be highly scalable.
+Both steps rely heavily on similarity computation, which should therefore be highly scalable.
 
@@
 The common weighting scheme is called \emph{Term Frequency Inverse Document Frequency} ($TFIDF$) weighting.
-An \emph{Inverse Document Frequency} ($IDF$) weight for the dictionary term $k$ is defined as $\log\left( \frac{N}{DF_k} \right)$, where $DF_k$ is the number of documents in the corpus which contain term $k$.
+An \emph{Inverse Document Frequency} ($IDF$) weight for the dictionary term $k$ is defined as $\log\left( \frac{N}{DF_k} \right)$, where $DF_k$ is the number of documents in the corpus which contain term $k$ and $N$ is the total number of documents in the corpus.
 When building cross-lingual models, the IDF scores were computed with respect to the Wikipedia corpus.
@@
 \begin{equation}\label{dLdx1}
- \frac{\partial}{\partial x_1} L = 0 \Rightarrow \sum_{i = 2}^m  \left(y_1^T G_{i} y_i \right) G_i y_i + \lambda_1 y_1 = 0,
+ \frac{\partial}{\partial y_1} L = 0 \Rightarrow \sum_{i = 2}^m  \left(y_1^T G_{i} y_i \right) G_i y_i + \lambda_1 y_1 = 0,
 \end{equation}
 \begin{equation}\label{dLdxi}
-\frac{\partial}{\partial x_i} L = 0 \Rightarrow \left(y_1^T G_{i} y_i \right) G_{i}^T y_1 + \lambda_i y_i = 0,~i > 1.
+\frac{\partial}{\partial y_i} L = 0 \Rightarrow \left(y_1^T G_{i} y_i \right) G_{i}^T y_1 + \lambda_i y_i = 0,~i > 1.
 \end{equation}
@@
 In this paper we consider the problem of matching events from different languages.
-We do not address the problem of detection of events and instead base our evaluation on an online system for detection of world events, Event Registry.
+We do not address the problem of detection of events but instead base our evaluation on an online system for detection of world events, Event Registry.
 The events are represented by clusters of articles and so ultimately our problem reduces to finding suitable matchings between clusters with articles in different languages.
@@
 
-The problem of cross-lingual event linking is to match monolingual clusters of news articles that describing the same event across languages.
+The problem of cross-lingual event linking is to match monolingual clusters of news articles that describe the same event across languages.
 For example, we want to match a cluster of Spanish news articles and a cluster of English news articles that both describe the same earthquake.
@@
 
-We are also seeking for an algorithm which does not do exhaustive comparison of all clusters, since that can become prohibitively expensive when working in a real-time setting.
+We also seek an algorithm which does not do exhaustive comparison of all clusters, since that can become prohibitively expensive when working in a real-time setting.
 More specifically, we wish to avoid testing cluster $c_i$ with all the clusters from all the other languages.
@@
 establishes that the hub based approach can deal with language pairs where little or no training data is available. The second set of experiments compares the main approaches
-that we presented on the task of mate retrieval and the task of event linking. Finally, we examine how different choices of features impact the event linking performance.
+that we presented on the task of mate retrieval and the task of event linking.
+In the mate retrieval task we are given a test set of document pairs, where each pair consists of a document and its translation.
+Given a query document from the test set, the goal is to retrieve its translation in the other language, which is also referred to as its \emph{mate} document.
+Finally, we examine how different choices of features impact the event linking performance.
 
@@
 Wikipedia is a large source of multilingual data that is especially important for the languages for which no translation tools, multilingual dictionaries as Eurovoc~\cite{eurovoc}, or strongly aligned multilingual corpora as Europarl~\cite{europarl} are available.
-Documents in different languages are related with so called 'inter-language' links that can be found on the left of the Wikipedia page.
+Documents in different languages are related with inter-language links that can be found on the left of the Wikipedia page.
 The Wikipedia is constantly growing.
@@
 We created a manually labelled dataset in order to evaluate cross-lingual event linking using two human annotators.
-The annotators were provided with an interface listing the articles, their content from and top concepts for a pair of clusters and their task was to determine if the clusters were equivalent or not (i.e., discuss same event).
+The annotators were provided with an interface listing the articles, their content and top concepts for a pair of clusters.
+Their task was to determine if the clusters were equivalent or not (i.e., discuss same event).
 To obtain a pair of clusters $(c_i, c_j)$ to annotate, we first randomly chose a cluster $c_i$, used Algorithm~\ref{cluster_merge_algo1} to compute a set of potentially equivalent clusters $C$ and randomly chose a cluster $c_j \in C$.
@@
 We also analyzed how cluster size influences the accuracy of cluster linking.
-We would expect that if the tested pair of clusters has a larger number of articles then the classifier should be able to more accurately predict whether the clusters should be linked or not.
+Our intuition was that linking large cluster pairs is easier than linking clusters with few articles.
 The reasoning is that the large clusters would provide more document linking information (more articles mean more links to other similar articles) as well as more accurately aggregated semantic information.
@@
 We have made  both the code and data that were used in the experiments publicly available at \url{https://github.com/rupnikj/jair_paper.git}.
-The manually labelled dataset used in the evaluation of event linking is available at in the ``dataset`` subfolder of the github repository.
+The manually labelled dataset used in the evaluation of event linking is available at in the ``dataset'' subfolder of the github repository.
 The included archive contains two folders: ``positive'' and ``negative'', where the first folder includes examples of cluster pairs in two languages that represent the same event and the second folder contains pairs of clusters in two languages that do not represent different events.

### 2016-01-26 rupnik2015.tex
final fixes

@@
 
-\jairheading{55}{2016}{x-xx}{3/24}{1/16}
+\jairheading{55}{2016}{283-xx}{03/24}{01/16}
 \ShortHeadings{Cross-Lingual Document Similarity and Event Tracking}
@@
 In addition, OPCA forces the vocabulary sizes for all languages to be the same, which is less intuitive.
-For our setting, the method in~\cite{nonnegfactor_lsi} has a prohibitively high computational cost when building models (it uses dense matrices whose dimensions are a product of the training set size and the vocabulary size).
+For our setting, the method by ~\citeauthor{nonnegfactor_lsi}~\citeyear{nonnegfactor_lsi} has a prohibitively high computational cost when building models (it uses dense matrices whose dimensions are a product of the training set size and the vocabulary size).
 Our proposed approach combines CCA and CL-LSI.
@@
 The second step involves solving a generalized version of canonical correlation analysis on the matrices $Y_i$ in order to find the mappings $P_i$.
-The approach is based on the sum of squares of correlations formulation by Kettenring~\cite{Kettenring}, where we consider only correlations between pairs $(Y_1, Y_i), i >1$ due to the hub language problem characteristic.
+The approach is based on the sum of squares of correlations formulation by Kettenring~\citeyear{Kettenring}, where we consider only correlations between pairs $(Y_1, Y_i), i >1$ due to the hub language problem characteristic.
 We will present the original unconstrained optimization problem, then a constrained formulation based on the hub language problem characteristic.
